### 2024-02-04 ms/simulations.qmd
added random effect of id to simulations and model

---
title: "Simulating and fitting synthetic data"
format: pdf
editor: visual
---

## Purpose

We are simulating and fitting synthetic data sets to develop and validate the statistical model for later analysis. This will eventually go in the supporting information.

## Hyperparameters

Here we explaining hyperparameter choices.

`mu_intercept`: this is the value of $A$ where $\text{log}(g_text{sw}) = 0$ or $g_text{sw} = 1$, so this should be close to $A_\text{max}$, which typically varies between 10 and 50 $\mu~\text{mol}~\text{m}^{-2}~\text{s}^{-1}$. So let's say mean of 30 (for high light).

`sigma_intercept_rep`: variation in intercept among replicates within accession. Let's say 2?

, 20 / 3)

### 2024-02-04 r/000_update-data.R

@@ -4,7 +4,9 @@ source("r/header.R")
 # For now, data on UH Google Drive.
 # Change path to  "../../data/adaptive-amphistomy" after data collection is complete
 
-c("plant-info", "stomata") |>
+
+# rh_curves should come from filtered data eventually
+c("plant-info", "rh_curves", "stomata") |>
   walk(\(.x) {
     file.copy(
       glue(

### 2024-02-04 r/101_define-hyperparameters.R

@@ -11,7 +11,7 @@ aa_hyperpars = list(
   
   # Experimental design hyperparameters
   n_acc = 1, # number of accessions
-  n_rep = 1e1, # number of replicates per accession per treatment
+  n_id  = 1e1, # number of replicates per accession per treatment
   n_pts = 1e1, # number of points per curve
   
   # Chamber environment hyperparameters
@@ -35,9 +35,11 @@ aa_hyperpars = list(
   max_gsw_amphi = 0.50,
   min_gsw_pseudohypo = 0.05,
   max_gsw_pseudohypo = 0.25,
-  mu_intercept = 0, # later on, change to mu_intercept, sigma_intercept
-  mu_slope = 1,  # later on, change to mu_slope, sigma_slope
-  sigma_error_intercept = 1
+  mu_intercept = 30, # later on, change to mu_intercept, sigma_intercept
+  sigma_intercept_id = 2,
+  mu_slope = 9, 
+  sigma_slope_id = 1,
+  sigma_intercept_error = 1
 )
 
 # Calculate the decay, per s, of autocorrelation between data points given the

### 2024-02-04 r/102_define-parameters.R

@@ -2,73 +2,88 @@
 source("r/header.R")
 set.seed(20231227)
 
-
 aa_hyperpars = read_rds("objects/aa_hyperpars.rds")
 
 aa_hyperpars |>
   transpose() |>
   iwalk(\(.hpar, .i) {
     
-    rep_vector = with(.hpar, str_c("r", str_pad(
-      seq_len(n_rep), floor(log10(n_rep)) + 1, "left", "0"
-    )))
+    id_vector = LETTERS[seq_len(.hpar$n_id)]
     pts_vector = with(.hpar, str_c("p", str_pad(
       seq_len(n_pts), floor(log10(n_pts)) + 1, "left", "0"
     )))
     leaf_type_vector = c("amphi", "pseudohypo")
     
-    aa_pars = with(.hpar,
-                   crossing(nesting(
-                     crossing(rep = rep_vector,
-                              leaf_type = leaf_type_vector),
-                     error_intercept = rnorm(2 * n_rep, 0, sigma_error_intercept)
-                   ),
-                   pts = pts_vector) |>
-                     split( ~ rep + leaf_type) |>
-                     map_dfr(\(x) {
-                       n = nrow(x)
-                       x |>
-                         mutate(
-                           elapsed = seq(0, length.out = n, by = interval),
-                           R_c = list(make_autocorr_matrix(elapsed, b_autocorr_c)),
-                           R_w = list(make_autocorr_matrix(elapsed, b_autocorr_w)),
-                           b_autocorr_c = b_autocorr_c,
-                           b_autocorr_w = b_autocorr_w,
-                           c_a = c_a,
-                           flow = flow,
-                           g_bw = g_bw,
-                           P = P,
-                           RH = RH,
-                           s = s,
-                           T_air = T_air,
-                           T_leaf = T_leaf,
-                           sigma_c = sigma_c,
-                           sigma_w = sigma_w,
-                           error_CO2r = sigma_c * mvnfast::rmvn(1, rep(0, n), R_c[[1]])[1, ],
-                           error_CO2s = sigma_c * mvnfast::rmvn(1, rep(0, n), R_c[[1]])[1, ],
-                           error_H2Or = sigma_w * mvnfast::rmvn(1, rep(0, n), R_w[[1]])[1, ],
-                           error_H2Os = sigma_w * mvnfast::rmvn(1, rep(0, n), R_w[[1]])[1, ]
-                         )
[…]

### 2024-02-04 r/103_prepare-stan-data.R

@@ -10,7 +10,7 @@ list.files("synthetic-data", pattern = "df_sim[0-9]{4}.rds", full.names = TRUE) 
     
     stan_sim = list(
       n_pts = length(unique(df_sim$pts)),
-      n_rep = length(unique(df_sim$rep)),
+      n_id = length(unique(df_sim$id)),
       n_leaf_type = length(unique(df_sim$leaf_type))
     )
     
@@ -34,7 +34,7 @@ list.files("synthetic-data", pattern = "df_sim[0-9]{4}.rds", full.names = TRUE) 
         glue(
           "stan_sim${var} = array(df_sim${var}, dim = c({i}, {j}, {k}))",
           i = stan_sim$n_pts,
-          j = stan_sim$n_rep,
+          j = stan_sim$n_id,
           k = stan_sim$n_leaf_type
         )
       }) |>

### 2024-02-04 r/201_fit-synthetic.R

@@ -13,7 +13,7 @@ list.files("synthetic-data", pattern = "stan_sim[0-9]{4}.rds", full.names = TRUE
       chains = 1L,
       parallel_chains = 1L,
       init = list(
-        list(c_a = stan_sim$CO2_s, g_sw = stan_sim$g_sw)
+        list(c_a = stan_sim$CO2_s, log_gsw = log(stan_sim$g_sw))
       ),
       seed = 20240203
     )

### 2024-02-04 r/202_compare-CO2.R

@@ -8,6 +8,7 @@ n_fit = str_extract(fit, "[0-9]{4}")
 
 assert_set_equal(n_dat, n_fit)
 
+# c_0 and CO2_r ----
 n_dat |>
   map_dfr(\(.x) {
     df_sim = read_rds(glue("synthetic-data/df_sim{.x}.rds"))
@@ -15,11 +16,10 @@ n_dat |>
     
     par_string = "^c_0\\[([0-9]+),([0-9]+),([0-9]+)\\]$"
     
-    # c_0 and CO2_r ----
     df_c0 = full_join(
       # Simulated c_0
       df_sim |>
-        select(leaf_type, rep, pts, c_0, CO2r_sim = CO2_r),
+        select(leaf_type, id, pts, c_0, CO2r_sim = CO2_r),
       
       # Estimated c_0
       fit_sim$draws("c_0") |>
@@ -31,18 +31,15 @@ n_dat |>
             str_replace(name, par_string, "\\1") |>
               str_pad(2L, "left", "0")
           ),
-          rep = str_c(
-            "r",
-            str_replace(name, par_string, "\\2") |>
-              str_pad(2L, "left", "0")
-          ),
+          id = LETTERS[str_replace(name, par_string, "\\2") |>
+                         as.numeric()],
           lt = str_replace(name, par_string, "\\3"),
           leaf_type = case_when(lt == 1 ~  "amphi",
                                 lt == 2 ~ "pseudohypo")
         ) |>
         select(-name,-lt) |>
-        summarize(c0_est = median(c_0), .by = c(pts, rep, leaf_type)),
-      by = join_by(leaf_type, rep, pts)
+        summarize(c0_est = median(c_0), .by = c(pts, id, leaf_type)),
+      by = join_by(leaf_type, id, pts)
     )
     
     # Summarize fit

### 2024-02-04 r/203_compare-H2O.R

@@ -1,110 +1,155 @@
 # Compare simulated to estimated [H2O] values
 source("r/header.R")
 
-df_sim = read_rds("synthetic-data/df_sim.rds")
-fit_sim = read_rds("objects/fit_sim.rds")
+dat = list.files("synthetic-data", "df_sim[0-9]{4}.rds", full.names = TRUE)
+fit = list.files("objects", "fit_sim[0-9]{4}.rds", full.names = TRUE)
+n_dat = str_extract(dat, "[0-9]{4}")
+n_fit = str_extract(fit, "[0-9]{4}")
 
-# w_0 and H2O_r
-par_string = "^w_0\\[([0-9]+),([0-9]+),([0-9]+)\\]$"
+assert_set_equal(n_dat, n_fit)
 
-df_w0 = full_join(
-  # Simulated w_0
-  df_sim |>
-    select(leaf_type, rep, pts, w_0, H2Or_sim = H2O_r),
-  
-  # Estimated w_0
-  fit_sim$draws("w_0") |>
-    as_draws_df() |>
-    pivot_longer(starts_with("w_0"), values_to = "w_0") |>
-    mutate(
-      pts = str_c(
-        "p",
-        str_replace(name, par_string, "\\1") |>
-          str_pad(2L, "left", "0")
-      ),
-      rep = str_c(
-        "r",
-        str_replace(name, par_string, "\\2") |>
-          str_pad(2L, "left", "0")
-      ),
-      lt = str_replace(name, par_string, "\\3"),
-      leaf_type = case_when(lt == 1 ~  "amphi",
-                            lt == 2 ~ "pseudohypo")
-    ) |>
-    select(-name, -lt) |>
-    summarize(w0_est = median(w_0), .by = c(pts, rep, leaf_type)),
-  by = join_by(leaf_type, rep, pts)
-)
+# w_0 and H2O_r ----
+n_dat |>
+  map_dfr(\(.x) {
+    df_sim = read_rds(glue("synthetic-data/df_sim{.x}.rds"))
+    fit_sim = read_rds(glue("objects/fit_sim{.x}.rds"))
+    
+    par_string = "^w_0\\[([0-9]+),([0-9]+),([0-9]+)\\]$"
+    
+    df_w0 = full_join(
+      # Simulated w_0
+      df_sim |>
+        select(leaf_type, id, pts, w_0, H2Or_sim = H2O_r),
+      
+      # Estimated w_0
+      fit_sim$draws("w_0") |>
+        as_draws_df() |>
[…]

### 2024-02-04 r/204_compare-gsw.R

@@ -1,57 +1,83 @@
 # Compare simulated to estimated g_sw values
 source("r/header.R")
 
-df_sim = read_rds("synthetic-data/df_sim.rds")
-fit_sim = read_rds("objects/fit_sim.rds")
+dat = list.files("synthetic-data", "df_sim[0-9]{4}.rds", full.names = TRUE)
+fit = list.files("objects", "fit_sim[0-9]{4}.rds", full.names = TRUE)
+n_dat = str_extract(dat, "[0-9]{4}")
+n_fit = str_extract(fit, "[0-9]{4}")
 
-par_string = "^g_sw\\[([0-9]+),([0-9]+),([0-9]+)\\]$"
+assert_set_equal(n_dat, n_fit)
 
-df_gsw = full_join(
-  # Simulated g_sw
-  df_sim |>
-    select(leaf_type, rep, pts, g_sw, gsw_sim = gsw_hat),
-  
-  # Estimated g_sw
-  fit_sim$draws("g_sw") |>
-    as_draws_df() |>
-    pivot_longer(starts_with("g_sw"), values_to = "g_sw") |>
-    mutate(
-      pts = str_c(
-        "p",
-        str_replace(name, par_string, "\\1") |>
-          str_pad(2L, "left", "0")
-      ),
-      rep = str_c(
-        "r",
-        str_replace(name, par_string, "\\2") |>
-          str_pad(2L, "left", "0")
-      ),
-      lt = str_replace(name, par_string, "\\3"),
-      leaf_type = case_when(lt == 1 ~  "amphi",
-                            lt == 2 ~ "pseudohypo")
-    ) |>
-    select(-name, -lt) |>
-    summarize(gsw_est = median(g_sw), .by = c(pts, rep, leaf_type)),
-  by = join_by(leaf_type, rep, pts)
-)
+n_dat |>
+  map_dfr(\(.x) {
+    df_sim = read_rds(glue("synthetic-data/df_sim{.x}.rds"))
+    fit_sim = read_rds(glue("objects/fit_sim{.x}.rds"))
+    
+    par_string = "^g_sw\\[([0-9]+),([0-9]+),([0-9]+)\\]$"
+    
+    df_gsw = full_join(
+      # Simulated g_sw
+      df_sim |>
+        select(leaf_type, id, pts, g_sw, gsw_sim = gsw_hat),
+      
+      # Estimated g_sw
+      fit_sim$draws("g_sw") |>
+        as_draws_df() |>
[…]

### 2024-02-04 r/205_compare-A.R

@@ -1,56 +1,79 @@
 # Compare simulated to estimated A values
 source("r/header.R")
 
-df_sim = read_rds("synthetic-data/df_sim.rds")
-fit_sim = read_rds("objects/fit_sim.rds")
+dat = list.files("synthetic-data", "df_sim[0-9]{4}.rds", full.names = TRUE)
+fit = list.files("objects", "fit_sim[0-9]{4}.rds", full.names = TRUE)
+n_dat = str_extract(dat, "[0-9]{4}")
+n_fit = str_extract(fit, "[0-9]{4}")
 
-par_string = "^A\\[([0-9]+),([0-9]+),([0-9]+)\\]$"
-df_A = full_join(
-  # Simulated A
-  df_sim |>
-    select(leaf_type, rep, pts, A, A_sim = A_hat),
-  
-  # Estimated A
-  fit_sim$draws("A") |>
-    as_draws_df() |>
-    pivot_longer(starts_with("A"), values_to = "A") |>
-    mutate(
-      pts = str_c(
-        "p",
-        str_replace(name, par_string, "\\1") |>
-          str_pad(2L, "left", "0")
-      ),
-      rep = str_c(
-        "r",
-        str_replace(name, par_string, "\\2") |>
-          str_pad(2L, "left", "0")
-      ),
-      lt = str_replace(name, par_string, "\\3"),
-      leaf_type = case_when(lt == 1 ~  "amphi",
-                            lt == 2 ~ "pseudohypo")
-    ) |>
-    select(-name, -lt) |>
-    summarize(A_est = median(A), .by = c(pts, rep, leaf_type)),
-  by = join_by(leaf_type, rep, pts)
-)
+assert_set_equal(n_dat, n_fit)
 
-# True A versus simulated A
-ggplot(df_A, aes(A, A_sim, color = leaf_type)) +
-  facet_wrap(~ rep) +
-  geom_abline(slope = 1, intercept = 0) +
-  geom_point() +
-  coord_equal()
+n_dat |>
+  map_dfr(\(.x) {
+    df_sim = read_rds(glue("synthetic-data/df_sim{.x}.rds"))
+    fit_sim = read_rds(glue("objects/fit_sim{.x}.rds"))
+    
+    par_string = "^A\\[([0-9]+),([0-9]+),([0-9]+)\\]$"
+    df_A = full_join(
+      # Simulated A
+      df_sim |>
[…]

### 2024-02-04 r/207_compare-hyperpars.R

@@ -2,25 +2,59 @@
 source("r/header.R")
 
 aa_hyperpars = read_rds("objects/aa_hyperpars.rds")
-fit_sim = read_rds("objects/fit_sim.rds")
+n_sim = lapply(aa_hyperpars, length)
 
-pdf("figures/fit_sim_areaplot.pdf",
-    width = 4,
-    height = 4)
+assert_true(length(unique(n_sim)) == 1L)
+n_sim = n_sim[1]
 
-c(
+fit = list.files("objects", "fit_sim[0-9]{4}.rds", full.names = TRUE)
+n_fit = str_extract(fit, "[0-9]{4}")
+
+assert_true(length(n_fit) == n_sim)
+
+pars = c(
   "sigma_c",
   "sigma_w",
   "mu_intercept",
+  "sigma_intercept_id",
+  "sigma_intercept_error",
   "mu_slope",
-  "sigma_error_intercept",
+  "sigma_slope_id",
   "b_autocorr_c",
   "b_autocorr_w"
-) |>
-  map(\(.hpar) {
-    fit_sim$draws(.hpar) |>
-      mcmc_areas(prob = 0.95) +
-      geom_vline(xintercept = aa_hyperpars[[.hpar]])
-  })
+)
 
-dev.off()
+full_join(
+  # Estimated values
+  n_fit |>
+    map_dfr(\(.x) {
+      fit_sim = read_rds(glue("objects/fit_sim{.x}.rds"))
+      fit_sim$summary(pars) |>
+        mutate(sim = glue("sim{.x}"))
+    }),
+  
+  # Simulated values
+  aa_hyperpars |>
+    as_tibble() |>
+    select(all_of(pars)) |>
+    mutate(sim = glue("sim{n_fit}")) |>
+    pivot_longer(-sim, names_to = "variable", values_to = "simulated"),
+  
+  by = join_by(sim, variable)
+  
+) |>
+  write_rds("objects/fit_sim_summary_hyperpars.rds")
+
+# Area plots (NOT SURE IF IʻLL USE FOR SIMULATED DATA, BUT MIGHT BE USEFUL FOR LATER)
+# pdf("figures/fit_sim_areaplot.pdf",
+#     width = 4,
+#     height = 4)
+# 
+# pars |>
+#   map(\(.hpar) {
+#     fit_sim$draws(.hpar) |>
+#       mcmc_areas(prob = 0.95) +
+#       geom_vline(xintercept = aa_hyperpars[[.hpar]])
+#   })
+# 
+# dev.off()

### 2024-02-04 r/header.R

@@ -10,6 +10,7 @@ library(ggplot2)
 library(glue)
 library(lubridate)
 library(magrittr)
+library(mvnfast)
 library(purrr)
 library(readr)
 library(scales)

### 2024-02-26 ms/notes.qmd
profiling and improving performance of stan model

---
title: "Notes"
format: html
editor: visual
execute:
  enabled: true
---

Making some notes on methods that might be useful for writing later

## 2024-02-25

Actually, I did this profiling wrong, but I'm going to try model without autocorrelation anyway to compare.

I profiled the `solanum-aa` code and found that most (90.6%) of the time is spent on the `multi_normal` calculations.

Next, I profiled the same model, ignoring residual autocorrelation. The speed-up is substantial (80\%). The time in this model is pretty evenly divided among various calculations and likelihood.

Note that total runtime was 5871.4 seconds for 400 iterations

| name         | total_time |
|--------------|------------|
| CO2_r likelihood | 297.691 |
| CO2_s likelihood | 238.332 |
| H2O_r likelihood | 304.786 |
| H2O_s likelihood | 306.439 |
| LICOR calcs | 1047.940 |
| priors | 749.983 |  
| regression | 645.282 |

Given the efficiency of model without residual autocorrelatiln, I will try fitting actual data with uncorrelated residuals first. If the residuals are highly correlated, then I will try a combination of thinning or accounting for autocorrelation, perhaps finding a more efficient approach.

### 2024-02-26 r/101_define-hyperparameters.R

@@ -48,13 +48,13 @@ post = as_draws(fit_preliminary)[[1]]
 set.seed(20240202)
 
 # Calculate realistic quantities from data
-n_id = rh_curves |>
-  summarise(n_id = length(unique(acc_id)), .by = "acc") |>
-  pull(n_id) |>
-  mean() |>
-  round()
+n_id = 10 #rh_curves |>
+  # summarise(n_id = length(unique(acc_id)), .by = "acc") |>
+  # pull(n_id) |>
+  # mean() |>
+  # round()
 
-n_pts = 20 #rh_curves |>
+lambda = 10 #rh_curves |>
   # summarise(n_pts = n(),
   #           .by = c("acc", "acc_id", "light_treatment", "light_intensity")) |>
   # pull(n_pts) |>
@@ -93,7 +93,7 @@ aa_hyperpars = list(
   # Experimental design hyperparameters
   n_acc = 1, # number of accessions
   n_id  = n_id, # number of replicates per accession per treatment
-  n_pts = n_pts, # number of points per curve
+  lambda = lambda, # average number of points per curve
   
   # Chamber environment hyperparameters
   c_a = 415, # CO2 [umol / mol]

### 2024-02-26 r/102_define-parameters.R

@@ -4,6 +4,10 @@ set.seed(20231227)
 
 aa_hyperpars = read_rds("objects/aa_hyperpars.rds")
 
+# .hpar = aa_hyperpars |>
+#   transpose() |>
+#   extract2(1)
+
 aa_hyperpars |>
   transpose() |>
   iwalk(\(.hpar, .i) {
@@ -11,9 +15,6 @@ aa_hyperpars |>
     light_treatment_vector = c("low", "high")
     leaf_type_vector = c("amphi", "pseudohypo")
     light_intensity_vector = c("150", "2000")
-    pts_vector = with(.hpar, str_c("p", str_pad(
-      seq_len(n_pts), floor(log10(n_pts)) + 1, "left", "0"
-    )))
     
     aa_pars_base = tibble(
       id = id_vector,
@@ -23,8 +24,13 @@ aa_hyperpars |>
       )
     ) |>
       crossing(leaf_type = leaf_type_vector,
-               light_intensity = light_intensity_vector,
-               pts = pts_vector)
+               light_intensity = light_intensity_vector) %>%
+      split(~ seq_len(nrow(.))) |>
+      map_dfr(\(.x, lambda) {
+        n_pts = rpois(1, lambda)
+        .x |>
+          reframe(across(everything(), ~ rep(.x, n_pts)))
+      }, lambda = .hpar$lambda)
     
     log_gsw_ranges = .hpar[str_detect(names(.hpar), "^m(ax|in)_log_gsw_")] |>
       as_tibble() |>

### 2024-02-26 r/201_fit-synthetic.R

@@ -1,10 +1,12 @@
 # Fit Stan model to synthetic data
 source("r/header.R")
 
-solanum_aa = cmdstan_model("stan/solanum-aa2.stan", dir = "stan/bin")
+solanum_aa = cmdstan_model("stan/solanum-aa3.stan", dir = "stan/bin")
 
 library(furrr)
 plan(multisession(workers = 3))
+
+# .x = "synthetic-data/stan_sim0001.rds"
 
 list.files("synthetic-data", pattern = "stan_sim[0-9]{4}.rds", full.names = TRUE) |>
   future_walk(\(.x) {
@@ -23,9 +25,16 @@ list.files("synthetic-data", pattern = "stan_sim[0-9]{4}.rds", full.names = TRUE
         sd_A = rep(sd(stan_sim$A), stan_sim$n_comp),
         mean_A = rep(mean(stan_sim$A), stan_sim$n_comp)
       )),
-      seed = 20240203
+      seed = 20240203,
+      iter_warmup = 2e2,
+      iter_sampling = 2e2,
+      refresh = 1e0
+      
     )
     
     fit_sim$save_object(glue("objects/fit_sim{n}.rds"))
-    
+    # some checks I can remove later
+    # fit_sim$profiles()
+    # fit_sim$summary("w")$median -> tmp
+    # hist(tmp[80 + 1:80])
   })

### 2024-02-28 htc/htc.sh
fitting actual data on htc and analyzing results from new model on synthetic data

@@ -33,7 +33,7 @@ cd ..
 tar -czf cmdstan-2.34.1.tar.gz cmdstan-2.34.1/
 
 
-
+# SYNTHETIC DATA
 
 # 1. Compress local directory before transfering
 tar --exclude='packages.tar.gz' -czf solanum-aa.tar.gz htc/
@@ -57,6 +57,30 @@ condor_submit solanum-aa/htc/fit_0001.sub # 19164174
 condor_submit solanum-aa/htc/fit_0002.sub # 19164175
 condor_submit solanum-aa/htc/fit_0003.sub # 19164176
 
-
 # check status
 condor_q 19164174
+
+# 8. Retrieve results
+scp cdmuir@submit2.chtc.wisc.edu:/home/cdmuir/fit_sim* objects/
+
+# clean up on submit node
+rm fit_*
+
+# ACTUAL DATA
+
+# 1. Transfer files before logging in using scp
+scp data/stan_rh_curves.rds htc/fit_dat.R htc/fit_dat.sh htc/fit_dat.sub stan/solanum-aa3.stan cdmuir@submit2.chtc.wisc.edu:/home/cdmuir/solanum-aa/htc
+
+# 2. Login to submit node
+
+# 3. Submit jobs
+condor_submit solanum-aa/htc/fit_dat.sub
+
+# check status
+condor_q 19228592
+
+# 4. Retrieve results
+scp cdmuir@submit2.chtc.wisc.edu:/home/cdmuir/fit_dat.rds objects/
+
+# clean up on submit node
+rm fit_*

### 2024-02-28 r/102_define-parameters.R

@@ -197,7 +197,7 @@ aa_hyperpars |>
         K = .hpar$K_amphi * 1 / (leaf_type == "amphi")
       )
     
-    
+    # Add LI-COR style estimates
     df_sim = df_sim |>
       li6800_simulate() |>
       # add autocorrelated error
@@ -208,6 +208,15 @@ aa_hyperpars |>
         H2O_s = w_a + error_H2Os
       ) |>
       li6800_estimate()
+    
+    # Add indexing information
+    df_sim = df_sim |>
+      # make unique ID for each leaf_type within id (change to acc_id later)
+      unite("leaftype_x_id", id, leaf_type, remove = FALSE) |>
+      # make unique ID for each curve
+      unite("curve", id, leaf_type, light_intensity, remove = FALSE) |>
+      mutate(curve_number = as.character(as.numeric(as.factor(curve))),
+             row = row_number())
     
     # Quick plot for checking
     # ggplot(df_sim, aes(gsw_hat, A_hat)) +

### 2024-02-28 r/103_prepare-stan-data.R

@@ -5,11 +5,7 @@ source("r/header.R")
 list.files("synthetic-data", pattern = "df_sim[0-9]{4}.rds", full.names = TRUE) |>
   walk(\(.x) {
     n = str_extract(.x, "[0-9]{4}")
-    df_sim = read_rds(.x) |>
-      # make unique ID for each leaf_type within id (change to acc_id later)
-      unite("leaftype_x_id", id, leaf_type, remove = FALSE) |>
-      # make unique ID for each curve
-      unite("curve", id, leaf_type, light_intensity, remove = FALSE)
+    df_sim = read_rds(.x)
     
     stan_sim = df_sim |>
       # select variables needed for Stan

### 2024-02-28 r/302_fit-data.R

@@ -1,0 +1,35 @@
+# Fit model to actual data
+source("r/header.R")
+
+solanum_aa = cmdstan_model("stan/solanum-aa3.stan", dir = "stan/bin")
+
+stan_rh_curves = read_rds("data/stan_rh_curves.rds")
+
+ad = c(stan_rh_curves$n, stan_rh_curves$n_comp)
+
+# started at 9:52 PM
+# 1% at 9:55 PM, 11.1 hours
+# 2% at 10:04 PM, 10 hours
+# 3% at 10:16 PM, 13.3 hours
+fit_dat = solanum_aa$sample(
+  data = stan_rh_curves,
+  chains = 2L,
+  parallel_chains = 2L,
+  init = list(list(
+    c_a = array(rep(stan_rh_curves$CO2_s, stan_rh_curves$n_comp), dim = ad),
+    log_gsw = array(rep(log(stan_rh_curves$g_sw), stan_rh_curves$n_comp), dim = ad),
+    sd_A = rep(sd(stan_rh_curves$A), stan_rh_curves$n_comp),
+    mean_A = rep(mean(stan_rh_curves$A), stan_rh_curves$n_comp)
+  )),
+  seed = 898932814,
+  iter_warmup = 1e3,
+  iter_sampling = 1e3,
+  refresh = 2e1
+  
+)
+
+fit_dat$save_object("objects/fit_dat.rds")
+    # some checks I can remove later
+    # fit_sim$profiles()
+    # fit_sim$summary("w")$median -> tmp
+    # hist(tmp[80 + 1:80])

### 2024-02-28 ms/notes.qmd
fitting actual data on htc and analyzing results from new model on synthetic data
updated H2O comparisons for synthetic data and fit

@@
 If the residuals are highly correlated, then I will try a combination of thinning or accounting for autocorrelation, perhaps finding a more efficient approach.
+
+## 2024-02-27
+
+Current model (no autocorrelation) is taking less than two hours to run on HTC, with average of 10 points per curve. 40 total curves.
+
+Current model (no autocorrelation) took ~8.33 hours to run LA2172 data on HTC, with average of 29 points per curve. 80 total curves.
+
+Actual data usage was 2,669,308 KB (33,554,432 requested)
+Actual memory usage was 2,580 MB (65,536 requested)

### 2024-02-28 r/203_compare-H2O.R

@@ -12,37 +12,64 @@ assert_set_equal(n_dat, n_fit)
 n_dat |>
   map_dfr(\(.x) {
     df_sim = read_rds(glue("synthetic-data/df_sim{.x}.rds"))
+    stan_sim = read_rds(glue("synthetic-data/stan_sim{.x}.rds"))
+    
+    # check order
+    assert_true(all(df_sim$curve_number == stan_sim$curve))
+    
     fit_sim = read_rds(glue("objects/fit_sim{.x}.rds"))
     
-    par_string = "^w_0\\[([0-9]+),([0-9]+),([0-9]+),([0-9]+)\\]$"
+    par_string = "^(w|w_0)\\[([0-9]+),([0-9]+)\\]$"
     
+    # Component weights per each curve
+    df_w = fit_sim$draws("w") |>
+      as_draws_df() |>
+      pivot_longer(matches(par_string), values_to = "w") |>
+      mutate(
+        curve_number = str_replace(name, par_string, "\\2"),
+        comp = str_replace(name, par_string, "\\3")
+      ) |>
+      select(-name) |>
+      full_join(
+        select(df_sim, row, curve_number),
+        by = join_by(curve_number),
+        relationship = "many-to-many"
+      )
+    
+    # w_0 estimates
+    df_w0_est = fit_sim$draws("w_0") |>
+      as_draws_df() |>
+      pivot_longer(matches(par_string), values_to = "w_0") |>
+      mutate(
+        row = as.integer(str_replace(name, par_string, "\\2")),
+        comp = str_replace(name, par_string, "\\3")) |>
+      select(-name) |>
+      full_join(select(df_sim, row, curve_number),
+                by = join_by(row),
+                relationship = "many-to-many")
+    
+    # Join weights and c_0 estimates
+    df_w_w0 = full_join(df_w,
+                        df_w0_est,
+                        join_by(.chain, .iteration, .draw, curve_number, comp, row)) |>
+      pivot_wider(
+        id_cols = c(starts_with("."), "row"),
+        names_from = "comp",
+        values_from = c("w", "w_0")
+      ) |>
+      mutate(w0_est = (w_1 * w_0_1 + w_2 * w_0_2) / (w_1 + w_2)) |>
+      group_by(row) |>
+      point_interval(w0_est)
+    
[…]

### 2024-03-01 ms/ms.qmd
outlining ms

---
title: "Untitled"
format: pdf
editor: visual
---

Stomata are important
Stomatal ratio is important unsolved problem
Amphi leaves are more common in sunny habitats and confer a great benefit. (this pattern is also true in Solanum? could we analyze herbarium data here?)

[maybe make distinction between amphi leaves and var in SR here since all tomato species are amphi]

Leaves will greater stomatal density ratio are more in open, sunny habitats because they deliver the greatest benefit in those circumstances. An amphistomatous leaf increases photosynthetic carbon gain compared to an otherwise identical hypostomatous leaf by increasing conductance through the leaf intercellular airspaces and boudary layers; the additional water loss through a second boundary layer is typically small [cites]. We quantify this benefit as the amphistomy advantage (AA = log(Aamphi/Ahypo). Why would AA be greater in sun than shade? There are three nonmutually exclusive hypotheses

[conceptual figure could also show differential benefit of amphi leaves in sun/shade]

to plants growing there.  in those
It has been hypothesized that amphistomy increases photosynthesis more in sunny places. Three nonmutually exclusive adaptive explanations for why amphi leaves:

Conceptual figure explaining each hypothesis

- acclimation: greater demand, higher gs increase AA
- developmental plasticity: light-induced changes in leaf anatomy modulate AA
- constitutive: genetic differences in leaves adapted to different light habitats

We distinguished among these hypotheses by comparing AA among wild tomato species from different native light habitats, grown under simulated sun and shade, and measured under contrasting light intenstiy (Figure of hypotheses and predictions).

[discussion]
- sun/shade has long been appreciated and this shows new trait that should be considered and that CO2 diffusion becomes major limitation

### 2024-03-01 htc/htc.sh

@@ -77,7 +77,7 @@ scp data/stan_rh_curves.rds htc/fit_dat.R htc/fit_dat.sh htc/fit_dat.sub stan/so
 condor_submit solanum-aa/htc/fit_dat.sub
 
 # check status
-condor_q 19228592
+condor_q 19234376 # max tree depth, iterations, and adapt delta increased
 
 # 4. Retrieve results
 scp cdmuir@submit2.chtc.wisc.edu:/home/cdmuir/fit_dat.rds objects/

### 2024-03-01 r/303_compare-parameters.R

@@ -1,0 +1,106 @@
+# Compare measured to estimated parameter values
+source("r/header.R")
+
+pars = c("A", "c_0", "g_sw", "w_0")
+
+fit_dat = read_rds("objects/fit_dat.rds")
+
+thinned_rh_curves = read_rds("data/thinned_rh_curves.rds") |>
+  # for now, filter to just LA2172
+  filter(str_detect(acc_id, "LA2172"))
+stan_rh_curves = read_rds("data/stan_rh_curves.rds")
+    
+assert_true(stan_rh_curves$n == nrow(thinned_rh_curves))
+# another assertion to make sure order is correct?
+
+par_string = glue("^({.s}|w)\\[([0-9]+),([0-9]+)\\]$", .s = str_c(pars, collapse = "|"))
+    
+# Component weights per each curve
+df_w = fit_dat$draws("w") |>
+  as_draws_df() |>
+  pivot_longer(matches(par_string), values_to = "w") |>
+  mutate(
+    curve_number = str_replace(name, par_string, "\\2"),
+    comp = str_replace(name, par_string, "\\3")
+  ) |>
+  select(-name)
+
+fit_dat$draws("lp__")     |> mcmc_trace()
+    # |>
+      full_join(
+        select(df_sim, row, curve_number),
+        by = join_by(curve_number),
+        relationship = "many-to-many"
+      )
+    
+    # parameter estimates
+    df_par_est = fit_sim$draws(pars) |>
+      as_draws_df() |>
+      pivot_longer(matches(par_string)) |>
+      mutate(
+        par = str_replace(name, par_string, "\\1"),
+        row = as.integer(str_replace(name, par_string, "\\2")),
+        comp = str_replace(name, par_string, "\\3")
+      ) |>
+      select(-name) |>
+      pivot_wider(names_from = "par") |>
+      full_join(select(df_sim, row, curve_number),
+                by = join_by(row),
+                relationship = "many-to-many")
+    
+    # Join weights and parameter estimates
+    df_w_par = full_join(df_w,
+                        df_par_est,
+                        join_by(.chain, .iteration, .draw, curve_number, comp, row)) |>
+      pivot_wider(
+        id_cols = c(starts_with("."), "row"),
+        names_from = "comp",
+        values_from = c("w", all_of(pars))
+      )
+    
+    .e = environment()
+    
+    pars |>
+      map(\(.p, n_comp) {
[…]

### 2024-03-04 ms/ms.qmd
outlining ms

@@
 ---
+
+In preamble:
+- short codes for commonly used symbols
 
@@
 We distinguished among these hypotheses by comparing AA among wild tomato species from different native light habitats, grown under simulated sun and shade, and measured under contrasting light intenstiy (Figure of hypotheses and predictions).
+We measured AA on XXX individual plants from XXX accessions (average of X replicates per light treatment) using a recently developed method [@triplett_amphistomy_2024].
+With this method, we directly compare the photosynthetic rate of an untreated amphistomatous leaf to that of the same leaf with gas exchange blocked through the adaxial (upper) surface by transparent plastic, which we refer to as 'pseudohypostomy'.
+To compare amphi- and pseudohypostomatous leaves at identical whole-leaf $g_\text{sc}$, we measure $A$ over a range of $g_\text{sc}$, inducing stomatal opening and closure by modulating humidity (see Materials and Methods for further details).
+
+[results]
+Amphistomy increases $A$ in all accessions, in both sun and shade leaves, and light intensities.
+We infer this from the fact that blocking gas exchange in pseudohypostomatous leaves reduced $A$ by X-X% depending on the accession, light treatment, and light intensity (Table/figure).
+The AA is equivalent to an X-X% change in total $g_\text{sc}$ (see SI section gs equivalency).
+But whereas increasing $g_\text{sc}$ would increase water loss as a necessary by-product, amphistomy can increase $A$ without any appreciable affect on transpiration.
+
+Sun leaves from high light habitats [not sure if htis result is true yet] benefit the most from amphistomy because of a both developmental plasticity and constitutive differences among accessions.
+[quantify difference in AA and contribution of different affects].
+Surprisingly, light intensity had a little effect...
+(should this be in this paragraph, or it's own?).
 

### 2024-03-04 ms/ms.qmd
outlining ms

--- 2024-03-01 ms/ms.qmd
@@
 
+Terminology to standardize:
+
+-   the native light habitat or environment
+
+-   growth condition versus measurement conditions
+
+-   ?
+
 Stomata are important
+
 Stomatal ratio is important unsolved problem
+
 Amphi leaves are more common in sunny habitats and confer a great benefit. (this pattern is also true in Solanum? could we analyze herbarium data here?)
 
-[maybe make distinction between amphi leaves and var in SR here since all tomato species are amphi]
+\[maybe make distinction between amphi leaves and var in SR here since all tomato species are amphi\]
 
-Leaves will greater stomatal density ratio are more in open, sunny habitats because they deliver the greatest benefit in those circumstances.
-An amphistomatous leaf increases photosynthetic carbon gain compared to an otherwise identical hypostomatous leaf by increasing conductance through the leaf intercellular airspaces and boudary layers; the additional water loss through a second boundary layer is typically small [cites].
+Leaves with greater stomatal density ratio are more in open, sunny habitats because they deliver the greatest benefit in those circumstances.
+An amphistomatous leaf increases photosynthetic carbon gain compared to an otherwise identical hypostomatous leaf by increasing conductance through the leaf intercellular airspaces and boudary layers; the additional water loss through a second boundary layer is typically small \[cites\].
 We quantify this benefit as the amphistomy advantage (AA = log(Aamphi/Ahypo).
 Why would AA be greater in sun than shade?
-There are three nonmutually exclusive hypotheses
+There are three nonmutually exclusive hypotheses that we classify as 'acclimatory', 'plastic', and 'constitutive'.
 
-[conceptual figure could also show differential benefit of amphi leaves in sun/shade]
+Acclimatory hypothesis: Leaves acclimated to high light intensity typically increase total leaf stomatal conductance (increased CO\$\_2\$ supply) and upregulate Rubisco activation <!--# I don't love the wording here.
+Is it just Rubisco activation, or are there are other things?
+--> (increased CO\$*2\$* demand).
+A one-dimensional circuit model using the Farquhar-von Caemmerer-Berry biochemical model of C\$\_3\$ photosynthesis shows that both increased stomatal conductance and Rubisco activity should increase AA, all else being equal (Supporting Information).
+If the acclimatory hypothesis is correct, we predict that AAhigh \> AAlow for all species regardless of native habitat or growth environment.
+Plants adapted to sunny, open habitats will evolve greater stomatal density ratio to take advantage of regular exposure to high light intensity.
 
-to plants growing there.  in those
-It has been hypothesized that amphistomy increases photosynthesis more in sunny places. Three nonmutually exclusive adaptive explanations for why amphi leaves:
+Plastic hypothesis: Individuals of the same genotype often develop dramatically different leaves in sun and shade conditions (cite).
+Plastic responses are likely adaptations to optimize photosynthesis at different light intensities (cites), but changes in leaf anatomy and biochemistry could modulate AA as a byproduct.
+Thicker or less porous leaves, both of which are associated with high leaf mass per area (LMA), will have lower g_ias; leaves with increased total stomatal density and Rubisco concentration have greater potential CO2 supply and demand.
+Under the plastic hypothesis, we predict that AAsun \> AAshade for all species and light intensities.
+AAsun and gsmax,sun should be positively associated with native light habitat.
+\[transition\] We assume that genotypes adapted to sunny, open habitats will express a phenotype best adapted to that environment when leaves develop under high light intensity; genotypes adapted to shaded closed habitats may be plastic, but suboptimal for light intensities they do not regularly experience in nature.
+
+\[conceptual figure could also show differential benefit of amphi leaves in sun/shade\]
+
+to plants growing there. in those It has been hypothesized that amphistomy increases photosynthesis more in sunny places. Three nonmutually exclusive adaptive explanations for why amphi leaves:
 
@@
 
-[discussion]
-- sun/shade has long been appreciated and this shows new trait that should be considered and that CO2 diffusion becomes major limitation
+\[discussion\] - sun/shade has long been appreciated and this shows new trait that should be considered and that CO2 diffusion becomes major limitation

### 2024-03-04 ms/ms.qmd

@@
 
+[introduction]
 Stomata are important

### 2024-03-04 ms/ms.qmd
working on ms

@@
 
+```{r load-objects, echo=FALSE, eval=TRUE}
+
+# I am mostly putting placeholders here for objects/values that will need to be
+# loaded eventually to use in paper.
+n_plant = 600
+n_acc = 30
+mean_n_rep = 10 # average number of individual replicates for accession per light treatment
+
+# environmental condition in growth room
+```
+
+## Notes
+
 In preamble:
 - short codes for commonly used symbols
+- ?
 
@@
 
--   the native light habitat or environment
+-   the native light habitat or environment: the SPLASH paper refers to what they calculated as 'habitat' PPFD, so I am going to use that terminology
 
@@
 -   ?
+
+## Main
 
@@
 
-We distinguished among these hypotheses by comparing AA among wild tomato species from different native light habitats, grown under simulated sun and shade, and measured under contrasting light intenstiy (Figure of hypotheses and predictions).
-We measured AA on XXX individual plants from XXX accessions (average of X replicates per light treatment) using a recently developed method [@triplett_amphistomy_2024].
+We distinguished among these hypotheses by comparing AA among wild tomato species from different native light habitats, grown under simulated sun and shade light treatments, and measured under contrasting light intensity (Figure of hypotheses and predictions).
+We measured AA on `r n_plant` individual plants from `r n_acc` accessions (average of `r mean_n_rep` replicates per light treatment) using a recently developed method [@triplett_amphistomy_2024].
 With this method, we directly compare the photosynthetic rate of an untreated amphistomatous leaf to that of the same leaf with gas exchange blocked through the adaxial (upper) surface by transparent plastic, which we refer to as 'pseudohypostomy'.
 To compare amphi- and pseudohypostomatous leaves at identical whole-leaf $g_\text{sc}$, we measure $A$ over a range of $g_\text{sc}$, inducing stomatal opening and closure by modulating humidity (see Materials and Methods for further details).
+
+Table of directional predictions (with table summarizing results? part of conceptual figure? in supplement?)
+
+caption: Directional predictions associated with each hypothesis explaining why amphistomy advantage (AA) might be greater for leaves in sunny, open habitats.
+For each hypothesis, we make predictions for how native light habitat, light treatment, and light intensity would affect AA.
+
+| hypothesis | native light habitat | light treatment | light intensity |
+|:----------:|:--------------------:|:---------------:|:---------------:|
+| acclimatory | cor(PAR, AA) = 0 | AAsun = AAshade | AA2000 > AA150 |
+| plastic | cor(PAR, AAsun) > 0 | AAsun > AAshade | AA2000 = AA150 |
+| constitutive | cor(PAR, AA) > 0 | AAsun = AAshade | AA2000 = AA150 |
+
+[add rows for when multiple hypotheses are supported simultaneously? or put that in SI?]
 
@@
 \[discussion\] - sun/shade has long been appreciated and this shows new trait that should be considered and that CO2 diffusion becomes major limitation
+
+## Materials and Methods
+
+[this will be moved to SI eventually]
+
+<!-- work until 10:15, then email Sarah about scheduling meeting -->
+
+### Accessions
+
+We compared AA among `r n_acc` ecologically diverse accessions of wild tomato, including representatives of all described species of *Solanum* sect.
+*Lycopersicon* and sect.
+*Lycopersicoides* [@peralta_taxonomy_2008] and the cultivated tomato *S.
+lycopersicum* var.
+*lycopersicum* (tab:accessions).
+Seeds provided by the Tomato Genetics Resource Center germinated on moist paper in plastic boxes after soaking for 30-60 minutes in a 50\% (volume per volume) solution of household bleach and water, followed by a thorough rinse.
+We transferred seedlings to cell-pack flats containing Pro-Mix BX potting mix (Premier Tech, Rivière-du-Loup, Quebec, Canada) once cotyledons fully emerged, typically within 1-2 weeks of sowing.
+After seedlings established in cell-pack flats for $\approx 2$ weeks, we transplanted them to 3.78 L plastic pots containing Pro-Mix BX potting mix, coral sand (Pro-Pak, Honolulu, Hawaiʻi, USA), and cinders (Niu Nursery, Honolulu, Hawaiʻi, USA) [look up exact proportions of each].
+
+outline of tab:accessions
+LA number,Species name, lat, lon, elev, SPLASH light value

### 2024-03-05 ms/ms.qmd
drafted a bunch of methods

@@
 ---
-title: "Untitled"
-format: pdf
-editor: visual
+title: Unknown
+format:
+  pdf:
+    keep-tex: true  
+    toc: false
+header-includes: |
+  \addtokomafont{disposition}{\rmfamily}
+  \usepackage{siunitx}
+mainfont: Times New Roman
+csl: https://www.zotero.org/styles/science
 ---
@@
 
-# environmental condition in growth room
+n_weeks = 80 # number of weeks for LICOR measurements
+start_licor_date = "May 1, 2022"
+end_licor_date = "October 31, 2023"
+
+# environmental condition in germination room
+germ_temp_day = 25
+germ_temp_night = 20
+germ_rh_day = 40
+germ_rh_night = 60
+germ_ppfd = 200
+
+ppfd_sun = 800
+ppfd_shade = 125
+temp_diff = 1 # degrees warmer that the sun treatment air was
+rh_diff = 5 # rh_sun - rh_shade
+leaf_temp_diff = 1 # difference in leaf temperature between treatments
+n_leaf_temp = 600 # number of leaves for which we have measured leaf temp under ambient conditions
+
 ```
@@
 
-In preamble:
-- short codes for commonly used symbols
-- ?
+In preamble: - short codes for commonly used symbols - ?
 
@@
 
-[introduction]
-Stomata are important
+\[introduction\] Stomata are important
 
@@
 |  hypothesis  | native light habitat | light treatment  | light intensity |
-|:----------:|:--------------------:|:---------------:|:---------------:|
-| acclimatory | cor(PAR, AA) = 0 | AAsun = AAshade | AA2000 > AA150 |
-| plastic | cor(PAR, AAsun) > 0 | AAsun > AAshade | AA2000 = AA150 |
-| constitutive | cor(PAR, AA) > 0 | AAsun = AAshade | AA2000 = AA150 |
+|:------------:|:--------------------:|:----------------:|:---------------:|
+| acclimatory  |   cor(PAR, AA) = 0   | AAsun = AAshade  | AA2000 \> AA150 |
+|   plastic    | cor(PAR, AAsun) \> 0 | AAsun \> AAshade | AA2000 = AA150  |
+| constitutive |  cor(PAR, AA) \> 0   | AAsun = AAshade  | AA2000 = AA150  |
 
-[add rows for when multiple hypotheses are supported simultaneously? or put that in SI?]
+\[add rows for when multiple hypotheses are supported simultaneously? or put that in SI?\]
 
-[results]
-Amphistomy increases $A$ in all accessions, in both sun and shade leaves, and light intensities.
+\[results\] Amphistomy increases $A$ in all accessions, in both sun and shade leaves, and light intensities.
 We infer this from the fact that blocking gas exchange in pseudohypostomatous leaves reduced $A$ by X-X% depending on the accession, light treatment, and light intensity (Table/figure).
@@
 
-Sun leaves from high light habitats [not sure if htis result is true yet] benefit the most from amphistomy because of a both developmental plasticity and constitutive differences among accessions.
-[quantify difference in AA and contribution of different affects].
+Sun leaves from high light habitats \[not sure if htis result is true yet\] benefit the most from amphistomy because of a both developmental plasticity and constitutive differences among accessions.
+\[quantify difference in AA and contribution of different affects\].
 Surprisingly, light intensity had a little effect...
@@
 
-[this will be moved to SI eventually]
-
-<!-- work until 10:15, then email Sarah about scheduling meeting -->
+\[this will be moved to SI eventually\]
 
@@
 *lycopersicum* (tab:accessions).
-Seeds provided by the Tomato Genetics Resource Center germinated on moist paper in plastic boxes after soaking for 30-60 minutes in a 50\% (volume per volume) solution of household bleach and water, followed by a thorough rinse.
+Due to constraints on growth space and time, we spread out measurements over `r n_weeks` weeks from `r start_licor_date` to `r end_licor_date`.
+Replicates within accession were evenly spread out over this period to prevent confounding of temporal variation in growth conditions with accession.
+[anything else to say here?
+maybe explain accession selection and phylogeny?]
+
+### Plant growth conditions
+
+In all growth spaces, we recorded $\mathrm{PPFD}$ using full spectrum quantum sensors (SQ-500-SS, Apogee Instruments, Logan, Utah, USA); we recorded temperature, RH, and [CO$_2$] using an EE894 sensor (E+E Elektronik, Engerwitzdorf, Austria) protected by a radiation shield.
+All environmental measurements were taken every 10 minutes from the middle of plants racks at approximately the same height as the leaves we measured.
+We measured leaf temperature of focal leaves prior to measurement using an infrared radiometer (SI-111-SS, Apogee Instruments, Logan, Utah, USA).
+
+#### Germination and seedling stage
+
+Seeds provided by the Tomato Genetics Resource Center germinated on moist paper in plastic boxes after soaking for 30-60 minutes in a 50% (volume per volume) solution of household bleach and water, followed by a thorough rinse.
 We transferred seedlings to cell-pack flats containing Pro-Mix BX potting mix (Premier Tech, Rivière-du-Loup, Quebec, Canada) once cotyledons fully emerged, typically within 1-2 weeks of sowing.
-After seedlings established in cell-pack flats for $\approx 2$ weeks, we transplanted them to 3.78 L plastic pots containing Pro-Mix BX potting mix, coral sand (Pro-Pak, Honolulu, Hawaiʻi, USA), and cinders (Niu Nursery, Honolulu, Hawaiʻi, USA) [look up exact proportions of each].
+We grew seeds and seedlings for both sun and shade treatments under the same environmental conditions (12:12 h, `r germ_temp_day`:$\qty{`r germ_temp_night`}{\degreeCelsius}$, `r germ_rh_day`:`r germ_rh_night` RH day:night cycle).
+LED light provided $\mathrm{PPFD} = \qty{`r germ_ppfd`}{\micro\mol\raiseto{-2}\meter\raiseto{-1}\second}$ (Fluence RAZRx, Austin, Texas, USA).
 
-outline of tab:accessions
-LA number,Species name, lat, lon, elev, SPLASH light value
+#### Light treatments
+
+Seedlings were randomly assigned in alternating order within accession to the sun or shade treatment during transplanting.
+After seedlings established in cell-pack flats for $\approx 2$ weeks, we transplanted them to 3.78 L plastic pots containing 60% Pro-Mix BX potting mix, 20% coral sand (Pro-Pak, Honolulu, Hawaiʻi, USA), and 20% cinders (Niu Nursery, Honolulu, Hawaiʻi, USA).
+Percentage composition is on a volume basis.
+The soil mixture contained slow release NPK fertilizer following manufacturer instructions (Osmocote Smart-Release Plant Food Flower & Vegetable, The Scotts Company, Marysville, Ohio, USA).
+We determined pot field capacity one week after transplanting using a scale (Ohaus V12P15 Valor 1000, Parsippany, New Jersey, USA) and watered to field capacity three times per week to prevent drought stress.
+
+We assigned sun and shade treatment to lower and upper racks of a $\qty{1.22}{\meter} \times \qty{2.44}{\meter}$ shelving unit in a climate-controlled growth room.
+We assigned the sun treatment to the lower rack to limit diffuse light from reaching the shade treatment.
+The average daytime $\mathrm{PPFD}$ was $\qty{`r ppfd_sun`}{\micro\mol\raiseto{-2}\meter\raiseto{-1}\second}$ and $\qty{`r ppfd_shade`}{\micro\mol\raiseto{-2}\meter\raiseto{-1}\second}$ for sun and shade treatments, respectively.
+To isolate the effect of light intensity from quality, we used the same LED model with the the same spectrum (Fluence SPYDR 2i, Austin, Texas, USAS), but dimmed the lights in the shade treatment.
+To maintain homogeneous environmental conditions other than light, we mixed air within the growth room using an air circulator (Vornado 693DC, Andover, Kansas, USA) and within racks using a miniature oscillating air circulator (Vornado Atom 1, Andover, Kansas, USA).
+Despite these efforts, the air in the sun treatment was on average $\qty{`r temp_diff`}{\degreeCelsius}$ warmer and the average RH was consequently `r rh_diff` lower.
+However, because of evaporative cooling, the leaves in the sun treatment were only $\qty{`r leaf_temp_diff`}{\degreeCelsius}$ on average ($n = `r n_leaf_temp`$ leaves).
+
+outline of tab:accessions LA number,Species name, lat, lon, elev, SPLASH light value

### 2024-03-05 ms/ms.tex
drafted a bunch of methods

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  letterpaper,
  DIV=11,
  numbers=noendperiod]{scrartcl}

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
  \setmainfont[]{Times New Roman}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother

\KOMAoption{captions}{tableheading}
\addtokomafont{disposition}{\rmfamily}
\usepackage{siunitx}
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Unknown},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{Unknown}
\author{}
\date{}

\begin{document}
\maketitle
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, borderline west={3pt}{0pt}{shadecolor}, breakable, boxrule=0pt, interior hidden, enhanced, sharp corners]}{\end{tcolorbox}}\fi

\hypertarget{notes}{%
\subsection{Notes}\label{notes}}

In preamble: - short codes for commonly used symbols - ?

Terminology to standardize:

\begin{itemize}
\item
  the native light habitat or environment: the SPLASH paper refers to
  what they calculated as `habitat' PPFD, so I am going to use that
  terminology
\item
  growth condition versus measurement conditions
\item
  ?
\end{itemize}

\hypertarget{main}{%
\subsection{Main}\label{main}}

{[}introduction{]} Stomata are important

Stomatal ratio is important unsolved problem

Amphi leaves are more common in sunny habitats and confer a great
benefit. (this pattern is also true in Solanum? could we analyze
herbarium data here?)

{[}maybe make distinction between amphi leaves and var in SR here since
all tomato species are amphi{]}

Leaves with greater stomatal density ratio are more in open, sunny
habitats because they deliver the greatest benefit in those
circumstances. An amphistomatous leaf increases photosynthetic carbon
gain compared to an otherwise identical hypostomatous leaf by increasing
conductance through the leaf intercellular airspaces and boudary layers;
the additional water loss through a second boundary layer is typically
small {[}cites{]}. We quantify this benefit as the amphistomy advantage
(AA = log(Aamphi/Ahypo). Why would AA be greater in sun than shade?
There are three nonmutually exclusive hypotheses that we classify as
`acclimatory', `plastic', and `constitutive'.

Acclimatory hypothesis: Leaves acclimated to high light intensity
typically increase total leaf stomatal conductance (increased CO\$\_2\$
supply) and upregulate Rubisco activation (increased CO\$\emph{2\$}
demand). A one-dimensional circuit model using the Farquhar-von
Caemmerer-Berry biochemical model of C\$\_3\$ photosynthesis shows that
both increased stomatal conductance and Rubisco activity should increase
AA, all else being equal (Supporting Information). If the acclimatory
hypothesis is correct, we predict that AAhigh \textgreater{} AAlow for
all species regardless of native habitat or growth environment. Plants
adapted to sunny, open habitats will evolve greater stomatal density
ratio to take advantage of regular exposure to high light intensity.

Plastic hypothesis: Individuals of the same genotype often develop
dramatically different leaves in sun and shade conditions (cite).
Plastic responses are likely adaptations to optimize photosynthesis at
different light intensities (cites), but changes in leaf anatomy and
biochemistry could modulate AA as a byproduct. Thicker or less porous
leaves, both of which are associated with high leaf mass per area (LMA),
will have lower g\_ias; leaves with increased total stomatal density and
Rubisco concentration have greater potential CO2 supply and demand.
Under the plastic hypothesis, we predict that AAsun \textgreater{}
AAshade for all species and light intensities. AAsun and gsmax,sun
should be positively associated with native light habitat.
{[}transition{]} We assume that genotypes adapted to sunny, open
habitats will express a phenotype best adapted to that environment when
leaves develop under high light intensity; genotypes adapted to shaded
closed habitats may be plastic, but suboptimal for light intensities
they do not regularly experience in nature.

{[}conceptual figure could also show differential benefit of amphi
leaves in sun/shade{]}

to plants growing there. in those It has been hypothesized that
amphistomy increases photosynthesis more in sunny places. Three
nonmutually exclusive adaptive explanations for why amphi leaves:

Conceptual figure explaining each hypothesis

\begin{itemize}
\tightlist
\item
  acclimation: greater demand, higher gs increase AA
\item
  developmental plasticity: light-induced changes in leaf anatomy
  modulate AA
\item
  constitutive: genetic differences in leaves adapted to different light
  habitats
\end{itemize}

We distinguished among these hypotheses by comparing AA among wild
tomato species from different native light habitats, grown under
simulated sun and shade light treatments, and measured under contrasting
light intensity (Figure of hypotheses and predictions). We measured AA
on 600 individual plants from 30 accessions (average of 10 replicates
per light treatment) using a recently developed method
{[}@triplett\_amphistomy\_2024{]}. With this method, we directly compare
the photosynthetic rate of an untreated amphistomatous leaf to that of
the same leaf with gas exchange blocked through the adaxial (upper)
surface by transparent plastic, which we refer to as `pseudohypostomy'.
To compare amphi- and pseudohypostomatous leaves at identical whole-leaf
\(g_\text{sc}\), we measure \(A\) over a range of \(g_\text{sc}\),
inducing stomatal opening and closure by modulating humidity (see
Materials and Methods for further details).

Table of directional predictions (with table summarizing results? part
of conceptual figure? in supplement?)

caption: Directional predictions associated with each hypothesis
explaining why amphistomy advantage (AA) might be greater for leaves in
sunny, open habitats. For each hypothesis, we make predictions for how
native light habitat, light treatment, and light intensity would affect
AA.

\begin{longtable}[]{@{}
  >{\centering\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.1972}}
  >{\centering\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.3099}}
  >{\centering\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.2535}}
  >{\centering\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.2394}}@{}}
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\centering
hypothesis
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
native light habitat
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
light treatment
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
light intensity
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
acclimatory & cor(PAR, AA) = 0 & AAsun = AAshade & AA2000 \textgreater{}
AA150 \\
plastic & cor(PAR, AAsun) \textgreater{} 0 & AAsun \textgreater{}
AAshade & AA2000 = AA150 \\
constitutive & cor(PAR, AA) \textgreater{} 0 & AAsun = AAshade & AA2000
= AA150 \\
\end{longtable}

{[}add rows for when multiple hypotheses are supported simultaneously?
or put that in SI?{]}

{[}results{]} Amphistomy increases \(A\) in all accessions, in both sun
and shade leaves, and light intensities. We infer this from the fact
that blocking gas exchange in pseudohypostomatous leaves reduced \(A\)
by X-X\% depending on the accession, light treatment, and light
intensity (Table/figure). The AA is equivalent to an X-X\% change in
total \(g_\text{sc}\) (see SI section gs equivalency). But whereas
increasing \(g_\text{sc}\) would increase water loss as a necessary
by-product, amphistomy can increase \(A\) without any appreciable affect
on transpiration.

Sun leaves from high light habitats {[}not sure if htis result is true
yet{]} benefit the most from amphistomy because of a both developmental
plasticity and constitutive differences among accessions. {[}quantify
difference in AA and contribution of different affects{]}. Surprisingly,
light intensity had a little effect\ldots{} (should this be in this
paragraph, or it's own?).

{[}discussion{]} - sun/shade has long been appreciated and this shows
new trait that should be considered and that CO2 diffusion becomes major
limitation

\hypertarget{materials-and-methods}{%
\subsection{Materials and Methods}\label{materials-and-methods}}

{[}this will be moved to SI eventually{]}

\hypertarget{accessions}{%
\subsubsection{Accessions}\label{accessions}}

We compared AA among 30 ecologically diverse accessions of wild tomato,
including representatives of all described species of \emph{Solanum}
sect. \emph{Lycopersicon} and sect. \emph{Lycopersicoides}
{[}@peralta\_taxonomy\_2008{]} and the cultivated tomato \emph{S.
lycopersicum} var. \emph{lycopersicum} (tab:accessions). Seeds provided
by the Tomato Genetics Resource Center germinated on moist paper in
plastic boxes after soaking for 30-60 minutes in a 50\% (volume per
volume) solution of household bleach and water, followed by a thorough
rinse. We transferred seedlings to cell-pack flats containing Pro-Mix BX
potting mix (Premier Tech, Rivière-du-Loup, Quebec, Canada) once
cotyledons fully emerged, typically within 1-2 weeks of sowing. We grew
seeds and seedlings for both sun and shade treatments under the same
environmental conditions (12:12 h, 25:\(\qty{20}{\degreeCelsius}\),
40:60 RH day:night cycle). LED light provided
\(\mathrm{PPFD} = \qty{200}{\micro\mol\raiseto{-2}\meter\raiseto{-1}\second}\).

After seedlings established in cell-pack flats for \(\approx 2\) weeks,
we transplanted them to 3.78 L plastic pots containing 60\% Pro-Mix BX
potting mix, 20\% coral sand (Pro-Pak, Honolulu, Hawaiʻi, USA), and 20\%
cinders (Niu Nursery, Honolulu, Hawaiʻi, USA). Percentage composition is
on a volume basis. The soil mixture contained slow release NPK
fertilizer following manufacturer instructions (Osmocote Smart-Release
Plant Food Flower \& Vegetable, The Scotts Company, Marysville, Ohio,
USA).

outline of tab:accessions LA number,Species name, lat, lon, elev, SPLASH
light value

\end{document}

### 2024-03-05 ms/test.qmd
drafted a bunch of methods

---
title: Unknown
format:
  pdf:
    keep-tex: true  
    toc: false
header-includes: |
  \addtokomafont{disposition}{\rmfamily}
  \usepackage{siunitx}
mainfont: Times New Roman
csl: https://www.zotero.org/styles/american-journal-of-botany
---

words $\qty{4}{\micro\meter}$

(12:12 $\qty{25}{\degreeCelsius}$

and then this $\qty{200}{\micro\mol\raiseto{-2}\meter\raiseto{-1}\second}$

$\qty{0.1}{\hertz}$

### 2024-03-06 ms/ms.qmd
working on materials and methods

@@
 
+percent_terminal = "70.0" # percent of LICOR measurements on terminal leaflet
+percent_lateral = "30.0"
+percent_change_leaflet = "10.0" # percent of time that anatomical traits were measured on different leaflet than LICOR measurements
 ```
@@
 
+```{r, accessions, eval=FALSE,echo=FALSE,message=FALSE,warning=FALSE}
+#| label: tbl-accessions
+#| tbl-cap: "Solanum accessions"
+
+# outline of tab:accessions LA number,Species name, lat, lon, elev, SPLASH light value
+
+# copied code from stomata-ilima to emulate
+# read_rds("../processed-data/site.rds") |>
+#   filter(site_code != "nnpl") |>
+#   mutate(Island = case_when(
+#     island == "oahu" ~ "Oʻahu",
+#     island == "hawaii" ~ "Hawaiʻi"
+#   )) |>
+#   arrange(site_type, site) |>
+#   select(Site = site, Island, Habitat = site_type, Latitude = latitude_degree,
+#          Longitude = longitude_degree, `Elevation (mas)` = elevation_m) |>
+#   knitr::kable(format = "latex", digits = c(rep(0, 3), 3, 3, 0),
+#                booktabs = TRUE) |>
+#   kableExtra::kable_styling(latex_options = c("striped", "no_vertical_lines")) |>
+#   kableExtra::column_spec(1, width = "4cm") |>
+#   kableExtra::column_spec(2, width = "1.5cm") |>
+#   kableExtra::column_spec(3, width = "1.5cm") |>
+#   kableExtra::column_spec(4, width = "1.5cm") |>
+#   kableExtra::column_spec(5, width = "1.5cm") |>
+#   kableExtra::column_spec(6, width = "1.5cm")
+
+```
+
 ### Plant growth conditions
@@
 
-outline of tab:accessions LA number,Species name, lat, lon, elev, SPLASH light value
+### Leaf trait measurements
+
+We selected a fully expanded, unshaded leaf at least six leaves above the cotyledons during early vegetative growth.
+This typically meant that plants had grown in light treatments for $\approx$ 4 weeks, ensuring they had time to sense and respond developmentally to the light intensity of the treatment rather than the seedling conditions [@schoch_dependence_1980].
+Shade plants grew slower than sun plants, hence leaves at the same developmental stage were measured on chronologically older plants in the shade treatment.
+In some sun plants, we had to use leaves higher on the stem because short internodes made lower leaves inaccessible with the gas exchange equipment.
+We measured terminal leaflets in `r percent_terminal`\% of cases, but used the lateral leaflet closest to the terminal leaflet when it was damaged or difficult to clamp into the gas exchange chamber.
+When a leaflet was damanged during gas exchange measurements, we collected anatomical data from the nearest leaflet on the same leaf (`r percent_change_leaflet`)\%.
+We measured chlorophyll concentration index (CCI) using a chrolophyll concentration meter (MC-100, Apogee Instruments, Logan, Utah, USA) on the lamina of focal leaflets before gas exchange measurements at the same time we measured leaf temperature.
+
+### Amphistomy advantage
+
+We estimated 'amphistomy advantage' ($\mathrm{AA}$) *sensu* @parkhurst_adaptive_1978, but with modifications previously described in @triplett_amphistomy_2024.
+$\mathrm{AA}$ is calculated as the log-response ratio of $A$ compared at the same total $g_\text{sw}$:
+
+$$\mathrm{AA} = \mathrm{log}(A_{\mathrm{amphi}} / A_{\mathrm{hypo}})$$
+
+We measured the photosynthetic rate of an untreated amphistomatous leaf ($A_{\mathrm{amphi}}$) over a range of $g_\text{sw}$ values.
+We refer to this as an $A \textendash g_\text{sw}$ curve.
+We compared the $A \textendash g_\text{sw}$ curve of the untreated leaf to the photosynthetic rate of pseudohypostomatous leaf ($A_\mathrm{hypo}$), which is the same leaf but with gas exchange through the upper surface blocked by a neutral density plastic (propafilm).
+
+We measured $A \textendash g_\text{sw}$ curves using a portable infrared gas analyzer (LI-6800PF, LI-COR Biosciences, Lincoln, Nebraska, USA).
+Light-acclimated plants were placed under LEDs dimmed to match their light treatment during gas exchange measurements.
+We estimated the photosynthetic rate ($A$) and stomatal conductance to water vapor ($g_\text{sw}$) at ambient CO$_2$ (415 ppm) and $T_\mathrm{leaf} = \qty{25.0}{\degreeCelsius}$.
+The irradiance of the light source in the pseudohypo leaf was higher because the propafilm reduces transmission.
+To compensate for reduced transmission, we increased incident $\mathrm{PPFD}$ for pseudohypo leaves by a factor 1/0.91, the inverse of the measured transmissivity of the propafilm.
+We also set the stomatal conductance ratio, for purposes of calculating boundary layer conductance, to 0 for pseudohypo leaves following manufacturer directions.
+
+We collected four $A \textendash g_\text{sw}$ curves per leaf, an amphi (untreated) curve and a pseudohypo (treated) curve at high light-intensity ($\mathrm{PPFD} = \qty{2000}{\micro\mol\raiseto{-2}\meter\raiseto{-1}\second}$) and low light-intensity ($\mathrm{PPFD} = \qty{150}{\micro\mol\raiseto{-2}\meter\raiseto{-1}\second}$).
+We always measured high light-intensity curves first because photosynthetic downregulation is faster than upregulation in these species.
+To control for order effects, we alternated between starting with amphi or pseudohypo leaf measurements.
+Unlike @triplett_amphistomy_2024, preliminary experiments with *Solanum* indicated a strong order effect in that $A$ declined in the second curve.
+Therefore, we made measurements over two days.
+On the first day, we measured high and low light-intensity curves for either amphi or pseudohypo leaves; on the second day, we measured high and low light-intensity curves on the other leaf type.
+
+In all cases, we acclimated the focal leaf to high light ($\mathrm{PPFD} = \qty{2000}{\micro\mol\raiseto{-2}\meter\raiseto{-1}\second}$) and high relative humidity ($\mathrm{RH} = 70\%$) until $A$ and $g_\text{sw}$ reach their maximum.
+After that, we decreased $\mathrm{RH}$ to $\approx 10\%$ to induce rapid stomatal closure without biochemical downregulation.
+Hence, $A_\text{amphi}$ and $A_\text{hypo}$ were both measured at low chamber humidity after the leaf had acclimated to high humidity.
+All other environmental conditions in the leaf chamber remained the same.
+We logged data until $g_\text{sw}$ reached its nadir.
+We then acclimated the leaf to low light ($\mathrm{PPFD} = \qty{150}{\micro\mol\raiseto{-2}\meter\raiseto{-1}\second}$) and 70\% $\mathrm{RH}$ before inducing stomatal closure with low $\mathrm{RH}$.

### 2024-03-06 ms/ms.qmd
started adding latex shortcuts to ms

@@
   \usepackage{siunitx}
+  \newcommand{\agcurve}{$A \textendash g_\text{sc}$}
+  \newcommand{\gsc}{$g_\text{sc}$}
+  \newcommand{\ppfd}{$\mathrm{PPFD}$}
+  \newcommand{\ppfdequals}[1]{$\mathrm{PPFD} = \qty{#1}{\micro\mol\raiseto{-2}\meter\raiseto{-1}\second}$}
+  \newcommand{\rh}{$\mathrm{RH}$}
 mainfont: Times New Roman
@@
 
-### Amphistomy advantage
+#### Amphistomy advantage
 
 We estimated 'amphistomy advantage' ($\mathrm{AA}$) *sensu* @parkhurst_adaptive_1978, but with modifications previously described in @triplett_amphistomy_2024.
-$\mathrm{AA}$ is calculated as the log-response ratio of $A$ compared at the same total $g_\text{sw}$:
+$\mathrm{AA}$ is calculated as the log-response ratio of $A$ compared at the same total \gsc:
 
@@
 
-We measured the photosynthetic rate of an untreated amphistomatous leaf ($A_{\mathrm{amphi}}$) over a range of $g_\text{sw}$ values.
-We refer to this as an $A \textendash g_\text{sw}$ curve.
-We compared the $A \textendash g_\text{sw}$ curve of the untreated leaf to the photosynthetic rate of pseudohypostomatous leaf ($A_\mathrm{hypo}$), which is the same leaf but with gas exchange through the upper surface blocked by a neutral density plastic (propafilm).
+We measured the photosynthetic rate of an untreated amphistomatous leaf ($A_{\mathrm{amphi}}$) over a range of \gsc values.
+We refer to this as an \agcurve curve.
+We compared the \agcurve curve of the untreated leaf to the photosynthetic rate of pseudohypostomatous leaf ($A_\mathrm{hypo}$), which is the same leaf but with gas exchange through the upper surface blocked by a neutral density plastic (propafilm).
 
-We measured $A \textendash g_\text{sw}$ curves using a portable infrared gas analyzer (LI-6800PF, LI-COR Biosciences, Lincoln, Nebraska, USA).
+We measured \agcurve curves using a portable infrared gas analyzer (LI-6800PF, LI-COR Biosciences, Lincoln, Nebraska, USA).
 Light-acclimated plants were placed under LEDs dimmed to match their light treatment during gas exchange measurements.
-We estimated the photosynthetic rate ($A$) and stomatal conductance to water vapor ($g_\text{sw}$) at ambient CO$_2$ (415 ppm) and $T_\mathrm{leaf} = \qty{25.0}{\degreeCelsius}$.
+We estimated the photosynthetic rate ($A$) and stomatal conductance to CO$_2$ (\gsc) at ambient CO$_2$ (415 ppm) and $T_\mathrm{leaf} = \qty{25.0}{\degreeCelsius}$.
 The irradiance of the light source in the pseudohypo leaf was higher because the propafilm reduces transmission.
-To compensate for reduced transmission, we increased incident $\mathrm{PPFD}$ for pseudohypo leaves by a factor 1/0.91, the inverse of the measured transmissivity of the propafilm.
+To compensate for reduced transmission, we increased incident \ppfd for pseudohypo leaves by a factor 1/0.91, the inverse of the measured transmissivity of the propafilm.
 We also set the stomatal conductance ratio, for purposes of calculating boundary layer conductance, to 0 for pseudohypo leaves following manufacturer directions.
 
-We collected four $A \textendash g_\text{sw}$ curves per leaf, an amphi (untreated) curve and a pseudohypo (treated) curve at high light-intensity ($\mathrm{PPFD} = \qty{2000}{\micro\mol\raiseto{-2}\meter\raiseto{-1}\second}$) and low light-intensity ($\mathrm{PPFD} = \qty{150}{\micro\mol\raiseto{-2}\meter\raiseto{-1}\second}$).
+We collected four \agcurve curves per leaf, an amphi (untreated) curve and a pseudohypo (treated) curve at high light-intensity (\ppfdequals{2000}) and low light-intensity (\ppfdequals{150}).
 We always measured high light-intensity curves first because photosynthetic downregulation is faster than upregulation in these species.
@@
 
-In all cases, we acclimated the focal leaf to high light ($\mathrm{PPFD} = \qty{2000}{\micro\mol\raiseto{-2}\meter\raiseto{-1}\second}$) and high relative humidity ($\mathrm{RH} = 70\%$) until $A$ and $g_\text{sw}$ reach their maximum.
-After that, we decreased $\mathrm{RH}$ to $\approx 10\%$ to induce rapid stomatal closure without biochemical downregulation.
+In all cases, we acclimated the focal leaf to high light (\ppfdequals{2000}) and high relative humidity ($\mathrm{RH} = 70\%$) until $A$ and \gsc reach their maximum.
+After that, we decreased \rh to $\approx 10\%$ to induce rapid stomatal closure without biochemical downregulation.
 Hence, $A_\text{amphi}$ and $A_\text{hypo}$ were both measured at low chamber humidity after the leaf had acclimated to high humidity.
 All other environmental conditions in the leaf chamber remained the same.
-We logged data until $g_\text{sw}$ reached its nadir.
-We then acclimated the leaf to low light ($\mathrm{PPFD} = \qty{150}{\micro\mol\raiseto{-2}\meter\raiseto{-1}\second}$) and 70\% $\mathrm{RH}$ before inducing stomatal closure with low $\mathrm{RH}$.
+We logged data until \gsc reached its nadir.
+We then acclimated the leaf to low light (\ppfdequals{150}) and 70\% \rh before inducing stomatal closure with low \rh and logging data as described above.
+
+#### Light-response curves
+
+We measured light-response curves on the same leaflets as \agcurve curves
+
+testing \gsc

### 2024-03-06 ms/ms.tex
started adding latex shortcuts to ms

@@
 \usepackage{siunitx}
+\newcommand{\gsc}{$\g_text{sc}$}
 \makeatletter
@@
 \maketitle
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, borderline west={3pt}{0pt}{shadecolor}, breakable, boxrule=0pt, interior hidden, enhanced, sharp corners]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, sharp corners, enhanced, breakable, borderline west={3pt}{0pt}{shadecolor}, frame hidden, boxrule=0pt]}{\end{tcolorbox}}\fi
 
@@
 {[}@peralta\_taxonomy\_2008{]} and the cultivated tomato \emph{S.
-lycopersicum} var. \emph{lycopersicum} (tab:accessions). Seeds provided
-by the Tomato Genetics Resource Center germinated on moist paper in
-plastic boxes after soaking for 30-60 minutes in a 50\% (volume per
-volume) solution of household bleach and water, followed by a thorough
-rinse. We transferred seedlings to cell-pack flats containing Pro-Mix BX
-potting mix (Premier Tech, Rivière-du-Loup, Quebec, Canada) once
-cotyledons fully emerged, typically within 1-2 weeks of sowing. We grew
-seeds and seedlings for both sun and shade treatments under the same
-environmental conditions (12:12 h, 25:\(\qty{20}{\degreeCelsius}\),
+lycopersicum} var. \emph{lycopersicum} (tab:accessions). Due to
+constraints on growth space and time, we spread out measurements over 80
+weeks from May 1, 2022 to October 31, 2023. Replicates within accession
+were evenly spread out over this period to prevent confounding of
+temporal variation in growth conditions with accession. {[}anything else
+to say here? maybe explain accession selection and phylogeny?{]}
+
+\begin{table}
+
+\end{table}
+
+\hypertarget{plant-growth-conditions}{%
+\subsubsection{Plant growth conditions}\label{plant-growth-conditions}}
+
+In all growth spaces, we recorded \(\mathrm{PPFD}\) using full spectrum
+quantum sensors (SQ-500-SS, Apogee Instruments, Logan, Utah, USA); we
+recorded temperature, RH, and {[}CO\(_2\){]} using an EE894 sensor (E+E
+Elektronik, Engerwitzdorf, Austria) protected by a radiation shield. All
+environmental measurements were taken every 10 minutes from the middle
+of plants racks at approximately the same height as the leaves we
+measured. We measured leaf temperature of focal leaves prior to
+measurement using an infrared radiometer (SI-111-SS, Apogee Instruments,
+Logan, Utah, USA).
+
+\hypertarget{germination-and-seedling-stage}{%
+\paragraph{Germination and seedling
+stage}\label{germination-and-seedling-stage}}
+
+Seeds provided by the Tomato Genetics Resource Center germinated on
+moist paper in plastic boxes after soaking for 30-60 minutes in a 50\%
+(volume per volume) solution of household bleach and water, followed by
+a thorough rinse. We transferred seedlings to cell-pack flats containing
+Pro-Mix BX potting mix (Premier Tech, Rivière-du-Loup, Quebec, Canada)
+once cotyledons fully emerged, typically within 1-2 weeks of sowing. We
+grew seeds and seedlings for both sun and shade treatments under the
+same environmental conditions (12:12 h, 25:\(\qty{20}{\degreeCelsius}\),
 40:60 RH day:night cycle). LED light provided
-\(\mathrm{PPFD} = \qty{200}{\micro\mol\raiseto{-2}\meter\raiseto{-1}\second}\).
+\(\mathrm{PPFD} = \qty{200}{\micro\mol\raiseto{-2}\meter\raiseto{-1}\second}\)
+(Fluence RAZRx, Austin, Texas, USA).
 
-After seedlings established in cell-pack flats for \(\approx 2\) weeks,
-we transplanted them to 3.78 L plastic pots containing 60\% Pro-Mix BX
-potting mix, 20\% coral sand (Pro-Pak, Honolulu, Hawaiʻi, USA), and 20\%
-cinders (Niu Nursery, Honolulu, Hawaiʻi, USA). Percentage composition is
-on a volume basis. The soil mixture contained slow release NPK
-fertilizer following manufacturer instructions (Osmocote Smart-Release
-Plant Food Flower \& Vegetable, The Scotts Company, Marysville, Ohio,
-USA).
+\hypertarget{light-treatments}{%
+\paragraph{Light treatments}\label{light-treatments}}
 
-outline of tab:accessions LA number,Species name, lat, lon, elev, SPLASH
-light value
+Seedlings were randomly assigned in alternating order within accession
+to the sun or shade treatment during transplanting. After seedlings
+established in cell-pack flats for \(\approx 2\) weeks, we transplanted
+them to 3.78 L plastic pots containing 60\% Pro-Mix BX potting mix, 20\%
+coral sand (Pro-Pak, Honolulu, Hawaiʻi, USA), and 20\% cinders (Niu
+Nursery, Honolulu, Hawaiʻi, USA). Percentage composition is on a volume
+basis. The soil mixture contained slow release NPK fertilizer following
+manufacturer instructions (Osmocote Smart-Release Plant Food Flower \&
+Vegetable, The Scotts Company, Marysville, Ohio, USA). We determined pot
+field capacity one week after transplanting using a scale (Ohaus V12P15
+Valor 1000, Parsippany, New Jersey, USA) and watered to field capacity
+three times per week to prevent drought stress.
+
+We assigned sun and shade treatment to lower and upper racks of a
+\(\qty{1.22}{\meter} \times \qty{2.44}{\meter}\) shelving unit in a
+climate-controlled growth room. We assigned the sun treatment to the
+lower rack to limit diffuse light from reaching the shade treatment. The
+average daytime \(\mathrm{PPFD}\) was
+\(\qty{800}{\micro\mol\raiseto{-2}\meter\raiseto{-1}\second}\) and
+\(\qty{125}{\micro\mol\raiseto{-2}\meter\raiseto{-1}\second}\) for sun
+and shade treatments, respectively. To isolate the effect of light
+intensity from quality, we used the same LED model with the the same
+spectrum (Fluence SPYDR 2i, Austin, Texas, USAS), but dimmed the lights
+in the shade treatment. To maintain homogeneous environmental conditions
+other than light, we mixed air within the growth room using an air
+circulator (Vornado 693DC, Andover, Kansas, USA) and within racks using
+a miniature oscillating air circulator (Vornado Atom 1, Andover, Kansas,
+USA). Despite these efforts, the air in the sun treatment was on average
+\(\qty{1}{\degreeCelsius}\) warmer and the average RH was consequently 5
+lower. However, because of evaporative cooling, the leaves in the sun
+treatment were only \(\qty{1}{\degreeCelsius}\) on average (\(n = 600\)
+leaves).
+
+\hypertarget{leaf-trait-measurements}{%
+\subsubsection{Leaf trait measurements}\label{leaf-trait-measurements}}
+
+We selected a fully expanded, unshaded leaf at least six leaves above
+the cotyledons during early vegetative growth. This typically meant that
+plants had grown in light treatments for \(\approx\) 4 weeks, ensuring
+they had time to sense and respond developmentally to the light
+intensity of the treatment rather than the seedling conditions
+{[}@schoch\_dependence\_1980{]}. Shade plants grew slower than sun
+plants, hence leaves at the same developmental stage were measured on
+chronologically older plants in the shade treatment. In some sun plants,
+we had to use leaves higher on the stem because short internodes made
+lower leaves inaccessible with the gas exchange equipment. We measured
+terminal leaflets in 70.0\% of cases, but used the lateral leaflet
+closest to the terminal leaflet when it was damaged or difficult to
+clamp into the gas exchange chamber. When a leaflet was damanged during
+gas exchange measurements, we collected anatomical data from the nearest
+leaflet on the same leaf (10.0)\%. We measured chlorophyll concentration
+index (CCI) using a chrolophyll concentration meter (MC-100, Apogee
+Instruments, Logan, Utah, USA) on the lamina of focal leaflets before
+gas exchange measurements at the same time we measured leaf temperature.
+
+\hypertarget{amphistomy-advantage}{%
+\paragraph{Amphistomy advantage}\label{amphistomy-advantage}}
+
+We estimated `amphistomy advantage' (\(\mathrm{AA}\)) \emph{sensu}
+@parkhurst\_adaptive\_1978, but with modifications previously described
+in @triplett\_amphistomy\_2024. \(\mathrm{AA}\) is calculated as the
+log-response ratio of \(A\) compared at the same total \(g_\text{sw}\):
+
+\[\mathrm{AA} = \mathrm{log}(A_{\mathrm{amphi}} / A_{\mathrm{hypo}})\]
+
+We measured the photosynthetic rate of an untreated amphistomatous leaf
+(\(A_{\mathrm{amphi}}\)) over a range of \(g_\text{sw}\) values. We
+refer to this as an \(A \textendash g_\text{sw}\) curve. We compared the
+\(A \textendash g_\text{sw}\) curve of the untreated leaf to the
+photosynthetic rate of pseudohypostomatous leaf (\(A_\mathrm{hypo}\)),
+which is the same leaf but with gas exchange through the upper surface
+blocked by a neutral density plastic (propafilm).
+
+We measured \(A \textendash g_\text{sw}\) curves using a portable
+infrared gas analyzer (LI-6800PF, LI-COR Biosciences, Lincoln, Nebraska,
+USA). Light-acclimated plants were placed under LEDs dimmed to match
+their light treatment during gas exchange measurements. We estimated the
+photosynthetic rate (\(A\)) and stomatal conductance to water vapor
+(\(g_\text{sw}\)) at ambient CO\(_2\) (415 ppm) and
+\(T_\mathrm{leaf} = \qty{25.0}{\degreeCelsius}\). The irradiance of the
+light source in the pseudohypo leaf was higher because the propafilm
+reduces transmission. To compensate for reduced transmission, we
+increased incident \(\mathrm{PPFD}\) for pseudohypo leaves by a factor
+1/0.91, the inverse of the measured transmissivity of the propafilm. We
+also set the stomatal conductance ratio, for purposes of calculating
+boundary layer conductance, to 0 for pseudohypo leaves following
+manufacturer directions.
+
+We collected four \(A \textendash g_\text{sc}\) curves per leaf, an
+amphi (untreated) curve and a pseudohypo (treated) curve at high
+light-intensity
+(\(\mathrm{PPFD} = \qty{2000}{\micro\mol\raiseto{-2}\meter\raiseto{-1}\second}\))
+and low light-intensity
+(\(\mathrm{PPFD} = \qty{150}{\micro\mol\raiseto{-2}\meter\raiseto{-1}\second}\)).
+We always measured high light-intensity curves first because
+photosynthetic downregulation is faster than upregulation in these
+species. To control for order effects, we alternated between starting
+with amphi or pseudohypo leaf measurements. Unlike
+@triplett\_amphistomy\_2024, preliminary experiments with \emph{Solanum}
+indicated a strong order effect in that \(A\) declined in the second
+curve. Therefore, we made measurements over two days. On the first day,
+we measured high and low light-intensity curves for either amphi or
+pseudohypo leaves; on the second day, we measured high and low
+light-intensity curves on the other leaf type.
+
+In all cases, we acclimated the focal leaf to high light
+(\(\mathrm{PPFD} = \qty{2000}{\micro\mol\raiseto{-2}\meter\raiseto{-1}\second}\))
+and high relative humidity (\(\mathrm{RH} = 70\%\)) until \(A\) and
+\(g_\text{sw}\) reach their maximum. After that, we decreased
+\(\mathrm{RH}\) to \(\approx 10\%\) to induce rapid stomatal closure
+without biochemical downregulation. Hence, \(A_\text{amphi}\) and
+\(A_\text{hypo}\) were both measured at low chamber humidity after the
+leaf had acclimated to high humidity. All other environmental conditions
+in the leaf chamber remained the same. We logged data until
+\(g_\text{sc}\) reached its nadir. We then acclimated the leaf to low
+light
+(\(\mathrm{PPFD} = \qty{150}{\micro\mol\raiseto{-2}\meter\raiseto{-1}\second}\))
+and 70\% \(\mathrm{RH}\) before inducing stomatal closure with low
+\(\mathrm{RH}\) and logging data as described above.
+
+\hypertarget{light-response-curves}{%
+\paragraph{Light-response curves}\label{light-response-curves}}
+
+We measured light-response curves on the same leaflets as
+\(A \textendash g_\text{sc}\)
+
+testing \gsc
 

### 2024-03-06 ms/test.qmd
started adding latex shortcuts to ms

@@
   \usepackage{siunitx}
+  \newcommand{\gsc}{$g_\text{sc}$}
+  \newcommand{\ppfd}[1]{$\mathrm{PPFD} = \qty{#1}{\micro\mol\raiseto{-2}\meter\raiseto{-1}\second}$}
 mainfont: Times New Roman
@@
 
-words $\qty{4}{\micro\meter}$
+words \gsc
 
-(12:12 $\qty{25}{\degreeCelsius}$
-
-and then this $\qty{200}{\micro\mol\raiseto{-2}\meter\raiseto{-1}\second}$
-
-$\qty{0.1}{\hertz}$
+this \ppfd{2000}

### 2024-03-06 ms/ms.qmd
working on materials and methods

@@
   \usepackage{siunitx}
-  \newcommand{\agcurve}{$A \textendash g_\text{sc}$}
-  \newcommand{\gsc}{$g_\text{sc}$}
+  \newcommand{\aax}{$\mathrm{AA}$}
+  \newcommand{\Aamphi}{$A_{\mathrm{amphi}}$}
+  \newcommand{\Ahypo}{$A_{\mathrm{hypo}}$}
+  \newcommand{\agcurve}{$A \textendash g_\text{sw}$}
+  \newcommand{\caequals}[1]{$C_\mathrm{a} = \qty{#1}{\micro\mol\raiseto{-1}\mol}$}
+  \newcommand{\gsw}{$g_\text{sw}$}
   \newcommand{\ppfd}{$\mathrm{PPFD}$}
@@
   \newcommand{\rh}{$\mathrm{RH}$}
+  \newcommand{\rhequals}[1]{$\mathrm{RH} = #1\%$}
+  \newcommand{\tleafequals}[1]{$T_\mathrm{leaf} = \qty{#1}{\degreeCelsius}$}
 mainfont: Times New Roman
+bibliography: solanum-aa.bib
 csl: https://www.zotero.org/styles/science
@@
 percent_change_leaflet = "10.0" # percent of time that anatomical traits were measured on different leaflet than LICOR measurements
+percent_aa_and_aq = "90.0" # percent of times where AA and AQ curves were done on same leaflet
+p_red_high = "97.7" # percent red LED in high light-intensity treatment
+p_blue_high = "2.3"
+p_red_low = "87.0"
+p_blue_low = "13.0"
+
+percent_lma_switch = "5.0" # percent measure LMA on alternate leaflet
+percent_lma_missing = "3.0" # percent leaves missing LMA
+
 ```
@@
 We assigned the sun treatment to the lower rack to limit diffuse light from reaching the shade treatment.
-The average daytime $\mathrm{PPFD}$ was $\qty{`r ppfd_sun`}{\micro\mol\raiseto{-2}\meter\raiseto{-1}\second}$ and $\qty{`r ppfd_shade`}{\micro\mol\raiseto{-2}\meter\raiseto{-1}\second}$ for sun and shade treatments, respectively.
+The average daytime \ppfd{} was \ppfdequals{`r ppfd_sun`} ($\qty{`r ppfd_sun`}{\micro\mol\raiseto{-2}\meter\raiseto{-1}\second}$) and \ppfdequals{`r ppfd_shade`} ($\qty{`r ppfd_shade`}{\micro\mol\raiseto{-2}\meter\raiseto{-1}\second}$) for sun and shade treatments, respectively.
 To isolate the effect of light intensity from quality, we used the same LED model with the the same spectrum (Fluence SPYDR 2i, Austin, Texas, USAS), but dimmed the lights in the shade treatment.
@@
 We measured terminal leaflets in `r percent_terminal`\% of cases, but used the lateral leaflet closest to the terminal leaflet when it was damaged or difficult to clamp into the gas exchange chamber.
-When a leaflet was damanged during gas exchange measurements, we collected anatomical data from the nearest leaflet on the same leaf (`r percent_change_leaflet`)\%.
+When a leaflet was damaged during gas exchange measurements, we collected anatomical data from the nearest leaflet on the same leaf (`r percent_change_leaflet`)\%.
 We measured chlorophyll concentration index (CCI) using a chrolophyll concentration meter (MC-100, Apogee Instruments, Logan, Utah, USA) on the lamina of focal leaflets before gas exchange measurements at the same time we measured leaf temperature.
@@
 
-We estimated 'amphistomy advantage' ($\mathrm{AA}$) *sensu* @parkhurst_adaptive_1978, but with modifications previously described in @triplett_amphistomy_2024.
-$\mathrm{AA}$ is calculated as the log-response ratio of $A$ compared at the same total \gsc:
+We estimated 'amphistomy advantage' (\aax) *sensu* @parkhurst_adaptive_1978, but with modifications previously described in @triplett_amphistomy_2024.
+\aax{} is calculated as the log-response ratio of $A$ compared at the same total \gsw:
 
@@
 
-We measured the photosynthetic rate of an untreated amphistomatous leaf ($A_{\mathrm{amphi}}$) over a range of \gsc values.
-We refer to this as an \agcurve curve.
-We compared the \agcurve curve of the untreated leaf to the photosynthetic rate of pseudohypostomatous leaf ($A_\mathrm{hypo}$), which is the same leaf but with gas exchange through the upper surface blocked by a neutral density plastic (propafilm).
+We measured the photosynthetic rate of an untreated amphistomatous leaf (\Aamphi) over a range of \gsw{} values.
+We refer to this as an \agcurve{} curve.
+We compared the \agcurve{} curve of the untreated leaf to the photosynthetic rate of pseudohypostomatous leaf (\Ahypo), which is the same leaf but with gas exchange through the upper surface blocked by a neutral density plastic (propafilm).
 
-We measured \agcurve curves using a portable infrared gas analyzer (LI-6800PF, LI-COR Biosciences, Lincoln, Nebraska, USA).
+We measured \agcurve{} curves using a portable infrared gas analyzer (LI-6800PF, LI-COR Biosciences, Lincoln, Nebraska, USA).
 Light-acclimated plants were placed under LEDs dimmed to match their light treatment during gas exchange measurements.
-We estimated the photosynthetic rate ($A$) and stomatal conductance to CO$_2$ (\gsc) at ambient CO$_2$ (415 ppm) and $T_\mathrm{leaf} = \qty{25.0}{\degreeCelsius}$.
+We estimated the photosynthetic rate ($A$) and stomatal conductance to CO$_2$ (\gsw) at ambient CO$_2$ (\caequals{415}) and \tleafequals{25.0}.
 The irradiance of the light source in the pseudohypo leaf was higher because the propafilm reduces transmission.
@@
 
-We collected four \agcurve curves per leaf, an amphi (untreated) curve and a pseudohypo (treated) curve at high light-intensity (\ppfdequals{2000}) and low light-intensity (\ppfdequals{150}).
+We collected four \agcurve{} curves per leaf, an amphi (untreated) curve and a pseudohypo (treated) curve at high light-intensity (\ppfdequals{2000}; `r p_red_high`:`r p_blue_high` red:blue) and low light-intensity (\ppfdequals{150}; `r p_red_low`:`r p_blue_low` red:blue).
 We always measured high light-intensity curves first because photosynthetic downregulation is faster than upregulation in these species.
@@
 
-In all cases, we acclimated the focal leaf to high light (\ppfdequals{2000}) and high relative humidity ($\mathrm{RH} = 70\%$) until $A$ and \gsc reach their maximum.
-After that, we decreased \rh to $\approx 10\%$ to induce rapid stomatal closure without biochemical downregulation.
-Hence, $A_\text{amphi}$ and $A_\text{hypo}$ were both measured at low chamber humidity after the leaf had acclimated to high humidity.
+In all cases, we acclimated the focal leaf to high light (\ppfdequals{2000}) and high relative humidity (\rhequals{70}) until $A$ and \gsw{} reach their maximum.
+After that, we decreased \rh{} to $\approx 10\%$ to induce rapid stomatal closure without biochemical downregulation.
+Hence, \Aamphi{} and \Ahypo{} were both measured at low chamber humidity after the leaf had acclimated to high humidity.
 All other environmental conditions in the leaf chamber remained the same.
-We logged data until \gsc reached its nadir.
-We then acclimated the leaf to low light (\ppfdequals{150}) and 70\% \rh before inducing stomatal closure with low \rh and logging data as described above.
+We logged data until \gsw{} reached its nadir.
+We then acclimated the leaf to low light (\ppfdequals{150}) and \rhequals{70} before inducing stomatal closure with low \rh and logging data as described above.
 
@@
 
-We measured light-response curves on the same leaflets as \agcurve curves
+In `r percent_aa_and_aq`\% of plants, we measured light-response curves on the same leaflets as \agcurve{} curves.
+However, when a leaflet was damaged during \agcurve{} curves, we used the next closest leaflet for light-response curves.
+Leaves acclimated to high light-intensity (\ppfdequals{2000}), ambient CO$_2$ (\caequals{415}), \rhequals{50}, and \tleafequals{25}.
+After $A$ and \gsw{} stabilized, we measured $A$ at 20 light-intensity levels between $0$ and $\qty{2000}{\micro\mol\raiseto{-2}\meter\raiseto{-1}\second}$ in descending order.
 
-testing \gsc
+#### Stomatal anatomy
+
+<!-- are there any cases where we did not get stomatal anatomy data? -->
+
+We estimated the stomatal density and size on ab- and adaxial leaf surfaces from all leaves, using guard cell length as a proxy for stomatal size since it proportional to maximum conductance [@sack_developmental_2016].
+We made surface impressions of leaf lamina from the same area used for gas exchange measurements using a-silicone impression material (Zhermack elite HD+, light body, fast set, Rovigo, Italy).
+We applied clear nail polish to make positive replicas of the impression.
+After nail polish dried, we mounted replicas on a microscope slide using transparent tape [@mott_amphistomy_1991].
+We digitized a portion of each leaf surface replica using a brightfield microscope (Leica DM2000, Wetzlar, Germany).
+We counted and measured guard cell length on all stomata using the FIJI implementation of ImageJ2 version 2.3.0 [@schindelin_fiji_2012], then divided the count by the visible leaf area ($\qty{0.890}{\raiseto{2}\milli\meter}$) to estimate stomatal density.
+
+#### Internal leaf anatomy
+
+Add information on internal leaf anatomy, if using
+
+#### Leaf mass per area
+
+Leaf mass per area (LMA) is the dry mass divided by the leaflet area.
+We scanned fresh leaflets on a flat bed scanner (Epson V600, Los Alamitos, California.
+USA) and measured leaflet area from digital images using the FIJI implementation of ImageJ2 version 2.3.0 [@schindelin_fiji_2012].
+We dried leaves for 72 hours at $\qty{74}{\degreeCelsius}$ in a food dehydrator (Cosori CP267-FD, Vesync Co., Anaheim, California, USA) and weighed using a benchtop analytical balance (Ohaus PR64 Analytical Balance, Parsippany, New Jersey, USA).
+In $`r percent_lma_switch`\%$ we measured LMA on the adjacent leaflet because the focal leaflet was damaged or wilted while making surface impressions and we could not reliably estimate area.
+ LMA data are missing from $`r percent_lma_missing`\%$ of individuals because the area or mass was not recorded at all or recorded incorrectly.
+
+## References

### 2024-03-06 ms/ms.tex
working on materials and methods

@@
 \makeatother
+% definitions for citeproc citations
+\NewDocumentCommand\citeproctext{}{}
+\NewDocumentCommand\citeproc{mm}{%
+  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
+\makeatletter
+ % allow citations to break across lines
+ \let\@cite@ofmt\@firstofone
+ % avoid brackets around text for \cite:
+ \def\@biblabel#1{}
+ \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
+\makeatother
+\newlength{\cslhangindent}
+\setlength{\cslhangindent}{1.5em}
+\newlength{\csllabelwidth}
+\setlength{\csllabelwidth}{3em}
+\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
+ {\begin{list}{}{%
+  \setlength{\itemindent}{0pt}
+  \setlength{\leftmargin}{0pt}
+  \setlength{\parsep}{0pt}
+  % turn on hanging indent if param 1 is 1
+  \ifodd #1
+   \setlength{\leftmargin}{\cslhangindent}
+   \setlength{\itemindent}{-1\cslhangindent}
+  \fi
+  % set entry spacing
+  \setlength{\itemsep}{#2\baselineskip}}}
+ {\end{list}}
+\usepackage{calc}
+\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
+\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
+\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
+\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
 
@@
 \usepackage{siunitx}
-\newcommand{\gsc}{$\g_text{sc}$}
-\makeatletter
-\makeatother
-\makeatletter
-\makeatother
+\newcommand{\aax}{$\mathrm{AA}$}
+\newcommand{\Aamphi}{$A_{\mathrm{amphi}}$}
+\newcommand{\Ahypo}{$A_{\mathrm{hypo}}$}
+\newcommand{\agcurve}{$A \textendash g_\text{sw}$}
+\newcommand{\caequals}[1]{$C_\mathrm{a} = \qty{#1}{\micro\mol\raiseto{-1}\mol}$}
+\newcommand{\gsw}{$g_\text{sw}$}
+\newcommand{\ppfd}{$\mathrm{PPFD}$}
+\newcommand{\ppfdequals}[1]{$\mathrm{PPFD} = \qty{#1}{\micro\mol\raiseto{-2}\meter\raiseto{-1}\second}$}
+\newcommand{\rh}{$\mathrm{RH}$}
+\newcommand{\rhequals}[1]{$\mathrm{RH} = #1\%$}
+\newcommand{\tleafequals}[1]{$T_\mathrm{leaf} = \qty{#1}{\degreeCelsius}$}
 \makeatletter
@@
 \makeatletter
+\makeatother
+\makeatletter
 \@ifpackageloaded{caption}{}{\usepackage{caption}}
 \@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
-\makeatother
-\makeatletter
-\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
-\makeatother
-\makeatletter
-\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
-\makeatother
-\makeatletter
-\makeatother
-\makeatletter
 \makeatother
@@
 \fi
-\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
+\usepackage{bookmark}
+
 \IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
@@
 \maketitle
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, sharp corners, enhanced, breakable, borderline west={3pt}{0pt}{shadecolor}, frame hidden, boxrule=0pt]}{\end{tcolorbox}}\fi
 
-\hypertarget{notes}{%
-\subsection{Notes}\label{notes}}
+\subsection{Notes}\label{notes}
 
@@
 
-\hypertarget{main}{%
-\subsection{Main}\label{main}}
+\subsection{Main}\label{main}
 
@@
 on 600 individual plants from 30 accessions (average of 10 replicates
-per light treatment) using a recently developed method
-{[}@triplett\_amphistomy\_2024{]}. With this method, we directly compare
-the photosynthetic rate of an untreated amphistomatous leaf to that of
-the same leaf with gas exchange blocked through the adaxial (upper)
-surface by transparent plastic, which we refer to as `pseudohypostomy'.
-To compare amphi- and pseudohypostomatous leaves at identical whole-leaf
-\(g_\text{sc}\), we measure \(A\) over a range of \(g_\text{sc}\),
-inducing stomatal opening and closure by modulating humidity (see
-Materials and Methods for further details).
+per light treatment) using a recently developed method (\emph{1}). With
+this method, we directly compare the photosynthetic rate of an untreated
+amphistomatous leaf to that of the same leaf with gas exchange blocked
+through the adaxial (upper) surface by transparent plastic, which we
+refer to as `pseudohypostomy'. To compare amphi- and pseudohypostomatous
+leaves at identical whole-leaf \(g_\text{sc}\), we measure \(A\) over a
+range of \(g_\text{sc}\), inducing stomatal opening and closure by
+modulating humidity (see Materials and Methods for further details).
 
@@
 
-\hypertarget{materials-and-methods}{%
-\subsection{Materials and Methods}\label{materials-and-methods}}
+\subsection{Materials and Methods}\label{materials-and-methods}
 
@@
 
-\hypertarget{accessions}{%
-\subsubsection{Accessions}\label{accessions}}
+\subsubsection{Accessions}\label{accessions}
 
@@
 including representatives of all described species of \emph{Solanum}
-sect. \emph{Lycopersicon} and sect. \emph{Lycopersicoides}
-{[}@peralta\_taxonomy\_2008{]} and the cultivated tomato \emph{S.
-lycopersicum} var. \emph{lycopersicum} (tab:accessions). Due to
-constraints on growth space and time, we spread out measurements over 80
-weeks from May 1, 2022 to October 31, 2023. Replicates within accession
-were evenly spread out over this period to prevent confounding of
-temporal variation in growth conditions with accession. {[}anything else
-to say here? maybe explain accession selection and phylogeny?{]}
+sect. \emph{Lycopersicon} and sect. \emph{Lycopersicoides} (\emph{2})
+and the cultivated tomato \emph{S. lycopersicum} var.
+\emph{lycopersicum} (tab:accessions). Due to constraints on growth space
+and time, we spread out measurements over 80 weeks from May 1, 2022 to
+October 31, 2023. Replicates within accession were evenly spread out
+over this period to prevent confounding of temporal variation in growth
+conditions with accession. {[}anything else to say here? maybe explain
+accession selection and phylogeny?{]}
 
@@
 
-\end{table}
+\caption{\label{tbl-accessions}Solanum accessions}
 
-\hypertarget{plant-growth-conditions}{%
-\subsubsection{Plant growth conditions}\label{plant-growth-conditions}}
+\centering{
+
+}
+
+\end{table}%
+
+\subsubsection{Plant growth conditions}\label{plant-growth-conditions}
 
@@
 
-\hypertarget{germination-and-seedling-stage}{%
 \paragraph{Germination and seedling
-stage}\label{germination-and-seedling-stage}}
+stage}\label{germination-and-seedling-stage}
 
@@
 
-\hypertarget{light-treatments}{%
-\paragraph{Light treatments}\label{light-treatments}}
+\paragraph{Light treatments}\label{light-treatments}
 
@@
 lower rack to limit diffuse light from reaching the shade treatment. The
-average daytime \(\mathrm{PPFD}\) was
-\(\qty{800}{\micro\mol\raiseto{-2}\meter\raiseto{-1}\second}\) and
-\(\qty{125}{\micro\mol\raiseto{-2}\meter\raiseto{-1}\second}\) for sun
+average daytime \ppfd{} was \ppfdequals{800}
+(\(\qty{800}{\micro\mol\raiseto{-2}\meter\raiseto{-1}\second}\)) and
+\ppfdequals{125}
+(\(\qty{125}{\micro\mol\raiseto{-2}\meter\raiseto{-1}\second}\)) for sun
 and shade treatments, respectively. To isolate the effect of light
@@
 
-\hypertarget{leaf-trait-measurements}{%
-\subsubsection{Leaf trait measurements}\label{leaf-trait-measurements}}
+\subsubsection{Leaf trait measurements}\label{leaf-trait-measurements}
 
@@
 intensity of the treatment rather than the seedling conditions
-{[}@schoch\_dependence\_1980{]}. Shade plants grew slower than sun
-plants, hence leaves at the same developmental stage were measured on
-chronologically older plants in the shade treatment. In some sun plants,
-we had to use leaves higher on the stem because short internodes made
-lower leaves inaccessible with the gas exchange equipment. We measured
-terminal leaflets in 70.0\% of cases, but used the lateral leaflet
-closest to the terminal leaflet when it was damaged or difficult to
-clamp into the gas exchange chamber. When a leaflet was damanged during
-gas exchange measurements, we collected anatomical data from the nearest
-leaflet on the same leaf (10.0)\%. We measured chlorophyll concentration
-index (CCI) using a chrolophyll concentration meter (MC-100, Apogee
+(\emph{3}). Shade plants grew slower than sun plants, hence leaves at
+the same developmental stage were measured on chronologically older
+plants in the shade treatment. In some sun plants, we had to use leaves
+higher on the stem because short internodes made lower leaves
+inaccessible with the gas exchange equipment. We measured terminal
+leaflets in 70.0\% of cases, but used the lateral leaflet closest to the
+terminal leaflet when it was damaged or difficult to clamp into the gas
+exchange chamber. When a leaflet was damaged during gas exchange
+measurements, we collected anatomical data from the nearest leaflet on
+the same leaf (10.0)\%. We measured chlorophyll concentration index
+(CCI) using a chrolophyll concentration meter (MC-100, Apogee
 Instruments, Logan, Utah, USA) on the lamina of focal leaflets before
@@
 
-\hypertarget{amphistomy-advantage}{%
-\paragraph{Amphistomy advantage}\label{amphistomy-advantage}}
+\paragraph{Amphistomy advantage}\label{amphistomy-advantage}
 
-We estimated `amphistomy advantage' (\(\mathrm{AA}\)) \emph{sensu}
-@parkhurst\_adaptive\_1978, but with modifications previously described
-in @triplett\_amphistomy\_2024. \(\mathrm{AA}\) is calculated as the
-log-response ratio of \(A\) compared at the same total \(g_\text{sw}\):
+We estimated `amphistomy advantage' (\aax) \emph{sensu} (\emph{4}), but
+with modifications previously described in (\emph{1}). \aax{} is
+calculated as the log-response ratio of \(A\) compared at the same total
+\gsw:
 
@@
 We measured the photosynthetic rate of an untreated amphistomatous leaf
-(\(A_{\mathrm{amphi}}\)) over a range of \(g_\text{sw}\) values. We
-refer to this as an \(A \textendash g_\text{sw}\) curve. We compared the
-\(A \textendash g_\text{sw}\) curve of the untreated leaf to the
-photosynthetic rate of pseudohypostomatous leaf (\(A_\mathrm{hypo}\)),
-which is the same leaf but with gas exchange through the upper surface
-blocked by a neutral density plastic (propafilm).
+(\Aamphi) over a range of \gsw{} values. We refer to this as an
+\agcurve{} curve. We compared the \agcurve{} curve of the untreated leaf
+to the photosynthetic rate of pseudohypostomatous leaf (\Ahypo), which
+is the same leaf but with gas exchange through the upper surface blocked
+by a neutral density plastic (propafilm).
 
-We measured \(A \textendash g_\text{sw}\) curves using a portable
-infrared gas analyzer (LI-6800PF, LI-COR Biosciences, Lincoln, Nebraska,
-USA). Light-acclimated plants were placed under LEDs dimmed to match
-their light treatment during gas exchange measurements. We estimated the
-photosynthetic rate (\(A\)) and stomatal conductance to water vapor
-(\(g_\text{sw}\)) at ambient CO\(_2\) (415 ppm) and
-\(T_\mathrm{leaf} = \qty{25.0}{\degreeCelsius}\). The irradiance of the
-light source in the pseudohypo leaf was higher because the propafilm
-reduces transmission. To compensate for reduced transmission, we
-increased incident \(\mathrm{PPFD}\) for pseudohypo leaves by a factor
-1/0.91, the inverse of the measured transmissivity of the propafilm. We
-also set the stomatal conductance ratio, for purposes of calculating
-boundary layer conductance, to 0 for pseudohypo leaves following
-manufacturer directions.
+We measured \agcurve{} curves using a portable infrared gas analyzer
+(LI-6800PF, LI-COR Biosciences, Lincoln, Nebraska, USA).
+Light-acclimated plants were placed under LEDs dimmed to match their
+light treatment during gas exchange measurements. We estimated the
+photosynthetic rate (\(A\)) and stomatal conductance to CO\(_2\) (\gsw)
+at ambient CO\(_2\) (\caequals{415}) and \tleafequals{25.0}. The
+irradiance of the light source in the pseudohypo leaf was higher because
+the propafilm reduces transmission. To compensate for reduced
+transmission, we increased incident \ppfd for pseudohypo leaves by a
+factor 1/0.91, the inverse of the measured transmissivity of the
+propafilm. We also set the stomatal conductance ratio, for purposes of
+calculating boundary layer conductance, to 0 for pseudohypo leaves
+following manufacturer directions.
 
-We collected four \(A \textendash g_\text{sc}\) curves per leaf, an
-amphi (untreated) curve and a pseudohypo (treated) curve at high
-light-intensity
-(\(\mathrm{PPFD} = \qty{2000}{\micro\mol\raiseto{-2}\meter\raiseto{-1}\second}\))
-and low light-intensity
-(\(\mathrm{PPFD} = \qty{150}{\micro\mol\raiseto{-2}\meter\raiseto{-1}\second}\)).
-We always measured high light-intensity curves first because
-photosynthetic downregulation is faster than upregulation in these
-species. To control for order effects, we alternated between starting
-with amphi or pseudohypo leaf measurements. Unlike
-@triplett\_amphistomy\_2024, preliminary experiments with \emph{Solanum}
-indicated a strong order effect in that \(A\) declined in the second
-curve. Therefore, we made measurements over two days. On the first day,
-we measured high and low light-intensity curves for either amphi or
-pseudohypo leaves; on the second day, we measured high and low
+We collected four \agcurve{} curves per leaf, an amphi (untreated) curve
+and a pseudohypo (treated) curve at high light-intensity
+(\ppfdequals{2000}; 97.7:2.3 red:blue) and low light-intensity
+(\ppfdequals{150}; 87.0:13.0 red:blue). We always measured high
+light-intensity curves first because photosynthetic downregulation is
+faster than upregulation in these species. To control for order effects,
+we alternated between starting with amphi or pseudohypo leaf
+measurements. Unlike (\emph{1}), preliminary experiments with
+\emph{Solanum} indicated a strong order effect in that \(A\) declined in
+the second curve. Therefore, we made measurements over two days. On the
+first day, we measured high and low light-intensity curves for either
+amphi or pseudohypo leaves; on the second day, we measured high and low
 light-intensity curves on the other leaf type.
@@
 In all cases, we acclimated the focal leaf to high light
-(\(\mathrm{PPFD} = \qty{2000}{\micro\mol\raiseto{-2}\meter\raiseto{-1}\second}\))
-and high relative humidity (\(\mathrm{RH} = 70\%\)) until \(A\) and
-\(g_\text{sw}\) reach their maximum. After that, we decreased
-\(\mathrm{RH}\) to \(\approx 10\%\) to induce rapid stomatal closure
-without biochemical downregulation. Hence, \(A_\text{amphi}\) and
-\(A_\text{hypo}\) were both measured at low chamber humidity after the
-leaf had acclimated to high humidity. All other environmental conditions
-in the leaf chamber remained the same. We logged data until
-\(g_\text{sc}\) reached its nadir. We then acclimated the leaf to low
-light
-(\(\mathrm{PPFD} = \qty{150}{\micro\mol\raiseto{-2}\meter\raiseto{-1}\second}\))
-and 70\% \(\mathrm{RH}\) before inducing stomatal closure with low
-\(\mathrm{RH}\) and logging data as described above.
+(\ppfdequals{2000}) and high relative humidity (\rhequals{70}) until
+\(A\) and \gsw{} reach their maximum. After that, we decreased \rh{} to
+\(\approx 10\%\) to induce rapid stomatal closure without biochemical
+downregulation. Hence, \Aamphi{} and \Ahypo{} were both measured at low
+chamber humidity after the leaf had acclimated to high humidity. All
+other environmental conditions in the leaf chamber remained the same. We
+logged data until \gsw{} reached its nadir. We then acclimated the leaf
+to low light (\ppfdequals{150}) and \rhequals{70} before inducing
+stomatal closure with low \rh and logging data as described above.
 
-\hypertarget{light-response-curves}{%
-\paragraph{Light-response curves}\label{light-response-curves}}
+\paragraph{Light-response curves}\label{light-response-curves}
 
-We measured light-response curves on the same leaflets as
-\(A \textendash g_\text{sc}\)
+In 90.0\% of plants, we measured light-response curves on the same
+leaflets as \agcurve{} curves. However, when a leaflet was damaged
+during \agcurve{} curves, we used the next closest leaflet for
+light-response curves. Leaves acclimated to high light-intensity
+(\ppfdequals{2000}), ambient CO\(_2\) (\caequals{415}), \rhequals{50},
+and \tleafequals{25}. After \(A\) and \gsw{} stabilized, we measured
+\(A\) at 20 light-intensity levels between \(0\) and
+\(\qty{2000}{\micro\mol\raiseto{-2}\meter\raiseto{-1}\second}\) in
+descending order.
 
-testing \gsc
+\paragraph{Stomatal anatomy}\label{stomatal-anatomy}
+
+We estimated the stomatal density and size on ab- and adaxial leaf
+surfaces from all leaves, using guard cell length as a proxy for
+stomatal size since it proportional to maximum conductance (\emph{5}).
+We made surface impressions of leaf lamina from the same area used for
+gas exchange measurements using a-silicone impression material (Zhermack
+elite HD+, light body, fast set, Rovigo, Italy). We applied clear nail
+polish to make positive replicas of the impression. After nail polish
+dried, we mounted replicas on a microscope slide using transparent tape
+(\emph{6}). We digitized a portion of each leaf surface replica using a
+brightfield microscope (Leica DM2000, Wetzlar, Germany). We counted and
+measured guard cell length on all stomata using the FIJI implementation
+of ImageJ2 version 2.3.0 (\emph{7}), then divided the count by the
+visible leaf area (\(\qty{0.890}{\raiseto{2}\milli\meter}\)) to estimate
+stomatal density.
+
+\paragraph{Internal leaf anatomy}\label{internal-leaf-anatomy}
+
+Add information on internal leaf anatomy, if using
+
+\paragraph{Leaf mass per area}\label{leaf-mass-per-area}
+
+Leaf mass per area (LMA) is the dry mass divided by the leaflet area. We
+scanned fresh leaflets on a flat bed scanner (Epson V600, Los Alamitos,
+California. USA) and measured leaflet area from digital images using the
+FIJI implementation of ImageJ2 version 2.3.0 (\emph{7}). We dried leaves
+for 72 hours at \(\qty{74}{\degreeCelsius}\) in a food dehydrator
+(Cosori CP267-FD, Vesync Co., Anaheim, California, USA) and weighed
+using a benchtop analytical balance (Ohaus PR64 Analytical Balance,
+Parsippany, New Jersey, USA). In \(5.0\%\) we measured LMA on the
+adjacent leaflet because the focal leaflet was damaged or wilted while
+making surface impressions and we could not reliably estimate area. LMA
+data are missing from \(3.0\%\) of individuals because the area or mass
+was not recorded at all or recorded incorrectly.
+
+\subsection*{References}\label{references}
+\addcontentsline{toc}{subsection}{References}
+
+\phantomsection\label{refs}
+\begin{CSLReferences}{0}{1}
+\bibitem[\citeproctext]{ref-triplett_amphistomy_2024}
+\CSLLeftMargin{1. }%
+\CSLRightInline{G. Triplett, T. N. Buckley, C. D. Muir,
+\href{https://doi.org/10.1002/ajb2.16284}{Amphistomy increases leaf
+photosynthesis more in coastal than montane plants of {Hawaiian} ʻilima
+( \emph{{Sida} fallax} )}. \emph{American Journal of Botany}
+\textbf{111}, e16284 (2024).}
+
+\bibitem[\citeproctext]{ref-peralta_taxonomy_2008}
+\CSLLeftMargin{2. }%
+\CSLRightInline{I. E. Peralta, D. M. Spooner, S. Knapp, Taxonomy of wild
+tomatoes and their relatives (\emph{solanum} sect.
+\emph{Lycopersicoides}, sect. \emph{Juglandifolia}, sect.
+\emph{Lycopersicon}; {Solanaceae}). \textbf{84} (2008).}
+
+\bibitem[\citeproctext]{ref-schoch_dependence_1980}
+\CSLLeftMargin{3. }%
+\CSLRightInline{P.-G. Schoch, C. Zinsou, M. Sibi,
+\href{https://doi.org/10.1093/jxb/31.5.1211}{Dependence of the stomatal
+index on environmental factors during stomatal differentiation in leaves
+of \emph{{Vigna} sinensis} {L}.: 1. {Effect} of light intensity}.
+\emph{Journal of Experimental Botany} \textbf{31}, 1211--1216 (1980).}
+
+\bibitem[\citeproctext]{ref-parkhurst_adaptive_1978}
+\CSLLeftMargin{4. }%
+\CSLRightInline{D. F. Parkhurst,
+\href{https://doi.org/10.2307/2259142}{The adaptive significance of
+stomatal occurrence on one or both surfaces of leaves}. \emph{The
+Journal of Ecology} \textbf{66}, 367--383 (1978).}
+
+\bibitem[\citeproctext]{ref-sack_developmental_2016}
+\CSLLeftMargin{5. }%
+\CSLRightInline{L. Sack, T. N. Buckley,
+\href{https://doi.org/10.1104/pp.16.00476}{The developmental basis of
+stomatal density and flux}. \emph{Plant Physiology} \textbf{171},
+2358--2363 (2016).}
+
+\bibitem[\citeproctext]{ref-mott_amphistomy_1991}
+\CSLLeftMargin{6. }%
+\CSLRightInline{K. A. Mott, O. Michaelson, Amphistomy as an adaptation
+to high light intensity in \emph{{Ambrosia} cordifolia} ({Compositae}).
+\emph{American Journal of Botany} \textbf{78}, 76--79 (1991).}
+
+\bibitem[\citeproctext]{ref-schindelin_fiji_2012}
+\CSLLeftMargin{7. }%
+\CSLRightInline{J. Schindelin, I. Arganda-Carreras, E. Frise, V. Kaynig,
+M. Longair, T. Pietzsch, S. Preibisch, C. Rueden, S. Saalfeld, B.
+Schmid, J.-Y. Tinevez, D. J. White, V. Hartenstein, K. Eliceiri, P.
+Tomancak, A. Cardona, \href{https://doi.org/10.1038/nmeth.2019}{Fiji: An
+open-source platform for biological-image analysis}. \emph{Nature
+Methods} \textbf{9}, 676--682 (2012).}
+
+\end{CSLReferences}
 

### 2024-03-07 ms/ms.qmd
methods and outlining predictions

@@
 
+library(readr)
 # I am mostly putting placeholders here for objects/values that will need to be
 # loaded eventually to use in paper.
+fit_aa = read_rds("../objects/fit_dummy.rds")
+dim_fit_aa = dim(fit_aa$fit)
+
 n_plant = 600
@@
 -   ?
+
+## Authors
+
+definite: Wei Shen Lim, Dachuan Wang, Chris Muir
+question: Sam McKlin, Tom Buckley
 
@@
 | acclimatory  | cor(PAR, AA) = 0   | AAsun = AAshade  | AA2000 \> AA150 |
-|   plastic    | cor(PAR, AAsun) \> 0 | AAsun \> AAshade | AA2000 = AA150  |
+|   plastic    | cor(PAR, AA) = 0 | AAsun \> AAshade | AA2000 = AA150  |
 | constitutive | cor(PAR, AA) \> 0   | AAsun = AAshade  | AA2000 = AA150  |
@@
 
+<!-- 
 #### Internal leaf anatomy
@@
 Add information on internal leaf anatomy, if using 
+-->
 
@@
 
+### Data analysis
+
+#### General points (not sure where this goes yet)
+
+- models fit in Stan/brms
+- warmup and sampling iterations were chosen so that parameter ESS > 1000 and  $\hat{R} < 1.01$ [@gelman_inference_1992] and the effective sample size (ESS) was greater than $10^3$.
+
+We fit Bayesian models using the *R* package **brms** version `r fit_aa$version$brms` [@burkner_brms_2017] with MCMC sampling in *Stan* [@stan_development_team_stan_2023].
+We used CmdStan version `r fit_aa$version$cmdstan` and **cmdstanr** version `r fit_aa$version$cmdstanr` [@gabry_cmdstanr_2023] to interface with *R* version `r getRversion()` [@r_core_team_r:_2023].
+We sampled the posterior distribution from `r dim_fit_aa[2]` chains with `r dim_fit_aa[1]` iterations each after `r dim_fit_aa[1]` warmup iterations per chain.
+We estimated parameters and confidence intervals as the median and 95% quantile intervals of the posterior, respectively.
+
+#### Parameter estimates (this is a provisional section)
+
+* Assimilatory: 
+  + average AA2000 - AA150
+  + among accession variation in AA2000 - AA150
+* Plastic
+  + average AAsun - AAshade
+  + among accession variation in AAsun - AAshade
+* Constitutive
+  + cor(PAR, AA) at accession level
+
+Reinforcing interactions (we only consider these because these are ones which could explain why amphi leaves occur in high light habitats)
+
+* Assimilatory * Plastic (reinforcing interaction)
+  + AAsun - AAshade greater at 2000 than 150
+  + AA2000 - AA150 greater in sun leaves than shade leaves
+  + there may be var among accessions, but not correlated with PAR
+
+* Assimilatory * Constitutive (reinforcing interaction)
+  + cor(AA2000, PAR) > 0; cor(AA150, PAR) = 0
+  + AA2000 - AA150 varies among accessions, correlated with PAR
+
+* Plastic * Constitutive (reinforcing interaction)
+  + cor(AAsun, PAR) > 0; cor(AAshade, PAR) = 0
+  + AAsun - AAshade varies among accessions, correlated with PAR
+
+* Assimilatory * Plastic * Constitutive (reinforcing interaction)
+  + AAsun - AAshade greater at 2000 than 150
+  + AA2000 - AA150 greater in sun leaves than shade leaves
+  + AAsun - AAshade varies among accession, correlated with PAR
+  + AA2000 - AA150 varies among accession, correlated with PAR
+  + cor(AAsun2000, PAR) > cor(AAsun150, PAR) and > cor(AAshade2000, PAR) and > cor(AAshade150, PAR) in nonadditive way
+## Acknowledgements
+
+Justin Alter, Max Gatlin, Joana Kim, Jenna Matsuyama, Brandon Najarian, Kai Yasuda
+
 ## References

### 2024-03-07 ms/ms.tex
methods and outlining predictions

@@
 \makeatother
-% definitions for citeproc citations
-\NewDocumentCommand\citeproctext{}{}
-\NewDocumentCommand\citeproc{mm}{%
-  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
-\makeatletter
- % allow citations to break across lines
- \let\@cite@ofmt\@firstofone
- % avoid brackets around text for \cite:
- \def\@biblabel#1{}
- \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
-\makeatother
 \newlength{\cslhangindent}
@@
 \setlength{\csllabelwidth}{3em}
-\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
- {\begin{list}{}{%
-  \setlength{\itemindent}{0pt}
-  \setlength{\leftmargin}{0pt}
-  \setlength{\parsep}{0pt}
+\newlength{\cslentryspacingunit} % times entry-spacing
+\setlength{\cslentryspacingunit}{\parskip}
+\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
+ {% don't indent paragraphs
+  \setlength{\parindent}{0pt}
   % turn on hanging indent if param 1 is 1
   \ifodd #1
-   \setlength{\leftmargin}{\cslhangindent}
-   \setlength{\itemindent}{-1\cslhangindent}
+  \let\oldpar\par
+  \def\par{\hangindent=\cslhangindent\oldpar}
   \fi
   % set entry spacing
-  \setlength{\itemsep}{#2\baselineskip}}}
- {\end{list}}
+  \setlength{\parskip}{#2\cslentryspacingunit}
+ }%
+ {}
 \usepackage{calc}
-\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
-\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
-\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
+\newcommand{\CSLBlock}[1]{#1\hfill\break}
+\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
+\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
 \newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
@@
 \newcommand{\tleafequals}[1]{$T_\mathrm{leaf} = \qty{#1}{\degreeCelsius}$}
+\makeatletter
+\makeatother
+\makeatletter
+\makeatother
 \makeatletter
@@
 \makeatletter
+\@ifpackageloaded{caption}{}{\usepackage{caption}}
+\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
 \makeatother
 \makeatletter
-\@ifpackageloaded{caption}{}{\usepackage{caption}}
-\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
+\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
+\makeatother
+\makeatletter
+\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
+\makeatother
+\makeatletter
+\makeatother
+\makeatletter
 \makeatother
@@
 \fi
-\usepackage{bookmark}
-
+\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
 \IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
@@
 \maketitle
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, interior hidden, borderline west={3pt}{0pt}{shadecolor}, breakable, sharp corners, frame hidden, enhanced]}{\end{tcolorbox}}\fi
 
-\subsection{Notes}\label{notes}
+\hypertarget{notes}{%
+\subsection{Notes}\label{notes}}
 
@@
 
-\subsection{Main}\label{main}
+\hypertarget{main}{%
+\subsection{Main}\label{main}}
 
@@
 
-\subsection{Materials and Methods}\label{materials-and-methods}
+\hypertarget{materials-and-methods}{%
+\subsection{Materials and Methods}\label{materials-and-methods}}
 
@@
 
-\subsubsection{Accessions}\label{accessions}
+\hypertarget{accessions}{%
+\subsubsection{Accessions}\label{accessions}}
 
@@
 
-\caption{\label{tbl-accessions}Solanum accessions}
+\end{table}
 
-\centering{
-
-}
-
-\end{table}%
-
-\subsubsection{Plant growth conditions}\label{plant-growth-conditions}
+\hypertarget{plant-growth-conditions}{%
+\subsubsection{Plant growth conditions}\label{plant-growth-conditions}}
 
@@
 
+\hypertarget{germination-and-seedling-stage}{%
 \paragraph{Germination and seedling
-stage}\label{germination-and-seedling-stage}
+stage}\label{germination-and-seedling-stage}}
 
@@
 
-\paragraph{Light treatments}\label{light-treatments}
+\hypertarget{light-treatments}{%
+\paragraph{Light treatments}\label{light-treatments}}
 
@@
 
-\subsubsection{Leaf trait measurements}\label{leaf-trait-measurements}
+\hypertarget{leaf-trait-measurements}{%
+\subsubsection{Leaf trait measurements}\label{leaf-trait-measurements}}
 
@@
 
-\paragraph{Amphistomy advantage}\label{amphistomy-advantage}
+\hypertarget{amphistomy-advantage}{%
+\paragraph{Amphistomy advantage}\label{amphistomy-advantage}}
 
@@
 
-\paragraph{Light-response curves}\label{light-response-curves}
+\hypertarget{light-response-curves}{%
+\paragraph{Light-response curves}\label{light-response-curves}}
 
@@
 
-\paragraph{Stomatal anatomy}\label{stomatal-anatomy}
+\hypertarget{stomatal-anatomy}{%
+\paragraph{Stomatal anatomy}\label{stomatal-anatomy}}
 
@@
 
-\paragraph{Internal leaf anatomy}\label{internal-leaf-anatomy}
-
-Add information on internal leaf anatomy, if using
-
-\paragraph{Leaf mass per area}\label{leaf-mass-per-area}
+\hypertarget{leaf-mass-per-area}{%
+\paragraph{Leaf mass per area}\label{leaf-mass-per-area}}
 
@@
 
-\subsection*{References}\label{references}
+\hypertarget{data-analysis}{%
+\subsubsection{Data analysis}\label{data-analysis}}
+
+\hypertarget{general-points-not-sure-where-this-goes-yet}{%
+\paragraph{General points (not sure where this goes
+yet)}\label{general-points-not-sure-where-this-goes-yet}}
+
+\begin{itemize}
+\tightlist
+\item
+  models fit in Stan/brms
+\item
+  warmup and sampling iterations were chosen so that parameter ESS
+  \textgreater{} 1000 and \(\hat{R} < 1.01\) (\emph{8}) and the
+  effective sample size (ESS) was greater than \(10^3\).
+\end{itemize}
+
+We fit Bayesian models using the \emph{R} package \textbf{brms} version
+2.20.4 (\emph{9}) with MCMC sampling in \emph{Stan} version XXX
+(\emph{10}). We used CmdStan version 2.33.1 and \textbf{cmdstanr}
+version 0.7.1 (\emph{11}) to interface with \emph{R} version 4.3.2
+(\emph{12}). We sampled the posterior distribution from 4 chains with
+1000 iterations each after 1000 warmup iterations per chain. We
+estimated parameters and confidence intervals as the median and 95\%
+quantile intervals of the posterior, respectively.
+
+\hypertarget{references}{%
+\subsection*{References}\label{references}}
 \addcontentsline{toc}{subsection}{References}
 
-\phantomsection\label{refs}
-\begin{CSLReferences}{0}{1}
-\bibitem[\citeproctext]{ref-triplett_amphistomy_2024}
+\hypertarget{refs}{}
+\begin{CSLReferences}{0}{0}
+\leavevmode\vadjust pre{\hypertarget{ref-triplett_amphistomy_2024}{}}%
 \CSLLeftMargin{1. }%
@@
 
-\bibitem[\citeproctext]{ref-peralta_taxonomy_2008}
+\leavevmode\vadjust pre{\hypertarget{ref-peralta_taxonomy_2008}{}}%
 \CSLLeftMargin{2. }%
@@
 
-\bibitem[\citeproctext]{ref-schoch_dependence_1980}
+\leavevmode\vadjust pre{\hypertarget{ref-schoch_dependence_1980}{}}%
 \CSLLeftMargin{3. }%
@@
 
-\bibitem[\citeproctext]{ref-parkhurst_adaptive_1978}
+\leavevmode\vadjust pre{\hypertarget{ref-parkhurst_adaptive_1978}{}}%
 \CSLLeftMargin{4. }%
@@
 
-\bibitem[\citeproctext]{ref-sack_developmental_2016}
+\leavevmode\vadjust pre{\hypertarget{ref-sack_developmental_2016}{}}%
 \CSLLeftMargin{5. }%
@@
 
-\bibitem[\citeproctext]{ref-mott_amphistomy_1991}
+\leavevmode\vadjust pre{\hypertarget{ref-mott_amphistomy_1991}{}}%
 \CSLLeftMargin{6. }%
@@
 
-\bibitem[\citeproctext]{ref-schindelin_fiji_2012}
+\leavevmode\vadjust pre{\hypertarget{ref-schindelin_fiji_2012}{}}%
 \CSLLeftMargin{7. }%
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-gelman_inference_1992}{}}%
+\CSLLeftMargin{8. }%
+\CSLRightInline{A. Gelman, D. B. Rubin, Inference from iterative
+simulation using multiple sequences. \emph{Statistical Science}
+\textbf{7}, 457--472 (1992).}
+
+\leavevmode\vadjust pre{\hypertarget{ref-burkner_brms_2017}{}}%
+\CSLLeftMargin{9. }%
+\CSLRightInline{P.-C. Bürkner,
+\href{https://doi.org/10.18637/jss.v080.i01}{\textbf{Brms} : {An}
+\emph{r} {Package} for {Bayesian} {Multilevel} {Models} {Using}
+\emph{stan}}. \emph{Journal of Statistical Software} \textbf{80}
+(2017).}
+
+\leavevmode\vadjust pre{\hypertarget{ref-stan_development_team_stan_2023}{}}%
+\CSLLeftMargin{10. }%
+\CSLRightInline{Stan Development Team, \emph{Stan {Modeling} {Language}
+{Users} {Guide} and {Reference} {Manual}} (2023;
+\url{https://mc-stan.org}).}
+
+\leavevmode\vadjust pre{\hypertarget{ref-gabry_cmdstanr_2023}{}}%
+\CSLLeftMargin{11. }%
+\CSLRightInline{J. Gabry, R. Češnovar, \emph{Cmdstanr: {R} {Interface}
+to '{CmdStan}'} (2023;
+\href{https://mc-stan.org/cmdstanr,\%20https://discourse.mc-stan.org}{https://mc-stan.org/cmdstanr,
+https://discourse.mc-stan.org}).}
+
+\leavevmode\vadjust pre{\hypertarget{ref-r_core_team_r:_2023}{}}%
+\CSLLeftMargin{12. }%
+\CSLRightInline{R Core Team, \emph{R: {A} {Language} and {Environment}
+for {Statistical} {Computing}} (R Foundation for Statistical Computing,
+Vienna, Austria, 2023; \url{http://www.R-project.org/}).}
+
 \end{CSLReferences}

### 2024-03-07 ms/ms.qmd
drafting parameter estimation subsection

@@
 
+Base model (need to add indices):
+
+$$A = \beta_0 + \beta_1 \text{log}(g_\text{sw}) + \beta_2 \text{log}(g_\text{sw}) ^ 2 + \epsilon$$
+Estimate \aax{} by integrating over range of \gsw{} overlap (need to add indices for accession, replicate, light intensity):
+
+$$\mathrm{AA} = \int_{\text{min(log(}g_\text{sw}))}^{\text{max(log(}g_\text{sw}))} \text{log}\bigg(\frac{A_\text{amphi}(x; \theta_\text{amphi})}{A_\text{hypo}(x; \theta_\text{hypo})}\bigg) dx$$
+
+where $\text{min(log(}g_\text{sw}))$ and $\text{max(log(}g_\text{sw}))$ are the minumum and maximum values of $\text{log(}g_\text{sw})$ where amphi and pseudohypo curves overlap.
+\Aamphi{} is calculated from parameter estimates $\theta_\text{amphi} \in \{\beta_{0,\text{amphi}}, \beta_{1,\text{amphi}}, \beta_{2,\text{amphi}}\}$; \Ahypo{} is calculated from parameter estimates $\theta_\text{hypo} \in \{\beta_{0,\text{hypo}}, \beta_{1,\text{hypo}}, \beta_{2,\text{hypo}}\}$
+
 * Assimilatory: 

### 2024-03-14 ms/ms.qmd
working on model comparison

--- 2024-03-07 ms/ms.qmd
@@
 
+loo_version = "2.7.0"
 ```
@@
 
+Base model (need to add indices):
+
+$$\text{log(A)} = \beta_0 + \beta_1 \text{log}(g_\text{sw}) + \beta_2 \text{log}(g_\text{sw}) ^ 2 + \epsilon$$
+Estimate \aax{} by integrating over range of \gsw{} overlap (need to add indices for accession, replicate, light intensity):
+
+$$\mathrm{AA} = \int_{\text{min(log(}g_\text{sw}))}^{\text{max(log(}g_\text{sw}))} \text{log}\bigg(\frac{A_\text{amphi}(x; \theta_\text{amphi})}{A_\text{hypo}(x; \theta_\text{hypo})}\bigg) dx$$
+
+where $\text{min(log(}g_\text{sw}))$ and $\text{max(log(}g_\text{sw}))$ are the minimum and maximum values of $\text{log(}g_\text{sw})$ where amphi and pseudohypo curves overlap.
+\Aamphi{} is calculated from parameter estimates $\theta_\text{amphi} \in \{\beta_{0,\text{amphi}}, \beta_{1,\text{amphi}}, \beta_{2,\text{amphi}}\}$; \Ahypo{} is calculated from parameter estimates $\theta_\text{hypo} \in \{\beta_{0,\text{hypo}}, \beta_{1,\text{hypo}}, \beta_{2,\text{hypo}}\}$ for each curve.
+
+#### Predictions
+
 * Assimilatory: 
@@
   + cor(AAsun2000, PAR) > cor(AAsun150, PAR) and > cor(AAshade2000, PAR) and > cor(AAshade150, PAR) in nonadditive way
+
+We tested competing predictions through a combination of model selection and parameter estimation from the posterior of selected models.
+We started with a base model that included fixed effects of light intensity and light treatment on \aax, a random effect of accession on \aax, and a linear regression of PAR on accession-level \aax.
+We next compared the fit of the base model to a more complex with an interaction between light intensity and light treatment using LOOIC with the \textit{R} package **loo** version `r loo_version` [@vehtari_practical_2017].
+These comparisons allowed us to test the main predictions of each hypothesis and for a reinforcing interactions between assimilatory and plastic hypotheses (tab:predictions).
+However, these models cannot test predictions when there is an interaction with the constitutive hypothesis, because this hypothesis posits accession-levels variation in responses to light intensity and/or light treatment.
+Therefore, we tested for significant variation among accessions in light intensity and light treatment by comparing LOOIC values of models with and without random interactions between these factors and accession.
+
 ## Acknowledgements

### 2024-03-14 htc/fit_aa2.sh

@@ -1,0 +1,19 @@
+#!/bin/bash
+
+# untar R installation
+tar -xzf R413.tar.gz
+
+# untar Stan installation
+tar -xzf cmdstan-2.34.1.tar.gz
+
+# untar R Packages
+tar -xzf packages.tar.gz
+
+# make sure the script will use your R installation, 
+# and the working directory as its home location
+export PATH=$PWD/R/bin:$PATH
+export RHOME=$PWD/R
+export R_LIBS=$PWD/packages
+
+# run your script
+Rscript fit_aa2.R

### 2024-03-14 htc/htc.sh

@@ -77,6 +77,7 @@ rm fit_*
 
 # 1. Transfer files before logging in using scp
 scp data/prepared_rh_curves.rds data/stan_rh_curves.rds htc/fit_aa1.R htc/fit_aa1.sh htc/fit_aa1.sub stan/solanum-aa1.stan cdmuir@ap2002.chtc.wisc.edu:/home/cdmuir/solanum-aa/htc
+scp data/prepared_rh_curves.rds data/stan_rh_curves.rds htc/fit_aa2.R htc/fit_aa2.sh htc/fit_aa2.sub stan/solanum-aa2.stan cdmuir@ap2002.chtc.wisc.edu:/home/cdmuir/solanum-aa/htc
 # scp htc/fit_aa1.sub cdmuir@ap2002.chtc.wisc.edu:/home/cdmuir/solanum-aa/htc
 
 # 2. Login to submit node
@@ -84,12 +85,14 @@ ssh cdmuir@ap2002.chtc.wisc.edu
 
 # 3. Submit jobs
 condor_submit solanum-aa/htc/fit_aa1.sub
+condor_submit solanum-aa/htc/fit_aa2.sub
 
 # check status
-condor_q 12740
+condor_q 13584
 
 # 4. Retrieve results
 scp cdmuir@ap2002.chtc.wisc.edu:/home/cdmuir/fit_aa1.rds objects/
+scp cdmuir@ap2002.chtc.wisc.edu:/home/cdmuir/fit_aa2.rds objects/
 
 # clean up on submit node
 rm fit_*

### 2024-03-14 r/303_compare-parameters.R

@@ -17,15 +17,12 @@ s = fit_aa1$summary()
 fit_aa1$summary(c( "log_sigma_resid", "rho_resid", "b0_aa",
 "b_aa_light_intensity_2000",
 "b_aa_light_treatment_high",
-"log_sigma_aa_light_intensity_2000_acc",
-"log_sigma_aa_light_intensity_2000_acc_id",
-"log_sigma_aa_light_treatment_high_acc",
-"log_sigma_aa_light_treatment_high_acc_id",
 "log_sigma_aa_acc",
 "log_sigma_aa_acc_id",
 "b0_log_sigma_aa",
 "b_log_sigma_aa_light_intensity_2000",
-"b_log_sigma_aa_light_treatment_high"
+"b_log_sigma_aa_light_treatment_high",
+"b1_ppfd"
 )) |>
   select(variable, median, q5, q95, rhat)
 

### 2024-03-18 ms/ms.qmd
working on model description

@@
 We tested competing predictions through a combination of model selection and parameter estimation from the posterior of selected models.
-We started with a base model that included fixed effects of light intensity and light treatment on \aax, a random effect of accession on \aax, and a linear regression of PAR on accession-level \aax.
+We started with a base model that includes fixed effects of light intensity and light treatment on \aax, a random effect of accession on \aax, and a linear regression of PAR on accession-level \aax.
 We next compared the fit of the base model to a more complex with an interaction between light intensity and light treatment using LOOIC with the \textit{R} package **loo** version `r loo_version` [@vehtari_practical_2017].
@@
 Therefore, we tested for significant variation among accessions in light intensity and light treatment by comparing LOOIC values of models with and without random interactions between these factors and accession.
+
+| Hypothesis   | Prediction | Model |
+|--------------|------------|------------------|
+| Assimilatory |  | base model |
+| Plastic      |  | base model |
+| Constitutive |  | base model |
+| Assimilatory * Plastic |  | + intensity $\times$ treatment |
+| Assimilatory * Constitutive |  | + accession $\times$ intensity (random) |
+| Plastic * Constitutive |  | + accession $\times$ treatment (random) |
+| Assimilatory * Plastic * Constitutive |  | + accession $\times$ intensity $\times$ treatment (random) |
 

### 2024-03-18 htc/htc.sh

@@ -92,7 +92,11 @@ condor_submit solanum-aa/htc/fit_aa4.sub # ?
 condor_q 23163
 condor_q 23164
 condor_q 23165
-condor_q 23167
+condor_q 23167 # aa4
+condor_q 23168 # aa1 after some changes
+condor_q 23169 # aa2 after some changes
+condor_q 23170 # aa3 after some changes
+condor_q 23171 # aa4 after some changes
 
 # 4. Retrieve results
 scp cdmuir@ap2002.chtc.wisc.edu:/home/cdmuir/fit_aa1.rds objects/

### 2024-04-01 ms/ms.qmd
working on draft of data cleaning methods

@@
   \newcommand{\agcurve}{$A \textendash g_\text{sw}$}
+  \newcommand{\ca}{$C_\mathrm{a}$}
   \newcommand{\caequals}[1]{$C_\mathrm{a} = \qty{#1}{\micro\mol\raiseto{-1}\mol}$}
+  \newcommand{\cabetween}[2]{#1 < $C_\mathrm{a} < \qty{#2}{\micro\mol\raiseto{-1}\mol}$}
   \newcommand{\gsw}{$g_\text{sw}$}
@@
 percent_lma_missing = "3.0" # percent leaves missing LMA
+
+### Cleaning RH curves
+n_rh_curve_raw = 2400 # export from adaptive-amphistomy/r/07_process-licor.R
+n_points_per_rh_curves_raw = 100 # export from adaptive-amphistomy/r/07_process-licor.R
+ca_min = 410 # export from adaptive-amphistomy/r/10_filter-rh-curves.R
+ca_max = 420 # export from adaptive-amphistomy/r/10_filter-rh-curves.R
+n_min_points_per_rh_curve = 10 # export from adaptive-amphistomy/r/10_filter-rh-curves.R
 
@@
 
+### Cleaning \agcurve{} curves
+
+The raw data set consisted of `r n_rh_curve_raw` \agcurve{} curves with an average of `r n_points_per_rh_curves_raw`.
+Principled, consistent, manual curation of a data set this size is not feasible.
+Therefore, we automated data cleaning using custom *R* scripts.
+Cleaning is divided into XXX sequential steps:
+
+1. remove unreliable data points
+2. remove hysteretic portion of \agcurve{} curves at low \gsw{}
+3. remove outliers within each \agcurve{} curve
+4. remove replicates with no overlap between amphi and pseudohypo \agcurve{} curves
+5. remove replicates with outlier \aax values
+
+#### Remove unreliable data points
+
+*Rationale*: Unreliable data points consisted of those where chamber [CO$_2$] was unstable and therefore measurements are not biologically meaningful.
+
+*Procedure*: We retained data points where \cabetween{`r ca_min`}{`r ca_max`}.
+
+#### Remove hysteretic portion of \agcurve{} curves at low \gsw{}
+
+*Rationale*: In most \agcurve{} curves, we observed a hysteretic response at low \gsw.
+After \gsw{} and $A$ declined simultaneously, $A$ increased slightly as \gsw{} continued to decline or stabilize, indicating some leaf acclimation to low \rh.
+We removed this portion of the curve to focus curve-fitting on the primary domain where $A$ increases monotonically with \gsw{}.
+
+*Procedure*: For each curve, we removed data points after \gsw{} had reached its minimum unless there were fewer than `r n_min_points_per_rh_curve` data points remaining.
+
+WORKING HERE
+
 ### Data analysis
@@
 - models fit in Stan/brms
-- warmup and sampling iterations were chosen so that parameter ESS > 1000 and  $\hat{R} < 1.01$ [@gelman_inference_1992] and the effective sample size (ESS) was greater than $10^3$.
+- we chose the number of chains, warmup and sampling iterations, and maximum treedepth so that parameter estimates converged ($\hat{R} < 1.01$ [@gelman_inference_1992]) and the effective sample size (ESS) for each parameter was greater than $10^3$.
 

### 2024-04-01 ms/ms.tex
working on draft of data cleaning methods

--- 2024-03-06 ms/ms.tex
@@
 \newcommand{\agcurve}{$A \textendash g_\text{sw}$}
+\newcommand{\ca}{$C_\mathrm{a}$}
 \newcommand{\caequals}[1]{$C_\mathrm{a} = \qty{#1}{\micro\mol\raiseto{-1}\mol}$}
+\newcommand{\cabetween}[2]{#1 < $C_\mathrm{a} < \qty{#2}{\micro\mol\raiseto{-1}\mol}$}
 \newcommand{\gsw}{$g_\text{sw}$}
@@
 \end{itemize}
+
+\subsection{Authors}\label{authors}
+
+definite: Wei Shen Lim, Dachuan Wang, Chris Muir question: Sam McKlin,
+Tom Buckley
 
@@
 AA150 \\
-plastic & cor(PAR, AAsun) \textgreater{} 0 & AAsun \textgreater{}
-AAshade & AA2000 = AA150 \\
+plastic & cor(PAR, AA) = 0 & AAsun \textgreater{} AAshade & AA2000 =
+AA150 \\
 constitutive & cor(PAR, AA) \textgreater{} 0 & AAsun = AAshade & AA2000
@@
 
-\paragraph{Internal leaf anatomy}\label{internal-leaf-anatomy}
-
-Add information on internal leaf anatomy, if using
-
 \paragraph{Leaf mass per area}\label{leaf-mass-per-area}
@@
 was not recorded at all or recorded incorrectly.
+
+\subsubsection{\texorpdfstring{Cleaning \agcurve{}
+curves}{Cleaning  curves}}\label{cleaning-curves}
+
+The raw data set consisted of 2400 \agcurve{} curves with an average of
+100. Principled, consistent, manual curation of a data set this size is
+not feasible. Therefore, we automated data cleaning using custom
+\emph{R} scripts. Cleaning is divided into XXX sequential steps:
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\tightlist
+\item
+  remove unreliable data points
+\item
+  remove hysteretic portion of \agcurve{} curves at low \gsw{}
+\item
+  remove outliers within each \agcurve{} curve
+\item
+  remove replicates with no overlap between amphi and pseudohypo
+  \agcurve{} curves
+\item
+  remove replicates with outlier \aax values
+\end{enumerate}
+
+\paragraph{Remove unreliable data
+points}\label{remove-unreliable-data-points}
+
+\emph{Rationale}: Unreliable data points consisted of those where
+chamber {[}CO\(_2\){]} was unstable and therefore measurements are not
+biologically meaningful.
+
+\emph{Procedure}: We retained data points where \cabetween{410}{420}.
+
+\paragraph{\texorpdfstring{Remove hysteretic portion of \agcurve{}
+curves at low
+\gsw{}}{Remove hysteretic portion of  curves at low }}\label{remove-hysteretic-portion-of-curves-at-low}
+
+\emph{Rationale}: In most \agcurve{} curves, we observed a hysteretic
+response at low \gsw. After \gsw{} and \(A\) declined simultaneously,
+\(A\) increased slightly as \gsw{} continued to decline or stabilize,
+indicating some leaf acclimation to low \rh. We removed this portion of
+the curve to focus curve-fitting on the primary domain where \(A\)
+increases monotonically with \gsw{}.
+
+\emph{Procedure}: For each
+
+\subsubsection{Data analysis}\label{data-analysis}
+
+\paragraph{General points (not sure where this goes
+yet)}\label{general-points-not-sure-where-this-goes-yet}
+
+\begin{itemize}
+\tightlist
+\item
+  models fit in Stan/brms
+\item
+  we chose the number of chains, warmup and sampling iterations, and
+  maximum treedepth so that parameter estimates converged
+  (\(\hat{R} < 1.01\) (\emph{8})) and the effective sample size (ESS)
+  for each parameter was greater than \(10^3\).
+\end{itemize}
+
+We fit Bayesian models using the \emph{R} package \textbf{brms} version
+2.20.4 (\emph{9}) with MCMC sampling in \emph{Stan} (\emph{10}). We used
+CmdStan version 2.33.1 and \textbf{cmdstanr} version 0.7.1 (\emph{11})
+to interface with \emph{R} version 4.3.2 (\emph{12}). We sampled the
+posterior distribution from 4 chains with 1000 iterations each after
+1000 warmup iterations per chain. We estimated parameters and confidence
+intervals as the median and 95\% quantile intervals of the posterior,
+respectively.
+
+\paragraph{Parameter estimates (this is a provisional
+section)}\label{parameter-estimates-this-is-a-provisional-section}
+
+Base model (need to add indices):
+
+\[\text{log(A)} = \beta_0 + \beta_1 \text{log}(g_\text{sw}) + \beta_2 \text{log}(g_\text{sw}) ^ 2 + \epsilon\]
+Estimate \aax{} by integrating over range of \gsw{} overlap (need to add
+indices for accession, replicate, light intensity):
+
+\[\mathrm{AA} = \int_{\text{min(log(}g_\text{sw}))}^{\text{max(log(}g_\text{sw}))} \text{log}\bigg(\frac{A_\text{amphi}(x; \theta_\text{amphi})}{A_\text{hypo}(x; \theta_\text{hypo})}\bigg) dx\]
+
+where \(\text{min(log(}g_\text{sw}))\) and
+\(\text{max(log(}g_\text{sw}))\) are the minimum and maximum values of
+\(\text{log(}g_\text{sw})\) where amphi and pseudohypo curves overlap.
+\Aamphi{} is calculated from parameter estimates
+\(\theta_\text{amphi} \in \{\beta_{0,\text{amphi}}, \beta_{1,\text{amphi}}, \beta_{2,\text{amphi}}\}\);
+\Ahypo{} is calculated from parameter estimates
+\(\theta_\text{hypo} \in \{\beta_{0,\text{hypo}}, \beta_{1,\text{hypo}}, \beta_{2,\text{hypo}}\}\)
+for each curve.
+
+\paragraph{Predictions}\label{predictions}
+
+\begin{itemize}
+\tightlist
+\item
+  Assimilatory:
+
+  \begin{itemize}
+  \tightlist
+  \item
+    average AA2000 - AA150
+  \item
+    among accession variation in AA2000 - AA150
+  \end{itemize}
+\item
+  Plastic
+
+  \begin{itemize}
+  \tightlist
+  \item
+    average AAsun - AAshade
+  \item
+    among accession variation in AAsun - AAshade
+  \end{itemize}
+\item
+  Constitutive
+
+  \begin{itemize}
+  \tightlist
+  \item
+    cor(PAR, AA) at accession level
+  \end{itemize}
+\end{itemize}
+
+Reinforcing interactions (we only consider these because these are ones
+which could explain why amphi leaves occur in high light habitats)
+
+\begin{itemize}
+\tightlist
+\item
+  Assimilatory * Plastic (reinforcing interaction)
+
+  \begin{itemize}
+  \tightlist
+  \item
+    AAsun - AAshade greater at 2000 than 150
+  \item
+    AA2000 - AA150 greater in sun leaves than shade leaves
+  \item
+    there may be var among accessions, but not correlated with PAR
+  \end{itemize}
+\item
+  Assimilatory * Constitutive (reinforcing interaction)
+
+  \begin{itemize}
+  \tightlist
+  \item
+    cor(AA2000, PAR) \textgreater{} 0; cor(AA150, PAR) = 0
+  \item
+    AA2000 - AA150 varies among accessions, correlated with PAR
+  \end{itemize}
+\item
+  Plastic * Constitutive (reinforcing interaction)
+
+  \begin{itemize}
+  \tightlist
+  \item
+    cor(AAsun, PAR) \textgreater{} 0; cor(AAshade, PAR) = 0
+  \item
+    AAsun - AAshade varies among accessions, correlated with PAR
+  \end{itemize}
+\item
+  Assimilatory * Plastic * Constitutive (reinforcing interaction)
+
+  \begin{itemize}
+  \tightlist
+  \item
+    AAsun - AAshade greater at 2000 than 150
+  \item
+    AA2000 - AA150 greater in sun leaves than shade leaves
+  \item
+    AAsun - AAshade varies among accession, correlated with PAR
+  \item
+    AA2000 - AA150 varies among accession, correlated with PAR
+  \item
+    cor(AAsun2000, PAR) \textgreater{} cor(AAsun150, PAR) and
+    \textgreater{} cor(AAshade2000, PAR) and \textgreater{}
+    cor(AAshade150, PAR) in nonadditive way
+  \end{itemize}
+\end{itemize}
+
+We tested competing predictions through a combination of model selection
+and parameter estimation from the posterior of selected models. We
+started with a base model that includes fixed effects of light intensity
+and light treatment on \aax, a random effect of accession on \aax, and a
+linear regression of PAR on accession-level \aax. We next compared the
+fit of the base model to a more complex with an interaction between
+light intensity and light treatment using LOOIC with the \textit{R}
+package \textbf{loo} version 2.7.0 (\textbf{vehtari\_practical\_2017?}).
+These comparisons allowed us to test the main predictions of each
+hypothesis and for a reinforcing interactions between assimilatory and
+plastic hypotheses (tab:predictions). However, these models cannot test
+predictions when there is an interaction with the constitutive
+hypothesis, because this hypothesis posits accession-levels variation in
+responses to light intensity and/or light treatment. Therefore, we
+tested for significant variation among accessions in light intensity and
+light treatment by comparing LOOIC values of models with and without
+random interactions between these factors and accession.
+
+\begin{longtable}[]{@{}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.3182}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.2727}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.4091}}@{}}
+\toprule\noalign{}
+\begin{minipage}[b]{\linewidth}\raggedright
+Hypothesis
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+Prediction
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+Model
+\end{minipage} \\
+\midrule\noalign{}
+\endhead
+\bottomrule\noalign{}
+\endlastfoot
+Assimilatory & & base model \\
+Plastic & & base model \\
+Constitutive & & base model \\
+Assimilatory * Plastic & & + intensity \(\times\) treatment \\
+Assimilatory * Constitutive & & + accession \(\times\) intensity
+(random) \\
+Plastic * Constitutive & & + accession \(\times\) treatment (random) \\
+Assimilatory * Plastic * Constitutive & & + accession \(\times\)
+intensity \(\times\) treatment (random) \\
+\end{longtable}
+
+\subsection{Acknowledgements}\label{acknowledgements}
+
+Justin Alter, Max Gatlin, Joana Kim, Jenna Matsuyama, Brandon Najarian,
+Kai Yasuda
 
@@
 
+\bibitem[\citeproctext]{ref-gelman_inference_1992}
+\CSLLeftMargin{8. }%
+\CSLRightInline{A. Gelman, D. B. Rubin, Inference from iterative
+simulation using multiple sequences. \emph{Statistical Science}
+\textbf{7}, 457--472 (1992).}
+
+\bibitem[\citeproctext]{ref-burkner_brms_2017}
+\CSLLeftMargin{9. }%
+\CSLRightInline{P.-C. Bürkner,
+\href{https://doi.org/10.18637/jss.v080.i01}{\textbf{Brms} : {An}
+\emph{r} {Package} for {Bayesian} {Multilevel} {Models} {Using}
+\emph{stan}}. \emph{Journal of Statistical Software} \textbf{80}
+(2017).}
+
+\bibitem[\citeproctext]{ref-stan_development_team_stan_2023}
+\CSLLeftMargin{10. }%
+\CSLRightInline{Stan Development Team, \emph{Stan {Modeling} {Language}
+{Users} {Guide} and {Reference} {Manual}} (2023;
+\url{https://mc-stan.org}).}
+
+\bibitem[\citeproctext]{ref-gabry_cmdstanr_2023}
+\CSLLeftMargin{11. }%
+\CSLRightInline{J. Gabry, R. Češnovar, \emph{Cmdstanr: {R} {Interface}
+to '{CmdStan}'} (2023;
+\href{https://mc-stan.org/cmdstanr,\%20https://discourse.mc-stan.org}{https://mc-stan.org/cmdstanr,
+https://discourse.mc-stan.org}).}
+
+\bibitem[\citeproctext]{ref-r_core_team_r:_2023}
+\CSLLeftMargin{12. }%
+\CSLRightInline{R Core Team, \emph{R: {A} {Language} and {Environment}
+for {Statistical} {Computing}} (R Foundation for Statistical Computing,
+Vienna, Austria, 2023; \url{http://www.R-project.org/}).}
+
 \end{CSLReferences}

### 2024-04-02 ms/ms.qmd
wrote methods for outlier removal from RH curves

@@
 # loaded eventually to use in paper.
+aa_args = read_rds("../objects/aa_args.rds")
 fit_aa = read_rds("../objects/fit_dummy.rds")
@@
 n_points_per_rh_curves_raw = 100 # export from adaptive-amphistomy/r/07_process-licor.R
-ca_min = 410 # export from adaptive-amphistomy/r/10_filter-rh-curves.R
-ca_max = 420 # export from adaptive-amphistomy/r/10_filter-rh-curves.R
-n_min_points_per_rh_curve = 10 # export from adaptive-amphistomy/r/10_filter-rh-curves.R
+ca_min = aa_args$ca_min
+ca_max = aa_args$ca_max
+n_min_points_per_rh_curve = aa_args$n_min_points_per_rh_curve
+outlier_threshold = aa_args$outlier_threshold
 
@@
 
+#### Remove outliers within each \agcurve{} curve
+
+*Rationale*: Individual outliers within \agcurve{} curves, usually caused by transitory changes in chamber conditions, exert undue leverage on parameter estimates and cause bias and/or low precision in parameter estimates.
+
+*Procedure*: We fit provisional quadratic regressions for each curve using ordinary least squares with the `lm()` function in *R*.
+We sequentially removed data points with an external studentized residual greater than `r outlier_threshold` until none remained.
+
+were likely due to measurement error or leaf damage.
 WORKING HERE

### 2024-04-03 ms/ms.qmd
finished drafting data cleaning methods

@@
 # loaded eventually to use in paper.
-aa_args = read_rds("../objects/aa_args.rds")
+aa_args = read_rds("../objects/aa_args1.rds")
 fit_aa = read_rds("../objects/fit_dummy.rds")
@@
 n_min_points_per_rh_curve = aa_args$n_min_points_per_rh_curve
-outlier_threshold = aa_args$outlier_threshold
+point_outlier_threshold = aa_args$point_outlier_threshold
+aa_outlier_threshold = aa_args$aa_outlier_threshold
+thinning_interval = aa_args$thinning_interval
 
@@
 3. remove outliers within each \agcurve{} curve
-4. remove replicates with no overlap between amphi and pseudohypo \agcurve{} curves
-5. remove replicates with outlier \aax values
+4. thin redundant data points within each \agcurve{} curve
+5. remove replicates with no overlap between amphi and pseudohypo \agcurve{} curves
+6. Trim extreme \aax{} values
+
+**Table showing number of curves and points before and after each cleaning step**
 
@@
 *Procedure*: We fit provisional quadratic regressions for each curve using ordinary least squares with the `lm()` function in *R*.
-We sequentially removed data points with an external studentized residual greater than `r outlier_threshold` until none remained.
+We sequentially removed data points with an absolute external studentized residual greater than `r point_outlier_threshold` until none remained.
 
-were likely due to measurement error or leaf damage.
-WORKING HERE
+#### Remove replicates with no overlap between amphi and pseudohypo \agcurve{} curves
+
+*Rationale*: We could not estimate \aax{} for replicates where amphi and pseudohypo \agcurve{} curves did not overlap.
+
+*Procedure*: We removed replicates where the range of \gsw{} values for amphi and pseudohypo \agcurve{} curves did not overlap.
+
+#### Thin redundant data points within each \agcurve{} curve
+
+*Rationale*: Data points closely spaced along the \agcurve{} curve provide redundant information and may be highly correlated (i.e.
+pseudoreplication).
+This occurred because data was logged at a constant temporal interval, but the rate at which \gsw{} declined was not constant.
+Thinning reduces parameter estimation bias toward densely sampled regions of the curve which may not be the most biologically informative.
+
+*Procedure*: We retained the maxima and minima \gsw{} for each curve and thinned all but one point per thinning interval of $`r thinning_interval`~\log(\si{\mol\raiseto{-2}\meter\raiseto{-1}\second})$, retaining the point nearest the midpoint of the interval.
+
+#### Trim extreme \aax{} values
+
+*Rationale*: Extreme \aax{} values were likely due to measurement error or leaf damage.
+Since amphi and pseudohypo \agcurve{} curves are measured on consecutive days, a poor calibration or a damaged leaf could cause a large difference in $A$ between days, which would appear as an extreme \aax{} value.
+
+*Procedure*: We privionsally estimated \aax{} for each replicate by integrating over the range of \gsw{} values where amphi and pseudohypo \agcurve{} curves overlap.
+In this procedure, curve parameters were provisionally estimated using ordinary least squares with the `lm()` function in *R*.
+We then used point estimates of \aax{} for each replicate as the response variable in a linear model with light treatment, light intensity, accession, and all interactions as explanatory variables.
+This model was also fit using ordinary least squares with the `lm()` function in *R*.
+We classified  extreme \aax{} values as those with an absolute internal studentized residual $> `r aa_outlier_threshold`$.
+Because these values likely indicate significant measurement error or leaf damage, we removed \agcurve{} curves at both light intensities if either was classified as extreme.
 

### 2024-04-03 ms/ms.tex
finished drafting data cleaning methods

@@
 \item
+  thin redundant data points within each \agcurve{} curve
+\item
   remove replicates with no overlap between amphi and pseudohypo
@@
 \item
-  remove replicates with outlier \aax values
+  Trim extreme \aax{} values
 \end{enumerate}
+
+\textbf{Table showing number of curves and points before and after each
+cleaning step}
 
@@
 
-\emph{Procedure}: For each
+\emph{Procedure}: For each curve, we removed data points after \gsw{}
+had reached its minimum unless there were fewer than 10 data points
+remaining.
+
+\paragraph{\texorpdfstring{Remove outliers within each \agcurve{}
+curve}{Remove outliers within each  curve}}\label{remove-outliers-within-each-curve}
+
+\emph{Rationale}: Individual outliers within \agcurve{} curves, usually
+caused by transitory changes in chamber conditions, exert undue leverage
+on parameter estimates and cause bias and/or low precision in parameter
+estimates.
+
+\emph{Procedure}: We fit provisional quadratic regressions for each
+curve using ordinary least squares with the \texttt{lm()} function in
+\emph{R}. We sequentially removed data points with an absolute external
+studentized residual greater than until none remained.
+
+\paragraph{\texorpdfstring{Remove replicates with no overlap between
+amphi and pseudohypo \agcurve{}
+curves}{Remove replicates with no overlap between amphi and pseudohypo  curves}}\label{remove-replicates-with-no-overlap-between-amphi-and-pseudohypo-curves}
+
+\emph{Rationale}: We could not estimate \aax{} for replicates where
+amphi and pseudohypo \agcurve{} curves did not overlap.
+
+\emph{Procedure}: We removed replicates where the range of \gsw{} values
+for amphi and pseudohypo \agcurve{} curves did not overlap.
+
+\paragraph{\texorpdfstring{Thin redundant data points within each
+\agcurve{}
+curve}{Thin redundant data points within each  curve}}\label{thin-redundant-data-points-within-each-curve}
+
+\emph{Rationale}: Data points closely spaced along the \agcurve{} curve
+provide redundant information and may be highly correlated
+(i.e.~pseudoreplication). This occurred because data was logged at a
+constant temporal interval, but the rate at which \gsw{} declined was
+not constant. Thinning reduces parameter estimation bias toward densely
+sampled regions of the curve which may not be the most biologically
+informative.
+
+\emph{Procedure}: We retained the maxima and minima \gsw{} for each
+curve and thinned all but one point per thinning interval of
+\(0.05~\log(\si{\mol\raiseto{-2}\meter\raiseto{-1}\second})\), retaining
+the point nearest the midpoint of the interval.
+
+\paragraph{\texorpdfstring{Trim extreme \aax{}
+values}{Trim extreme  values}}\label{trim-extreme-values}
+
+\emph{Rationale}: Extreme \aax{} values were likely due to measurement
+error or leaf damage. Since amphi and pseudohypo \agcurve{} curves are
+measured on consecutive days, a poor calibration or a damaged leaf could
+cause a large difference in \(A\) between days, which would appear as an
+extreme \aax{} value.
+
+\emph{Procedure}: We privionsally estimated \aax{} for each replicate by
+integrating over the range of \gsw{} values where amphi and pseudohypo
+\agcurve{} curves overlap. In this procedure, curve parameters were
+provisionally estimated using ordinary least squares with the
+\texttt{lm()} function in \emph{R}. We then used point estimates of
+\aax{} for each replicate as the response variable in a linear model
+with light treatment, light intensity, accession, and all interactions
+as explanatory variables. This model was also fit using ordinary least
+squares with the \texttt{lm()} function in \emph{R}. We classified
+extreme \aax{} values as those with an absolute internal studentized
+residual \(> 3.5\). Because these values likely indicate significant
+measurement error or leaf damage, we removed \agcurve{} curves at both
+light intensities if either was classified as extreme.
 

### 2024-04-03 r/001_thin-data.R

@@ -1,12 +1,11 @@
 # Thin data to reduce oversampling small regions
 source("r/header.R")
 
-rh_curves = read_rds("data/rh_curves.rds") |>
-  filter(assumed_K == 0.5)
+rh_curves = read_rds("data/rh_curves.rds")
 
 thinned_rh_curves = rh_curves |>
   split( ~ acc_id + curve_type + light_intensity) |>
-  map_dfr(thin_data, bin_width = 0.05, .progress = TRUE)
+  map_dfr(thin_data, bin_width = aa_args$thinning_interval, .progress = TRUE)
 
 # stats on thinning
 thinning_summary = thinned_rh_curves |>

### 2024-04-03 r/003_trim-outliers.R

@@ -32,7 +32,7 @@ fit1 = lm(aa ~ light_treatment * light_intensity * acc, data = rh_curves1)
 rh_curves1$resid = rstudent(fit1)
 
 acc_id_outlier = rh_curves1 |>
-  filter(abs(resid) > aa_outlier_threshold) |>
+  filter(abs(resid) > aa_pars$aa_outlier_threshold) |>
   select(acc_id, aa, resid) |>
   pull(acc_id)
 

### 2024-04-03 r/header.R

@@ -31,5 +31,11 @@ id_string = "^(LA[0-9]{4}A*|nelsonii|sandwicense)-[A-Z]{1}[A]{0,1}$"
 
 theme_set(theme_cowplot())
 
-# options (could this be a separate file?)
-aa_outlier_threshold = 3.5
+# options 
+aa_args = read_rds("objects/aa_args.rds") |>
+  c(
+    aa_outlier_threshold = 3.5,
+    thinning_interval = 0.05
+  )
+
+write_rds(aa_args, "objects/aa_args1.rds")

### 2024-04-03 ms/ms.qmd
drafted tbl-cleaning

@@
 header-includes: |
+  \usepackage{hyperref}
   \addtokomafont{disposition}{\rmfamily}
@@
 library(readr)
+library(tibble)
 # I am mostly putting placeholders here for objects/values that will need to be
@@
 aa_args = read_rds("../objects/aa_args1.rds")
+aa_stats = read_rds("../objects/aa_stats.rds")
 fit_aa = read_rds("../objects/fit_dummy.rds")
@@
 lycopersicum* var.
-*lycopersicum* (tab:accessions).
+*lycopersicum* \autoref{tab:accessions}.
 Due to constraints on growth space and time, we spread out measurements over `r n_weeks` weeks from `r start_licor_date` to `r end_licor_date`.
@@
 The raw data set consisted of `r n_rh_curve_raw` \agcurve{} curves with an average of `r n_points_per_rh_curves_raw`.
-Principled, consistent, manual curation of a data set this size is not feasible.
+Manual curation of a data set this size in a principled, consistent manner is not feasible.
 Therefore, we automated data cleaning using custom *R* scripts.
-Cleaning is divided into XXX sequential steps:
+Cleaning is divided into six sequential steps \autoref{tbl-cleaning}.
 
-1. remove unreliable data points
-2. remove hysteretic portion of \agcurve{} curves at low \gsw{}
-3. remove outliers within each \agcurve{} curve
-4. thin redundant data points within each \agcurve{} curve
-5. remove replicates with no overlap between amphi and pseudohypo \agcurve{} curves
-6. Trim extreme \aax{} values
+```{r, cleaning, echo=FALSE,message=FALSE,warning=FALSE}
+#| label: tbl-cleaning
+#| tbl-cap: "Six sequential steps for cleaning \\agcurve{} curves.
+The rationale and procedure for each step are described in the text.
+The rightmost columns summarize the number of curves and mean number of points per curve remaining after each step.
+For reference, there are four possible \\agcurve{} curves per replicate: all combinations of leaf type (amphi or pseudohypo) and light intensity (high or low)."
 
-**Table showing number of curves and points before and after each cleaning step**
+step_description_col = c(
+  "1. remove unreliable and unusable data points",
+  "2. remove hysteretic portion of \\agcurve{} curves at low \\gsw{}",
+  "3. remove outliers within each \\agcurve{} curve",
+  "4. thin redundant data points within each \\agcurve{} curve",
+  "5. remove replicates with no overlap between amphi and pseudohypo \\agcurve{} curves",
+  "6. Trim extreme \\aax{} values"
+)
 
-#### Remove unreliable data points
+tibble(
+  `Step: description` = step_description_col,
+  `Number of curves` = 100 + c(1,2,3,4,5,6),
+  `Number of points per curve` = c(10.1,2,3,4,5,6)
+) |>
+  knitr::kable(format = "latex",
+               align = c("l", "c", "c"),
+               digits = c(0, 0, 1),
+               booktabs = TRUE, escape = FALSE) |>
+  kableExtra::kable_styling(latex_options = c("striped", "no_vertical_lines")) |>
+  kableExtra::column_spec(1, width = "9cm") |>
+  kableExtra::column_spec(2, width = "2.5cm") |>
+  kableExtra::column_spec(3, width = "2.5cm")
+
+```
+
+#### Remove unreliable and unusable data points
 
 *Rationale*: Unreliable data points consisted of those where chamber [CO$_2$] was unstable and therefore measurements are not biologically meaningful.
+Unusable data points were those where $A < 0$ because the logarithm of a negative number is undefined.
 
-*Procedure*: We retained data points where \cabetween{`r ca_min`}{`r ca_max`}.
+*Procedure*: We retained data points where \cabetween{`r ca_min`}{`r ca_max`} and $A> 0$.
 

### 2024-04-03 ms/test.qmd
drafted tbl-cleaning

@@
 header-includes: |
+  \usepackage{hyperref}
   \addtokomafont{disposition}{\rmfamily}
   \usepackage{siunitx}
-  \newcommand{\gsc}{$g_\text{sc}$}
-  \newcommand{\ppfd}[1]{$\mathrm{PPFD} = \qty{#1}{\micro\mol\raiseto{-2}\meter\raiseto{-1}\second}$}
+  \newcommand{\aax}{$\mathrm{AA}$}
+  \newcommand{\Aamphi}{$A_{\mathrm{amphi}}$}
+  \newcommand{\Ahypo}{$A_{\mathrm{hypo}}$}
+  \newcommand{\agcurve}{$A \textendash g_\text{sw}$}
+  \newcommand{\ca}{$C_\mathrm{a}$}
+  \newcommand{\caequals}[1]{$C_\mathrm{a} = \qty{#1}{\micro\mol\raiseto{-1}\mol}$}
+  \newcommand{\cabetween}[2]{#1 < $C_\mathrm{a} < \qty{#2}{\micro\mol\raiseto{-1}\mol}$}
+  \newcommand{\gsw}{$g_\text{sw}$}
+  \newcommand{\ppfd}{$\mathrm{PPFD}$}
+  \newcommand{\ppfdequals}[1]{$\mathrm{PPFD} = \qty{#1}{\micro\mol\raiseto{-2}\meter\raiseto{-1}\second}$}
+  \newcommand{\rh}{$\mathrm{RH}$}
+  \newcommand{\rhequals}[1]{$\mathrm{RH} = #1\%$}
+  \newcommand{\tleafequals}[1]{$T_\mathrm{leaf} = \qty{#1}{\degreeCelsius}$}
 mainfont: Times New Roman
-csl: https://www.zotero.org/styles/american-journal-of-botany
+bibliography: solanum-aa.bib
+csl: https://www.zotero.org/styles/science
 ---
 
-words \gsc
+```{r load-objects, echo=FALSE, eval=TRUE}
 
-this \ppfd{2000}
+library(readr)
+library(tibble)
+```
+
+Manual curation of a data set this size in a principled, consistent manner is not feasible.
+Therefore, we automated data cleaning using custom *R* scripts.
+Cleaning is divided into six sequential steps \autoref{tbl-cleaning}.
+
+```{r, cleaning, echo=FALSE,message=FALSE,warning=FALSE}
+#| label: tbl-cleaning
+#| tbl-cap: "Six sequential steps for cleaning \\agcurve{} curves.
+The rationale and procedure for each step are described in the text.
+The rightmost columns summarize the number of curves and mean number of points per curve remaining after each step.
+For reference, there are four possible \\agcurve{} curves per replicate: all combinations of leaf type (amphi or pseudohypo) and light intensity (high or low)."
+
+step_description_col = c(
+  "1. remove unreliable data points",
+  "2. remove hysteretic portion of \\agcurve{} curves at low \\gsw{}",
+  "3. remove outliers within each \\agcurve{} curve",
+  "4. thin redundant data points within each \\agcurve{} curve",
+  "5. remove replicates with no overlap between amphi and pseudohypo \\agcurve{} curves",
+  "6. Trim extreme \\aax{} values"
+)
+
+tibble(
+  `Step: description` = step_description_col,
+  `Number of curves` = 100 + c(1,2,3,4,5,6),
+  `Number of points per curve` = c(10.1,2,3,4,5,6)
+) |>
+  knitr::kable(format = "latex",
+               align = c("l", "c", "c"),
+               digits = c(0, 0, 1),
+               booktabs = TRUE, escape = FALSE) |>
+  kableExtra::kable_styling(latex_options = c("striped", "no_vertical_lines")) |>
+  kableExtra::column_spec(1, width = "9cm") |>
+  kableExtra::column_spec(2, width = "2.5cm") |>
+  kableExtra::column_spec(3, width = "2.5cm")
+
+```

### 2024-04-03 r/000_update-data.R

@@ -13,7 +13,7 @@ c("processed-data/accession-climate",
     )
   })
 
-c("objects/aa_args") |>
+c("objects/aa_args", "objects/aa_stats.rds") |>
   walk(\(.x) {
     file.copy(
       glue("../../data/adaptive-amphistomy/{.x}.rds"),

### 2024-04-03 ms/ms.qmd
updated rh cleaning table

@@
   \newcommand{\agcurve}{$A \textendash g_\text{sw}$}
+  \newcommand{\aqcurve}{$A \textendash Q$}
   \newcommand{\ca}{$C_\mathrm{a}$}
@@
   \newcommand{\ppfdequals}[1]{$\mathrm{PPFD} = \qty{#1}{\micro\mol\raiseto{-2}\meter\raiseto{-1}\second}$}
+  \newcommand{\ppfdqty}[1]{$\qty{#1}{\micro\mol\raiseto{-2}\meter\raiseto{-1}\second}$}
   \newcommand{\rh}{$\mathrm{RH}$}
@@
 library(readr)
+library(stringr)
 library(tibble)
+
 # I am mostly putting placeholders here for objects/values that will need to be
@@
 aa_stats = read_rds("../objects/aa_stats.rds")
+# sort list by name
+aa_stats = aa_stats[order(names(aa_stats))]
+
 fit_aa = read_rds("../objects/fit_dummy.rds")
@@
 We assigned the sun treatment to the lower rack to limit diffuse light from reaching the shade treatment.
-The average daytime \ppfd{} was \ppfdequals{`r ppfd_sun`} ($\qty{`r ppfd_sun`}{\micro\mol\raiseto{-2}\meter\raiseto{-1}\second}$) and \ppfdequals{`r ppfd_shade`} ($\qty{`r ppfd_shade`}{\micro\mol\raiseto{-2}\meter\raiseto{-1}\second}$) for sun and shade treatments, respectively.
+The average daytime \ppfd{} was \ppfdqty{`r ppfd_sun`} and \ppfdqty{`r ppfd_shade`} for sun and shade treatments, respectively.
 To isolate the effect of light intensity from quality, we used the same LED model with the the same spectrum (Fluence SPYDR 2i, Austin, Texas, USAS), but dimmed the lights in the shade treatment.
@@
 
-#### Light-response curves
+#### Light-response (\aqcurve) curves
 
-In `r percent_aa_and_aq`\% of plants, we measured light-response curves on the same leaflets as \agcurve{} curves.
-However, when a leaflet was damaged during \agcurve{} curves, we used the next closest leaflet for light-response curves.
+In `r percent_aa_and_aq`\% of plants, we measured light-response (\aqcurve) curves on the same leaflets as \agcurve{} curves.
+However, when a leaflet was damaged during \agcurve{} curves, we used the next closest leaflet for \aqcurve{} curves.
 Leaves acclimated to high light-intensity (\ppfdequals{2000}), ambient CO$_2$ (\caequals{415}), \rhequals{50}, and \tleafequals{25}.
@@
 
-The raw data set consisted of `r n_rh_curve_raw` \agcurve{} curves with an average of `r n_points_per_rh_curves_raw`.
+The raw data set consisted of `r prettyNum(aa_stats$n_rh_curve0)` \agcurve{} curves with an average of `r formatC(aa_stats$n_point_per_rh_curve0, digits = 3)` points per curve.
 Manual curation of a data set this size in a principled, consistent manner is not feasible.
 Therefore, we automated data cleaning using custom *R* scripts.
-Cleaning is divided into six sequential steps \autoref{tbl-cleaning}.
+Cleaning is divided into six sequential steps \autoref{tbl-cleaning1}.
 
-```{r, cleaning, echo=FALSE,message=FALSE,warning=FALSE}
-#| label: tbl-cleaning
+```{r, cleaning1, echo=FALSE,message=FALSE,warning=FALSE}
+#| label: tbl-cleaning1
 #| tbl-cap: "Six sequential steps for cleaning \\agcurve{} curves.
@@
   "3. remove outliers within each \\agcurve{} curve",
-  "4. thin redundant data points within each \\agcurve{} curve",
-  "5. remove replicates with no overlap between amphi and pseudohypo \\agcurve{} curves",
+  "4. remove replicates with no overlap between amphi and pseudohypo \\agcurve{} curves",
+  "5. thin redundant data points within each \\agcurve{} curve",
   "6. Trim extreme \\aax{} values"
@@
 
+n_curve_col = unlist(aa_stats[str_detect(names(aa_stats), "^n_rh_curve[1-9]$")])
+
+n_point_col = unlist(aa_stats[str_detect(names(aa_stats), "^n_point_per_rh_curve[1-9]$")])
+
 tibble(
   `Step: description` = step_description_col,
-  `Number of curves` = 100 + c(1,2,3,4,5,6),
-  `Number of points per curve` = c(10.1,2,3,4,5,6)
+  `Number of curves` = n_curve_col,
+  `Number of points per curve` = n_point_col
 ) |>
@@
 
-#### Remove replicates with no overlap between amphi and pseudohypo \agcurve{} curves
-
-*Rationale*: We could not estimate \aax{} for replicates where amphi and pseudohypo \agcurve{} curves did not overlap.
-
-*Procedure*: We removed replicates where the range of \gsw{} values for amphi and pseudohypo \agcurve{} curves did not overlap.
-
 #### Thin redundant data points within each \agcurve{} curve
@@
 
+#### Remove replicates with no overlap between amphi and pseudohypo \agcurve{} curves
+
+*Rationale*: We could not estimate \aax{} for replicates where amphi and pseudohypo \agcurve{} curves did not overlap.
+
+*Procedure*: We removed replicates where the range of \gsw{} values for amphi and pseudohypo \agcurve{} curves did not overlap.
+
 #### Trim extreme \aax{} values
@@
 
-*Procedure*: We privionsally estimated \aax{} for each replicate by integrating over the range of \gsw{} values where amphi and pseudohypo \agcurve{} curves overlap.
+*Procedure*: We provionsally estimated \aax{} for each replicate by integrating over the range of \gsw{} values where amphi and pseudohypo \agcurve{} curves overlap.
 In this procedure, curve parameters were provisionally estimated using ordinary least squares with the `lm()` function in *R*.
@@
 Because these values likely indicate significant measurement error or leaf damage, we removed \agcurve{} curves at both light intensities if either was classified as extreme.
+
+### Cleaning \aqcurve{} curves
+
+WORKING HERE!
+
+The raw data set consisted of `r prettyNum(aa_stats$n_aq_curve0)` \aqcurve{} curves with an average of `r formatC(aa_stats$n_point_per_aq_curve0, digits = 3)`.
+Manual curation of a data set this size in a principled, consistent manner is not feasible.
+Therefore, we automated data cleaning using custom *R* scripts.
+Cleaning is divided into six sequential steps \autoref{tbl-cleaning2}.
+
+```{r, cleaning2, eval=FALSE,echo=FALSE,message=FALSE,warning=FALSE}
+#| label: tbl-cleaning2
+#| tbl-cap: "Six sequential steps for cleaning \\agcurve{} curves.
+The rationale and procedure for each step are described in the text.
+The rightmost columns summarize the number of curves and mean number of points per curve remaining after each step.
+For reference, there are four possible \\agcurve{} curves per replicate: all combinations of leaf type (amphi or pseudohypo) and light intensity (high or low)."
+
+step_description_col = c(
+  "1. remove unreliable and unusable data points",
+  "2. remove hysteretic portion of \\agcurve{} curves at low \\gsw{}",
+  "3. remove outliers within each \\agcurve{} curve",
+  "4. remove replicates with no overlap between amphi and pseudohypo \\agcurve{} curves",
+  "5. thin redundant data points within each \\agcurve{} curve",
+  "6. Trim extreme \\aax{} values"
+)
+
+n_curve_col = unlist(aa_stats[str_detect(names(aa_stats), "^n_rh_curve[0-9]$")])
+
+n_point_col = unlist(aa_stats[str_detect(names(aa_stats), "^n_point_per_rh_curve[0-9]$")])
+
+tibble(
+  `Step: description` = step_description_col,
+  `Number of curves` = n_curve_col,
+  `Number of points per curve` = n_point_col
+) |>
+  knitr::kable(format = "latex",
+               align = c("l", "c", "c"),
+               digits = c(0, 0, 1),
+               booktabs = TRUE, escape = FALSE) |>
+  kableExtra::kable_styling(latex_options = c("striped", "no_vertical_lines")) |>
+  kableExtra::column_spec(1, width = "9cm") |>
+  kableExtra::column_spec(2, width = "2.5cm") |>
+  kableExtra::column_spec(3, width = "2.5cm")
+
+```
 

### 2024-04-03 ms/ms.tex
updated rh cleaning table

@@
 
+\usepackage{booktabs}
+\usepackage{longtable}
+\usepackage{array}
+\usepackage{multirow}
+\usepackage{wrapfig}
+\usepackage{float}
+\usepackage{colortbl}
+\usepackage{pdflscape}
+\usepackage{tabu}
+\usepackage{threeparttable}
+\usepackage{threeparttablex}
+\usepackage[normalem]{ulem}
+\usepackage{makecell}
+\usepackage{xcolor}
 \KOMAoption{captions}{tableheading}
+\usepackage{hyperref}
 \addtokomafont{disposition}{\rmfamily}
@@
 \newcommand{\agcurve}{$A \textendash g_\text{sw}$}
+\newcommand{\aqcurve}{$A \textendash Q$}
 \newcommand{\ca}{$C_\mathrm{a}$}
@@
 \newcommand{\ppfdequals}[1]{$\mathrm{PPFD} = \qty{#1}{\micro\mol\raiseto{-2}\meter\raiseto{-1}\second}$}
+\newcommand{\ppfdqty}[1]{$\qty{#1}{\micro\mol\raiseto{-2}\meter\raiseto{-1}\second}$}
 \newcommand{\rh}{$\mathrm{RH}$}
@@
 and the cultivated tomato \emph{S. lycopersicum} var.
-\emph{lycopersicum} (tab:accessions). Due to constraints on growth space
-and time, we spread out measurements over 80 weeks from May 1, 2022 to
-October 31, 2023. Replicates within accession were evenly spread out
-over this period to prevent confounding of temporal variation in growth
-conditions with accession. {[}anything else to say here? maybe explain
-accession selection and phylogeny?{]}
+\emph{lycopersicum} \autoref{tab:accessions}. Due to constraints on
+growth space and time, we spread out measurements over 80 weeks from May
+1, 2022 to October 31, 2023. Replicates within accession were evenly
+spread out over this period to prevent confounding of temporal variation
+in growth conditions with accession. {[}anything else to say here? maybe
+explain accession selection and phylogeny?{]}
 
@@
 lower rack to limit diffuse light from reaching the shade treatment. The
-average daytime \ppfd{} was \ppfdequals{800}
-(\(\qty{800}{\micro\mol\raiseto{-2}\meter\raiseto{-1}\second}\)) and
-\ppfdequals{125}
-(\(\qty{125}{\micro\mol\raiseto{-2}\meter\raiseto{-1}\second}\)) for sun
-and shade treatments, respectively. To isolate the effect of light
-intensity from quality, we used the same LED model with the the same
-spectrum (Fluence SPYDR 2i, Austin, Texas, USAS), but dimmed the lights
-in the shade treatment. To maintain homogeneous environmental conditions
-other than light, we mixed air within the growth room using an air
-circulator (Vornado 693DC, Andover, Kansas, USA) and within racks using
-a miniature oscillating air circulator (Vornado Atom 1, Andover, Kansas,
-USA). Despite these efforts, the air in the sun treatment was on average
+average daytime \ppfd{} was \ppfdqty{800} and \ppfdqty{125} for sun and
+shade treatments, respectively. To isolate the effect of light intensity
+from quality, we used the same LED model with the the same spectrum
+(Fluence SPYDR 2i, Austin, Texas, USAS), but dimmed the lights in the
+shade treatment. To maintain homogeneous environmental conditions other
+than light, we mixed air within the growth room using an air circulator
+(Vornado 693DC, Andover, Kansas, USA) and within racks using a miniature
+oscillating air circulator (Vornado Atom 1, Andover, Kansas, USA).
+Despite these efforts, the air in the sun treatment was on average
 \(\qty{1}{\degreeCelsius}\) warmer and the average RH was consequently 5
@@
 
-\paragraph{Light-response curves}\label{light-response-curves}
+\paragraph{\texorpdfstring{Light-response (\aqcurve)
+curves}{Light-response () curves}}\label{light-response-curves}
 
-In 90.0\% of plants, we measured light-response curves on the same
-leaflets as \agcurve{} curves. However, when a leaflet was damaged
+In 90.0\% of plants, we measured light-response (\aqcurve) curves on the
+same leaflets as \agcurve{} curves. However, when a leaflet was damaged
 during \agcurve{} curves, we used the next closest leaflet for
-light-response curves. Leaves acclimated to high light-intensity
+\aqcurve{} curves. Leaves acclimated to high light-intensity
 (\ppfdequals{2000}), ambient CO\(_2\) (\caequals{415}), \rhequals{50},
@@
 
-The raw data set consisted of 2400 \agcurve{} curves with an average of
-100. Principled, consistent, manual curation of a data set this size is
-not feasible. Therefore, we automated data cleaning using custom
-\emph{R} scripts. Cleaning is divided into XXX sequential steps:
+The raw data set consisted of 2370 \agcurve{} curves with an average of
+63.2. Manual curation of a data set this size in a principled,
+consistent manner is not feasible. Therefore, we automated data cleaning
+using custom \emph{R} scripts. Cleaning is divided into six sequential
+steps \autoref{tbl-cleaning1}.
 
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\tightlist
-\item
-  remove unreliable data points
-\item
-  remove hysteretic portion of \agcurve{} curves at low \gsw{}
-\item
-  remove outliers within each \agcurve{} curve
-\item
-  thin redundant data points within each \agcurve{} curve
-\item
-  remove replicates with no overlap between amphi and pseudohypo
-  \agcurve{} curves
-\item
-  Trim extreme \aax{} values
-\end{enumerate}
+\begin{table}
 
-\textbf{Table showing number of curves and points before and after each
-cleaning step}
+\caption{\label{tbl-cleaning1}Six sequential steps for cleaning
+\agcurve{} curves. The rationale and procedure for each step are
+described in the text. The rightmost columns summarize the number of
+curves and mean number of points per curve remaining after each step.
+For reference, there are four possible \agcurve{} curves per replicate:
+all combinations of leaf type (amphi or pseudohypo) and light intensity
+(high or low).}
 
-\paragraph{Remove unreliable data
-points}\label{remove-unreliable-data-points}
+\centering{
+
+\centering
+\begin{tabular}{>{\raggedright\arraybackslash}p{9cm}>{\centering\arraybackslash}p{2.5cm}>{\centering\arraybackslash}p{2.5cm}}
+\toprule
+Step: description & Number of curves & Number of points per curve\\
+\midrule
+\cellcolor{gray!10}{1. remove unreliable and unusable data points} & \cellcolor{gray!10}{2361} & \cellcolor{gray!10}{63.0}\\
+2. remove hysteretic portion of \agcurve{} curves at low \gsw{} & 2360 & 59.2\\
+\cellcolor{gray!10}{3. remove outliers within each \agcurve{} curve} & \cellcolor{gray!10}{2360} & \cellcolor{gray!10}{58.7}\\
+4. remove replicates with no overlap between amphi and pseudohypo \agcurve{} curves & 2268 & 58.5\\
+\cellcolor{gray!10}{5. thin redundant data points within each \agcurve{} curve} & \cellcolor{gray!10}{2268} & \cellcolor{gray!10}{28.1}\\
+\addlinespace
+6. Trim extreme \aax{} values & 2210 & 28.0\\
+\bottomrule
+\end{tabular}
+
+}
+
+\end{table}%
+
+\paragraph{Remove unreliable and unusable data
+points}\label{remove-unreliable-and-unusable-data-points}
 
@@
 chamber {[}CO\(_2\){]} was unstable and therefore measurements are not
-biologically meaningful.
+biologically meaningful. Unusable data points were those where \(A < 0\)
+because the logarithm of a negative number is undefined.
 
-\emph{Procedure}: We retained data points where \cabetween{410}{420}.
+\emph{Procedure}: We retained data points where \cabetween{410}{420} and
+\(A> 0\).
 
@@
 \emph{R}. We sequentially removed data points with an absolute external
-studentized residual greater than until none remained.
-
-\paragraph{\texorpdfstring{Remove replicates with no overlap between
-amphi and pseudohypo \agcurve{}
-curves}{Remove replicates with no overlap between amphi and pseudohypo  curves}}\label{remove-replicates-with-no-overlap-between-amphi-and-pseudohypo-curves}
-
-\emph{Rationale}: We could not estimate \aax{} for replicates where
-amphi and pseudohypo \agcurve{} curves did not overlap.
-
-\emph{Procedure}: We removed replicates where the range of \gsw{} values
-for amphi and pseudohypo \agcurve{} curves did not overlap.
+studentized residual greater than 3 until none remained.
 
@@
 
+\paragraph{\texorpdfstring{Remove replicates with no overlap between
+amphi and pseudohypo \agcurve{}
+curves}{Remove replicates with no overlap between amphi and pseudohypo  curves}}\label{remove-replicates-with-no-overlap-between-amphi-and-pseudohypo-curves}
+
+\emph{Rationale}: We could not estimate \aax{} for replicates where
+amphi and pseudohypo \agcurve{} curves did not overlap.
+
+\emph{Procedure}: We removed replicates where the range of \gsw{} values
+for amphi and pseudohypo \agcurve{} curves did not overlap.
+
 \paragraph{\texorpdfstring{Trim extreme \aax{}
@@
 
-\emph{Procedure}: We privionsally estimated \aax{} for each replicate by
+\emph{Procedure}: We provionsally estimated \aax{} for each replicate by
 integrating over the range of \gsw{} values where amphi and pseudohypo
@@
 light intensities if either was classified as extreme.
+
+\subsubsection{\texorpdfstring{Cleaning \aqcurve{}
+curves}{Cleaning  curves}}\label{cleaning-curves-1}
+
+WORKING HERE!
+
+The raw data set consisted of \aqcurve{} curves with an average of .
+Manual curation of a data set this size in a principled, consistent
+manner is not feasible. Therefore, we automated data cleaning using
+custom \emph{R} scripts. Cleaning is divided into six sequential steps
+\autoref{tbl-cleaning2}.
+
+\begin{table}
+
+\caption{\label{tbl-cleaning2}Six sequential steps for cleaning
+\agcurve{} curves. The rationale and procedure for each step are
+described in the text. The rightmost columns summarize the number of
+curves and mean number of points per curve remaining after each step.
+For reference, there are four possible \agcurve{} curves per replicate:
+all combinations of leaf type (amphi or pseudohypo) and light intensity
+(high or low).}
+
+\centering{
+
+}
+
+\end{table}%
 

### 2024-04-03 r/000_update-data.R

@@ -9,15 +9,15 @@ c("processed-data/accession-climate",
     file.copy(
       glue("../../data/adaptive-amphistomy/{.x}.rds"),
       glue("data/{.y}.rds", .y = str_remove(.x, "(process|filter)ed-data/")),
-      overwrite = FALSE
+      overwrite = TRUE
     )
   })
 
-c("objects/aa_args", "objects/aa_stats.rds") |>
+c("objects/aa_args", "objects/aa_stats") |>
   walk(\(.x) {
     file.copy(
       glue("../../data/adaptive-amphistomy/{.x}.rds"),
       glue("{.x}.rds"),
-      overwrite = FALSE
+      overwrite = TRUE
     )
   })

### 2024-04-03 r/001_thin-data.R

@@ -16,9 +16,21 @@ thinning_summary = thinned_rh_curves |>
   ) |>
   mutate(percent_thin = 100 * (1 - n_keep / n_total))
 
-# ggplot(thinning_summary, aes(percent_thin)) +
-#   facet_grid(vars(light_intensity, light_treatment), vars(curve_type)) +
-#   geom_histogram()
+# add stats on data cleaning
+do.call(
+  "add_to_stats",
+  thinned_rh_curves |>
+    filter(keep) |>
+    dplyr::summarize(
+      n_point = length(obs),
+      .by = c("acc_id", "curve_type", "light_intensity")
+    ) |>
+    dplyr::summarize(
+      n_rh_curve5 = n(),
+      n_point_per_rh_curve5 = mean(n_point)
+    ) |>
+    as.list()
+)
 
 thinned_rh_curves |>
   filter(keep) |>

### 2024-04-03 r/002_prepare-data.R

@@ -9,12 +9,9 @@ read_rds("data/thinned_rh_curves.rds") |>
       curve_type == "2-sided RH" ~ "amphi"
     )
   ) |>
-  separate_wider_delim(acc_id,
-                       "-",
-                       names = c("acc", "id"),
-                       cols_remove = FALSE) |>
+  # this step is doing in adaptive-amphistomy/10_filter-rh-curves.R
   # filter out Hawaiian, cultivated Solanum, and LA3909 (not enough reps)
-  dplyr::filter(str_detect(acc, "^LA[0-9]{4}"), acc != "LA3475", acc != "LA3909") |>
+  # dplyr::filter(str_detect(acc, "^LA[0-9]{4}"), acc != "LA3475", acc != "LA3909") |>
   # make unique ID for each leaf_type within acc_id
   unite("lightintensity_x_acc_id", acc_id, light_intensity, remove = FALSE) |>
   # make unique ID for each curve

### 2024-04-03 r/003_trim-outliers.R

@@ -32,9 +32,24 @@ fit1 = lm(aa ~ light_treatment * light_intensity * acc, data = rh_curves1)
 rh_curves1$resid = rstudent(fit1)
 
 acc_id_outlier = rh_curves1 |>
-  filter(abs(resid) > aa_pars$aa_outlier_threshold) |>
+  filter(abs(resid) > aa_args$aa_outlier_threshold) |>
   select(acc_id, aa, resid) |>
   pull(acc_id)
+
+do.call(
+  "add_to_stats",
+  rh_curves |>
+    filter(!(acc_id %in% acc_id_outlier)) |>
+    dplyr::summarize(
+      n_point = length(obs),
+      .by = c("acc_id", "curve_type", "light_intensity")
+    ) |>
+    dplyr::summarize(
+      n_rh_curve6 = n(),
+      n_point_per_rh_curve6 = mean(n_point)
+    ) |>
+    as.list()
+)
 
 rh_curves |> 
   filter(!(acc_id %in% acc_id_outlier)) |>

### 2024-04-03 r/304_fit-data.R

@@ -2,30 +2,38 @@
 # This version runs locally; I have other code for running on HPC
 source("r/header.R")
 
-m = "aa2"
+args = commandArgs(trailingOnly = TRUE)
+# args = list("aa1", "0")
+
+m = args[1]
+chain = args[2]
 
 chkpt_path = glue("checkpoints/chkpt_folder_{m}")
 
 # Only untar if path doesn't exist
 if (!dir.exists(chkpt_path)) {
   untar(paste0(chkpt_path, ".tar"))
-  file.remove(paste0(chkpt_path, ".tar"))
+  options(cmdstanr_force_recompile = TRUE)
+  # file.remove(paste0(chkpt_path, ".tar"))
 }
 
-dir.create("checkpoints/another_name")
-untar(paste0(chkpt_path, ".tar"), "checkpoints/another_name")
+chkpt_path1 = glue("checkpoints/chkpt_folder_{m}_{chain}")
+
+if (!dir.exists(chkpt_path1)) {
+  R.utils::copyDirectory(from = chkpt_path, to = chkpt_path1)
+}
 
 fit_m = chkpt_stan1(
-  chkpt_path,
-  iter_warmup = 40, #4e3,
-  iter_sampling = 40, #4e3,
-  thin = 1, #4e0,
-  iter_per_chkpt = 10, #2e0,
+  chkpt_path1,
+  iter_warmup = 4e3,
+  iter_sampling = 4e3,
+  thin = 4e0,
+  iter_per_chkpt = 2e2,
   chkpt_progress = TRUE,
-  max_treedepth = 10L #12L
+  max_treedepth = 12L
 )
 
-draws = combine_chkpt_draws1(path = chkpt_path)
+draws = combine_chkpt_draws1(path = chkpt_path1)
 
-write_rds(draws, glue("objects/draws_{m}.rds"))
-tar(paste0(chkpt_path, ".tar"), chkpt_path)
+write_rds(draws, glue("objects/draws_{m}_{chain}.rds"))
+tar(paste0(chkpt_path1, ".tar"), chkpt_path1)

### 2024-04-03 r/functions.R

@@ -1,4 +1,12 @@
 # 0. Import data ----
+add_to_stats = function(...) {
+  old = read_rds("objects/aa_stats.rds")
+  new = list(...)
+  if (any(names(new) %in% names(old))) {
+    old[intersect(names(old), names(new))] = NULL
+  }
+  write_rds(c(old, new), "objects/aa_stats.rds")
+}
 
 ## Function to thin data before analysis
 thin_data = function(.d, bin_width, min_n = 20L) {

### 2024-04-04 ms/ms.qmd
drafted light response cleaning methods and table

@@
 thinning_interval = aa_args$thinning_interval
+aq_r2_threshold = aa_args$aq_r2_threshold
 
-loo_version = "2.7.0"
 ```
@@
 
-The raw data set consisted of `r prettyNum(aa_stats$n_rh_curve0)` \agcurve{} curves with an average of `r formatC(aa_stats$n_point_per_rh_curve0, digits = 3)` points per curve.
+The raw data set consisted of `r prettyNum(aa_stats$n_rh_curve0, big.mark = ",")` \agcurve{} curves with an average of `r formatC(aa_stats$n_point_per_rh_curve0, digits = 3)` points per curve.
 Manual curation of a data set this size in a principled, consistent manner is not feasible.
@@
 
-n_curve_col = unlist(aa_stats[str_detect(names(aa_stats), "^n_rh_curve[1-9]$")])
+n_curve_col = unlist(aa_stats[str_detect(names(aa_stats), "^n_rh_curve[1-9]$")]) |>
+  prettyNum(big.mark = ",")
 
@@
 *Procedure*: We fit provisional quadratic regressions for each curve using ordinary least squares with the `lm()` function in *R*.
-We sequentially removed data points with an absolute external studentized residual greater than `r point_outlier_threshold` until none remained.
+We sequentially removed data points with an absolute external studentized residual $> `r point_outlier_threshold`$ until none remained.
 
@@
 
-WORKING HERE!
-
-The raw data set consisted of `r prettyNum(aa_stats$n_aq_curve0)` \aqcurve{} curves with an average of `r formatC(aa_stats$n_point_per_aq_curve0, digits = 3)`.
+The raw data set consisted of `r prettyNum(aa_stats$n_aq_curve0, big.mark = ",")` \aqcurve{} curves with an average of `r formatC(aa_stats$n_point_per_aq_curve0, digits = 3)`.
 Manual curation of a data set this size in a principled, consistent manner is not feasible.
 Therefore, we automated data cleaning using custom *R* scripts.
-Cleaning is divided into six sequential steps \autoref{tbl-cleaning2}.
+Cleaning is divided into two sequential steps \autoref{tbl-cleaning2}.
 
-```{r, cleaning2, eval=FALSE,echo=FALSE,message=FALSE,warning=FALSE}
+```{r, cleaning2,echo=FALSE,message=FALSE,warning=FALSE}
 #| label: tbl-cleaning2
-#| tbl-cap: "Six sequential steps for cleaning \\agcurve{} curves.
+#| tbl-cap: "Two sequential steps for cleaning \\agcurve{} curves.
 The rationale and procedure for each step are described in the text.
-The rightmost columns summarize the number of curves and mean number of points per curve remaining after each step.
-For reference, there are four possible \\agcurve{} curves per replicate: all combinations of leaf type (amphi or pseudohypo) and light intensity (high or low)."
+The rightmost columns summarize the number of curves and mean number of points per curve remaining after each step."
 
 step_description_col = c(
-  "1. remove unreliable and unusable data points",
-  "2. remove hysteretic portion of \\agcurve{} curves at low \\gsw{}",
-  "3. remove outliers within each \\agcurve{} curve",
-  "4. remove replicates with no overlap between amphi and pseudohypo \\agcurve{} curves",
-  "5. thin redundant data points within each \\agcurve{} curve",
-  "6. Trim extreme \\aax{} values"
+  "1. remove outliers within each \\aqcurve{} curve",
+  "2. remove \\aqcurve{} curves with poor fit"
 )
 
-n_curve_col = unlist(aa_stats[str_detect(names(aa_stats), "^n_rh_curve[0-9]$")])
+n_curve_col = unlist(aa_stats[str_detect(names(aa_stats), "^n_aq_curve[1-9]$")]) |>
+  prettyNum(big.mark = ",")
 
-n_point_col = unlist(aa_stats[str_detect(names(aa_stats), "^n_point_per_rh_curve[0-9]$")])
+n_point_col = unlist(aa_stats[str_detect(names(aa_stats), "^n_point_per_aq_curve[1-9]$")])
 
@@
 
+#### Remove outliers within each \aqcurve{} curve
+
+*Rationale*: Individual outliers within \agcurve{} curves, usually caused by transitory changes in chamber conditions, exert undue leverage on parameter estimates and cause bias and/or low precision in parameter estimates.
+
+*Procedure*: We fit provisional nonrectangular hyperbola [@marshall_model_1980] to each \aqcurve{} curve using nonlinear regression with the `nlsLM()` function from the *R* package **minpack.lm** version `r packageVersion("minpack.lm")` [@Elzhov_minpack.lm_2023].
+We sequentially removed data points with an absolute external studentized residual $> `r point_outlier_threshold`$ until none remained.
+
+#### Remove \aqcurve{} curves with poor fit
+
+*Rationale*: \aqcurve{} curves with a poor fit to the nonrectangular hyperbola most likely indicate systematic measurement error and/or the leaf was not fully acclimated to the chamber environment.
+
+*Procedure*: As described above, we fit provisional nonrectangular hyperbola to each \aqcurve{} curve and calculated the model $r^2$.
+There was a clear break between typical curves and poorly fitting curves where $r^2 < `r aq_r2_threshold`$.
+We therefore removed \aqcurve{} curves with $r^2 < `r aq_r2_threshold`$.
+
 ### Data analysis
@@
 - models fit in Stan/brms
-- we chose the number of chains, warmup and sampling iterations, and maximum treedepth so that parameter estimates converged ($\hat{R} < 1.01$ [@gelman_inference_1992]) and the effective sample size (ESS) for each parameter was greater than $10^3$.
+- we chose the number of chains, warmup and sampling iterations, and maximum treedepth so that parameter estimates converged ($\hat{R} < 1.01$ [@gelman_inference_1992]) and the effective sample size (ESS) for each parameter was $> 10^3$.
 
@@
 We started with a base model that includes fixed effects of light intensity and light treatment on \aax, a random effect of accession on \aax, and a linear regression of PAR on accession-level \aax.
-We next compared the fit of the base model to a more complex with an interaction between light intensity and light treatment using LOOIC with the \textit{R} package **loo** version `r loo_version` [@vehtari_practical_2017].
+We next compared the fit of the base model to a more complex with an interaction between light intensity and light treatment using LOOIC with the \textit{R} package **loo** version `r packageVersion("loo")` [@vehtari_practical_2017].
 These comparisons allowed us to test the main predictions of each hypothesis and for a reinforcing interactions between assimilatory and plastic hypotheses (tab:predictions).

### 2024-04-04 ms/ms.tex
drafted light response cleaning methods and table

@@
 
-The raw data set consisted of 2370 \agcurve{} curves with an average of
-63.2. Manual curation of a data set this size in a principled,
-consistent manner is not feasible. Therefore, we automated data cleaning
-using custom \emph{R} scripts. Cleaning is divided into six sequential
-steps \autoref{tbl-cleaning1}.
+The raw data set consisted of 2,370 \agcurve{} curves with an average of
+63.2 points per curve. Manual curation of a data set this size in a
+principled, consistent manner is not feasible. Therefore, we automated
+data cleaning using custom \emph{R} scripts. Cleaning is divided into
+six sequential steps \autoref{tbl-cleaning1}.
 
@@
 \midrule
-\cellcolor{gray!10}{1. remove unreliable and unusable data points} & \cellcolor{gray!10}{2361} & \cellcolor{gray!10}{63.0}\\
-2. remove hysteretic portion of \agcurve{} curves at low \gsw{} & 2360 & 59.2\\
-\cellcolor{gray!10}{3. remove outliers within each \agcurve{} curve} & \cellcolor{gray!10}{2360} & \cellcolor{gray!10}{58.7}\\
-4. remove replicates with no overlap between amphi and pseudohypo \agcurve{} curves & 2268 & 58.5\\
-\cellcolor{gray!10}{5. thin redundant data points within each \agcurve{} curve} & \cellcolor{gray!10}{2268} & \cellcolor{gray!10}{28.1}\\
+\cellcolor{gray!10}{1. remove unreliable and unusable data points} & \cellcolor{gray!10}{2,361} & \cellcolor{gray!10}{63.0}\\
+2. remove hysteretic portion of \agcurve{} curves at low \gsw{} & 2,360 & 59.2\\
+\cellcolor{gray!10}{3. remove outliers within each \agcurve{} curve} & \cellcolor{gray!10}{2,360} & \cellcolor{gray!10}{58.7}\\
+4. remove replicates with no overlap between amphi and pseudohypo \agcurve{} curves & 2,268 & 58.5\\
+\cellcolor{gray!10}{5. thin redundant data points within each \agcurve{} curve} & \cellcolor{gray!10}{2,268} & \cellcolor{gray!10}{28.1}\\
 \addlinespace
-6. Trim extreme \aax{} values & 2210 & 28.0\\
+6. Trim extreme \aax{} values & 2,210 & 28.0\\
 \bottomrule
@@
 \emph{R}. We sequentially removed data points with an absolute external
-studentized residual greater than 3 until none remained.
+studentized residual \(> 3\) until none remained.
 
@@
 
-WORKING HERE!
-
-The raw data set consisted of \aqcurve{} curves with an average of .
-Manual curation of a data set this size in a principled, consistent
-manner is not feasible. Therefore, we automated data cleaning using
-custom \emph{R} scripts. Cleaning is divided into six sequential steps
-\autoref{tbl-cleaning2}.
+The raw data set consisted of 658 \aqcurve{} curves with an average of
+19.4. Manual curation of a data set this size in a principled,
+consistent manner is not feasible. Therefore, we automated data cleaning
+using custom \emph{R} scripts. Cleaning is divided into two sequential
+steps \autoref{tbl-cleaning2}.
 
@@
 
-\caption{\label{tbl-cleaning2}Six sequential steps for cleaning
+\caption{\label{tbl-cleaning2}Two sequential steps for cleaning
 \agcurve{} curves. The rationale and procedure for each step are
 described in the text. The rightmost columns summarize the number of
-curves and mean number of points per curve remaining after each step.
-For reference, there are four possible \agcurve{} curves per replicate:
-all combinations of leaf type (amphi or pseudohypo) and light intensity
-(high or low).}
+curves and mean number of points per curve remaining after each step.}
 
@@
 
+\centering
+\begin{tabular}{>{\raggedright\arraybackslash}p{9cm}>{\centering\arraybackslash}p{2.5cm}>{\centering\arraybackslash}p{2.5cm}}
+\toprule
+Step: description & Number of curves & Number of points per curve\\
+\midrule
+\cellcolor{gray!10}{1. remove outliers within each \aqcurve{} curve} & \cellcolor{gray!10}{658} & \cellcolor{gray!10}{19.1}\\
+2. remove \aqcurve{} curves with poor fit & 652 & 19.1\\
+\bottomrule
+\end{tabular}
+
 }
@@
 \end{table}%
+
+\paragraph{\texorpdfstring{Remove outliers within each \aqcurve{}
+curve}{Remove outliers within each  curve}}\label{remove-outliers-within-each-curve-1}
+
+\emph{Rationale}: Individual outliers within \agcurve{} curves, usually
+caused by transitory changes in chamber conditions, exert undue leverage
+on parameter estimates and cause bias and/or low precision in parameter
+estimates.
+
+\emph{Procedure}: We fit provisional nonrectangular hyperbola
+(\textbf{marshall\_model\_1980?}) to each \aqcurve{} curve using
+nonlinear regression with the \texttt{nlsLM()} function from the
+\emph{R} package \textbf{minpack.lm} version 1.2.4
+(\textbf{Elzhov\_minpack.lm\_2023?}). We sequentially removed data
+points with an absolute external studentized residual \(> 3\) until none
+remained.
+
+\paragraph{\texorpdfstring{Remove \aqcurve{} curves with poor
+fit}{Remove  curves with poor fit}}\label{remove-curves-with-poor-fit}
+
+\emph{Rationale}: \aqcurve{} curves with a poor fit to the
+nonrectangular hyperbola most likely indicate systematic measurement
+error and/or the leaf was not fully acclimated to the chamber
+environment.
+
+\emph{Procedure}: As described above, we fit provisional nonrectangular
+hyperbola to each \aqcurve{} curve and calculated the model \(r^2\).
+There was a clear break between typical curves and poorly fitting curves
+where \(r^2 < 0.99\). We therefore removed \aqcurve{} curves with
+\(r^2 < 0.99\).
 
@@
   (\(\hat{R} < 1.01\) (\emph{8})) and the effective sample size (ESS)
-  for each parameter was greater than \(10^3\).
+  for each parameter was \(> 10^3\).
 \end{itemize}

### 2024-04-04 ms/ms.qmd
drafted some methods on estimating parameters

@@
   \newcommand{\gsw}{$g_\text{sw}$}
+  \newcommand{\loggsw}{$\log(g_\text{sw})$}
+  \newcommand{\logA}{$\log(A)$}
   \newcommand{\ppfd}{$\mathrm{PPFD}$}
@@
 
-*Procedure*: We fit provisional nonrectangular hyperbola [@marshall_model_1980] to each \aqcurve{} curve using nonlinear regression with the `nlsLM()` function from the *R* package **minpack.lm** version `r packageVersion("minpack.lm")` [@Elzhov_minpack.lm_2023].
+*Procedure*: We fit provisional nonrectangular hyperbola [@marshall_model_1980] to each \aqcurve{} curve using nonlinear regression with the `nlsLM()` function from the *R* package **minpack.lm** version `r packageVersion("minpack.lm")` [@elzhov_minpacklm_2023].
 We sequentially removed data points with an absolute external studentized residual $> `r point_outlier_threshold`$ until none remained.
@@
 
-We fit Bayesian models using the *R* package **brms** version `r fit_aa$version$brms` [@burkner_brms_2017] with MCMC sampling in *Stan* [@stan_development_team_stan_2023].
+We fit Bayesian models with MCMC sampling in the probabilistic programming language *Stan* [@stan_development_team_stan_2023].
 We used CmdStan version `r fit_aa$version$cmdstan` and **cmdstanr** version `r fit_aa$version$cmdstanr` [@gabry_cmdstanr_2023] to interface with *R* version `r getRversion()` [@r_core_team_r:_2023].
@@
 
-#### Parameter estimates (this is a provisional section)
+#### Parameter estimation
+
+There were four levels of parameter estimation in our analysis:
+
+1. Estimate \agcurve{} curve parameters for each leaf
+2. Estimate \aax{} for each light intensity with leaf using \agcurve{} curve parameters
+3. Estimate the effects of light intensity, light treatment, and accession on \aax{} (assimilatory and plasticity hypotheses)
+4. Estimate the effects of native light habitat on accession-level \aax{} (constitutive hypothesis)
+
+Although the higher-level parameter estimates depend on the lower-level parameter estimates, we fit all models simultaneously to ensure that the uncertainty in lower-level estimates propagated to higher-level estimates.
+
+##### \agcurve{} curve parameters
+
+We modeled \logA as a quadratic function of \loggsw for each leaf using the following equation:
+
+$$\log(A_{ij}) = (\beta_0 + b_{0,j}) + (\beta_1 + b_{1,j}) \log{g_{\text{sw},i}} + (\beta_2 + b_{2,j}) \log{g_{\text{sw},i}}^2 + \epsilon_{i}$$
+
+where $\beta_0$, $\beta_1$, and $\beta_2$ are the average intercept, linear, and quadratic coefficients, respectively.
+We used diffuse normal priors with mean $0$ and standard deviation $10$ on these parameters.
+We estimated random effects of curve $j$ on the intercept ($b_{0,j}$), linear ($b_{1,j}$), and quadratic ($b_{2,j}$) coefficients.
+We assumed that the $j \times 3$ array of coefficients was multivariate normal a mean vector of $\vec{0}$ and covariance $\Sigma_\text{curve}$.
+We used a weakly informative normal prior with mean $0$ and standard deviation $1$ on the log-transformed standard deviations (i.e.
+the diagonal of $\Sigma_\text{curve}$).
+We used a weakly informative $\mathrm{LJK}(2)$ prior on the correlation matrix.
+The off diagonal elements of $\Sigma_\text{curve}$ can be calculated from its diagonal elements and the correlation matrix.
+
+The residuals $\epsilon_{i}$ were modeled as a lag-1 autocorrelated time-series.
+We further assumed that the residual standard deviation of the $j^{\text{th}}$ curve ($\sigma_{\epsilon,j}$) was inversely proportional to the leaf surface area ($S_j$) within the chamber:
+
+$$\log (\sigma_{\epsilon,j}) = \log (\sigma_{\qty{6}{\raiseto{2}\centi\meter},\epsilon}) + (6 - S_j) \beta_{S,\epsilon}$$
+where $\sigma_{\qty{6}{\raiseto{2}\centi\meter},\epsilon}$ is the minimum residual standard deviation when the $\qty{6}{\raiseto{2}\centi\meter}$ chamber is completely filled.
+The residual standard deviation increases on log-linear scale by $\beta_{S,\epsilon}$.
+We used a weakly informative normal prior with mean $-3$ and standard deviation $5$ on $\log (\sigma_{\qty{6}{\raiseto{2}\centi\meter},\epsilon})$ and a weakly informative normal prior with mean $0$ and standard deviation $1$ on $\beta_{S,\epsilon}$.
+
+##### \aax{} for each light intensity with leaf using \agcurve{} curve parameters
+
+Within the $k^{\text{th}}$ leaf, we estimated \aax{} for each light intensity by integrating the difference in \logA between the amphi and pseudohypo \agcurve{} curves over the range of \gsw{} values where the curves overlap.
+We used the following equation to estimate \aax{} for the $k^{\text{th}}$ leaf at light intensity $l$ in accession $m$:
+
+$$\widehat{\mathrm{AA}}_{klm} = \int_{\text{min(log(}g_\text{sw}))}^{\text{max(log(}g_\text{sw}))} \text{log}\bigg(\frac{\hat{A}_\text{amphi}(x; \theta_{klm,\text{amphi})}}{\hat{A}_\text{hypo}(x; \theta_{klm,\text{hypo})}}\bigg) dx$$
+where:
+
+$$\theta_\text{amphi} \in \{\hat{b}_{0, f(\text{amphi}, k,l,m)}, \hat{b}_{1, f(\text{amphi}, k,l,m)}, \hat{b}_{2, f(\text{amphi}, k,l,m)}\}, \text{and}$$
+
+$$\theta_\text{hypo} \in \{\hat{b}_{0, f(\text{hypo}, k,l,m)}, \hat{b}_{1, f(\text{hypo}, k,l,m)}, \hat{b}_{2, f(\text{hypo}, k,l,m)}\}.$$ 
+The function $f : \Theta_1 \rightarrow \Theta_2$ maps the set $\Theta_1$ indexed by leaf type (amphi or pseudohypo), leaf replicate, light intensity, and accession to set $\Theta_2$ indexed by individual \agcurve{} curve.
+
+##### Effects of light intensity, light treatment, and accession on \aax
+
+WORKING HERE
+
+We estimated these parameters for each leaf using the posterior distribution of the parameters from the \agcurve{} curve fitting procedure.
 
@@
 Estimate \aax{} by integrating over range of \gsw{} overlap (need to add indices for accession, replicate, light intensity):
-
-$$\mathrm{AA} = \int_{\text{min(log(}g_\text{sw}))}^{\text{max(log(}g_\text{sw}))} \text{log}\bigg(\frac{A_\text{amphi}(x; \theta_\text{amphi})}{A_\text{hypo}(x; \theta_\text{hypo})}\bigg) dx$$
 

### 2024-04-04 ms/ms.tex
drafted some methods on estimating parameters

@@
 \newcommand{\gsw}{$g_\text{sw}$}
+\newcommand{\loggsw}{$\log(g_\text{sw})$}
+\newcommand{\logA}{$\log(A)$}
 \newcommand{\ppfd}{$\mathrm{PPFD}$}
@@
 
-\emph{Procedure}: We fit provisional nonrectangular hyperbola
-(\textbf{marshall\_model\_1980?}) to each \aqcurve{} curve using
-nonlinear regression with the \texttt{nlsLM()} function from the
-\emph{R} package \textbf{minpack.lm} version 1.2.4
-(\textbf{Elzhov\_minpack.lm\_2023?}). We sequentially removed data
-points with an absolute external studentized residual \(> 3\) until none
-remained.
+\emph{Procedure}: We fit provisional nonrectangular hyperbola (\emph{8})
+to each \aqcurve{} curve using nonlinear regression with the
+\texttt{nlsLM()} function from the \emph{R} package \textbf{minpack.lm}
+version 1.2.4 (\emph{9}). We sequentially removed data points with an
+absolute external studentized residual \(> 3\) until none remained.
 
@@
   maximum treedepth so that parameter estimates converged
-  (\(\hat{R} < 1.01\) (\emph{8})) and the effective sample size (ESS)
+  (\(\hat{R} < 1.01\) (\emph{10})) and the effective sample size (ESS)
   for each parameter was \(> 10^3\).
@@
 
-We fit Bayesian models using the \emph{R} package \textbf{brms} version
-2.20.4 (\emph{9}) with MCMC sampling in \emph{Stan} (\emph{10}). We used
-CmdStan version 2.33.1 and \textbf{cmdstanr} version 0.7.1 (\emph{11})
-to interface with \emph{R} version 4.3.2 (\emph{12}). We sampled the
-posterior distribution from 4 chains with 1000 iterations each after
-1000 warmup iterations per chain. We estimated parameters and confidence
-intervals as the median and 95\% quantile intervals of the posterior,
+We fit Bayesian models with MCMC sampling in the probabilistic
+programming language \emph{Stan} (\emph{11}). We used CmdStan version
+2.33.1 and \textbf{cmdstanr} version 0.7.1 (\emph{12}) to interface with
+\emph{R} version 4.3.2 (\emph{13}). We sampled the posterior
+distribution from 4 chains with 1000 iterations each after 1000 warmup
+iterations per chain. We estimated parameters and confidence intervals
+as the median and 95\% quantile intervals of the posterior,
 respectively.
 
-\paragraph{Parameter estimates (this is a provisional
-section)}\label{parameter-estimates-this-is-a-provisional-section}
+\paragraph{Parameter estimation}\label{parameter-estimation}
+
+There were four levels of parameter estimation in our analysis:
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\tightlist
+\item
+  Estimate \agcurve{} curve parameters for each leaf
+\item
+  Estimate \aax{} for each light intensity with leaf using \agcurve{}
+  curve parameters
+\item
+  Estimate the effects of light intensity, light treatment, and
+  accession on \aax{} (assimilatory and plasticity hypotheses)
+\item
+  Estimate the effects of native light habitat on accession-level \aax{}
+  (constitutive hypothesis)
+\end{enumerate}
+
+Although the higher-level parameter estimates depend on the lower-level
+parameter estimates, we fit all models simultaneously to ensure that the
+uncertainty in lower-level estimates propagated to higher-level
+estimates.
+
+\subparagraph{\texorpdfstring{\agcurve{} curve
+parameters}{ curve parameters}}\label{curve-parameters}
+
+We modeled \logA as a quadratic function of \loggsw for each leaf using
+the following equation:
+
+\[\log(A_{ij}) = (\beta_0 + b_{0,j}) + (\beta_1 + b_{1,j}) \log{g_{\text{sw},i}} + (\beta_2 + b_{2,j}) \log{g_{\text{sw},i}}^2 + \epsilon_{i}\]
+
+where \(\beta_0\), \(\beta_1\), and \(\beta_2\) are the average
+intercept, linear, and quadratic coefficients, respectively. We used
+diffuse normal priors with mean \(0\) and standard deviation \(10\) on
+these parameters. We estimated random effects of curve \(j\) on the
+intercept (\(b_{0,j}\)), linear (\(b_{1,j}\)), and quadratic
+(\(b_{2,j}\)) coefficients. We assumed that the \(j \times 3\) array of
+coefficients was multivariate normal a mean vector of \(\vec{0}\) and
+covariance \(\Sigma_\text{curve}\). We used a weakly informative normal
+prior with mean \(0\) and standard deviation \(1\) on the
+log-transformed standard deviations (i.e.~the diagonal of
+\(\Sigma_\text{curve}\)). We used a weakly informative
+\(\mathrm{LJK}(2)\) prior on the correlation matrix. The off diagonal
+elements of \(\Sigma_\text{curve}\) can be calculated from its diagonal
+elements and the correlation matrix.
+
+The residuals \(\epsilon_{i}\) were modeled as a lag-1 autocorrelated
+time-series. We further assumed that the residual standard deviation of
+the \(j^{\text{th}}\) curve (\(\sigma_{\epsilon,j}\)) was inversely
+proportional to the leaf surface area (\(S_j\)) within the chamber:
+
+\[\log (\sigma_{\epsilon,j}) = \log (\sigma_{\qty{6}{\raiseto{2}\centi\meter},\epsilon}) + (6 - S_j) \beta_{S,\epsilon}\]
+where \(\sigma_{\qty{6}{\raiseto{2}\centi\meter},\epsilon}\) is the
+minimum residual standard deviation when the
+\(\qty{6}{\raiseto{2}\centi\meter}\) chamber is completely filled. The
+residual standard deviation increases on log-linear scale by
+\(\beta_{S,\epsilon}\). We used a weakly informative normal prior with
+mean \(-3\) and standard deviation \(5\) on
+\(\log (\sigma_{\qty{6}{\raiseto{2}\centi\meter},\epsilon})\) and a
+weakly informative normal prior with mean \(0\) and standard deviation
+\(1\) on \(\beta_{S,\epsilon}\).
+
+\subparagraph{\texorpdfstring{\aax{} for each light intensity with leaf
+using \agcurve{} curve
+parameters}{ for each light intensity with leaf using  curve parameters}}\label{for-each-light-intensity-with-leaf-using-curve-parameters}
+
+Within the \(k^{\text{th}}\) leaf, we estimated \aax{} for each light
+intensity by integrating the difference in \logA between the amphi and
+pseudohypo \agcurve{} curves over the range of \gsw{} values where the
+curves overlap. We used the following equation to estimate \aax{} for
+the \(k^{\text{th}}\) leaf at light intensity \(l\) in accession \(m\):
+
+\[\widehat{\mathrm{AA}}_{klm} = \int_{\text{min(log(}g_\text{sw}))}^{\text{max(log(}g_\text{sw}))} \text{log}\bigg(\frac{\hat{A}_\text{amphi}(x; \theta_{klm,\text{amphi})}}{\hat{A}_\text{hypo}(x; \theta_{klm,\text{hypo})}}\bigg) dx\]
+where:
+
+\[\theta_\text{amphi} \in \{\hat{b}_{0, f(\text{amphi}, k,l,m)}, \hat{b}_{1, f(\text{amphi}, k,l,m)}, \hat{b}_{2, f(\text{amphi}, k,l,m)}\}, \text{and}\]
+
+\[\theta_\text{hypo} \in \{\hat{b}_{0, f(\text{hypo}, k,l,m)}, \hat{b}_{1, f(\text{hypo}, k,l,m)}, \hat{b}_{2, f(\text{hypo}, k,l,m)}\}.\]
+The function \(f : \Theta_1 \rightarrow \Theta_2\) maps the set
+\(\Theta_1\) indexed by leaf type (amphi or pseudohypo), leaf replicate,
+light intensity, and accession to set \(\Theta_2\) indexed by \agcurve{}
+curve.
+
+WORKING HERE
+
+We estimated these parameters for each leaf using the posterior
+distribution of the parameters from the \agcurve{} curve fitting
+procedure.
 
@@
 indices for accession, replicate, light intensity):
-
-\[\mathrm{AA} = \int_{\text{min(log(}g_\text{sw}))}^{\text{max(log(}g_\text{sw}))} \text{log}\bigg(\frac{A_\text{amphi}(x; \theta_\text{amphi})}{A_\text{hypo}(x; \theta_\text{hypo})}\bigg) dx\]
 
@@
 light intensity and light treatment using LOOIC with the \textit{R}
-package \textbf{loo} version 2.7.0 (\textbf{vehtari\_practical\_2017?}).
-These comparisons allowed us to test the main predictions of each
-hypothesis and for a reinforcing interactions between assimilatory and
-plastic hypotheses (tab:predictions). However, these models cannot test
-predictions when there is an interaction with the constitutive
-hypothesis, because this hypothesis posits accession-levels variation in
-responses to light intensity and/or light treatment. Therefore, we
-tested for significant variation among accessions in light intensity and
-light treatment by comparing LOOIC values of models with and without
-random interactions between these factors and accession.
+package \textbf{loo} version 2.7.0 (\emph{14}). These comparisons
+allowed us to test the main predictions of each hypothesis and for a
+reinforcing interactions between assimilatory and plastic hypotheses
+(tab:predictions). However, these models cannot test predictions when
+there is an interaction with the constitutive hypothesis, because this
+hypothesis posits accession-levels variation in responses to light
+intensity and/or light treatment. Therefore, we tested for significant
+variation among accessions in light intensity and light treatment by
+comparing LOOIC values of models with and without random interactions
+between these factors and accession.
 
@@
 
+\bibitem[\citeproctext]{ref-marshall_model_1980}
+\CSLLeftMargin{8. }%
+\CSLRightInline{B. Marshall, P. V. Biscoe,
+\href{https://doi.org/10.1093/jxb/31.1.29}{A {Model} for
+{C}\(_{\textrm{3}}\) {Leaves} {Describing} the {Dependence} of {Net}
+{Photosynthesis} on {Irradiance}}. \emph{Journal of Experimental Botany}
+\textbf{31}, 29--39 (1980).}
+
+\bibitem[\citeproctext]{ref-elzhov_minpacklm_2023}
+\CSLLeftMargin{9. }%
+\CSLRightInline{T. V. Elzhov, K. M. Mullen, A.-N. Spiess, B. M. Bolker,
+\emph{Minpack.lm: {R} {Interface} to the {Levenberg}-{Marquardt}
+{Nonlinear} {Least}-{Squares} {Algorithm} {Found} in {MINPACK}, {Plus}
+{Support} for {Bounds}} (2023;
+\url{https://CRAN.R-project.org/package=minpack.lm}).}
+
 \bibitem[\citeproctext]{ref-gelman_inference_1992}
-\CSLLeftMargin{8. }%
+\CSLLeftMargin{10. }%
 \CSLRightInline{A. Gelman, D. B. Rubin, Inference from iterative
@@
 
-\bibitem[\citeproctext]{ref-burkner_brms_2017}
-\CSLLeftMargin{9. }%
-\CSLRightInline{P.-C. Bürkner,
-\href{https://doi.org/10.18637/jss.v080.i01}{\textbf{Brms} : {An}
-\emph{r} {Package} for {Bayesian} {Multilevel} {Models} {Using}
-\emph{stan}}. \emph{Journal of Statistical Software} \textbf{80}
-(2017).}
-
 \bibitem[\citeproctext]{ref-stan_development_team_stan_2023}
-\CSLLeftMargin{10. }%
+\CSLLeftMargin{11. }%
 \CSLRightInline{Stan Development Team, \emph{Stan {Modeling} {Language}
@@
 \bibitem[\citeproctext]{ref-gabry_cmdstanr_2023}
-\CSLLeftMargin{11. }%
+\CSLLeftMargin{12. }%
 \CSLRightInline{J. Gabry, R. Češnovar, \emph{Cmdstanr: {R} {Interface}
@@
 \bibitem[\citeproctext]{ref-r_core_team_r:_2023}
-\CSLLeftMargin{12. }%
+\CSLLeftMargin{13. }%
 \CSLRightInline{R Core Team, \emph{R: {A} {Language} and {Environment}
@@
 
+\bibitem[\citeproctext]{ref-vehtari_practical_2017}
+\CSLLeftMargin{14. }%
+\CSLRightInline{A. Vehtari, A. Gelman, J. Gabry,
+\href{https://doi.org/10.1007/s11222-016-9696-4}{Practical {Bayesian}
+model evaluation using leave-one-out cross-validation and {WAIC}}.
+\emph{Statistics and Computing} \textbf{27}, 1413--1432 (2017).}
+
 \end{CSLReferences}

### 2024-04-04 ms/ms.qmd
moar methods

@@
 
-Within the $k^{\text{th}}$ leaf, we estimated \aax{} for each light intensity by integrating the difference in \logA between the amphi and pseudohypo \agcurve{} curves over the range of \gsw{} values where the curves overlap.
-We used the following equation to estimate \aax{} for the $k^{\text{th}}$ leaf at light intensity $l$ in accession $m$:
+Within the $k^{\text{th}}$ leaf, we estimated \aax{} for each light intensity by integrating the difference in \logA between the amphi and pseudohypo \agcurve{} curves over the range of \gsw{} values where the curves overlap (from $\min(\log(g_\text{sw}))$ to $\max(\log(g_\text{sw}))$).
+The estimate of \aax{} for the $k^{\text{th}}$ leaf at light intensity $l$ in accession $m$ is:
 
@@
 The function $f : \Theta_1 \rightarrow \Theta_2$ maps the set $\Theta_1$ indexed by leaf type (amphi or pseudohypo), leaf replicate, light intensity, and accession to set $\Theta_2$ indexed by individual \agcurve{} curve.
+This mapping is necessary because the random effects structure differs between models of \loggsw on \logA and that of models predicting \aax{} described in the next section.
 
 ##### Effects of light intensity, light treatment, and accession on \aax
+
+We tested for effects of light intensity, light treatment, accession, and their interactions on \aax{} by estimating parameters in the following equation:
+
+$$\mathrm{AA} = ... + \epsilon$$
 
@@
 We estimated these parameters for each leaf using the posterior distribution of the parameters from the \agcurve{} curve fitting procedure.
-
-Base model (need to add indices):
-
-$$\text{log(A)} = \beta_0 + \beta_1 \text{log}(g_\text{sw}) + \beta_2 \text{log}(g_\text{sw}) ^ 2 + \epsilon$$
-Estimate \aax{} by integrating over range of \gsw{} overlap (need to add indices for accession, replicate, light intensity):
-
-where $\text{min(log(}g_\text{sw}))$ and $\text{max(log(}g_\text{sw}))$ are the minimum and maximum values of $\text{log(}g_\text{sw})$ where amphi and pseudohypo curves overlap.
-\Aamphi{} is calculated from parameter estimates $\theta_\text{amphi} \in \{\beta_{0,\text{amphi}}, \beta_{1,\text{amphi}}, \beta_{2,\text{amphi}}\}$; \Ahypo{} is calculated from parameter estimates $\theta_\text{hypo} \in \{\beta_{0,\text{hypo}}, \beta_{1,\text{hypo}}, \beta_{2,\text{hypo}}\}$ for each curve.
 

### 2024-04-05 ms/ms.qmd
worked on stat methods

@@
 
-We tested for effects of light intensity, light treatment, accession, and their interactions on \aax{} by estimating parameters in the following equation:
+We tested for effects of light intensity, light treatment, accession, and their interactions on \aax.
+All models included effects of light intensity and light treatment, as well as random effects of accession and replicate within accession.
+More complex models included interactions between light intensity and light treatment, as well as random effects of accession on the effects of light intensity and light treatment.
+We used a weakly informative normal prior with mean $0$ and standard deviation $10$ for fixed effects light intensity and treatment.
+We used a weakly informative normal prior with mean $-5$ and standard deviation $5$ for the the random effect standard deviation of replicate within accession.
+We accounted for the phylogenetic structure among the random effects of accession using an Ornstein-Uhlenbeck (OU) process [@hansen_stabilizing_1997].
+We used a weakly informative normal prior with mean $0$ and standard deviation $10$ on OU parameters.
+We modeled the residual standard deviation on a log-link scale with effects of light intensity and light treatment.
+We used a weakly informative normal prior with mean $-3$ and standard deviation $5$ on the residual standard deviation intercept and a weakly informative normal prior with mean $0$ and standard deviation $1$ on the effects of light intensity and light treatment on the residual standard deviation.
 
-$$\mathrm{AA} = ... + \epsilon$$
-
-WORKING HERE
+WORKING HERE - TABLE OR SOMETHING SUMMARIZING MODELS?
 

### 2024-04-05 ms/ms.qmd
drafted methods for regression of light habitat on aa

@@
 We used a weakly informative normal prior with mean $0$ and standard deviation $10$ for fixed effects light intensity and treatment.
-We used a weakly informative normal prior with mean $-5$ and standard deviation $5$ for the the random effect standard deviation of replicate within accession.
+We used a weakly informative normal prior with mean $-3$ and standard deviation $5$ for the the random effect standard deviation of replicate within accession.
 We accounted for the phylogenetic structure among the random effects of accession using an Ornstein-Uhlenbeck (OU) process [@hansen_stabilizing_1997].
@@
 
+##### Effects of native light habitat on accession-level \aax
+
+We tested a linear effect of native light habitat on accession-level \aax.
+In Models 1 and 2, we used the random intercept of \aax{} as a response variable; in model 3 we used the random intercept plus the random effect of accession at high light intensity; in model 4 we used the random intercept plus the random effect of accession in the sun treatment; in model 5 we used the random intercept plus the random effects of accession at high light intensity and sun treatment.
+We used a weakly informative normal prior with mean $0$ and standard deviation $1$ for the slope and intercept.
+We accounted for the phylogenetic structure among the model residuals using an (OU) process.
+We used a weakly informative normal prior with mean $0$ and standard deviation $10$ on OU parameters.
+
 WORKING HERE - TABLE OR SOMETHING SUMMARIZING MODELS?
+- TABLE SUMMARIZING PARAMETERS
 

### 2024-04-06 ms/ms.qmd
working on table of estimated parameters

@@
 ```{r load-objects, echo=FALSE, eval=TRUE}
-
+library(dplyr)
 library(readr)
@@
 
+```{r, parameters, echo=FALSE,message=FALSE,warning=FALSE}
+#| label: tbl-parameters
+#| tbl-cap: "caption"
+
+df_parameters = read_csv("../data/parameters1.csv", col_types = "cccc", n_max = 6)
+
+# WORKING ON TABLE IN test.gmd
+df_parameters |>
+  dplyr::select(Parameter = latex, Description = description) |>
+  knitr::kable(
+    format = "latex",
+    align = c("c", "l"),
+    booktabs = TRUE,
+    escape = FALSE
+  ) |>
+  kableExtra::kable_styling(latex_options = c("striped", "no_vertical_lines")) |>
+  kableExtra::column_spec(1, width = "1in") |>
+  kableExtra::column_spec(2, width = "5in") |>
+  kableExtra::pack_rows("\\agcurve{} curve parameters",
+                        min(which(df_parameters$par_type == "ags")),
+                        max(which(df_parameters$par_type == "ags")),
+                        escape = FALSE) #|>
+                        # kableExtra::pack_rows("x", 2, 2)
+
+```
+
 ##### \agcurve{} curve parameters

### 2024-04-06 ms/test.qmd
working on table of estimated parameters

--- 2024-03-06 ms/test.qmd
@@
 header-includes: |
+  \usepackage{hyperref}
   \addtokomafont{disposition}{\rmfamily}
   \usepackage{siunitx}
-  \newcommand{\gsc}{$g_\text{sc}$}
-  \newcommand{\ppfd}[1]{$\mathrm{PPFD} = \qty{#1}{\micro\mol\raiseto{-2}\meter\raiseto{-1}\second}$}
+  \newcommand{\aax}{$\mathrm{AA}$}
+  \newcommand{\Aamphi}{$A_{\mathrm{amphi}}$}
+  \newcommand{\Ahypo}{$A_{\mathrm{hypo}}$}
+  \newcommand{\agcurve}{$A \textendash g_\text{sw}$}
+  \newcommand{\ca}{$C_\mathrm{a}$}
+  \newcommand{\caequals}[1]{$C_\mathrm{a} = \qty{#1}{\micro\mol\raiseto{-1}\mol}$}
+  \newcommand{\cabetween}[2]{#1 < $C_\mathrm{a} < \qty{#2}{\micro\mol\raiseto{-1}\mol}$}
+  \newcommand{\gsw}{$g_\text{sw}$}
+  \newcommand{\ppfd}{$\mathrm{PPFD}$}
+  \newcommand{\ppfdequals}[1]{$\mathrm{PPFD} = \qty{#1}{\micro\mol\raiseto{-2}\meter\raiseto{-1}\second}$}
+  \newcommand{\rh}{$\mathrm{RH}$}
+  \newcommand{\rhequals}[1]{$\mathrm{RH} = #1\%$}
+  \newcommand{\tleafequals}[1]{$T_\mathrm{leaf} = \qty{#1}{\degreeCelsius}$}
 mainfont: Times New Roman
-csl: https://www.zotero.org/styles/american-journal-of-botany
+bibliography: solanum-aa.bib
+csl: https://www.zotero.org/styles/science
 ---
 
-words \gsc
+```{r load-objects, echo=FALSE, eval=TRUE}
+library(dplyr)
+library(readr)
+library(tibble)
+```
 
-this \ppfd{2000}
+Manual curation of a data set this size in a principled, consistent manner is not feasible.
+Therefore, we automated data cleaning using custom *R* scripts.
+Cleaning is divided into six sequential steps \autoref{tbl-parameters}.
+
+```{r, parameters, echo=FALSE,message=FALSE,warning=FALSE}
+#| label: tbl-parameters
+#| tbl-cap: "caption"
+
+df_parameters = read_csv("../data/parameters1.csv", col_types = "cccc", n_max = 6)
+
+df_parameters |>
+  dplyr::select(Parameter = latex, Description = description) |>
+  knitr::kable(
+    format = "latex",
+    align = c("c", "l"),
+    booktabs = TRUE,
+    escape = FALSE
+  ) |>
+  kableExtra::kable_styling(latex_options = c("striped", "no_vertical_lines")) |>
+  kableExtra::column_spec(1, width = "1in") |>
+  kableExtra::column_spec(2, width = "5in") |>
+  kableExtra::pack_rows("\\agcurve{} curve parameters",
+                        min(which(df_parameters$par_type == "ags")),
+                        max(which(df_parameters$par_type == "ags")),
+                        escape = FALSE) #|>
+                        # kableExtra::pack_rows("x", 2, 2)
+
+```

### 2024-04-08 ms/ms.qmd
added draft of tbl-parameters

--- 2024-04-05 ms/ms.qmd
@@
 
-1. Estimate \agcurve{} curve parameters for each leaf
+1. Estimate \agcurve{} curve parameters
 2. Estimate \aax{} for each light intensity with leaf using \agcurve{} curve parameters
@@
 Although the higher-level parameter estimates depend on the lower-level parameter estimates, we fit all models simultaneously to ensure that the uncertainty in lower-level estimates propagated to higher-level estimates.
+
+```{r, parameters, echo=FALSE,message=FALSE,warning=FALSE}
+#| label: tbl-parameters
+#| tbl-cap: "Description of parameters estimated in the hierarchical Bayesian model.
+The `Parameter` column lists the parameter name as it appears in text.
+The `Description` column provides a brief description of the parameter."
+
+df_parameters = read_csv("../data/parameters1.csv", col_types = "ccccl") |>
+    dplyr::filter(include)
+
+df_parameters |>
+  dplyr::select(Parameter = latex, Description = description) |>
+  knitr::kable(
+    format = "latex",
+    align = c("l", "l"),
+    booktabs = TRUE,
+    escape = FALSE,
+    longtable = TRUE
+  ) |>
+  kableExtra::kable_styling(latex_options = c("striped", "no_vertical_lines")) |>
+  kableExtra::column_spec(1, width = "1in") |>
+  kableExtra::column_spec(2, width = "5in") |>
+  kableExtra::pack_rows("\\agcurve{} curve parameters",
+                        min(which(df_parameters$par_type == "ags")),
+                        max(which(df_parameters$par_type == "ags")),
+                        escape = FALSE) |>
+  kableExtra::pack_rows("\\aax{} for each light intensity with leaf using \\agcurve{} curve parameters",
+                        min(which(df_parameters$par_type == "aa_est")),
+                        max(which(df_parameters$par_type == "aa_est")),
+                        escape = FALSE) |>
+  kableExtra::pack_rows("effects light intensity, light treatments, and accession on \\aax",
+                        min(which(df_parameters$par_type == "aa")),
+                        max(which(df_parameters$par_type == "aa")),
+                        escape = FALSE) |>
+  kableExtra::pack_rows("effects of native light habitat on accession-level \\aax",
+                        min(which(df_parameters$par_type == "ppfd_aa")),
+                        max(which(df_parameters$par_type == "ppfd_aa")),
+                        escape = FALSE)
+
+```
 
@@
 We accounted for the phylogenetic structure among the random effects of accession using an Ornstein-Uhlenbeck (OU) process [@hansen_stabilizing_1997].
+The expected covariance between accessions $i$ and $j$ ($\text{Cov}(i, j)$ ) is:
+
+$$\text{Cov}(i, j) = \frac{\sigma^2}{2 \alpha} \exp(-\alpha D_{ij})$$
+where $\sigma^2$ is the variance of the random effect, $\alpha$ is the rate of decay of the covariance with phylogenetic distance, $D_{ij}$, the phylogenetic distance between accessions $i$ and $j$.
+We estimated $\sigma^2 / (2 \alpha)$ and $\alpha$ as a separate parameters and reparameterized them as $\sigma^2$ and $\alpha$.
 We used a weakly informative normal prior with mean $0$ and standard deviation $10$ on OU parameters.
-We modeled the residual standard deviation on a log-link scale with effects of light intensity and light treatment.
+
+We modeled the residual standard deviation of \aax, (i.e.
+phylogenetically unstructured variation unaccounted for by explanatory variables) on a log-link scale with effects of light intensity and light treatment.
 We used a weakly informative normal prior with mean $-3$ and standard deviation $5$ on the residual standard deviation intercept and a weakly informative normal prior with mean $0$ and standard deviation $1$ on the effects of light intensity and light treatment on the residual standard deviation.

### 2024-04-08 ms/ms.tex
added draft of tbl-parameters

@@
 
+\begin{verbatim}
+
+Attaching package: 'dplyr'
+\end{verbatim}
+
+\begin{verbatim}
+The following objects are masked from 'package:stats':
+
+    filter, lag
+\end{verbatim}
+
+\begin{verbatim}
+The following objects are masked from 'package:base':
+
+    intersect, setdiff, setequal, union
+\end{verbatim}
+
 \subsection{Notes}\label{notes}
@@
 \item
-  Estimate \agcurve{} curve parameters for each leaf
+  Estimate \agcurve{} curve parameters
 \item
@@
 estimates.
+
+\begin{longtable}{>{\raggedright\arraybackslash}p{1in}>{\raggedright\arraybackslash}p{5in}}
+
+\caption{\label{tbl-parameters}Description of parameters estimated in
+the hierarchical Bayesian model. The \texttt{Parameter} column lists the
+parameter name as it appears in text. The \texttt{Description} column
+provides a brief description of the parameter.}
+
+\tabularnewline
+
+\toprule
+Parameter & Description\\
+\midrule
+\addlinespace[0.3em]
+\multicolumn{2}{l}{\textbf{\agcurve{} curve parameters}}\\
+\hspace{1em}\cellcolor{gray!10}{$\mathbf{B}_\text{curve}$} & \cellcolor{gray!10}{$n_\text{curve} \times 3$ array of random \agcurve{} curve-level coefficients ($b_{0,j}$, $b_{1,j}$, $b_{2,j}$); $\mathbf{B}_\text{curve} \sim \text{MVN}(\vec{0}, \symbf{\Sigma}_\text{curve})$}\\
+\hspace{1em}$\vec{\beta}_\text{curve}$ & vector of mean quadratic coefficients ($\beta_0$, $\beta_1$, $\beta_2$)\\
+\hspace{1em}\cellcolor{gray!10}{$\symbf{\Sigma}_\text{curve}$} & \cellcolor{gray!10}{$3 \times 3$ covariance matrix of curve-level coefficients}\\
+\hspace{1em}$\sigma_{\qty{6}{\raiseto{2}\centi\meter},\epsilon}$ & minimum residual standard deviation when the measured leaf surface area is $\qty{6}{\raiseto{2}\centi\meter}$\\
+\hspace{1em}\cellcolor{gray!10}{$\beta_{S,\epsilon}$} & \cellcolor{gray!10}{slope of the relationship between residual standard deviation and measured leaf surface area (log-link scale)}\\
+\hspace{1em}$\rho_{\epsilon}$ & lag-1 residual autocorrelation\\
+\addlinespace[0.3em]
+\multicolumn{2}{l}{\textbf{\aax{} for each light intensity with leaf using agcurve{} curve parameters}}\\
+\hspace{1em}\cellcolor{gray!10}{$\widehat{\mathrm{AA}}_{klm}$} & \cellcolor{gray!10}{estimate of \aax{} for the $k^{\text{th}}$ leaf at light intensity $l$ in accession $m$}\\
+\addlinespace[0.3em]
+\multicolumn{2}{l}{\textbf{effects light intensity, light treatments, and accession on \aax}}\\
+\hspace{1em}$\beta_{\mathrm{AA},0}$ & intercept of \aax{} at low light intensity in shade treatment\\
+\hspace{1em}\cellcolor{gray!10}{$\beta_{\mathrm{AA},2000}$} & \cellcolor{gray!10}{effect of high light intensity at \ppfdequals{2000} on \aax}\\
+\hspace{1em}$\beta_{\mathrm{AA},\text{sun}}$ & effect of sun treatment on \aax{}\\
+\hspace{1em}\cellcolor{gray!10}{$\beta_{\mathrm{AA},\text{2000,high}}$} & \cellcolor{gray!10}{effect of high light intensity at \ppfdequals{2000} on \aax{} in sun treatment}\\
+\hspace{1em}$\vec{\beta}_{\mathrm{AA},\text{acc}}$ & vector of $n_\text{acc}$ phylogenetically structured random accession-level effects on \aax; $\vec{\beta}_{\mathrm{AA},\text{acc}} \sim \text{MVN}(\vec{0}, \symbf{\Sigma}_{\text{AA,acc}})$\\
+\hspace{1em}\cellcolor{gray!10}{$\symbf{\Sigma}_{\text{AA,acc}}$} & \cellcolor{gray!10}{$n_\text{acc} \times n_\text{acc}$ covariance matrix of phylogenetically structured random accession-level effects on \aax;}\\
+\hspace{1em}$\vec{\beta}_{\mathrm{AA},\text{rep}}$ & vector of $n_\text{rep}$ random replicate-level effects on \aax; $\vec{\beta}_{\mathrm{AA},\text{rep}} \sim \text{Normal}(0, \sigma_{\mathrm{AA},\text{rep}})$\\
+\hspace{1em}\cellcolor{gray!10}{$\vec{\beta}_{\mathrm{AA},\text{2000,acc}}$} & \cellcolor{gray!10}{vector of $n_\text{acc}$ phylogenetically structured random accession-level effects of high light intensity at \ppfdequals{2000} on \aax; $\vec{\beta}_{\mathrm{AA},\text{2000,acc}} \sim \text{MVN}(\vec{0}, \symbf{\Sigma}_{\mathrm{AA},\text{2000,acc}})$}\\
+\hspace{1em}$\vec{\beta}_{\mathrm{AA},\text{sun,acc}}$ & vector of $n_\text{acc}$ phylogenetically structured random accession-level effects of sun treatment on \aax; $\vec{\beta}_{\mathrm{AA},\text{sun,acc}} \sim \text{MVN}(\vec{0}, \symbf{\Sigma}_{\mathrm{AA},\text{sun,acc}})$\\
+\hspace{1em}\cellcolor{gray!10}{$\sigma_{\mathrm{AA},\epsilon,0}$} & \cellcolor{gray!10}{intercept of phylogenetically unstructured residual standard deviation of \aax}\\
+\hspace{1em}$\beta_{\mathrm{AA},\epsilon,2000}$ & effect of high light intensity at \ppfdequals{2000} on phylogenetically unstructured residual standard deviation of \aax{} (log-link scale)\\
+\hspace{1em}\cellcolor{gray!10}{$\beta_{\mathrm{AA},\epsilon,\text{sun}}$} & \cellcolor{gray!10}{effect of sun treatment on phylogenetically unstructured residual standard deviation of \aax{} (log-link scale)}\\
+\hspace{1em}$\sigma_{\mathrm{AA},\text{rep}}$ & standard deviation of random replicate-level effects on \aax\\
+\hspace{1em}\cellcolor{gray!10}{$\alpha_{\mathrm{AA},\text{acc}}$} & \cellcolor{gray!10}{decay rate of phylogenetic covariance in random accession-level effects on \aax}\\
+\hspace{1em}$\sigma^2_{\mathrm{AA},\text{acc}}$ & phylogenetic diffusion rate in random accession-level effects on \aax\\
+\hspace{1em}\cellcolor{gray!10}{$\alpha_{\mathrm{AA},\text{2000,acc}}$} & \cellcolor{gray!10}{decay rate of phylogenetic covariance in random accession-level effects of high light intensity at \ppfdequals{2000} on \aax}\\
+\hspace{1em}$\sigma^2_{\mathrm{AA},\text{2000,acc}}$ & phylogenetic diffusion rate in random accession-level effects of high light intensity at \ppfdequals{2000} on \aax\\
+\hspace{1em}\cellcolor{gray!10}{$\alpha_{\mathrm{AA},\text{sun,acc}}$} & \cellcolor{gray!10}{decay rate of phylogenetic covariance in random accession-level effects of sun treatment on \aax}\\
+\hspace{1em}$\sigma^2_{\mathrm{AA},\text{sun,acc}}$ & phylogenetic diffusion rate in random accession-level effects of sun treatment on \aax\\
+\addlinespace[0.3em]
+\multicolumn{2}{l}{\textbf{effects of native light habitat on accession-level \aax}}\\
+\hspace{1em}\cellcolor{gray!10}{$\beta_{\mathrm{AA},\text{PPFD},0}$} & \cellcolor{gray!10}{intercept of accession-level \aax{} when native \ppfdequals{0}}\\
+\hspace{1em}$\beta_{\mathrm{AA},\text{PPFD},1}$ & slope of native \ppfd{} on accession-level \aax{}\\
+\bottomrule
+
+\end{longtable}
 
@@
 pseudohypo \agcurve{} curves over the range of \gsw{} values where the
-curves overlap. We used the following equation to estimate \aax{} for
-the \(k^{\text{th}}\) leaf at light intensity \(l\) in accession \(m\):
+curves overlap (from \(\min(\log(g_\text{sw}))\) to
+\(\max(\log(g_\text{sw}))\)). The estimate of \aax{} for the
+\(k^{\text{th}}\) leaf at light intensity \(l\) in accession \(m\) is:
 
@@
 \(\Theta_1\) indexed by leaf type (amphi or pseudohypo), leaf replicate,
-light intensity, and accession to set \(\Theta_2\) indexed by \agcurve{}
-curve.
-
-WORKING HERE
+light intensity, and accession to set \(\Theta_2\) indexed by individual
+\agcurve{} curve. This mapping is necessary because the random effects
+structure differs between models of \loggsw on \logA and that of models
+predicting \aax{} described in the next section.
+
+\subparagraph{\texorpdfstring{Effects of light intensity, light
+treatment, and accession on
+\aax}{Effects of light intensity, light treatment, and accession on }}\label{effects-of-light-intensity-light-treatment-and-accession-on}
+
+We tested for effects of light intensity, light treatment, accession,
+and their interactions on \aax. All models included effects of light
+intensity and light treatment, as well as random effects of accession
+and replicate within accession. More complex models included
+interactions between light intensity and light treatment, as well as
+random effects of accession on the effects of light intensity and light
+treatment. We used a weakly informative normal prior with mean \(0\) and
+standard deviation \(10\) for fixed effects light intensity and
+treatment. We used a weakly informative normal prior with mean \(-3\)
+and standard deviation \(5\) for the the random effect standard
+deviation of replicate within accession. We accounted for the
+phylogenetic structure among the random effects of accession using an
+Ornstein-Uhlenbeck (OU) process (\textbf{hansen\_stabilizing\_1997?}).
+The expected covariance between accessions \(i\) and \(j\)
+(\(\text{Cov}(i, j)\) ) is:
+
+\[\text{Cov}(i, j) = \frac{\sigma^2}{2 \alpha} \exp(-\alpha D_{ij})\]
+where \(\sigma^2\) is the variance of the random effect, \(\alpha\) is
+the rate of decay of the covariance with phylogenetic distance,
+\(D_{ij}\), the phylogenetic distance between accessions \(i\) and
+\(j\). We estimated \(\sigma^2 / (2 \alpha)\) and \(\alpha\) as a
+separate parameters and reparameterized them as \(\sigma^2\) and
+\(\alpha\). We used a weakly informative normal prior with mean \(0\)
+and standard deviation \(10\) on OU parameters.
+
+We modeled the residual standard deviation of \aax,
+(i.e.~phylogenetically unstructured variation unaccounted for by
+explanatory variables) on a log-link scale with effects of light
+intensity and light treatment. We used a weakly informative normal prior
+with mean \(-3\) and standard deviation \(5\) on the residual standard
+deviation intercept and a weakly informative normal prior with mean
+\(0\) and standard deviation \(1\) on the effects of light intensity and
+light treatment on the residual standard deviation.
+
+\subparagraph{\texorpdfstring{Effects of native light habitat on
+accession-level
+\aax}{Effects of native light habitat on accession-level }}\label{effects-of-native-light-habitat-on-accession-level}
+
+We tested a linear effect of native light habitat on accession-level
+\aax. In Models 1 and 2, we used the random intercept of \aax{} as a
+response variable; in model 3 we used the random intercept plus the
+random effect of accession at high light intensity; in model 4 we used
+the random intercept plus the random effect of accession in the sun
+treatment; in model 5 we used the random intercept plus the random
+effects of accession at high light intensity and sun treatment. We used
+a weakly informative normal prior with mean \(0\) and standard deviation
+\(1\) for the slope and intercept. We accounted for the phylogenetic
+structure among the model residuals using an (OU) process. We used a
+weakly informative normal prior with mean \(0\) and standard deviation
+\(10\) on OU parameters.
+
+WORKING HERE - TABLE OR SOMETHING SUMMARIZING MODELS? - TABLE
+SUMMARIZING PARAMETERS
 
@@
 procedure.
-
-Base model (need to add indices):
-
-\[\text{log(A)} = \beta_0 + \beta_1 \text{log}(g_\text{sw}) + \beta_2 \text{log}(g_\text{sw}) ^ 2 + \epsilon\]
-Estimate \aax{} by integrating over range of \gsw{} overlap (need to add
-indices for accession, replicate, light intensity):
-
-where \(\text{min(log(}g_\text{sw}))\) and
-\(\text{max(log(}g_\text{sw}))\) are the minimum and maximum values of
-\(\text{log(}g_\text{sw})\) where amphi and pseudohypo curves overlap.
-\Aamphi{} is calculated from parameter estimates
-\(\theta_\text{amphi} \in \{\beta_{0,\text{amphi}}, \beta_{1,\text{amphi}}, \beta_{2,\text{amphi}}\}\);
-\Ahypo{} is calculated from parameter estimates
-\(\theta_\text{hypo} \in \{\beta_{0,\text{hypo}}, \beta_{1,\text{hypo}}, \beta_{2,\text{hypo}}\}\)
-for each curve.
 

### 2024-04-08 ms/test.qmd
added draft of tbl-parameters

@@
 
-df_parameters = read_csv("../data/parameters1.csv", col_types = "cccc", n_max = 6)
+df_parameters = read_csv("../data/parameters1.csv", col_types = "ccccl") |>
+    dplyr::filter(include)
 
@@
     format = "latex",
-    align = c("c", "l"),
+    align = c("l", "l"),
     booktabs = TRUE,
-    escape = FALSE
+    escape = FALSE,
+    longtable = TRUE
   ) |>
@@
                         max(which(df_parameters$par_type == "ags")),
-                        escape = FALSE) #|>
-                        # kableExtra::pack_rows("x", 2, 2)
+                        escape = FALSE) |>
+  kableExtra::pack_rows("\\aax{} for each light intensity with leaf using \\agcurve{} curve parameters",
+                        min(which(df_parameters$par_type == "aa_est")),
+                        max(which(df_parameters$par_type == "aa_est")),
+                        escape = FALSE) |>
+  kableExtra::pack_rows("effects light intensity, light treatments, and accession on \\aax",
+                        min(which(df_parameters$par_type == "aa")),
+                        max(which(df_parameters$par_type == "aa")),
+                        escape = FALSE) |>
+  kableExtra::pack_rows("effects of native light habitat on accession-level \\aax",
+                        min(which(df_parameters$par_type == "ppfd_aa")),
+                        max(which(df_parameters$par_type == "ppfd_aa")),
+                        escape = FALSE)
 

### 2024-04-08 r/311_summarize-ou.R

@@ -1,0 +1,5 @@
+# Summarize OU parameters
+# Not sure what goes here, just leaving notes on how to map from {eta^2, rho^2} -> {simga^2, alpha} OU parameterization
+# 
+# alpha = rho ^ 2
+# sigma ^ 2 = 2 * eta ^ 2 * rho ^ 2

### 2024-04-09 ms/ms.qmd
drafting tbl-predictions

@@
 - TABLE SUMMARIZING PARAMETERS
+model | fixed effects | phylogenetic random effects | response to native PPFD
+1 | light intensity | varying intercept among accessions | AA_0,acc
+1 | light treatment | | 
+2 | light intensity | varying intercept among accessions | AA_0,acc
+2 | light treatment | | 
+2 | light intensity x light treatment | | 
+3 | light intensity | varying intercept among accessions | AA_2000,acc
+3 | light treatment | varying effect of high light intensity among accessions | 
+4 | light intensity | varying intercept among accessions | AA_sun,acc
+4 | light treatment | varying effect of sun treatment among accessions | 
+5 | light intensity | varying intercept among accessions | AA_2000,high,acc
+5 | light treatment | varying effect of high light intensity among accessions | 
+5 | light intensity x light treatment | varying effect of sun treatment among accessions |
 

### 2024-04-09 ms/test.qmd
drafting tbl-predictions

--- 2024-04-06 ms/test.qmd
@@
 
-Manual curation of a data set this size in a principled, consistent manner is not feasible.
-Therefore, we automated data cleaning using custom *R* scripts.
-Cleaning is divided into six sequential steps \autoref{tbl-parameters}.
+blah \autoref{tbl-predictions}.
 
-```{r, parameters, echo=FALSE,message=FALSE,warning=FALSE}
-#| label: tbl-parameters
+```{r, predictions, echo=FALSE,message=FALSE,warning=FALSE}
+#| label: tbl-predictions
 #| tbl-cap: "caption"
 
-df_parameters = read_csv("../data/parameters1.csv", col_types = "cccc", n_max = 6)
-
 df_parameters |>
-  dplyr::select(Parameter = latex, Description = description) |>
+  tribble(
+    Parameter = latex, Description = description) |>
   knitr::kable(
     format = "latex",
-    align = c("c", "l"),
+    align = c("l", "l"),
     booktabs = TRUE,
-    escape = FALSE
+    escape = FALSE,
+    longtable = TRUE
   ) |>

### 2024-04-09 ms/ms.qmd
drafted tbl-models

--- 2024-04-08 ms/ms.qmd
@@
 We estimated parameters and confidence intervals as the median and 95% quantile intervals of the posterior, respectively.
+#### *Stan* models
+
+- high level overview of differences among models \autoref{tbl-models}
+- next section describes common elements of all models and additional details about parameters
+
+```{r, models, echo=FALSE,message=FALSE,warning=FALSE}
+#| label: tbl-models
+#| tbl-cap: "Summary of differences among competing models of how \\aax{} varies with light intensity, light treatment, and among accessions as a function of native \\ppfd.
+The models are numbered from simpler to more complex.
+All models include fixed effects of light intensity and light treatment; some models include interactions between.
+All models include a phylogenetic random effect of accession on \\aax; some models include varying effects of light intensity and light treatment among accessions.
+The last column indicates the accession-level \\aax{} variable we used as a response to native \\ppfd."
+
+tribble(
+    ~"Model", ~"Fixed effects", ~"Phylogenetic random effects", ~"Response to native \\ppfd",
+    "1", "light intensity", "varying intercept among accessions", "$\\mathrm{AA}_{0, \\text{acc}}$",
+    "1", "light treatment", "", "",
+    "2", "light intensity", "varying intercept among accessions", "$\\mathrm{AA}_{0, \\text{acc}}$",
+    "2", "light treatment", "", "",
+    "2", "intensity $\\times$ treatment", "", "",
+    "3", "light intensity", "varying intercept among accessions", "$\\mathrm{AA}_{2000, \\text{acc}}$",
+    "3", "light treatment", "varying effect of high light intensity among accessions", "",
+    "4", "light intensity", "varying intercept among accessions", "$\\mathrm{AA}_{\\text{sun, acc}}$",
+    "4", "light treatment", "varying effect of sun treatment among accessions", "",
+    "5", "light intensity", "varying intercept among accessions", "$\\mathrm{AA}_{2000, \\text{sun, acc}}$",
+    "5", "light treatment", "varying effect of high light intensity among accessions", "",
+    "5", "intensity $\\times$ treatment", "varying effect of sun treatment among accessions", ""
+  ) |>
+  knitr::kable(
+    format = "latex",
+    align = c("c", "l", "l", "l"),
+    booktabs = TRUE,
+    escape = FALSE,
+    longtable = FALSE
+  ) |>
+    kableExtra::collapse_rows(columns = 1, latex_hline = "major", row_group_label_position = "first") |>
+  kableExtra::column_spec(1, width = "0.3in") |>
+  kableExtra::column_spec(2, width = "1.3in") |>
+  kableExtra::column_spec(3, width = "3.4in") |>
+  kableExtra::column_spec(4, width = "1in")
+
+```
 
@@
 
-WORKING HERE - TABLE OR SOMETHING SUMMARIZING MODELS?
-- TABLE SUMMARIZING PARAMETERS
-
 We estimated these parameters for each leaf using the posterior distribution of the parameters from the \agcurve{} curve fitting procedure.

### 2024-04-09 ms/ms.tex
drafted tbl-models

@@
 as the median and 95\% quantile intervals of the posterior,
-respectively.
+respectively. \#\#\#\# \emph{Stan} models
+
+\begin{itemize}
+\tightlist
+\item
+  high level overview of differences among models \autoref{tbl-models}
+\item
+  next section describes common elements of all models and additional
+  details about parameters
+\end{itemize}
+
+\begin{table}
+
+\caption{\label{tbl-models}Summary of differences among competing models
+of how \aax{} varies with light intensity, light treatment, and among
+accessions as a function of native \ppfd. The models are numbered from
+simpler to more complex. All models include fixed effects of light
+intensity and light treatment; some models include interactions between.
+All models include a phylogenetic random effect of accession on \aax;
+some models include varying effects of light intensity and light
+treatment among accessions. The last column indicates the
+accession-level \aax{} variable we used as a response to native \ppfd.}
+
+\centering{
+
+\begin{tabular}{>{\centering\arraybackslash}p{0.3in}>{\raggedright\arraybackslash}p{1.3in}>{\raggedright\arraybackslash}p{3.4in}>{\raggedright\arraybackslash}p{1in}}
+\toprule
+Model & Fixed effects & Phylogenetic random effects & Response to native \ppfd\\
+\midrule
+1 & light intensity & varying intercept among accessions & $\mathrm{AA}_{0, \text{acc}}$\\
+
+ & light treatment &  & \\
+\cmidrule{1-4}
+2 & light intensity & varying intercept among accessions & $\mathrm{AA}_{0, \text{acc}}$\\
+
+ & light treatment &  & \\
+
+ & intensity $\times$ treatment &  & \\
+\cmidrule{1-4}
+3 & light intensity & varying intercept among accessions & $\mathrm{AA}_{2000, \text{acc}}$\\
+
+ & light treatment & varying effect of high light intensity among accessions & \\
+\cmidrule{1-4}
+4 & light intensity & varying intercept among accessions & $\mathrm{AA}_{\text{sun, acc}}$\\
+
+ & light treatment & varying effect of sun treatment among accessions & \\
+\cmidrule{1-4}
+5 & light intensity & varying intercept among accessions & $\mathrm{AA}_{2000, \text{sun, acc}}$\\
+
+ & light treatment & varying effect of high light intensity among accessions & \\
+
+ & intensity $\times$ treatment & varying effect of sun treatment among accessions & \\
+\bottomrule
+\end{tabular}
+
+}
+
+\end{table}%
 
@@
 
-WORKING HERE - TABLE OR SOMETHING SUMMARIZING MODELS? - TABLE
-SUMMARIZING PARAMETERS
-
 We estimated these parameters for each leaf using the posterior

### 2024-04-09 ms/test.qmd
drafted tbl-models

--- 2024-04-06 ms/test.qmd
@@
 
-Manual curation of a data set this size in a principled, consistent manner is not feasible.
-Therefore, we automated data cleaning using custom *R* scripts.
-Cleaning is divided into six sequential steps \autoref{tbl-parameters}.
+blah \autoref{tbl-models}.
 
-```{r, parameters, echo=FALSE,message=FALSE,warning=FALSE}
-#| label: tbl-parameters
-#| tbl-cap: "caption"
+```{r, models, echo=FALSE,message=FALSE,warning=FALSE}
+#| label: tbl-models
+#| tbl-cap: "Summary of differences among competing models of how \\aax{} varies with light intensity, light treatment, and among accessions as a function of native \\ppfd.
+The models are numbered from simpler to more complex.
+All models include fixed effects of light intensity and light treatment; some models include interactions between.
+All models include a phylogenetic random effect of accession on \\aax; some models include varying effects of light intensity and light treatment among accessions.
+The last column indicates the accession-level \\aax{} variable we used as a response to native \\ppfd."
 
-df_parameters = read_csv("../data/parameters1.csv", col_types = "cccc", n_max = 6)
-
-df_parameters |>
-  dplyr::select(Parameter = latex, Description = description) |>
+tribble(
+    ~"Model", ~"Fixed effects", ~"Phylogenetic random effects", ~"Response to native \\ppfd",
+    "1", "light intensity", "varying intercept among accessions", "$\\mathrm{AA}_{0, \\text{acc}}$",
+    "1", "light treatment", "", "",
+    "2", "light intensity", "varying intercept among accessions", "$\\mathrm{AA}_{0, \\text{acc}}$",
+    "2", "light treatment", "", "",
+    "2", "intensity $\\times$ treatment", "", "",
+    "3", "light intensity", "varying intercept among accessions", "$\\mathrm{AA}_{2000, \\text{acc}}$",
+    "3", "light treatment", "varying effect of high light intensity among accessions", "",
+    "4", "light intensity", "varying intercept among accessions", "$\\mathrm{AA}_{\\text{sun, acc}}$",
+    "4", "light treatment", "varying effect of sun treatment among accessions", "",
+    "5", "light intensity", "varying intercept among accessions", "$\\mathrm{AA}_{2000, \\text{sun, acc}}$",
+    "5", "light treatment", "varying effect of high light intensity among accessions", "",
+    "5", "intensity $\\times$ treatment", "varying effect of sun treatment among accessions", ""
+  ) |>
   knitr::kable(
     format = "latex",
-    align = c("c", "l"),
+    align = c("c", "l", "l", "l"),
     booktabs = TRUE,
-    escape = FALSE
+    escape = FALSE,
+    longtable = FALSE
   ) |>
-  kableExtra::kable_styling(latex_options = c("striped", "no_vertical_lines")) |>
-  kableExtra::column_spec(1, width = "1in") |>
-  kableExtra::column_spec(2, width = "5in") |>
-  kableExtra::pack_rows("\\agcurve{} curve parameters",
-                        min(which(df_parameters$par_type == "ags")),
-                        max(which(df_parameters$par_type == "ags")),
-                        escape = FALSE) #|>
-                        # kableExtra::pack_rows("x", 2, 2)
+    kableExtra::collapse_rows(columns = 1, latex_hline = "major", row_group_label_position = "first") |>
+  kableExtra::column_spec(1, width = "0.3in") |>
+  kableExtra::column_spec(2, width = "1.3in") |>
+  kableExtra::column_spec(3, width = "3.4in") |>
+  kableExtra::column_spec(4, width = "1in")
 

### 2024-04-10 ms/ms.qmd
revised structure of data analysis section

@@
 
-```{r load-objects, echo=FALSE, eval=TRUE}
+```{r load-objects, echo=FALSE, eval=TRUE, message=FALSE, warning=FALSE}
 library(dplyr)
@@
 library(tibble)
+
+source("../r/functions.R")
 
@@
 aq_r2_threshold = aa_args$aq_r2_threshold
+
+### Cleaning AQ curves
+
+### Bayesian data analysis
+n_model = 5 |>
+  number_to_word()
 
@@
 Therefore, we automated data cleaning using custom *R* scripts.
-Cleaning is divided into six sequential steps \autoref{tbl-cleaning1}.
+Cleaning is divided into six sequential steps (\autoref{tbl-cleaning1}).
 
@@
 Therefore, we automated data cleaning using custom *R* scripts.
-Cleaning is divided into two sequential steps \autoref{tbl-cleaning2}.
+Cleaning is divided into two sequential steps (\autoref{tbl-cleaning2}).
 
@@
 
-### Data analysis
+### Bayesian data analysis in *Stan*
 
-#### General points (not sure where this goes yet)
-
-- models fit in Stan/brms
-- we chose the number of chains, warmup and sampling iterations, and maximum treedepth so that parameter estimates converged ($\hat{R} < 1.01$ [@gelman_inference_1992]) and the effective sample size (ESS) for each parameter was $> 10^3$.
-
-We fit Bayesian models with MCMC sampling in the probabilistic programming language *Stan* [@stan_development_team_stan_2023].
-We used CmdStan version `r fit_aa$version$cmdstan` and **cmdstanr** version `r fit_aa$version$cmdstanr` [@gabry_cmdstanr_2023] to interface with *R* version `r getRversion()` [@r_core_team_r:_2023].
-We sampled the posterior distribution from `r dim_fit_aa[2]` chains with `r dim_fit_aa[1]` iterations each after `r dim_fit_aa[1]` warmup iterations per chain.
-We estimated parameters and confidence intervals as the median and 95% quantile intervals of the posterior, respectively.
-#### *Stan* models
-
-- high level overview of differences among models \autoref{tbl-models}
-- next section describes common elements of all models and additional details about parameters
+We fit `r n_model` models to test predictions of competing hypotheses about why amphistomy advantage (\aax) might be greater for leaves in sunny, open habitats.
+This section provides an overview of differences among models (\autoref{tbl-models}).
+The next sections describe how we fit models in *Stan*, all model parameters and priors, and specific predictions about parameter values for each hypothesis.
 
@@
 
-#### Parameter estimation
+#### Fitting models in *Stan*
+
+We fit Bayesian models with MCMC sampling in the probabilistic programming language *Stan* [@stan_development_team_stan_2023].
+We used CmdStan version `r fit_aa$version$cmdstan` and **cmdstanr** version `r fit_aa$version$cmdstanr` [@gabry_cmdstanr_2023] to interface with *R* version `r getRversion()` [@r_core_team_r:_2023].
+We sampled the posterior distribution from `r dim_fit_aa[2]` chains with `r dim_fit_aa[1]` iterations each after `r dim_fit_aa[1]` warmup iterations per chain.
+We estimated parameters and confidence intervals as the median and 95% quantile intervals of the posterior, respectively.
+We chose the number of chains, warmup and sampling iterations, and maximum treedepth so that parameter estimates converged ($\hat{R} < 1.01$ [@gelman_inference_1992]) and the effective sample size (ESS) for each parameter was $> 10^3$.
+
+#### Parameter estimation and priors
 

### 2024-04-10 ms/ms.tex
revised structure of data analysis section

@@
 
-\begin{verbatim}
-
-Attaching package: 'dplyr'
-\end{verbatim}
-
-\begin{verbatim}
-The following objects are masked from 'package:stats':
-
-    filter, lag
-\end{verbatim}
-
-\begin{verbatim}
-The following objects are masked from 'package:base':
-
-    intersect, setdiff, setequal, union
-\end{verbatim}
-
 \subsection{Notes}\label{notes}
@@
 data cleaning using custom \emph{R} scripts. Cleaning is divided into
-six sequential steps \autoref{tbl-cleaning1}.
+six sequential steps (\autoref{tbl-cleaning1}).
 
@@
 using custom \emph{R} scripts. Cleaning is divided into two sequential
-steps \autoref{tbl-cleaning2}.
+steps (\autoref{tbl-cleaning2}).
 
@@
 
-\subsubsection{Data analysis}\label{data-analysis}
-
-\paragraph{General points (not sure where this goes
-yet)}\label{general-points-not-sure-where-this-goes-yet}
-
-\begin{itemize}
-\tightlist
-\item
-  models fit in Stan/brms
-\item
-  we chose the number of chains, warmup and sampling iterations, and
-  maximum treedepth so that parameter estimates converged
-  (\(\hat{R} < 1.01\) (\emph{10})) and the effective sample size (ESS)
-  for each parameter was \(> 10^3\).
-\end{itemize}
-
-We fit Bayesian models with MCMC sampling in the probabilistic
-programming language \emph{Stan} (\emph{11}). We used CmdStan version
-2.33.1 and \textbf{cmdstanr} version 0.7.1 (\emph{12}) to interface with
-\emph{R} version 4.3.2 (\emph{13}). We sampled the posterior
-distribution from 4 chains with 1000 iterations each after 1000 warmup
-iterations per chain. We estimated parameters and confidence intervals
-as the median and 95\% quantile intervals of the posterior,
-respectively. \#\#\#\# \emph{Stan} models
-
-\begin{itemize}
-\tightlist
-\item
-  high level overview of differences among models \autoref{tbl-models}
-\item
-  next section describes common elements of all models and additional
-  details about parameters
-\end{itemize}
+\subsubsection{\texorpdfstring{Bayesian data analysis in
+\emph{Stan}}{Bayesian data analysis in Stan}}\label{bayesian-data-analysis-in-stan}
+
+We fit five models to test predictions of competing hypotheses about why
+amphistomy advantage (\aax) might be greater for leaves in sunny, open
+habitats. This section provides an overview of differences among models
+(\autoref{tbl-models}). The next sections describe how we fit models in
+\emph{Stan}, all model parameters and priors, and specific predictions
+about parameter values for each hypothesis.
 
@@
 
-\paragraph{Parameter estimation}\label{parameter-estimation}
+\paragraph{\texorpdfstring{Fitting models in
+\emph{Stan}}{Fitting models in Stan}}\label{fitting-models-in-stan}
+
+We fit Bayesian models with MCMC sampling in the probabilistic
+programming language \emph{Stan} (\emph{10}). We used CmdStan version
+2.33.1 and \textbf{cmdstanr} version 0.7.1 (\emph{11}) to interface with
+\emph{R} version 4.3.2 (\emph{12}). We sampled the posterior
+distribution from 4 chains with 1000 iterations each after 1000 warmup
+iterations per chain. We estimated parameters and confidence intervals
+as the median and 95\% quantile intervals of the posterior,
+respectively. We chose the number of chains, warmup and sampling
+iterations, and maximum treedepth so that parameter estimates converged
+(\(\hat{R} < 1.01\) (\emph{13})) and the effective sample size (ESS) for
+each parameter was \(> 10^3\).
+
+\paragraph{Parameter estimation and
+priors}\label{parameter-estimation-and-priors}
 
@@
 \multicolumn{2}{l}{\textbf{effects of native light habitat on accession-level \aax}}\\
-\hspace{1em}\cellcolor{gray!10}{$\beta_{\mathrm{AA},\text{PPFD},0}$} & \cellcolor{gray!10}{intercept of accession-level \aax{} when native \ppfdequals{0}}\\
-\hspace{1em}$\beta_{\mathrm{AA},\text{PPFD},1}$ & slope of native \ppfd{} on accession-level \aax{}\\
+\hspace{1em}\hspace{1em}\cellcolor{gray!10}{$\beta_{\mathrm{AA},\text{PPFD},0}$} & \cellcolor{gray!10}{intercept of accession-level \aax{} when native \ppfdequals{0}}\\
+\hspace{1em}\hspace{1em}$\beta_{\mathrm{AA},\text{PPFD},1}$ & slope of native \ppfd{} on accession-level \aax{}\\
+\hspace{1em}\cellcolor{gray!10}{$\alpha_{\mathrm{AA},\text{PPFD}}$} & \cellcolor{gray!10}{decay rate of phylogenetic covariance in residuals of model testing effect of native \ppfd{} on accession-level \aax}\\
+\hspace{1em}$\sigma^2_{\mathrm{AA},\text{PPFD}}$ & phylogenetic diffusion rate in residuals of model testing effect of native \ppfd{} on accession-level \aax\\
 \bottomrule
@@
 phylogenetic structure among the random effects of accession using an
-Ornstein-Uhlenbeck (OU) process (\textbf{hansen\_stabilizing\_1997?}).
-The expected covariance between accessions \(i\) and \(j\)
-(\(\text{Cov}(i, j)\) ) is:
+Ornstein-Uhlenbeck (OU) process (\emph{14}). The expected covariance
+between accessions \(i\) and \(j\) (\(\text{Cov}(i, j)\) ) is:
 
@@
 light intensity and light treatment using LOOIC with the \textit{R}
-package \textbf{loo} version 2.7.0 (\emph{14}). These comparisons
+package \textbf{loo} version 2.7.0 (\emph{15}). These comparisons
 allowed us to test the main predictions of each hypothesis and for a
@@
 
-\bibitem[\citeproctext]{ref-gelman_inference_1992}
+\bibitem[\citeproctext]{ref-stan_development_team_stan_2023}
 \CSLLeftMargin{10. }%
-\CSLRightInline{A. Gelman, D. B. Rubin, Inference from iterative
-simulation using multiple sequences. \emph{Statistical Science}
-\textbf{7}, 457--472 (1992).}
-
-\bibitem[\citeproctext]{ref-stan_development_team_stan_2023}
-\CSLLeftMargin{11. }%
 \CSLRightInline{Stan Development Team, \emph{Stan {Modeling} {Language}
@@
 \bibitem[\citeproctext]{ref-gabry_cmdstanr_2023}
-\CSLLeftMargin{12. }%
+\CSLLeftMargin{11. }%
 \CSLRightInline{J. Gabry, R. Češnovar, \emph{Cmdstanr: {R} {Interface}
@@
 \bibitem[\citeproctext]{ref-r_core_team_r:_2023}
-\CSLLeftMargin{13. }%
+\CSLLeftMargin{12. }%
 \CSLRightInline{R Core Team, \emph{R: {A} {Language} and {Environment}
@@
 
+\bibitem[\citeproctext]{ref-gelman_inference_1992}
+\CSLLeftMargin{13. }%
+\CSLRightInline{A. Gelman, D. B. Rubin, Inference from iterative
+simulation using multiple sequences. \emph{Statistical Science}
+\textbf{7}, 457--472 (1992).}
+
+\bibitem[\citeproctext]{ref-hansen_stabilizing_1997}
+\CSLLeftMargin{14. }%
+\CSLRightInline{T. F. Hansen,
+\href{https://doi.org/10.1111/j.1558-5646.1997.tb01457.x}{Stabilizing
+selection and the comparative analysis of adaptation}. \emph{Evolution}
+\textbf{51}, 1341--1351 (1997).}
+
 \bibitem[\citeproctext]{ref-vehtari_practical_2017}
-\CSLLeftMargin{14. }%
+\CSLLeftMargin{15. }%
 \CSLRightInline{A. Vehtari, A. Gelman, J. Gabry,

### 2024-04-10 r/functions.R

@@ -1,3 +1,13 @@
+# Utility functions ----
+number_to_word = function(num) {
+  if (num >= 0 && num <= 10) {
+    words = c("zero", "one", "two", "three", "four", "five", "six", "seven", "eight", "nine", "ten")
+    return(words[num + 1])
+  } else {
+    return(num)
+  }
+}
+
 # 0. Import data ----
 add_to_stats = function(...) {
   old = read_rds("objects/aa_stats.rds")

### 2024-04-10 ms/ms.qmd
working on predictions subsection

@@
     keep-tex: true  
-    toc: false
+    toc: true
+number-sections: true
 header-includes: |
@@
 
-#### Fitting models in *Stan*
+#### Fitting models in *Stan* {#sec-fitting}
 
@@
 
-Within the $k^{\text{th}}$ leaf, we estimated \aax{} for each light intensity by integrating the difference in \logA between the amphi and pseudohypo \agcurve{} curves over the range of \gsw{} values where the curves overlap (from $\min(\log(g_\text{sw}))$ to $\max(\log(g_\text{sw}))$).
+Within the $k^{\text{th}}$ leaf, we estimated \aax{} for each light intensity by integrating the difference in \logA{} between the amphi and pseudohypo \agcurve{} curves over the range of \gsw{} values where the curves overlap (from $\min(\log(g_\text{sw}))$ to $\max(\log(g_\text{sw}))$).
 The estimate of \aax{} for the $k^{\text{th}}$ leaf at light intensity $l$ in accession $m$ is:
@@
 The function $f : \Theta_1 \rightarrow \Theta_2$ maps the set $\Theta_1$ indexed by leaf type (amphi or pseudohypo), leaf replicate, light intensity, and accession to set $\Theta_2$ indexed by individual \agcurve{} curve.
-This mapping is necessary because the random effects structure differs between models of \loggsw on \logA and that of models predicting \aax{} described in the next section.
+This mapping is necessary because the random effects structure differs between models of \loggsw{} on \logA{} and that of models predicting \aax{} described in the next section.
 
@@
 
-We estimated these parameters for each leaf using the posterior distribution of the parameters from the \agcurve{} curve fitting procedure.
+#### Predictions {#sec-predictions}
 
-#### Predictions
+The assimilatory, plastic, and constitutive hypotheses make different predictions about the relationship between \aax{} and light intensity, light treatment, and native \ppfd{} among accessions.
+Since these hypotheses are not mutually exclusive, we describe how we assessed support for one, two, or all three hypotheses simultaneously in \autoref{tbl-predictions}.
+
+We evaluated predictions using a combination of parameter estimation and model selection.
+In all cases, we estimated parameters and confidence intervals as described in @sec-fitting.
+If the 95% confidence intervals for a parameter did not overlap zero, we considered the parameter to be significantly different from zero.
+We used leave-one-out cross-validation (LOOIC) to compare the fit of models (\autoref{tbl-models}) using the *R* package **loo** version `r packageVersion("loo")` [@vehtari_practical_2017] to calculate LOOIC values.
+We considered models with two standard errors of the mean lower LOOIC value to be a better fit to the data; we considered models with LOOIC values within two standard errors of the mean to be have similar support.
+
+```{r, predictions, eval=FALSE, echo=FALSE,message=FALSE,warning=FALSE}
+#| label: tbl-predictions
+#| tbl-cap: "caption"
 
@@
 
-We tested competing predictions through a combination of model selection and parameter estimation from the posterior of selected models.
-We started with a base model that includes fixed effects of light intensity and light treatment on \aax, a random effect of accession on \aax, and a linear regression of PAR on accession-level \aax.
-We next compared the fit of the base model to a more complex with an interaction between light intensity and light treatment using LOOIC with the \textit{R} package **loo** version `r packageVersion("loo")` [@vehtari_practical_2017].
-These comparisons allowed us to test the main predictions of each hypothesis and for a reinforcing interactions between assimilatory and plastic hypotheses (tab:predictions).
-However, these models cannot test predictions when there is an interaction with the constitutive hypothesis, because this hypothesis posits accession-levels variation in responses to light intensity and/or light treatment.
-Therefore, we tested for significant variation among accessions in light intensity and light treatment by comparing LOOIC values of models with and without random interactions between these factors and accession.
-
 | Hypothesis   | Prediction | Model |
@@
 
+```
+
 ## Acknowledgements

### 2024-04-10 ms/ms.tex
working on predictions subsection

@@
 \setlength{\emergencystretch}{3em} % prevent overfull lines
-\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
+\setcounter{secnumdepth}{5}
 % Make \paragraph and \subparagraph free-standing
@@
 
-\subsection{Notes}\label{notes}
+\renewcommand*\contentsname{Table of contents}
+{
+\hypersetup{linkcolor=}
+\setcounter{tocdepth}{3}
+\tableofcontents
+}
+\section{Notes}\label{notes}
 
@@
 
-\subsection{Authors}\label{authors}
+\section{Authors}\label{authors}
 
@@
 
-\subsection{Main}\label{main}
+\section{Main}\label{main}
 
@@
 
-\subsection{Materials and Methods}\label{materials-and-methods}
+\section{Materials and Methods}\label{materials-and-methods}
 
@@
 
-\subsubsection{Accessions}\label{accessions}
+\subsection{Accessions}\label{accessions}
 
@@
 
-\subsubsection{Plant growth conditions}\label{plant-growth-conditions}
+\subsection{Plant growth conditions}\label{plant-growth-conditions}
 
@@
 
-\paragraph{Germination and seedling
+\subsubsection{Germination and seedling
 stage}\label{germination-and-seedling-stage}
@@
 
-\paragraph{Light treatments}\label{light-treatments}
+\subsubsection{Light treatments}\label{light-treatments}
 
@@
 
-\subsubsection{Leaf trait measurements}\label{leaf-trait-measurements}
+\subsection{Leaf trait measurements}\label{leaf-trait-measurements}
 
@@
 
-\paragraph{Amphistomy advantage}\label{amphistomy-advantage}
+\subsubsection{Amphistomy advantage}\label{amphistomy-advantage}
 
@@
 
-\paragraph{\texorpdfstring{Light-response (\aqcurve)
+\subsubsection{\texorpdfstring{Light-response (\aqcurve)
 curves}{Light-response () curves}}\label{light-response-curves}
@@
 
-\paragraph{Stomatal anatomy}\label{stomatal-anatomy}
+\subsubsection{Stomatal anatomy}\label{stomatal-anatomy}
 
@@
 
-\paragraph{Leaf mass per area}\label{leaf-mass-per-area}
+\subsubsection{Leaf mass per area}\label{leaf-mass-per-area}
 
@@
 
-\subsubsection{\texorpdfstring{Cleaning \agcurve{}
+\subsection{\texorpdfstring{Cleaning \agcurve{}
 curves}{Cleaning  curves}}\label{cleaning-curves}
@@
 
-\paragraph{Remove unreliable and unusable data
+\subsubsection{Remove unreliable and unusable data
 points}\label{remove-unreliable-and-unusable-data-points}
@@
 
-\paragraph{\texorpdfstring{Remove hysteretic portion of \agcurve{}
+\subsubsection{\texorpdfstring{Remove hysteretic portion of \agcurve{}
 curves at low
@@
 
-\paragraph{\texorpdfstring{Remove outliers within each \agcurve{}
+\subsubsection{\texorpdfstring{Remove outliers within each \agcurve{}
 curve}{Remove outliers within each  curve}}\label{remove-outliers-within-each-curve}
@@
 
-\paragraph{\texorpdfstring{Thin redundant data points within each
+\subsubsection{\texorpdfstring{Thin redundant data points within each
 \agcurve{}
@@
 
-\paragraph{\texorpdfstring{Remove replicates with no overlap between
+\subsubsection{\texorpdfstring{Remove replicates with no overlap between
 amphi and pseudohypo \agcurve{}
@@
 
-\paragraph{\texorpdfstring{Trim extreme \aax{}
+\subsubsection{\texorpdfstring{Trim extreme \aax{}
 values}{Trim extreme  values}}\label{trim-extreme-values}
@@
 
-\subsubsection{\texorpdfstring{Cleaning \aqcurve{}
+\subsection{\texorpdfstring{Cleaning \aqcurve{}
 curves}{Cleaning  curves}}\label{cleaning-curves-1}
@@
 
-\paragraph{\texorpdfstring{Remove outliers within each \aqcurve{}
+\subsubsection{\texorpdfstring{Remove outliers within each \aqcurve{}
 curve}{Remove outliers within each  curve}}\label{remove-outliers-within-each-curve-1}
@@
 
-\paragraph{\texorpdfstring{Remove \aqcurve{} curves with poor
+\subsubsection{\texorpdfstring{Remove \aqcurve{} curves with poor
 fit}{Remove  curves with poor fit}}\label{remove-curves-with-poor-fit}
@@
 
-\subsubsection{\texorpdfstring{Bayesian data analysis in
+\subsection{\texorpdfstring{Bayesian data analysis in
 \emph{Stan}}{Bayesian data analysis in Stan}}\label{bayesian-data-analysis-in-stan}
@@
 
-\paragraph{\texorpdfstring{Fitting models in
-\emph{Stan}}{Fitting models in Stan}}\label{fitting-models-in-stan}
+\subsubsection{\texorpdfstring{Fitting models in
+\emph{Stan}}{Fitting models in Stan}}\label{sec-fitting}
 
@@
 
-\paragraph{Parameter estimation and
+\subsubsection{Parameter estimation and
 priors}\label{parameter-estimation-and-priors}
@@
 
-\subparagraph{\texorpdfstring{\agcurve{} curve
+\paragraph{\texorpdfstring{\agcurve{} curve
 parameters}{ curve parameters}}\label{curve-parameters}
@@
 
-\subparagraph{\texorpdfstring{\aax{} for each light intensity with leaf
+\paragraph{\texorpdfstring{\aax{} for each light intensity with leaf
 using \agcurve{} curve
@@
 Within the \(k^{\text{th}}\) leaf, we estimated \aax{} for each light
-intensity by integrating the difference in \logA between the amphi and
+intensity by integrating the difference in \logA{} between the amphi and
 pseudohypo \agcurve{} curves over the range of \gsw{} values where the
@@
 \agcurve{} curve. This mapping is necessary because the random effects
-structure differs between models of \loggsw on \logA and that of models
-predicting \aax{} described in the next section.
-
-\subparagraph{\texorpdfstring{Effects of light intensity, light
-treatment, and accession on
+structure differs between models of \loggsw{} on \logA{} and that of
+models predicting \aax{} described in the next section.
+
+\paragraph{\texorpdfstring{Effects of light intensity, light treatment,
+and accession on
 \aax}{Effects of light intensity, light treatment, and accession on }}\label{effects-of-light-intensity-light-treatment-and-accession-on}
@@
 
-\subparagraph{\texorpdfstring{Effects of native light habitat on
+\paragraph{\texorpdfstring{Effects of native light habitat on
 accession-level
@@
 
-We estimated these parameters for each leaf using the posterior
-distribution of the parameters from the \agcurve{} curve fitting
-procedure.
-
-\paragraph{Predictions}\label{predictions}
+\subsubsection{Predictions}\label{sec-predictions}
+
+WORKING HERE TO ADD PREDICTIONS AND TABLE
+
+The assimilatory, plastic, and constitutive hypotheses make different
+predictions about the relationship between \aax{} and light intensity,
+light treatment, and native \ppfd{} among accessions. Since these
+hypotheses are not mutually exclusive, we describe how we assessed
+support for one, two, or all three hypotheses simultaneously.
+
+We tested competing predictions through a combination of model selection
+and parameter estimation from the posterior of fitted models. In all
+cases, we estimated parameters and confidence intervals as described in
+Section~\ref{sec-fitting}.
+
+We started with a base model that includes fixed effects of light
+intensity and light treatment on \aax, a random effect of accession on
+\aax, and a linear regression of PAR on accession-level \aax. We next
+compared the fit of the base model to a more complex with an interaction
+between light intensity and light treatment using LOOIC with the
+\textit{R} package \textbf{loo} version 2.7.0 (\emph{15}). These
+comparisons allowed us to test the main predictions of each hypothesis
+and for a reinforcing interactions between assimilatory and plastic
+hypotheses (tab:predictions). However, these models cannot test
+predictions when there is an interaction with the constitutive
+hypothesis, because this hypothesis posits accession-levels variation in
+responses to light intensity and/or light treatment. Therefore, we
+tested for significant variation among accessions in light intensity and
+light treatment by comparing LOOIC values of models with and without
+random interactions between these factors and accession.
 
@@
 \end{itemize}
-
-We tested competing predictions through a combination of model selection
-and parameter estimation from the posterior of selected models. We
-started with a base model that includes fixed effects of light intensity
-and light treatment on \aax, a random effect of accession on \aax, and a
-linear regression of PAR on accession-level \aax. We next compared the
-fit of the base model to a more complex with an interaction between
-light intensity and light treatment using LOOIC with the \textit{R}
-package \textbf{loo} version 2.7.0 (\emph{15}). These comparisons
-allowed us to test the main predictions of each hypothesis and for a
-reinforcing interactions between assimilatory and plastic hypotheses
-(tab:predictions). However, these models cannot test predictions when
-there is an interaction with the constitutive hypothesis, because this
-hypothesis posits accession-levels variation in responses to light
-intensity and/or light treatment. Therefore, we tested for significant
-variation among accessions in light intensity and light treatment by
-comparing LOOIC values of models with and without random interactions
-between these factors and accession.
 
@@
 
-\subsection{Acknowledgements}\label{acknowledgements}
+\section{Acknowledgements}\label{acknowledgements}
 
@@
 
-\subsection*{References}\label{references}
-\addcontentsline{toc}{subsection}{References}
+\section*{References}\label{references}
+\addcontentsline{toc}{section}{References}
 

### 2024-04-10 ms/ms.qmd
drafted predictions section

@@
 
-#### Parameter estimation and priors
+#### Parameter estimation and priors {#sec-parameters}
 
@@
                         escape = FALSE) |>
-  kableExtra::pack_rows("effects light intensity, light treatments, and accession on \\aax",
+  kableExtra::pack_rows("effects of light intensity, light treatments, and accession on \\aax",
                         min(which(df_parameters$par_type == "aa")),
@@
 More complex models included interactions between light intensity and light treatment, as well as random effects of accession on the effects of light intensity and light treatment.
-We used a weakly informative normal prior with mean $0$ and standard deviation $10$ for fixed effects light intensity and treatment.
+We used a weakly informative normal prior with mean $0$ and standard deviation $10$ for fixed effects of light intensity and treatment.
 We used a weakly informative normal prior with mean $-3$ and standard deviation $5$ for the the random effect standard deviation of replicate within accession.
@@
 Since these hypotheses are not mutually exclusive, we describe how we assessed support for one, two, or all three hypotheses simultaneously in \autoref{tbl-predictions}.
+In general, the assimilatory hypothesis was supported if \aax{} was greater at high light intensity than low light intensity.
+The plastic hypothesis was supported if \aax{} was greater in sun leaves than shade leaves.
+The constitutive hypothesis was supported if accession-level \aax{} increased with native \ppfd.
+In interactive models, we only consider positively reinforcing interactions between high light intensity, sun leaves, and native \ppfd{} because these are the only interactions which could explain why amphistomatous leaves are advantageous in high light habitats.
 
@@
 If the 95% confidence intervals for a parameter did not overlap zero, we considered the parameter to be significantly different from zero.
-We used leave-one-out cross-validation (LOOIC) to compare the fit of models (\autoref{tbl-models}) using the *R* package **loo** version `r packageVersion("loo")` [@vehtari_practical_2017] to calculate LOOIC values.
+We used the leave-one-out cross-validation information criterion (LOOIC) to compare the fit of models (\autoref{tbl-models}) using the *R* package **loo** version `r packageVersion("loo")` [@vehtari_practical_2017] to calculate LOOIC values.
+Note that LOOIC was calculated only from pointwise likelihood values of the submodel estimating effects of light intensity, light treatments, and accession on \aax{} (see @sec-parameters).
 We considered models with two standard errors of the mean lower LOOIC value to be a better fit to the data; we considered models with LOOIC values within two standard errors of the mean to be have similar support.
 
-```{r, predictions, eval=FALSE, echo=FALSE,message=FALSE,warning=FALSE}
+```{r, predictions, eval=TRUE, echo=FALSE,message=FALSE,warning=FALSE}
 #| label: tbl-predictions
-#| tbl-cap: "caption"
+#| tbl-cap: "Predictions of competing hypotheses about the relationship between \\aax{} and light intensity, light treatment, and native \\ppfd{} among accessions.
+The middle column lists specific directional predictions about parameter values and model fit according to leave-one-out cross-validation information criterion (LOOIC), where $\\text{LOOIC}_i$ is the LOOIC value for model $i$.
+The rightmost column describes the predictions in words and explains how accession-level \\aax{} is calculated in the relevant model."
 
-* Assimilatory: 
-  + average AA2000 - AA150
-  + among accession variation in AA2000 - AA150
-* Plastic
-  + average AAsun - AAshade
-  + among accession variation in AAsun - AAshade
-* Constitutive
-  + cor(PAR, AA) at accession level
+tribble(
+  ~"\\textbf{Hypothesis}", ~"\\textbf{Prediction(s)}", ~"\\textbf{Description}",
+  "Assimilatory", "$\\beta_{\\mathrm{AA},2000} > 0$", "Average \\aax{} at high light intensity is greater than that at low light intensity",
+  "Plastic", "$\\beta_{\\mathrm{AA},\\text{sun}} > 0$", "Average \\aax{} in sun leaves is greater than that in shade leaves",
+  "Constitutive", "$\\beta_{\\mathrm{PPFD,AA}} > 0$", "Accession-level \\aax{} ($\\mathrm{AA}_\\text{acc}$) increases with native \\ppfd",
+  "Constitutive", "", "$\\mathrm{AA}_\\text{acc} = \\beta_{\\mathrm{AA}, 0} + \\vec{\\beta}_{\\mathrm{AA}, \\text{acc}}$",
+  "Assimilatory $\\times$ Plastic", "$\\beta_{\\mathrm{AA},2000,\\text{sun}} > 0$", "Average \\aax{} is highest at high light intensity in sun leaves",
+  "Assimilatory $\\times$ Plastic", "$\\text{LOOIC}_\\text{1} > \\text{LOOIC}_\\text{2}$", "",
+  "Assimilatory $\\times$ Constitutive", "$\\beta_{\\mathrm{AA},2000} > 0$", "Average \\aax{} at high light intensity is greater than that at low light intensity",
+  "Assimilatory $\\times$ Constitutive", "$\\beta_{\\mathrm{PPFD,AA}} > 0$", "Accession-level \\aax{} at high light intensity ($\\mathrm{AA}_{\\text{acc},2000}$) increases with native \\ppfd",
+  "Assimilatory $\\times$ Constitutive", "$\\text{LOOIC}_\\text{1} > \\text{LOOIC}_\\text{3}$", "$\\mathrm{AA}_{\\text{acc},2000} = \\beta_{\\mathrm{AA}, 0} + \\vec{\\beta}_{\\mathrm{AA}, \\text{acc}} + \\vec{\\beta}_{\\mathrm{AA}, 2000, \\text{acc}}$",
+  "Plastic $\\times$ Constitutive", "$\\beta_{\\mathrm{AA},\\text{sun}} > 0$", "Average \\aax{} in sun leaves is greater than that in shade leaves",
+  "Plastic $\\times$ Constitutive", "$\\beta_{\\mathrm{PPFD,AA}} > 0$", "Accession-level \\aax{} in sun leaves ($\\mathrm{AA}_{\\text{acc},\\text{sun}}$) increases with native \\ppfd",
+  "Plastic $\\times$ Constitutive", "$\\text{LOOIC}_\\text{1} > \\text{LOOIC}_\\text{4}$", "$\\mathrm{AA}_{\\text{acc},\\text{sun}} = \\beta_{\\mathrm{AA}, 0} + \\vec{\\beta}_{\\mathrm{AA}, \\text{acc}} + \\vec{\\beta}_{\\mathrm{AA}, \\text{sun}, \\text{acc}}$",
+  "Assimilatory $\\times$ Plastic $\\times$ Constitutive", "$\\beta_{\\mathrm{AA},2000,\\text{sun}} > 0$", "Average \\aax{} is highest at high light intensity in sun leaves",
+  "Assimilatory $\\times$ Plastic $\\times$ Constitutive", "$\\beta_{\\mathrm{PPFD,AA}} > 0$", "Accession-level \\aax{} at high light intensity in sun leaves ($\\mathrm{AA}_{\\text{acc},2000,\\text{sun}}$) increases with native \\ppfd",
+  "Assimilatory $\\times$ Plastic $\\times$ Constitutive", "$\\text{LOOIC}_\\text{3} > \\text{LOOIC}_\\text{5}$", "$\\mathrm{AA}_{\\text{acc},2000,\\text{sun}} = \\beta_{\\mathrm{AA}, 0} + \\vec{\\beta}_{\\mathrm{AA}, \\text{acc}} + \\vec{\\beta}_{\\mathrm{AA}, 2000, \\text{acc}} + \\vec{\\beta}_{\\mathrm{AA}, \\text{sun}, \\text{acc}}$"
 
-Reinforcing interactions (we only consider these because these are ones which could explain why amphi leaves occur in high light habitats)
+) |>
+  knitr::kable(
+    format = "latex",
+    align = c("l", "l", "l"),
+    booktabs = TRUE,
+    escape = FALSE,
+    longtable = TRUE
+  ) |>
+  kableExtra::add_indent(positions = seq_len(15), level_of_indent = -1, target_cols = 3) |>
+  kableExtra::collapse_rows(1, latex_hline = "major", valign = "top",
+                            row_group_label_position = "first") |>
+  kableExtra::column_spec(1, width = "1in") |>
+  kableExtra::column_spec(2, width = "1.5in") |>
+  kableExtra::column_spec(3, width = "3in")
 
-* Assimilatory * Plastic (reinforcing interaction)
-  + AAsun - AAshade greater at 2000 than 150
-  + AA2000 - AA150 greater in sun leaves than shade leaves
-  + there may be var among accessions, but not correlated with PAR
-
-* Assimilatory * Constitutive (reinforcing interaction)
-  + cor(AA2000, PAR) > 0; cor(AA150, PAR) = 0
-  + AA2000 - AA150 varies among accessions, correlated with PAR
-
-* Plastic * Constitutive (reinforcing interaction)
-  + cor(AAsun, PAR) > 0; cor(AAshade, PAR) = 0
-  + AAsun - AAshade varies among accessions, correlated with PAR
-
-* Assimilatory * Plastic * Constitutive (reinforcing interaction)
-  + AAsun - AAshade greater at 2000 than 150
-  + AA2000 - AA150 greater in sun leaves than shade leaves
-  + AAsun - AAshade varies among accession, correlated with PAR
-  + AA2000 - AA150 varies among accession, correlated with PAR
-  + cor(AAsun2000, PAR) > cor(AAsun150, PAR) and > cor(AAshade2000, PAR) and > cor(AAshade150, PAR) in nonadditive way
-
-| Hypothesis   | Prediction | Model |
-|--------------|------------|------------------|
-| Assimilatory |  | base model |
-| Plastic      |  | base model |
-| Constitutive |  | base model |
-| Assimilatory * Plastic |  | + intensity $\times$ treatment |
-| Assimilatory * Constitutive |  | + accession $\times$ intensity (random) |
-| Plastic * Constitutive |  | + accession $\times$ treatment (random) |
-| Assimilatory * Plastic * Constitutive |  | + accession $\times$ intensity $\times$ treatment (random) |
+# * Assimilatory: 
+#   + average AA2000 - AA150
+#   + among accession variation in AA2000 - AA150
+# * Plastic
+#   + average AAsun - AAshade
+#   + among accession variation in AAsun - AAshade
+# * Constitutive
+#   + cor(PAR, AA) at accession level
+#   
+# Reinforcing interactions 
+# 
+# * Assimilatory * Plastic (reinforcing interaction)
+#   + AAsun - AAshade greater at 2000 than 150
+#   + AA2000 - AA150 greater in sun leaves than shade leaves
+#   + there may be var among accessions, but not correlated with PAR
+#   
+# * Assimilatory * Constitutive (reinforcing interaction)
+#   + cor(AA2000, PAR) > 0; cor(AA150, PAR) = 0
+#   + AA2000 - AA150 varies among accessions, correlated with PAR
+#   
+# * Plastic * Constitutive (reinforcing interaction)
+#   + cor(AAsun, PAR) > 0; cor(AAshade, PAR) = 0
+#   + AAsun - AAshade varies among accessions, correlated with PAR
+#   
+# * Assimilatory * Plastic * Constitutive (reinforcing interaction)
+#   + AAsun - AAshade greater at 2000 than 150
+#   + AA2000 - AA150 greater in sun leaves than shade leaves
+#   + AAsun - AAshade varies among accession, correlated with PAR
+#   + AA2000 - AA150 varies among accession, correlated with PAR
+#   + cor(AAsun2000, PAR) > cor(AAsun150, PAR) and > cor(AAshade2000, PAR) and > cor(AAshade150, PAR) in nonadditive way
+#   
+# 
+# | Hypothesis   | Prediction | Model |
+# |--------------|------------|------------------|
+# | Assimilatory |  | base model |
+# | Plastic      |  | base model |
+# | Constitutive |  | base model |
+# | Assimilatory * Plastic |  | + intensity $\times$ treatment |
+# | Assimilatory * Constitutive |  | + accession $\times$ intensity (random) |
+# | Plastic * Constitutive |  | + accession $\times$ treatment (random) |
+# | Assimilatory * Plastic * Constitutive |  | + accession $\times$ intensity $\times$ treatment (random) |
 

### 2024-04-10 ms/ms.tex
drafted predictions section

@@
 
-\subsubsection{Parameter estimation and
-priors}\label{parameter-estimation-and-priors}
+\subsubsection{Parameter estimation and priors}\label{sec-parameters}
 
@@
 \addlinespace[0.3em]
-\multicolumn{2}{l}{\textbf{effects light intensity, light treatments, and accession on \aax}}\\
+\multicolumn{2}{l}{\textbf{effects of light intensity, light treatments, and accession on \aax}}\\
 \hspace{1em}$\beta_{\mathrm{AA},0}$ & intercept of \aax{} at low light intensity in shade treatment\\
@@
 \hspace{1em}$\beta_{\mathrm{AA},\text{sun}}$ & effect of sun treatment on \aax{}\\
-\hspace{1em}\cellcolor{gray!10}{$\beta_{\mathrm{AA},\text{2000,high}}$} & \cellcolor{gray!10}{effect of high light intensity at \ppfdequals{2000} on \aax{} in sun treatment}\\
+\hspace{1em}\cellcolor{gray!10}{$\beta_{\mathrm{AA},2000,\text{sun}}$} & \cellcolor{gray!10}{effect of high light intensity at \ppfdequals{2000} on \aax{} in sun treatment}\\
 \hspace{1em}$\vec{\beta}_{\mathrm{AA},\text{acc}}$ & vector of $n_\text{acc}$ phylogenetically structured random accession-level effects on \aax; $\vec{\beta}_{\mathrm{AA},\text{acc}} \sim \text{MVN}(\vec{0}, \symbf{\Sigma}_{\text{AA,acc}})$\\
@@
 treatment. We used a weakly informative normal prior with mean \(0\) and
-standard deviation \(10\) for fixed effects light intensity and
+standard deviation \(10\) for fixed effects of light intensity and
 treatment. We used a weakly informative normal prior with mean \(-3\)
@@
 
-WORKING HERE TO ADD PREDICTIONS AND TABLE
-
 The assimilatory, plastic, and constitutive hypotheses make different
@@
 hypotheses are not mutually exclusive, we describe how we assessed
-support for one, two, or all three hypotheses simultaneously.
-
-We tested competing predictions through a combination of model selection
-and parameter estimation from the posterior of fitted models. In all
-cases, we estimated parameters and confidence intervals as described in
-Section~\ref{sec-fitting}.
-
-We started with a base model that includes fixed effects of light
-intensity and light treatment on \aax, a random effect of accession on
-\aax, and a linear regression of PAR on accession-level \aax. We next
-compared the fit of the base model to a more complex with an interaction
-between light intensity and light treatment using LOOIC with the
-\textit{R} package \textbf{loo} version 2.7.0 (\emph{15}). These
-comparisons allowed us to test the main predictions of each hypothesis
-and for a reinforcing interactions between assimilatory and plastic
-hypotheses (tab:predictions). However, these models cannot test
-predictions when there is an interaction with the constitutive
-hypothesis, because this hypothesis posits accession-levels variation in
-responses to light intensity and/or light treatment. Therefore, we
-tested for significant variation among accessions in light intensity and
-light treatment by comparing LOOIC values of models with and without
-random interactions between these factors and accession.
-
-\begin{itemize}
-\tightlist
-\item
-  Assimilatory:
-
-  \begin{itemize}
-  \tightlist
-  \item
-    average AA2000 - AA150
-  \item
-    among accession variation in AA2000 - AA150
-  \end{itemize}
-\item
-  Plastic
-
-  \begin{itemize}
-  \tightlist
-  \item
-    average AAsun - AAshade
-  \item
-    among accession variation in AAsun - AAshade
-  \end{itemize}
-\item
-  Constitutive
-
-  \begin{itemize}
-  \tightlist
-  \item
-    cor(PAR, AA) at accession level
-  \end{itemize}
-\end{itemize}
-
-Reinforcing interactions (we only consider these because these are ones
-which could explain why amphi leaves occur in high light habitats)
-
-\begin{itemize}
-\tightlist
-\item
-  Assimilatory * Plastic (reinforcing interaction)
-
-  \begin{itemize}
-  \tightlist
-  \item
-    AAsun - AAshade greater at 2000 than 150
-  \item
-    AA2000 - AA150 greater in sun leaves than shade leaves
-  \item
-    there may be var among accessions, but not correlated with PAR
-  \end{itemize}
-\item
-  Assimilatory * Constitutive (reinforcing interaction)
-
-  \begin{itemize}
-  \tightlist
-  \item
-    cor(AA2000, PAR) \textgreater{} 0; cor(AA150, PAR) = 0
-  \item
-    AA2000 - AA150 varies among accessions, correlated with PAR
-  \end{itemize}
-\item
-  Plastic * Constitutive (reinforcing interaction)
-
-  \begin{itemize}
-  \tightlist
-  \item
-    cor(AAsun, PAR) \textgreater{} 0; cor(AAshade, PAR) = 0
-  \item
-    AAsun - AAshade varies among accessions, correlated with PAR
-  \end{itemize}
-\item
-  Assimilatory * Plastic * Constitutive (reinforcing interaction)
-
-  \begin{itemize}
-  \tightlist
-  \item
-    AAsun - AAshade greater at 2000 than 150
-  \item
-    AA2000 - AA150 greater in sun leaves than shade leaves
-  \item
-    AAsun - AAshade varies among accession, correlated with PAR
-  \item
-    AA2000 - AA150 varies among accession, correlated with PAR
-  \item
-    cor(AAsun2000, PAR) \textgreater{} cor(AAsun150, PAR) and
-    \textgreater{} cor(AAshade2000, PAR) and \textgreater{}
-    cor(AAshade150, PAR) in nonadditive way
-  \end{itemize}
-\end{itemize}
-
-\begin{longtable}[]{@{}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.3182}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.2727}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.4091}}@{}}
-\toprule\noalign{}
-\begin{minipage}[b]{\linewidth}\raggedright
-Hypothesis
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-Prediction
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-Model
-\end{minipage} \\
-\midrule\noalign{}
-\endhead
-\bottomrule\noalign{}
-\endlastfoot
-Assimilatory & & base model \\
-Plastic & & base model \\
-Constitutive & & base model \\
-Assimilatory * Plastic & & + intensity \(\times\) treatment \\
-Assimilatory * Constitutive & & + accession \(\times\) intensity
-(random) \\
-Plastic * Constitutive & & + accession \(\times\) treatment (random) \\
-Assimilatory * Plastic * Constitutive & & + accession \(\times\)
-intensity \(\times\) treatment (random) \\
+support for one, two, or all three hypotheses simultaneously in
+\autoref{tbl-predictions}. In general, the assimilatory hypothesis was
+supported if \aax{} was greater at high light intensity than low light
+intensity. The plastic hypothesis was supported if \aax{} was greater in
+sun leaves than shade leaves. The constitutive hypothesis was supported
+if accession-level \aax{} increased with native \ppfd. In interactive
+models, we only consider positively reinforcing interactions between
+high light intensity, sun leaves, and native \ppfd{} because these are
+the only interactions which could explain why amphistomatous leaves are
+advantageous in high light habitats.
+
+We evaluated predictions using a combination of parameter estimation and
+model selection. In all cases, we estimated parameters and confidence
+intervals as described in Section~\ref{sec-fitting}. If the 95\%
+confidence intervals for a parameter did not overlap zero, we considered
+the parameter to be significantly different from zero. We used the
+leave-one-out cross-validation information criterion (LOOIC) to compare
+the fit of models (\autoref{tbl-models}) using the \emph{R} package
+\textbf{loo} version 2.7.0 (\emph{15}) to calculate LOOIC values. Note
+that LOOIC was calculated only from pointwise likelihood values of the
+submodel estimating effects of light intensity, light treatments, and
+accession on \aax{} (see Section~\ref{sec-parameters}). We considered
+models with two standard errors of the mean lower LOOIC value to be a
+better fit to the data; we considered models with LOOIC values within
+two standard errors of the mean to be have similar support.
+
+\begin{longtable}{>{\raggedright\arraybackslash}p{1in}>{\raggedright\arraybackslash}p{1.5in}>{\raggedright\arraybackslash}p{3in}}
+
+\caption{\label{tbl-predictions}Predictions of competing hypotheses
+about the relationship between \aax{} and light intensity, light
+treatment, and native \ppfd{} among accessions. The middle column lists
+specific directional predictions about parameter values and model fit
+according to leave-one-out cross-validation information criterion
+(LOOIC), where \(\text{LOOIC}_i\) is the LOOIC value for model \(i\).
+The rightmost column describes the predictions in words and explains how
+accession-level \aax{} is calculated in the relevant model.}
+
+\tabularnewline
+
+\toprule
+\textbf{Hypothesis} & \textbf{Prediction(s)} & \textbf{Description}\\
+\midrule
+Assimilatory & $\beta_{\mathrm{AA},2000} > 0$ & \hspace{-1em}Average \aax{} at high light intensity is greater than that at low light intensity\\
+\cmidrule{1-3}\pagebreak[0]
+Plastic & $\beta_{\mathrm{AA},\text{sun}} > 0$ & \hspace{-1em}Average \aax{} in sun leaves is greater than that in shade leaves\\
+\cmidrule{1-3}\pagebreak[0]
+Constitutive & $\beta_{\mathrm{PPFD,AA}} > 0$ & \hspace{-1em}Accession-level \aax{} ($\mathrm{AA}_\text{acc}$) increases with native \ppfd\\
+\nopagebreak
+ &  & \hspace{-1em}$\mathrm{AA}_\text{acc} = \beta_{\mathrm{AA}, 0} + \vec{\beta}_{\mathrm{AA}, \text{acc}}$\\
+\cmidrule{1-3}\pagebreak[0]
+Assimilatory $\times$ Plastic & $\beta_{\mathrm{AA},2000,\text{sun}} > 0$ & \hspace{-1em}Average \aax{} is highest at high light intensity in sun leaves\\
+\nopagebreak
+ & $\text{LOOIC}_\text{1} > \text{LOOIC}_\text{2}$ & \hspace{-1em}\\
+\cmidrule{1-3}\pagebreak[0]
+Assimilatory $\times$ Constitutive & $\beta_{\mathrm{AA},2000} > 0$ & \hspace{-1em}Average \aax{} at high light intensity is greater than that at low light intensity\\
+\nopagebreak
+ & $\beta_{\mathrm{PPFD,AA}} > 0$ & \hspace{-1em}Accession-level \aax{} at high light intensity ($\mathrm{AA}_{\text{acc},2000}$) increases with native \ppfd\\
+\nopagebreak
+ & $\text{LOOIC}_\text{1} > \text{LOOIC}_\text{3}$ & \hspace{-1em}$\mathrm{AA}_{\text{acc},2000} = \beta_{\mathrm{AA}, 0} + \vec{\beta}_{\mathrm{AA}, \text{acc}} + \vec{\beta}_{\mathrm{AA}, 2000, \text{acc}}$\\
+\cmidrule{1-3}\pagebreak[0]
+Plastic $\times$ Constitutive & $\beta_{\mathrm{AA},\text{sun}} > 0$ & \hspace{-1em}Average \aax{} in sun leaves is greater than that in shade leaves\\
+\nopagebreak
+ & $\beta_{\mathrm{PPFD,AA}} > 0$ & \hspace{-1em}Accession-level \aax{} in sun leaves ($\mathrm{AA}_{\text{acc},\text{sun}}$) increases with native \ppfd\\
+\nopagebreak
+ & $\text{LOOIC}_\text{1} > \text{LOOIC}_\text{4}$ & \hspace{-1em}$\mathrm{AA}_{\text{acc},\text{sun}} = \beta_{\mathrm{AA}, 0} + \vec{\beta}_{\mathrm{AA}, \text{acc}} + \vec{\beta}_{\mathrm{AA}, \text{sun}, \text{acc}}$\\
+\cmidrule{1-3}\pagebreak[0]
+Assimilatory $\times$ Plastic $\times$ Constitutive & $\beta_{\mathrm{AA},2000,\text{sun}} > 0$ & \hspace{-1em}Average \aax{} is highest at high light intensity in sun leaves\\
+\nopagebreak
+ & $\beta_{\mathrm{PPFD,AA}} > 0$ & \hspace{-1em}Accession-level \aax{} at high light intensity in sun leaves ($\mathrm{AA}_{\text{acc},2000,\text{sun}}$) increases with native \ppfd\\
+\nopagebreak
+ & $\text{LOOIC}_\text{3} > \text{LOOIC}_\text{5}$ & \hspace{-1em}$\mathrm{AA}_{\text{acc},2000,\text{sun}} = \beta_{\mathrm{AA}, 0} + \vec{\beta}_{\mathrm{AA}, \text{acc}} + \vec{\beta}_{\mathrm{AA}, 2000, \text{acc}} + \vec{\beta}_{\mathrm{AA}, \text{sun}, \text{acc}}$\\
+\bottomrule
+
 \end{longtable}

### 2024-04-11 ms/ms.qmd
working on main text outline

@@
 
-In preamble: - short codes for commonly used symbols - ?
-
 Terminology to standardize:
@@
 definite: Wei Shen Lim, Dachuan Wang, Chris Muir
+
 question: Sam McKlin, Tom Buckley
 
-## Main
+I am using this approach to drafting the paper:
 
-\[introduction\] Stomata are important
+https://blogs.nature.com/nyc/2011/08/10/how-to-write-a-paper-one-possible-answer
 
-Stomatal ratio is important unsolved problem
+## Punchline
+
+Plastic changes in the structure of sun leaves can increase the intercellular CO$_2$ diffusion path length.
+Gas exchange through stomata on both upper and lower leaf surfaces, amphistomy, may be a compensatory mechanism to reduce the impact of increased diffusion path length on photosynthesis without increasing water loss.
+We demonstrate for the first time that sun leaves benefit most from amphistomy because it compensates for plastic structural changes in leaf anatomy.
+
+## Figures
+
+1. competing hypotheses and predictions (or predictions after methods?) for the benefit of amphistomy
+2. design of experiment to test competing hypotheses (with actual #'s of replicate, curves, etc.)
+3. Results on acclimatory, plastic, constitutive hypotheses
+4. Additional supporting figure on change in LMA, reduction of Amass, and relationship between LMA and AA? Modeling?
+
+## Outline
+
+1. Stomata are important
+  * organismal level, they help plants respond in variable environments (mention as key innovation of land plants, getting tall on land)
+  * ecosystem level, collectively stomata contribute to global carbon and water cycles
+  * because of their importance, stomata are useful for understanding adaptation, improving crops, predicting paleoclimate, and modeling future climate
+
+2. Despite extensive theoretical and empirical progress understanding stomata, stomatal ratio is important unsolved problem
+  * SR varies both qualitatively and quantitatively among leaves
+  * Presumably there is some relationship SR and conductance ratio
+  * More even conductance ratio increase CO2 supply without increasing water loss
+  * But we don't understand why SR varies so much even among dorsiventral leaves
+
+3. Here we focus on the idea amphistomy may be more common in sunny habitats because sun leaves benefit more.
+  * three nonmutually exclusive hypotheses: acclimatory, plastic, constitutive
+  * WORKING HERE
+
+## Intro
 
@@
 
-# * Assimilatory: 
-#   + average AA2000 - AA150
-#   + among accession variation in AA2000 - AA150
-# * Plastic
-#   + average AAsun - AAshade
-#   + among accession variation in AAsun - AAshade
-# * Constitutive
-#   + cor(PAR, AA) at accession level
-#   
-# Reinforcing interactions 
-# 
-# * Assimilatory * Plastic (reinforcing interaction)
-#   + AAsun - AAshade greater at 2000 than 150
-#   + AA2000 - AA150 greater in sun leaves than shade leaves
-#   + there may be var among accessions, but not correlated with PAR
-#   
-# * Assimilatory * Constitutive (reinforcing interaction)
-#   + cor(AA2000, PAR) > 0; cor(AA150, PAR) = 0
-#   + AA2000 - AA150 varies among accessions, correlated with PAR
-#   
-# * Plastic * Constitutive (reinforcing interaction)
-#   + cor(AAsun, PAR) > 0; cor(AAshade, PAR) = 0
-#   + AAsun - AAshade varies among accessions, correlated with PAR
-#   
-# * Assimilatory * Plastic * Constitutive (reinforcing interaction)
-#   + AAsun - AAshade greater at 2000 than 150
-#   + AA2000 - AA150 greater in sun leaves than shade leaves
-#   + AAsun - AAshade varies among accession, correlated with PAR
-#   + AA2000 - AA150 varies among accession, correlated with PAR
-#   + cor(AAsun2000, PAR) > cor(AAsun150, PAR) and > cor(AAshade2000, PAR) and > cor(AAshade150, PAR) in nonadditive way
-#   
-# 
-# | Hypothesis   | Prediction | Model |
-# |--------------|------------|------------------|
-# | Assimilatory |  | base model |
-# | Plastic      |  | base model |
-# | Constitutive |  | base model |
-# | Assimilatory * Plastic |  | + intensity $\times$ treatment |
-# | Assimilatory * Constitutive |  | + accession $\times$ intensity (random) |
-# | Plastic * Constitutive |  | + accession $\times$ treatment (random) |
-# | Assimilatory * Plastic * Constitutive |  | + accession $\times$ intensity $\times$ treatment (random) |
-
 ```
@@
 
-Justin Alter, Max Gatlin, Joana Kim, Jenna Matsuyama, Brandon Najarian, Kai Yasuda
+Justin Alter, Max Gatlin, Joana Kim, Jenna Matsuyama, Brandon Najarian, Kai Yasuda. Sarah Friedrich helped with figures.
 

### 2024-04-11 ms/ms.qmd
drafted intro more

@@
   * three nonmutually exclusive hypotheses: acclimatory, plastic, constitutive
-  * WORKING HERE
+  * I've already drafted brief explanations of these hypotheses below
 
-## Intro
+4.
+(a lot of this is AI generated text that I need to refine) We tested these hypotheses by comparing AA among wild tomato species from different native light habitats, grown under simulated sun and shade light treatments, and measured under contrasting light intensity
+  * We measured AA on `r n_plant` individual plants from `r n_acc` accessions (average of `r mean_n_rep` replicates per light treatment) using a recently developed method [@triplett_amphistomy_2024]
+  * We directly compare the photosynthetic rate of an untreated amphistomatous leaf to that of the same leaf with gas exchange blocked through the adaxial (upper) surface by transparent plastic, which we refer to as 'pseudohypostomy'
+  * We measured AA over a range of $g_\text{sc}$, inducing stomatal opening and closure by modulating humidity
+  * We estimated 'amphistomy advantage' (\aax) *sensu* @parkhurst_adaptive_1978, but with modifications previously described in @triplett_amphistomy_2024
+  * We collected four \agcurve{} curves per leaf, an amphi (untreated) curve and a pseudohypo (treated) curve at high light-intensity (\ppfdequals{2000}; `r p_red_high`:`r p_blue_high` red:blue) and low light-intensity (\ppfdequals{150}; `r p_red_low`:`r p_blue_low` red:blue)
+  * We acclimated the focal leaf to high light (\ppfdequals{2000}) and high relative humidity (\rhequals{70}) until $A$ and \gsw{} reach their maximum.
+After that, we decreased \rh{} to $\approx 10\%$ to induce rapid stomatal closure without biochemical down
+
+5. Sun leaves have consistently higher SD and LMA and AQ parameters, but only accessions from low light habitats increased gmax ratio
+  * Large plastic responses to light intensity
+  * gmax ratio is positively associated with native light habitat of shade leaves (what is the take-away here?)
+
+6. Data support the plasticity hypothesis
+  * AA was actually greater at low light intensity
+  * No association between native PPFD and AA
+  * Consistent increase in AA in sun leaves across all accessions
+
+7. What it all means
+  * Our study reveals an important role for plasticity in explaining why amphi leaves are more common in sunny habitats
+  * Confirms that leaf structure of sun leaves incurs costs that amphistomy can mitigate
+  * It will be useful to consider coordination between leaf anatomy, stomatal density, and ratio in future studies of leaf structure-function and in crops (or something)
+
+## Intro - WORKING HERE
+
+<!-- Paragraph 1 -->
+Stomata are microscopic pores on the surfaces of leaves and other photosynthetic organs formed by a pair of guard cells.
+They are essential for growing tall on land because they allow plants to access CO$_2$ for photosynthetic carbon assimilation while preventing hydraulic failure in variable environments (citations).
+Optimal stomatal function depends on both dynamic changes in aperture on the scale of minutes to hours, as well as static anatomy determined by developmental plasticity and constitutive genetic differences [@boer__2016; harrison__2020; more citations].
+Understanding how stomata respond to environmental change over daily, seasonal, and evolutionary time is important for understanding adaptation (cites), predicting paleoclimate from fossil cuticles (cites), and improving crops (cites).
+Stomatal function contributes to global carbon and water cycles and therefore predicting future climate (cites).
+
+<!-- Paragraph 2 -->
 

### 2024-04-12 ms/ms.qmd
finished draft of intro

@@
 
-2. Despite extensive theoretical and empirical progress understanding stomata, stomatal ratio is important unsolved problem
+2. Describe amphistomy
+
+3. Despite extensive theoretical and empirical progress understanding stomata, stomatal ratio is important unsolved problem
   * SR varies both qualitatively and quantitatively among leaves
@@
 
-3. Here we focus on the idea amphistomy may be more common in sunny habitats because sun leaves benefit more.
+4. Here we focus on the idea amphistomy may be more common in sunny habitats because sun leaves benefit more.
   * three nonmutually exclusive hypotheses: acclimatory, plastic, constitutive
@@
 
-4.
+5.
 (a lot of this is AI generated text that I need to refine) We tested these hypotheses by comparing AA among wild tomato species from different native light habitats, grown under simulated sun and shade light treatments, and measured under contrasting light intensity
@@
 
-5. Sun leaves have consistently higher SD and LMA and AQ parameters, but only accessions from low light habitats increased gmax ratio
+6. Sun leaves have consistently higher SD and LMA and AQ parameters, but only accessions from low light habitats increased gmax ratio
   * Large plastic responses to light intensity
@@
 
-6. Data support the plasticity hypothesis
+8. Data support the plasticity hypothesis
   * AA was actually greater at low light intensity
@@
 
-7. What it all means
+8. What it all means
   * Our study reveals an important role for plasticity in explaining why amphi leaves are more common in sunny habitats
@@
 
-## Intro - WORKING HERE
+## Intro
 
@@
 <!-- Paragraph 2 -->
+Despite extensive theoretical and empirical progress understanding stomata function and anatomy from molecular to ecosystem levels, the adaptive significance of amphistomatous leaves remains an important unsolved problem in leaf structure-function relationships (parkhurst, grubb, gibson, drake).
+Amphistomatous leaves develop abaxial and adaxial stomata that can be independently regulated (reich, mott) to allow gas exchange through each surface (wall, marquez).
+In theory, gas exchange through stomata on each surface increases CO$_2$ supply to chloroplasts, all else being equal, enhancing photosynthesis (parkhurst, gutschick, fig.
+1).
+Amphistomatous leaves also lose more water through evaporation because of a second boundary layer conductance (Smith), but the additional carbon gain should be enough to offset this cost in most realistic scenarios (Muir 2019).
+
+<!-- Paragraph 3 -->
+The surprising fact is that despite the potential photosynthetic benefit, most leaves are not amphistomatous.
+Many vertically oriented and/or isobilateral leaves are amphistomatous (arcto paper, drake); among dorsiventral leaves, herbaceous plants in open, high light habitats tend to have have amphistomatous leaves (salisbury, bucher, jordan, etc.).
+Most other leaves, except those from aquatic habitats, are hypostomatous, producing stomata only on the lower, abaxial surface.
+The covariation between stomatal density ratio and light habitat is both qualitative and quantitative.
+A higher proportion of sun leaves are amphistomatous (salisbury) and the proportion of stomata on the upper, adaxial surface increases with light (muir 2018, bucher?).
+[end more strongly with statement that this is an important problem to be solved??]
+
+## Outtro - WORKING HERE
 

### 2024-04-17 ms/ms.qmd
updated plant-info with LMA for preliminary analyses and worked on outtro

@@
   \newcommand{\cabetween}[2]{#1 < $C_\mathrm{a} < \qty{#2}{\micro\mol\raiseto{-1}\mol}$}
+  \newcommand{\gmaxratio}{$g_\text{max,ratio}$}
   \newcommand{\gsw}{$g_\text{sw}$}
@@
 <!-- Paragraph 3 -->
+<!-- should i add analysis of herbarium data here? -->
 The surprising fact is that despite the potential photosynthetic benefit, most leaves are not amphistomatous.
@@
 
-Amphi leaves are more common in sunny habitats and confer a great benefit. (this pattern is also true in Solanum? could we analyze herbarium data here?)
+It is commonly assumed in comparative studies that most trait variation between species is constitutive and genetically determined, but our results suggest that plastic responses to light are necessary to explain why \gmaxratio{} increases with light habitat.
+Reaction norms to light intensity among wild tomatoes indicate that accessions from low light habitats express low \gmaxratio{} when grown in shade, whereas accessions from high light habitats have constitutively higher \gmaxratio{} (Fig.
+rxn-norm).
+The larger response to light intensity is driven by greater plasticty in leaf structure in accessions from low-light habitats (Fig/table - I haven't tested this yet, so this is a placeholder.) We infer that these structural changes in sun leaves are costly because they increase the diffusion path length for CO$_2$ and reduce photosynthesis.
+As a consequence, sun leaves benefit more from amphistomy.
+This is the first demonstration that sun leaves benefit more from amphistomy because it compensates for developmental plasticity in leaf structure.
 
-\[maybe make distinction between amphi leaves and var in SR here since all tomato species are amphi\]
+Two outstanding areas for further research are 1) the specific anatomical changes, such as leaf thickness or mesophyll porosity, that increase the diffusion path length for CO$_2$ in sun leaves, and 2) the costs of amphistomy that could explain plasticity in \gmaxratio{} in accessions from low-light habitats.
+Resolving the first question will require detailed anatomical measurements on a large number of shade and sun leaves with concomitant estimates of \aax.
+The fact that accessions from low-light habitats are more plastic suggests that costs of amphistomy are greatest in these environments.
+WORKING HERE
+
+## Middle
 

### 2024-04-17 r/003_trim-outliers.R

@@ -55,6 +55,10 @@ rh_curves |>
   filter(!(acc_id %in% acc_id_outlier)) |>
   write_rds("data/trimmed_rh_curves.rds")
 
+# Writing rh_curves1 to use in preliminary analyses of AA and LMA, gmax_ratio, etc.
+write_rds(rh_curves1, "objects/rh_curves1.rds")
+
+
 # The rest of this is things I tried but probably won't use
 # ggplot(rh_curves1, aes(light_treatment, aa, color = light_intensity)) +
 #   geom_boxplot()

### 2025-05-29 ms/ms.qmd

@@
 # loaded eventually to use in paper.
+# See code in 'solanum-aq' which can probably be reused here.
 aa_args = read_rds("../objects/aa_args1.rds")
@@
 aa_stats = aa_stats[order(names(aa_stats))]
+
+plant_info = read_rds("../data/plant-info.rds")
 
@@
 
+# Environmental data
+df_germ_summary = read_rds("../data/df_germ_summary.rds") |>
+  ungroup() |>
+  summarize(value = median(value), .by = c("period", "variable", "parameter"))
+
+df_growth_summary = read_rds("data/df_growth_summary.rds") |>
+  ungroup() |>
+  filter(!is.na(value))
+
 # environmental condition in germination room
-germ_temp_day = 25
-germ_temp_night = 20
-germ_rh_day = 40
-germ_rh_night = 60
-germ_ppfd = 200
 
-ppfd_sun = 800
-ppfd_shade = 125
-temp_diff = 1 # degrees warmer that the sun treatment air was
-rh_diff = 5 # rh_sun - rh_shade
-leaf_temp_diff = 1 # difference in leaf temperature between treatments
-n_leaf_temp = 600 # number of leaves for which we have measured leaf temp under ambient conditions
+germ_temp_day = df_germ_summary |>
+  filter(period == "day", variable == "degC", parameter == "mu") |>
+  pull(value) |>
+  prettyNum(digits = 3)
 
-percent_terminal = "70.0" # percent of LICOR measurements on terminal leaflet
-percent_lateral = "30.0"
+germ_temp_night = df_germ_summary |>
+  filter(period == "night", variable == "degC", parameter == "mu") |>
+  pull(value) |>
+  prettyNum(digits = 3)
+
+germ_rh_day = df_germ_summary |>
+  filter(period == "day", variable == "RH", parameter == "mu") |>
+  pull(value) |>
+  prettyNum(digits = 3)
+
+germ_rh_night = df_germ_summary |>
+  filter(period == "night", variable == "RH", parameter == "mu") |>
+  pull(value) |>
+  prettyNum(digits = 3)
+
+germ_ppfd = df_germ_summary |>
+  filter(period == "day", variable == "PAR", parameter == "mu") |>
+  pull(value) |>
+  prettyNum(digits = 3)
+
+# environmental conditions in growth room
+
+d1 = df_growth_summary |>
+  summarize(value = median(value),
+            .by = c("period", "rack", "variable", "parameter"))
+
+ppfd_sun = d1 |>
+  filter(period == "day", rack == "lower", variable == "PAR", parameter == "mu") |>
+  pull(value) |>
+  prettyNum(digits = 3)
+
+ppfd_shade = d1 |>
+  filter(period == "day", rack == "upper", variable == "PAR", parameter == "mu") |>
+  pull(value) |>
+  prettyNum(digits = 3)
+
+# degrees warmer that the sun treatment air was
+temp_diff = df_growth_summary |>
+  filter(period == "day", rack != "hobo", variable == "degC", parameter == "mu") |>
+  pivot_wider(names_from = "rack") |>
+  mutate(diff = lower - upper) |>
+  summarize(value = median(diff, na.rm = TRUE)) |>
+  pull(value) |>
+  prettyNum(digits = 3)
+
+# difference in RH
+rh_diff = df_growth_summary |>
+  filter(period == "day", rack != "hobo", variable == "RH", parameter == "mu") |>
+  pivot_wider(names_from = "rack") |>
+  mutate(diff = upper - lower) |>
+  summarize(value = median(diff, na.rm = TRUE)) |>
+  pull(value) |>
+  prettyNum(digits = 3)
+
+df_leaf_temp = plant_info |>
+  filter(!is.na(leaf_temperature_C)) |>
+  summarize(
+    leaf_temperature_C = mean(leaf_temperature_C),
+    n = n(),
+    .by = "light_treatment"
+  )
+
+# difference in leaf temperature between treatments
+leaf_temp_diff = df_leaf_temp$leaf_temperature_C |>
+  diff() |>
+  prettyNum(digits = 3)
+
+# number of leaves for which we have measured leaf temp under ambient conditions
+n_leaf_temp = sum(df_leaf_temp$n)
+
+# data on leaflet type
+df_leaflet_type = plant_info |>
+  summarize(n = n(), .by = "leaflet") |>
+  pivot_wider(names_from = "leaflet", values_from = "n") |>
+  mutate(
+    percent_terminal = prettyNum(terminal / (terminal + lateral) * 100, digits = 3),
+    percent_lateral = prettyNum(lateral / (terminal + lateral) * 100, digits = 3)
+  )
+
+percent_terminal = df_leaflet_type$percent_terminal
+percent_lateral = df_leaflet_type$percent_lateral
 percent_change_leaflet = "10.0" # percent of time that anatomical traits were measured on different leaflet than LICOR measurements
 percent_aa_and_aq = "90.0" # percent of times where AA and AQ curves were done on same leaflet
+
+# data on chamber light conditions
 p_red_high = "97.7" # percent red LED in high light-intensity treatment
@@
 4. Additional supporting figure on change in LMA, reduction of Amass, and relationship between LMA and AA? Modeling?
+
+## Abstract
+
+(use for GRC)
+
+Developmental plasticity to light intensity modulates amphistomy advantage
+
+The presence of stomata on both leaf surfaces (amphistomy) can increase photosynthesis in C$_3$ plants by reducing the path length for CO$_2$ diffusion between substomatal cavities and chloroplasts.
+Amphistomatous leaves are most common among herbaceous plants growing in sunny habitats, including many crop species.
+The distribution of amphistomatous leaves in nature may result from an increased photosynthetic benefit of amphistomy under high light intensity, either because of acclimatory, plastic, or constitutive variation in CO$_2$ supply or demand.
+We used a recently developed method to quantify amphistomy advantage, the photosynthetic rate of an amphistomatous leaf relative to an otherwise identical hypostomatous leaf, in 29 diverse populations representing 15 species of wild tomatoes (*Solanum* sect.
+*Lycopersicon* and sect.
+*Lycopersicoides*).
+Plant grown under high light intensity benefit more from amphostomy than those grown under low light, regardless of acclimation or native light intensity.
+Curvature of light response curves indicates that high-light leaves benefit more from amphistomy because of greater resistance to CO$_2$ diffusion within the mesophyll.
+Developmental plasticity in leaf thickness and/or cell arrangement to optimize light interception likely increases under varying light intensity modulates CO$_2$ diffusion within the leaf and, hence, amphistomy advantage as a byproduct.
+Developmental plasticity may therefore play an important and previously unnoticed role in explaining the adaptive significance of amphistomy and the distribution of amphistomatous leaves in nature.
 
@@
 
-\[this will be moved to SI eventually\]
+\[this will be moved to SI eventually. note that much of it also the same in solanum-aq project, so might change to cite that paper?\]
 
@@
 
-```{r, accessions, eval=FALSE,echo=FALSE,message=FALSE,warning=FALSE}
+```{r, accessions, eval=TRUE,echo=FALSE,message=FALSE,warning=FALSE}
 #| label: tbl-accessions
-#| tbl-cap: "Solanum accessions"
+#| tbl-cap: "Accessions of *Solanum* used in this study.
+The species name, accession number, collection latitude, longitude, elevation, and daily solar radiation estimated using the SPLASH algorithm (see 'Climate data').
+TGRC: Tomato Genetics Resource Center; PPFD: Photosynthetic Photon Flux Density."
 
-# outline of tab:accessions LA number,Species name, lat, lon, elev, SPLASH light value
+# Copied from solanum-aq project
 
-# copied code from stomata-ilima to emulate
-# read_rds("../processed-data/site.rds") |>
-#   filter(site_code != "nnpl") |>
-#   mutate(Island = case_when(
-#     island == "oahu" ~ "Oʻahu",
-#     island == "hawaii" ~ "Hawaiʻi"
-#   )) |>
-#   arrange(site_type, site) |>
-#   select(Site = site, Island, Habitat = site_type, Latitude = latitude_degree,
-#          Longitude = longitude_degree, `Elevation (mas)` = elevation_m) |>
-#   knitr::kable(format = "latex", digits = c(rep(0, 3), 3, 3, 0),
-#                booktabs = TRUE) |>
-#   kableExtra::kable_styling(latex_options = c("striped", "no_vertical_lines")) |>
-#   kableExtra::column_spec(1, width = "4cm") |>
-#   kableExtra::column_spec(2, width = "1.5cm") |>
-#   kableExtra::column_spec(3, width = "1.5cm") |>
-#   kableExtra::column_spec(4, width = "1.5cm") |>
-#   kableExtra::column_spec(5, width = "1.5cm") |>
-#   kableExtra::column_spec(6, width = "1.5cm")
+accession_info = read_rds("../data/accession-info.rds") |>
+  filter(accession %in% light_curves$acc) |>
+  left_join(read_rds("../data/accession-climate.rds") |>
+              select(accession, ppfd_mol_m2), by = join_by(accession)) |>
+  arrange(species, accession) |>
+  select(
+    Species = species,
+    `TGRC accession` = accession,
+    `Latitude` = lat,
+    `Longitude` = lon,
+    `Elevation (mas)` = elev,
+    `$\\mathrm{PPFD}~(\\mathrm{mol}~\\mathrm{m}^2)$` = ppfd_mol_m2
+  )
+
+accession_info |>
+  knitr::kable(
+    format = "latex",
+    digits = c(rep(0, 2), 3, 3, 0, 1),
+    booktabs = TRUE,
+    longtable = TRUE,
+    escape = FALSE
+  ) |>
+  kableExtra::kable_styling(latex_options = c("striped", "no_vertical_lines")) |>
+  kableExtra::column_spec(1, width = "3.25cm", italic = TRUE)
 

### 2025-06-20 ms/ms.qmd

@@
 
+What's surprising: role of developmental plasticity in explaining comparative results; diffusional limitations through palisade ias is strong and needs to be considered with light gradients in understanding leaf strucuture- funciton
+
 ## Outline
@@
 Optimal stomatal function depends on both dynamic changes in aperture on the scale of minutes to hours, as well as static anatomy determined by developmental plasticity and constitutive genetic differences [@boer__2016; harrison__2020; more citations].
-Understanding how stomata respond to environmental change over daily, seasonal, and evolutionary time is important for understanding adaptation (cites), predicting paleoclimate from fossil cuticles (cites), and improving crops (cites).
+Understanding how stomata respond to environmental change over daily, developmental, and evolutionary time is important for understanding adaptation (cites), predicting paleoclimate from fossil cuticles (cites), and improving crops (cites).
 Stomatal function contributes to global carbon and water cycles and therefore predicting future climate (cites).
@@
 <!-- Paragraph 3 -->
-<!-- should i add analysis of herbarium data here? -->
-The surprising fact is that despite the potential photosynthetic benefit, most leaves are not amphistomatous.
-Many vertically oriented and/or isobilateral leaves are amphistomatous (arcto paper, drake); among dorsiventral leaves, herbaceous plants in open, high light habitats tend to have have amphistomatous leaves (salisbury, bucher, jordan, etc.).
+The paradoxical fact is that, despite the potential photosynthetic benefit, most leaves are not amphistomatous.
+Many vertically oriented and/or isobilateral leaves are amphistomatous (arcto paper, drake).
+But among dorsiventral leaves, herbaceous plants in open, high light habitats tend to have have amphistomatous leaves (salisbury, bucher, jordan, etc.).
 Most other leaves, except those from aquatic habitats, are hypostomatous, producing stomata only on the lower, abaxial surface.
+Even resupinate leaves develop stomata on the lower, albeit adaxial surface (Lesheyde?), suggesting that leaf orientation (lower vs.
+upper) rather than leaf polarity (abaxial vs.
+adaxial) is causal.
 The covariation between stomatal density ratio and light habitat is both qualitative and quantitative.
 A higher proportion of sun leaves are amphistomatous (salisbury) and the proportion of stomata on the upper, adaxial surface increases with light (muir 2018, bucher?).
-[end more strongly with statement that this is an important problem to be solved??]
+Resolving why high light intensity favors amphistomatous dorsivental leaves is an important first step toward understanding variation in stomatal density ratio and leaf structure-function relationships more generally.
 
-## Outtro - WORKING HERE
+<!-- Paragraph 4+, hypotheses and predictions -->
+The overarching hypothesis is that leaves with greater stomatal density ratio are more common in open, sunny habitats because they increase photosynthesis most in those circumstances.
+An amphistomatous leaf increases photosynthetic carbon gain compared to an otherwise identical hypostomatous leaf by increasing conductance through the leaf intercellular airspaces and boudary layers; the additional water loss through a second boundary layer is typically small (smith paper, muir 2019).
+We quantify this benefit as the amphistomy advantage (AA = log(Aamphi/Ahypo) (Parkhurst, Triplett).
+Why would AA be greater in sun than shade?
+We consider three nonmutually exclusive hypotheses that we classify as 'acclimatory', 'plastic', and 'constitutive'.
 
-It is commonly assumed in comparative studies that most trait variation between species is constitutive and genetically determined, but our results suggest that plastic responses to light are necessary to explain why \gmaxratio{} increases with light habitat.
-Reaction norms to light intensity among wild tomatoes indicate that accessions from low light habitats express low \gmaxratio{} when grown in shade, whereas accessions from high light habitats have constitutively higher \gmaxratio{} (Fig.
-rxn-norm).
-The larger response to light intensity is driven by greater plasticty in leaf structure in accessions from low-light habitats (Fig/table - I haven't tested this yet, so this is a placeholder.) We infer that these structural changes in sun leaves are costly because they increase the diffusion path length for CO$_2$ and reduce photosynthesis.
-As a consequence, sun leaves benefit more from amphistomy.
-This is the first demonstration that sun leaves benefit more from amphistomy because it compensates for developmental plasticity in leaf structure.
-
-Two outstanding areas for further research are 1) the specific anatomical changes, such as leaf thickness or mesophyll porosity, that increase the diffusion path length for CO$_2$ in sun leaves, and 2) the costs of amphistomy that could explain plasticity in \gmaxratio{} in accessions from low-light habitats.
-Resolving the first question will require detailed anatomical measurements on a large number of shade and sun leaves with concomitant estimates of \aax.
-The fact that accessions from low-light habitats are more plastic suggests that costs of amphistomy are greatest in these environments.
-WORKING HERE
-
-## Middle
-
-Leaves with greater stomatal density ratio are more in open, sunny habitats because they deliver the greatest benefit in those circumstances.
-An amphistomatous leaf increases photosynthetic carbon gain compared to an otherwise identical hypostomatous leaf by increasing conductance through the leaf intercellular airspaces and boudary layers; the additional water loss through a second boundary layer is typically small \[cites\].
-We quantify this benefit as the amphistomy advantage (AA = log(Aamphi/Ahypo).
-Why would AA be greater in sun than shade?
-There are three nonmutually exclusive hypotheses that we classify as 'acclimatory', 'plastic', and 'constitutive'.
-
-Acclimatory hypothesis: Leaves acclimated to high light intensity typically increase total leaf stomatal conductance (increased CO\$\_2\$ supply) and upregulate Rubisco activation <!--# I don't love the wording here.
-Is it just Rubisco activation, or are there are other things?
---> (increased CO\$*2\$* demand).
+Acclimatory hypothesis: Photosynthetic induction to high light intensity typically involves increases in total leaf stomatal conductance (increased CO\$\_2\$ supply), the concentration of active Rubisco, and electron transport capacity (increased CO\$*2\$* demand).
 A one-dimensional circuit model using the Farquhar-von Caemmerer-Berry biochemical model of C\$\_3\$ photosynthesis shows that both increased stomatal conductance and Rubisco activity should increase AA, all else being equal (Supporting Information).
@@
 Plastic hypothesis: Individuals of the same genotype often develop dramatically different leaves in sun and shade conditions (cite).
-Plastic responses are likely adaptations to optimize photosynthesis at different light intensities (cites), but changes in leaf anatomy and biochemistry could modulate AA as a byproduct.
-Thicker or less porous leaves, both of which are associated with high leaf mass per area (LMA), will have lower g_ias; leaves with increased total stomatal density and Rubisco concentration have greater potential CO2 supply and demand.
+Plastic responses are likely adaptations to optimize photosynthesis at different light intensities in variable environments (cites).
+Plastic changes in leaf anatomy and biochemistry could modulate AA as a byproduct.
+Thicker or less porous leaves, both of which are associated with high leaf mass per area (LMA), will have lower $g_\mathrm{ias}$; leaves with increased total stomatal density and photosynthetic capacity have greater potential CO$_2$ supply and demand.
 Under the plastic hypothesis, we predict that AAsun \> AAshade for all species and light intensities.
-AAsun and gsmax,sun should be positively associated with native light habitat.
-\[transition\] We assume that genotypes adapted to sunny, open habitats will express a phenotype best adapted to that environment when leaves develop under high light intensity; genotypes adapted to shaded closed habitats may be plastic, but suboptimal for light intensities they do not regularly experience in nature.
+Secondarily, AAsun and gsmax,sun should also be positively associated with native light habitat if genotypes adapted to sunny, open habitats if they can express a phenotype best adapted to that environment when leaves develop under high light intensity.
+Genotypes adapted to shaded, closed habitats may be plastic, but limits on the width of their reaction norms prevent them from developing traits optimal for conditions they do not regularly experience in nature.
 
-\[conceptual figure could also show differential benefit of amphi leaves in sun/shade\]
+Constitutive hypothesis: In environments that are relatively constant or where environmental change cannot be anticipated by a reliable cue, natural selection will favor constitutive expression of optimal phenotypes.
+We therefore predict genotypes from more sunny, open habitats will have consistently greater AA under also measurement and growth light intensities.
+For herbaceous plants, light intensity is largely a function of the tree canopy (cites?).
+Herbs growing in the open will regularly experience high light intensity; herbs growing under a forest canopy will often experience low light intensity.
 
-to plants growing there. in those It has been hypothesized that amphistomy increases photosynthesis more in sunny places. Three nonmutually exclusive adaptive explanations for why amphi leaves:
-
-Conceptual figure explaining each hypothesis
-
--   acclimation: greater demand, higher gs increase AA
--   developmental plasticity: light-induced changes in leaf anatomy modulate AA
--   constitutive: genetic differences in leaves adapted to different light habitats
+CONCEPTUAL FIGURE HERE
 
@@
 \[discussion\] - sun/shade has long been appreciated and this shows new trait that should be considered and that CO2 diffusion becomes major limitation
+
+## Outtro - I think this goes at the end??
+It is commonly assumed in comparative studies that most trait variation between species is constitutive and genetically determined, but our results suggest that plastic responses to light are necessary to explain why \gmaxratio{} increases with light habitat.
+Reaction norms to light intensity among wild tomatoes indicate that accessions from low light habitats express low \gmaxratio{} when grown in shade, whereas accessions from high light habitats have constitutively higher \gmaxratio{} (Fig.
+rxn-norm).
+The larger response to light intensity is driven by greater plasticty in leaf structure in accessions from low-light habitats (Fig/table - I haven't tested this yet, so this is a placeholder.) We infer that these structural changes in sun leaves are costly because they increase the diffusion path length for CO$_2$ and reduce photosynthesis.
+As a consequence, sun leaves benefit more from amphistomy.
+This is the first demonstration that sun leaves benefit more from amphistomy because it compensates for developmental plasticity in leaf structure.
+
+Two outstanding areas for further research are 1) the specific anatomical changes, such as leaf thickness or mesophyll porosity, that increase the diffusion path length for CO$_2$ in sun leaves, and 2) the costs of amphistomy that could explain plasticity in \gmaxratio{} in accessions from low-light habitats.
+Resolving the first question will require detailed anatomical measurements on a large number of shade and sun leaves with concomitant estimates of \aax.
+The fact that accessions from low-light habitats are more plastic suggests that costs of amphistomy are greatest in these environments.
+WORKING HERE
 

### 2025-06-20 ms/ms.qmd
working on ms

@@
 
-We distinguished among these hypotheses by comparing AA among wild tomato species from different native light habitats, grown under simulated sun and shade light treatments, and measured under contrasting light intensity (Figure of hypotheses and predictions).
+caption: Directional predictions associated with each hypothesis explaining why amphistomy advantage (AA) might be greater for leaves in sunny, open habitats.
+For each hypothesis, we make predictions for how native plant area index (PAI), growth light treatment, and measurement light intensity would affect AA.
+
+|  hypothesis  | measurement light intensity | growth light intensity  | native PAI |
+|:------------:|:---------------------------:|:------------------------:|:----------:|
+| acclimatory  | AA2000 > AA150             | AAsun = AAshade          | cor(PAI, AA) = 0 |
+| plastic      | AA2000 = AA150             | AAsun > AAshade          | cor(PAI, AA) = 0 |
+| constitutive | AA2000 = AA150             | AAsun = AAshade          | cor(PAI, AA) > 0 |
+
+<!-- Paragraph 5 (summarize methods) -->
+We tested these hypotheses by comparing AA among amphistomatous wild tomato species (systematic cites) from different native light habitats, grown under simulated sun and shade light treatments, and measured under contrasting light intensity (Figure of hypotheses and predictions).
 We measured AA on `r n_plant` individual plants from `r n_acc` accessions (average of `r mean_n_rep` replicates per light treatment) using a recently developed method [@triplett_amphistomy_2024].
@@
 To compare amphi- and pseudohypostomatous leaves at identical whole-leaf $g_\text{sc}$, we measure $A$ over a range of $g_\text{sc}$, inducing stomatal opening and closure by modulating humidity (see Materials and Methods for further details).
+We estimated 'amphistomy advantage' (\aax) *sensu* @parkhurst_adaptive_1978, but with modifications previously described in @triplett_amphistomy_2024 and here (Materials and Methods).
+The native light intensity was represented by plant area index (PAI $\unit{\meter\squared\per\meter\squared}$), estimated using a global gridded X-m2 resolution data set derived from the Global Ecosystem Dynamics Investigation (GEDI; cite) and georeferenced accession collection information (cite TGRC).
+The growth light intensities were `r ppfd_sun` (sun treatment) and `r ppfd_shade` (shade treatment) while all other environment conditions were nearly identical (see @sec-methods for supporting detail).
+The high and low measurements intensities were \ppfdequals{2000} (`r p_red_high`:`r p_blue_high` red:blue) and \ppfdequals{150} (`r p_red_low`:`r p_blue_low` red:blue), respectively.
 
-Table of directional predictions (with table summarizing results? part of conceptual figure? in supplement?)
-
-caption: Directional predictions associated with each hypothesis explaining why amphistomy advantage (AA) might be greater for leaves in sunny, open habitats.
-For each hypothesis, we make predictions for how native light habitat, light treatment, and light intensity would affect AA.
-
-|  hypothesis  | native light habitat | light treatment  | light intensity |
-|:------------:|:--------------------:|:----------------:|:---------------:|
-| acclimatory  | cor(PAR, AA) = 0   | AAsun = AAshade  | AA2000 \> AA150 |
-|   plastic    | cor(PAR, AA) = 0 | AAsun \> AAshade | AA2000 = AA150  |
-| constitutive | cor(PAR, AA) \> 0   | AAsun = AAshade  | AA2000 = AA150  |
+#### MARKER
+(not sure where this goes)  * We acclimated the focal leaf to high light (\ppfdequals{2000}) and high relative humidity (\rhequals{70}) until $A$ and \gsw{} reach their maximum.
+After that, we decreased \rh{} to $\approx 10\%$ to induce rapid stomatal closure without biochemical down
 
@@
 The fact that accessions from low-light habitats are more plastic suggests that costs of amphistomy are greatest in these environments.
-WORKING HERE
+(not finished yet)
 
-## Materials and Methods
+## Materials and Methods [#sec-methods]
 

### 2025-06-22 ms/ms.qmd
testing lma as mediator

@@
 
-## Outline
-
-1. Stomata are important
-  * organismal level, they help plants respond in variable environments (mention as key innovation of land plants, getting tall on land)
-  * ecosystem level, collectively stomata contribute to global carbon and water cycles
-  * because of their importance, stomata are useful for understanding adaptation, improving crops, predicting paleoclimate, and modeling future climate
-
-2. Describe amphistomy
-
-3. Despite extensive theoretical and empirical progress understanding stomata, stomatal ratio is important unsolved problem
-  * SR varies both qualitatively and quantitatively among leaves
-  * Presumably there is some relationship SR and conductance ratio
-  * More even conductance ratio increase CO2 supply without increasing water loss
-  * But we don't understand why SR varies so much even among dorsiventral leaves
-
-4. Here we focus on the idea amphistomy may be more common in sunny habitats because sun leaves benefit more.
-  * three nonmutually exclusive hypotheses: acclimatory, plastic, constitutive
-  * I've already drafted brief explanations of these hypotheses below
-
-5.
-(a lot of this is AI generated text that I need to refine) We tested these hypotheses by comparing AA among wild tomato species from different native light habitats, grown under simulated sun and shade light treatments, and measured under contrasting light intensity
-  * We measured AA on `r n_plant` individual plants from `r n_acc` accessions (average of `r mean_n_rep` replicates per light treatment) using a recently developed method [@triplett_amphistomy_2024]
-  * We directly compare the photosynthetic rate of an untreated amphistomatous leaf to that of the same leaf with gas exchange blocked through the adaxial (upper) surface by transparent plastic, which we refer to as 'pseudohypostomy'
-  * We measured AA over a range of $g_\text{sc}$, inducing stomatal opening and closure by modulating humidity
-  * We estimated 'amphistomy advantage' (\aax) *sensu* @parkhurst_adaptive_1978, but with modifications previously described in @triplett_amphistomy_2024
-  * We collected four \agcurve{} curves per leaf, an amphi (untreated) curve and a pseudohypo (treated) curve at high light-intensity (\ppfdequals{2000}; `r p_red_high`:`r p_blue_high` red:blue) and low light-intensity (\ppfdequals{150}; `r p_red_low`:`r p_blue_low` red:blue)
-  * We acclimated the focal leaf to high light (\ppfdequals{2000}) and high relative humidity (\rhequals{70}) until $A$ and \gsw{} reach their maximum.
-After that, we decreased \rh{} to $\approx 10\%$ to induce rapid stomatal closure without biochemical down
-
-6. Sun leaves have consistently higher SD and LMA and AQ parameters, but only accessions from low light habitats increased gmax ratio
-  * Large plastic responses to light intensity
-  * gmax ratio is positively associated with native light habitat of shade leaves (what is the take-away here?)
-
-8. Data support the plasticity hypothesis
-  * AA was actually greater at low light intensity
-  * No association between native PPFD and AA
-  * Consistent increase in AA in sun leaves across all accessions
-
-8. What it all means
-  * Our study reveals an important role for plasticity in explaining why amphi leaves are more common in sunny habitats
-  * Confirms that leaf structure of sun leaves incurs costs that amphistomy can mitigate
-  * It will be useful to consider coordination between leaf anatomy, stomatal density, and ratio in future studies of leaf structure-function and in crops (or something)
-
 ## Intro
@@
 Stomata are microscopic pores on the surfaces of leaves and other photosynthetic organs formed by a pair of guard cells.
-They are essential for growing tall on land because they allow plants to access CO$_2$ for photosynthetic carbon assimilation while preventing hydraulic failure in variable environments (citations).
-Optimal stomatal function depends on both dynamic changes in aperture on the scale of minutes to hours, as well as static anatomy determined by developmental plasticity and constitutive genetic differences [@boer__2016; harrison__2020; more citations].
+They are essential for balancing carbon gained per unit water lost and permitted vascular plants to grow tall on land by enabling access to CO$_2$ for photosynthetic carbon assimilation while preventing hydraulic failure in variable environments [@raven_selection_2002; @mccadam_stomata_2021; @clark_origin_2022].
+Optimal stomatal function depends on both dynamic changes in aperture on the scale of minutes to hours, as well as static anatomy determined by developmental plasticity and constitutive genetic differences [@boer_optimal_2016; @harrison_influence_2020; more citations].
 Understanding how stomata respond to environmental change over daily, developmental, and evolutionary time is important for understanding adaptation (cites), predicting paleoclimate from fossil cuticles (cites), and improving crops (cites).
@@
 
+The primary directional predictions for each hypothesis are summarized in Table X; detailed predictions for results that would indicate support for multiple hypotheses are in Table SX.
+
 CONCEPTUAL FIGURE HERE
@@
 
-#### MARKER
-(not sure where this goes)  * We acclimated the focal leaf to high light (\ppfdequals{2000}) and high relative humidity (\rhequals{70}) until $A$ and \gsw{} reach their maximum.
-After that, we decreased \rh{} to $\approx 10\%$ to induce rapid stomatal closure without biochemical down
-
-\[add rows for when multiple hypotheses are supported simultaneously? or put that in SI?\]
-
-\[results\] Amphistomy increases $A$ in all accessions, in both sun and shade leaves, and light intensities.
-We infer this from the fact that blocking gas exchange in pseudohypostomatous leaves reduced $A$ by X-X% depending on the accession, light treatment, and light intensity (Table/figure).
+<!-- Paragraph 6 (summarize main results) -->
+Consistent with biophysical theory of CO2 diffusion within leaves, AA > 0 for all accessions (Figure X).
+AA varied substantially between measurement light intensities, growth light intensities, and among accessions (point to model comparison evidence).
+Measured under high light intensity, AA was consistently greater for sun plants.
+The average AA among accessions in the shade treat was X (X of 29 accessions significantly greater than 0); however, the same accessions grown at high light intensity had an AA of X (X of 29 accession greater than 0).
+Contrary to the predictions of the assimilatory hypothesis, AA was consistently greater under low measurement light intensity.
+Among sun and shade grown plants, AA was greater under low light intensity in X of 29 accessions (significantly greater than 0 in X of 29 accessions).
+The overall average AA of shade and sun grown plants measured under low light intensity was X (+/-) and X (+/-), respectively.
+There was a slight tendency for accessions from more closed habitats (greater PAI), but AA varied widely among accessions from open habitats (low PAI) regardless of growth and measurement light intensities (Figure X).
+The pattern of AA across wild tomatoes strongly supports the plasticity hypothesis, argues against the acclimatory hypothesis, and provides only weak support for the constitutive hypothesis.
+Not sure where this goes: We infer this from the fact that blocking gas exchange in pseudohypostomatous leaves reduced $A$ by X-X% depending on the accession, light treatment, and light intensity (Table/figure).
 The AA is equivalent to an X-X% change in total $g_\text{sc}$ (see SI section gs equivalency).
@@
 
-Sun leaves from high light habitats \[not sure if htis result is true yet\] benefit the most from amphistomy because of a both developmental plasticity and constitutive differences among accessions.
-\[quantify difference in AA and contribution of different affects\].
-Surprisingly, light intensity had a little effect...
-(should this be in this paragraph, or it's own?).
-
-\[discussion\] - sun/shade has long been appreciated and this shows new trait that should be considered and that CO2 diffusion becomes major limitation
+<!-- Paragraph 7 (summarize secondary results) -->
+These results suggest that developmental plasticity, as opposed to acclimation or adaptation to open habitats, may explain the explain the long-standing observation taht amphistomatous leaves are more common in sunny habitats.
+Our results add to existing explanations by showing high light intensity *per se* does not increase the benefit of amphistomy, but rather than anatomical and biochemical change associated with higher light intensity modulate AA.
+We cannot yet attribute specific changes wrought by plasticity, but correlational evidence suggests that changes in leaf anatomy may be important.
+Species responded consistently by increasing stomatal density, especially on the adaxial surface, increasing stomatal density ratio.
+Sun leaves also had greater gs, Amax.
+However, these traits were not associated with change in AA (table/figures).
 
@@
 
+Problems to bring up in discussion:
+- all our species were amphi, so they may all live in PAI below threshold to favor amphi
+- other microclimate diffs not considered
+- don't know specific anatomical change
+
+8. What it all means
+  * Our study reveals an important role for plasticity in explaining why amphi leaves are more common in sunny habitats
+  * Confirms that leaf structure of sun leaves incurs costs that amphistomy can mitigate
+  * It will be useful to consider coordination between leaf anatomy, stomatal density, and ratio in future studies of leaf structure-function and in crops (or something)
+
 ## Materials and Methods [#sec-methods]
@@
 \[this will be moved to SI eventually. note that much of it also the same in solanum-aq project, so might change to cite that paper?\]
+
+(not sure where this goes)  * We acclimated the focal leaf to high light (\ppfdequals{2000}) and high relative humidity (\rhequals{70}) until $A$ and \gsw{} reach their maximum.
+After that, we decreased \rh{} to $\approx 10\%$ to induce rapid stomatal closure without biochemical down
 

### 2025-06-22 r/203_prepare-stan-data.R

@@ -2,7 +2,8 @@
 source("r/header.R")
 
 plant_info = read_rds("data/plant-info.rds") |>
-  dplyr::select(accession, acc_id, light_treatment) 
+  mutate(llma = log(lma_gm2)) |>
+  dplyr::select(accession, acc_id, light_treatment, llma) 
 
 trimmed_amphi_first = read_rds("data/trimmed_amphi_first.rds") |>
   summarize(

### 2025-06-22 r/204_select-model1.R

@@ -1,4 +1,4 @@
-# Fit nonphylogenetic model on posterior estimate of AA to determine model form  for phylogenetic model ----
+# Fit phylogenetic model on posterior estimate of AA 
 source("r/header.R")
 
 d1 = read_rds("objects/stan_data_df.rds")
@@ -12,7 +12,7 @@ fixed_effects = c(
   "light_treatment",
   "light_intensity",
   "light_treatment + light_intensity",
-  "light_treatment * light_intensity"
+  "light_treatment * light_intensity",
 )
 
 random_effects = c(
@@ -31,7 +31,7 @@ sigma_effects = c(
 )
 
 # Cross all combinations
-set.seed(602275201)
+set.seed(622232590)
 model_forms = expand.grid(fixed = fixed_effects,
                           random = random_effects,
                           sigma = sigma_effects,
@@ -39,7 +39,7 @@ model_forms = expand.grid(fixed = fixed_effects,
   mutate(seed = sample(1e9, nrow(.)))
 
 # Build and fit each model
-plan(multisession, workers = 19)
+plan(multisession, workers = 9)
 
 aa_models = future_pmap(model_forms, function(fixed, random, sigma, seed) {
   fml = bf(as.formula(paste("aa ~", fixed, "+", random)), as.formula(paste("sigma ~", sigma)))
@@ -52,12 +52,21 @@ aa_models = future_pmap(model_forms, function(fixed, random, sigma, seed) {
     family = student(),
     save_pars = save_pars(all = TRUE),
     seed = seed
-  ) |>
-    add_criterion("loo", moment_match = TRUE, reloo = TRUE)
+  )
 }, .progress = TRUE, .options = furrr_options(seed = TRUE))
 
 # Extract loo objects
-loos = lapply(aa_models, function(m) m$criteria$loo)
+loos = future_map(
+  aa_models,
+  \(m) loo(
+    m,
+    resp = "aa",
+    moment_match = FALSE,
+    reloo = FALSE
+  ),
+  .progress = TRUE,
+  .options = furrr_options(seed = TRUE)
+)
 
 # Name the models
 names(loos) = paste0("model_", seq_along(loos))
@@ -74,52 +83,7 @@ aa_loo_table = tibble(
     best_model = delta_looic == 0
   )
 
-write_rds(aa_loo_table, "objects/aa_loo_table.rds")
[…]

### 2025-06-22 r/205_select-model2.R

@@ -1,0 +1,99 @@
+# Test whether LMA mediates affect growth light treatment on AA
+source("r/header.R")
+
+d1 = read_rds("objects/stan_data_df.rds")
+tr = read_rds("data/phylogeny.rds")
+A = vcv(tr, corr = TRUE)
+
+fixed_effects = c(
+  "1",
+  "light_treatment",
+  "light_intensity",
+  "mi(llma)",
+  "light_treatment + light_intensity",
+  "light_treatment * light_intensity",
+  "light_treatment + mi(llma)",
+  "light_treatment * mi(llma)",
+  "light_intensity + mi(llma)",
+  "light_intensity * mi(llma)",
+  "light_treatment + light_intensity + mi(llma)",
+  "light_treatment * light_intensity + mi(llma)",
+  "light_treatment * mi(llma) + light_intensity",
+  "light_treatment + light_intensity * mi(llma)",
+  "light_treatment * light_intensity * mi(llma)"
+)
+
+random_effects = c(
+  "(1 | gr(acc, cov = A))",
+  "(1 + light_treatment | gr(acc, cov = A))",
+  "(1 + light_intensity | gr(acc, cov = A))",
+  "(1 + light_treatment + light_intensity | gr(acc, cov = A))",
+  "(1 + light_treatment * light_intensity | gr(acc, cov = A))"
+)
+
+sigma_effects = c(
+  "1",
+  "light_treatment",
+  "light_intensity",
+  "light_treatment + light_intensity"
+)
+
+# Cross all combinations
+set.seed(602275201)
+model_forms = expand.grid(fixed = fixed_effects,
+                          random = random_effects,
+                          sigma = sigma_effects,
+                          stringsAsFactors = FALSE) %>%
+  mutate(seed = sample(1e9, nrow(.)))
+
+# Build and fit each model
+plan(multisession, workers = 19)
+
+aa_models = future_pmap(model_forms, function(fixed, random, sigma, seed) {
+  fml = bf(as.formula(paste("aa ~", fixed, "+", random)), as.formula(paste("sigma ~", sigma)))
+  brm(
+    formula = fml + bf(llma |
+                         mi() ~ light_treatment + (1 + light_treatment |
+                                                     gr(acc, cov = A))),
+    data = d1,
+    data2 = list(A = A),
+    backend = "cmdstanr",
+    chains = 1,
[…]

### 2025-06-22 ms/ms.qmd
working on ms intro

@@
 They are essential for balancing carbon gained per unit water lost and permitted vascular plants to grow tall on land by enabling access to CO$_2$ for photosynthetic carbon assimilation while preventing hydraulic failure in variable environments [@raven_selection_2002; @mccadam_stomata_2021; @clark_origin_2022].
-Optimal stomatal function depends on both dynamic changes in aperture on the scale of minutes to hours, as well as static anatomy determined by developmental plasticity and constitutive genetic differences [@boer_optimal_2016; @harrison_influence_2020; more citations].
-Understanding how stomata respond to environmental change over daily, developmental, and evolutionary time is important for understanding adaptation (cites), predicting paleoclimate from fossil cuticles (cites), and improving crops (cites).
-Stomatal function contributes to global carbon and water cycles and therefore predicting future climate (cites).
+Optimal stomatal function depends on both dynamic changes in aperture on the scale of minutes to hours, as well as static anatomy determined by developmental plasticity and constitutive genetic differences [@hetherington_role_2003; @boer_optimal_2016; @harrison_influence_2020].
+Understanding how stomata respond to environmental change over daily, developmental, and evolutionary time is important for understanding adaptation [@woodward_stomatal_1987; @buckley_modeling_2013; @raven_selection_2002; @haworth_coordination_2013; @liang_stomatal_2023; @chua_stomatal_2024; @lang_century_2024], predicting paleoclimate from fossil cuticles [@franks_new_2014; @mcelwain_paleoecology_2017; @honisch_toward_2023], and improving crops [@hofmann_impact_2025].
+Stomatal function contributes to global carbon and water cycles [@berry_stomata_2010] and therefore predicting future climate [@franks_stomatal_2017].
 

### 2025-06-22 ms/ms.qmd
working on paragraph 2

@@
 <!-- Paragraph 2 -->
-Despite extensive theoretical and empirical progress understanding stomata function and anatomy from molecular to ecosystem levels, the adaptive significance of amphistomatous leaves remains an important unsolved problem in leaf structure-function relationships (parkhurst, grubb, gibson, drake).
-Amphistomatous leaves develop abaxial and adaxial stomata that can be independently regulated (reich, mott) to allow gas exchange through each surface (wall, marquez).
+Despite extensive theoretical and empirical progress understanding stomata function and anatomy from molecular to ecosystem levels, the adaptive significance of amphistomatous leaves remains an important unsolved problem in leaf structure-function relationships [@grubb_leaf_1977; @parkhurst_adaptive_1978; @mott_adaptive_1982;@gibson_structure_1996; @oguchi_leaf_2018; @drake_two_2019; @grubb_leaf_2020].
+Amphistomatous leaves develop abaxial and adaxial stomata that can be independently regulated [@mott_stomatal_1984; @reich_effects_1985; @mott_asymmetric_1993] (reich, mott) to allow gas exchange through each surface (wall, marquez).
 In theory, gas exchange through stomata on each surface increases CO$_2$ supply to chloroplasts, all else being equal, enhancing photosynthesis (parkhurst, gutschick, fig.

### 2025-06-22 ms/ms.qmd
working on ms

@@
 Despite extensive theoretical and empirical progress understanding stomata function and anatomy from molecular to ecosystem levels, the adaptive significance of amphistomatous leaves remains an important unsolved problem in leaf structure-function relationships [@grubb_leaf_1977; @parkhurst_adaptive_1978; @mott_adaptive_1982;@gibson_structure_1996; @oguchi_leaf_2018; @drake_two_2019; @grubb_leaf_2020].
-Amphistomatous leaves develop abaxial and adaxial stomata that can be independently regulated [@mott_stomatal_1984; @reich_effects_1985; @mott_asymmetric_1993] (reich, mott) to allow gas exchange through each surface (wall, marquez).
-In theory, gas exchange through stomata on each surface increases CO$_2$ supply to chloroplasts, all else being equal, enhancing photosynthesis (parkhurst, gutschick, fig.
-1).
+Amphistomatous leaves develop abaxial and adaxial stomata that can be independently regulated [@pospisilova_stomatal_1980; @mott_stomatal_1984; @reich_effects_1985; @mott_asymmetric_1993; @wall_stomata_2022] to allow gas exchange through each surface.
+In theory, gas exchange through stomata on each surface increases CO$_2$ supply to chloroplasts, all else being equal, enhancing photosynthesis [@parkhurst_adaptive_1978; @gutschick_photosynthesis_1984].
 Amphistomatous leaves also lose more water through evaporation because of a second boundary layer conductance (Smith), but the additional carbon gain should be enough to offset this cost in most realistic scenarios (Muir 2019).
@@
 <!-- Paragraph 3 -->
+<!-- Need to mention here or in previous grafs, estimates of r_ias Mott & Oleary 1984; Marquez et al., etc. -->
+<!-- "Farquhar and Raschke (2).
+However, using a technique similar to ours, Sharkey et al.
+(11) estimate r, in Xan:hium to be closer to 1.0 m2 s mol-', and state that they believe the estimate of 3.2 m2 s mol-' to be high" -->
+
 The paradoxical fact is that, despite the potential photosynthetic benefit, most leaves are not amphistomatous.
@@
 However, these traits were not associated with change in AA (table/figures).
+
+from Mott and Oleary: "However, Jones and Slatyer (5) reported a higher mesophyll resistance for CO, entering through the upper stomata than for the lower, and the data of Vaclavik (12) appear to support this conclusion"
 

### 2025-06-22 r/207_plot-rxn-norm.R

@@ -1,6 +1,6 @@
 source("r/header.R")
 
-fit_aa = read_rds("objects/fit_aa.rds")
+fit_aa = read_rds("objects/fit_aa1.rds")
 d1 = fit_aa$data
 
 # code for brms model

### 2025-06-23 ms/ms.qmd
working on intro

--- 2025-06-22 ms/ms.qmd
@@
 <!-- Paragraph 2 -->
-Despite extensive theoretical and empirical progress understanding stomata function and anatomy from molecular to ecosystem levels, the adaptive significance of amphistomatous leaves remains an important unsolved problem in leaf structure-function relationships (parkhurst, grubb, gibson, drake).
-Amphistomatous leaves develop abaxial and adaxial stomata that can be independently regulated (reich, mott) to allow gas exchange through each surface (wall, marquez).
-In theory, gas exchange through stomata on each surface increases CO$_2$ supply to chloroplasts, all else being equal, enhancing photosynthesis (parkhurst, gutschick, fig.
-1).
-Amphistomatous leaves also lose more water through evaporation because of a second boundary layer conductance (Smith), but the additional carbon gain should be enough to offset this cost in most realistic scenarios (Muir 2019).
+<!-- Parkhurst gradients 1988 -->
+<!-- Syvertsen et al 1995 - may be more here to check about gi in sun and shade-->
+<!-- "Farquhar and Raschke (2).
+However, using a technique similar to ours, Sharkey et al.
+(11) estimate r, in Xan:hium to be closer to 1.0 m2 s mol-', and state that they believe the estimate of 3.2 m2 s mol-' to be high" -->
+Despite extensive theoretical and empirical progress understanding stomata function and anatomy from molecular to ecosystem levels, the adaptive significance of amphistomatous leaves remains an important unsolved problem in leaf structure-function relationships [@grubb_leaf_1977; @parkhurst_adaptive_1978; @mott_adaptive_1982; @gibson_structure_1996; @smith_leaf_1997; @oguchi_leaf_2018; @drake_two_2019; @grubb_leaf_2020].
+Amphistomatous leaves develop abaxial and adaxial stomata whose aperture can be independently regulated [@pospisilova_stomatal_1980; @mott_stomatal_1984; @reich_effects_1985; @mott_asymmetric_1993; @wall_stomata_2022] to control gas exchange through each surface.
+All else being equal, gas exchange through stomata on both surfaces increases CO$_2$ supply to chloroplasts by providing a second parallel pathway through leaf intercellular airspaces, enhancing photosynthesis [@parkhurst_adaptive_1978; @gutschick_photosynthesis_1984].
+The extent to which amphistomy increases CO$_2$ supply depends on resistance to diffusion in intercellular airspaces.
+This resistance can be low in thin, porous, amphistomatous leaves [@mott_stomatal_1984; @marquez_assessing_2023], but may be more substantial in thick, dense, hypostomatous leaves [@syvertsen_relationship_1995].
+Amphistomatous leaves also lose more water through evaporation because of a second boundary layer conductance [@foster_influence_1986], but the additional carbon gain should be enough to offset this cost in most realistic scenarios [@muir_amphistomy_2019].
 
 <!-- Paragraph 3 -->
-The paradoxical fact is that, despite the potential photosynthetic benefit, most leaves are not amphistomatous.
-Many vertically oriented and/or isobilateral leaves are amphistomatous (arcto paper, drake).
-But among dorsiventral leaves, herbaceous plants in open, high light habitats tend to have have amphistomatous leaves (salisbury, bucher, jordan, etc.).
+The paradoxical fact is that, despite the photosynthetic benefit, most leaves are not amphistomatous.
+Many vertically oriented and/or isobilateral leaves are amphistomatous [@howell_concerning_1945, @drake_two_2019].
+But among dorsiventral leaves, herbaceous plants in open, high light habitats tend to have have amphistomatous leaves [salisbury, @jordan_using_2014; @bucher_stomatal_2017; @muir_light_2018].
 Most other leaves, except those from aquatic habitats, are hypostomatous, producing stomata only on the lower, abaxial surface.

### 2025-06-23 ms/ms.qmd
working on ms

@@
 
-# I am mostly putting placeholders here for objects/values that will need to be
-# loaded eventually to use in paper.
-# See code in 'solanum-aq' which can probably be reused here.
 aa_args = read_rds("../objects/aa_args1.rds")
@@
 
+rh_curves = read_rds("../data/trimmed_rh_curves.rds")
+
+stomata = read_rds("../data/stomata.rds")
+
 fit_aa = read_rds("../objects/fit_dummy.rds")
@@
 
-n_plant = 600
-n_acc = 30
-mean_n_rep = 10 # average number of individual replicates for accession per light treatment
+df_aa_pred1 = read_rds("../objects/df_aa_pred1.rds")
 
-n_weeks = 80 # number of weeks for LICOR measurements
-start_licor_date = "May 1, 2022"
-end_licor_date = "October 31, 2023"
+n_plant = length(unique(rh_curves$acc_id))
+n_acc = length(unique(rh_curves$acc))
+# average number of individual replicates for accession per light treatment
+mean_n_rep = rh_curves |>
+  summarize(
+    n = length(unique(acc_id)),
+    .by = c("acc", "light_treatment")
+  ) |>
+  pull(n) |>
+  mean() |>
+  prettyNum(digits = 3)
+
+licor_date_range = plant_info |>
+  filter(acc_id %in% unique(rh_curves$acc_id)) |>
+  dplyr::select(`1s_rh_response_date`, `2s_rh_response_date`) |>
+  pivot_longer(cols = everything()) |>
+  pull(value) |>
+  ymd() |>
+  range(na.rm = TRUE)
+
+n_weeks = prettyNum(as.numeric(difftime(licor_date_range[2], licor_date_range[1], units = "days")) / 7, digits = 3)
+ # number of weeks for LICOR measurements
+start_licor_date = format(licor_date_range[1], "%B %d, %Y")
+end_licor_date = format(licor_date_range[2], "%B %d, %Y")
 
@@
 percent_lateral = df_leaflet_type$percent_lateral
-percent_change_leaflet = "10.0" # percent of time that anatomical traits were measured on different leaflet than LICOR measurements
-percent_aa_and_aq = "90.0" # percent of times where AA and AQ curves were done on same leaflet
+# percent of time that anatomical traits were measured on different leaflet than LICOR measurements
+percent_change_leaflet = full_join(
+  plant_info |>
+    filter(acc_id %in% unique(rh_curves$acc_id)) |>
+    select(acc_id, licor = leaflet),
+  stomata |>
+    mutate(acc_id = str_c(accession, replicate, sep = "-")) |>
+    filter(acc_id %in% unique(rh_curves$acc_id)) |>
+    select(acc_id, stomata = leaflet),
+  by = join_by(acc_id)
+) |>
+  summarize(x = mean(licor != stomata, na.rm = TRUE)) |>
+  pull(x) |>
+  multiply_by(100) |>
+  prettyNum(digits = 3)
+
+ # percent of times where AA and AQ curves were done on same leaflet
+percent_aa_and_aq = plant_info |>
+  filter(acc_id %in% unique(rh_curves$acc_id)) |>
+  mutate(same_leaflets = !switch_leaflets) |>
+  pull(same_leaflets) |>
+  mean() |>
+  multiply_by(100) |>
+  prettyNum(digits = 3)
 
 # data on chamber light conditions
-p_red_high = "97.7" # percent red LED in high light-intensity treatment
-p_blue_high = "2.3"
-p_red_low = "87.0"
-p_blue_low = "13.0"
+percent_red = rh_curves |>
+  mutate(p_red = Q_red / Q) |>
+  summarize(p_red = mean(p_red, na.rm = TRUE),
+            .by = c(acc_id, light_intensity)) |>
+  summarize(p_red = mean(p_red, na.rm = TRUE), .by = light_intensity) |>
+  pivot_wider(names_from = light_intensity, values_from = p_red)
 
-percent_lma_switch = "5.0" # percent measure LMA on alternate leaflet
-percent_lma_missing = "3.0" # percent leaves missing LMA
+p_red_high = format(signif(100 * percent_red$`2000`, 3), nsmall = 1)
+p_blue_high = format(signif(100 * (1 - percent_red$`2000`), 3), nsmall = 1)
+p_red_low = format(signif(100 * percent_red$`150`, 3), nsmall = 1)
+p_blue_low = format(signif(100 * (1 - percent_red$`150`), 3), nsmall = 1)
+
+# percent measure LMA on alternate leaflet
+percent_lma_switch = plant_info |>
+  filter(acc_id %in% rh_curves$acc_id) |> 
+  summarize(x = mean(leaflet != lma_leaflet, na.rm = TRUE)) |>
+  pull(x) |>
+  multiply_by(100) |>
+  prettyNum(digits = 3)
+
+ # percent leaves missing LMA
+percent_lma_missing = plant_info |>
+  filter(acc_id %in% rh_curves$acc_id) |> 
+  summarize(x = mean(is.na(lma_leaflet))) |>
+  pull(x) |>
+  multiply_by(100) |>
+  prettyNum(digits = 3)
 
 ### Cleaning RH curves
-n_rh_curve_raw = 2400 # export from adaptive-amphistomy/r/07_process-licor.R
-n_points_per_rh_curves_raw = 100 # export from adaptive-amphistomy/r/07_process-licor.R
 ca_min = aa_args$ca_min
@@
 thinning_interval = aa_args$thinning_interval
+
+### Cleaning AQ curves
 aq_r2_threshold = aa_args$aq_r2_threshold
 
-### Cleaning AQ curves
+### Bayesian data analysis
+n_model = 5 |> # i don't think I need this anymore
+  number_to_word()
 
-### Bayesian data analysis
-n_model = 5 |>
-  number_to_word()
+### Summarizing AA results
+aa_mean_range = df_aa_pred1 |>
+  ungroup() |>
+  mutate(signif = sign(.lower) == sign(.upper)) |>
+  summarize(
+    min = min(aa),
+    max = max(aa),
+    aa = mean(aa),
+    n_signif = sum(signif),
+    .by = c(light_treatment, light_intensity)
+  ) |>
+  mutate(x = sprintf(
+    "%.3f (range: %.3f–%.3f; %.0f of 29 accessions significant)",
+    aa,
+    min,
+    max,
+    n_signif
+  ))
 
@@
 The paradoxical fact is that, despite the photosynthetic benefit, most leaves are not amphistomatous.
-Many vertically oriented and/or isobilateral leaves are amphistomatous [@howell_concerning_1945, @drake_two_2019].
-But among dorsiventral leaves, herbaceous plants in open, high light habitats tend to have have amphistomatous leaves [salisbury, @jordan_using_2014; @bucher_stomatal_2017; @muir_light_2018].
+Many vertically oriented and/or isobilateral leaves are amphistomatous [@wood_physiology_1934; @howell_concerning_1945, @drake_two_2019].
+But among dorsiventral leaves, herbaceous plants in open, high light habitats tend to have have amphistomatous leaves [@salisbury_causes_1928; @peat_comparative_1994; @gibson_structure_1996; @jordan_using_2014; @bucher_stomatal_2017; @muir_light_2018; @triplett_stomatal_2025].
 Most other leaves, except those from aquatic habitats, are hypostomatous, producing stomata only on the lower, abaxial surface.
-Even resupinate leaves develop stomata on the lower, albeit adaxial surface (Lesheyde?), suggesting that leaf orientation (lower vs.
+Even resupinate leaves develop stomata on the lower, albeit adaxial surface [@lyshede_comparative_2002], suggesting that leaf orientation (lower vs.
 upper) rather than leaf polarity (abaxial vs.
@@
 The covariation between stomatal density ratio and light habitat is both qualitative and quantitative.
-A higher proportion of sun leaves are amphistomatous (salisbury) and the proportion of stomata on the upper, adaxial surface increases with light (muir 2018, bucher?).
+A higher proportion of sun leaves are amphistomatous [@salisbury_causes_1928] and the proportion of stomata on the upper, adaxial surface increases with light [@bucher_stomatal_2017; @muir_light_2018].
 Resolving why high light intensity favors amphistomatous dorsivental leaves is an important first step toward understanding variation in stomatal density ratio and leaf structure-function relationships more generally.
@@
 The overarching hypothesis is that leaves with greater stomatal density ratio are more common in open, sunny habitats because they increase photosynthesis most in those circumstances.
-An amphistomatous leaf increases photosynthetic carbon gain compared to an otherwise identical hypostomatous leaf by increasing conductance through the leaf intercellular airspaces and boudary layers; the additional water loss through a second boundary layer is typically small (smith paper, muir 2019).
-We quantify this benefit as the amphistomy advantage (AA = log(Aamphi/Ahypo) (Parkhurst, Triplett).
+An amphistomatous leaf increases photosynthetic carbon gain compared to an otherwise identical hypostomatous leaf by increasing conductance through the leaf intercellular airspaces and boudary layers; the additional water loss through a second boundary layer is typically small [@foster_influence_1986].
+We quantify this benefit as the amphistomy advantage (AA = log(Aamphi/Ahypo) [@parkhurst_adaptive_1978; @triplett_amphistomy_2024].
 Why would AA be greater in sun than shade?
@@
 
-Acclimatory hypothesis: Photosynthetic induction to high light intensity typically involves increases in total leaf stomatal conductance (increased CO\$\_2\$ supply), the concentration of active Rubisco, and electron transport capacity (increased CO\$*2\$* demand).
-A one-dimensional circuit model using the Farquhar-von Caemmerer-Berry biochemical model of C\$\_3\$ photosynthesis shows that both increased stomatal conductance and Rubisco activity should increase AA, all else being equal (Supporting Information).
+Acclimatory hypothesis: Photosynthetic induction to high light intensity typically involves increases in total leaf stomatal conductance (increased CO$_2$ supply), the concentration of active Rubisco, and electron transport capacity (increased CO$_2$ demand).
+A one-dimensional circuit model using the Farquhar-von Caemmerer-Berry biochemical model of C$_3$ photosynthesis [@farquhar_biochemical_1980] shows that both increased stomatal conductance and Rubisco activity should increase AA, all else being equal (Supporting Information).
 If the acclimatory hypothesis is correct, we predict that AAhigh \> AAlow for all species regardless of native habitat or growth environment.
@@
 
-Plastic hypothesis: Individuals of the same genotype often develop dramatically different leaves in sun and shade conditions (cite).
-Plastic responses are likely adaptations to optimize photosynthesis at different light intensities in variable environments (cites).
+Plastic hypothesis: Individuals of the same genotype often develop dramatically different leaves in sun and shade conditions [@poorter_meta-analysis_2019].
+Plastic responses are likely adaptations to optimize photosynthesis at different light intensities in variable environments [@givnish_common-garden_2014].
 Plastic changes in leaf anatomy and biochemistry could modulate AA as a byproduct.
@@
 We therefore predict genotypes from more sunny, open habitats will have consistently greater AA under also measurement and growth light intensities.
-For herbaceous plants, light intensity is largely a function of the tree canopy (cites?).
+For herbaceous plants, light intensity is largely a function of the tree canopy [@engelbrecht_evaluation_2001].
 Herbs growing in the open will regularly experience high light intensity; herbs growing under a forest canopy will often experience low light intensity.
@@
 <!-- Paragraph 5 (summarize methods) -->
-We tested these hypotheses by comparing AA among amphistomatous wild tomato species (systematic cites) from different native light habitats, grown under simulated sun and shade light treatments, and measured under contrasting light intensity (Figure of hypotheses and predictions).
+We tested these hypotheses by comparing AA among amphistomatous wild tomato species [@peralta_taxonomy_2008] from different native light habitats, grown under simulated sun and shade light treatments, and measured under contrasting light intensity (Figure of hypotheses and predictions).
 We measured AA on `r n_plant` individual plants from `r n_acc` accessions (average of `r mean_n_rep` replicates per light treatment) using a recently developed method [@triplett_amphistomy_2024].
@@
 We estimated 'amphistomy advantage' (\aax) *sensu* @parkhurst_adaptive_1978, but with modifications previously described in @triplett_amphistomy_2024 and here (Materials and Methods).
-The native light intensity was represented by plant area index (PAI $\unit{\meter\squared\per\meter\squared}$), estimated using a global gridded X-m2 resolution data set derived from the Global Ecosystem Dynamics Investigation (GEDI; cite) and georeferenced accession collection information (cite TGRC).
+The native light intensity was represented by plant area index (PAI $\unit{\meter\squared\per\meter\squared}$), estimated using a global gridded X-m2 resolution data set derived from the Global Ecosystem Dynamics Investigation [GEDI; @burns_multi-resolution_2024] and georeferenced accession collection information from the Tomato Genetics Resource Center.
 The growth light intensities were `r ppfd_sun` (sun treatment) and `r ppfd_shade` (shade treatment) while all other environment conditions were nearly identical (see @sec-methods for supporting detail).
@@
 
+#### MARKER
 <!-- Paragraph 6 (summarize main results) -->
-Consistent with biophysical theory of CO2 diffusion within leaves, AA > 0 for all accessions (Figure X).
+Consistent with biophysical theory of CO$_2$ diffusion within leaves, AA > 0 for all accessions (Figure X).
 AA varied substantially between measurement light intensities, growth light intensities, and among accessions (point to model comparison evidence).
 Measured under high light intensity, AA was consistently greater for sun plants.
-The average AA among accessions in the shade treat was X (X of 29 accessions significantly greater than 0); however, the same accessions grown at high light intensity had an AA of X (X of 29 accession greater than 0).
+The average AA among accessions in the shade treatment was `r filter(aa_mean_range, light_treatment == "low", light_intensity == "2000")$x`; however, the same accessions grown at high light intensity showed a mean AA of `r filter(aa_mean_range, light_treatment == "high", light_intensity == "2000")$x`.
 Contrary to the predictions of the assimilatory hypothesis, AA was consistently greater under low measurement light intensity.
 Among sun and shade grown plants, AA was greater under low light intensity in X of 29 accessions (significantly greater than 0 in X of 29 accessions).
-The overall average AA of shade and sun grown plants measured under low light intensity was X (+/-) and X (+/-), respectively.
+The overall average AA of shade and sun grown plants measured under low light intensity was `r filter(aa_mean_range, light_treatment == "low", light_intensity == "150")$x` and `r filter(aa_mean_range, light_treatment == "high", light_intensity == "150")$x`, respectively.
 There was a slight tendency for accessions from more closed habitats (greater PAI), but AA varied widely among accessions from open habitats (low PAI) regardless of growth and measurement light intensities (Figure X).
 The pattern of AA across wild tomatoes strongly supports the plasticity hypothesis, argues against the acclimatory hypothesis, and provides only weak support for the constitutive hypothesis.
+
 Not sure where this goes: We infer this from the fact that blocking gas exchange in pseudohypostomatous leaves reduced $A$ by X-X% depending on the accession, light treatment, and light intensity (Table/figure).

### 2025-06-23 r/200_fit-curves.R

@@ -9,7 +9,7 @@ plan(multisession, workers = 19)
 rh_curves |>
   split( ~ curve) |>
   future_iwalk(\(df, curve_id) {
-    # if (!file.exists(paste0("objects/curve-fits/", curve_id, ".rds"))) {
+    if (!file.exists(paste0("objects/curve-fits/", curve_id, ".rds"))) {
       x = 2
       ad = 0.8
       n_divergent = Inf
@@ -35,7 +35,7 @@ rh_curves |>
       
       write_rds(fit_curve,
                 paste0("objects/curve-fits/", curve_id, ".rds"))
-    # } else {
-    #   message(paste0("Curve fit for ", curve_id, " already exists. Skipping."))
-    # }
+    } else {
+      message(paste0("Curve fit for ", curve_id, " already exists. Skipping."))
+    }
   }, .progress = TRUE, .options = furrr_options(seed = TRUE))

### 2025-06-23 r/207_plot-rxn-norm.R

@@ -4,7 +4,6 @@ fit_aa = read_rds("objects/fit_aa1.rds")
 d1 = fit_aa$data
 
 # code for brms model
-
 df_new = crossing(
   acc = unique(d1$acc),
   light_treatment = c("low", "high"),
@@ -12,6 +11,7 @@ df_new = crossing(
 ) |>
   mutate(row = row_number())
 
+# Estimates of AA for each accession in each light treatment and light intensity
 df_aa_pred1 = posterior_epred(fit_aa, newdata = df_new) |>
   t() |>
   as_draws_df() |>
@@ -31,6 +31,54 @@ df_aa_pred1 = posterior_epred(fit_aa, newdata = df_new) |>
       factor(levels = c("150", "2000")) |>
       fct_recode(low = "150", high = "2000")
 )
+
+# Estimate difference in AA between high and low light treatment
+# This would need to be changed for model where effect of light_treatment varies 
+# by accession
+df_new = crossing(
+  acc = first(d1$acc),
+  light_treatment = c("low", "high"),
+  light_intensity = c("150", "2000")
+) |>
+  mutate(row = row_number())
+
+df_aa_pred2 = posterior_epred(fit_aa, newdata = df_new) |>
+  t() |>
+  as_draws_df() |>
+  mutate(row = row_number()) |>
+  pivot_longer(cols = -row, names_to = "draw", values_to = "aa") |>
+  full_join(df_new, by = "row") |>
+  dplyr::select(-row) |>
+  pivot_wider(    
+    names_from = light_treatment,
+    values_from = aa
+  ) |>
+  mutate(d_aa = high - low) |>
+  group_by(light_intensity) |>
+  point_interval(d_aa) 
+
+# Estimate difference in AA between high and low light intensity
+df_new = crossing(
+  acc = unique(d1$acc),
+  light_treatment = c("low", "high"),
+  light_intensity = c("150", "2000")
+) |>
+  mutate(row = row_number())
+
+df_aa_pred3 = posterior_epred(fit_aa, newdata = df_new) |>
+  t() |>
+  as_draws_df() |>
+  mutate(row = row_number()) |>
+  pivot_longer(cols = -row, names_to = "draw", values_to = "aa") |>
+  full_join(df_new, by = "row") |>
+  dplyr::select(-row) |>
+  pivot_wider(    
[…]

### 2025-06-24 ms/ms.qmd
updated bib and aa_stats

@@
 library(dplyr)
+library(lubridate)
+library(magrittr)
 library(readr)
@@
 library(tibble)
+library(tidyr)
 
@@
 
-fit_aa = read_rds("../objects/fit_dummy.rds")
+fit_aa = read_rds("../objects/fit_aa1.rds")
 dim_fit_aa = dim(fit_aa$fit)
@@
 df_aa_pred1 = read_rds("../objects/df_aa_pred1.rds")
+# df_aa_pred2 = read_rds("../objects/df_aa_pred2.rds")
+df_aa_pred3 = read_rds("../objects/df_aa_pred3.rds")
 
@@
 
-df_growth_summary = read_rds("data/df_growth_summary.rds") |>
+df_growth_summary = read_rds("../data/df_growth_summary.rds") |>
   ungroup() |>
@@
     n_signif = sum(signif),
+    n_acc = n(),
     .by = c(light_treatment, light_intensity)
@@
   mutate(x = sprintf(
-    "%.3f (range: %.3f–%.3f; %.0f of 29 accessions significant)",
+    "%.3f (range: %.3f–%.3f; %.0f of %.0f accessions significant)",
     aa,
@@
     max,
-    n_signif
+    n_signif,
+    n_acc
   ))
+
+aa_summary3 = df_aa_pred3 |>
+  mutate(signif = (.upper < 0)) |>
+  summarize(n_signif = sum(signif), .by = light_treatment) |>
+  pull(n_signif, light_treatment)
 
@@
 Stomata are microscopic pores on the surfaces of leaves and other photosynthetic organs formed by a pair of guard cells.
-They are essential for balancing carbon gained per unit water lost and permitted vascular plants to grow tall on land by enabling access to CO$_2$ for photosynthetic carbon assimilation while preventing hydraulic failure in variable environments [@raven_selection_2002; @mccadam_stomata_2021; @clark_origin_2022].
-Optimal stomatal function depends on both dynamic changes in aperture on the scale of minutes to hours, as well as static anatomy determined by developmental plasticity and constitutive genetic differences [@hetherington_role_2003; @boer_optimal_2016; @harrison_influence_2020].
-Understanding how stomata respond to environmental change over daily, developmental, and evolutionary time is important for understanding adaptation [@woodward_stomatal_1987; @buckley_modeling_2013; @raven_selection_2002; @haworth_coordination_2013; @liang_stomatal_2023; @chua_stomatal_2024; @lang_century_2024], predicting paleoclimate from fossil cuticles [@franks_new_2014; @mcelwain_paleoecology_2017; @honisch_toward_2023], and improving crops [@hofmann_impact_2025].
-Stomatal function contributes to global carbon and water cycles [@berry_stomata_2010] and therefore predicting future climate [@franks_stomatal_2017].
+They are essential for balancing carbon gained per unit water lost and permitted vascular plants to grow tall on land by enabling access to CO$_2$ for photosynthetic carbon assimilation while preventing hydraulic failure in variable environments [@raven_selection_2002; @mcadam_stomata_2021; @clark_origin_2022].
+Optimal stomatal function depends on both dynamic changes in aperture on the scale of minutes to hours, as well as static anatomy determined by developmental plasticity and constitutive genetic differences [@hetherington_role_2003; @de_boer_optimal_2016; @harrison_influence_2020].
+Understanding how stomata respond to environmental change over daily, developmental, and evolutionary time is important for understanding adaptation [@woodward_stomatal_1987; @buckley_modelling_2013; @raven_selection_2002; @haworth_co-ordination_2013; @liang_stomatal_2023; @chua_stomatal_2024; @lang_century-long_2024], predicting paleoclimate from fossil cuticles [@franks_new_2014; @mcelwain_paleoecology_2017; @the_cenozoic_co_proxy_integration_project_cencopip_consortium_toward_2023], and improving crops [@hofmann_impact_2025].
+Stomatal function contributes to global carbon and water cycles [@berry_stomata:_2010] and therefore predicting future climate [@franks_stomatal_2017].
 
@@
 (11) estimate r, in Xan:hium to be closer to 1.0 m2 s mol-', and state that they believe the estimate of 3.2 m2 s mol-' to be high" -->
-Despite extensive theoretical and empirical progress understanding stomata function and anatomy from molecular to ecosystem levels, the adaptive significance of amphistomatous leaves remains an important unsolved problem in leaf structure-function relationships [@grubb_leaf_1977; @parkhurst_adaptive_1978; @mott_adaptive_1982; @gibson_structure_1996; @smith_leaf_1997; @oguchi_leaf_2018; @drake_two_2019; @grubb_leaf_2020].
-Amphistomatous leaves develop abaxial and adaxial stomata whose aperture can be independently regulated [@pospisilova_stomatal_1980; @mott_stomatal_1984; @reich_effects_1985; @mott_asymmetric_1993; @wall_stomata_2022] to control gas exchange through each surface.
+Despite extensive theoretical and empirical progress understanding stomata function and anatomy from molecular to ecosystem levels, the adaptive significance of amphistomatous leaves remains an important unsolved problem in leaf structure-function relationships [@grubb_leaf_1977; @parkhurst_adaptive_1978; @mott_adaptive_1982; @gibson_structure-function_1996; @smith_leaf_1997; @oguchi_leaf_2018; @drake_two_2019; @grubb_leaf_2020].
+Amphistomatous leaves develop abaxial and adaxial stomata whose aperture can be independently regulated [@pospisilova_environmental_1980; @mott_stomatal_1984; @reich_effects_1985; @mott_asymmetric_1993; @wall_stomata_2022] to control gas exchange through each surface.
 All else being equal, gas exchange through stomata on both surfaces increases CO$_2$ supply to chloroplasts by providing a second parallel pathway through leaf intercellular airspaces, enhancing photosynthesis [@parkhurst_adaptive_1978; @gutschick_photosynthesis_1984].
 The extent to which amphistomy increases CO$_2$ supply depends on resistance to diffusion in intercellular airspaces.
-This resistance can be low in thin, porous, amphistomatous leaves [@mott_stomatal_1984; @marquez_assessing_2023], but may be more substantial in thick, dense, hypostomatous leaves [@syvertsen_relationship_1995].
-Amphistomatous leaves also lose more water through evaporation because of a second boundary layer conductance [@foster_influence_1986], but the additional carbon gain should be enough to offset this cost in most realistic scenarios [@muir_amphistomy_2019].
+This resistance can be low in thin, porous, amphistomatous leaves [@mott_stomatal_1984; @marquez_assessing_2023], but may be more substantial in thick, dense, hypostomatous leaves [@parkhurst_intercellular_1990].
+Amphistomatous leaves also lose more water through evaporation because of a second boundary layer conductance [@foster_influence_1986], but the additional carbon gain should be enough to offset this cost in most realistic scenarios [@muir_is_2019].
 
@@
 Many vertically oriented and/or isobilateral leaves are amphistomatous [@wood_physiology_1934; @howell_concerning_1945, @drake_two_2019].
-But among dorsiventral leaves, herbaceous plants in open, high light habitats tend to have have amphistomatous leaves [@salisbury_causes_1928; @peat_comparative_1994; @gibson_structure_1996; @jordan_using_2014; @bucher_stomatal_2017; @muir_light_2018; @triplett_stomatal_2025].
+But among dorsiventral leaves, herbaceous plants in open, high light habitats tend to have have amphistomatous leaves [@salisbury_i_1928; @peat_comparative_1994; @gibson_structure-function_1996; @jordan_using_2014; @bucher_stomatal_2017; @muir_light_2018; @triplett_stomatal_2025].
 Most other leaves, except those from aquatic habitats, are hypostomatous, producing stomata only on the lower, abaxial surface.
@@
 The covariation between stomatal density ratio and light habitat is both qualitative and quantitative.
-A higher proportion of sun leaves are amphistomatous [@salisbury_causes_1928] and the proportion of stomata on the upper, adaxial surface increases with light [@bucher_stomatal_2017; @muir_light_2018].
+A higher proportion of sun leaves are amphistomatous [@salisbury_i_1928] and the proportion of stomata on the upper, adaxial surface increases with light [@bucher_stomatal_2017; @muir_light_2018].
 Resolving why high light intensity favors amphistomatous dorsivental leaves is an important first step toward understanding variation in stomatal density ratio and leaf structure-function relationships more generally.
@@
 
-Plastic hypothesis: Individuals of the same genotype often develop dramatically different leaves in sun and shade conditions [@poorter_meta-analysis_2019].
+Plastic hypothesis: Individuals of the same genotype often develop dramatically different leaves in sun and shade conditions [@poorter_metaanalysis_2019].
 Plastic responses are likely adaptations to optimize photosynthesis at different light intensities in variable environments [@givnish_common-garden_2014].
@@
 The average AA among accessions in the shade treatment was `r filter(aa_mean_range, light_treatment == "low", light_intensity == "2000")$x`; however, the same accessions grown at high light intensity showed a mean AA of `r filter(aa_mean_range, light_treatment == "high", light_intensity == "2000")$x`.
-Contrary to the predictions of the assimilatory hypothesis, AA was consistently greater under low measurement light intensity.
-Among sun and shade grown plants, AA was greater under low light intensity in X of 29 accessions (significantly greater than 0 in X of 29 accessions).
+Contrary to the predictions of the assimilatory hypothesis, AA was greater in all accessions under low measurement light intensity for both sun and shade grown plants.
+The effect of low light on AA was more pronounced in the sun-grown plants, where AA was significantly greater under low measurement light intensity in `r aa_summary3["high"]` of `r n_acc` accessions compared to to `r aa_summary3["low"]` accessions for shade-grown plants.
 The overall average AA of shade and sun grown plants measured under low light intensity was `r filter(aa_mean_range, light_treatment == "low", light_intensity == "150")$x` and `r filter(aa_mean_range, light_treatment == "high", light_intensity == "150")$x`, respectively.
@@
 accession_info = read_rds("../data/accession-info.rds") |>
-  filter(accession %in% light_curves$acc) |>
-  left_join(read_rds("../data/accession-climate.rds") |>
-              select(accession, ppfd_mol_m2), by = join_by(accession)) |>
+  filter(accession %in% rh_curves$acc) |>
+  left_join(read_rds("../data/accession-gedi.rds") |>
+              select(accession, pai), by = join_by(accession)) |>
   arrange(species, accession) |>
@@
     `Elevation (mas)` = elev,
-    `$\\mathrm{PPFD}~(\\mathrm{mol}~\\mathrm{m}^2)$` = ppfd_mol_m2
+    `$\\mathrm{PAI}~(\\unit{\\meter\\squared\\per\\meter\\squared})$` = pai
   )
@@
     format = "latex",
-    digits = c(rep(0, 2), 3, 3, 0, 1),
+    digits = c(rep(0, 2), 3, 3, 0, 2),
     booktabs = TRUE,
@@
   "5. thin redundant data points within each \\agcurve{} curve",
-  "6. Trim extreme \\aax{} values"
+  "6. trim extreme \\aax{} values"
 )
@@
 
-We fit Bayesian models with MCMC sampling in the probabilistic programming language *Stan* [@stan_development_team_stan_2023].
-We used CmdStan version `r fit_aa$version$cmdstan` and **cmdstanr** version `r fit_aa$version$cmdstanr` [@gabry_cmdstanr_2023] to interface with *R* version `r getRversion()` [@r_core_team_r:_2023].
+We fit Bayesian models with MCMC sampling in the probabilistic programming language *Stan* [@stan_development_team_stan_2025] using the *R* package **brms** version `r fit_aa$version$brms` [@burkner_brms_2017].
+We used CmdStan version `r fit_aa$version$cmdstan` and **cmdstanr** version `r fit_aa$version$cmdstanr` [@gabry_cmdstanr_2025] to interface with *R* version `r getRversion()` [@r_core_team_r:_2025].
 We sampled the posterior distribution from `r dim_fit_aa[2]` chains with `r dim_fit_aa[1]` iterations each after `r dim_fit_aa[1]` warmup iterations per chain.

### 2025-06-24 ms/ms.tex
updated bib and aa_stats

@@
 \newcommand{\cabetween}[2]{#1 < $C_\mathrm{a} < \qty{#2}{\micro\mol\raiseto{-1}\mol}$}
+\newcommand{\gmaxratio}{$g_\text{max,ratio}$}
 \newcommand{\gsw}{$g_\text{sw}$}
@@
 
-definite: Wei Shen Lim, Dachuan Wang, Chris Muir question: Sam McKlin,
-Tom Buckley
-
-\section{Main}\label{main}
-
-{[}introduction{]} Stomata are important
-
-Stomatal ratio is important unsolved problem
-
-Amphi leaves are more common in sunny habitats and confer a great
-benefit. (this pattern is also true in Solanum? could we analyze
-herbarium data here?)
-
-{[}maybe make distinction between amphi leaves and var in SR here since
-all tomato species are amphi{]}
-
-Leaves with greater stomatal density ratio are more in open, sunny
-habitats because they deliver the greatest benefit in those
-circumstances. An amphistomatous leaf increases photosynthetic carbon
-gain compared to an otherwise identical hypostomatous leaf by increasing
-conductance through the leaf intercellular airspaces and boudary layers;
-the additional water loss through a second boundary layer is typically
-small {[}cites{]}. We quantify this benefit as the amphistomy advantage
-(AA = log(Aamphi/Ahypo). Why would AA be greater in sun than shade?
-There are three nonmutually exclusive hypotheses that we classify as
-`acclimatory', `plastic', and `constitutive'.
-
-Acclimatory hypothesis: Leaves acclimated to high light intensity
-typically increase total leaf stomatal conductance (increased CO\$\_2\$
-supply) and upregulate Rubisco activation (increased CO\$\emph{2\$}
-demand). A one-dimensional circuit model using the Farquhar-von
-Caemmerer-Berry biochemical model of C\$\_3\$ photosynthesis shows that
-both increased stomatal conductance and Rubisco activity should increase
-AA, all else being equal (Supporting Information). If the acclimatory
+definite: Wei Shen Lim, Dachuan Wang, Chris Muir
+
+question: Sam McKlin, Tom Buckley
+
+I am using this approach to drafting the paper:
+
+https://blogs.nature.com/nyc/2011/08/10/how-to-write-a-paper-one-possible-answer
+
+\section{Punchline}\label{punchline}
+
+Plastic changes in the structure of sun leaves can increase the
+intercellular CO\(_2\) diffusion path length. Gas exchange through
+stomata on both upper and lower leaf surfaces, amphistomy, may be a
+compensatory mechanism to reduce the impact of increased diffusion path
+length on photosynthesis without increasing water loss. We demonstrate
+for the first time that sun leaves benefit most from amphistomy because
+it compensates for plastic structural changes in leaf anatomy.
+
+\section{Figures}\label{figures}
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\tightlist
+\item
+  competing hypotheses and predictions (or predictions after methods?)
+  for the benefit of amphistomy
+\item
+  design of experiment to test competing hypotheses (with actual \#'s of
+  replicate, curves, etc.)
+\item
+  Results on acclimatory, plastic, constitutive hypotheses
+\item
+  Additional supporting figure on change in LMA, reduction of Amass, and
+  relationship between LMA and AA? Modeling?
+\end{enumerate}
+
+\section{Abstract}\label{abstract}
+
+(use for GRC)
+
+Developmental plasticity to light intensity modulates amphistomy
+advantage
+
+The presence of stomata on both leaf surfaces (amphistomy) can increase
+photosynthesis in C\(_3\) plants by reducing the path length for
+CO\(_2\) diffusion between substomatal cavities and chloroplasts.
+Amphistomatous leaves are most common among herbaceous plants growing in
+sunny habitats, including many crop species. The distribution of
+amphistomatous leaves in nature may result from an increased
+photosynthetic benefit of amphistomy under high light intensity, either
+because of acclimatory, plastic, or constitutive variation in CO\(_2\)
+supply or demand. We used a recently developed method to quantify
+amphistomy advantage, the photosynthetic rate of an amphistomatous leaf
+relative to an otherwise identical hypostomatous leaf, in 29 diverse
+populations representing 15 species of wild tomatoes (\emph{Solanum}
+sect. \emph{Lycopersicon} and sect. \emph{Lycopersicoides}). Plant grown
+under high light intensity benefit more from amphostomy than those grown
+under low light, regardless of acclimation or native light intensity.
+Curvature of light response curves indicates that high-light leaves
+benefit more from amphistomy because of greater resistance to CO\(_2\)
+diffusion within the mesophyll. Developmental plasticity in leaf
+thickness and/or cell arrangement to optimize light interception likely
+increases under varying light intensity modulates CO\(_2\) diffusion
+within the leaf and, hence, amphistomy advantage as a byproduct.
+Developmental plasticity may therefore play an important and previously
+unnoticed role in explaining the adaptive significance of amphistomy and
+the distribution of amphistomatous leaves in nature.
+
+What's surprising: role of developmental plasticity in explaining
+comparative results; diffusional limitations through palisade ias is
+strong and needs to be considered with light gradients in understanding
+leaf strucuture- funciton
+
+\section{Intro}\label{intro}
+
+Stomata are microscopic pores on the surfaces of leaves and other
+photosynthetic organs formed by a pair of guard cells. They are
+essential for balancing carbon gained per unit water lost and permitted
+vascular plants to grow tall on land by enabling access to CO\(_2\) for
+photosynthetic carbon assimilation while preventing hydraulic failure in
+variable environments (\emph{1}--\emph{3}). Optimal stomatal function
+depends on both dynamic changes in aperture on the scale of minutes to
+hours, as well as static anatomy determined by developmental plasticity
+and constitutive genetic differences (\emph{4}--\emph{6}). Understanding
+how stomata respond to environmental change over daily, developmental,
+and evolutionary time is important for understanding adaptation
+(\emph{1}, \emph{7}--\emph{12}), predicting paleoclimate from fossil
+cuticles (\emph{13}--\emph{15}), and improving crops (\emph{16}).
+Stomatal function contributes to global carbon and water cycles
+(\emph{17}) and therefore predicting future climate (\emph{18}).
+
+Despite extensive theoretical and empirical progress understanding
+stomata function and anatomy from molecular to ecosystem levels, the
+adaptive significance of amphistomatous leaves remains an important
+unsolved problem in leaf structure-function relationships
+(\emph{19}--\emph{26}). Amphistomatous leaves develop abaxial and
+adaxial stomata whose aperture can be independently regulated
+(\emph{27}--\emph{31}) to control gas exchange through each surface. All
+else being equal, gas exchange through stomata on both surfaces
+increases CO\(_2\) supply to chloroplasts by providing a second parallel
+pathway through leaf intercellular airspaces, enhancing photosynthesis
+(\emph{20}, \emph{32}). The extent to which amphistomy increases
+CO\(_2\) supply depends on resistance to diffusion in intercellular
+airspaces. This resistance can be low in thin, porous, amphistomatous
+leaves (\emph{28}, \emph{33}), but may be more substantial in thick,
+dense, hypostomatous leaves (\emph{34}). Amphistomatous leaves also lose
+more water through evaporation because of a second boundary layer
+conductance (\emph{35}), but the additional carbon gain should be enough
+to offset this cost in most realistic scenarios (\emph{36}).
+
+The paradoxical fact is that, despite the photosynthetic benefit, most
+leaves are not amphistomatous. Many vertically oriented and/or
+isobilateral leaves are amphistomatous (\emph{25}). But among
+dorsiventral leaves, herbaceous plants in open, high light habitats tend
+to have have amphistomatous leaves (\emph{22}, \emph{39}--\emph{44}).
+Most other leaves, except those from aquatic habitats, are
+hypostomatous, producing stomata only on the lower, abaxial surface.
+Even resupinate leaves develop stomata on the lower, albeit adaxial
+surface (\emph{45}), suggesting that leaf orientation (lower vs.~upper)
+rather than leaf polarity (abaxial vs.~adaxial) is causal. The
+covariation between stomatal density ratio and light habitat is both
+qualitative and quantitative. A higher proportion of sun leaves are
+amphistomatous (\textbf{salisbury\_causes\_1928?}) and the proportion of
+stomata on the upper, adaxial surface increases with light (\emph{42},
+\emph{43}). Resolving why high light intensity favors amphistomatous
+dorsivental leaves is an important first step toward understanding
+variation in stomatal density ratio and leaf structure-function
+relationships more generally.
+
+The overarching hypothesis is that leaves with greater stomatal density
+ratio are more common in open, sunny habitats because they increase
+photosynthesis most in those circumstances. An amphistomatous leaf
+increases photosynthetic carbon gain compared to an otherwise identical
+hypostomatous leaf by increasing conductance through the leaf
+intercellular airspaces and boudary layers; the additional water loss
+through a second boundary layer is typically small (\emph{35}). We
+quantify this benefit as the amphistomy advantage (AA =
+log(Aamphi/Ahypo) (\emph{20}, \emph{46}). Why would AA be greater in sun
+than shade? We consider three nonmutually exclusive hypotheses that we
+classify as `acclimatory', `plastic', and `constitutive'.
+
+Acclimatory hypothesis: Photosynthetic induction to high light intensity
+typically involves increases in total leaf stomatal conductance
+(increased CO\(_2\) supply), the concentration of active Rubisco, and
+electron transport capacity (increased CO\(_2\) demand). A
+one-dimensional circuit model using the Farquhar-von Caemmerer-Berry
+biochemical model of C\(_3\) photosynthesis (\emph{47}) shows that both
+increased stomatal conductance and Rubisco activity should increase AA,
+all else being equal (Supporting Information). If the acclimatory
 hypothesis is correct, we predict that AAhigh \textgreater{} AAlow for
@@
 Plastic hypothesis: Individuals of the same genotype often develop
-dramatically different leaves in sun and shade conditions (cite).
+dramatically different leaves in sun and shade conditions (\emph{48}).
 Plastic responses are likely adaptations to optimize photosynthesis at
-different light intensities (cites), but changes in leaf anatomy and
-biochemistry could modulate AA as a byproduct. Thicker or less porous
-leaves, both of which are associated with high leaf mass per area (LMA),
-will have lower g\_ias; leaves with increased total stomatal density and
-Rubisco concentration have greater potential CO2 supply and demand.
-Under the plastic hypothesis, we predict that AAsun \textgreater{}
-AAshade for all species and light intensities. AAsun and gsmax,sun
-should be positively associated with native light habitat.
-{[}transition{]} We assume that genotypes adapted to sunny, open
-habitats will express a phenotype best adapted to that environment when
-leaves develop under high light intensity; genotypes adapted to shaded
-closed habitats may be plastic, but suboptimal for light intensities
-they do not regularly experience in nature.
-
-{[}conceptual figure could also show differential benefit of amphi
-leaves in sun/shade{]}
-
-to plants growing there. in those It has been hypothesized that
-amphistomy increases photosynthesis more in sunny places. Three
-nonmutually exclusive adaptive explanations for why amphi leaves:
-
-Conceptual figure explaining each hypothesis
-
-\begin{itemize}
-\tightlist
-\item
-  acclimation: greater demand, higher gs increase AA
-\item
-  developmental plasticity: light-induced changes in leaf anatomy
-  modulate AA
-\item
-  constitutive: genetic differences in leaves adapted to different light
-  habitats
-\end{itemize}
-
-We distinguished among these hypotheses by comparing AA among wild
-tomato species from different native light habitats, grown under
-simulated sun and shade light treatments, and measured under contrasting
-light intensity (Figure of hypotheses and predictions). We measured AA
-on 600 individual plants from 30 accessions (average of 10 replicates
-per light treatment) using a recently developed method (\emph{1}). With
-this method, we directly compare the photosynthetic rate of an untreated
-amphistomatous leaf to that of the same leaf with gas exchange blocked
-through the adaxial (upper) surface by transparent plastic, which we
-refer to as `pseudohypostomy'. To compare amphi- and pseudohypostomatous
-leaves at identical whole-leaf \(g_\text{sc}\), we measure \(A\) over a
-range of \(g_\text{sc}\), inducing stomatal opening and closure by
-modulating humidity (see Materials and Methods for further details).
-
-Table of directional predictions (with table summarizing results? part
-of conceptual figure? in supplement?)
+different light intensities in variable environments (\emph{49}).
+Plastic changes in leaf anatomy and biochemistry could modulate AA as a
+byproduct. Thicker or less porous leaves, both of which are associated
+with high leaf mass per area (LMA), will have lower \(g_\mathrm{ias}\);
+leaves with increased total stomatal density and photosynthetic capacity
+have greater potential CO\(_2\) supply and demand. Under the plastic
+hypothesis, we predict that AAsun \textgreater{} AAshade for all species
+and light intensities. Secondarily, AAsun and gsmax,sun should also be
+positively associated with native light habitat if genotypes adapted to
+sunny, open habitats if they can express a phenotype best adapted to
+that environment when leaves develop under high light intensity.
+Genotypes adapted to shaded, closed habitats may be plastic, but limits
+on the width of their reaction norms prevent them from developing traits
+optimal for conditions they do not regularly experience in nature.
+
+Constitutive hypothesis: In environments that are relatively constant or
+where environmental change cannot be anticipated by a reliable cue,
+natural selection will favor constitutive expression of optimal
+phenotypes. We therefore predict genotypes from more sunny, open
+habitats will have consistently greater AA under also measurement and
+growth light intensities. For herbaceous plants, light intensity is
+largely a function of the tree canopy (\emph{50}). Herbs growing in the
+open will regularly experience high light intensity; herbs growing under
+a forest canopy will often experience low light intensity.
+
+The primary directional predictions for each hypothesis are summarized
+in Table X; detailed predictions for results that would indicate support
+for multiple hypotheses are in Table SX.
+
+CONCEPTUAL FIGURE HERE
 
@@
 sunny, open habitats. For each hypothesis, we make predictions for how
-native light habitat, light treatment, and light intensity would affect
-AA.
+native plant area index (PAI), growth light treatment, and measurement
+light intensity would affect AA.
 
 \begin{longtable}[]{@{}
-  >{\centering\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.1972}}
-  >{\centering\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.3099}}
-  >{\centering\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.2535}}
-  >{\centering\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.2394}}@{}}
+  >{\centering\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.1728}}
+  >{\centering\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.3580}}
+  >{\centering\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.3210}}
+  >{\centering\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.1481}}@{}}
 \toprule\noalign{}
@@
 \end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-native light habitat
+measurement light intensity
 \end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-light treatment
+growth light intensity
 \end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-light intensity
+native PAI
 \end{minipage} \\
@@
 \endlastfoot
-acclimatory & cor(PAR, AA) = 0 & AAsun = AAshade & AA2000 \textgreater{}
-AA150 \\
-plastic & cor(PAR, AA) = 0 & AAsun \textgreater{} AAshade & AA2000 =
-AA150 \\
-constitutive & cor(PAR, AA) \textgreater{} 0 & AAsun = AAshade & AA2000
-= AA150 \\
+acclimatory & AA2000 \textgreater{} AA150 & AAsun = AAshade & cor(PAI,
+AA) = 0 \\
+plastic & AA2000 = AA150 & AAsun \textgreater{} AAshade & cor(PAI, AA) =
+0 \\
+constitutive & AA2000 = AA150 & AAsun = AAshade & cor(PAI, AA)
+\textgreater{} 0 \\
 \end{longtable}
 
-{[}add rows for when multiple hypotheses are supported simultaneously?
-or put that in SI?{]}
-
-{[}results{]} Amphistomy increases \(A\) in all accessions, in both sun
-and shade leaves, and light intensities. We infer this from the fact
-that blocking gas exchange in pseudohypostomatous leaves reduced \(A\)
-by X-X\% depending on the accession, light treatment, and light
-intensity (Table/figure). The AA is equivalent to an X-X\% change in
-total \(g_\text{sc}\) (see SI section gs equivalency). But whereas
-increasing \(g_\text{sc}\) would increase water loss as a necessary
-by-product, amphistomy can increase \(A\) without any appreciable affect
-on transpiration.
-
-Sun leaves from high light habitats {[}not sure if htis result is true
-yet{]} benefit the most from amphistomy because of a both developmental
-plasticity and constitutive differences among accessions. {[}quantify
-difference in AA and contribution of different affects{]}. Surprisingly,
-light intensity had a little effect\ldots{} (should this be in this
-paragraph, or it's own?).
-
-{[}discussion{]} - sun/shade has long been appreciated and this shows
-new trait that should be considered and that CO2 diffusion becomes major
-limitation
-
-\section{Materials and Methods}\label{materials-and-methods}
-
-{[}this will be moved to SI eventually{]}
+We tested these hypotheses by comparing AA among amphistomatous wild
+tomato species (\emph{51}) from different native light habitats, grown
+under simulated sun and shade light treatments, and measured under
+contrasting light intensity (Figure of hypotheses and predictions). We
+measured AA on 572 individual plants from 29 accessions (average of 9.86
+replicates per light treatment) using a recently developed method
+(\emph{46}). With this method, we directly compare the photosynthetic
+rate of an untreated amphistomatous leaf to that of the same leaf with
+gas exchange blocked through the adaxial (upper) surface by transparent
+plastic, which we refer to as `pseudohypostomy'. To compare amphi- and
+pseudohypostomatous leaves at identical whole-leaf \(g_\text{sc}\), we
+measure \(A\) over a range of \(g_\text{sc}\), inducing stomatal opening
+and closure by modulating humidity (see Materials and Methods for
+further details). We estimated `amphistomy advantage' (\aax)
+\emph{sensu} (\emph{20}), but with modifications previously described in
+(\emph{46}) and here (Materials and Methods). The native light intensity
+was represented by plant area index (PAI
+\(\unit{\meter\squared\per\meter\squared}\)), estimated using a global
+gridded X-m2 resolution data set derived from the Global Ecosystem
+Dynamics Investigation {[}GEDI; (\emph{52}){]} and georeferenced
+accession collection information from the Tomato Genetics Resource
+Center. The growth light intensities were 761 (sun treatment) and 115
+(shade treatment) while all other environment conditions were nearly
+identical (see \textbf{?@sec-methods} for supporting detail). The high
+and low measurements intensities were \ppfdequals{2000} (97.8:2.24
+red:blue) and \ppfdequals{150} (87.0:13.0 red:blue), respectively.
+
+\subsubsection{MARKER}\label{marker}
+
+Consistent with biophysical theory of CO\(_2\) diffusion within leaves,
+AA \textgreater{} 0 for all accessions (Figure X). AA varied
+substantially between measurement light intensities, growth light
+intensities, and among accessions (point to model comparison evidence).
+Measured under high light intensity, AA was consistently greater for sun
+plants. The average AA among accessions in the shade treatment was 0.039
+(range: 0.012--0.109; 21 of 29 accessions significant); however, the
+same accessions grown at high light intensity showed a mean AA of 0.050
+(range: 0.024--0.120; 25 of 29 accessions significant). Contrary to the
+predictions of the assimilatory hypothesis, AA was greater in all
+accessions under low measurement light intensity for both sun and shade
+grown plants. The effect of low light on AA was more pronounced in the
+sun-grown plants, where AA was significantly greater under low
+measurement light intensity in 28 of 29 accessions compared to to 11
+accessions for shade-grown plants. The overall average AA of shade and
+sun grown plants measured under low light intensity was 0.066 (range:
+0.019--0.143; 27 of 29 accessions significant) and 0.098 (range:
+0.051--0.175; 29 of 29 accessions significant), respectively. There was
+a slight tendency for accessions from more closed habitats (greater
+PAI), but AA varied widely among accessions from open habitats (low PAI)
+regardless of growth and measurement light intensities (Figure X). The
+pattern of AA across wild tomatoes strongly supports the plasticity
+hypothesis, argues against the acclimatory hypothesis, and provides only
+weak support for the constitutive hypothesis.
+
+Not sure where this goes: We infer this from the fact that blocking gas
+exchange in pseudohypostomatous leaves reduced \(A\) by X-X\% depending
+on the accession, light treatment, and light intensity (Table/figure).
+The AA is equivalent to an X-X\% change in total \(g_\text{sc}\) (see SI
+section gs equivalency). But whereas increasing \(g_\text{sc}\) would
+increase water loss as a necessary by-product, amphistomy can increase
+\(A\) without any appreciable affect on transpiration.
+
+These results suggest that developmental plasticity, as opposed to
+acclimation or adaptation to open habitats, may explain the explain the
+long-standing observation taht amphistomatous leaves are more common in
+sunny habitats. Our results add to existing explanations by showing high
+light intensity \emph{per se} does not increase the benefit of
+amphistomy, but rather than anatomical and biochemical change associated
+with higher light intensity modulate AA. We cannot yet attribute
+specific changes wrought by plasticity, but correlational evidence
+suggests that changes in leaf anatomy may be important. Species
+responded consistently by increasing stomatal density, especially on the
+adaxial surface, increasing stomatal density ratio. Sun leaves also had
+greater gs, Amax. However, these traits were not associated with change
+in AA (table/figures).
+
+from Mott and Oleary: ``However, Jones and Slatyer (5) reported a higher
+mesophyll resistance for CO, entering through the upper stomata than for
+the lower, and the data of Vaclavik (12) appear to support this
+conclusion''
+
+\section{Outtro - I think this goes at the
+end??}\label{outtro---i-think-this-goes-at-the-end}
+
+It is commonly assumed in comparative studies that most trait variation
+between species is constitutive and genetically determined, but our
+results suggest that plastic responses to light are necessary to explain
+why \gmaxratio{} increases with light habitat. Reaction norms to light
+intensity among wild tomatoes indicate that accessions from low light
+habitats express low \gmaxratio{} when grown in shade, whereas
+accessions from high light habitats have constitutively higher
+\gmaxratio{} (Fig. rxn-norm). The larger response to light intensity is
+driven by greater plasticty in leaf structure in accessions from
+low-light habitats (Fig/table - I haven't tested this yet, so this is a
+placeholder.) We infer that these structural changes in sun leaves are
+costly because they increase the diffusion path length for CO\(_2\) and
+reduce photosynthesis. As a consequence, sun leaves benefit more from
+amphistomy. This is the first demonstration that sun leaves benefit more
+from amphistomy because it compensates for developmental plasticity in
+leaf structure.
+
+Two outstanding areas for further research are 1) the specific
+anatomical changes, such as leaf thickness or mesophyll porosity, that
+increase the diffusion path length for CO\(_2\) in sun leaves, and 2)
+the costs of amphistomy that could explain plasticity in \gmaxratio{} in
+accessions from low-light habitats. Resolving the first question will
+require detailed anatomical measurements on a large number of shade and
+sun leaves with concomitant estimates of \aax. The fact that accessions
+from low-light habitats are more plastic suggests that costs of
+amphistomy are greatest in these environments. (not finished yet)
+
+Problems to bring up in discussion: - all our species were amphi, so
+they may all live in PAI below threshold to favor amphi - other
+microclimate diffs not considered - don't know specific anatomical
+change
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\setcounter{enumi}{7}
+\tightlist
+\item
+  What it all means
+\end{enumerate}
+
+\begin{itemize}
+\tightlist
+\item
+  Our study reveals an important role for plasticity in explaining why
+  amphi leaves are more common in sunny habitats
+\item
+  Confirms that leaf structure of sun leaves incurs costs that
+  amphistomy can mitigate
+\item
+  It will be useful to consider coordination between leaf anatomy,
+  stomatal density, and ratio in future studies of leaf
+  structure-function and in crops (or something)
+\end{itemize}
+
+\section{Materials and Methods
+{[}\#sec-methods{]}}\label{materials-and-methods-sec-methods}
+
+{[}this will be moved to SI eventually. note that much of it also the
+same in solanum-aq project, so might change to cite that paper?{]}
+
+(not sure where this goes) * We acclimated the focal leaf to high light
+(\ppfdequals{2000}) and high relative humidity (\rhequals{70}) until
+\(A\) and \gsw{} reach their maximum. After that, we decreased \rh{} to
+\(\approx 10\%\) to induce rapid stomatal closure without biochemical
+down
 
@@
 
-We compared AA among 30 ecologically diverse accessions of wild tomato,
+We compared AA among 29 ecologically diverse accessions of wild tomato,
 including representatives of all described species of \emph{Solanum}
-sect. \emph{Lycopersicon} and sect. \emph{Lycopersicoides} (\emph{2})
+sect. \emph{Lycopersicon} and sect. \emph{Lycopersicoides} (\emph{51})
 and the cultivated tomato \emph{S. lycopersicum} var.
 \emph{lycopersicum} \autoref{tab:accessions}. Due to constraints on
-growth space and time, we spread out measurements over 80 weeks from May
-1, 2022 to October 31, 2023. Replicates within accession were evenly
-spread out over this period to prevent confounding of temporal variation
-in growth conditions with accession. {[}anything else to say here? maybe
-explain accession selection and phylogeny?{]}
-
-\begin{table}
-
-\caption{\label{tbl-accessions}Solanum accessions}
-
-\centering{
-
-}
-
-\end{table}%
+growth space and time, we spread out measurements over 61.1 weeks from
+August 29, 2022 to October 31, 2023. Replicates within accession were
+evenly spread out over this period to prevent confounding of temporal
+variation in growth conditions with accession. {[}anything else to say
+here? maybe explain accession selection and phylogeny?{]}
+
+\begin{longtable}{>{\raggedright\arraybackslash}p{3.25cm}lrrrr}
+
+\caption{\label{tbl-accessions}Accessions of \emph{Solanum} used in this
+study. The species name, accession number, collection latitude,
+longitude, elevation, and daily solar radiation estimated using the
+SPLASH algorithm (see `Climate data'). TGRC: Tomato Genetics Resource
+Center; PPFD: Photosynthetic Photon Flux Density.}
+
+\tabularnewline
+
+\toprule
+Species & TGRC accession & Latitude & Longitude & Elevation (mas) & $\mathrm{PAI}~(\unit{\meter\squared\per\meter\squared}$\\
+\midrule
+\em{\cellcolor{gray!10}{S. arcanum}} & \cellcolor{gray!10}{LA2172} & \cellcolor{gray!10}{-6.008} & \cellcolor{gray!10}{-78.858} & \cellcolor{gray!10}{662} & \cellcolor{gray!10}{0.7}\\
+\em{S. cheesmaniae} & LA0429 & -0.644 & -90.329 & 800 & 0.6\\
+\em{\cellcolor{gray!10}{S. cheesmaniae}} & \cellcolor{gray!10}{LA3124} & \cellcolor{gray!10}{-0.804} & \cellcolor{gray!10}{-90.042} & \cellcolor{gray!10}{1} & \cellcolor{gray!10}{0.4}\\
+\em{S. chilense} & LA1782 & -15.267 & -74.633 & 1000 & 0.3\\
+\em{\cellcolor{gray!10}{S. chilense}} & \cellcolor{gray!10}{LA4117A} & \cellcolor{gray!10}{-22.907} & \cellcolor{gray!10}{-67.941} & \cellcolor{gray!10}{3540} & \cellcolor{gray!10}{0.2}\\
+\addlinespace
+\em{S. chmielewskii} & LA1028 & -13.883 & -73.017 & 3000 & 0.4\\
+\em{\cellcolor{gray!10}{S. chmielewskii}} & \cellcolor{gray!10}{LA1316} & \cellcolor{gray!10}{-13.400} & \cellcolor{gray!10}{-73.906} & \cellcolor{gray!10}{2920} & \cellcolor{gray!10}{1.1}\\
+\em{S. corneliomulleri} & LA0107 & -13.117 & -76.383 & 60 & 0.0\\
+\em{\cellcolor{gray!10}{S. corneliomulleri}} & \cellcolor{gray!10}{LA0444} & \cellcolor{gray!10}{-13.433} & \cellcolor{gray!10}{-76.133} & \cellcolor{gray!10}{100} & \cellcolor{gray!10}{0.5}\\
+\em{S. galapagense} & LA0436 & -0.953 & -90.978 & 40 & 0.2\\
+\addlinespace
+\em{\cellcolor{gray!10}{S. galapagense}} & \cellcolor{gray!10}{LA1044} & \cellcolor{gray!10}{-0.284} & \cellcolor{gray!10}{-90.548} & \cellcolor{gray!10}{0} & \cellcolor{gray!10}{0.2}\\
+\em{S. habrochaites} & LA0407 & -2.181 & -79.884 & 70 & 0.5\\
+\em{\cellcolor{gray!10}{S. habrochaites}} & \cellcolor{gray!10}{LA1777} & \cellcolor{gray!10}{-9.550} & \cellcolor{gray!10}{-77.700} & \cellcolor{gray!10}{3216} & \cellcolor{gray!10}{0.4}\\
+\em{S. huaylasense} & LA1358 & -9.533 & -77.967 & 750 & 0.6\\
+\em{\cellcolor{gray!10}{S. huaylasense}} & \cellcolor{gray!10}{LA1360} & \cellcolor{gray!10}{-9.546} & \cellcolor{gray!10}{-77.929} & \cellcolor{gray!10}{1490} & \cellcolor{gray!10}{0.5}\\
+\addlinespace
+\em{S. huaylasense} & LA1364 & -10.133 & -77.383 & 2920 & 0.9\\
+\em{\cellcolor{gray!10}{S. lycopersicoides}} & \cellcolor{gray!10}{LA2951} & \cellcolor{gray!10}{-19.317} & \cellcolor{gray!10}{-69.450} & \cellcolor{gray!10}{2200} & \cellcolor{gray!10}{0.5}\\
+\em{S. lycopersicoides} & LA4126 & -19.287 & -69.396 & 3120 & 0.4\\
+\em{\cellcolor{gray!10}{S. neorickii}} & \cellcolor{gray!10}{LA1322} & \cellcolor{gray!10}{-13.483} & \cellcolor{gray!10}{-72.442} & \cellcolor{gray!10}{2380} & \cellcolor{gray!10}{0.7}\\
+\em{S. neorickii} & LA2133 & -3.400 & -79.183 & 1980 & 1.1\\
+\addlinespace
+\em{\cellcolor{gray!10}{S. pennellii}} & \cellcolor{gray!10}{LA0716} & \cellcolor{gray!10}{-16.225} & \cellcolor{gray!10}{-73.617} & \cellcolor{gray!10}{50} & \cellcolor{gray!10}{0.2}\\
+\em{S. pennellii} & LA0750 & -14.775 & -75.034 & 550 & 0.1\\
+\em{\cellcolor{gray!10}{S. pennellii}} & \cellcolor{gray!10}{LA3778} & \cellcolor{gray!10}{-14.775} & \cellcolor{gray!10}{-75.034} & \cellcolor{gray!10}{616} & \cellcolor{gray!10}{0.1}\\
+\em{S. peruvianum} & LA2744 & -18.550 & -70.150 & 400 & 0.2\\
+\em{\cellcolor{gray!10}{S. peruvianum}} & \cellcolor{gray!10}{LA2964} & \cellcolor{gray!10}{-18.028} & \cellcolor{gray!10}{-70.835} & \cellcolor{gray!10}{75} & \cellcolor{gray!10}{1.2}\\
+\addlinespace
+\em{S. pimpinellifolium} & LA1269 & -11.483 & -77.075 & 400 & 0.5\\
+\em{\cellcolor{gray!10}{S. pimpinellifolium}} & \cellcolor{gray!10}{LA1589} & \cellcolor{gray!10}{-8.433} & \cellcolor{gray!10}{-78.817} & \cellcolor{gray!10}{30} & \cellcolor{gray!10}{0.1}\\
+\em{S. pimpinellifolium} & LA2933 & -1.442 & -80.562 & 375 & 1.6\\
+\em{\cellcolor{gray!10}{S. sitiens}} & \cellcolor{gray!10}{LA4116} & \cellcolor{gray!10}{-22.159} & \cellcolor{gray!10}{-68.782} & \cellcolor{gray!10}{2960} & \cellcolor{gray!10}{0.1}\\
+\bottomrule
+
+\end{longtable}
 
@@
 grew seeds and seedlings for both sun and shade treatments under the
-same environmental conditions (12:12 h, 25:\(\qty{20}{\degreeCelsius}\),
-40:60 RH day:night cycle). LED light provided
-\(\mathrm{PPFD} = \qty{200}{\micro\mol\raiseto{-2}\meter\raiseto{-1}\second}\)
+same environmental conditions (12:12 h,
+24.3:\(\qty{21.7}{\degreeCelsius}\), 49.6:58.4 RH day:night cycle). LED
+light provided
+\(\mathrm{PPFD} = \qty{267}{\micro\mol\raiseto{-2}\meter\raiseto{-1}\second}\)
 (Fluence RAZRx, Austin, Texas, USA).
@@
 lower rack to limit diffuse light from reaching the shade treatment. The
-average daytime \ppfd{} was \ppfdqty{800} and \ppfdqty{125} for sun and
+average daytime \ppfd{} was \ppfdqty{761} and \ppfdqty{115} for sun and
 shade treatments, respectively. To isolate the effect of light intensity
@@
 Despite these efforts, the air in the sun treatment was on average
-\(\qty{1}{\degreeCelsius}\) warmer and the average RH was consequently 5
-lower. However, because of evaporative cooling, the leaves in the sun
-treatment were only \(\qty{1}{\degreeCelsius}\) on average (\(n = 600\)
-leaves).
+\(\qty{2.56}{\degreeCelsius}\) warmer and the average RH was
+consequently 5.75 lower. However, because of evaporative cooling, the
+leaves in the sun treatment were only \(\qty{0.886}{\degreeCelsius}\) on
+average (\(n = 699\) leaves).
 
@@
 intensity of the treatment rather than the seedling conditions
-(\emph{3}). Shade plants grew slower than sun plants, hence leaves at
+(\emph{53}). Shade plants grew slower than sun plants, hence leaves at
 the same developmental stage were measured on chronologically older
@@
 inaccessible with the gas exchange equipment. We measured terminal
-leaflets in 70.0\% of cases, but used the lateral leaflet closest to the
+leaflets in 82.6\% of cases, but used the lateral leaflet closest to the
 terminal leaflet when it was damaged or difficult to clamp into the gas
@@
 measurements, we collected anatomical data from the nearest leaflet on
-the same leaf (10.0)\%. We measured chlorophyll concentration index
+the same leaf (1.58)\%. We measured chlorophyll concentration index
 (CCI) using a chrolophyll concentration meter (MC-100, Apogee
@@
 
-We estimated `amphistomy advantage' (\aax) \emph{sensu} (\emph{4}), but
-with modifications previously described in (\emph{1}). \aax{} is
+We estimated `amphistomy advantage' (\aax) \emph{sensu} (\emph{20}), but
+with modifications previously described in (\emph{46}). \aax{} is
 calculated as the log-response ratio of \(A\) compared at the same total
@@
 and a pseudohypo (treated) curve at high light-intensity
-(\ppfdequals{2000}; 97.7:2.3 red:blue) and low light-intensity
+(\ppfdequals{2000}; 97.8:2.24 red:blue) and low light-intensity
 (\ppfdequals{150}; 87.0:13.0 red:blue). We always measured high
@@
 we alternated between starting with amphi or pseudohypo leaf
-measurements. Unlike (\emph{1}), preliminary experiments with
+measurements. Unlike (\emph{46}), preliminary experiments with
 \emph{Solanum} indicated a strong order effect in that \(A\) declined in
@@
 
-In 90.0\% of plants, we measured light-response (\aqcurve) curves on the
+In 91.3\% of plants, we measured light-response (\aqcurve) curves on the
 same leaflets as \agcurve{} curves. However, when a leaflet was damaged
@@
 surfaces from all leaves, using guard cell length as a proxy for
-stomatal size since it proportional to maximum conductance (\emph{5}).
+stomatal size since it proportional to maximum conductance (\emph{54}).
 We made surface impressions of leaf lamina from the same area used for
@@
 dried, we mounted replicas on a microscope slide using transparent tape
-(\emph{6}). We digitized a portion of each leaf surface replica using a
+(\emph{55}). We digitized a portion of each leaf surface replica using a
 brightfield microscope (Leica DM2000, Wetzlar, Germany). We counted and
 measured guard cell length on all stomata using the FIJI implementation
-of ImageJ2 version 2.3.0 (\emph{7}), then divided the count by the
+of ImageJ2 version 2.3.0 (\emph{56}), then divided the count by the
 visible leaf area (\(\qty{0.890}{\raiseto{2}\milli\meter}\)) to estimate
@@
 California. USA) and measured leaflet area from digital images using the
-FIJI implementation of ImageJ2 version 2.3.0 (\emph{7}). We dried leaves
-for 72 hours at \(\qty{74}{\degreeCelsius}\) in a food dehydrator
+FIJI implementation of ImageJ2 version 2.3.0 (\emph{56}). We dried
+leaves for 72 hours at \(\qty{74}{\degreeCelsius}\) in a food dehydrator
 (Cosori CP267-FD, Vesync Co., Anaheim, California, USA) and weighed
 using a benchtop analytical balance (Ohaus PR64 Analytical Balance,
-Parsippany, New Jersey, USA). In \(5.0\%\) we measured LMA on the
+Parsippany, New Jersey, USA). In \(10.5\%\) we measured LMA on the
 adjacent leaflet because the focal leaflet was damaged or wilted while
 making surface impressions and we could not reliably estimate area. LMA
-data are missing from \(3.0\%\) of individuals because the area or mass
+data are missing from \(3.32\%\) of individuals because the area or mass
 was not recorded at all or recorded incorrectly.
@@
 \addlinespace
-6. Trim extreme \aax{} values & 2,210 & 28.0\\
+6. trim extreme \aax{} values & 2,214 & 28.1\\
 \bottomrule
@@
 extreme \aax{} values as those with an absolute internal studentized
-residual \(> 3.5\). Because these values likely indicate significant
+residual \(> 3\). Because these values likely indicate significant
 measurement error or leaf damage, we removed \agcurve{} curves at both
@@
 
-\emph{Procedure}: We fit provisional nonrectangular hyperbola (\emph{8})
-to each \aqcurve{} curve using nonlinear regression with the
+\emph{Procedure}: We fit provisional nonrectangular hyperbola
+(\emph{57}) to each \aqcurve{} curve using nonlinear regression with the
 \texttt{nlsLM()} function from the \emph{R} package \textbf{minpack.lm}
-version 1.2.4 (\emph{9}). We sequentially removed data points with an
+version 1.2.4 (\emph{58}). We sequentially removed data points with an
 absolute external studentized residual \(> 3\) until none remained.
@@
 We fit Bayesian models with MCMC sampling in the probabilistic
-programming language \emph{Stan} (\emph{10}). We used CmdStan version
-2.33.1 and \textbf{cmdstanr} version 0.7.1 (\emph{11}) to interface with
-\emph{R} version 4.3.2 (\emph{12}). We sampled the posterior
-distribution from 4 chains with 1000 iterations each after 1000 warmup
+programming language \emph{Stan} (\emph{59}) using the \emph{R} package
+\textbf{brms} version 2.22.0 (\emph{60}). We used CmdStan version 2.36.0
+and \textbf{cmdstanr} version 0.9.0 (\emph{61}) to interface with
+\emph{R} version 4.5.0 (\emph{62}). We sampled the posterior
+distribution from 1 chains with 1000 iterations each after 1000 warmup
 iterations per chain. We estimated parameters and confidence intervals
@@
 iterations, and maximum treedepth so that parameter estimates converged
-(\(\hat{R} < 1.01\) (\emph{13})) and the effective sample size (ESS) for
+(\(\hat{R} < 1.01\) (\emph{63})) and the effective sample size (ESS) for
 each parameter was \(> 10^3\).
@@
 phylogenetic structure among the random effects of accession using an
-Ornstein-Uhlenbeck (OU) process (\emph{14}). The expected covariance
+Ornstein-Uhlenbeck (OU) process (\emph{64}). The expected covariance
 between accessions \(i\) and \(j\) (\(\text{Cov}(i, j)\) ) is:
@@
 the fit of models (\autoref{tbl-models}) using the \emph{R} package
-\textbf{loo} version 2.7.0 (\emph{15}) to calculate LOOIC values. Note
+\textbf{loo} version 2.8.0 (\emph{65}) to calculate LOOIC values. Note
 that LOOIC was calculated only from pointwise likelihood values of the
@@
 Justin Alter, Max Gatlin, Joana Kim, Jenna Matsuyama, Brandon Najarian,
-Kai Yasuda
+Kai Yasuda. Sarah Friedrich helped with figures.
 
@@
 \begin{CSLReferences}{0}{1}
+\bibitem[\citeproctext]{ref-raven_selection_2002}
+\CSLLeftMargin{1. }%
+\CSLRightInline{J. A. Raven,
+\href{https://doi.org/10.1046/j.0028-646X.2001.00334.x}{Selection
+pressures on stomatal evolution}. \emph{New Phytologist} \textbf{153},
+371--386 (2002).}
+
+\bibitem[\citeproctext]{ref-mcadam_stomata_2021}
+\CSLLeftMargin{2. }%
+\CSLRightInline{S. A. M. McAdam, J. G. Duckett, F. C. Sussmilch, S.
+Pressel, K. S. Renzaglia, R. Hedrich, T. J. Brodribb, A. Merced,
+\href{https://doi.org/10.1002/ajb2.1619}{Stomata: The holey grail of
+plant evolution}. \emph{American Journal of Botany} \textbf{108},
+366--371 (2021).}
+
+\bibitem[\citeproctext]{ref-clark_origin_2022}
+\CSLLeftMargin{3. }%
+\CSLRightInline{J. W. Clark, B. J. Harris, A. J. Hetherington, N.
+Hurtado-Castano, R. A. Brench, S. Casson, T. A. Williams, J. E. Gray, A.
+M. Hetherington, \href{https://doi.org/10.1016/j.cub.2022.04.040}{The
+origin and evolution of stomata}. \emph{Current Biology} \textbf{32},
+R539--R553 (2022).}
+
+\bibitem[\citeproctext]{ref-hetherington_role_2003}
+\CSLLeftMargin{4. }%
+\CSLRightInline{A. M. Hetherington, F. I. Woodward,
+\href{https://doi.org/10.1038/nature01843}{The role of stomata in
+sensing and driving environmental change}. \emph{Nature} \textbf{424},
+901--908 (2003).}
+
+\bibitem[\citeproctext]{ref-de_boer_optimal_2016}
+\CSLLeftMargin{5. }%
+\CSLRightInline{H. J. de Boer, C. A. Price, F. Wagner‐Cremer, S. C.
+Dekker, P. J. Franks, E. J. Veneklaas,
+\href{https://doi.org/10.1111/nph.13929}{Optimal allocation of leaf
+epidermal area for gas exchange}. \emph{New Phytologist} \textbf{210},
+1219--1228 (2016).}
+
+\bibitem[\citeproctext]{ref-harrison_influence_2020}
+\CSLLeftMargin{6. }%
+\CSLRightInline{E. L. Harrison, L. Arce Cubas, J. E. Gray, C. Hepworth,
+\href{https://doi.org/10.1111/tpj.14560}{The influence of stomatal
+morphology and distribution on photosynthetic gas exchange}. \emph{The
+Plant Journal} \textbf{101}, 768--779 (2020).}
+
+\bibitem[\citeproctext]{ref-woodward_stomatal_1987}
+\CSLLeftMargin{7. }%
+\CSLRightInline{F. I. Woodward,
+\href{https://doi.org/10.1038/327617a0}{Stomatal numbers are sensitive
+to increases in {CO2} from pre-industrial levels}. \emph{Nature}
+\textbf{327}, 617--618 (1987).}
+
+\bibitem[\citeproctext]{ref-buckley_modelling_2013}
+\CSLLeftMargin{8. }%
+\CSLRightInline{T. N. Buckley, K. A. Mott,
+\href{https://doi.org/10.1111/pce.12140}{Modelling stomatal conductance
+in response to environmental factors: {Modelling} stomatal conductance}.
+\emph{Plant, Cell \& Environment} \textbf{36}, 1691--1699 (2013).}
+
+\bibitem[\citeproctext]{ref-haworth_co-ordination_2013}
+\CSLLeftMargin{9. }%
+\CSLRightInline{M. Haworth, C. Elliott-Kingston, J. C. McElwain,
+\href{https://doi.org/10.1007/s00442-012-2406-9}{Co-ordination of
+physiological and morphological responses of stomata to elevated
+{[}{CO2}{]} in vascular plants}. \emph{Oecologia} \textbf{171}, 71--82
+(2013).}
+
+\bibitem[\citeproctext]{ref-liang_stomatal_2023}
+\CSLLeftMargin{10. }%
+\CSLRightInline{X. Liang, D. Wang, Q. Ye, J. Zhang, M. Liu, H. Liu, K.
+Yu, Y. Wang, E. Hou, B. Zhong, L. Xu, T. Lv, S. Peng, H. Lu, P. Sicard,
+A. Anav, D. S. Ellsworth,
+\href{https://doi.org/10.1038/s41467-023-37934-7}{Stomatal responses of
+terrestrial plants to global change}. \emph{Nature Communications}
+\textbf{14}, 2188 (2023).}
+
+\bibitem[\citeproctext]{ref-chua_stomatal_2024}
+\CSLLeftMargin{11. }%
+\CSLRightInline{L. C. Chua, O. S. Lau,
+\href{https://doi.org/10.1242/dev.202681}{Stomatal development in the
+changing climate}. \emph{Development} \textbf{151}, dev202681 (2024).}
+
+\bibitem[\citeproctext]{ref-lang_century-long_2024}
+\CSLLeftMargin{12. }%
+\CSLRightInline{P. L. M. Lang, J. M. Erberich, L. Lopez, C. L. Weiß, G.
+Amador, H. F. Fung, S. M. Latorre, J. R. Lasky, H. A. Burbano, M.
+Expósito-Alonso, D. C. Bergmann,
+\href{https://doi.org/10.1038/s41559-024-02481-x}{Century-long timelines
+of herbarium genomes predict plant stomatal response to climate change}.
+\emph{Nature Ecology \& Evolution} \textbf{8}, 1641--1653 (2024).}
+
+\bibitem[\citeproctext]{ref-franks_new_2014}
+\CSLLeftMargin{13. }%
+\CSLRightInline{P. J. Franks, D. L. Royer, D. J. Beerling, P. K. Van de
+Water, D. J. Cantrill, M. M. Barbour, J. A. Berry,
+\href{https://doi.org/10.1002/2014GL060457}{New constraints on
+atmospheric {CO}\(_{\textrm{2}}\) concentration for the {Phanerozoic}}.
+\emph{Geophysical Research Letters} \textbf{41}, 4685--4694 (2014).}
+
+\bibitem[\citeproctext]{ref-mcelwain_paleoecology_2017}
+\CSLLeftMargin{14. }%
+\CSLRightInline{J. C. McElwain, M. Steinthorsdottir, Paleoecology,
+ploidy, paleoatmospheric composition, and developmental biology: A
+review of the multiple uses of fossil stomata. \emph{Plant Physiology}
+\textbf{174}, 650--664 (2017).}
+
+\bibitem[\citeproctext]{ref-the_cenozoic_co_proxy_integration_project_cencopip_consortium_toward_2023}
+\CSLLeftMargin{15. }%
+\CSLRightInline{The Cenozoic CO Proxy Integration Project (CenCOPIP)
+Consortium*†, B. Hönisch, D. L. Royer, D. O. Breecker, P. J. Polissar,
+G. J. Bowen, M. J. Henehan, Y. Cui, M. Steinthorsdottir, J. C. McElwain,
+M. J. Kohn, A. Pearson, S. R. Phelps, K. T. Uno, A. Ridgwell, E.
+Anagnostou, J. Austermann, M. P. S. Badger, R. S. Barclay, P. K. Bijl,
+T. B. Chalk, C. R. Scotese, E. De La Vega, R. M. DeConto, K. A. Dyez, V.
+Ferrini, P. J. Franks, C. F. Giulivi, M. Gutjahr, D. T. Harper, L. L.
+Haynes, M. Huber, K. E. Snell, B. A. Keisling, W. Konrad, T. K.
+Lowenstein, A. Malinverno, M. Guillermic, L. M. Mejía, J. N. Milligan,
+J. J. Morton, L. Nordt, R. Whiteford, A. Roth-Nebelsick, J. K. C.
+Rugenstein, M. F. Schaller, N. D. Sheldon, S. Sosdian, E. B. Wilkes, C.
+R. Witkowski, Y. G. Zhang, L. Anderson, D. J. Beerling, C. Bolton, T. E.
+Cerling, J. M. Cotton, J. Da, D. D. Ekart, G. L. Foster, D. R.
+Greenwood, E. G. Hyland, E. A. Jagniecki, J. P. Jasper, J. B. Kowalczyk,
+L. Kunzmann, W. M. Kürschner, C. E. Lawrence, C. H. Lear, M. A.
+Martínez-Botí, D. P. Maxbauer, P. Montagna, B. D. A. Naafs, J. W. B.
+Rae, M. Raitzsch, G. J. Retallack, S. J. Ring, O. Seki, J. Sepúlveda, A.
+Sinha, T. F. Tesfamichael, A. Tripati, J. Van Der Burgh, J. Yu, J. C.
+Zachos, L. Zhang, \href{https://doi.org/10.1126/science.adi5177}{Toward
+a {Cenozoic} history of atmospheric {CO}\(_{\textrm{2}}\)}.
+\emph{Science} \textbf{382}, eadi5177 (2023).}
+
+\bibitem[\citeproctext]{ref-hofmann_impact_2025}
+\CSLLeftMargin{16. }%
+\CSLRightInline{T. A. Hofmann, W. Atkinson, M. Fan, A. J. Simkin, P.
+Jindal, T. Lawson, \href{https://doi.org/10.1098/rstb.2024.0244}{Impact
+of climate-driven changes in temperature on stomatal anatomy and
+physiology}. \emph{Philosophical Transactions of the Royal Society B:
+Biological Sciences} \textbf{380}, 20240244 (2025).}
+
+\bibitem[\citeproctext]{ref-berry_stomata:_2010}
+\CSLLeftMargin{17. }%
+\CSLRightInline{J. A. Berry, D. J. Beerling, P. J. Franks,
+\href{https://doi.org/10.1016/j.pbi.2010.04.013}{Stomata: Key players in
+the earth system, past and present}. \emph{Current Opinion in Plant
+Biology} \textbf{13}, 232--239 (2010).}
+
+\bibitem[\citeproctext]{ref-franks_stomatal_2017}
+\CSLLeftMargin{18. }%
+\CSLRightInline{P. J. Franks, J. A. Berry, D. L. Lombardozzi, G. B.
+Bonan, \href{https://doi.org/10.1104/pp.17.00287}{Stomatal {Function}
+across {Temporal} and {Spatial} {Scales}: {Deep}-{Time} {Trends},
+{Land}-{Atmosphere} {Coupling} and {Global} {Models}}. \emph{Plant
+Physiology} \textbf{174}, 583--602 (2017).}
+
+\bibitem[\citeproctext]{ref-grubb_leaf_1977}
+\CSLLeftMargin{19. }%
+\CSLRightInline{P. J. Grubb, {``Leaf structure and function''} in
+\emph{The Encyclopedia of Ignorance}, R. Duncan, M. Weston-Smith, Eds.
+(Pergamon, Oxford, 1977)vol. 2, pp. 317--330.}
+
+\bibitem[\citeproctext]{ref-parkhurst_adaptive_1978}
+\CSLLeftMargin{20. }%
+\CSLRightInline{D. F. Parkhurst,
+\href{https://doi.org/10.2307/2259142}{The adaptive significance of
+stomatal occurrence on one or both surfaces of leaves}. \emph{The
+Journal of Ecology} \textbf{66}, 367--383 (1978).}
+
+\bibitem[\citeproctext]{ref-mott_adaptive_1982}
+\CSLLeftMargin{21. }%
+\CSLRightInline{K. A. Mott, A. C. Gibson, J. W. O'Leary,
+\href{https://doi.org/10.1111/1365-3040.ep11611750}{The adaptive
+significance of amphistomatic leaves}. \emph{Plant, Cell \& Environment}
+\textbf{5}, 455--460 (1982).}
+
+\bibitem[\citeproctext]{ref-gibson_structure-function_1996}
+\CSLLeftMargin{22. }%
+\CSLRightInline{A. C. Gibson, \emph{Structure-{Function} {Relations} of
+{Warm} {Desert} {Plants}} (Springer Berlin / Heidelberg, Berlin,
+Heidelberg, 1996;
+\url{http://public.eblib.com/choice/PublicFullRecord.aspx?p=6495247}).}
+
+\bibitem[\citeproctext]{ref-smith_leaf_1997}
+\CSLLeftMargin{23. }%
+\CSLRightInline{W. K. Smith, T. C. Vogelmann, E. H. DeLucia, D. T. Bell,
+K. A. Shepherd, \href{https://doi.org/10.2307/1313100}{Leaf {Form} and
+{Photosynthesis}}. \emph{BioScience} \textbf{47}, 785--793 (1997).}
+
+\bibitem[\citeproctext]{ref-oguchi_leaf_2018}
+\CSLLeftMargin{24. }%
+\CSLRightInline{R. Oguchi, Y. Onoda, I. Terashima, D. Tholen, {``Leaf
+{Anatomy} and {Function}''} in \emph{The {Leaf}: {A} {Platform} for
+{Performing} {Photosynthesis}}, W. W. Adams III, I. Terashima, Eds.
+(Springer International Publishing, Cham, 2018;
+\url{https://doi.org/10.1007/978-3-319-93594-2_5})\emph{Advances in
+{Photosynthesis} and {Respiration}}, pp. 97--139.}
+
+\bibitem[\citeproctext]{ref-drake_two_2019}
+\CSLLeftMargin{25. }%
+\CSLRightInline{P. L. Drake, H. J. de Boer, S. J. Schymanski, E. J.
+Veneklaas, \href{https://doi.org/10.1111/nph.15652}{Two sides to every
+leaf: Water and {CO}\(_{\textrm{2}}\) transport in hypostomatous and
+amphistomatous leaves}. \emph{New Phytologist} \textbf{222}, 1179--1187
+(2019).}
+
+\bibitem[\citeproctext]{ref-grubb_leaf_2020}
+\CSLLeftMargin{26. }%
+\CSLRightInline{P. J. Grubb, {``Leaf structure and function''} in
+\emph{Unsolved {Problems} in {Ecology}}, A. Dobson, D. Tilman, R. D.
+Holt, Eds. (Princeton University Press, Princeton, 2020), pp. 124--144.}
+
+\bibitem[\citeproctext]{ref-pospisilova_environmental_1980}
+\CSLLeftMargin{27. }%
+\CSLRightInline{J. Pospíŝilová, J. Solárová, Environmental and
+biological control of diffusive conductances of adaxial and abaxial leaf
+epidermes. \emph{Photosynthetica} \textbf{14}, 90--127 (1980).}
+
+\bibitem[\citeproctext]{ref-mott_stomatal_1984}
+\CSLLeftMargin{28. }%
+\CSLRightInline{K. A. Mott, J. W. O'Leary,
+\href{http://www.jstor.org/stable/4268406}{Stomatal {Behavior} and {CO}₂
+{Exchange} {Characteristics} in {Amphistomatous} {Leaves}}. \emph{Plant
+Physiology} \textbf{74}, 47--51 (1984).}
+
+\bibitem[\citeproctext]{ref-reich_effects_1985}
+\CSLLeftMargin{29. }%
+\CSLRightInline{P. B. Reich, A. W. Schoettle, R. G. Amundson,
+\href{https://doi.org/10.1111/j.1399-3054.1985.tb02818.x}{Effects of low
+concentrations of {O3}, leaf age and water stress on leaf diffusive
+conductance and water use efficiency in soybean}. \emph{Physiologia
+Plantarum} \textbf{63}, 58--64 (1985).}
+
+\bibitem[\citeproctext]{ref-mott_asymmetric_1993}
+\CSLLeftMargin{30. }%
+\CSLRightInline{K. A. Mott, Z. G. Cardon, J. A. Berry,
+\href{https://doi.org/10.1111/j.1365-3040.1993.tb00841.x}{Asymmetric
+patchy stomatal closure for the two surfaces of \emph{{Xanthium}
+strumarium} {L}. Leaves at low humidity}. \emph{Plant, Cell \&
+Environment} \textbf{16}, 25--34 (1993).}
+
+\bibitem[\citeproctext]{ref-wall_stomata_2022}
+\CSLLeftMargin{31. }%
+\CSLRightInline{S. Wall, S. Vialet‐Chabrand, P. Davey, J. Van Rie, A.
+Galle, J. Cockram, T. Lawson,
+\href{https://doi.org/10.1111/nph.18257}{Stomata on the abaxial and
+adaxial leaf surfaces contribute differently to leaf gas exchange and
+photosynthesis in wheat}. \emph{New Phytologist} \textbf{235},
+1743--1756 (2022).}
+
+\bibitem[\citeproctext]{ref-gutschick_photosynthesis_1984}
+\CSLLeftMargin{32. }%
+\CSLRightInline{V. P. Gutschick, Photosynthesis model for
+{C}\(_{\textrm{3}}\) leaves incorporating {CO}\(_{\textrm{2}}\)
+transport, propagation of radiation, and biochemistry 2. Ecological and
+agricultural utility. \emph{Photosynthetica} \textbf{18}, 569--595
+(1984).}
+
+\bibitem[\citeproctext]{ref-marquez_assessing_2023}
+\CSLLeftMargin{33. }%
+\CSLRightInline{D. A. Márquez, H. Stuart‐Williams, L. A. Cernusak, G. D.
+Farquhar, \href{https://doi.org/10.1111/nph.18784}{Assessing the
+{\textless{}}span style="font-variant:small-caps;"{\textgreater{}}
+{CO}\(_{\textrm{2}}\) {\textless{}}/span{\textgreater{}} concentration
+at the surface of photosynthetic mesophyll cells}. \emph{New
+Phytologist} \textbf{238}, 1446--1460 (2023).}
+
+\bibitem[\citeproctext]{ref-parkhurst_intercellular_1990}
+\CSLLeftMargin{34. }%
+\CSLRightInline{D. F. Parkhurst, K. A. Mott,
+\href{https://doi.org/10.1104/pp.94.3.1024}{Intercellular diffusion
+limits to {CO}\(_{\textrm{2}}\) uptake in leaves: Studies in air and
+helox}. \emph{Plant Physiology} \textbf{94}, 1024--1032 (1990).}
+
+\bibitem[\citeproctext]{ref-foster_influence_1986}
+\CSLLeftMargin{35. }%
+\CSLRightInline{J. R. Foster, W. K. Smith,
+\href{https://doi.org/10.1111/j.1365-3040.1986.tb02108.x}{Influence of
+stomatal distribution on transpiration in low-wind environments}.
+\emph{Plant, Cell and Environment} \textbf{9}, 751--759 (1986).}
+
+\bibitem[\citeproctext]{ref-muir_is_2019}
+\CSLLeftMargin{36. }%
+\CSLRightInline{C. D. Muir, \href{https://doi.org/10.1093/icb/icz085}{Is
+amphistomy an adaptation to high light? {Optimality} models of stomatal
+traits along light gradients}. \emph{Integrative and Comparative
+Biology} \textbf{59}, 571--584 (2019).}
+
+\bibitem[\citeproctext]{ref-wood_physiology_1934}
+\CSLLeftMargin{37. }%
+\CSLRightInline{J. G. Wood, The physiology of xerophytism in
+{Australian} plants: The stomatal frequencies, transpiration and osmotic
+pressures of sclerophyll and tomentose-succulent leaved plants.
+\emph{Journal of Ecology} \textbf{22}, 69--87 (1934).}
+
+\bibitem[\citeproctext]{ref-howell_concerning_1945}
+\CSLLeftMargin{38. }%
+\CSLRightInline{J. T. Howell, Concerning stomata on leaves in
+\emph{a}rctostaphylos. \emph{The Wasmann Collector} \textbf{6}, 57--65
+(1945).}
+
+\bibitem[\citeproctext]{ref-salisbury_i_1928}
+\CSLLeftMargin{39. }%
+\CSLRightInline{E. J. Salisbury,
+\href{https://doi.org/10.1098/rstb.1928.0001}{I. {On} the causes and
+ecological significance of stomatal frequency, with special reference to
+the woodland flora}. \emph{Philosophical Transactions of the Royal
+Society of London. Series B, Containing Papers of a Biological
+Character} \textbf{216}, 1--65 (1928).}
+
+\bibitem[\citeproctext]{ref-peat_comparative_1994}
+\CSLLeftMargin{40. }%
+\CSLRightInline{H. J. Peat, A. H. Fitter, A comparative study of the
+distribution and density of stomata in the {British} flora.
+\emph{Biological Journal of the Linnean Society} \textbf{52}, 377--393
+(1994).}
+
+\bibitem[\citeproctext]{ref-jordan_using_2014}
+\CSLLeftMargin{41. }%
+\CSLRightInline{G. J. Jordan, R. J. Carpenter, T. J. Brodribb,
+\href{https://doi.org/10.1016/j.palaeo.2013.12.035}{Using fossil leaves
+as evidence for open vegetation}. \emph{Palaeogeography,
+Palaeoclimatology, Palaeoecology} \textbf{395}, 168--175 (2014).}
+
+\bibitem[\citeproctext]{ref-bucher_stomatal_2017}
+\CSLLeftMargin{42. }%
+\CSLRightInline{S. F. Bucher, K. Auerswald, C. Grün-Wenzel, S. I.
+Higgins, J. Garcia Jorge, C. Römermann,
+\href{https://doi.org/10.1016/j.flora.2017.02.011}{Stomatal traits
+relate to habitat preferences of herbaceous species in a temperate
+climate}. \emph{Flora} \textbf{229}, 107--115 (2017).}
+
+\bibitem[\citeproctext]{ref-muir_light_2018}
+\CSLLeftMargin{43. }%
+\CSLRightInline{C. D. Muir, Light and growth form interact to shape
+stomatal ratio among {British} angiosperms. \emph{New Phytologist}
+\textbf{218}, 242--252 (2018).}
+
+\bibitem[\citeproctext]{ref-triplett_stomatal_2025}
+\CSLLeftMargin{44. }%
+\CSLRightInline{G. Triplett, A. S. David,
+\href{https://doi.org/10.1002/ajb2.70050}{Stomatal distribution and
+post‐fire recovery: {Intra}‐ and interspecific variation in plants of
+the pyrogenic {Florida} scrub}. \emph{American Journal of Botany},
+e70050 (2025).}
+
+\bibitem[\citeproctext]{ref-lyshede_comparative_2002}
+\CSLLeftMargin{45. }%
+\CSLRightInline{O. B. Lyshede, Comparative and functional leaf anatomy
+of selected {Alstroemeriaceae} of mainly {Chilean} origin.
+\emph{Botanical Journal of the Linnean Society} \textbf{140}, 261--272
+(2002).}
+
 \bibitem[\citeproctext]{ref-triplett_amphistomy_2024}
-\CSLLeftMargin{1. }%
+\CSLLeftMargin{46. }%
 \CSLRightInline{G. Triplett, T. N. Buckley, C. D. Muir,
@@
 photosynthesis more in coastal than montane plants of {Hawaiian} ʻilima
-( \emph{{Sida} fallax} )}. \emph{American Journal of Botany}
-\textbf{111}, e16284 (2024).}
+(\emph{{Sida} fallax})}. \emph{American Journal of Botany} \textbf{111},
+e16284 (2024).}
+
+\bibitem[\citeproctext]{ref-farquhar_biochemical_1980}
+\CSLLeftMargin{47. }%
+\CSLRightInline{G. D. Farquhar, S. von Caemmerer, J. A. Berry,
+\href{https://doi.org/10.1007/BF00386231}{A biochemical model of
+photosynthetic {CO}\(_{\textrm{2}}\) assimilation in leaves of
+{C}\(_{\textrm{3}}\) species}. \emph{Planta} \textbf{149}, 78--90
+(1980).}
+
+\bibitem[\citeproctext]{ref-poorter_metaanalysis_2019}
+\CSLLeftMargin{48. }%
+\CSLRightInline{H. Poorter, Ü. Niinemets, N. Ntagkas, A. Siebenkäs, M.
+Mäenpää, S. Matsubara, T. L. Pons,
+\href{https://doi.org/10.1111/nph.15754}{A meta‐analysis of plant
+responses to light intensity for 70 traits ranging from molecules to
+whole plant performance}. \emph{New Phytologist} \textbf{223},
+1073--1105 (2019).}
+
+\bibitem[\citeproctext]{ref-givnish_common-garden_2014}
+\CSLLeftMargin{49. }%
+\CSLRightInline{T. J. Givnish, R. A. Montgomery,
+\href{https://doi.org/10.1098/rspb.2013.2944}{Common-garden studies on
+adaptive radiation of photosynthetic physiology among {Hawaiian}
+lobeliads}. \emph{Proceedings of the Royal Society B: Biological
+Sciences} \textbf{281}, 20132944--20132944 (2014).}
+
+\bibitem[\citeproctext]{ref-engelbrecht_evaluation_2001}
+\CSLLeftMargin{50. }%
+\CSLRightInline{B. M. J. Engelbrecht, H. M. Herz,
+\href{https://doi.org/10.1017/S0266467401001146}{Evaluation of different
+methods to estimate understorey light conditions in tropical forests}.
+\emph{Journal of Tropical Ecology} \textbf{17}, 207--224 (2001).}
 
 \bibitem[\citeproctext]{ref-peralta_taxonomy_2008}
-\CSLLeftMargin{2. }%
+\CSLLeftMargin{51. }%
 \CSLRightInline{I. E. Peralta, D. M. Spooner, S. Knapp, Taxonomy of wild
@@
 
+\bibitem[\citeproctext]{ref-burns_multi-resolution_2024}
+\CSLLeftMargin{52. }%
+\CSLRightInline{P. Burns, C. R. Hakkenberg, S. J. Goetz,
+\href{https://doi.org/10.1038/s41597-024-03668-4}{Multi-resolution
+gridded maps of vegetation structure from {GEDI}}. \emph{Scientific
+Data} \textbf{11}, 881 (2024).}
+
 \bibitem[\citeproctext]{ref-schoch_dependence_1980}
-\CSLLeftMargin{3. }%
+\CSLLeftMargin{53. }%
 \CSLRightInline{P.-G. Schoch, C. Zinsou, M. Sibi,
@@
 
-\bibitem[\citeproctext]{ref-parkhurst_adaptive_1978}
-\CSLLeftMargin{4. }%
-\CSLRightInline{D. F. Parkhurst,
-\href{https://doi.org/10.2307/2259142}{The adaptive significance of
-stomatal occurrence on one or both surfaces of leaves}. \emph{The
-Journal of Ecology} \textbf{66}, 367--383 (1978).}
-
 \bibitem[\citeproctext]{ref-sack_developmental_2016}
-\CSLLeftMargin{5. }%
+\CSLLeftMargin{54. }%
 \CSLRightInline{L. Sack, T. N. Buckley,
@@
 \bibitem[\citeproctext]{ref-mott_amphistomy_1991}
-\CSLLeftMargin{6. }%
+\CSLLeftMargin{55. }%
 \CSLRightInline{K. A. Mott, O. Michaelson, Amphistomy as an adaptation
@@
 \bibitem[\citeproctext]{ref-schindelin_fiji_2012}
-\CSLLeftMargin{7. }%
+\CSLLeftMargin{56. }%
 \CSLRightInline{J. Schindelin, I. Arganda-Carreras, E. Frise, V. Kaynig,
@@
 \bibitem[\citeproctext]{ref-marshall_model_1980}
-\CSLLeftMargin{8. }%
+\CSLLeftMargin{57. }%
 \CSLRightInline{B. Marshall, P. V. Biscoe,
@@
 \bibitem[\citeproctext]{ref-elzhov_minpacklm_2023}
-\CSLLeftMargin{9. }%
+\CSLLeftMargin{58. }%
 \CSLRightInline{T. V. Elzhov, K. M. Mullen, A.-N. Spiess, B. M. Bolker,
@@
 
-\bibitem[\citeproctext]{ref-stan_development_team_stan_2023}
-\CSLLeftMargin{10. }%
+\bibitem[\citeproctext]{ref-stan_development_team_stan_2025}
+\CSLLeftMargin{59. }%
 \CSLRightInline{Stan Development Team, \emph{Stan {Modeling} {Language}
-{Users} {Guide} and {Reference} {Manual}} (2023;
+{Users} {Guide} and {Reference} {Manual}} (2025;
 \url{https://mc-stan.org}).}
 
-\bibitem[\citeproctext]{ref-gabry_cmdstanr_2023}
-\CSLLeftMargin{11. }%
-\CSLRightInline{J. Gabry, R. Češnovar, \emph{Cmdstanr: {R} {Interface}
-to '{CmdStan}'} (2023;
+\bibitem[\citeproctext]{ref-burkner_brms_2017}
+\CSLLeftMargin{60. }%
+\CSLRightInline{P.-C. Bürkner,
+\href{https://doi.org/10.18637/jss.v080.i01}{\textbf{Brms} : {An}
+\emph{r} {Package} for {Bayesian} {Multilevel} {Models} {Using}
+\emph{stan}}. \emph{Journal of Statistical Software} \textbf{80}
+(2017).}
+
+\bibitem[\citeproctext]{ref-gabry_cmdstanr_2025}
+\CSLLeftMargin{61. }%
+\CSLRightInline{J. Gabry, R. Češnovar, A. Johnson, S. Bronder,
+\emph{Cmdstanr: {R} {Interface} to '{CmdStan}'} (2025;
 \href{https://mc-stan.org/cmdstanr,\%20https://discourse.mc-stan.org}{https://mc-stan.org/cmdstanr,
@@
 
-\bibitem[\citeproctext]{ref-r_core_team_r:_2023}
-\CSLLeftMargin{12. }%
+\bibitem[\citeproctext]{ref-r_core_team_r:_2025}
+\CSLLeftMargin{62. }%
 \CSLRightInline{R Core Team, \emph{R: {A} {Language} and {Environment}
 for {Statistical} {Computing}} (R Foundation for Statistical Computing,
-Vienna, Austria, 2023; \url{http://www.R-project.org/}).}
+Vienna, Austria, 2025; \url{http://www.R-project.org/}).}
 
 \bibitem[\citeproctext]{ref-gelman_inference_1992}
-\CSLLeftMargin{13. }%
+\CSLLeftMargin{63. }%
 \CSLRightInline{A. Gelman, D. B. Rubin, Inference from iterative
@@
 \bibitem[\citeproctext]{ref-hansen_stabilizing_1997}
-\CSLLeftMargin{14. }%
+\CSLLeftMargin{64. }%
 \CSLRightInline{T. F. Hansen,
@@
 \bibitem[\citeproctext]{ref-vehtari_practical_2017}
-\CSLLeftMargin{15. }%
+\CSLLeftMargin{65. }%
 \CSLRightInline{A. Vehtari, A. Gelman, J. Gabry,

### 2025-06-24 r/000_update-data.R

@@ -3,6 +3,7 @@ source("r/header.R")
 
 c("processed-data/accession-climate",
   "processed-data/accession-gedi",
+  "processed-data/accession-info",
   "processed-data/plant-info",
   "filtered-data/rh_curves",
   "filtered-data/rh_hi_curves",

### 2025-06-24 r/001_thin-data.R

@@ -26,8 +26,8 @@ do.call(
       .by = c("acc_id", "curve_type", "light_intensity")
     ) |>
     dplyr::summarize(
-      n_rh_curve6 = n(),
-      n_point_per_rh_curve6 = mean(n_point)
+      n_rh_curve5 = n(),
+      n_point_per_rh_curve5 = mean(n_point)
     ) |>
     as.list()
 )

### 2025-06-24 r/003_trim-outliers.R

@@ -47,8 +47,8 @@ do.call(
       .by = c("acc_id", "curve_type", "light_intensity")
     ) |>
     dplyr::summarize(
-      n_rh_curve7 = n(),
-      n_point_per_rh_curve7 = mean(n_point)
+      n_rh_curve6 = n(),
+      n_point_per_rh_curve6 = mean(n_point)
     ) |>
     as.list()
 )

### 2025-06-25 ms/ms.qmd
working on ms results

@@
 
-#### MARKER
 <!-- Paragraph 6 (summarize main results) -->
@@
 There was a slight tendency for accessions from more closed habitats (greater PAI), but AA varied widely among accessions from open habitats (low PAI) regardless of growth and measurement light intensities (Figure X).
+
+<!-- Paragraph 7 (summarize secondary results) -->
 The pattern of AA across wild tomatoes strongly supports the plasticity hypothesis, argues against the acclimatory hypothesis, and provides only weak support for the constitutive hypothesis.
-
-Not sure where this goes: We infer this from the fact that blocking gas exchange in pseudohypostomatous leaves reduced $A$ by X-X% depending on the accession, light treatment, and light intensity (Table/figure).
-The AA is equivalent to an X-X% change in total $g_\text{sc}$ (see SI section gs equivalency).
-But whereas increasing $g_\text{sc}$ would increase water loss as a necessary by-product, amphistomy can increase $A$ without any appreciable affect on transpiration.
-
-<!-- Paragraph 7 (summarize secondary results) -->
+Plastic changes in leaf thickness and/or packing density, summarized by the bulk leaf mass per area (LMA), likely mediate the effect of growth light intensity on AA.
+When LMA is included in the model, the effect of growth light intensity on AA is no longer significant based on model comparison (Table ?
+- need to check out what the interaction with measurement light intensity means).
+In contrast, LMA did not mediate indirect effects of native PAI on AA (Table ?).
+Many anatomical traits underlie LMA [@john_anatomical_2017] and future research will be needed identify which particular traits, such as leaf thickness or mesophyll porosity, are responsible for mediating AA.
+
+#### MARKER
 These results suggest that developmental plasticity, as opposed to acclimation or adaptation to open habitats, may explain the explain the long-standing observation taht amphistomatous leaves are more common in sunny habitats.
@@
 However, these traits were not associated with change in AA (table/figures).
+
+Not sure where this goes: We infer this from the fact that blocking gas exchange in pseudohypostomatous leaves reduced $A$ by X-X% depending on the accession, light treatment, and light intensity (Table/figure).
+The AA is equivalent to an X-X% change in total $g_\text{sc}$ (see SI section gs equivalency).
+But whereas increasing $g_\text{sc}$ would increase water loss as a necessary by-product, amphistomy can increase $A$ without any appreciable affect on transpiration.
 

### 2025-06-25 ms/ms.qmd
lma-aa figure

@@
 Plastic changes in leaf thickness and/or packing density, summarized by the bulk leaf mass per area (LMA), likely mediate the effect of growth light intensity on AA.
+LMA increased in sun grown plants in all accessions by an average of X% [CI], similar to plastic responses in many species [@poorter_metaanalysis_2019].
 When LMA is included in the model, the effect of growth light intensity on AA is no longer significant based on model comparison (Table ?
-- need to check out what the interaction with measurement light intensity means).
+- need to check out what the interaction with measurement light intensity means) and LMA is weakly positively correlated with AA (Figure X).
 In contrast, LMA did not mediate indirect effects of native PAI on AA (Table ?).

### 2025-06-25 r/209_plot-lma.R

@@ -5,11 +5,13 @@ fit_aa = read_rds("objects/fit_aa2.rds")
 
 # Estimated llma for each acc, light_intensity, and light_treatment
 df_new1 = crossing(
-    acc = unique(fit_aa$data$acc),
+    # acc = unique(fit_aa$data$acc),
+    acc_id = unique(fit_aa$data$acc_id),
     light_intensity = unique(fit_aa$data$light_intensity),
     light_treatment = unique(fit_aa$data$light_treatment)
   ) |>
-  mutate(row = row_number())
+  mutate(acc = str_extract(acc_id, acc_string),
+         row = row_number())
 
 df_pred1 = posterior_epred(fit_aa, newdata = df_new1, resp = "llma") |>
   as_draws_df() |>
@@ -64,21 +66,92 @@ df1 = full_join(
   )
 
 # Regression lines
-fit_aa$fit$draws
-
-as_draws_df(fit_aa) %>%
-  colnames()
-  dplyr::select("b_aa_Intercept", "b_aa_light_intensity2000", "bsp_aa_millma",
-                "bsp_aa_millma:light_intensity2000")
+ce = conditional_effects(fit_aa)
+df2 = ce[["aa.aa_llma:light_intensity"]] |>
+  mutate(lma = exp(llma), aa = estimate__, .lower = lower__, .upper = upper__,
+         Measurement = light_intensity |>
+           factor(levels = c("150", "2000")) |>
+           fct_recode(low = "150", high = "2000"), Growth = NA)
   
 # Plot
-ggplot(df1, aes(x = lma, y = aa, color = Growth, shape = Measurement)) +
+ggplot(df1, aes(x = lma, y = aa, shape = Measurement)) +
+  geom_ribbon(data = df2, mapping = aes(ymin = .lower, ymax = .upper, group = Measurement), alpha = 0.5, color = NA, fill = "grey") + 
+  geom_line(data = df2, aes(linetype = Measurement), linewidth = 1.1) +
   # geom_interval(aes(ymin = aa_lower, ymax = aa_upper), linewidth = 0.5) +
   # geom_interval(aes(xmin = lma_lower, xmax = lma_upper), linewidth = 0.5) +
-  geom_point(fill = "white") +
+  geom_point(mapping = aes(color = Growth), fill = "white", size = 2) +
   scale_color_manual(values = c("shade" = "tomato4", "sun" = "tomato")) +
[…]

### 2025-06-28 ms/ms.qmd

@@
 Many anatomical traits underlie LMA [@john_anatomical_2017] and future research will be needed identify which particular traits, such as leaf thickness or mesophyll porosity, are responsible for mediating AA.
+Our study is also limited by the fact that we could not directly measure the stomatal conductance ratio and our method for estimating AA requires large sample sizes to overcome several sources of measurement error (see supporting information).
+Future experiments measuring AA with a dual sided chamber [@wall; @marquez] can overcome these limitations.
 
@@
 
-Not sure where this goes: We infer this from the fact that blocking gas exchange in pseudohypostomatous leaves reduced $A$ by X-X% depending on the accession, light treatment, and light intensity (Table/figure).
-The AA is equivalent to an X-X% change in total $g_\text{sc}$ (see SI section gs equivalency).
-But whereas increasing $g_\text{sc}$ would increase water loss as a necessary by-product, amphistomy can increase $A$ without any appreciable affect on transpiration.
-
-from Mott and Oleary: "However, Jones and Slatyer (5) reported a higher mesophyll resistance for CO, entering through the upper stomata than for the lower, and the data of Vaclavik (12) appear to support this conclusion"
+big implications - if trees were amphi, more more carbon sink or reduced transpiration
 
@@
 
+### MANUSCRAPS
+Problems to bring up in discussion:
+- all our species were amphi, so they may all live in PAI below threshold to favor amphi
+- other microclimate diffs not considered
+- don't know specific anatomical change
+
 Two outstanding areas for further research are 1) the specific anatomical changes, such as leaf thickness or mesophyll porosity, that increase the diffusion path length for CO$_2$ in sun leaves, and 2) the costs of amphistomy that could explain plasticity in \gmaxratio{} in accessions from low-light habitats.
@@
 
-Problems to bring up in discussion:
-- all our species were amphi, so they may all live in PAI below threshold to favor amphi
-- other microclimate diffs not considered
-- don't know specific anatomical change
+Not sure where this goes: We infer this from the fact that blocking gas exchange in pseudohypostomatous leaves reduced $A$ by X-X% depending on the accession, light treatment, and light intensity (Table/figure).
+The AA is equivalent to an X-X% change in total $g_\text{sc}$ (see SI section gs equivalency).
+But whereas increasing $g_\text{sc}$ would increase water loss as a necessary by-product, amphistomy can increase $A$ without any appreciable affect on transpiration.
+
+from Mott and Oleary: "However, Jones and Slatyer (5) reported a higher mesophyll resistance for CO, entering through the upper stomata than for the lower, and the data of Vaclavik (12) appear to support this conclusion"
 
@@
 We estimated parameters and confidence intervals as the median and 95% quantile intervals of the posterior, respectively.
-We chose the number of chains, warmup and sampling iterations, and maximum treedepth so that parameter estimates converged ($\hat{R} < 1.01$ [@gelman_inference_1992]) and the effective sample size (ESS) for each parameter was $> 10^3$.
+We chose the number of chains, warmup and sampling iterations, and maximum treedepth so that parameter estimates converged ($\hat{R} < `r aa_args['max_rhat']`$ [@gelman_inference_1992]) and the effective sample size (ESS) for each parameter was $> `r aa_args['min_ess']`$.
 

### 2025-07-03 ms/ms.qmd
revising conclusions

@@
 
+Our results indicate that developmental plasticity, as opposed to acclimation or adaptation to open habitats, may explain the long-standing observation that amphistomatous leaves are more common in sunny habitats.
+If correct, this implies that high light intensity *per se* does not increase the benefit of amphistomy, but rather than anatomical and/or biochemical changes associated with higher light intensity modulate AA.
+We cannot yet attribute specific changes wrought by plasticity, but the correlation with laminar LMA suggests that changes in leaf thickness and/or density mediate the relationship between plasticity and AA.
+While we found little evidence for native light habitat in driving AA, unexplained constitutuve differences among accessions may be byproducts of adaptive evolutionary divergence.
+Hence, we cannot yet rule out an important role for environmental adaptation in modulating AA and this will require future research to resolve.
+
 #### MARKER
-These results suggest that developmental plasticity, as opposed to acclimation or adaptation to open habitats, may explain the explain the long-standing observation taht amphistomatous leaves are more common in sunny habitats.
-Our results add to existing explanations by showing high light intensity *per se* does not increase the benefit of amphistomy, but rather than anatomical and biochemical change associated with higher light intensity modulate AA.
-We cannot yet attribute specific changes wrought by plasticity, but correlational evidence suggests that changes in leaf anatomy may be important.
-Species responded consistently by increasing stomatal density, especially on the adaxial surface, increasing stomatal density ratio.
-Sun leaves also had greater gs, Amax.
-However, these traits were not associated with change in AA (table/figures).
-
-big implications - if trees were amphi, more more carbon sink or reduced transpiration
+Amphistomy likely plays a significant, but under appreciated, role in the carbon and water balance of plant groups where this trait is common, such as herbaceous plants in sunny habitats and woody species in arid biomes (cites).
+If our estimates of AA from the wild tomato clades
+big implications 
+- amphistomy plays role in carbon/water balance. do equivalent gs
+- if trees were amphi, more more carbon sink or reduced transpiration
+- comparative studies should consider plasticity
 

### 2025-07-04 ms/ms.qmd
first draft

--- 2025-06-25 ms/ms.qmd
@@
 ---
-title: Unknown
+title: Developmental plasticity to high light intensity amplifies the advantage of amphistomous leaves
 format:
@@
     keep-tex: true  
-    toc: true
-number-sections: true
+    toc: false
+author:
+  - name: "Christopher D. Muir$^{1,2*}$"
+    orcid: 0000-0003-2555-3878
+  - name: "Wei Shen Lim$^{1}$"
+  - name: "Dachuan Wang$^{1,2}$"
+affiliations: false
+number-sections: false
 header-includes: |
@@
   \usepackage{siunitx}
+  \newcommand{\aahigh}{\mathrm{AA}_{2000}}
+  \newcommand{\aalow}{\mathrm{AA}_{150}}
+  \newcommand{\aashade}{\mathrm{AA}_{\text{shade}}}
+  \newcommand{\aasun}{\mathrm{AA}_{\text{sun}}}
   \newcommand{\aax}{$\mathrm{AA}$}
+  \newcommand{\aay}{\mathrm{AA}}
   \newcommand{\Aamphi}{$A_{\mathrm{amphi}}$}
@@
   \newcommand{\gmaxratio}{$g_\text{max,ratio}$}
+  \newcommand{\gmaxshade}{$g_\text{max,shade}$}
+  \newcommand{\gmaxsun}{$g_\text{max,sun}$}
   \newcommand{\gsw}{$g_\text{sw}$}
@@
 
+```{=latex}
+\begin{center}
+$^1$School of Life Sciences, University of Hawaiʻi, Mānoa, Hawaiʻi, USA\\
+$^2$Department of Botany, University of Wisconsin, Madison, Wisconsin, USA\\
+\medskip
+*Corresponding author: cdmuir@wisc.edu\\
+\medskip
+ORCID: Christopher D. Muir — \href{https://orcid.org/0000-0003-2555-3878}{0000-0003-2555-3878}\\
+\end{center}
+```
+
 ```{r load-objects, echo=FALSE, eval=TRUE, message=FALSE, warning=FALSE}
+library(brms)
+library(cowplot)
 library(dplyr)
+library(ggplot2)
+library(kableExtra)
+library(knitr)
 library(lubridate)
@@
 
+theme_set(theme_cowplot())
+
 source("../r/functions.R")
@@
 
+accession_info = read_rds("../data/accession-info.rds") |>
+  dplyr::select(species, accession)
+
 rh_curves = read_rds("../data/trimmed_rh_curves.rds")
@@
 
-fit_aa = read_rds("../objects/fit_aa1.rds")
-dim_fit_aa = dim(fit_aa$fit)
+fit_aa1 = read_rds("../objects/fit_aa1.rds")
+fit_aa2 = read_rds("../objects/fit_aa2.rds")
+dim_fit_aa = dim(fit_aa1$fit)
 
@@
 n_acc = length(unique(rh_curves$acc))
+n_sp = accession_info |>
+  filter(accession %in% unique(rh_curves$acc)) |>
+  pull(species) |>
+  unique() |>
+  length()
+
 # average number of individual replicates for accession per light treatment
@@
   mutate(x = sprintf(
-    "%.3f (range: %.3f–%.3f; %.0f of %.0f accessions significant)",
+    "%.3f (range: %.3f–%.3f; %.0f of %.0f populations significant)",
     aa,
@@
 
-```
-
-## Notes
-
-Terminology to standardize:
-
--   the native light habitat or environment: the SPLASH paper refers to what they calculated as 'habitat' PPFD, so I am going to use that terminology
-
--   growth condition versus measurement conditions
-
--   ?
-
-## Authors
-
-definite: Wei Shen Lim, Dachuan Wang, Chris Muir
-
-question: Sam McKlin, Tom Buckley
-
-I am using this approach to drafting the paper:
-
-https://blogs.nature.com/nyc/2011/08/10/how-to-write-a-paper-one-possible-answer
-
-## Punchline
-
-Plastic changes in the structure of sun leaves can increase the intercellular CO$_2$ diffusion path length.
-Gas exchange through stomata on both upper and lower leaf surfaces, amphistomy, may be a compensatory mechanism to reduce the impact of increased diffusion path length on photosynthesis without increasing water loss.
-We demonstrate for the first time that sun leaves benefit most from amphistomy because it compensates for plastic structural changes in leaf anatomy.
-
-## Figures
-
-1. competing hypotheses and predictions (or predictions after methods?) for the benefit of amphistomy
-2. design of experiment to test competing hypotheses (with actual #'s of replicate, curves, etc.)
-3. Results on acclimatory, plastic, constitutive hypotheses
-4. Additional supporting figure on change in LMA, reduction of Amass, and relationship between LMA and AA? Modeling?
+# Effect of light treatment on LMA
+post_lma = posterior_summary(fit_aa2)["b_llma_light_treatmenthigh", ]
+
+```
 
@@
 
-(use for GRC)
-
-Developmental plasticity to light intensity modulates amphistomy advantage
-
 The presence of stomata on both leaf surfaces (amphistomy) can increase photosynthesis in C$_3$ plants by reducing the path length for CO$_2$ diffusion between substomatal cavities and chloroplasts.
-Amphistomatous leaves are most common among herbaceous plants growing in sunny habitats, including many crop species.
-The distribution of amphistomatous leaves in nature may result from an increased photosynthetic benefit of amphistomy under high light intensity, either because of acclimatory, plastic, or constitutive variation in CO$_2$ supply or demand.
-We used a recently developed method to quantify amphistomy advantage, the photosynthetic rate of an amphistomatous leaf relative to an otherwise identical hypostomatous leaf, in 29 diverse populations representing 15 species of wild tomatoes (*Solanum* sect.
+Amphistomatous leaves are common among herbaceous plants growing in sunny habitats, including many crop species.
+This distribution of amphistomatous leaves in nature may result from an increased photosynthetic benefit of amphistomy under high light intensity, either because of acclimatory, plastic, or constitutive variation in CO$_2$ supply or demand.
+We used a recently developed method to quantify amphistomy advantage, the photosynthetic rate of an amphistomatous leaf relative to an otherwise identical hypostomatous leaf, in `r n_acc` diverse populations representing `r n_sp` species of wild tomatoes (*Solanum* sect.
 *Lycopersicon* and sect.
 *Lycopersicoides*).
-Plant grown under high light intensity benefit more from amphostomy than those grown under low light, regardless of acclimation or native light intensity.
-Curvature of light response curves indicates that high-light leaves benefit more from amphistomy because of greater resistance to CO$_2$ diffusion within the mesophyll.
-Developmental plasticity in leaf thickness and/or cell arrangement to optimize light interception likely increases under varying light intensity modulates CO$_2$ diffusion within the leaf and, hence, amphistomy advantage as a byproduct.
+Plant grown under high light intensity benefit more in terms of CO$_2$ assimilated for a given stomatal conductance than plants grown under low light, regardless of measurement light intensity or native light habitat.
+Developmental plasticity in leaf thickness and/or density, an adaptive response to optimize light interception, likely affects CO$_2$ diffusion within the leaf, altering amphistomy advantage as a byproduct.
 Developmental plasticity may therefore play an important and previously unnoticed role in explaining the adaptive significance of amphistomy and the distribution of amphistomatous leaves in nature.
 
-What's surprising: role of developmental plasticity in explaining comparative results; diffusional limitations through palisade ias is strong and needs to be considered with light gradients in understanding leaf strucuture- funciton
-
-## Intro
+## Main text
 
@@
 The overarching hypothesis is that leaves with greater stomatal density ratio are more common in open, sunny habitats because they increase photosynthesis most in those circumstances.
-An amphistomatous leaf increases photosynthetic carbon gain compared to an otherwise identical hypostomatous leaf by increasing conductance through the leaf intercellular airspaces and boudary layers; the additional water loss through a second boundary layer is typically small [@foster_influence_1986].
-We quantify this benefit as the amphistomy advantage (AA = log(Aamphi/Ahypo) [@parkhurst_adaptive_1978; @triplett_amphistomy_2024].
-Why would AA be greater in sun than shade?
-We consider three nonmutually exclusive hypotheses that we classify as 'acclimatory', 'plastic', and 'constitutive'.
-
-Acclimatory hypothesis: Photosynthetic induction to high light intensity typically involves increases in total leaf stomatal conductance (increased CO$_2$ supply), the concentration of active Rubisco, and electron transport capacity (increased CO$_2$ demand).
-A one-dimensional circuit model using the Farquhar-von Caemmerer-Berry biochemical model of C$_3$ photosynthesis [@farquhar_biochemical_1980] shows that both increased stomatal conductance and Rubisco activity should increase AA, all else being equal (Supporting Information).
-If the acclimatory hypothesis is correct, we predict that AAhigh \> AAlow for all species regardless of native habitat or growth environment.
-Plants adapted to sunny, open habitats will evolve greater stomatal density ratio to take advantage of regular exposure to high light intensity.
-
-Plastic hypothesis: Individuals of the same genotype often develop dramatically different leaves in sun and shade conditions [@poorter_metaanalysis_2019].
+An amphistomatous leaf increases photosynthetic carbon gain compared to an otherwise identical hypostomatous leaf by increasing conductance through the leaf intercellular airspaces and boundary layers; the additional water loss through a second boundary layer is typically small [@foster_influence_1986].
+We quantify this benefit as the amphistomy advantage (\aax{}), the log-response ratio of photosynthesis in an amphistomatous leaf compared to an otherwise identical pseudohypostomatous leaf [@parkhurst_adaptive_1978; @triplett_amphistomy_2024].
+Why would \aax{} be greater in sun than shade?
+We consider three nonmutually exclusive hypotheses that we classify as 'acclimatory', 'plastic', and 'constitutive' ([Fig.
+@fig-concept]).
+
+*Acclimatory hypothesis*: Photosynthetic induction to high light intensity typically involves increases in total leaf stomatal conductance (increased CO$_2$ supply), the concentration of active Rubisco, and electron transport capacity (increased CO$_2$ demand).
+A one-dimensional circuit model using the Farquhar-von Caemmerer-Berry biochemical model of C$_3$ photosynthesis [@farquhar_biochemical_1980] shows that both increased stomatal conductance and Rubisco activity should increase \aax{}, all else being equal (Supporting Information).
+If the acclimatory hypothesis is correct, we predict that $\aahigh{} > \aalow{}$ for all populations regardless of native habitat or growth environment.
+
+*Plasticity hypothesis*: Individuals of the same genotype often develop dramatically different leaves in sun and shade conditions [@poorter_metaanalysis_2019].
 Plastic responses are likely adaptations to optimize photosynthesis at different light intensities in variable environments [@givnish_common-garden_2014].
-Plastic changes in leaf anatomy and biochemistry could modulate AA as a byproduct.
+Plastic changes in leaf anatomy and biochemistry could modulate \aax{} as a byproduct.
 Thicker or less porous leaves, both of which are associated with high leaf mass per area (LMA), will have lower $g_\mathrm{ias}$; leaves with increased total stomatal density and photosynthetic capacity have greater potential CO$_2$ supply and demand.
-Under the plastic hypothesis, we predict that AAsun \> AAshade for all species and light intensities.
-Secondarily, AAsun and gsmax,sun should also be positively associated with native light habitat if genotypes adapted to sunny, open habitats if they can express a phenotype best adapted to that environment when leaves develop under high light intensity.
-Genotypes adapted to shaded, closed habitats may be plastic, but limits on the width of their reaction norms prevent them from developing traits optimal for conditions they do not regularly experience in nature.
-
-Constitutive hypothesis: In environments that are relatively constant or where environmental change cannot be anticipated by a reliable cue, natural selection will favor constitutive expression of optimal phenotypes.
-We therefore predict genotypes from more sunny, open habitats will have consistently greater AA under also measurement and growth light intensities.
+Under the plastic hypothesis, we predict that $\aasun{} > \aashade{}$ for all populations and light intensities.
+Secondarily, $\aasun{}$ and \gmaxsun{} should also be positively associated with native light habitat if genotypes adapted to sunny, open habitats if they can express a phenotype best adapted to that environment when leaves develop under high light intensity.
+
+*Constitutive hypothesis*: In environments that are relatively constant or where environmental change cannot be anticipated by a reliable cue, natural selection will favor constitutive expression of optimal phenotypes.
+We therefore predict genotypes from more sunny, open habitats will have consistently greater \aax under also measurement and growth light intensities.
 For herbaceous plants, light intensity is largely a function of the tree canopy [@engelbrecht_evaluation_2001].
@@
 
-The primary directional predictions for each hypothesis are summarized in Table X; detailed predictions for results that would indicate support for multiple hypotheses are in Table SX.
-
-CONCEPTUAL FIGURE HERE
-
-caption: Directional predictions associated with each hypothesis explaining why amphistomy advantage (AA) might be greater for leaves in sunny, open habitats.
-For each hypothesis, we make predictions for how native plant area index (PAI), growth light treatment, and measurement light intensity would affect AA.
-
-|  hypothesis  | measurement light intensity | growth light intensity  | native PAI |
-|:------------:|:---------------------------:|:------------------------:|:----------:|
-| acclimatory  | AA2000 > AA150             | AAsun = AAshade          | cor(PAI, AA) = 0 |
-| plastic      | AA2000 = AA150             | AAsun > AAshade          | cor(PAI, AA) = 0 |
-| constitutive | AA2000 = AA150             | AAsun = AAshade          | cor(PAI, AA) > 0 |
+```{r, concept, echo = FALSE}
+#| label: fig-concept
+#| fig-align: center
+#| fig-cap: "\\textbf{Conceptual outline of three nonmutually exclusive hypotheses explaining why amphistomy advantage (\\aax{}) might be greater for leaves in sunny, open habitats.} The acclimatory hypothesis predicts that \\aax{} is greater under high measurement light intensity (\\ppfdqty{2000}) than low measurement light intensity (\\ppfdequals{150}), regardless of growth light intensity or native plant area index (PAI).
+The plasticity hypothesis predicts that \\aax{} is greater for plants grown in sun than shade, regardless of measurement light intensity or native PAI.
+The constitutive hypothesis predicts that \\aax{} is greater for plants adapted to sunny habitats (lower native PAI) than shaded habitats, regardless of measurement light intensity or growth light intensity."
+
+knitr::include_graphics("../figures/amphisomy draft 2024-03-28.jpg")
+```
+
+The primary directional predictions for each hypothesis are summarized in @tbl-hypotheses; detailed predictions for results that would indicate support for multiple hypotheses are in @tbl-predictions.
+
+```{r, hypotheses, echo = FALSE}
+#| label: tbl-hypotheses
+#| tbl-cap: "Three nonmutually exclusive hypotheses and directional predictions explaining why amphistomy advantage (\\aax{}) might be greater for leaves in sunny, open habitats.
+For each hypothesis, we make predictions for how measurement light intensity (\\ppfdequals{150} vs.
+\\ppfdqty{2000}), growth light intensity (sun vs.
+shade), and native plant area index (PAI) would affect \\aax{}.
+\\ppfd{}: photosynthetic photon flux density."
+
+kableExtra::kable(
+  data.frame(
+    hypothesis = c("acclimatory", "plastic", "constitutive"),
+    measurement = c(
+      "$\\mathrm{AA}_{2000} > \\mathrm{AA}_{150}$",
+      "$\\mathrm{AA}_{2000} = \\mathrm{AA}_{150}$",
+      "$\\mathrm{AA}_{2000} = \\mathrm{AA}_{150}$"
+    ),
+    growth = c(
+      "$\\mathrm{AA}_{\\text{sun}} = \\mathrm{AA}_{\\text{shade}}$",
+      "$\\mathrm{AA}_{\\text{sun}} > \\mathrm{AA}_{\\text{shade}}$",
+      "$\\mathrm{AA}_{\\text{sun}} = \\mathrm{AA}_{\\text{shade}}$"
+    ),
+    native = c(
+      "$\\operatorname{cor}(\\mathrm{PAI}, \\mathrm{AA}) = 0$",
+      "$\\operatorname{cor}(\\mathrm{PAI}, \\mathrm{AA}) = 0$",
+      "$\\operatorname{cor}(\\mathrm{PAI}, \\mathrm{AA}) > 0$"
+    )
+  ),
+  col.names = c(
+    "Hypothesis",
+    "Measurement light intensity",
+    "Growth light intensity",
+    "Native PAI"
+  ),
+  escape = FALSE,
+  booktabs = TRUE,
+  format = "latex",
+  align = "c"
+) |>
+  kable_styling() |>
+  column_spec(1, width = "1in") |>
+  column_spec(2, width = "1.5in") |>
+  column_spec(3, width = "1.5in") |>
+  column_spec(4, width = "1.25in")
+```
 
 <!-- Paragraph 5 (summarize methods) -->
-We tested these hypotheses by comparing AA among amphistomatous wild tomato species [@peralta_taxonomy_2008] from different native light habitats, grown under simulated sun and shade light treatments, and measured under contrasting light intensity (Figure of hypotheses and predictions).
-We measured AA on `r n_plant` individual plants from `r n_acc` accessions (average of `r mean_n_rep` replicates per light treatment) using a recently developed method [@triplett_amphistomy_2024].
+We tested these hypotheses by comparing \aax{} among amphistomatous wild tomato species [@peralta_taxonomy_2008] from different native light habitats, grown under simulated sun and shade light treatments, and measured under contrasting light intensity ().
+We measured AA on `r n_plant` individual plants from `r n_acc` populations (average of `r mean_n_rep` replicates per light treatment) using a recently developed method [@triplett_amphistomy_2024].
 With this method, we directly compare the photosynthetic rate of an untreated amphistomatous leaf to that of the same leaf with gas exchange blocked through the adaxial (upper) surface by transparent plastic, which we refer to as 'pseudohypostomy'.
@@
 The native light intensity was represented by plant area index (PAI $\unit{\meter\squared\per\meter\squared}$), estimated using a global gridded X-m2 resolution data set derived from the Global Ecosystem Dynamics Investigation [GEDI; @burns_multi-resolution_2024] and georeferenced accession collection information from the Tomato Genetics Resource Center.
-The growth light intensities were `r ppfd_sun` (sun treatment) and `r ppfd_shade` (shade treatment) while all other environment conditions were nearly identical (see @sec-methods for supporting detail).
+The growth light intensities were \ppfdequals{`r ppfd_sun`} (sun treatment) and \ppfdqty{`r ppfd_shade`} (shade treatment) while all other environment conditions were nearly identical (see materials and methods in supplementary materials).
 The high and low measurements intensities were \ppfdequals{2000} (`r p_red_high`:`r p_blue_high` red:blue) and \ppfdequals{150} (`r p_red_low`:`r p_blue_low` red:blue), respectively.
@@
 <!-- Paragraph 6 (summarize main results) -->
-Consistent with biophysical theory of CO$_2$ diffusion within leaves, AA > 0 for all accessions (Figure X).
-AA varied substantially between measurement light intensities, growth light intensities, and among accessions (point to model comparison evidence).
-Measured under high light intensity, AA was consistently greater for sun plants.
-The average AA among accessions in the shade treatment was `r filter(aa_mean_range, light_treatment == "low", light_intensity == "2000")$x`; however, the same accessions grown at high light intensity showed a mean AA of `r filter(aa_mean_range, light_treatment == "high", light_intensity == "2000")$x`.
-Contrary to the predictions of the assimilatory hypothesis, AA was greater in all accessions under low measurement light intensity for both sun and shade grown plants.
-The effect of low light on AA was more pronounced in the sun-grown plants, where AA was significantly greater under low measurement light intensity in `r aa_summary3["high"]` of `r n_acc` accessions compared to to `r aa_summary3["low"]` accessions for shade-grown plants.
-The overall average AA of shade and sun grown plants measured under low light intensity was `r filter(aa_mean_range, light_treatment == "low", light_intensity == "150")$x` and `r filter(aa_mean_range, light_treatment == "high", light_intensity == "150")$x`, respectively.
-There was a slight tendency for accessions from more closed habitats (greater PAI), but AA varied widely among accessions from open habitats (low PAI) regardless of growth and measurement light intensities (Figure X).
+Consistent with biophysical theory of CO$_2$ diffusion within leaves, $\aay{} > 0$ for all populations ([Fig.
+@fig-aa]A).
+\aax{} varied substantially between measurement light intensities, growth light intensities, and among populations (point to model comparison evidence).
+Measured under high light intensity, \aax{} was consistently greater for sun plants.
+The average \aax{} among populations in the shade treatment was `r filter(aa_mean_range, light_treatment == "low", light_intensity == "2000")$x`; however, the same populations grown at high light intensity showed a mean \aax{} of `r filter(aa_mean_range, light_treatment == "high", light_intensity == "2000")$x`.
+Contrary to the predictions of the assimilatory hypothesis, \aax{} was greater in all populations under low measurement light intensity for both sun and shade grown plants.
+The effect of low light on \aax{} was more pronounced in the sun-grown plants, where \aax{} was significantly greater under low measurement light intensity in `r aa_summary3["high"]` of `r n_acc` populations compared to to `r aa_summary3["low"]` populations for shade-grown plants.
+The overall average \aax{} of shade and sun grown plants measured under low light intensity was `r filter(aa_mean_range, light_treatment == "low", light_intensity == "150")$x` and `r filter(aa_mean_range, light_treatment == "high", light_intensity == "150")$x`, respectively.
+There was a slight tendency for populations from more closed habitats (greater PAI), but \aax{} varied widely among populations from open habitats (low PAI) regardless of growth and measurement light intensities ([Fig.
+@fig-aa]B).
+
+```{r, aa, echo = FALSE, warning=FALSE, message=FALSE} 
+#| label: fig-aa
+#| fig-align: center
+#| fig-width: 6
+#| fig-height: 8
+#| fig-cap: "Caption on following page."
+
+plot_grid(
+  read_rds("../objects/fig_aa.rds"),
+  read_rds("../objects/fig_aa_pai.rds"),
+  ncol = 1,
+  labels = "AUTO",
+  label_size = 20,
+  label_fontface = "bold",
+  align = "hv"
+)
+
+```
+
+```{=latex}
+\addtocounter{figure}{-1}
+```
+
+```{r, aa_cap, echo = FALSE, warning=FALSE, message=FALSE} 
+#| label: fig-aacap
+#| fig-align: center
+#| fig-width: 0.1
+#| fig-height: 0.1
+#| fig-cap: "\\textbf{Amphistomy advantage (\\aax{}) in wild tomatoes is amplified under simulated sunny growth conditions.} \\aax{} ($y$-axis) is the log-response ratio of photosynthesis in an amphistomatous leaf compared to an otherwise identical pseudohypostomatous leaf.
+In both panels, estimates are shown for plants grown under simulated shade or sun and measured under high light intensity (\\ppfdequals{2000}) and low light intensity (\\ppfdequals{150}).
+The dashed line at zero indicates no difference in photosynthesis between amphi- and pseudohypostomatous leaves.
+(\\textbf{A}) The points are the posterior median \\aax{} for each population, with error bars showing 95% confidence intervals.
+Within each facet, the populations are arranged by \\aax{} estimated in that growth and measurement condition.
+(\\textbf{B}) The same estimates of \\aax{} in each combination of growth and measurement light intensity ploted against native plant area index (PAI; $x$-axis, log-scale).
+The confidence intervals were omitted for visual clarity."
+
+ggplot()
+```
 
 <!-- Paragraph 7 (summarize secondary results) -->
-The pattern of AA across wild tomatoes strongly supports the plasticity hypothesis, argues against the acclimatory hypothesis, and provides only weak support for the constitutive hypothesis.
-Plastic changes in leaf thickness and/or packing density, summarized by the bulk leaf mass per area (LMA), likely mediate the effect of growth light intensity on AA.
-LMA increased in sun grown plants in all accessions by an average of X% [CI], similar to plastic responses in many species [@poorter_metaanalysis_2019].
-When LMA is included in the model, the effect of growth light intensity on AA is no longer significant based on model comparison (Table ?
-- need to check out what the interaction with measurement light intensity means) and LMA is weakly positively correlated with AA (Figure X).
-In contrast, LMA did not mediate indirect effects of native PAI on AA (Table ?).
-Many anatomical traits underlie LMA [@john_anatomical_2017] and future research will be needed identify which particular traits, such as leaf thickness or mesophyll porosity, are responsible for mediating AA.
-
-#### MARKER
-These results suggest that developmental plasticity, as opposed to acclimation or adaptation to open habitats, may explain the explain the long-standing observation taht amphistomatous leaves are more common in sunny habitats.
-Our results add to existing explanations by showing high light intensity *per se* does not increase the benefit of amphistomy, but rather than anatomical and biochemical change associated with higher light intensity modulate AA.
-We cannot yet attribute specific changes wrought by plasticity, but correlational evidence suggests that changes in leaf anatomy may be important.
-Species responded consistently by increasing stomatal density, especially on the adaxial surface, increasing stomatal density ratio.
-Sun leaves also had greater gs, Amax.
-However, these traits were not associated with change in AA (table/figures).
-
-Not sure where this goes: We infer this from the fact that blocking gas exchange in pseudohypostomatous leaves reduced $A$ by X-X% depending on the accession, light treatment, and light intensity (Table/figure).
-The AA is equivalent to an X-X% change in total $g_\text{sc}$ (see SI section gs equivalency).
-But whereas increasing $g_\text{sc}$ would increase water loss as a necessary by-product, amphistomy can increase $A$ without any appreciable affect on transpiration.
-
-from Mott and Oleary: "However, Jones and Slatyer (5) reported a higher mesophyll resistance for CO, entering through the upper stomata than for the lower, and the data of Vaclavik (12) appear to support this conclusion"
-
-## Outtro - I think this goes at the end??
-It is commonly assumed in comparative studies that most trait variation between species is constitutive and genetically determined, but our results suggest that plastic responses to light are necessary to explain why \gmaxratio{} increases with light habitat.
-Reaction norms to light intensity among wild tomatoes indicate that accessions from low light habitats express low \gmaxratio{} when grown in shade, whereas accessions from high light habitats have constitutively higher \gmaxratio{} (Fig.
-rxn-norm).
-The larger response to light intensity is driven by greater plasticty in leaf structure in accessions from low-light habitats (Fig/table - I haven't tested this yet, so this is a placeholder.) We infer that these structural changes in sun leaves are costly because they increase the diffusion path length for CO$_2$ and reduce photosynthesis.
-As a consequence, sun leaves benefit more from amphistomy.
-This is the first demonstration that sun leaves benefit more from amphistomy because it compensates for developmental plasticity in leaf structure.
-
-Two outstanding areas for further research are 1) the specific anatomical changes, such as leaf thickness or mesophyll porosity, that increase the diffusion path length for CO$_2$ in sun leaves, and 2) the costs of amphistomy that could explain plasticity in \gmaxratio{} in accessions from low-light habitats.
-Resolving the first question will require detailed anatomical measurements on a large number of shade and sun leaves with concomitant estimates of \aax.
-The fact that accessions from low-light habitats are more plastic suggests that costs of amphistomy are greatest in these environments.
-(not finished yet)
-
-Problems to bring up in discussion:
-- all our species were amphi, so they may all live in PAI below threshold to favor amphi
-- other microclimate diffs not considered
-- don't know specific anatomical change
-
-8. What it all means
-  * Our study reveals an important role for plasticity in explaining why amphi leaves are more common in sunny habitats
-  * Confirms that leaf structure of sun leaves incurs costs that amphistomy can mitigate
-  * It will be useful to consider coordination between leaf anatomy, stomatal density, and ratio in future studies of leaf structure-function and in crops (or something)
-
-## Materials and Methods [#sec-methods]
-
-\[this will be moved to SI eventually. note that much of it also the same in solanum-aq project, so might change to cite that paper?\]
-
-(not sure where this goes)  * We acclimated the focal leaf to high light (\ppfdequals{2000}) and high relative humidity (\rhequals{70}) until $A$ and \gsw{} reach their maximum.
-After that, we decreased \rh{} to $\approx 10\%$ to induce rapid stomatal closure without biochemical down
-
-### Accessions
-
-We compared AA among `r n_acc` ecologically diverse accessions of wild tomato, including representatives of all described species of *Solanum* sect.
+The pattern of \aax{} across wild tomatoes strongly supports the plasticity hypothesis, argues against the acclimatory hypothesis, and provides only weak support for the constitutive hypothesis.
+Plastic changes in leaf thickness and/or packing density, summarized by the bulk leaf mass per area (LMA), likely mediate the effect of growth light intensity on \aax{}.
+LMA increased in sun grown plants in all populations by an average of `r signif(100 * (exp(post_lma["Estimate"]) - 1), 3)`% [95% CI: `r signif(100 * (exp(post_lma["Q2.5"]) - 1), 3)` to `r signif(100 * (exp(post_lma["Q97.5"]) - 1), 3)`%], similar to plastic responses in many species [@poorter_metaanalysis_2019].
+When LMA is included in the model, the effect of growth light intensity on \aax{} is no longer significant based on model comparison (Table S8) and LMA is weakly positively correlated with \aax{} ([Fig.
+@fig-lma_aa]).
+In contrast, LMA did not mediate an indirect effect of native PAI on \aax{} (Table S8).
+Many anatomical traits underlie LMA [@john_anatomical_2017] and future research will be needed identify which particular traits, such as leaf thickness or mesophyll porosity, are responsible for mediating \aax{}.
+The fact that the \aax{} of sun plants was greater under low measurement light intensity supports a long-standing hypothesis that resistance to CO$_2$ diffusion is greater in the upper than lower portions of the leaf interior [@jones_effects_1972].
+At low light intensity, the bulk properties of the leaf are weighed toward the upper palisade where most light is intercepted.
+Hence, amphistomy may, expectedly, be particularly beneficial for sun leaves experiencing intermittant shade or cloud cover.
+Our study is limited in testing this because we could not directly measure the stomatal conductance ratio and intercellular resistance on each surface.
+Future experiments measuring \aax{} with a dual sided chamber [@wall_stomata_2022; @marquez_assessing_2023] can overcome these limitations.
+
+```{r, aa, echo = FALSE,warning=FALSE, message=FALSE} 
+#| label: fig-lma_aa
+#| fig-align: center
+#| fig-width: 6
+#| fig-height: 4
+#| fig-cap: "\\textbf{Plastic changes in leaf mass per area (LMA) mediate the effect of sun- to shade-grown plants on amphistomy advantage (\\aax{}).} Each point is the estimated population-level LMA ($x$-axis) and \\aax{} ($y$-axis) for each population grown under high light intensity (\\ppfdequals{2000}) and low light intensity (\\ppfdequals{150}).
+The points are colored by growth light treatment.
+The solid line is the posterior median of the linear regression of individual-level LMA on $\\aahigh$  with the shaded region showing the 95% confidence ribbon; the dashed line is the same for $\\aalow$.
+Confidence intervals for each population estimate and individual-level estimates are omitted for visual clarity."
+
+read_rds("../objects/fig_lma_aa.rds")
+
+```
+
+We conclude that developmental plasticity, as opposed to acclimation or adaptation to open habitats, may explain the explain the long-standing observation that amphistomatous leaves are more common in sunny habitats, at least among herbaceous plants including crop relatives.
+Our results advance existing explanations by showing high light intensity *per se* does not increase the benefit of amphistomy, but rather that anatomical and biochemical change associated with higher light intensity modulate \aax{}.
+Developmental plasticity needs to be considered when interpreting phylogenetic comparative analyses of adaptation, because plastic differences may be mistaken for genetic differences.
+
+A second major conclusion is that the significant photosynthetic gain of amphistomatous leaves implies that the many hypostomatous leaves that dominate mesic to wet forests globally are giving up 'free' carbon that would require little to no additional water loss.
+Under realistic assumptions, we estimate that hypostomatous leaves could achieve the same $A$ with X to X% less water loss if they were amphistomatous (Table S9).
+Ours is the first experimental demonstration that amphistomy is beneficial for sun leaves because of plastic changes, but future research needs to estimates the tradeoffs associated with amphistomatous leaves in order to fully explain their rarity among many woody plants and shade-tolerant herbs.
+
+## Acknowledgements
+
+Sam McKlin and Tom Buckley helped with protocol development.
+Justin Alter, Max Gatlin, Joana Kim, Jenna Matsuyama, Brandon Najarian, and Kai Yasuda contributed to data collection.
+Sarah Friedrich helped with conceptual figure.
+
+### Funding:
+
+US National Science Foundation OIA-1929167 to C.D.M.
+
+### Author contributions:
+
+Conceptualization: C.D.M.; Methodology: C.D.M., W.S.L.; Investigation: C.D.M., W.S.L., D.W.; Visualization: C.D.M.; Funding acquisition: C.D.M.; Writing – original draft: C.D.M.; Writing – review & editing: C.D.M., W.S.L., D.W.
+
+## References
+
+::: {#refs}
+:::
+
+{{< pagebreak >}}
+
+# Supplementary Materials
+
+```{=latex}
+\renewcommand{\thefigure}{S\arabic{figure}}
+\renewcommand{\thetable}{S\arabic{table}}
+\renewcommand{\theequation}{S\arabic{equation}}
+\setcounter{figure}{0}
+\setcounter{table}{0}
+\setcounter{equation}{0}
+```
+
+## Materials and Methods {#sec-methods}
+
+### Populations
+
+We compared \aax{} among `r n_acc` ecologically diverse populations of wild tomato, including representatives of all described species of *Solanum* sect.
 *Lycopersicon* and sect.
@@
 lycopersicum* var.
-*lycopersicum* \autoref{tab:accessions}.
+*lycopersicum* (@tbl-populations).
 Due to constraints on growth space and time, we spread out measurements over `r n_weeks` weeks from `r start_licor_date` to `r end_licor_date`.
-Replicates within accession were evenly spread out over this period to prevent confounding of temporal variation in growth conditions with accession.
+Replicates within population were evenly spread out over this period to prevent confounding of temporal variation in growth conditions with accession.
 [anything else to say here?
-maybe explain accession selection and phylogeny?]
-
-```{r, accessions, eval=TRUE,echo=FALSE,message=FALSE,warning=FALSE}
-#| label: tbl-accessions
-#| tbl-cap: "Accessions of *Solanum* used in this study.
+maybe explain population selection and phylogeny?]
+
+```{r, populations, eval=TRUE,echo=FALSE,message=FALSE,warning=FALSE}
+#| label: tbl-populations
+#| tbl-cap: "Accession information of *Solanum* populations used in this study.
 The species name, accession number, collection latitude, longitude, elevation, and daily solar radiation estimated using the SPLASH algorithm (see 'Climate data').
-TGRC: Tomato Genetics Resource Center; PPFD: Photosynthetic Photon Flux Density."
+TGRC: Tomato Genetics Resource Center; PAI: Plant area index."
 
@@
 
-Seedlings were randomly assigned in alternating order within accession to the sun or shade treatment during transplanting.
+Seedlings were randomly assigned in alternating order within population to the sun or shade treatment during transplanting.
 After seedlings established in cell-pack flats for $\approx 2$ weeks, we transplanted them to 3.78 L plastic pots containing 60% Pro-Mix BX potting mix, 20% coral sand (Pro-Pak, Honolulu, Hawaiʻi, USA), and 20% cinders (Niu Nursery, Honolulu, Hawaiʻi, USA).
@@
 To maintain homogeneous environmental conditions other than light, we mixed air within the growth room using an air circulator (Vornado 693DC, Andover, Kansas, USA) and within racks using a miniature oscillating air circulator (Vornado Atom 1, Andover, Kansas, USA).
-Despite these efforts, the air in the sun treatment was on average $\qty{`r temp_diff`}{\degreeCelsius}$ warmer and the average RH was consequently `r rh_diff` lower.
+Despite these efforts, the air in the sun treatment was on average $\qty{`r temp_diff`}{\degreeCelsius}$ warmer and the average RH was consequently `r rh_diff`% lower.
 However, because of evaporative cooling, the leaves in the sun treatment were only $\qty{`r leaf_temp_diff`}{\degreeCelsius}$ on average ($n = `r n_leaf_temp`$ leaves).
@@
 In this procedure, curve parameters were provisionally estimated using ordinary least squares with the `lm()` function in *R*.
-We then used point estimates of \aax{} for each replicate as the response variable in a linear model with light treatment, light intensity, accession, and all interactions as explanatory variables.
+We then used point estimates of \aax{} for each replicate as the response variable in a linear model with light treatment, light intensity, population, and all interactions as explanatory variables.
 This model was also fit using ordinary least squares with the `lm()` function in *R*.
@@
 #| label: tbl-models
-#| tbl-cap: "Summary of differences among competing models of how \\aax{} varies with light intensity, light treatment, and among accessions as a function of native \\ppfd.
+#| tbl-cap: "Summary of differences among competing models of how \\aax{} varies with light intensity, light treatment, and among populations as a function of native \\ppfd.
 The models are numbered from simpler to more complex.
 All models include fixed effects of light intensity and light treatment; some models include interactions between.
-All models include a phylogenetic random effect of accession on \\aax; some models include varying effects of light intensity and light treatment among accessions.
-The last column indicates the accession-level \\aax{} variable we used as a response to native \\ppfd."
+All models include a phylogenetic random effect of population on \\aax; some models include varying effects of light intensity and light treatment among populations.
+The last column indicates the population-level \\aax{} variable we used as a response to native \\ppfd."
 
@@
     ~"Model", ~"Fixed effects", ~"Phylogenetic random effects", ~"Response to native \\ppfd",
-    "1", "light intensity", "varying intercept among accessions", "$\\mathrm{AA}_{0, \\text{acc}}$",
+    "1", "light intensity", "varying intercept among populations", "$\\mathrm{AA}_{0, \\text{acc}}$",
     "1", "light treatment", "", "",
-    "2", "light intensity", "varying intercept among accessions", "$\\mathrm{AA}_{0, \\text{acc}}$",
+    "2", "light intensity", "varying intercept among populations", "$\\mathrm{AA}_{0, \\text{acc}}$",
     "2", "light treatment", "", "",
     "2", "intensity $\\times$ treatment", "", "",
-    "3", "light intensity", "varying intercept among accessions", "$\\mathrm{AA}_{2000, \\text{acc}}$",
-    "3", "light treatment", "varying effect of high light intensity among accessions", "",
-    "4", "light intensity", "varying intercept among accessions", "$\\mathrm{AA}_{\\text{sun, acc}}$",
-    "4", "light treatment", "varying effect of sun treatment among accessions", "",
-    "5", "light intensity", "varying intercept among accessions", "$\\mathrm{AA}_{2000, \\text{sun, acc}}$",
-    "5", "light treatment", "varying effect of high light intensity among accessions", "",
-    "5", "intensity $\\times$ treatment", "varying effect of sun treatment among accessions", ""
+    "3", "light intensity", "varying intercept among populations", "$\\mathrm{AA}_{2000, \\text{acc}}$",
+    "3", "light treatment", "varying effect of high light intensity among populations", "",
+    "4", "light intensity", "varying intercept among populations", "$\\mathrm{AA}_{\\text{sun, acc}}$",
+    "4", "light treatment", "varying effect of sun treatment among populations", "",
+    "5", "light intensity", "varying intercept among populations", "$\\mathrm{AA}_{2000, \\text{sun, acc}}$",
+    "5", "light treatment", "varying effect of high light intensity among populations", "",
+    "5", "intensity $\\times$ treatment", "varying effect of sun treatment among populations", ""
   ) |>
@@
 
-We fit Bayesian models with MCMC sampling in the probabilistic programming language *Stan* [@stan_development_team_stan_2025] using the *R* package **brms** version `r fit_aa$version$brms` [@burkner_brms_2017].
-We used CmdStan version `r fit_aa$version$cmdstan` and **cmdstanr** version `r fit_aa$version$cmdstanr` [@gabry_cmdstanr_2025] to interface with *R* version `r getRversion()` [@r_core_team_r:_2025].
+We fit Bayesian models with MCMC sampling in the probabilistic programming language *Stan* [@stan_development_team_stan_2025] using the *R* package **brms** version `r fit_aa1$version$brms` [@burkner_brms_2017].
+We used *CmdStan* version `r fit_aa1$version$cmdstan` and **cmdstanr** version `r fit_aa1$version$cmdstanr` [@gabry_cmdstanr_2025] to interface with *R* version `r getRversion()` [@r_core_team_r:_2025].
 We sampled the posterior distribution from `r dim_fit_aa[2]` chains with `r dim_fit_aa[1]` iterations each after `r dim_fit_aa[1]` warmup iterations per chain.
 We estimated parameters and confidence intervals as the median and 95% quantile intervals of the posterior, respectively.
-We chose the number of chains, warmup and sampling iterations, and maximum treedepth so that parameter estimates converged ($\hat{R} < 1.01$ [@gelman_inference_1992]) and the effective sample size (ESS) for each parameter was $> 10^3$.
+We chose the number of chains, warmup and sampling iterations, and maximum treedepth so that parameter estimates converged ($\hat{R} < `r aa_args['max_rhat']`$ [@gelman_inference_1992]) and the effective sample size (ESS) for each parameter was $> `r aa_args['min_ess']`$.
 
@@
 2. Estimate \aax{} for each light intensity with leaf using \agcurve{} curve parameters
-3. Estimate the effects of light intensity, light treatment, and accession on \aax{} (assimilatory and plasticity hypotheses)
-4. Estimate the effects of native light habitat on accession-level \aax{} (constitutive hypothesis)
+3. Estimate the effects of light intensity, light treatment, and population on \aax{} (assimilatory and plasticity hypotheses)
+4. Estimate the effects of native light habitat on population-level \aax{} (constitutive hypothesis)
 
@@
                         escape = FALSE) |>
-  kableExtra::pack_rows("effects of light intensity, light treatments, and accession on \\aax",
+  kableExtra::pack_rows("effects of light intensity, light treatments, and population on \\aax",
                         min(which(df_parameters$par_type == "aa")),
@@
                         escape = FALSE) |>
-  kableExtra::pack_rows("effects of native light habitat on accession-level \\aax",
+  kableExtra::pack_rows("effects of native light habitat on population-level \\aax",
                         min(which(df_parameters$par_type == "ppfd_aa")),
@@
 Within the $k^{\text{th}}$ leaf, we estimated \aax{} for each light intensity by integrating the difference in \logA{} between the amphi and pseudohypo \agcurve{} curves over the range of \gsw{} values where the curves overlap (from $\min(\log(g_\text{sw}))$ to $\max(\log(g_\text{sw}))$).
-The estimate of \aax{} for the $k^{\text{th}}$ leaf at light intensity $l$ in accession $m$ is:
+The estimate of \aax{} for the $k^{\text{th}}$ leaf at light intensity $l$ in population $m$ is:
 
@@
 $$\theta_\text{hypo} \in \{\hat{b}_{0, f(\text{hypo}, k,l,m)}, \hat{b}_{1, f(\text{hypo}, k,l,m)}, \hat{b}_{2, f(\text{hypo}, k,l,m)}\}.$$ 
-The function $f : \Theta_1 \rightarrow \Theta_2$ maps the set $\Theta_1$ indexed by leaf type (amphi or pseudohypo), leaf replicate, light intensity, and accession to set $\Theta_2$ indexed by individual \agcurve{} curve.
+The function $f : \Theta_1 \rightarrow \Theta_2$ maps the set $\Theta_1$ indexed by leaf type (amphi or pseudohypo), leaf replicate, light intensity, and population to set $\Theta_2$ indexed by individual \agcurve{} curve.
 This mapping is necessary because the random effects structure differs between models of \loggsw{} on \logA{} and that of models predicting \aax{} described in the next section.
 
-##### Effects of light intensity, light treatment, and accession on \aax
-
-We tested for effects of light intensity, light treatment, accession, and their interactions on \aax.
-All models included effects of light intensity and light treatment, as well as random effects of accession and replicate within accession.
-More complex models included interactions between light intensity and light treatment, as well as random effects of accession on the effects of light intensity and light treatment.
+##### Effects of light intensity, light treatment, and population on \aax
+
+We tested for effects of light intensity, light treatment, population, and their interactions on \aax.
+All models included effects of light intensity and light treatment, as well as random effects of population and replicate within population.
+More complex models included interactions between light intensity and light treatment, as well as random effects of population on the effects of light intensity and light treatment.
 We used a weakly informative normal prior with mean $0$ and standard deviation $10$ for fixed effects of light intensity and treatment.
-We used a weakly informative normal prior with mean $-3$ and standard deviation $5$ for the the random effect standard deviation of replicate within accession.
-We accounted for the phylogenetic structure among the random effects of accession using an Ornstein-Uhlenbeck (OU) process [@hansen_stabilizing_1997].
-The expected covariance between accessions $i$ and $j$ ($\text{Cov}(i, j)$ ) is:
+We used a weakly informative normal prior with mean $-3$ and standard deviation $5$ for the the random effect standard deviation of replicate within population.
+We accounted for the phylogenetic structure among the random effects of population using an Ornstein-Uhlenbeck (OU) process [@hansen_stabilizing_1997].
+The expected covariance between populations $i$ and $j$ ($\text{Cov}(i, j)$ ) is:
 
 $$\text{Cov}(i, j) = \frac{\sigma^2}{2 \alpha} \exp(-\alpha D_{ij})$$
-where $\sigma^2$ is the variance of the random effect, $\alpha$ is the rate of decay of the covariance with phylogenetic distance, $D_{ij}$, the phylogenetic distance between accessions $i$ and $j$.
+where $\sigma^2$ is the variance of the random effect, $\alpha$ is the rate of decay of the covariance with phylogenetic distance, $D_{ij}$, the phylogenetic distance between populations $i$ and $j$.
 We estimated $\sigma^2 / (2 \alpha)$ and $\alpha$ as a separate parameters and reparameterized them as $\sigma^2$ and $\alpha$.
@@
 
-##### Effects of native light habitat on accession-level \aax
-
-We tested a linear effect of native light habitat on accession-level \aax.
-In Models 1 and 2, we used the random intercept of \aax{} as a response variable; in model 3 we used the random intercept plus the random effect of accession at high light intensity; in model 4 we used the random intercept plus the random effect of accession in the sun treatment; in model 5 we used the random intercept plus the random effects of accession at high light intensity and sun treatment.
+##### Effects of native light habitat on population-level \aax
+
+We tested a linear effect of native light habitat on population-level \aax.
+In Models 1 and 2, we used the random intercept of \aax{} as a response variable; in model 3 we used the random intercept plus the random effect of population at high light intensity; in model 4 we used the random intercept plus the random effect of population in the sun treatment; in model 5 we used the random intercept plus the random effects of population at high light intensity and sun treatment.
 We used a weakly informative normal prior with mean $0$ and standard deviation $1$ for the slope and intercept.
@@
 
-The assimilatory, plastic, and constitutive hypotheses make different predictions about the relationship between \aax{} and light intensity, light treatment, and native \ppfd{} among accessions.
+The assimilatory, plastic, and constitutive hypotheses make different predictions about the relationship between \aax{} and light intensity, light treatment, and native \ppfd{} among populations.
 Since these hypotheses are not mutually exclusive, we describe how we assessed support for one, two, or all three hypotheses simultaneously in \autoref{tbl-predictions}.
@@
 The plastic hypothesis was supported if \aax{} was greater in sun leaves than shade leaves.
-The constitutive hypothesis was supported if accession-level \aax{} increased with native \ppfd.
+The constitutive hypothesis was supported if population-level \aax{} increased with native \ppfd.
 In interactive models, we only consider positively reinforcing interactions between high light intensity, sun leaves, and native \ppfd{} because these are the only interactions which could explain why amphistomatous leaves are advantageous in high light habitats.
@@
 We used the leave-one-out cross-validation information criterion (LOOIC) to compare the fit of models (\autoref{tbl-models}) using the *R* package **loo** version `r packageVersion("loo")` [@vehtari_practical_2017] to calculate LOOIC values.
-Note that LOOIC was calculated only from pointwise likelihood values of the submodel estimating effects of light intensity, light treatments, and accession on \aax{} (see @sec-parameters).
+Note that LOOIC was calculated only from pointwise likelihood values of the submodel estimating effects of light intensity, light treatments, and population on \aax{} (see @sec-parameters).
 We considered models with two standard errors of the mean lower LOOIC value to be a better fit to the data; we considered models with LOOIC values within two standard errors of the mean to be have similar support.
@@
 #| label: tbl-predictions
-#| tbl-cap: "Predictions of competing hypotheses about the relationship between \\aax{} and light intensity, light treatment, and native \\ppfd{} among accessions.
+#| tbl-cap: "Predictions of competing hypotheses about the relationship between \\aax{} and light intensity, light treatment, and native \\ppfd{} among populations.
 The middle column lists specific directional predictions about parameter values and model fit according to leave-one-out cross-validation information criterion (LOOIC), where $\\text{LOOIC}_i$ is the LOOIC value for model $i$.
-The rightmost column describes the predictions in words and explains how accession-level \\aax{} is calculated in the relevant model."
+The rightmost column describes the predictions in words and explains how population-level \\aax{} is calculated in the relevant model."
 
@@
   "Plastic", "$\\beta_{\\mathrm{AA},\\text{sun}} > 0$", "Average \\aax{} in sun leaves is greater than that in shade leaves",
-  "Constitutive", "$\\beta_{\\mathrm{PPFD,AA}} > 0$", "Accession-level \\aax{} ($\\mathrm{AA}_\\text{acc}$) increases with native \\ppfd",
+  "Constitutive", "$\\beta_{\\mathrm{PPFD,AA}} > 0$", "Population-level \\aax{} ($\\mathrm{AA}_\\text{acc}$) increases with native \\ppfd",
   "Constitutive", "", "$\\mathrm{AA}_\\text{acc} = \\beta_{\\mathrm{AA}, 0} + \\vec{\\beta}_{\\mathrm{AA}, \\text{acc}}$",
@@
   "Assimilatory $\\times$ Constitutive", "$\\beta_{\\mathrm{AA},2000} > 0$", "Average \\aax{} at high light intensity is greater than that at low light intensity",
-  "Assimilatory $\\times$ Constitutive", "$\\beta_{\\mathrm{PPFD,AA}} > 0$", "Accession-level \\aax{} at high light intensity ($\\mathrm{AA}_{\\text{acc},2000}$) increases with native \\ppfd",
+  "Assimilatory $\\times$ Constitutive", "$\\beta_{\\mathrm{PPFD,AA}} > 0$", "Population-level \\aax{} at high light intensity ($\\mathrm{AA}_{\\text{acc},2000}$) increases with native \\ppfd",
   "Assimilatory $\\times$ Constitutive", "$\\text{LOOIC}_\\text{1} > \\text{LOOIC}_\\text{3}$", "$\\mathrm{AA}_{\\text{acc},2000} = \\beta_{\\mathrm{AA}, 0} + \\vec{\\beta}_{\\mathrm{AA}, \\text{acc}} + \\vec{\\beta}_{\\mathrm{AA}, 2000, \\text{acc}}$",
   "Plastic $\\times$ Constitutive", "$\\beta_{\\mathrm{AA},\\text{sun}} > 0$", "Average \\aax{} in sun leaves is greater than that in shade leaves",
-  "Plastic $\\times$ Constitutive", "$\\beta_{\\mathrm{PPFD,AA}} > 0$", "Accession-level \\aax{} in sun leaves ($\\mathrm{AA}_{\\text{acc},\\text{sun}}$) increases with native \\ppfd",
+  "Plastic $\\times$ Constitutive", "$\\beta_{\\mathrm{PPFD,AA}} > 0$", "Population-level \\aax{} in sun leaves ($\\mathrm{AA}_{\\text{acc},\\text{sun}}$) increases with native \\ppfd",
   "Plastic $\\times$ Constitutive", "$\\text{LOOIC}_\\text{1} > \\text{LOOIC}_\\text{4}$", "$\\mathrm{AA}_{\\text{acc},\\text{sun}} = \\beta_{\\mathrm{AA}, 0} + \\vec{\\beta}_{\\mathrm{AA}, \\text{acc}} + \\vec{\\beta}_{\\mathrm{AA}, \\text{sun}, \\text{acc}}$",
   "Assimilatory $\\times$ Plastic $\\times$ Constitutive", "$\\beta_{\\mathrm{AA},2000,\\text{sun}} > 0$", "Average \\aax{} is highest at high light intensity in sun leaves",
-  "Assimilatory $\\times$ Plastic $\\times$ Constitutive", "$\\beta_{\\mathrm{PPFD,AA}} > 0$", "Accession-level \\aax{} at high light intensity in sun leaves ($\\mathrm{AA}_{\\text{acc},2000,\\text{sun}}$) increases with native \\ppfd",
+  "Assimilatory $\\times$ Plastic $\\times$ Constitutive", "$\\beta_{\\mathrm{PPFD,AA}} > 0$", "Population-level \\aax{} at high light intensity in sun leaves ($\\mathrm{AA}_{\\text{acc},2000,\\text{sun}}$) increases with native \\ppfd",
   "Assimilatory $\\times$ Plastic $\\times$ Constitutive", "$\\text{LOOIC}_\\text{3} > \\text{LOOIC}_\\text{5}$", "$\\mathrm{AA}_{\\text{acc},2000,\\text{sun}} = \\beta_{\\mathrm{AA}, 0} + \\vec{\\beta}_{\\mathrm{AA}, \\text{acc}} + \\vec{\\beta}_{\\mathrm{AA}, 2000, \\text{acc}} + \\vec{\\beta}_{\\mathrm{AA}, \\text{sun}, \\text{acc}}$"
@@
 
-## Acknowledgements
-
-Justin Alter, Max Gatlin, Joana Kim, Jenna Matsuyama, Brandon Najarian, Kai Yasuda. Sarah Friedrich helped with figures.
-
-## References
+## Results
+
+* Variation in stomatal ratio among pops and light treatments
+* Table S7: Model comparison LOOIC tables
+* Table S8: Model comparison LOOIC tables with LMA
+* Table S9: Calculation of alternate scenario where amphi leaves can get same A with less R
+
+{{< pagebreak >}}
+
+### MANUSCRAPS
+
+(not sure where this goes)  * We acclimated the focal leaf to high light (\ppfdequals{2000}) and high relative humidity (\rhequals{70}) until $A$ and \gsw{} reach their maximum.
+After that, we decreased \rh{} to $\approx 10\%$ to induce rapid stomatal closure without biochemical down
+
+Problems to bring up in discussion:
+- all our species were amphi, so they may all live in PAI below threshold to favor amphi
+- other microclimate diffs not considered
+- don't know specific anatomical change
+
+Two outstanding areas for further research are 1) the specific anatomical changes, such as leaf thickness or mesophyll porosity, that increase the diffusion path length for CO$_2$ in sun leaves, and 2) the costs of amphistomy that could explain plasticity in \gmaxratio{} in populations from low-light habitats.
+Resolving the first question will require detailed anatomical measurements on a large number of shade and sun leaves with concomitant estimates of \aax.
+The fact that populations from low-light habitats are more plastic suggests that costs of amphistomy are greatest in these environments.
+(not finished yet)
+
+Not sure where this goes: We infer this from the fact that blocking gas exchange in pseudohypostomatous leaves reduced $A$ by X-X% depending on the population, light treatment, and light intensity (Table/figure).
+The AA is equivalent to an X-X% change in total $g_\text{sc}$ (see SI section gs equivalency).
+But whereas increasing $g_\text{sc}$ would increase water loss as a necessary by-product, amphistomy can increase $A$ without any appreciable affect on transpiration.
+
+from Mott and Oleary: "However, Jones and Slatyer (5) reported a higher mesophyll resistance for CO, entering through the upper stomata than for the lower, and the data of Vaclavik (12) appear to support this conclusion"
+
+8. What it all means
+  * Our study reveals an important role for plasticity in explaining why amphi leaves are more common in sunny habitats
+  * Confirms that leaf structure of sun leaves incurs costs that amphistomy can mitigate
+  * It will be useful to consider coordination between leaf anatomy, stomatal density, and ratio in future studies of leaf structure-function and in crops (or something)
+
+I am using this approach to drafting the paper:
+
+https://blogs.nature.com/nyc/2011/08/10/how-to-write-a-paper-one-possible-answer
+
+## Punchline
+
+Plastic changes in the structure of sun leaves can increase the intercellular CO$_2$ diffusion path length.
+Gas exchange through stomata on both upper and lower leaf surfaces, amphistomy, may be a compensatory mechanism to reduce the impact of increased diffusion path length on photosynthesis without increasing water loss.
+We demonstrate for the first time that sun leaves benefit most from amphistomy because it compensates for plastic structural changes in leaf anatomy.
+
+What's surprising: role of developmental plasticity in explaining comparative results; diffusional limitations through palisade ias is strong and needs to be considered with light gradients in understanding leaf strucuture- funciton
+
+big implications - if trees were amphi, more more carbon sink or reduced transpiration
+
+It is commonly assumed in comparative studies that most trait variation between species is constitutive and genetically determined, but our results suggest that plastic responses to light are necessary to explain why \gmaxratio{} increases with light habitat.
+Reaction norms to light intensity among wild tomatoes indicate that populations from low light habitats express low \gmaxratio{} when grown in shade, whereas populations from high light habitats have constitutively higher \gmaxratio{} (Fig.
+rxn-norm).
+The larger response to light intensity is driven by greater plasticty in leaf structure in populations from low-light habitats (Fig/table - I haven't tested this yet, so this is a placeholder.) We infer that these structural changes in sun leaves are costly because they increase the diffusion path length for CO$_2$ and reduce photosynthesis.
+As a consequence, sun leaves benefit more from amphistomy.
+This is the first demonstration that sun leaves benefit more from amphistomy because it compensates for developmental plasticity in leaf structure.

### 2025-07-04 ms/ms.tex
first draft

@@
 \setlength{\emergencystretch}{3em} % prevent overfull lines
-\setcounter{secnumdepth}{5}
+\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
 % Make \paragraph and \subparagraph free-standing
+\makeatletter
 \ifx\paragraph\undefined\else
   \let\oldparagraph\paragraph
-  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
+  \renewcommand{\paragraph}{
+    \@ifstar
+      \xxxParagraphStar
+      \xxxParagraphNoStar
+  }
+  \newcommand{\xxxParagraphStar}[1]{\oldparagraph*{#1}\mbox{}}
+  \newcommand{\xxxParagraphNoStar}[1]{\oldparagraph{#1}\mbox{}}
 \fi
@@
   \let\oldsubparagraph\subparagraph
-  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
+  \renewcommand{\subparagraph}{
+    \@ifstar
+      \xxxSubParagraphStar
+      \xxxSubParagraphNoStar
+  }
+  \newcommand{\xxxSubParagraphStar}[1]{\oldsubparagraph*{#1}\mbox{}}
+  \newcommand{\xxxSubParagraphNoStar}[1]{\oldsubparagraph{#1}\mbox{}}
 \fi
+\makeatother
 
@@
 \makeatletter
-\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
-\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
-\makeatother
-% Scale images if necessary, so that they will not overflow the page
-% margins by default, and it is still possible to overwrite the defaults
-% using explicit options in \includegraphics[width, height, ...]{}
-\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
+\newsavebox\pandoc@box
+\newcommand*\pandocbounded[1]{% scales image to fit in text height/width
+  \sbox\pandoc@box{#1}%
+  \Gscale@div\@tempa{\textheight}{\dimexpr\ht\pandoc@box+\dp\pandoc@box\relax}%
+  \Gscale@div\@tempb{\linewidth}{\wd\pandoc@box}%
+  \ifdim\@tempb\p@<\@tempa\p@\let\@tempa\@tempb\fi% select the smaller of both
+  \ifdim\@tempa\p@<\p@\scalebox{\@tempa}{\usebox\pandoc@box}%
+  \else\usebox{\pandoc@box}%
+  \fi%
+}
 % Set default figure placement to htbp
-\makeatletter
 \def\fps@figure{htbp}
@@
 \usepackage{siunitx}
+\newcommand{\aahigh}{\mathrm{AA}_{2000}}
+\newcommand{\aalow}{\mathrm{AA}_{150}}
+\newcommand{\aashade}{\mathrm{AA}_{\text{shade}}}
+\newcommand{\aasun}{\mathrm{AA}_{\text{sun}}}
 \newcommand{\aax}{$\mathrm{AA}$}
+\newcommand{\aay}{\mathrm{AA}}
 \newcommand{\Aamphi}{$A_{\mathrm{amphi}}$}
@@
 \newcommand{\gmaxratio}{$g_\text{max,ratio}$}
+\newcommand{\gmaxshade}{$g_\text{max,shade}$}
+\newcommand{\gmaxsun}{$g_\text{max,sun}$}
 \newcommand{\gsw}{$g_\text{sw}$}
@@
 \makeatother
-\ifLuaTeX
-  \usepackage{selnolig}  % disable illegal ligatures
-\fi
+
 \usepackage{bookmark}
@@
 \hypersetup{
-  pdftitle={Unknown},
+  pdftitle={Developmental plasticity to high light intensity amplifies the advantage of amphistomous leaves},
+  pdfauthor={Christopher D. Muir\^{}\{1,2*\}; Wei Shen Lim\^{}\{1\}; Dachuan Wang\^{}\{1,2\}},
   colorlinks=true,
@@
 
-\title{Unknown}
-\author{}
+\title{Developmental plasticity to high light intensity amplifies the
+advantage of amphistomous leaves}
+\author{Christopher D. Muir\(^{1,2*}\) \and Wei Shen
+Lim\(^{1}\) \and Dachuan Wang\(^{1,2}\)}
 \date{}
@@
 
-\renewcommand*\contentsname{Table of contents}
-{
-\hypersetup{linkcolor=}
-\setcounter{tocdepth}{3}
-\tableofcontents
-}
-\section{Notes}\label{notes}
-
-Terminology to standardize:
-
-\begin{itemize}
-\item
-  the native light habitat or environment: the SPLASH paper refers to
-  what they calculated as `habitat' PPFD, so I am going to use that
-  terminology
-\item
-  growth condition versus measurement conditions
-\item
-  ?
-\end{itemize}
-
-\section{Authors}\label{authors}
-
-definite: Wei Shen Lim, Dachuan Wang, Chris Muir
-
-question: Sam McKlin, Tom Buckley
-
-I am using this approach to drafting the paper:
-
-https://blogs.nature.com/nyc/2011/08/10/how-to-write-a-paper-one-possible-answer
-
-\section{Punchline}\label{punchline}
-
-Plastic changes in the structure of sun leaves can increase the
-intercellular CO\(_2\) diffusion path length. Gas exchange through
-stomata on both upper and lower leaf surfaces, amphistomy, may be a
-compensatory mechanism to reduce the impact of increased diffusion path
-length on photosynthesis without increasing water loss. We demonstrate
-for the first time that sun leaves benefit most from amphistomy because
-it compensates for plastic structural changes in leaf anatomy.
-
-\section{Figures}\label{figures}
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\tightlist
-\item
-  competing hypotheses and predictions (or predictions after methods?)
-  for the benefit of amphistomy
-\item
-  design of experiment to test competing hypotheses (with actual \#'s of
-  replicate, curves, etc.)
-\item
-  Results on acclimatory, plastic, constitutive hypotheses
-\item
-  Additional supporting figure on change in LMA, reduction of Amass, and
-  relationship between LMA and AA? Modeling?
-\end{enumerate}
-
-\section{Abstract}\label{abstract}
-
-(use for GRC)
-
-Developmental plasticity to light intensity modulates amphistomy
-advantage
+\begin{center}
+$^1$School of Life Sciences, University of Hawaiʻi, Mānoa, Hawaiʻi, USA\\
+$^2$Department of Botany, University of Wisconsin, Madison, Wisconsin, USA\\
+\medskip
+*Corresponding author: cdmuir@wisc.edu\\
+\medskip
+ORCID: Christopher D. Muir — \href{https://orcid.org/0000-0003-2555-3878}{0000-0003-2555-3878}\\
+\end{center}
+
+\subsection{Abstract}\label{abstract}
 
@@
 CO\(_2\) diffusion between substomatal cavities and chloroplasts.
-Amphistomatous leaves are most common among herbaceous plants growing in
-sunny habitats, including many crop species. The distribution of
+Amphistomatous leaves are common among herbaceous plants growing in
+sunny habitats, including many crop species. This distribution of
 amphistomatous leaves in nature may result from an increased
@@
 relative to an otherwise identical hypostomatous leaf, in 29 diverse
-populations representing 15 species of wild tomatoes (\emph{Solanum}
+populations representing 14 species of wild tomatoes (\emph{Solanum}
 sect. \emph{Lycopersicon} and sect. \emph{Lycopersicoides}). Plant grown
-under high light intensity benefit more from amphostomy than those grown
-under low light, regardless of acclimation or native light intensity.
-Curvature of light response curves indicates that high-light leaves
-benefit more from amphistomy because of greater resistance to CO\(_2\)
-diffusion within the mesophyll. Developmental plasticity in leaf
-thickness and/or cell arrangement to optimize light interception likely
-increases under varying light intensity modulates CO\(_2\) diffusion
-within the leaf and, hence, amphistomy advantage as a byproduct.
+under high light intensity benefit more in terms of CO\(_2\) assimilated
+for a given stomatal conductance than plants grown under low light,
+regardless of measurement light intensity or native light habitat.
+Developmental plasticity in leaf thickness and/or density, an adaptive
+response to optimize light interception, likely affects CO\(_2\)
+diffusion within the leaf, altering amphistomy advantage as a byproduct.
 Developmental plasticity may therefore play an important and previously
@@
 
-What's surprising: role of developmental plasticity in explaining
-comparative results; diffusional limitations through palisade ias is
-strong and needs to be considered with light gradients in understanding
-leaf strucuture- funciton
-
-\section{Intro}\label{intro}
+\subsection{Main text}\label{main-text}
 
@@
 qualitative and quantitative. A higher proportion of sun leaves are
-amphistomatous (\textbf{salisbury\_causes\_1928?}) and the proportion of
-stomata on the upper, adaxial surface increases with light (\emph{42},
-\emph{43}). Resolving why high light intensity favors amphistomatous
-dorsivental leaves is an important first step toward understanding
-variation in stomatal density ratio and leaf structure-function
-relationships more generally.
+amphistomatous (\emph{39}) and the proportion of stomata on the upper,
+adaxial surface increases with light (\emph{42}, \emph{43}). Resolving
+why high light intensity favors amphistomatous dorsivental leaves is an
+important first step toward understanding variation in stomatal density
+ratio and leaf structure-function relationships more generally.
 
@@
 hypostomatous leaf by increasing conductance through the leaf
-intercellular airspaces and boudary layers; the additional water loss
+intercellular airspaces and boundary layers; the additional water loss
 through a second boundary layer is typically small (\emph{35}). We
-quantify this benefit as the amphistomy advantage (AA =
-log(Aamphi/Ahypo) (\emph{20}, \emph{46}). Why would AA be greater in sun
-than shade? We consider three nonmutually exclusive hypotheses that we
-classify as `acclimatory', `plastic', and `constitutive'.
-
-Acclimatory hypothesis: Photosynthetic induction to high light intensity
-typically involves increases in total leaf stomatal conductance
-(increased CO\(_2\) supply), the concentration of active Rubisco, and
-electron transport capacity (increased CO\(_2\) demand). A
+quantify this benefit as the amphistomy advantage (\aax{}), the
+log-response ratio of photosynthesis in an amphistomatous leaf compared
+to an otherwise identical pseudohypostomatous leaf (\emph{20},
+\emph{46}). Why would \aax{} be greater in sun than shade? We consider
+three nonmutually exclusive hypotheses that we classify as
+`acclimatory', `plastic', and `constitutive' (Fig.~\ref{fig-concept}).
+
+\emph{Acclimatory hypothesis}: Photosynthetic induction to high light
+intensity typically involves increases in total leaf stomatal
+conductance (increased CO\(_2\) supply), the concentration of active
+Rubisco, and electron transport capacity (increased CO\(_2\) demand). A
 one-dimensional circuit model using the Farquhar-von Caemmerer-Berry
 biochemical model of C\(_3\) photosynthesis (\emph{47}) shows that both
-increased stomatal conductance and Rubisco activity should increase AA,
-all else being equal (Supporting Information). If the acclimatory
-hypothesis is correct, we predict that AAhigh \textgreater{} AAlow for
-all species regardless of native habitat or growth environment. Plants
-adapted to sunny, open habitats will evolve greater stomatal density
-ratio to take advantage of regular exposure to high light intensity.
-
-Plastic hypothesis: Individuals of the same genotype often develop
-dramatically different leaves in sun and shade conditions (\emph{48}).
-Plastic responses are likely adaptations to optimize photosynthesis at
-different light intensities in variable environments (\emph{49}).
-Plastic changes in leaf anatomy and biochemistry could modulate AA as a
-byproduct. Thicker or less porous leaves, both of which are associated
-with high leaf mass per area (LMA), will have lower \(g_\mathrm{ias}\);
-leaves with increased total stomatal density and photosynthetic capacity
-have greater potential CO\(_2\) supply and demand. Under the plastic
-hypothesis, we predict that AAsun \textgreater{} AAshade for all species
-and light intensities. Secondarily, AAsun and gsmax,sun should also be
-positively associated with native light habitat if genotypes adapted to
-sunny, open habitats if they can express a phenotype best adapted to
-that environment when leaves develop under high light intensity.
-Genotypes adapted to shaded, closed habitats may be plastic, but limits
-on the width of their reaction norms prevent them from developing traits
-optimal for conditions they do not regularly experience in nature.
-
-Constitutive hypothesis: In environments that are relatively constant or
-where environmental change cannot be anticipated by a reliable cue,
-natural selection will favor constitutive expression of optimal
-phenotypes. We therefore predict genotypes from more sunny, open
-habitats will have consistently greater AA under also measurement and
+increased stomatal conductance and Rubisco activity should increase
+\aax{}, all else being equal (Supporting Information). If the
+acclimatory hypothesis is correct, we predict that
+\(\aahigh{} > \aalow{}\) for all populations regardless of native
+habitat or growth environment.
+
+\emph{Plasticity hypothesis}: Individuals of the same genotype often
+develop dramatically different leaves in sun and shade conditions
+(\emph{48}). Plastic responses are likely adaptations to optimize
+photosynthesis at different light intensities in variable environments
+(\emph{49}). Plastic changes in leaf anatomy and biochemistry could
+modulate \aax{} as a byproduct. Thicker or less porous leaves, both of
+which are associated with high leaf mass per area (LMA), will have lower
+\(g_\mathrm{ias}\); leaves with increased total stomatal density and
+photosynthetic capacity have greater potential CO\(_2\) supply and
+demand. Under the plastic hypothesis, we predict that
+\(\aasun{} > \aashade{}\) for all populations and light intensities.
+Secondarily, \(\aasun{}\) and \gmaxsun{} should also be positively
+associated with native light habitat if genotypes adapted to sunny, open
+habitats if they can express a phenotype best adapted to that
+environment when leaves develop under high light intensity.
+
+\emph{Constitutive hypothesis}: In environments that are relatively
+constant or where environmental change cannot be anticipated by a
+reliable cue, natural selection will favor constitutive expression of
+optimal phenotypes. We therefore predict genotypes from more sunny, open
+habitats will have consistently greater \aax under also measurement and
 growth light intensities. For herbaceous plants, light intensity is
@@
 
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../figures/amphisomy draft 2024-03-28.jpg}}
+
+}
+
+\caption{\label{fig-concept}\textbf{Conceptual outline of three nonmutually exclusive hypotheses explaining why amphistomy advantage (\aax{}) might be greater for leaves in sunny, open habitats.}
+The acclimatory hypothesis predicts that \aax{} is greater under high
+measurement light intensity (\ppfdqty{2000}) than low measurement light
+intensity (\ppfdequals{150}), regardless of growth light intensity or
+native plant area index (PAI). The plasticity hypothesis predicts that
+\aax{} is greater for plants grown in sun than shade, regardless of
+measurement light intensity or native PAI. The constitutive hypothesis
+predicts that \aax{} is greater for plants adapted to sunny habitats
+(lower native PAI) than shaded habitats, regardless of measurement light
+intensity or growth light intensity.}
+
+\end{figure}%
+
 The primary directional predictions for each hypothesis are summarized
-in Table X; detailed predictions for results that would indicate support
-for multiple hypotheses are in Table SX.
-
-CONCEPTUAL FIGURE HERE
-
-caption: Directional predictions associated with each hypothesis
-explaining why amphistomy advantage (AA) might be greater for leaves in
-sunny, open habitats. For each hypothesis, we make predictions for how
-native plant area index (PAI), growth light treatment, and measurement
-light intensity would affect AA.
-
-\begin{longtable}[]{@{}
-  >{\centering\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.1728}}
-  >{\centering\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.3580}}
-  >{\centering\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.3210}}
-  >{\centering\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.1481}}@{}}
-\toprule\noalign{}
-\begin{minipage}[b]{\linewidth}\centering
-hypothesis
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-measurement light intensity
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-growth light intensity
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-native PAI
-\end{minipage} \\
-\midrule\noalign{}
-\endhead
-\bottomrule\noalign{}
-\endlastfoot
-acclimatory & AA2000 \textgreater{} AA150 & AAsun = AAshade & cor(PAI,
-AA) = 0 \\
-plastic & AA2000 = AA150 & AAsun \textgreater{} AAshade & cor(PAI, AA) =
-0 \\
-constitutive & AA2000 = AA150 & AAsun = AAshade & cor(PAI, AA)
-\textgreater{} 0 \\
-\end{longtable}
-
-We tested these hypotheses by comparing AA among amphistomatous wild
+in Table~\ref{tbl-hypotheses}; detailed predictions for results that
+would indicate support for multiple hypotheses are in
+Table~\ref{tbl-predictions}.
+
+\begin{table}
+
+\caption{\label{tbl-hypotheses}Three nonmutually exclusive hypotheses
+and directional predictions explaining why amphistomy advantage (\aax{})
+might be greater for leaves in sunny, open habitats. For each
+hypothesis, we make predictions for how measurement light intensity
+(\ppfdequals{150} vs.~\ppfdqty{2000}), growth light intensity (sun
+vs.~shade), and native plant area index (PAI) would affect \aax{}.
+\ppfd{}: photosynthetic photon flux density.}
+
+\centering{
+
+\centering
+\begin{tabular}{>{\centering\arraybackslash}p{1in}>{\centering\arraybackslash}p{1.5in}>{\centering\arraybackslash}p{1.5in}>{\centering\arraybackslash}p{1.25in}}
+\toprule
+Hypothesis & Measurement light intensity & Growth light intensity & Native PAI\\
+\midrule
+acclimatory & $\mathrm{AA}_{2000} > \mathrm{AA}_{150}$ & $\mathrm{AA}_{\text{sun}} = \mathrm{AA}_{\text{shade}}$ & $\operatorname{cor}(\mathrm{PAI}, \mathrm{AA}) = 0$\\
+plastic & $\mathrm{AA}_{2000} = \mathrm{AA}_{150}$ & $\mathrm{AA}_{\text{sun}} > \mathrm{AA}_{\text{shade}}$ & $\operatorname{cor}(\mathrm{PAI}, \mathrm{AA}) = 0$\\
+constitutive & $\mathrm{AA}_{2000} = \mathrm{AA}_{150}$ & $\mathrm{AA}_{\text{sun}} = \mathrm{AA}_{\text{shade}}$ & $\operatorname{cor}(\mathrm{PAI}, \mathrm{AA}) > 0$\\
+\bottomrule
+\end{tabular}
+
+}
+
+\end{table}%
+
+We tested these hypotheses by comparing \aax{} among amphistomatous wild
 tomato species (\emph{51}) from different native light habitats, grown
 under simulated sun and shade light treatments, and measured under
-contrasting light intensity (Figure of hypotheses and predictions). We
-measured AA on 572 individual plants from 29 accessions (average of 9.86
-replicates per light treatment) using a recently developed method
-(\emph{46}). With this method, we directly compare the photosynthetic
-rate of an untreated amphistomatous leaf to that of the same leaf with
-gas exchange blocked through the adaxial (upper) surface by transparent
-plastic, which we refer to as `pseudohypostomy'. To compare amphi- and
-pseudohypostomatous leaves at identical whole-leaf \(g_\text{sc}\), we
-measure \(A\) over a range of \(g_\text{sc}\), inducing stomatal opening
-and closure by modulating humidity (see Materials and Methods for
-further details). We estimated `amphistomy advantage' (\aax)
-\emph{sensu} (\emph{20}), but with modifications previously described in
-(\emph{46}) and here (Materials and Methods). The native light intensity
-was represented by plant area index (PAI
-\(\unit{\meter\squared\per\meter\squared}\)), estimated using a global
-gridded X-m2 resolution data set derived from the Global Ecosystem
-Dynamics Investigation {[}GEDI; (\emph{52}){]} and georeferenced
-accession collection information from the Tomato Genetics Resource
-Center. The growth light intensities were 761 (sun treatment) and 115
-(shade treatment) while all other environment conditions were nearly
-identical (see \textbf{?@sec-methods} for supporting detail). The high
-and low measurements intensities were \ppfdequals{2000} (97.8:2.24
-red:blue) and \ppfdequals{150} (87.0:13.0 red:blue), respectively.
-
-\subsubsection{MARKER}\label{marker}
+contrasting light intensity (). We measured AA on 572 individual plants
+from 29 populations (average of 9.86 replicates per light treatment)
+using a recently developed method (\emph{46}). With this method, we
+directly compare the photosynthetic rate of an untreated amphistomatous
+leaf to that of the same leaf with gas exchange blocked through the
+adaxial (upper) surface by transparent plastic, which we refer to as
+`pseudohypostomy'. To compare amphi- and pseudohypostomatous leaves at
+identical whole-leaf \(g_\text{sc}\), we measure \(A\) over a range of
+\(g_\text{sc}\), inducing stomatal opening and closure by modulating
+humidity (see Materials and Methods for further details). We estimated
+`amphistomy advantage' (\aax) \emph{sensu} (\emph{20}), but with
+modifications previously described in (\emph{46}) and here (Materials
+and Methods). The native light intensity was represented by plant area
+index (PAI \(\unit{\meter\squared\per\meter\squared}\)), estimated using
+a global gridded X-m2 resolution data set derived from the Global
+Ecosystem Dynamics Investigation {[}GEDI; (\emph{52}){]} and
+georeferenced accession collection information from the Tomato Genetics
+Resource Center. The growth light intensities were \ppfdequals{761} (sun
+treatment) and \ppfdqty{115} (shade treatment) while all other
+environment conditions were nearly identical (see materials and methods
+in supplementary materials). The high and low measurements intensities
+were \ppfdequals{2000} (97.8:2.24 red:blue) and \ppfdequals{150}
+(87.0:13.0 red:blue), respectively.
 
 Consistent with biophysical theory of CO\(_2\) diffusion within leaves,
-AA \textgreater{} 0 for all accessions (Figure X). AA varied
+\(\aay{} > 0\) for all populations (Fig.~\ref{fig-aa}A). \aax{} varied
 substantially between measurement light intensities, growth light
-intensities, and among accessions (point to model comparison evidence).
-Measured under high light intensity, AA was consistently greater for sun
-plants. The average AA among accessions in the shade treatment was 0.039
-(range: 0.012--0.109; 21 of 29 accessions significant); however, the
-same accessions grown at high light intensity showed a mean AA of 0.050
-(range: 0.024--0.120; 25 of 29 accessions significant). Contrary to the
-predictions of the assimilatory hypothesis, AA was greater in all
-accessions under low measurement light intensity for both sun and shade
-grown plants. The effect of low light on AA was more pronounced in the
-sun-grown plants, where AA was significantly greater under low
-measurement light intensity in 28 of 29 accessions compared to to 11
-accessions for shade-grown plants. The overall average AA of shade and
-sun grown plants measured under low light intensity was 0.066 (range:
-0.019--0.143; 27 of 29 accessions significant) and 0.098 (range:
-0.051--0.175; 29 of 29 accessions significant), respectively. There was
-a slight tendency for accessions from more closed habitats (greater
-PAI), but AA varied widely among accessions from open habitats (low PAI)
-regardless of growth and measurement light intensities (Figure X). The
-pattern of AA across wild tomatoes strongly supports the plasticity
-hypothesis, argues against the acclimatory hypothesis, and provides only
-weak support for the constitutive hypothesis.
-
-Not sure where this goes: We infer this from the fact that blocking gas
-exchange in pseudohypostomatous leaves reduced \(A\) by X-X\% depending
-on the accession, light treatment, and light intensity (Table/figure).
-The AA is equivalent to an X-X\% change in total \(g_\text{sc}\) (see SI
-section gs equivalency). But whereas increasing \(g_\text{sc}\) would
-increase water loss as a necessary by-product, amphistomy can increase
-\(A\) without any appreciable affect on transpiration.
-
-These results suggest that developmental plasticity, as opposed to
-acclimation or adaptation to open habitats, may explain the explain the
-long-standing observation taht amphistomatous leaves are more common in
-sunny habitats. Our results add to existing explanations by showing high
-light intensity \emph{per se} does not increase the benefit of
-amphistomy, but rather than anatomical and biochemical change associated
-with higher light intensity modulate AA. We cannot yet attribute
-specific changes wrought by plasticity, but correlational evidence
-suggests that changes in leaf anatomy may be important. Species
-responded consistently by increasing stomatal density, especially on the
-adaxial surface, increasing stomatal density ratio. Sun leaves also had
-greater gs, Amax. However, these traits were not associated with change
-in AA (table/figures).
-
-from Mott and Oleary: ``However, Jones and Slatyer (5) reported a higher
-mesophyll resistance for CO, entering through the upper stomata than for
-the lower, and the data of Vaclavik (12) appear to support this
-conclusion''
-
-\section{Outtro - I think this goes at the
-end??}\label{outtro---i-think-this-goes-at-the-end}
-
-It is commonly assumed in comparative studies that most trait variation
-between species is constitutive and genetically determined, but our
-results suggest that plastic responses to light are necessary to explain
-why \gmaxratio{} increases with light habitat. Reaction norms to light
-intensity among wild tomatoes indicate that accessions from low light
-habitats express low \gmaxratio{} when grown in shade, whereas
-accessions from high light habitats have constitutively higher
-\gmaxratio{} (Fig. rxn-norm). The larger response to light intensity is
-driven by greater plasticty in leaf structure in accessions from
-low-light habitats (Fig/table - I haven't tested this yet, so this is a
-placeholder.) We infer that these structural changes in sun leaves are
-costly because they increase the diffusion path length for CO\(_2\) and
-reduce photosynthesis. As a consequence, sun leaves benefit more from
-amphistomy. This is the first demonstration that sun leaves benefit more
-from amphistomy because it compensates for developmental plasticity in
-leaf structure.
-
-Two outstanding areas for further research are 1) the specific
-anatomical changes, such as leaf thickness or mesophyll porosity, that
-increase the diffusion path length for CO\(_2\) in sun leaves, and 2)
-the costs of amphistomy that could explain plasticity in \gmaxratio{} in
-accessions from low-light habitats. Resolving the first question will
-require detailed anatomical measurements on a large number of shade and
-sun leaves with concomitant estimates of \aax. The fact that accessions
-from low-light habitats are more plastic suggests that costs of
-amphistomy are greatest in these environments. (not finished yet)
-
-Problems to bring up in discussion: - all our species were amphi, so
-they may all live in PAI below threshold to favor amphi - other
-microclimate diffs not considered - don't know specific anatomical
-change
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\setcounter{enumi}{7}
-\tightlist
-\item
-  What it all means
-\end{enumerate}
-
-\begin{itemize}
-\tightlist
-\item
-  Our study reveals an important role for plasticity in explaining why
-  amphi leaves are more common in sunny habitats
-\item
-  Confirms that leaf structure of sun leaves incurs costs that
-  amphistomy can mitigate
-\item
-  It will be useful to consider coordination between leaf anatomy,
-  stomatal density, and ratio in future studies of leaf
-  structure-function and in crops (or something)
-\end{itemize}
-
-\section{Materials and Methods
-{[}\#sec-methods{]}}\label{materials-and-methods-sec-methods}
-
-{[}this will be moved to SI eventually. note that much of it also the
-same in solanum-aq project, so might change to cite that paper?{]}
-
-(not sure where this goes) * We acclimated the focal leaf to high light
-(\ppfdequals{2000}) and high relative humidity (\rhequals{70}) until
-\(A\) and \gsw{} reach their maximum. After that, we decreased \rh{} to
-\(\approx 10\%\) to induce rapid stomatal closure without biochemical
-down
-
-\subsection{Accessions}\label{accessions}
-
-We compared AA among 29 ecologically diverse accessions of wild tomato,
-including representatives of all described species of \emph{Solanum}
-sect. \emph{Lycopersicon} and sect. \emph{Lycopersicoides} (\emph{51})
-and the cultivated tomato \emph{S. lycopersicum} var.
-\emph{lycopersicum} \autoref{tab:accessions}. Due to constraints on
-growth space and time, we spread out measurements over 61.1 weeks from
-August 29, 2022 to October 31, 2023. Replicates within accession were
-evenly spread out over this period to prevent confounding of temporal
-variation in growth conditions with accession. {[}anything else to say
-here? maybe explain accession selection and phylogeny?{]}
-
-\begin{longtable}{>{\raggedright\arraybackslash}p{3.25cm}lrrrr}
-
-\caption{\label{tbl-accessions}Accessions of \emph{Solanum} used in this
-study. The species name, accession number, collection latitude,
-longitude, elevation, and daily solar radiation estimated using the
-SPLASH algorithm (see `Climate data'). TGRC: Tomato Genetics Resource
-Center; PPFD: Photosynthetic Photon Flux Density.}
-
-\tabularnewline
-
-\toprule
-Species & TGRC accession & Latitude & Longitude & Elevation (mas) & $\mathrm{PAI}~(\unit{\meter\squared\per\meter\squared}$\\
-\midrule
-\em{\cellcolor{gray!10}{S. arcanum}} & \cellcolor{gray!10}{LA2172} & \cellcolor{gray!10}{-6.008} & \cellcolor{gray!10}{-78.858} & \cellcolor{gray!10}{662} & \cellcolor{gray!10}{0.7}\\
-\em{S. cheesmaniae} & LA0429 & -0.644 & -90.329 & 800 & 0.6\\
-\em{\cellcolor{gray!10}{S. cheesmaniae}} & \cellcolor{gray!10}{LA3124} & \cellcolor{gray!10}{-0.804} & \cellcolor{gray!10}{-90.042} & \cellcolor{gray!10}{1} & \cellcolor{gray!10}{0.4}\\
-\em{S. chilense} & LA1782 & -15.267 & -74.633 & 1000 & 0.3\\
-\em{\cellcolor{gray!10}{S. chilense}} & \cellcolor{gray!10}{LA4117A} & \cellcolor{gray!10}{-22.907} & \cellcolor{gray!10}{-67.941} & \cellcolor{gray!10}{3540} & \cellcolor{gray!10}{0.2}\\
-\addlinespace
-\em{S. chmielewskii} & LA1028 & -13.883 & -73.017 & 3000 & 0.4\\
-\em{\cellcolor{gray!10}{S. chmielewskii}} & \cellcolor{gray!10}{LA1316} & \cellcolor{gray!10}{-13.400} & \cellcolor{gray!10}{-73.906} & \cellcolor{gray!10}{2920} & \cellcolor{gray!10}{1.1}\\
-\em{S. corneliomulleri} & LA0107 & -13.117 & -76.383 & 60 & 0.0\\
-\em{\cellcolor{gray!10}{S. corneliomulleri}} & \cellcolor{gray!10}{LA0444} & \cellcolor{gray!10}{-13.433} & \cellcolor{gray!10}{-76.133} & \cellcolor{gray!10}{100} & \cellcolor{gray!10}{0.5}\\
-\em{S. galapagense} & LA0436 & -0.953 & -90.978 & 40 & 0.2\\
-\addlinespace
-\em{\cellcolor{gray!10}{S. galapagense}} & \cellcolor{gray!10}{LA1044} & \cellcolor{gray!10}{-0.284} & \cellcolor{gray!10}{-90.548} & \cellcolor{gray!10}{0} & \cellcolor{gray!10}{0.2}\\
-\em{S. habrochaites} & LA0407 & -2.181 & -79.884 & 70 & 0.5\\
-\em{\cellcolor{gray!10}{S. habrochaites}} & \cellcolor{gray!10}{LA1777} & \cellcolor{gray!10}{-9.550} & \cellcolor{gray!10}{-77.700} & \cellcolor{gray!10}{3216} & \cellcolor{gray!10}{0.4}\\
-\em{S. huaylasense} & LA1358 & -9.533 & -77.967 & 750 & 0.6\\
-\em{\cellcolor{gray!10}{S. huaylasense}} & \cellcolor{gray!10}{LA1360} & \cellcolor{gray!10}{-9.546} & \cellcolor{gray!10}{-77.929} & \cellcolor{gray!10}{1490} & \cellcolor{gray!10}{0.5}\\
-\addlinespace
-\em{S. huaylasense} & LA1364 & -10.133 & -77.383 & 2920 & 0.9\\
-\em{\cellcolor{gray!10}{S. lycopersicoides}} & \cellcolor{gray!10}{LA2951} & \cellcolor{gray!10}{-19.317} & \cellcolor{gray!10}{-69.450} & \cellcolor{gray!10}{2200} & \cellcolor{gray!10}{0.5}\\
-\em{S. lycopersicoides} & LA4126 & -19.287 & -69.396 & 3120 & 0.4\\
-\em{\cellcolor{gray!10}{S. neorickii}} & \cellcolor{gray!10}{LA1322} & \cellcolor{gray!10}{-13.483} & \cellcolor{gray!10}{-72.442} & \cellcolor{gray!10}{2380} & \cellcolor{gray!10}{0.7}\\
-\em{S. neorickii} & LA2133 & -3.400 & -79.183 & 1980 & 1.1\\
-\addlinespace
-\em{\cellcolor{gray!10}{S. pennellii}} & \cellcolor{gray!10}{LA0716} & \cellcolor{gray!10}{-16.225} & \cellcolor{gray!10}{-73.617} & \cellcolor{gray!10}{50} & \cellcolor{gray!10}{0.2}\\
-\em{S. pennellii} & LA0750 & -14.775 & -75.034 & 550 & 0.1\\
-\em{\cellcolor{gray!10}{S. pennellii}} & \cellcolor{gray!10}{LA3778} & \cellcolor{gray!10}{-14.775} & \cellcolor{gray!10}{-75.034} & \cellcolor{gray!10}{616} & \cellcolor{gray!10}{0.1}\\
-\em{S. peruvianum} & LA2744 & -18.550 & -70.150 & 400 & 0.2\\
-\em{\cellcolor{gray!10}{S. peruvianum}} & \cellcolor{gray!10}{LA2964} & \cellcolor{gray!10}{-18.028} & \cellcolor{gray!10}{-70.835} & \cellcolor{gray!10}{75} & \cellcolor{gray!10}{1.2}\\
-\addlinespace
-\em{S. pimpinellifolium} & LA1269 & -11.483 & -77.075 & 400 & 0.5\\
-\em{\cellcolor{gray!10}{S. pimpinellifolium}} & \cellcolor{gray!10}{LA1589} & \cellcolor{gray!10}{-8.433} & \cellcolor{gray!10}{-78.817} & \cellcolor{gray!10}{30} & \cellcolor{gray!10}{0.1}\\
-\em{S. pimpinellifolium} & LA2933 & -1.442 & -80.562 & 375 & 1.6\\
-\em{\cellcolor{gray!10}{S. sitiens}} & \cellcolor{gray!10}{LA4116} & \cellcolor{gray!10}{-22.159} & \cellcolor{gray!10}{-68.782} & \cellcolor{gray!10}{2960} & \cellcolor{gray!10}{0.1}\\
-\bottomrule
-
-\end{longtable}
-
-\subsection{Plant growth conditions}\label{plant-growth-conditions}
-
-In all growth spaces, we recorded \(\mathrm{PPFD}\) using full spectrum
-quantum sensors (SQ-500-SS, Apogee Instruments, Logan, Utah, USA); we
-recorded temperature, RH, and {[}CO\(_2\){]} using an EE894 sensor (E+E
-Elektronik, Engerwitzdorf, Austria) protected by a radiation shield. All
-environmental measurements were taken every 10 minutes from the middle
-of plants racks at approximately the same height as the leaves we
-measured. We measured leaf temperature of focal leaves prior to
-measurement using an infrared radiometer (SI-111-SS, Apogee Instruments,
-Logan, Utah, USA).
-
-\subsubsection{Germination and seedling
-stage}\label{germination-and-seedling-stage}
-
-Seeds provided by the Tomato Genetics Resource Center germinated on
-moist paper in plastic boxes after soaking for 30-60 minutes in a 50\%
-(volume per volume) solution of household bleach and water, followed by
-a thorough rinse. We transferred seedlings to cell-pack flats containing
-Pro-Mix BX potting mix (Premier Tech, Rivière-du-Loup, Quebec, Canada)
-once cotyledons fully emerged, typically within 1-2 weeks of sowing. We
-grew seeds and seedlings for both sun and shade treatments under the
-same environmental conditions (12:12 h,
-24.3:\(\qty{21.7}{\degreeCelsius}\), 49.6:58.4 RH day:night cycle). LED
-light provided
-\(\mathrm{PPFD} = \qty{267}{\micro\mol\raiseto{-2}\meter\raiseto{-1}\second}\)
-(Fluence RAZRx, Austin, Texas, USA).
-
-\subsubsection{Light treatments}\label{light-treatments}
-
-Seedlings were randomly assigned in alternating order within accession
-to the sun or shade treatment during transplanting. After seedlings
-established in cell-pack flats for \(\approx 2\) weeks, we transplanted
-them to 3.78 L plastic pots containing 60\% Pro-Mix BX potting mix, 20\%
-coral sand (Pro-Pak, Honolulu, Hawaiʻi, USA), and 20\% cinders (Niu
-Nursery, Honolulu, Hawaiʻi, USA). Percentage composition is on a volume
-basis. The soil mixture contained slow release NPK fertilizer following
-manufacturer instructions (Osmocote Smart-Release Plant Food Flower \&
-Vegetable, The Scotts Company, Marysville, Ohio, USA). We determined pot
-field capacity one week after transplanting using a scale (Ohaus V12P15
-Valor 1000, Parsippany, New Jersey, USA) and watered to field capacity
-three times per week to prevent drought stress.
-
-We assigned sun and shade treatment to lower and upper racks of a
-\(\qty{1.22}{\meter} \times \qty{2.44}{\meter}\) shelving unit in a
-climate-controlled growth room. We assigned the sun treatment to the
-lower rack to limit diffuse light from reaching the shade treatment. The
-average daytime \ppfd{} was \ppfdqty{761} and \ppfdqty{115} for sun and
-shade treatments, respectively. To isolate the effect of light intensity
-from quality, we used the same LED model with the the same spectrum
-(Fluence SPYDR 2i, Austin, Texas, USAS), but dimmed the lights in the
-shade treatment. To maintain homogeneous environmental conditions other
-than light, we mixed air within the growth room using an air circulator
-(Vornado 693DC, Andover, Kansas, USA) and within racks using a miniature
-oscillating air circulator (Vornado Atom 1, Andover, Kansas, USA).
-Despite these efforts, the air in the sun treatment was on average
-\(\qty{2.56}{\degreeCelsius}\) warmer and the average RH was
-consequently 5.75 lower. However, because of evaporative cooling, the
-leaves in the sun treatment were only \(\qty{0.886}{\degreeCelsius}\) on
-average (\(n = 699\) leaves).
-
-\subsection{Leaf trait measurements}\label{leaf-trait-measurements}
-
-We selected a fully expanded, unshaded leaf at least six leaves above
-the cotyledons during early vegetative growth. This typically meant that
-plants had grown in light treatments for \(\approx\) 4 weeks, ensuring
-they had time to sense and respond developmentally to the light
-intensity of the treatment rather than the seedling conditions
-(\emph{53}). Shade plants grew slower than sun plants, hence leaves at
-the same developmental stage were measured on chronologically older
-plants in the shade treatment. In some sun plants, we had to use leaves
-higher on the stem because short internodes made lower leaves
-inaccessible with the gas exchange equipment. We measured terminal
-leaflets in 82.6\% of cases, but used the lateral leaflet closest to the
-terminal leaflet when it was damaged or difficult to clamp into the gas
-exchange chamber. When a leaflet was damaged during gas exchange
-measurements, we collected anatomical data from the nearest leaflet on
-the same leaf (1.58)\%. We measured chlorophyll concentration index
-(CCI) using a chrolophyll concentration meter (MC-100, Apogee
-Instruments, Logan, Utah, USA) on the lamina of focal leaflets before
-gas exchange measurements at the same time we measured leaf temperature.
-
-\subsubsection{Amphistomy advantage}\label{amphistomy-advantage}
-
-We estimated `amphistomy advantage' (\aax) \emph{sensu} (\emph{20}), but
-with modifications previously described in (\emph{46}). \aax{} is
-calculated as the log-response ratio of \(A\) compared at the same total
-\gsw:
-
-\[\mathrm{AA} = \mathrm{log}(A_{\mathrm{amphi}} / A_{\mathrm{hypo}})\]
-
-We measured the photosynthetic rate of an untreated amphistomatous leaf
-(\Aamphi) over a range of \gsw{} values. We refer to this as an
-\agcurve{} curve. We compared the \agcurve{} curve of the untreated leaf
-to the photosynthetic rate of pseudohypostomatous leaf (\Ahypo), which
-is the same leaf but with gas exchange through the upper surface blocked
-by a neutral density plastic (propafilm).
-
-We measured \agcurve{} curves using a portable infrared gas analyzer
-(LI-6800PF, LI-COR Biosciences, Lincoln, Nebraska, USA).
-Light-acclimated plants were placed under LEDs dimmed to match their
-light treatment during gas exchange measurements. We estimated the
-photosynthetic rate (\(A\)) and stomatal conductance to CO\(_2\) (\gsw)
-at ambient CO\(_2\) (\caequals{415}) and \tleafequals{25.0}. The
-irradiance of the light source in the pseudohypo leaf was higher because
-the propafilm reduces transmission. To compensate for reduced
-transmission, we increased incident \ppfd for pseudohypo leaves by a
-factor 1/0.91, the inverse of the measured transmissivity of the
-propafilm. We also set the stomatal conductance ratio, for purposes of
-calculating boundary layer conductance, to 0 for pseudohypo leaves
-following manufacturer directions.
-
-We collected four \agcurve{} curves per leaf, an amphi (untreated) curve
-and a pseudohypo (treated) curve at high light-intensity
-(\ppfdequals{2000}; 97.8:2.24 red:blue) and low light-intensity
-(\ppfdequals{150}; 87.0:13.0 red:blue). We always measured high
-light-intensity curves first because photosynthetic downregulation is
-faster than upregulation in these species. To control for order effects,
-we alternated between starting with amphi or pseudohypo leaf
-measurements. Unlike (\emph{46}), preliminary experiments with
-\emph{Solanum} indicated a strong order effect in that \(A\) declined in
-the second curve. Therefore, we made measurements over two days. On the
-first day, we measured high and low light-intensity curves for either
-amphi or pseudohypo leaves; on the second day, we measured high and low
-light-intensity curves on the other leaf type.
-
-In all cases, we acclimated the focal leaf to high light
-(\ppfdequals{2000}) and high relative humidity (\rhequals{70}) until
-\(A\) and \gsw{} reach their maximum. After that, we decreased \rh{} to
-\(\approx 10\%\) to induce rapid stomatal closure without biochemical
-downregulation. Hence, \Aamphi{} and \Ahypo{} were both measured at low
-chamber humidity after the leaf had acclimated to high humidity. All
-other environmental conditions in the leaf chamber remained the same. We
-logged data until \gsw{} reached its nadir. We then acclimated the leaf
-to low light (\ppfdequals{150}) and \rhequals{70} before inducing
-stomatal closure with low \rh and logging data as described above.
-
-\subsubsection{\texorpdfstring{Light-response (\aqcurve)
-curves}{Light-response () curves}}\label{light-response-curves}
-
-In 91.3\% of plants, we measured light-response (\aqcurve) curves on the
-same leaflets as \agcurve{} curves. However, when a leaflet was damaged
-during \agcurve{} curves, we used the next closest leaflet for
-\aqcurve{} curves. Leaves acclimated to high light-intensity
-(\ppfdequals{2000}), ambient CO\(_2\) (\caequals{415}), \rhequals{50},
-and \tleafequals{25}. After \(A\) and \gsw{} stabilized, we measured
-\(A\) at 20 light-intensity levels between \(0\) and
-\(\qty{2000}{\micro\mol\raiseto{-2}\meter\raiseto{-1}\second}\) in
-descending order.
-
-\subsubsection{Stomatal anatomy}\label{stomatal-anatomy}
-
-We estimated the stomatal density and size on ab- and adaxial leaf
-surfaces from all leaves, using guard cell length as a proxy for
-stomatal size since it proportional to maximum conductance (\emph{54}).
-We made surface impressions of leaf lamina from the same area used for
-gas exchange measurements using a-silicone impression material (Zhermack
-elite HD+, light body, fast set, Rovigo, Italy). We applied clear nail
-polish to make positive replicas of the impression. After nail polish
-dried, we mounted replicas on a microscope slide using transparent tape
-(\emph{55}). We digitized a portion of each leaf surface replica using a
-brightfield microscope (Leica DM2000, Wetzlar, Germany). We counted and
-measured guard cell length on all stomata using the FIJI implementation
-of ImageJ2 version 2.3.0 (\emph{56}), then divided the count by the
-visible leaf area (\(\qty{0.890}{\raiseto{2}\milli\meter}\)) to estimate
-stomatal density.
-
-\subsubsection{Leaf mass per area}\label{leaf-mass-per-area}
-
-Leaf mass per area (LMA) is the dry mass divided by the leaflet area. We
-scanned fresh leaflets on a flat bed scanner (Epson V600, Los Alamitos,
-California. USA) and measured leaflet area from digital images using the
-FIJI implementation of ImageJ2 version 2.3.0 (\emph{56}). We dried
-leaves for 72 hours at \(\qty{74}{\degreeCelsius}\) in a food dehydrator
-(Cosori CP267-FD, Vesync Co., Anaheim, California, USA) and weighed
-using a benchtop analytical balance (Ohaus PR64 Analytical Balance,
-Parsippany, New Jersey, USA). In \(10.5\%\) we measured LMA on the
-adjacent leaflet because the focal leaflet was damaged or wilted while
-making surface impressions and we could not reliably estimate area. LMA
-data are missing from \(3.32\%\) of individuals because the area or mass
-was not recorded at all or recorded incorrectly.
-
-\subsection{\texorpdfstring{Cleaning \agcurve{}
-curves}{Cleaning  curves}}\label{cleaning-curves}
-
-The raw data set consisted of 2,370 \agcurve{} curves with an average of
-63.2 points per curve. Manual curation of a data set this size in a
-principled, consistent manner is not feasible. Therefore, we automated
-data cleaning using custom \emph{R} scripts. Cleaning is divided into
-six sequential steps (\autoref{tbl-cleaning1}).
-
-\begin{table}
-
-\caption{\label{tbl-cleaning1}Six sequential steps for cleaning
-\agcurve{} curves. The rationale and procedure for each step are
-described in the text. The rightmost columns summarize the number of
-curves and mean number of points per curve remaining after each step.
-For reference, there are four possible \agcurve{} curves per replicate:
-all combinations of leaf type (amphi or pseudohypo) and light intensity
-(high or low).}
+intensities, and among populations (point to model comparison evidence).
+Measured under high light intensity, \aax{} was consistently greater for
+sun plants. The average \aax{} among populations in the shade treatment
+was 0.039 (range: 0.012--0.109; 21 of 29 populations significant);
+however, the same populations grown at high light intensity showed a
+mean \aax{} of 0.050 (range: 0.024--0.120; 25 of 29 populations
+significant). Contrary to the predictions of the assimilatory
+hypothesis, \aax{} was greater in all populations under low measurement
+light intensity for both sun and shade grown plants. The effect of low
+light on \aax{} was more pronounced in the sun-grown plants, where
+\aax{} was significantly greater under low measurement light intensity
+in 28 of 29 populations compared to to 11 populations for shade-grown
+plants. The overall average \aax{} of shade and sun grown plants
+measured under low light intensity was 0.066 (range: 0.019--0.143; 27 of
+29 populations significant) and 0.098 (range: 0.051--0.175; 29 of 29
+populations significant), respectively. There was a slight tendency for
+populations from more closed habitats (greater PAI), but \aax{} varied
+widely among populations from open habitats (low PAI) regardless of
+growth and measurement light intensities (Fig.~\ref{fig-aa}B).
+
+\begin{figure}
 
@@
 
-\centering
-\begin{tabular}{>{\raggedright\arraybackslash}p{9cm}>{\centering\arraybackslash}p{2.5cm}>{\centering\arraybackslash}p{2.5cm}}
-\toprule
-Step: description & Number of curves & Number of points per curve\\
-\midrule
-\cellcolor{gray!10}{1. remove unreliable and unusable data points} & \cellcolor{gray!10}{2,361} & \cellcolor{gray!10}{63.0}\\
-2. remove hysteretic portion of \agcurve{} curves at low \gsw{} & 2,360 & 59.2\\
-\cellcolor{gray!10}{3. remove outliers within each \agcurve{} curve} & \cellcolor{gray!10}{2,360} & \cellcolor{gray!10}{58.7}\\
-4. remove replicates with no overlap between amphi and pseudohypo \agcurve{} curves & 2,268 & 58.5\\
-\cellcolor{gray!10}{5. thin redundant data points within each \agcurve{} curve} & \cellcolor{gray!10}{2,268} & \cellcolor{gray!10}{28.1}\\
-\addlinespace
-6. trim extreme \aax{} values & 2,214 & 28.1\\
-\bottomrule
-\end{tabular}
+\pandocbounded{\includegraphics[keepaspectratio]{ms_files/figure-pdf/fig-aa-1.pdf}}
 
@@
 
-\end{table}%
-
-\subsubsection{Remove unreliable and unusable data
-points}\label{remove-unreliable-and-unusable-data-points}
-
-\emph{Rationale}: Unreliable data points consisted of those where
-chamber {[}CO\(_2\){]} was unstable and therefore measurements are not
-biologically meaningful. Unusable data points were those where \(A < 0\)
-because the logarithm of a negative number is undefined.
-
-\emph{Procedure}: We retained data points where \cabetween{410}{420} and
-\(A> 0\).
-
-\subsubsection{\texorpdfstring{Remove hysteretic portion of \agcurve{}
-curves at low
-\gsw{}}{Remove hysteretic portion of  curves at low }}\label{remove-hysteretic-portion-of-curves-at-low}
-
-\emph{Rationale}: In most \agcurve{} curves, we observed a hysteretic
-response at low \gsw. After \gsw{} and \(A\) declined simultaneously,
-\(A\) increased slightly as \gsw{} continued to decline or stabilize,
-indicating some leaf acclimation to low \rh. We removed this portion of
-the curve to focus curve-fitting on the primary domain where \(A\)
-increases monotonically with \gsw{}.
-
-\emph{Procedure}: For each curve, we removed data points after \gsw{}
-had reached its minimum unless there were fewer than 10 data points
-remaining.
-
-\subsubsection{\texorpdfstring{Remove outliers within each \agcurve{}
-curve}{Remove outliers within each  curve}}\label{remove-outliers-within-each-curve}
-
-\emph{Rationale}: Individual outliers within \agcurve{} curves, usually
-caused by transitory changes in chamber conditions, exert undue leverage
-on parameter estimates and cause bias and/or low precision in parameter
-estimates.
-
-\emph{Procedure}: We fit provisional quadratic regressions for each
-curve using ordinary least squares with the \texttt{lm()} function in
-\emph{R}. We sequentially removed data points with an absolute external
-studentized residual \(> 3\) until none remained.
-
-\subsubsection{\texorpdfstring{Thin redundant data points within each
-\agcurve{}
-curve}{Thin redundant data points within each  curve}}\label{thin-redundant-data-points-within-each-curve}
-
-\emph{Rationale}: Data points closely spaced along the \agcurve{} curve
-provide redundant information and may be highly correlated
-(i.e.~pseudoreplication). This occurred because data was logged at a
-constant temporal interval, but the rate at which \gsw{} declined was
-not constant. Thinning reduces parameter estimation bias toward densely
-sampled regions of the curve which may not be the most biologically
-informative.
-
-\emph{Procedure}: We retained the maxima and minima \gsw{} for each
-curve and thinned all but one point per thinning interval of
-\(0.05~\log(\si{\mol\raiseto{-2}\meter\raiseto{-1}\second})\), retaining
-the point nearest the midpoint of the interval.
-
-\subsubsection{\texorpdfstring{Remove replicates with no overlap between
-amphi and pseudohypo \agcurve{}
-curves}{Remove replicates with no overlap between amphi and pseudohypo  curves}}\label{remove-replicates-with-no-overlap-between-amphi-and-pseudohypo-curves}
-
-\emph{Rationale}: We could not estimate \aax{} for replicates where
-amphi and pseudohypo \agcurve{} curves did not overlap.
-
-\emph{Procedure}: We removed replicates where the range of \gsw{} values
-for amphi and pseudohypo \agcurve{} curves did not overlap.
-
-\subsubsection{\texorpdfstring{Trim extreme \aax{}
-values}{Trim extreme  values}}\label{trim-extreme-values}
-
-\emph{Rationale}: Extreme \aax{} values were likely due to measurement
-error or leaf damage. Since amphi and pseudohypo \agcurve{} curves are
-measured on consecutive days, a poor calibration or a damaged leaf could
-cause a large difference in \(A\) between days, which would appear as an
-extreme \aax{} value.
-
-\emph{Procedure}: We provionsally estimated \aax{} for each replicate by
-integrating over the range of \gsw{} values where amphi and pseudohypo
-\agcurve{} curves overlap. In this procedure, curve parameters were
-provisionally estimated using ordinary least squares with the
-\texttt{lm()} function in \emph{R}. We then used point estimates of
-\aax{} for each replicate as the response variable in a linear model
-with light treatment, light intensity, accession, and all interactions
-as explanatory variables. This model was also fit using ordinary least
-squares with the \texttt{lm()} function in \emph{R}. We classified
-extreme \aax{} values as those with an absolute internal studentized
-residual \(> 3\). Because these values likely indicate significant
-measurement error or leaf damage, we removed \agcurve{} curves at both
-light intensities if either was classified as extreme.
-
-\subsection{\texorpdfstring{Cleaning \aqcurve{}
-curves}{Cleaning  curves}}\label{cleaning-curves-1}
-
-The raw data set consisted of 658 \aqcurve{} curves with an average of
-19.4. Manual curation of a data set this size in a principled,
-consistent manner is not feasible. Therefore, we automated data cleaning
-using custom \emph{R} scripts. Cleaning is divided into two sequential
-steps (\autoref{tbl-cleaning2}).
-
-\begin{table}
-
-\caption{\label{tbl-cleaning2}Two sequential steps for cleaning
-\agcurve{} curves. The rationale and procedure for each step are
-described in the text. The rightmost columns summarize the number of
-curves and mean number of points per curve remaining after each step.}
+\caption{\label{fig-aa}Caption on following page.}
+
+\end{figure}%
+
+\addtocounter{figure}{-1}
+
+\begin{figure}
 
@@
 
-\centering
-\begin{tabular}{>{\raggedright\arraybackslash}p{9cm}>{\centering\arraybackslash}p{2.5cm}>{\centering\arraybackslash}p{2.5cm}}
-\toprule
-Step: description & Number of curves & Number of points per curve\\
-\midrule
-\cellcolor{gray!10}{1. remove outliers within each \aqcurve{} curve} & \cellcolor{gray!10}{658} & \cellcolor{gray!10}{19.1}\\
-2. remove \aqcurve{} curves with poor fit & 652 & 19.1\\
-\bottomrule
-\end{tabular}
+\pandocbounded{\includegraphics[keepaspectratio]{ms_files/figure-pdf/fig-aacap-1.pdf}}
 
@@
 
-\end{table}%
-
-\subsubsection{\texorpdfstring{Remove outliers within each \aqcurve{}
-curve}{Remove outliers within each  curve}}\label{remove-outliers-within-each-curve-1}
-
-\emph{Rationale}: Individual outliers within \agcurve{} curves, usually
-caused by transitory changes in chamber conditions, exert undue leverage
-on parameter estimates and cause bias and/or low precision in parameter
-estimates.
-
-\emph{Procedure}: We fit provisional nonrectangular hyperbola
-(\emph{57}) to each \aqcurve{} curve using nonlinear regression with the
-\texttt{nlsLM()} function from the \emph{R} package \textbf{minpack.lm}
-version 1.2.4 (\emph{58}). We sequentially removed data points with an
-absolute external studentized residual \(> 3\) until none remained.
-
-\subsubsection{\texorpdfstring{Remove \aqcurve{} curves with poor
-fit}{Remove  curves with poor fit}}\label{remove-curves-with-poor-fit}
-
-\emph{Rationale}: \aqcurve{} curves with a poor fit to the
-nonrectangular hyperbola most likely indicate systematic measurement
-error and/or the leaf was not fully acclimated to the chamber
-environment.
-
-\emph{Procedure}: As described above, we fit provisional nonrectangular
-hyperbola to each \aqcurve{} curve and calculated the model \(r^2\).
-There was a clear break between typical curves and poorly fitting curves
-where \(r^2 < 0.99\). We therefore removed \aqcurve{} curves with
-\(r^2 < 0.99\).
-
-\subsection{\texorpdfstring{Bayesian data analysis in
-\emph{Stan}}{Bayesian data analysis in Stan}}\label{bayesian-data-analysis-in-stan}
-
-We fit five models to test predictions of competing hypotheses about why
-amphistomy advantage (\aax) might be greater for leaves in sunny, open
-habitats. This section provides an overview of differences among models
-(\autoref{tbl-models}). The next sections describe how we fit models in
-\emph{Stan}, all model parameters and priors, and specific predictions
-about parameter values for each hypothesis.
-
-\begin{table}
-
-\caption{\label{tbl-models}Summary of differences among competing models
-of how \aax{} varies with light intensity, light treatment, and among
-accessions as a function of native \ppfd. The models are numbered from
-simpler to more complex. All models include fixed effects of light
-intensity and light treatment; some models include interactions between.
-All models include a phylogenetic random effect of accession on \aax;
-some models include varying effects of light intensity and light
-treatment among accessions. The last column indicates the
-accession-level \aax{} variable we used as a response to native \ppfd.}
+\caption{\label{fig-aacap}\textbf{Amphistomy advantage (\aax{}) in wild tomatoes is amplified under simulated sunny growth conditions.}
+\aax{} (\(y\)-axis) is the log-response ratio of photosynthesis in an
+amphistomatous leaf compared to an otherwise identical
+pseudohypostomatous leaf. In both panels, estimates are shown for plants
+grown under simulated shade or sun and measured under high light
+intensity (\ppfdequals{2000}) and low light intensity
+(\ppfdequals{150}). The dashed line at zero indicates no difference in
+photosynthesis between amphi- and pseudohypostomatous leaves.
+(\textbf{A}) The points are the posterior median \aax{} for each
+population, with error bars showing 95\% confidence intervals. Within
+each facet, the populations are arranged by \aax{} estimated in that
+growth and measurement condition. (\textbf{B}) The same estimates of
+\aax{} in each combination of growth and measurement light intensity
+ploted against native plant area index (PAI; \(x\)-axis, log-scale). The
+confidence intervals were omitted for visual clarity.}
+
+\end{figure}%
+
+The pattern of \aax{} across wild tomatoes strongly supports the
+plasticity hypothesis, argues against the acclimatory hypothesis, and
+provides only weak support for the constitutive hypothesis. Plastic
+changes in leaf thickness and/or packing density, summarized by the bulk
+leaf mass per area (LMA), likely mediate the effect of growth light
+intensity on \aax{}. LMA increased in sun grown plants in all
+populations by an average of 123\% {[}95\% CI: 50.5 to 245\%{]}, similar
+to plastic responses in many species (\emph{48}). When LMA is included
+in the model, the effect of growth light intensity on \aax{} is no
+longer significant based on model comparison (Table S8) and LMA is
+weakly positively correlated with \aax{} (Fig.~\ref{fig-lma_aa}). In
+contrast, LMA did not mediate an indirect effect of native PAI on \aax{}
+(Table S8). Many anatomical traits underlie LMA (\emph{53}) and future
+research will be needed identify which particular traits, such as leaf
+thickness or mesophyll porosity, are responsible for mediating \aax{}.
+The fact that the \aax{} of sun plants was greater under low measurement
+light intensity supports a long-standing hypothesis that resistance to
+CO\(_2\) diffusion is greater in the upper than lower portions of the
+leaf interior (\emph{54}). At low light intensity, the bulk properties
+of the leaf are weighed toward the upper palisade where most light is
+intercepted. Hence, amphistomy may, expectedly, be particularly
+beneficial for sun leaves experiencing intermittant shade or cloud
+cover. Our study is limited in testing this because we could not
+directly measure the stomatal conductance ratio and intercellular
+resistance on each surface. Future experiments measuring \aax{} with a
+dual sided chamber (\emph{31}, \emph{33}) can overcome these
+limitations.
+
+\begin{figure}
 
@@
 
-\begin{tabular}{>{\centering\arraybackslash}p{0.3in}>{\raggedright\arraybackslash}p{1.3in}>{\raggedright\arraybackslash}p{3.4in}>{\raggedright\arraybackslash}p{1in}}
-\toprule
-Model & Fixed effects & Phylogenetic random effects & Response to native \ppfd\\
-\midrule
-1 & light intensity & varying intercept among accessions & $\mathrm{AA}_{0, \text{acc}}$\\
-
- & light treatment &  & \\
-\cmidrule{1-4}
-2 & light intensity & varying intercept among accessions & $\mathrm{AA}_{0, \text{acc}}$\\
-
- & light treatment &  & \\
-
- & intensity $\times$ treatment &  & \\
-\cmidrule{1-4}
-3 & light intensity & varying intercept among accessions & $\mathrm{AA}_{2000, \text{acc}}$\\
-
- & light treatment & varying effect of high light intensity among accessions & \\
-\cmidrule{1-4}
-4 & light intensity & varying intercept among accessions & $\mathrm{AA}_{\text{sun, acc}}$\\
-
- & light treatment & varying effect of sun treatment among accessions & \\
-\cmidrule{1-4}
-5 & light intensity & varying intercept among accessions & $\mathrm{AA}_{2000, \text{sun, acc}}$\\
-
- & light treatment & varying effect of high light intensity among accessions & \\
-
- & intensity $\times$ treatment & varying effect of sun treatment among accessions & \\
-\bottomrule
-\end{tabular}
+\pandocbounded{\includegraphics[keepaspectratio]{ms_files/figure-pdf/fig-lma_aa-1.pdf}}
 
@@
 
-\end{table}%
-
-\subsubsection{\texorpdfstring{Fitting models in
-\emph{Stan}}{Fitting models in Stan}}\label{sec-fitting}
-
-We fit Bayesian models with MCMC sampling in the probabilistic
-programming language \emph{Stan} (\emph{59}) using the \emph{R} package
-\textbf{brms} version 2.22.0 (\emph{60}). We used CmdStan version 2.36.0
-and \textbf{cmdstanr} version 0.9.0 (\emph{61}) to interface with
-\emph{R} version 4.5.0 (\emph{62}). We sampled the posterior
-distribution from 1 chains with 1000 iterations each after 1000 warmup
-iterations per chain. We estimated parameters and confidence intervals
-as the median and 95\% quantile intervals of the posterior,
-respectively. We chose the number of chains, warmup and sampling
-iterations, and maximum treedepth so that parameter estimates converged
-(\(\hat{R} < 1.01\) (\emph{63})) and the effective sample size (ESS) for
-each parameter was \(> 10^3\).
-
-\subsubsection{Parameter estimation and priors}\label{sec-parameters}
-
-There were four levels of parameter estimation in our analysis:
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\tightlist
-\item
-  Estimate \agcurve{} curve parameters
-\item
-  Estimate \aax{} for each light intensity with leaf using \agcurve{}
-  curve parameters
-\item
-  Estimate the effects of light intensity, light treatment, and
-  accession on \aax{} (assimilatory and plasticity hypotheses)
-\item
-  Estimate the effects of native light habitat on accession-level \aax{}
-  (constitutive hypothesis)
-\end{enumerate}
-
-Although the higher-level parameter estimates depend on the lower-level
-parameter estimates, we fit all models simultaneously to ensure that the
-uncertainty in lower-level estimates propagated to higher-level
-estimates.
-
-\begin{longtable}{>{\raggedright\arraybackslash}p{1in}>{\raggedright\arraybackslash}p{5in}}
-
-\caption{\label{tbl-parameters}Description of parameters estimated in
-the hierarchical Bayesian model. The \texttt{Parameter} column lists the
-parameter name as it appears in text. The \texttt{Description} column
-provides a brief description of the parameter.}
-
-\tabularnewline
-
-\toprule
-Parameter & Description\\
-\midrule
-\addlinespace[0.3em]
-\multicolumn{2}{l}{\textbf{\agcurve{} curve parameters}}\\
-\hspace{1em}\cellcolor{gray!10}{$\mathbf{B}_\text{curve}$} & \cellcolor{gray!10}{$n_\text{curve} \times 3$ array of random \agcurve{} curve-level coefficients ($b_{0,j}$, $b_{1,j}$, $b_{2,j}$); $\mathbf{B}_\text{curve} \sim \text{MVN}(\vec{0}, \symbf{\Sigma}_\text{curve})$}\\
-\hspace{1em}$\vec{\beta}_\text{curve}$ & vector of mean quadratic coefficients ($\beta_0$, $\beta_1$, $\beta_2$)\\
-\hspace{1em}\cellcolor{gray!10}{$\symbf{\Sigma}_\text{curve}$} & \cellcolor{gray!10}{$3 \times 3$ covariance matrix of curve-level coefficients}\\
-\hspace{1em}$\sigma_{\qty{6}{\raiseto{2}\centi\meter},\epsilon}$ & minimum residual standard deviation when the measured leaf surface area is $\qty{6}{\raiseto{2}\centi\meter}$\\
-\hspace{1em}\cellcolor{gray!10}{$\beta_{S,\epsilon}$} & \cellcolor{gray!10}{slope of the relationship between residual standard deviation and measured leaf surface area (log-link scale)}\\
-\hspace{1em}$\rho_{\epsilon}$ & lag-1 residual autocorrelation\\
-\addlinespace[0.3em]
-\multicolumn{2}{l}{\textbf{\aax{} for each light intensity with leaf using agcurve{} curve parameters}}\\
-\hspace{1em}\cellcolor{gray!10}{$\widehat{\mathrm{AA}}_{klm}$} & \cellcolor{gray!10}{estimate of \aax{} for the $k^{\text{th}}$ leaf at light intensity $l$ in accession $m$}\\
-\addlinespace[0.3em]
-\multicolumn{2}{l}{\textbf{effects of light intensity, light treatments, and accession on \aax}}\\
-\hspace{1em}$\beta_{\mathrm{AA},0}$ & intercept of \aax{} at low light intensity in shade treatment\\
-\hspace{1em}\cellcolor{gray!10}{$\beta_{\mathrm{AA},2000}$} & \cellcolor{gray!10}{effect of high light intensity at \ppfdequals{2000} on \aax}\\
-\hspace{1em}$\beta_{\mathrm{AA},\text{sun}}$ & effect of sun treatment on \aax{}\\
-\hspace{1em}\cellcolor{gray!10}{$\beta_{\mathrm{AA},2000,\text{sun}}$} & \cellcolor{gray!10}{effect of high light intensity at \ppfdequals{2000} on \aax{} in sun treatment}\\
-\hspace{1em}$\vec{\beta}_{\mathrm{AA},\text{acc}}$ & vector of $n_\text{acc}$ phylogenetically structured random accession-level effects on \aax; $\vec{\beta}_{\mathrm{AA},\text{acc}} \sim \text{MVN}(\vec{0}, \symbf{\Sigma}_{\text{AA,acc}})$\\
-\hspace{1em}\cellcolor{gray!10}{$\symbf{\Sigma}_{\text{AA,acc}}$} & \cellcolor{gray!10}{$n_\text{acc} \times n_\text{acc}$ covariance matrix of phylogenetically structured random accession-level effects on \aax;}\\
-\hspace{1em}$\vec{\beta}_{\mathrm{AA},\text{rep}}$ & vector of $n_\text{rep}$ random replicate-level effects on \aax; $\vec{\beta}_{\mathrm{AA},\text{rep}} \sim \text{Normal}(0, \sigma_{\mathrm{AA},\text{rep}})$\\
-\hspace{1em}\cellcolor{gray!10}{$\vec{\beta}_{\mathrm{AA},\text{2000,acc}}$} & \cellcolor{gray!10}{vector of $n_\text{acc}$ phylogenetically structured random accession-level effects of high light intensity at \ppfdequals{2000} on \aax; $\vec{\beta}_{\mathrm{AA},\text{2000,acc}} \sim \text{MVN}(\vec{0}, \symbf{\Sigma}_{\mathrm{AA},\text{2000,acc}})$}\\
-\hspace{1em}$\vec{\beta}_{\mathrm{AA},\text{sun,acc}}$ & vector of $n_\text{acc}$ phylogenetically structured random accession-level effects of sun treatment on \aax; $\vec{\beta}_{\mathrm{AA},\text{sun,acc}} \sim \text{MVN}(\vec{0}, \symbf{\Sigma}_{\mathrm{AA},\text{sun,acc}})$\\
-\hspace{1em}\cellcolor{gray!10}{$\sigma_{\mathrm{AA},\epsilon,0}$} & \cellcolor{gray!10}{intercept of phylogenetically unstructured residual standard deviation of \aax}\\
-\hspace{1em}$\beta_{\mathrm{AA},\epsilon,2000}$ & effect of high light intensity at \ppfdequals{2000} on phylogenetically unstructured residual standard deviation of \aax{} (log-link scale)\\
-\hspace{1em}\cellcolor{gray!10}{$\beta_{\mathrm{AA},\epsilon,\text{sun}}$} & \cellcolor{gray!10}{effect of sun treatment on phylogenetically unstructured residual standard deviation of \aax{} (log-link scale)}\\
-\hspace{1em}$\sigma_{\mathrm{AA},\text{rep}}$ & standard deviation of random replicate-level effects on \aax\\
-\hspace{1em}\cellcolor{gray!10}{$\alpha_{\mathrm{AA},\text{acc}}$} & \cellcolor{gray!10}{decay rate of phylogenetic covariance in random accession-level effects on \aax}\\
-\hspace{1em}$\sigma^2_{\mathrm{AA},\text{acc}}$ & phylogenetic diffusion rate in random accession-level effects on \aax\\
-\hspace{1em}\cellcolor{gray!10}{$\alpha_{\mathrm{AA},\text{2000,acc}}$} & \cellcolor{gray!10}{decay rate of phylogenetic covariance in random accession-level effects of high light intensity at \ppfdequals{2000} on \aax}\\
-\hspace{1em}$\sigma^2_{\mathrm{AA},\text{2000,acc}}$ & phylogenetic diffusion rate in random accession-level effects of high light intensity at \ppfdequals{2000} on \aax\\
-\hspace{1em}\cellcolor{gray!10}{$\alpha_{\mathrm{AA},\text{sun,acc}}$} & \cellcolor{gray!10}{decay rate of phylogenetic covariance in random accession-level effects of sun treatment on \aax}\\
-\hspace{1em}$\sigma^2_{\mathrm{AA},\text{sun,acc}}$ & phylogenetic diffusion rate in random accession-level effects of sun treatment on \aax\\
-\addlinespace[0.3em]
-\multicolumn{2}{l}{\textbf{effects of native light habitat on accession-level \aax}}\\
-\hspace{1em}\hspace{1em}\cellcolor{gray!10}{$\beta_{\mathrm{AA},\text{PPFD},0}$} & \cellcolor{gray!10}{intercept of accession-level \aax{} when native \ppfdequals{0}}\\
-\hspace{1em}\hspace{1em}$\beta_{\mathrm{AA},\text{PPFD},1}$ & slope of native \ppfd{} on accession-level \aax{}\\
-\hspace{1em}\cellcolor{gray!10}{$\alpha_{\mathrm{AA},\text{PPFD}}$} & \cellcolor{gray!10}{decay rate of phylogenetic covariance in residuals of model testing effect of native \ppfd{} on accession-level \aax}\\
-\hspace{1em}$\sigma^2_{\mathrm{AA},\text{PPFD}}$ & phylogenetic diffusion rate in residuals of model testing effect of native \ppfd{} on accession-level \aax\\
-\bottomrule
-
-\end{longtable}
-
-\paragraph{\texorpdfstring{\agcurve{} curve
-parameters}{ curve parameters}}\label{curve-parameters}
-
-We modeled \logA as a quadratic function of \loggsw for each leaf using
-the following equation:
-
-\[\log(A_{ij}) = (\beta_0 + b_{0,j}) + (\beta_1 + b_{1,j}) \log{g_{\text{sw},i}} + (\beta_2 + b_{2,j}) \log{g_{\text{sw},i}}^2 + \epsilon_{i}\]
-
-where \(\beta_0\), \(\beta_1\), and \(\beta_2\) are the average
-intercept, linear, and quadratic coefficients, respectively. We used
-diffuse normal priors with mean \(0\) and standard deviation \(10\) on
-these parameters. We estimated random effects of curve \(j\) on the
-intercept (\(b_{0,j}\)), linear (\(b_{1,j}\)), and quadratic
-(\(b_{2,j}\)) coefficients. We assumed that the \(j \times 3\) array of
-coefficients was multivariate normal a mean vector of \(\vec{0}\) and
-covariance \(\Sigma_\text{curve}\). We used a weakly informative normal
-prior with mean \(0\) and standard deviation \(1\) on the
-log-transformed standard deviations (i.e.~the diagonal of
-\(\Sigma_\text{curve}\)). We used a weakly informative
-\(\mathrm{LJK}(2)\) prior on the correlation matrix. The off diagonal
-elements of \(\Sigma_\text{curve}\) can be calculated from its diagonal
-elements and the correlation matrix.
-
-The residuals \(\epsilon_{i}\) were modeled as a lag-1 autocorrelated
-time-series. We further assumed that the residual standard deviation of
-the \(j^{\text{th}}\) curve (\(\sigma_{\epsilon,j}\)) was inversely
-proportional to the leaf surface area (\(S_j\)) within the chamber:
-
-\[\log (\sigma_{\epsilon,j}) = \log (\sigma_{\qty{6}{\raiseto{2}\centi\meter},\epsilon}) + (6 - S_j) \beta_{S,\epsilon}\]
-where \(\sigma_{\qty{6}{\raiseto{2}\centi\meter},\epsilon}\) is the
-minimum residual standard deviation when the
-\(\qty{6}{\raiseto{2}\centi\meter}\) chamber is completely filled. The
-residual standard deviation increases on log-linear scale by
-\(\beta_{S,\epsilon}\). We used a weakly informative normal prior with
-mean \(-3\) and standard deviation \(5\) on
-\(\log (\sigma_{\qty{6}{\raiseto{2}\centi\meter},\epsilon})\) and a
-weakly informative normal prior with mean \(0\) and standard deviation
-\(1\) on \(\beta_{S,\epsilon}\).
-
-\paragraph{\texorpdfstring{\aax{} for each light intensity with leaf
-using \agcurve{} curve
-parameters}{ for each light intensity with leaf using  curve parameters}}\label{for-each-light-intensity-with-leaf-using-curve-parameters}
-
-Within the \(k^{\text{th}}\) leaf, we estimated \aax{} for each light
-intensity by integrating the difference in \logA{} between the amphi and
-pseudohypo \agcurve{} curves over the range of \gsw{} values where the
-curves overlap (from \(\min(\log(g_\text{sw}))\) to
-\(\max(\log(g_\text{sw}))\)). The estimate of \aax{} for the
-\(k^{\text{th}}\) leaf at light intensity \(l\) in accession \(m\) is:
-
-\[\widehat{\mathrm{AA}}_{klm} = \int_{\text{min(log(}g_\text{sw}))}^{\text{max(log(}g_\text{sw}))} \text{log}\bigg(\frac{\hat{A}_\text{amphi}(x; \theta_{klm,\text{amphi})}}{\hat{A}_\text{hypo}(x; \theta_{klm,\text{hypo})}}\bigg) dx\]
-where:
-
-\[\theta_\text{amphi} \in \{\hat{b}_{0, f(\text{amphi}, k,l,m)}, \hat{b}_{1, f(\text{amphi}, k,l,m)}, \hat{b}_{2, f(\text{amphi}, k,l,m)}\}, \text{and}\]
-
-\[\theta_\text{hypo} \in \{\hat{b}_{0, f(\text{hypo}, k,l,m)}, \hat{b}_{1, f(\text{hypo}, k,l,m)}, \hat{b}_{2, f(\text{hypo}, k,l,m)}\}.\]
-The function \(f : \Theta_1 \rightarrow \Theta_2\) maps the set
-\(\Theta_1\) indexed by leaf type (amphi or pseudohypo), leaf replicate,
-light intensity, and accession to set \(\Theta_2\) indexed by individual
-\agcurve{} curve. This mapping is necessary because the random effects
-structure differs between models of \loggsw{} on \logA{} and that of
-models predicting \aax{} described in the next section.
-
-\paragraph{\texorpdfstring{Effects of light intensity, light treatment,
-and accession on
-\aax}{Effects of light intensity, light treatment, and accession on }}\label{effects-of-light-intensity-light-treatment-and-accession-on}
-
-We tested for effects of light intensity, light treatment, accession,
-and their interactions on \aax. All models included effects of light
-intensity and light treatment, as well as random effects of accession
-and replicate within accession. More complex models included
-interactions between light intensity and light treatment, as well as
-random effects of accession on the effects of light intensity and light
-treatment. We used a weakly informative normal prior with mean \(0\) and
-standard deviation \(10\) for fixed effects of light intensity and
-treatment. We used a weakly informative normal prior with mean \(-3\)
-and standard deviation \(5\) for the the random effect standard
-deviation of replicate within accession. We accounted for the
-phylogenetic structure among the random effects of accession using an
-Ornstein-Uhlenbeck (OU) process (\emph{64}). The expected covariance
-between accessions \(i\) and \(j\) (\(\text{Cov}(i, j)\) ) is:
-
-\[\text{Cov}(i, j) = \frac{\sigma^2}{2 \alpha} \exp(-\alpha D_{ij})\]
-where \(\sigma^2\) is the variance of the random effect, \(\alpha\) is
-the rate of decay of the covariance with phylogenetic distance,
-\(D_{ij}\), the phylogenetic distance between accessions \(i\) and
-\(j\). We estimated \(\sigma^2 / (2 \alpha)\) and \(\alpha\) as a
-separate parameters and reparameterized them as \(\sigma^2\) and
-\(\alpha\). We used a weakly informative normal prior with mean \(0\)
-and standard deviation \(10\) on OU parameters.
-
-We modeled the residual standard deviation of \aax,
-(i.e.~phylogenetically unstructured variation unaccounted for by
-explanatory variables) on a log-link scale with effects of light
-intensity and light treatment. We used a weakly informative normal prior
-with mean \(-3\) and standard deviation \(5\) on the residual standard
-deviation intercept and a weakly informative normal prior with mean
-\(0\) and standard deviation \(1\) on the effects of light intensity and
-light treatment on the residual standard deviation.
-
-\paragraph{\texorpdfstring{Effects of native light habitat on
-accession-level
-\aax}{Effects of native light habitat on accession-level }}\label{effects-of-native-light-habitat-on-accession-level}
-
-We tested a linear effect of native light habitat on accession-level
-\aax. In Models 1 and 2, we used the random intercept of \aax{} as a
-response variable; in model 3 we used the random intercept plus the
-random effect of accession at high light intensity; in model 4 we used
-the random intercept plus the random effect of accession in the sun
-treatment; in model 5 we used the random intercept plus the random
-effects of accession at high light intensity and sun treatment. We used
-a weakly informative normal prior with mean \(0\) and standard deviation
-\(1\) for the slope and intercept. We accounted for the phylogenetic
-structure among the model residuals using an (OU) process. We used a
-weakly informative normal prior with mean \(0\) and standard deviation
-\(10\) on OU parameters.
-
-\subsubsection{Predictions}\label{sec-predictions}
-
-The assimilatory, plastic, and constitutive hypotheses make different
-predictions about the relationship between \aax{} and light intensity,
-light treatment, and native \ppfd{} among accessions. Since these
-hypotheses are not mutually exclusive, we describe how we assessed
-support for one, two, or all three hypotheses simultaneously in
-\autoref{tbl-predictions}. In general, the assimilatory hypothesis was
-supported if \aax{} was greater at high light intensity than low light
-intensity. The plastic hypothesis was supported if \aax{} was greater in
-sun leaves than shade leaves. The constitutive hypothesis was supported
-if accession-level \aax{} increased with native \ppfd. In interactive
-models, we only consider positively reinforcing interactions between
-high light intensity, sun leaves, and native \ppfd{} because these are
-the only interactions which could explain why amphistomatous leaves are
-advantageous in high light habitats.
-
-We evaluated predictions using a combination of parameter estimation and
-model selection. In all cases, we estimated parameters and confidence
-intervals as described in Section~\ref{sec-fitting}. If the 95\%
-confidence intervals for a parameter did not overlap zero, we considered
-the parameter to be significantly different from zero. We used the
-leave-one-out cross-validation information criterion (LOOIC) to compare
-the fit of models (\autoref{tbl-models}) using the \emph{R} package
-\textbf{loo} version 2.8.0 (\emph{65}) to calculate LOOIC values. Note
-that LOOIC was calculated only from pointwise likelihood values of the
-submodel estimating effects of light intensity, light treatments, and
-accession on \aax{} (see Section~\ref{sec-parameters}). We considered
-models with two standard errors of the mean lower LOOIC value to be a
-better fit to the data; we considered models with LOOIC values within
-two standard errors of the mean to be have similar support.
-
-\begin{longtable}{>{\raggedright\arraybackslash}p{1in}>{\raggedright\arraybackslash}p{1.5in}>{\raggedright\arraybackslash}p{3in}}
-
-\caption{\label{tbl-predictions}Predictions of competing hypotheses
-about the relationship between \aax{} and light intensity, light
-treatment, and native \ppfd{} among accessions. The middle column lists
-specific directional predictions about parameter values and model fit
-according to leave-one-out cross-validation information criterion
-(LOOIC), where \(\text{LOOIC}_i\) is the LOOIC value for model \(i\).
-The rightmost column describes the predictions in words and explains how
-accession-level \aax{} is calculated in the relevant model.}
-
-\tabularnewline
-
-\toprule
-\textbf{Hypothesis} & \textbf{Prediction(s)} & \textbf{Description}\\
-\midrule
-Assimilatory & $\beta_{\mathrm{AA},2000} > 0$ & \hspace{-1em}Average \aax{} at high light intensity is greater than that at low light intensity\\
-\cmidrule{1-3}\pagebreak[0]
-Plastic & $\beta_{\mathrm{AA},\text{sun}} > 0$ & \hspace{-1em}Average \aax{} in sun leaves is greater than that in shade leaves\\
-\cmidrule{1-3}\pagebreak[0]
-Constitutive & $\beta_{\mathrm{PPFD,AA}} > 0$ & \hspace{-1em}Accession-level \aax{} ($\mathrm{AA}_\text{acc}$) increases with native \ppfd\\
-\nopagebreak
- &  & \hspace{-1em}$\mathrm{AA}_\text{acc} = \beta_{\mathrm{AA}, 0} + \vec{\beta}_{\mathrm{AA}, \text{acc}}$\\
-\cmidrule{1-3}\pagebreak[0]
-Assimilatory $\times$ Plastic & $\beta_{\mathrm{AA},2000,\text{sun}} > 0$ & \hspace{-1em}Average \aax{} is highest at high light intensity in sun leaves\\
-\nopagebreak
- & $\text{LOOIC}_\text{1} > \text{LOOIC}_\text{2}$ & \hspace{-1em}\\
-\cmidrule{1-3}\pagebreak[0]
-Assimilatory $\times$ Constitutive & $\beta_{\mathrm{AA},2000} > 0$ & \hspace{-1em}Average \aax{} at high light intensity is greater than that at low light intensity\\
-\nopagebreak
- & $\beta_{\mathrm{PPFD,AA}} > 0$ & \hspace{-1em}Accession-level \aax{} at high light intensity ($\mathrm{AA}_{\text{acc},2000}$) increases with native \ppfd\\
-\nopagebreak
- & $\text{LOOIC}_\text{1} > \text{LOOIC}_\text{3}$ & \hspace{-1em}$\mathrm{AA}_{\text{acc},2000} = \beta_{\mathrm{AA}, 0} + \vec{\beta}_{\mathrm{AA}, \text{acc}} + \vec{\beta}_{\mathrm{AA}, 2000, \text{acc}}$\\
-\cmidrule{1-3}\pagebreak[0]
-Plastic $\times$ Constitutive & $\beta_{\mathrm{AA},\text{sun}} > 0$ & \hspace{-1em}Average \aax{} in sun leaves is greater than that in shade leaves\\
-\nopagebreak
- & $\beta_{\mathrm{PPFD,AA}} > 0$ & \hspace{-1em}Accession-level \aax{} in sun leaves ($\mathrm{AA}_{\text{acc},\text{sun}}$) increases with native \ppfd\\
-\nopagebreak
- & $\text{LOOIC}_\text{1} > \text{LOOIC}_\text{4}$ & \hspace{-1em}$\mathrm{AA}_{\text{acc},\text{sun}} = \beta_{\mathrm{AA}, 0} + \vec{\beta}_{\mathrm{AA}, \text{acc}} + \vec{\beta}_{\mathrm{AA}, \text{sun}, \text{acc}}$\\
-\cmidrule{1-3}\pagebreak[0]
-Assimilatory $\times$ Plastic $\times$ Constitutive & $\beta_{\mathrm{AA},2000,\text{sun}} > 0$ & \hspace{-1em}Average \aax{} is highest at high light intensity in sun leaves\\
-\nopagebreak
- & $\beta_{\mathrm{PPFD,AA}} > 0$ & \hspace{-1em}Accession-level \aax{} at high light intensity in sun leaves ($\mathrm{AA}_{\text{acc},2000,\text{sun}}$) increases with native \ppfd\\
-\nopagebreak
- & $\text{LOOIC}_\text{3} > \text{LOOIC}_\text{5}$ & \hspace{-1em}$\mathrm{AA}_{\text{acc},2000,\text{sun}} = \beta_{\mathrm{AA}, 0} + \vec{\beta}_{\mathrm{AA}, \text{acc}} + \vec{\beta}_{\mathrm{AA}, 2000, \text{acc}} + \vec{\beta}_{\mathrm{AA}, \text{sun}, \text{acc}}$\\
-\bottomrule
-
-\end{longtable}
-
-\section{Acknowledgements}\label{acknowledgements}
-
-Justin Alter, Max Gatlin, Joana Kim, Jenna Matsuyama, Brandon Najarian,
-Kai Yasuda. Sarah Friedrich helped with figures.
-
-\section*{References}\label{references}
-\addcontentsline{toc}{section}{References}
+\caption{\label{fig-lma_aa}\textbf{Plastic changes in leaf mass per area (LMA) mediate the effect of sun- to shade-grown plants on amphistomy advantage (\aax{}).}
+Each point is the estimated population-level LMA (\(x\)-axis) and \aax{}
+(\(y\)-axis) for each population grown under high light intensity
+(\ppfdequals{2000}) and low light intensity (\ppfdequals{150}). The
+points are colored by growth light treatment. The solid line is the
+posterior median of the linear regression of individual-level LMA on
+\(\aahigh\) with the shaded region showing the 95\% confidence ribbon;
+the dashed line is the same for \(\aalow\). Confidence intervals for
+each population estimate and individual-level estimates are omitted for
+visual clarity.}
+
+\end{figure}%
+
+We conclude that developmental plasticity, as opposed to acclimation or
+adaptation to open habitats, may explain the explain the long-standing
+observation that amphistomatous leaves are more common in sunny
+habitats, at least among herbaceous plants including crop relatives. Our
+results advance existing explanations by showing high light intensity
+\emph{per se} does not increase the benefit of amphistomy, but rather
+that anatomical and biochemical change associated with higher light
+intensity modulate \aax{}. Developmental plasticity needs to be
+considered when interpreting phylogenetic comparative analyses of
+adaptation, because plastic differences may be mistaken for genetic
+differences.
+
+A second major conclusion is that the significant photosynthetic gain of
+amphistomatous leaves implies that the many hypostomatous leaves that
+dominate mesic to wet forests globally are giving up `free' carbon that
+would require little to no additional water loss. Under realistic
+assumptions, we estimate that hypostomatous leaves could achieve the
+same \(A\) with X to X\% less water loss if they were amphistomatous
+(Table S9). Ours is the first experimental demonstration that amphistomy
+is beneficial for sun leaves because of plastic changes, but future
+research needs to estimates the tradeoffs associated with amphistomatous
+leaves in order to fully explain their rarity among many woody plants
+and shade-tolerant herbs.
+
+\subsection{Acknowledgements}\label{acknowledgements}
+
+Sam McKlin and Tom Buckley helped with protocol development. Justin
+Alter, Max Gatlin, Joana Kim, Jenna Matsuyama, Brandon Najarian, and Kai
+Yasuda contributed to data collection. Sarah Friedrich helped with
+conceptual figure.
+
+\subsubsection{Funding:}\label{funding}
+
+US National Science Foundation OIA-1929167 to C.D.M.
+
+\subsubsection{Author contributions:}\label{author-contributions}
+
+Conceptualization: C.D.M.; Methodology: C.D.M., W.S.L.; Investigation:
+C.D.M., W.S.L., D.W.; Visualization: C.D.M.; Funding acquisition:
+C.D.M.; Writing -- original draft: C.D.M.; Writing -- review \& editing:
+C.D.M., W.S.L., D.W.
+
+\subsection{References}\label{references}
 
@@
 
+\bibitem[\citeproctext]{ref-john_anatomical_2017}
+\CSLLeftMargin{53. }%
+\CSLRightInline{G. P. John, C. Scoffoni, T. N. Buckley, R. Villar, H.
+Poorter, L. Sack, \href{https://doi.org/10.1111/ele.12739}{The
+anatomical and compositional basis of leaf mass per area}. \emph{Ecology
+Letters} \textbf{20}, 412--425 (2017).}
+
+\bibitem[\citeproctext]{ref-jones_effects_1972}
+\CSLLeftMargin{54. }%
+\CSLRightInline{H. G. Jones, R. O. Slatyer,
+\href{https://doi.org/10.1071/BI9720443}{Effects of {Intercellular}
+{Resistances} on {Estimates} of the {Intracellular} {Resistance} to
+{Co2} {Uptake} by {Plant} {Leaves}}. \emph{Australian Journal of
+Biological Sciences} \textbf{25}, 443 (1972).}
+
 \bibitem[\citeproctext]{ref-schoch_dependence_1980}
-\CSLLeftMargin{53. }%
+\CSLLeftMargin{55. }%
 \CSLRightInline{P.-G. Schoch, C. Zinsou, M. Sibi,
@@
 \bibitem[\citeproctext]{ref-sack_developmental_2016}
-\CSLLeftMargin{54. }%
+\CSLLeftMargin{56. }%
 \CSLRightInline{L. Sack, T. N. Buckley,
@@
 \bibitem[\citeproctext]{ref-mott_amphistomy_1991}
-\CSLLeftMargin{55. }%
+\CSLLeftMargin{57. }%
 \CSLRightInline{K. A. Mott, O. Michaelson, Amphistomy as an adaptation
@@
 \bibitem[\citeproctext]{ref-schindelin_fiji_2012}
-\CSLLeftMargin{56. }%
+\CSLLeftMargin{58. }%
 \CSLRightInline{J. Schindelin, I. Arganda-Carreras, E. Frise, V. Kaynig,
@@
 \bibitem[\citeproctext]{ref-marshall_model_1980}
-\CSLLeftMargin{57. }%
+\CSLLeftMargin{59. }%
 \CSLRightInline{B. Marshall, P. V. Biscoe,
@@
 \bibitem[\citeproctext]{ref-elzhov_minpacklm_2023}
-\CSLLeftMargin{58. }%
+\CSLLeftMargin{60. }%
 \CSLRightInline{T. V. Elzhov, K. M. Mullen, A.-N. Spiess, B. M. Bolker,
@@
 \bibitem[\citeproctext]{ref-stan_development_team_stan_2025}
-\CSLLeftMargin{59. }%
+\CSLLeftMargin{61. }%
 \CSLRightInline{Stan Development Team, \emph{Stan {Modeling} {Language}
@@
 \bibitem[\citeproctext]{ref-burkner_brms_2017}
-\CSLLeftMargin{60. }%
+\CSLLeftMargin{62. }%
 \CSLRightInline{P.-C. Bürkner,
@@
 \bibitem[\citeproctext]{ref-gabry_cmdstanr_2025}
-\CSLLeftMargin{61. }%
+\CSLLeftMargin{63. }%
 \CSLRightInline{J. Gabry, R. Češnovar, A. Johnson, S. Bronder,
@@
 \bibitem[\citeproctext]{ref-r_core_team_r:_2025}
-\CSLLeftMargin{62. }%
+\CSLLeftMargin{64. }%
 \CSLRightInline{R Core Team, \emph{R: {A} {Language} and {Environment}
@@
 \bibitem[\citeproctext]{ref-gelman_inference_1992}
-\CSLLeftMargin{63. }%
+\CSLLeftMargin{65. }%
 \CSLRightInline{A. Gelman, D. B. Rubin, Inference from iterative
@@
 \bibitem[\citeproctext]{ref-hansen_stabilizing_1997}
-\CSLLeftMargin{64. }%
+\CSLLeftMargin{66. }%
 \CSLRightInline{T. F. Hansen,
@@
 \bibitem[\citeproctext]{ref-vehtari_practical_2017}
-\CSLLeftMargin{65. }%
+\CSLLeftMargin{67. }%
 \CSLRightInline{A. Vehtari, A. Gelman, J. Gabry,
@@
 
+\newpage{}
+
+\section{Supplementary Materials}\label{supplementary-materials}
+
+\renewcommand{\thefigure}{S\arabic{figure}}
+\renewcommand{\thetable}{S\arabic{table}}
+\renewcommand{\theequation}{S\arabic{equation}}
+\setcounter{figure}{0}
+\setcounter{table}{0}
+\setcounter{equation}{0}
+
+\subsection{Materials and Methods}\label{sec-methods}
+
+\subsubsection{Populations}\label{populations}
+
+We compared \aax{} among 29 ecologically diverse populations of wild
+tomato, including representatives of all described species of
+\emph{Solanum} sect. \emph{Lycopersicon} and sect.
+\emph{Lycopersicoides} (\emph{51}) and the cultivated tomato \emph{S.
+lycopersicum} var. \emph{lycopersicum} (Table~\ref{tbl-populations}).
+Due to constraints on growth space and time, we spread out measurements
+over 61.1 weeks from August 29, 2022 to October 31, 2023. Replicates
+within population were evenly spread out over this period to prevent
+confounding of temporal variation in growth conditions with accession.
+{[}anything else to say here? maybe explain population selection and
+phylogeny?{]}
+
+\begin{longtable}{>{\raggedright\arraybackslash}p{3.25cm}lrrrr}
+
+\caption{\label{tbl-populations}Accession information of \emph{Solanum}
+populations used in this study. The species name, accession number,
+collection latitude, longitude, elevation, and daily solar radiation
+estimated using the SPLASH algorithm (see `Climate data'). TGRC: Tomato
+Genetics Resource Center; PAI: Plant area index.}
+
+\tabularnewline
+
+\toprule
+Species & TGRC accession & Latitude & Longitude & Elevation (mas) & $\mathrm{PAI}~(\unit{\meter\squared\per\meter\squared})$\\
+\midrule
+\em{\cellcolor{gray!10}{S. arcanum}} & \cellcolor{gray!10}{LA2172} & \cellcolor{gray!10}{-6.008} & \cellcolor{gray!10}{-78.858} & \cellcolor{gray!10}{662} & \cellcolor{gray!10}{0.70}\\
+\em{S. cheesmaniae} & LA0429 & -0.644 & -90.329 & 800 & 0.60\\
+\em{\cellcolor{gray!10}{S. cheesmaniae}} & \cellcolor{gray!10}{LA3124} & \cellcolor{gray!10}{-0.804} & \cellcolor{gray!10}{-90.042} & \cellcolor{gray!10}{1} & \cellcolor{gray!10}{0.36}\\
+\em{S. chilense} & LA1782 & -15.267 & -74.633 & 1000 & 0.35\\
+\em{\cellcolor{gray!10}{S. chilense}} & \cellcolor{gray!10}{LA4117A} & \cellcolor{gray!10}{-22.907} & \cellcolor{gray!10}{-67.941} & \cellcolor{gray!10}{3540} & \cellcolor{gray!10}{0.22}\\
+\addlinespace
+\em{S. chmielewskii} & LA1028 & -13.883 & -73.017 & 3000 & 0.43\\
+\em{\cellcolor{gray!10}{S. chmielewskii}} & \cellcolor{gray!10}{LA1316} & \cellcolor{gray!10}{-13.400} & \cellcolor{gray!10}{-73.906} & \cellcolor{gray!10}{2920} & \cellcolor{gray!10}{1.07}\\
+\em{S. corneliomulleri} & LA0107 & -13.117 & -76.383 & 60 & 0.02\\
+\em{\cellcolor{gray!10}{S. corneliomulleri}} & \cellcolor{gray!10}{LA0444} & \cellcolor{gray!10}{-13.433} & \cellcolor{gray!10}{-76.133} & \cellcolor{gray!10}{100} & \cellcolor{gray!10}{0.49}\\
+\em{S. galapagense} & LA0436 & -0.953 & -90.978 & 40 & 0.20\\
+\addlinespace
+\em{\cellcolor{gray!10}{S. galapagense}} & \cellcolor{gray!10}{LA1044} & \cellcolor{gray!10}{-0.284} & \cellcolor{gray!10}{-90.548} & \cellcolor{gray!10}{0} & \cellcolor{gray!10}{0.18}\\
+\em{S. habrochaites} & LA0407 & -2.181 & -79.884 & 70 & 0.47\\
+\em{\cellcolor{gray!10}{S. habrochaites}} & \cellcolor{gray!10}{LA1777} & \cellcolor{gray!10}{-9.550} & \cellcolor{gray!10}{-77.700} & \cellcolor{gray!10}{3216} & \cellcolor{gray!10}{0.42}\\
+\em{S. huaylasense} & LA1358 & -9.533 & -77.967 & 750 & 0.58\\
+\em{\cellcolor{gray!10}{S. huaylasense}} & \cellcolor{gray!10}{LA1360} & \cellcolor{gray!10}{-9.546} & \cellcolor{gray!10}{-77.929} & \cellcolor{gray!10}{1490} & \cellcolor{gray!10}{0.48}\\
+\addlinespace
+\em{S. huaylasense} & LA1364 & -10.133 & -77.383 & 2920 & 0.88\\
+\em{\cellcolor{gray!10}{S. lycopersicoides}} & \cellcolor{gray!10}{LA2951} & \cellcolor{gray!10}{-19.317} & \cellcolor{gray!10}{-69.450} & \cellcolor{gray!10}{2200} & \cellcolor{gray!10}{0.49}\\
+\em{S. lycopersicoides} & LA4126 & -19.287 & -69.396 & 3120 & 0.39\\
+\em{\cellcolor{gray!10}{S. neorickii}} & \cellcolor{gray!10}{LA1322} & \cellcolor{gray!10}{-13.483} & \cellcolor{gray!10}{-72.442} & \cellcolor{gray!10}{2380} & \cellcolor{gray!10}{0.68}\\
+\em{S. neorickii} & LA2133 & -3.400 & -79.183 & 1980 & 1.05\\
+\addlinespace
+\em{\cellcolor{gray!10}{S. pennellii}} & \cellcolor{gray!10}{LA0716} & \cellcolor{gray!10}{-16.225} & \cellcolor{gray!10}{-73.617} & \cellcolor{gray!10}{50} & \cellcolor{gray!10}{0.19}\\
+\em{S. pennellii} & LA0750 & -14.775 & -75.034 & 550 & 0.05\\
+\em{\cellcolor{gray!10}{S. pennellii}} & \cellcolor{gray!10}{LA3778} & \cellcolor{gray!10}{-14.775} & \cellcolor{gray!10}{-75.034} & \cellcolor{gray!10}{616} & \cellcolor{gray!10}{0.05}\\
+\em{S. peruvianum} & LA2744 & -18.550 & -70.150 & 400 & 0.16\\
+\em{\cellcolor{gray!10}{S. peruvianum}} & \cellcolor{gray!10}{LA2964} & \cellcolor{gray!10}{-18.028} & \cellcolor{gray!10}{-70.835} & \cellcolor{gray!10}{75} & \cellcolor{gray!10}{1.16}\\
+\addlinespace
+\em{S. pimpinellifolium} & LA1269 & -11.483 & -77.075 & 400 & 0.50\\
+\em{\cellcolor{gray!10}{S. pimpinellifolium}} & \cellcolor{gray!10}{LA1589} & \cellcolor{gray!10}{-8.433} & \cellcolor{gray!10}{-78.817} & \cellcolor{gray!10}{30} & \cellcolor{gray!10}{0.11}\\
+\em{S. pimpinellifolium} & LA2933 & -1.442 & -80.562 & 375 & 1.62\\
+\em{\cellcolor{gray!10}{S. sitiens}} & \cellcolor{gray!10}{LA4116} & \cellcolor{gray!10}{-22.159} & \cellcolor{gray!10}{-68.782} & \cellcolor{gray!10}{2960} & \cellcolor{gray!10}{0.06}\\
+\bottomrule
+
+\end{longtable}
+
+\subsubsection{Plant growth conditions}\label{plant-growth-conditions}
+
+In all growth spaces, we recorded \(\mathrm{PPFD}\) using full spectrum
+quantum sensors (SQ-500-SS, Apogee Instruments, Logan, Utah, USA); we
+recorded temperature, RH, and {[}CO\(_2\){]} using an EE894 sensor (E+E
+Elektronik, Engerwitzdorf, Austria) protected by a radiation shield. All
+environmental measurements were taken every 10 minutes from the middle
+of plants racks at approximately the same height as the leaves we
+measured. We measured leaf temperature of focal leaves prior to
+measurement using an infrared radiometer (SI-111-SS, Apogee Instruments,
+Logan, Utah, USA).
+
+\paragraph{Germination and seedling
+stage}\label{germination-and-seedling-stage}
+
+Seeds provided by the Tomato Genetics Resource Center germinated on
+moist paper in plastic boxes after soaking for 30-60 minutes in a 50\%
+(volume per volume) solution of household bleach and water, followed by
+a thorough rinse. We transferred seedlings to cell-pack flats containing
+Pro-Mix BX potting mix (Premier Tech, Rivière-du-Loup, Quebec, Canada)
+once cotyledons fully emerged, typically within 1-2 weeks of sowing. We
+grew seeds and seedlings for both sun and shade treatments under the
+same environmental conditions (12:12 h,
+24.3:\(\qty{21.7}{\degreeCelsius}\), 49.6:58.4 RH day:night cycle). LED
+light provided
+\(\mathrm{PPFD} = \qty{267}{\micro\mol\raiseto{-2}\meter\raiseto{-1}\second}\)
+(Fluence RAZRx, Austin, Texas, USA).
+
+\paragraph{Light treatments}\label{light-treatments}
+
+Seedlings were randomly assigned in alternating order within population
+to the sun or shade treatment during transplanting. After seedlings
+established in cell-pack flats for \(\approx 2\) weeks, we transplanted
+them to 3.78 L plastic pots containing 60\% Pro-Mix BX potting mix, 20\%
+coral sand (Pro-Pak, Honolulu, Hawaiʻi, USA), and 20\% cinders (Niu
+Nursery, Honolulu, Hawaiʻi, USA). Percentage composition is on a volume
+basis. The soil mixture contained slow release NPK fertilizer following
+manufacturer instructions (Osmocote Smart-Release Plant Food Flower \&
+Vegetable, The Scotts Company, Marysville, Ohio, USA). We determined pot
+field capacity one week after transplanting using a scale (Ohaus V12P15
+Valor 1000, Parsippany, New Jersey, USA) and watered to field capacity
+three times per week to prevent drought stress.
+
+We assigned sun and shade treatment to lower and upper racks of a
+\(\qty{1.22}{\meter} \times \qty{2.44}{\meter}\) shelving unit in a
+climate-controlled growth room. We assigned the sun treatment to the
+lower rack to limit diffuse light from reaching the shade treatment. The
+average daytime \ppfd{} was \ppfdqty{761} and \ppfdqty{115} for sun and
+shade treatments, respectively. To isolate the effect of light intensity
+from quality, we used the same LED model with the the same spectrum
+(Fluence SPYDR 2i, Austin, Texas, USAS), but dimmed the lights in the
+shade treatment. To maintain homogeneous environmental conditions other
+than light, we mixed air within the growth room using an air circulator
+(Vornado 693DC, Andover, Kansas, USA) and within racks using a miniature
+oscillating air circulator (Vornado Atom 1, Andover, Kansas, USA).
+Despite these efforts, the air in the sun treatment was on average
+\(\qty{2.56}{\degreeCelsius}\) warmer and the average RH was
+consequently 5.75\% lower. However, because of evaporative cooling, the
+leaves in the sun treatment were only \(\qty{0.886}{\degreeCelsius}\) on
+average (\(n = 699\) leaves).
+
+\subsubsection{Leaf trait measurements}\label{leaf-trait-measurements}
+
+We selected a fully expanded, unshaded leaf at least six leaves above
+the cotyledons during early vegetative growth. This typically meant that
+plants had grown in light treatments for \(\approx\) 4 weeks, ensuring
+they had time to sense and respond developmentally to the light
+intensity of the treatment rather than the seedling conditions
+(\emph{55}). Shade plants grew slower than sun plants, hence leaves at
+the same developmental stage were measured on chronologically older
+plants in the shade treatment. In some sun plants, we had to use leaves
+higher on the stem because short internodes made lower leaves
+inaccessible with the gas exchange equipment. We measured terminal
+leaflets in 82.6\% of cases, but used the lateral leaflet closest to the
+terminal leaflet when it was damaged or difficult to clamp into the gas
+exchange chamber. When a leaflet was damaged during gas exchange
+measurements, we collected anatomical data from the nearest leaflet on
+the same leaf (1.58)\%. We measured chlorophyll concentration index
+(CCI) using a chrolophyll concentration meter (MC-100, Apogee
+Instruments, Logan, Utah, USA) on the lamina of focal leaflets before
+gas exchange measurements at the same time we measured leaf temperature.
+
+\paragraph{Amphistomy advantage}\label{amphistomy-advantage}
+
+We estimated `amphistomy advantage' (\aax) \emph{sensu} (\emph{20}), but
+with modifications previously described in (\emph{46}). \aax{} is
+calculated as the log-response ratio of \(A\) compared at the same total
+\gsw:
+
+\[\mathrm{AA} = \mathrm{log}(A_{\mathrm{amphi}} / A_{\mathrm{hypo}})\]
+
+We measured the photosynthetic rate of an untreated amphistomatous leaf
+(\Aamphi) over a range of \gsw{} values. We refer to this as an
+\agcurve{} curve. We compared the \agcurve{} curve of the untreated leaf
+to the photosynthetic rate of pseudohypostomatous leaf (\Ahypo), which
+is the same leaf but with gas exchange through the upper surface blocked
+by a neutral density plastic (propafilm).
+
+We measured \agcurve{} curves using a portable infrared gas analyzer
+(LI-6800PF, LI-COR Biosciences, Lincoln, Nebraska, USA).
+Light-acclimated plants were placed under LEDs dimmed to match their
+light treatment during gas exchange measurements. We estimated the
+photosynthetic rate (\(A\)) and stomatal conductance to CO\(_2\) (\gsw)
+at ambient CO\(_2\) (\caequals{415}) and \tleafequals{25.0}. The
+irradiance of the light source in the pseudohypo leaf was higher because
+the propafilm reduces transmission. To compensate for reduced
+transmission, we increased incident \ppfd for pseudohypo leaves by a
+factor 1/0.91, the inverse of the measured transmissivity of the
+propafilm. We also set the stomatal conductance ratio, for purposes of
+calculating boundary layer conductance, to 0 for pseudohypo leaves
+following manufacturer directions.
+
+We collected four \agcurve{} curves per leaf, an amphi (untreated) curve
+and a pseudohypo (treated) curve at high light-intensity
+(\ppfdequals{2000}; 97.8:2.24 red:blue) and low light-intensity
+(\ppfdequals{150}; 87.0:13.0 red:blue). We always measured high
+light-intensity curves first because photosynthetic downregulation is
+faster than upregulation in these species. To control for order effects,
+we alternated between starting with amphi or pseudohypo leaf
+measurements. Unlike (\emph{46}), preliminary experiments with
+\emph{Solanum} indicated a strong order effect in that \(A\) declined in
+the second curve. Therefore, we made measurements over two days. On the
+first day, we measured high and low light-intensity curves for either
+amphi or pseudohypo leaves; on the second day, we measured high and low
+light-intensity curves on the other leaf type.
+
+In all cases, we acclimated the focal leaf to high light
+(\ppfdequals{2000}) and high relative humidity (\rhequals{70}) until
+\(A\) and \gsw{} reach their maximum. After that, we decreased \rh{} to
+\(\approx 10\%\) to induce rapid stomatal closure without biochemical
+downregulation. Hence, \Aamphi{} and \Ahypo{} were both measured at low
+chamber humidity after the leaf had acclimated to high humidity. All
+other environmental conditions in the leaf chamber remained the same. We
+logged data until \gsw{} reached its nadir. We then acclimated the leaf
+to low light (\ppfdequals{150}) and \rhequals{70} before inducing
+stomatal closure with low \rh and logging data as described above.
+
+\paragraph{\texorpdfstring{Light-response (\aqcurve)
+curves}{Light-response () curves}}\label{light-response-curves}
+
+In 91.3\% of plants, we measured light-response (\aqcurve) curves on the
+same leaflets as \agcurve{} curves. However, when a leaflet was damaged
+during \agcurve{} curves, we used the next closest leaflet for
+\aqcurve{} curves. Leaves acclimated to high light-intensity
+(\ppfdequals{2000}), ambient CO\(_2\) (\caequals{415}), \rhequals{50},
+and \tleafequals{25}. After \(A\) and \gsw{} stabilized, we measured
+\(A\) at 20 light-intensity levels between \(0\) and
+\(\qty{2000}{\micro\mol\raiseto{-2}\meter\raiseto{-1}\second}\) in
+descending order.
+
+\paragraph{Stomatal anatomy}\label{stomatal-anatomy}
+
+We estimated the stomatal density and size on ab- and adaxial leaf
+surfaces from all leaves, using guard cell length as a proxy for
+stomatal size since it proportional to maximum conductance (\emph{56}).
+We made surface impressions of leaf lamina from the same area used for
+gas exchange measurements using a-silicone impression material (Zhermack
+elite HD+, light body, fast set, Rovigo, Italy). We applied clear nail
+polish to make positive replicas of the impression. After nail polish
+dried, we mounted replicas on a microscope slide using transparent tape
+(\emph{57}). We digitized a portion of each leaf surface replica using a
+brightfield microscope (Leica DM2000, Wetzlar, Germany). We counted and
+measured guard cell length on all stomata using the FIJI implementation
+of ImageJ2 version 2.3.0 (\emph{58}), then divided the count by the
+visible leaf area (\(\qty{0.890}{\raiseto{2}\milli\meter}\)) to estimate
+stomatal density.
+
+\paragraph{Leaf mass per area}\label{leaf-mass-per-area}
+
+Leaf mass per area (LMA) is the dry mass divided by the leaflet area. We
+scanned fresh leaflets on a flat bed scanner (Epson V600, Los Alamitos,
+California. USA) and measured leaflet area from digital images using the
+FIJI implementation of ImageJ2 version 2.3.0 (\emph{58}). We dried
+leaves for 72 hours at \(\qty{74}{\degreeCelsius}\) in a food dehydrator
+(Cosori CP267-FD, Vesync Co., Anaheim, California, USA) and weighed
+using a benchtop analytical balance (Ohaus PR64 Analytical Balance,
+Parsippany, New Jersey, USA). In \(10.5\%\) we measured LMA on the
+adjacent leaflet because the focal leaflet was damaged or wilted while
+making surface impressions and we could not reliably estimate area. LMA
+data are missing from \(2.97\%\) of individuals because the area or mass
+was not recorded at all or recorded incorrectly.
+
+\subsubsection{\texorpdfstring{Cleaning \agcurve{}
+curves}{Cleaning  curves}}\label{cleaning-curves}
+
+The raw data set consisted of 2,370 \agcurve{} curves with an average of
+63.2 points per curve. Manual curation of a data set this size in a
+principled, consistent manner is not feasible. Therefore, we automated
+data cleaning using custom \emph{R} scripts. Cleaning is divided into
+six sequential steps (\autoref{tbl-cleaning1}).
+
+\begin{table}
+
+\caption{\label{tbl-cleaning1}Six sequential steps for cleaning
+\agcurve{} curves. The rationale and procedure for each step are
+described in the text. The rightmost columns summarize the number of
+curves and mean number of points per curve remaining after each step.
+For reference, there are four possible \agcurve{} curves per replicate:
+all combinations of leaf type (amphi or pseudohypo) and light intensity
+(high or low).}
+
+\centering{
+
+\centering
+\begin{tabular}{>{\raggedright\arraybackslash}p{9cm}>{\centering\arraybackslash}p{2.5cm}>{\centering\arraybackslash}p{2.5cm}}
+\toprule
+Step: description & Number of curves & Number of points per curve\\
+\midrule
+\cellcolor{gray!10}{1. remove unreliable and unusable data points} & \cellcolor{gray!10}{2,361} & \cellcolor{gray!10}{63.0}\\
+2. remove hysteretic portion of \agcurve{} curves at low \gsw{} & 2,360 & 59.2\\
+\cellcolor{gray!10}{3. remove outliers within each \agcurve{} curve} & \cellcolor{gray!10}{2,360} & \cellcolor{gray!10}{58.7}\\
+4. remove replicates with no overlap between amphi and pseudohypo \agcurve{} curves & 2,268 & 58.5\\
+\cellcolor{gray!10}{5. thin redundant data points within each \agcurve{} curve} & \cellcolor{gray!10}{2,268} & \cellcolor{gray!10}{28.1}\\
+\addlinespace
+6. trim extreme \aax{} values & 2,214 & 28.1\\
+\bottomrule
+\end{tabular}
+
+}
+
+\end{table}%
+
+\paragraph{Remove unreliable and unusable data
+points}\label{remove-unreliable-and-unusable-data-points}
+
+\emph{Rationale}: Unreliable data points consisted of those where
+chamber {[}CO\(_2\){]} was unstable and therefore measurements are not
+biologically meaningful. Unusable data points were those where \(A < 0\)
+because the logarithm of a negative number is undefined.
+
+\emph{Procedure}: We retained data points where \cabetween{410}{420} and
+\(A> 0\).
+
+\paragraph{\texorpdfstring{Remove hysteretic portion of \agcurve{}
+curves at low
+\gsw{}}{Remove hysteretic portion of  curves at low }}\label{remove-hysteretic-portion-of-curves-at-low}
+
+\emph{Rationale}: In most \agcurve{} curves, we observed a hysteretic
+response at low \gsw. After \gsw{} and \(A\) declined simultaneously,
+\(A\) increased slightly as \gsw{} continued to decline or stabilize,
+indicating some leaf acclimation to low \rh. We removed this portion of
+the curve to focus curve-fitting on the primary domain where \(A\)
+increases monotonically with \gsw{}.
+
+\emph{Procedure}: For each curve, we removed data points after \gsw{}
+had reached its minimum unless there were fewer than 10 data points
+remaining.
+
+\paragraph{\texorpdfstring{Remove outliers within each \agcurve{}
+curve}{Remove outliers within each  curve}}\label{remove-outliers-within-each-curve}
+
+\emph{Rationale}: Individual outliers within \agcurve{} curves, usually
+caused by transitory changes in chamber conditions, exert undue leverage
+on parameter estimates and cause bias and/or low precision in parameter
+estimates.
+
+\emph{Procedure}: We fit provisional quadratic regressions for each
+curve using ordinary least squares with the \texttt{lm()} function in
+\emph{R}. We sequentially removed data points with an absolute external
+studentized residual \(> 3\) until none remained.
+
+\paragraph{\texorpdfstring{Thin redundant data points within each
+\agcurve{}
+curve}{Thin redundant data points within each  curve}}\label{thin-redundant-data-points-within-each-curve}
+
+\emph{Rationale}: Data points closely spaced along the \agcurve{} curve
+provide redundant information and may be highly correlated
+(i.e.~pseudoreplication). This occurred because data was logged at a
+constant temporal interval, but the rate at which \gsw{} declined was
+not constant. Thinning reduces parameter estimation bias toward densely
+sampled regions of the curve which may not be the most biologically
+informative.
+
+\emph{Procedure}: We retained the maxima and minima \gsw{} for each
+curve and thinned all but one point per thinning interval of
+\(0.05~\log(\si{\mol\raiseto{-2}\meter\raiseto{-1}\second})\), retaining
+the point nearest the midpoint of the interval.
+
+\paragraph{\texorpdfstring{Remove replicates with no overlap between
+amphi and pseudohypo \agcurve{}
+curves}{Remove replicates with no overlap between amphi and pseudohypo  curves}}\label{remove-replicates-with-no-overlap-between-amphi-and-pseudohypo-curves}
+
+\emph{Rationale}: We could not estimate \aax{} for replicates where
+amphi and pseudohypo \agcurve{} curves did not overlap.
+
+\emph{Procedure}: We removed replicates where the range of \gsw{} values
+for amphi and pseudohypo \agcurve{} curves did not overlap.
+
+\paragraph{\texorpdfstring{Trim extreme \aax{}
+values}{Trim extreme  values}}\label{trim-extreme-values}
+
+\emph{Rationale}: Extreme \aax{} values were likely due to measurement
+error or leaf damage. Since amphi and pseudohypo \agcurve{} curves are
+measured on consecutive days, a poor calibration or a damaged leaf could
+cause a large difference in \(A\) between days, which would appear as an
+extreme \aax{} value.
+
+\emph{Procedure}: We provionsally estimated \aax{} for each replicate by
+integrating over the range of \gsw{} values where amphi and pseudohypo
+\agcurve{} curves overlap. In this procedure, curve parameters were
+provisionally estimated using ordinary least squares with the
+\texttt{lm()} function in \emph{R}. We then used point estimates of
+\aax{} for each replicate as the response variable in a linear model
+with light treatment, light intensity, population, and all interactions
+as explanatory variables. This model was also fit using ordinary least
+squares with the \texttt{lm()} function in \emph{R}. We classified
+extreme \aax{} values as those with an absolute internal studentized
+residual \(> 3\). Because these values likely indicate significant
+measurement error or leaf damage, we removed \agcurve{} curves at both
+light intensities if either was classified as extreme.
+
+\subsubsection{\texorpdfstring{Cleaning \aqcurve{}
+curves}{Cleaning  curves}}\label{cleaning-curves-1}
+
+The raw data set consisted of 658 \aqcurve{} curves with an average of
+19.4. Manual curation of a data set this size in a principled,
+consistent manner is not feasible. Therefore, we automated data cleaning
+using custom \emph{R} scripts. Cleaning is divided into two sequential
+steps (\autoref{tbl-cleaning2}).
+
+\begin{table}
+
+\caption{\label{tbl-cleaning2}Two sequential steps for cleaning
+\agcurve{} curves. The rationale and procedure for each step are
+described in the text. The rightmost columns summarize the number of
+curves and mean number of points per curve remaining after each step.}
+
+\centering{
+
+\centering
+\begin{tabular}{>{\raggedright\arraybackslash}p{9cm}>{\centering\arraybackslash}p{2.5cm}>{\centering\arraybackslash}p{2.5cm}}
+\toprule
+Step: description & Number of curves & Number of points per curve\\
+\midrule
+\cellcolor{gray!10}{1. remove outliers within each \aqcurve{} curve} & \cellcolor{gray!10}{658} & \cellcolor{gray!10}{19.1}\\
+2. remove \aqcurve{} curves with poor fit & 652 & 19.1\\
+\bottomrule
+\end{tabular}
+
+}
+
+\end{table}%
+
+\paragraph{\texorpdfstring{Remove outliers within each \aqcurve{}
+curve}{Remove outliers within each  curve}}\label{remove-outliers-within-each-curve-1}
+
+\emph{Rationale}: Individual outliers within \agcurve{} curves, usually
+caused by transitory changes in chamber conditions, exert undue leverage
+on parameter estimates and cause bias and/or low precision in parameter
+estimates.
+
+\emph{Procedure}: We fit provisional nonrectangular hyperbola
+(\emph{59}) to each \aqcurve{} curve using nonlinear regression with the
+\texttt{nlsLM()} function from the \emph{R} package \textbf{minpack.lm}
+version 1.2.4 (\emph{60}). We sequentially removed data points with an
+absolute external studentized residual \(> 3\) until none remained.
+
+\paragraph{\texorpdfstring{Remove \aqcurve{} curves with poor
+fit}{Remove  curves with poor fit}}\label{remove-curves-with-poor-fit}
+
+\emph{Rationale}: \aqcurve{} curves with a poor fit to the
+nonrectangular hyperbola most likely indicate systematic measurement
+error and/or the leaf was not fully acclimated to the chamber
+environment.
+
+\emph{Procedure}: As described above, we fit provisional nonrectangular
+hyperbola to each \aqcurve{} curve and calculated the model \(r^2\).
+There was a clear break between typical curves and poorly fitting curves
+where \(r^2 < 0.99\). We therefore removed \aqcurve{} curves with
+\(r^2 < 0.99\).
+
+\subsubsection{\texorpdfstring{Bayesian data analysis in
+\emph{Stan}}{Bayesian data analysis in Stan}}\label{bayesian-data-analysis-in-stan}
+
+We fit five models to test predictions of competing hypotheses about why
+amphistomy advantage (\aax) might be greater for leaves in sunny, open
+habitats. This section provides an overview of differences among models
+(\autoref{tbl-models}). The next sections describe how we fit models in
+\emph{Stan}, all model parameters and priors, and specific predictions
+about parameter values for each hypothesis.
+
+\begin{table}
+
+\caption{\label{tbl-models}Summary of differences among competing models
+of how \aax{} varies with light intensity, light treatment, and among
+populations as a function of native \ppfd. The models are numbered from
+simpler to more complex. All models include fixed effects of light
+intensity and light treatment; some models include interactions between.
+All models include a phylogenetic random effect of population on \aax;
+some models include varying effects of light intensity and light
+treatment among populations. The last column indicates the
+population-level \aax{} variable we used as a response to native \ppfd.}
+
+\centering{
+
+\begin{tabular}{>{\centering\arraybackslash}p{0.3in}>{\raggedright\arraybackslash}p{1.3in}>{\raggedright\arraybackslash}p{3.4in}>{\raggedright\arraybackslash}p{1in}}
+\toprule
+Model & Fixed effects & Phylogenetic random effects & Response to native \ppfd\\
+\midrule
+1 & light intensity & varying intercept among populations & $\mathrm{AA}_{0, \text{acc}}$\\
+
+ & light treatment &  & \\
+\cmidrule{1-4}
+2 & light intensity & varying intercept among populations & $\mathrm{AA}_{0, \text{acc}}$\\
+
+ & light treatment &  & \\
+
+ & intensity $\times$ treatment &  & \\
+\cmidrule{1-4}
+3 & light intensity & varying intercept among populations & $\mathrm{AA}_{2000, \text{acc}}$\\
+
+ & light treatment & varying effect of high light intensity among populations & \\
+\cmidrule{1-4}
+4 & light intensity & varying intercept among populations & $\mathrm{AA}_{\text{sun, acc}}$\\
+
+ & light treatment & varying effect of sun treatment among populations & \\
+\cmidrule{1-4}
+5 & light intensity & varying intercept among populations & $\mathrm{AA}_{2000, \text{sun, acc}}$\\
+
+ & light treatment & varying effect of high light intensity among populations & \\
+
+ & intensity $\times$ treatment & varying effect of sun treatment among populations & \\
+\bottomrule
+\end{tabular}
+
+}
+
+\end{table}%
+
+\paragraph{\texorpdfstring{Fitting models in
+\emph{Stan}}{Fitting models in Stan}}\label{sec-fitting}
+
+We fit Bayesian models with MCMC sampling in the probabilistic
+programming language \emph{Stan} (\emph{61}) using the \emph{R} package
+\textbf{brms} version 2.22.0 (\emph{62}). We used \emph{CmdStan} version
+2.35.0 and \textbf{cmdstanr} version 0.9.0 (\emph{63}) to interface with
+\emph{R} version 4.5.0 (\emph{64}). We sampled the posterior
+distribution from 1 chains with 2000 iterations each after 2000 warmup
+iterations per chain. We estimated parameters and confidence intervals
+as the median and 95\% quantile intervals of the posterior,
+respectively. We chose the number of chains, warmup and sampling
+iterations, and maximum treedepth so that parameter estimates converged
+(\(\hat{R} < 1.01\) (\emph{65})) and the effective sample size (ESS) for
+each parameter was \(> 400\).
+
+\paragraph{Parameter estimation and priors}\label{sec-parameters}
+
+There were four levels of parameter estimation in our analysis:
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\tightlist
+\item
+  Estimate \agcurve{} curve parameters
+\item
+  Estimate \aax{} for each light intensity with leaf using \agcurve{}
+  curve parameters
+\item
+  Estimate the effects of light intensity, light treatment, and
+  population on \aax{} (assimilatory and plasticity hypotheses)
+\item
+  Estimate the effects of native light habitat on population-level
+  \aax{} (constitutive hypothesis)
+\end{enumerate}
+
+Although the higher-level parameter estimates depend on the lower-level
+parameter estimates, we fit all models simultaneously to ensure that the
+uncertainty in lower-level estimates propagated to higher-level
+estimates.
+
+\begin{longtable}{>{\raggedright\arraybackslash}p{1in}>{\raggedright\arraybackslash}p{5in}}
+
+\caption{\label{tbl-parameters}Description of parameters estimated in
+the hierarchical Bayesian model. The \texttt{Parameter} column lists the
+parameter name as it appears in text. The \texttt{Description} column
+provides a brief description of the parameter.}
+
+\tabularnewline
+
+\toprule
+Parameter & Description\\
+\midrule
+\addlinespace[0.3em]
+\multicolumn{2}{l}{\textbf{\agcurve{} curve parameters}}\\
+\hspace{1em}\cellcolor{gray!10}{$\mathbf{B}_\text{curve}$} & \cellcolor{gray!10}{$n_\text{curve} \times 3$ array of random \agcurve{} curve-level coefficients ($b_{0,j}$, $b_{1,j}$, $b_{2,j}$); $\mathbf{B}_\text{curve} \sim \text{MVN}(\vec{0}, \symbf{\Sigma}_\text{curve})$}\\
+\hspace{1em}$\vec{\beta}_\text{curve}$ & vector of mean quadratic coefficients ($\beta_0$, $\beta_1$, $\beta_2$)\\
+\hspace{1em}\cellcolor{gray!10}{$\symbf{\Sigma}_\text{curve}$} & \cellcolor{gray!10}{$3 \times 3$ covariance matrix of curve-level coefficients}\\
+\hspace{1em}$\sigma_{\qty{6}{\raiseto{2}\centi\meter},\epsilon}$ & minimum residual standard deviation when the measured leaf surface area is $\qty{6}{\raiseto{2}\centi\meter}$\\
+\hspace{1em}\cellcolor{gray!10}{$\beta_{S,\epsilon}$} & \cellcolor{gray!10}{slope of the relationship between residual standard deviation and measured leaf surface area (log-link scale)}\\
+\hspace{1em}$\rho_{\epsilon}$ & lag-1 residual autocorrelation\\
+\addlinespace[0.3em]
+\multicolumn{2}{l}{\textbf{\aax{} for each light intensity with leaf using agcurve{} curve parameters}}\\
+\hspace{1em}\cellcolor{gray!10}{$\widehat{\mathrm{AA}}_{klm}$} & \cellcolor{gray!10}{estimate of \aax{} for the $k^{\text{th}}$ leaf at light intensity $l$ in accession $m$}\\
+\addlinespace[0.3em]
+\multicolumn{2}{l}{\textbf{effects of light intensity, light treatments, and population on \aax}}\\
+\hspace{1em}$\beta_{\mathrm{AA},0}$ & intercept of \aax{} at low light intensity in shade treatment\\
+\hspace{1em}\cellcolor{gray!10}{$\beta_{\mathrm{AA},2000}$} & \cellcolor{gray!10}{effect of high light intensity at \ppfdequals{2000} on \aax}\\
+\hspace{1em}$\beta_{\mathrm{AA},\text{sun}}$ & effect of sun treatment on \aax{}\\
+\hspace{1em}\cellcolor{gray!10}{$\beta_{\mathrm{AA},2000,\text{sun}}$} & \cellcolor{gray!10}{effect of high light intensity at \ppfdequals{2000} on \aax{} in sun treatment}\\
+\hspace{1em}$\vec{\beta}_{\mathrm{AA},\text{acc}}$ & vector of $n_\text{acc}$ phylogenetically structured random accession-level effects on \aax; $\vec{\beta}_{\mathrm{AA},\text{acc}} \sim \text{MVN}(\vec{0}, \symbf{\Sigma}_{\text{AA,acc}})$\\
+\hspace{1em}\cellcolor{gray!10}{$\symbf{\Sigma}_{\text{AA,acc}}$} & \cellcolor{gray!10}{$n_\text{acc} \times n_\text{acc}$ covariance matrix of phylogenetically structured random accession-level effects on \aax;}\\
+\hspace{1em}$\vec{\beta}_{\mathrm{AA},\text{rep}}$ & vector of $n_\text{rep}$ random replicate-level effects on \aax; $\vec{\beta}_{\mathrm{AA},\text{rep}} \sim \text{Normal}(0, \sigma_{\mathrm{AA},\text{rep}})$\\
+\hspace{1em}\cellcolor{gray!10}{$\vec{\beta}_{\mathrm{AA},\text{2000,acc}}$} & \cellcolor{gray!10}{vector of $n_\text{acc}$ phylogenetically structured random accession-level effects of high light intensity at \ppfdequals{2000} on \aax; $\vec{\beta}_{\mathrm{AA},\text{2000,acc}} \sim \text{MVN}(\vec{0}, \symbf{\Sigma}_{\mathrm{AA},\text{2000,acc}})$}\\
+\hspace{1em}$\vec{\beta}_{\mathrm{AA},\text{sun,acc}}$ & vector of $n_\text{acc}$ phylogenetically structured random accession-level effects of sun treatment on \aax; $\vec{\beta}_{\mathrm{AA},\text{sun,acc}} \sim \text{MVN}(\vec{0}, \symbf{\Sigma}_{\mathrm{AA},\text{sun,acc}})$\\
+\hspace{1em}\cellcolor{gray!10}{$\sigma_{\mathrm{AA},\epsilon,0}$} & \cellcolor{gray!10}{intercept of phylogenetically unstructured residual standard deviation of \aax}\\
+\hspace{1em}$\beta_{\mathrm{AA},\epsilon,2000}$ & effect of high light intensity at \ppfdequals{2000} on phylogenetically unstructured residual standard deviation of \aax{} (log-link scale)\\
+\hspace{1em}\cellcolor{gray!10}{$\beta_{\mathrm{AA},\epsilon,\text{sun}}$} & \cellcolor{gray!10}{effect of sun treatment on phylogenetically unstructured residual standard deviation of \aax{} (log-link scale)}\\
+\hspace{1em}$\sigma_{\mathrm{AA},\text{rep}}$ & standard deviation of random replicate-level effects on \aax\\
+\hspace{1em}\cellcolor{gray!10}{$\alpha_{\mathrm{AA},\text{acc}}$} & \cellcolor{gray!10}{decay rate of phylogenetic covariance in random accession-level effects on \aax}\\
+\hspace{1em}$\sigma^2_{\mathrm{AA},\text{acc}}$ & phylogenetic diffusion rate in random accession-level effects on \aax\\
+\hspace{1em}\cellcolor{gray!10}{$\alpha_{\mathrm{AA},\text{2000,acc}}$} & \cellcolor{gray!10}{decay rate of phylogenetic covariance in random accession-level effects of high light intensity at \ppfdequals{2000} on \aax}\\
+\hspace{1em}$\sigma^2_{\mathrm{AA},\text{2000,acc}}$ & phylogenetic diffusion rate in random accession-level effects of high light intensity at \ppfdequals{2000} on \aax\\
+\hspace{1em}\cellcolor{gray!10}{$\alpha_{\mathrm{AA},\text{sun,acc}}$} & \cellcolor{gray!10}{decay rate of phylogenetic covariance in random accession-level effects of sun treatment on \aax}\\
+\hspace{1em}$\sigma^2_{\mathrm{AA},\text{sun,acc}}$ & phylogenetic diffusion rate in random accession-level effects of sun treatment on \aax\\
+\addlinespace[0.3em]
+\multicolumn{2}{l}{\textbf{effects of native light habitat on population-level \aax}}\\
+\hspace{1em}\hspace{1em}\cellcolor{gray!10}{$\beta_{\mathrm{AA},\text{PPFD},0}$} & \cellcolor{gray!10}{intercept of accession-level \aax{} when native \ppfdequals{0}}\\
+\hspace{1em}\hspace{1em}$\beta_{\mathrm{AA},\text{PPFD},1}$ & slope of native \ppfd{} on accession-level \aax{}\\
+\hspace{1em}\cellcolor{gray!10}{$\alpha_{\mathrm{AA},\text{PPFD}}$} & \cellcolor{gray!10}{decay rate of phylogenetic covariance in residuals of model testing effect of native \ppfd{} on accession-level \aax}\\
+\hspace{1em}$\sigma^2_{\mathrm{AA},\text{PPFD}}$ & phylogenetic diffusion rate in residuals of model testing effect of native \ppfd{} on accession-level \aax\\
+\bottomrule
+
+\end{longtable}
+
+\subparagraph{\texorpdfstring{\agcurve{} curve
+parameters}{ curve parameters}}\label{curve-parameters}
+
+We modeled \logA as a quadratic function of \loggsw for each leaf using
+the following equation:
+
+\[\log(A_{ij}) = (\beta_0 + b_{0,j}) + (\beta_1 + b_{1,j}) \log{g_{\text{sw},i}} + (\beta_2 + b_{2,j}) \log{g_{\text{sw},i}}^2 + \epsilon_{i}\]
+
+where \(\beta_0\), \(\beta_1\), and \(\beta_2\) are the average
+intercept, linear, and quadratic coefficients, respectively. We used
+diffuse normal priors with mean \(0\) and standard deviation \(10\) on
+these parameters. We estimated random effects of curve \(j\) on the
+intercept (\(b_{0,j}\)), linear (\(b_{1,j}\)), and quadratic
+(\(b_{2,j}\)) coefficients. We assumed that the \(j \times 3\) array of
+coefficients was multivariate normal a mean vector of \(\vec{0}\) and
+covariance \(\Sigma_\text{curve}\). We used a weakly informative normal
+prior with mean \(0\) and standard deviation \(1\) on the
+log-transformed standard deviations (i.e.~the diagonal of
+\(\Sigma_\text{curve}\)). We used a weakly informative
+\(\mathrm{LJK}(2)\) prior on the correlation matrix. The off diagonal
+elements of \(\Sigma_\text{curve}\) can be calculated from its diagonal
+elements and the correlation matrix.
+
+The residuals \(\epsilon_{i}\) were modeled as a lag-1 autocorrelated
+time-series. We further assumed that the residual standard deviation of
+the \(j^{\text{th}}\) curve (\(\sigma_{\epsilon,j}\)) was inversely
+proportional to the leaf surface area (\(S_j\)) within the chamber:
+
+\[\log (\sigma_{\epsilon,j}) = \log (\sigma_{\qty{6}{\raiseto{2}\centi\meter},\epsilon}) + (6 - S_j) \beta_{S,\epsilon}\]
+where \(\sigma_{\qty{6}{\raiseto{2}\centi\meter},\epsilon}\) is the
+minimum residual standard deviation when the
+\(\qty{6}{\raiseto{2}\centi\meter}\) chamber is completely filled. The
+residual standard deviation increases on log-linear scale by
+\(\beta_{S,\epsilon}\). We used a weakly informative normal prior with
+mean \(-3\) and standard deviation \(5\) on
+\(\log (\sigma_{\qty{6}{\raiseto{2}\centi\meter},\epsilon})\) and a
+weakly informative normal prior with mean \(0\) and standard deviation
+\(1\) on \(\beta_{S,\epsilon}\).
+
+\subparagraph{\texorpdfstring{\aax{} for each light intensity with leaf
+using \agcurve{} curve
+parameters}{ for each light intensity with leaf using  curve parameters}}\label{for-each-light-intensity-with-leaf-using-curve-parameters}
+
+Within the \(k^{\text{th}}\) leaf, we estimated \aax{} for each light
+intensity by integrating the difference in \logA{} between the amphi and
+pseudohypo \agcurve{} curves over the range of \gsw{} values where the
+curves overlap (from \(\min(\log(g_\text{sw}))\) to
+\(\max(\log(g_\text{sw}))\)). The estimate of \aax{} for the
+\(k^{\text{th}}\) leaf at light intensity \(l\) in population \(m\) is:
+
+\[\widehat{\mathrm{AA}}_{klm} = \int_{\text{min(log(}g_\text{sw}))}^{\text{max(log(}g_\text{sw}))} \text{log}\bigg(\frac{\hat{A}_\text{amphi}(x; \theta_{klm,\text{amphi})}}{\hat{A}_\text{hypo}(x; \theta_{klm,\text{hypo})}}\bigg) dx\]
+where:
+
+\[\theta_\text{amphi} \in \{\hat{b}_{0, f(\text{amphi}, k,l,m)}, \hat{b}_{1, f(\text{amphi}, k,l,m)}, \hat{b}_{2, f(\text{amphi}, k,l,m)}\}, \text{and}\]
+
+\[\theta_\text{hypo} \in \{\hat{b}_{0, f(\text{hypo}, k,l,m)}, \hat{b}_{1, f(\text{hypo}, k,l,m)}, \hat{b}_{2, f(\text{hypo}, k,l,m)}\}.\]
+The function \(f : \Theta_1 \rightarrow \Theta_2\) maps the set
+\(\Theta_1\) indexed by leaf type (amphi or pseudohypo), leaf replicate,
+light intensity, and population to set \(\Theta_2\) indexed by
+individual \agcurve{} curve. This mapping is necessary because the
+random effects structure differs between models of \loggsw{} on \logA{}
+and that of models predicting \aax{} described in the next section.
+
+\subparagraph{\texorpdfstring{Effects of light intensity, light
+treatment, and population on
+\aax}{Effects of light intensity, light treatment, and population on }}\label{effects-of-light-intensity-light-treatment-and-population-on}
+
+We tested for effects of light intensity, light treatment, population,
+and their interactions on \aax. All models included effects of light
+intensity and light treatment, as well as random effects of population
+and replicate within population. More complex models included
+interactions between light intensity and light treatment, as well as
+random effects of population on the effects of light intensity and light
+treatment. We used a weakly informative normal prior with mean \(0\) and
+standard deviation \(10\) for fixed effects of light intensity and
+treatment. We used a weakly informative normal prior with mean \(-3\)
+and standard deviation \(5\) for the the random effect standard
+deviation of replicate within population. We accounted for the
+phylogenetic structure among the random effects of population using an
+Ornstein-Uhlenbeck (OU) process (\emph{66}). The expected covariance
+between populations \(i\) and \(j\) (\(\text{Cov}(i, j)\) ) is:
+
+\[\text{Cov}(i, j) = \frac{\sigma^2}{2 \alpha} \exp(-\alpha D_{ij})\]
+where \(\sigma^2\) is the variance of the random effect, \(\alpha\) is
+the rate of decay of the covariance with phylogenetic distance,
+\(D_{ij}\), the phylogenetic distance between populations \(i\) and
+\(j\). We estimated \(\sigma^2 / (2 \alpha)\) and \(\alpha\) as a
+separate parameters and reparameterized them as \(\sigma^2\) and
+\(\alpha\). We used a weakly informative normal prior with mean \(0\)
+and standard deviation \(10\) on OU parameters.
+
+We modeled the residual standard deviation of \aax,
+(i.e.~phylogenetically unstructured variation unaccounted for by
+explanatory variables) on a log-link scale with effects of light
+intensity and light treatment. We used a weakly informative normal prior
+with mean \(-3\) and standard deviation \(5\) on the residual standard
+deviation intercept and a weakly informative normal prior with mean
+\(0\) and standard deviation \(1\) on the effects of light intensity and
+light treatment on the residual standard deviation.
+
+\subparagraph{\texorpdfstring{Effects of native light habitat on
+population-level
+\aax}{Effects of native light habitat on population-level }}\label{effects-of-native-light-habitat-on-population-level}
+
+We tested a linear effect of native light habitat on population-level
+\aax. In Models 1 and 2, we used the random intercept of \aax{} as a
+response variable; in model 3 we used the random intercept plus the
+random effect of population at high light intensity; in model 4 we used
+the random intercept plus the random effect of population in the sun
+treatment; in model 5 we used the random intercept plus the random
+effects of population at high light intensity and sun treatment. We used
+a weakly informative normal prior with mean \(0\) and standard deviation
+\(1\) for the slope and intercept. We accounted for the phylogenetic
+structure among the model residuals using an (OU) process. We used a
+weakly informative normal prior with mean \(0\) and standard deviation
+\(10\) on OU parameters.
+
+\paragraph{Predictions}\label{sec-predictions}
+
+The assimilatory, plastic, and constitutive hypotheses make different
+predictions about the relationship between \aax{} and light intensity,
+light treatment, and native \ppfd{} among populations. Since these
+hypotheses are not mutually exclusive, we describe how we assessed
+support for one, two, or all three hypotheses simultaneously in
+\autoref{tbl-predictions}. In general, the assimilatory hypothesis was
+supported if \aax{} was greater at high light intensity than low light
+intensity. The plastic hypothesis was supported if \aax{} was greater in
+sun leaves than shade leaves. The constitutive hypothesis was supported
+if population-level \aax{} increased with native \ppfd. In interactive
+models, we only consider positively reinforcing interactions between
+high light intensity, sun leaves, and native \ppfd{} because these are
+the only interactions which could explain why amphistomatous leaves are
+advantageous in high light habitats.
+
+We evaluated predictions using a combination of parameter estimation and
+model selection. In all cases, we estimated parameters and confidence
+intervals as described in Section~\ref{sec-fitting}. If the 95\%
+confidence intervals for a parameter did not overlap zero, we considered
+the parameter to be significantly different from zero. We used the
+leave-one-out cross-validation information criterion (LOOIC) to compare
+the fit of models (\autoref{tbl-models}) using the \emph{R} package
+\textbf{loo} version 2.8.0 (\emph{67}) to calculate LOOIC values. Note
+that LOOIC was calculated only from pointwise likelihood values of the
+submodel estimating effects of light intensity, light treatments, and
+population on \aax{} (see Section~\ref{sec-parameters}). We considered
+models with two standard errors of the mean lower LOOIC value to be a
+better fit to the data; we considered models with LOOIC values within
+two standard errors of the mean to be have similar support.
+
+\begin{longtable}{>{\raggedright\arraybackslash}p{1in}>{\raggedright\arraybackslash}p{1.5in}>{\raggedright\arraybackslash}p{3in}}
+
+\caption{\label{tbl-predictions}Predictions of competing hypotheses
+about the relationship between \aax{} and light intensity, light
+treatment, and native \ppfd{} among populations. The middle column lists
+specific directional predictions about parameter values and model fit
+according to leave-one-out cross-validation information criterion
+(LOOIC), where \(\text{LOOIC}_i\) is the LOOIC value for model \(i\).
+The rightmost column describes the predictions in words and explains how
+population-level \aax{} is calculated in the relevant model.}
+
+\tabularnewline
+
+\toprule
+\textbf{Hypothesis} & \textbf{Prediction(s)} & \textbf{Description}\\
+\midrule
+Assimilatory & $\beta_{\mathrm{AA},2000} > 0$ & \hspace{-1em}Average \aax{} at high light intensity is greater than that at low light intensity\\
+\cmidrule{1-3}\pagebreak[0]
+Plastic & $\beta_{\mathrm{AA},\text{sun}} > 0$ & \hspace{-1em}Average \aax{} in sun leaves is greater than that in shade leaves\\
+\cmidrule{1-3}\pagebreak[0]
+Constitutive & $\beta_{\mathrm{PPFD,AA}} > 0$ & \hspace{-1em}Population-level \aax{} ($\mathrm{AA}_\text{acc}$) increases with native \ppfd\\
+\nopagebreak
+ &  & \hspace{-1em}$\mathrm{AA}_\text{acc} = \beta_{\mathrm{AA}, 0} + \vec{\beta}_{\mathrm{AA}, \text{acc}}$\\
+\cmidrule{1-3}\pagebreak[0]
+Assimilatory $\times$ Plastic & $\beta_{\mathrm{AA},2000,\text{sun}} > 0$ & \hspace{-1em}Average \aax{} is highest at high light intensity in sun leaves\\
+\nopagebreak
+ & $\text{LOOIC}_\text{1} > \text{LOOIC}_\text{2}$ & \hspace{-1em}\\
+\cmidrule{1-3}\pagebreak[0]
+Assimilatory $\times$ Constitutive & $\beta_{\mathrm{AA},2000} > 0$ & \hspace{-1em}Average \aax{} at high light intensity is greater than that at low light intensity\\
+\nopagebreak
+ & $\beta_{\mathrm{PPFD,AA}} > 0$ & \hspace{-1em}Population-level \aax{} at high light intensity ($\mathrm{AA}_{\text{acc},2000}$) increases with native \ppfd\\
+\nopagebreak
+ & $\text{LOOIC}_\text{1} > \text{LOOIC}_\text{3}$ & \hspace{-1em}$\mathrm{AA}_{\text{acc},2000} = \beta_{\mathrm{AA}, 0} + \vec{\beta}_{\mathrm{AA}, \text{acc}} + \vec{\beta}_{\mathrm{AA}, 2000, \text{acc}}$\\
+\cmidrule{1-3}\pagebreak[0]
+Plastic $\times$ Constitutive & $\beta_{\mathrm{AA},\text{sun}} > 0$ & \hspace{-1em}Average \aax{} in sun leaves is greater than that in shade leaves\\
+\nopagebreak
+ & $\beta_{\mathrm{PPFD,AA}} > 0$ & \hspace{-1em}Population-level \aax{} in sun leaves ($\mathrm{AA}_{\text{acc},\text{sun}}$) increases with native \ppfd\\
+\nopagebreak
+ & $\text{LOOIC}_\text{1} > \text{LOOIC}_\text{4}$ & \hspace{-1em}$\mathrm{AA}_{\text{acc},\text{sun}} = \beta_{\mathrm{AA}, 0} + \vec{\beta}_{\mathrm{AA}, \text{acc}} + \vec{\beta}_{\mathrm{AA}, \text{sun}, \text{acc}}$\\
+\cmidrule{1-3}\pagebreak[0]
+Assimilatory $\times$ Plastic $\times$ Constitutive & $\beta_{\mathrm{AA},2000,\text{sun}} > 0$ & \hspace{-1em}Average \aax{} is highest at high light intensity in sun leaves\\
+\nopagebreak
+ & $\beta_{\mathrm{PPFD,AA}} > 0$ & \hspace{-1em}Population-level \aax{} at high light intensity in sun leaves ($\mathrm{AA}_{\text{acc},2000,\text{sun}}$) increases with native \ppfd\\
+\nopagebreak
+ & $\text{LOOIC}_\text{3} > \text{LOOIC}_\text{5}$ & \hspace{-1em}$\mathrm{AA}_{\text{acc},2000,\text{sun}} = \beta_{\mathrm{AA}, 0} + \vec{\beta}_{\mathrm{AA}, \text{acc}} + \vec{\beta}_{\mathrm{AA}, 2000, \text{acc}} + \vec{\beta}_{\mathrm{AA}, \text{sun}, \text{acc}}$\\
+\bottomrule
+
+\end{longtable}
+
+\subsection{Results}\label{results}
+
+\begin{itemize}
+\tightlist
+\item
+  Variation in stomatal ratio among pops and light treatments
+\item
+  Table S7: Model comparison LOOIC tables
+\item
+  Table S8: Model comparison LOOIC tables with LMA
+\item
+  Table S9: Calculation of alternate scenario where amphi leaves can get
+  same A with less R
+\end{itemize}
+
+\newpage{}
+
+\subsubsection{MANUSCRAPS}\label{manuscraps}
+
+(not sure where this goes) * We acclimated the focal leaf to high light
+(\ppfdequals{2000}) and high relative humidity (\rhequals{70}) until
+\(A\) and \gsw{} reach their maximum. After that, we decreased \rh{} to
+\(\approx 10\%\) to induce rapid stomatal closure without biochemical
+down
+
+Problems to bring up in discussion: - all our species were amphi, so
+they may all live in PAI below threshold to favor amphi - other
+microclimate diffs not considered - don't know specific anatomical
+change
+
+Two outstanding areas for further research are 1) the specific
+anatomical changes, such as leaf thickness or mesophyll porosity, that
+increase the diffusion path length for CO\(_2\) in sun leaves, and 2)
+the costs of amphistomy that could explain plasticity in \gmaxratio{} in
+populations from low-light habitats. Resolving the first question will
+require detailed anatomical measurements on a large number of shade and
+sun leaves with concomitant estimates of \aax. The fact that populations
+from low-light habitats are more plastic suggests that costs of
+amphistomy are greatest in these environments. (not finished yet)
+
+Not sure where this goes: We infer this from the fact that blocking gas
+exchange in pseudohypostomatous leaves reduced \(A\) by X-X\% depending
+on the population, light treatment, and light intensity (Table/figure).
+The AA is equivalent to an X-X\% change in total \(g_\text{sc}\) (see SI
+section gs equivalency). But whereas increasing \(g_\text{sc}\) would
+increase water loss as a necessary by-product, amphistomy can increase
+\(A\) without any appreciable affect on transpiration.
+
+from Mott and Oleary: ``However, Jones and Slatyer (5) reported a higher
+mesophyll resistance for CO, entering through the upper stomata than for
+the lower, and the data of Vaclavik (12) appear to support this
+conclusion''
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\setcounter{enumi}{7}
+\tightlist
+\item
+  What it all means
+\end{enumerate}
+
+\begin{itemize}
+\tightlist
+\item
+  Our study reveals an important role for plasticity in explaining why
+  amphi leaves are more common in sunny habitats
+\item
+  Confirms that leaf structure of sun leaves incurs costs that
+  amphistomy can mitigate
+\item
+  It will be useful to consider coordination between leaf anatomy,
+  stomatal density, and ratio in future studies of leaf
+  structure-function and in crops (or something)
+\end{itemize}
+
+I am using this approach to drafting the paper:
+
+https://blogs.nature.com/nyc/2011/08/10/how-to-write-a-paper-one-possible-answer
+
+\subsection{Punchline}\label{punchline}
+
+Plastic changes in the structure of sun leaves can increase the
+intercellular CO\(_2\) diffusion path length. Gas exchange through
+stomata on both upper and lower leaf surfaces, amphistomy, may be a
+compensatory mechanism to reduce the impact of increased diffusion path
+length on photosynthesis without increasing water loss. We demonstrate
+for the first time that sun leaves benefit most from amphistomy because
+it compensates for plastic structural changes in leaf anatomy.
+
+What's surprising: role of developmental plasticity in explaining
+comparative results; diffusional limitations through palisade ias is
+strong and needs to be considered with light gradients in understanding
+leaf strucuture- funciton
+
+big implications - if trees were amphi, more more carbon sink or reduced
+transpiration
+
+It is commonly assumed in comparative studies that most trait variation
+between species is constitutive and genetically determined, but our
+results suggest that plastic responses to light are necessary to explain
+why \gmaxratio{} increases with light habitat. Reaction norms to light
+intensity among wild tomatoes indicate that populations from low light
+habitats express low \gmaxratio{} when grown in shade, whereas
+populations from high light habitats have constitutively higher
+\gmaxratio{} (Fig. rxn-norm). The larger response to light intensity is
+driven by greater plasticty in leaf structure in populations from
+low-light habitats (Fig/table - I haven't tested this yet, so this is a
+placeholder.) We infer that these structural changes in sun leaves are
+costly because they increase the diffusion path length for CO\(_2\) and
+reduce photosynthesis. As a consequence, sun leaves benefit more from
+amphistomy. This is the first demonstration that sun leaves benefit more
+from amphistomy because it compensates for developmental plasticity in
+leaf structure.
+
 \end{document}

### 2025-07-04 r/208_plot-pai.R

@@ -36,18 +36,23 @@ df_aa_pred1 = posterior_epred(fit_aa1, newdata = df_new) |>
       fct_recode(low = "150", high = "2000")
   )
 
-ggplot(df_aa_pred1, aes(pai, aa, color = Growth, shape = Measurement)) +
+fig_aa_pai = ggplot(df_aa_pred1, aes(pai, aa, color = Growth, shape = Measurement)) +
   facet_grid(Measurement ~ Growth) +
   geom_point() +
   scale_color_manual(values = c("shade" = "tomato4", "sun" = "tomato")) +
   scale_shape_manual(values = c("low" = 19, "high" = 21)) +
   scale_x_continuous(breaks = c(0.01, 0.1, 1), trans = reverselog10_trans()) +
-  xlab(expression(paste("plant area index [", m^2~m^-2, "]"))) +
+  xlab(expression(paste("native plant area index [", m^2~m^-2, "]"))) +
   ylab("amphi advantage") +
+  ylim(0, 0.2) +
+  geom_hline(yintercept = 0, linetype = "dashed") +
   theme(legend.position = "none")
+
+write_rds(fig_aa_pai, "objects/fig_aa_pai.rds")
 
 ggsave(
   "figures/pai-aa.pdf",
+  fig_aa_pai,
   width = 6,
   height = 4,
   device = cairo_pdf,

### 2025-07-04 r/209_plot-lma.R

@@ -5,13 +5,12 @@ fit_aa = read_rds("objects/fit_aa2.rds")
 
 # Estimated llma for each acc, light_intensity, and light_treatment
 df_new1 = crossing(
-    # acc = unique(fit_aa$data$acc),
-    acc_id = unique(fit_aa$data$acc_id),
+    acc = unique(fit_aa$data$acc),
+    # acc_id = unique(fit_aa$data$acc_id),
     light_intensity = unique(fit_aa$data$light_intensity),
     light_treatment = unique(fit_aa$data$light_treatment)
   ) |>
-  mutate(acc = str_extract(acc_id, acc_string),
-         row = row_number())
+  mutate(row = row_number())
 
 df_pred1 = posterior_epred(fit_aa, newdata = df_new1, resp = "llma") |>
   as_draws_df() |>
@@ -74,7 +73,7 @@ df2 = ce[["aa.aa_llma:light_intensity"]] |>
            fct_recode(low = "150", high = "2000"), Growth = NA)
   
 # Plot
-ggplot(df1, aes(x = lma, y = aa, shape = Measurement)) +
+fig_lma_aa = ggplot(df1, aes(x = lma, y = aa, shape = Measurement)) +
   geom_ribbon(data = df2, mapping = aes(ymin = .lower, ymax = .upper, group = Measurement), alpha = 0.5, color = NA, fill = "grey") + 
   geom_line(data = df2, aes(linetype = Measurement), linewidth = 1.1) +
   # geom_interval(aes(ymin = aa_lower, ymax = aa_upper), linewidth = 0.5) +
@@ -89,7 +88,8 @@ ggplot(df1, aes(x = lma, y = aa, shape = Measurement)) +
   xlim(min(df1$lma), max(df1$lma)) +
   theme(legend.key.width = unit(1.5, "cm")) 
 
-ggsave("figures/lma-aa.pdf", width = 6, height = 4)
+write_rds(fig_lma_aa, "objects/fig_lma_aa.rds")
+ggsave("figures/lma-aa.pdf", fig_lma_aa, width = 6, height = 4)
 
 
 # other code. not using at the moment

### 2025-07-04 ms/ms.qmd

@@
 
-```{r, aa, echo = FALSE,warning=FALSE, message=FALSE} 
+```{r, lma_aa, echo = FALSE,warning=FALSE, message=FALSE} 
 #| label: fig-lma_aa
@@
 Developmental plasticity needs to be considered when interpreting phylogenetic comparative analyses of adaptation, because plastic differences may be mistaken for genetic differences.
+
+maybe this wording is better: Amphistomy likely plays a significant, but under appreciated, role in the carbon and water balance of plant groups where this trait is common, such as herbaceous plants in sunny habitats and woody species in arid biomes (cites).
+If our estimates of AA from the wild tomato clades
 

### 2025-07-04 ms/manuscraps.qmd
added aa_loo1 table to SI

--- 2025-07-03 ms/ms.qmd
@@
 ---
-title: Unknown
+title: manuscraps
 format:
   pdf:
-    keep-tex: true  
-    toc: true
-number-sections: true
-header-includes: |
-  \usepackage{hyperref}
-  \addtokomafont{disposition}{\rmfamily}
-  \usepackage{siunitx}
-  \newcommand{\aax}{$\mathrm{AA}$}
-  \newcommand{\Aamphi}{$A_{\mathrm{amphi}}$}
-  \newcommand{\Ahypo}{$A_{\mathrm{hypo}}$}
-  \newcommand{\agcurve}{$A \textendash g_\text{sw}$}
-  \newcommand{\aqcurve}{$A \textendash Q$}
-  \newcommand{\ca}{$C_\mathrm{a}$}
-  \newcommand{\caequals}[1]{$C_\mathrm{a} = \qty{#1}{\micro\mol\raiseto{-1}\mol}$}
-  \newcommand{\cabetween}[2]{#1 < $C_\mathrm{a} < \qty{#2}{\micro\mol\raiseto{-1}\mol}$}
-  \newcommand{\gmaxratio}{$g_\text{max,ratio}$}
-  \newcommand{\gsw}{$g_\text{sw}$}
-  \newcommand{\loggsw}{$\log(g_\text{sw})$}
-  \newcommand{\logA}{$\log(A)$}
-  \newcommand{\ppfd}{$\mathrm{PPFD}$}
-  \newcommand{\ppfdequals}[1]{$\mathrm{PPFD} = \qty{#1}{\micro\mol\raiseto{-2}\meter\raiseto{-1}\second}$}
-  \newcommand{\ppfdqty}[1]{$\qty{#1}{\micro\mol\raiseto{-2}\meter\raiseto{-1}\second}$}
-  \newcommand{\rh}{$\mathrm{RH}$}
-  \newcommand{\rhequals}[1]{$\mathrm{RH} = #1\%$}
-  \newcommand{\tleafequals}[1]{$T_\mathrm{leaf} = \qty{#1}{\degreeCelsius}$}
-mainfont: Times New Roman
+    keep-tex: false  
+    toc: false
+number-sections: false
 bibliography: solanum-aa.bib
@@
 
-```{r load-objects, echo=FALSE, eval=TRUE, message=FALSE, warning=FALSE}
-library(dplyr)
-library(lubridate)
-library(magrittr)
-library(readr)
-library(stringr)
-library(tibble)
-library(tidyr)
+### MANUSCRAPS
 
-source("../r/functions.R")
+(not sure where this goes)  * We acclimated the focal leaf to high light (\ppfdequals{2000}) and high relative humidity (\rhequals{70}) until $A$ and \gsw{} reach their maximum.
+After that, we decreased \rh{} to $\approx 10\%$ to induce rapid stomatal closure without biochemical down
 
-aa_args = read_rds("../objects/aa_args1.rds")
-aa_stats = read_rds("../objects/aa_stats.rds")
-# sort list by name
-aa_stats = aa_stats[order(names(aa_stats))]
+Problems to bring up in discussion:
+- all our species were amphi, so they may all live in PAI below threshold to favor amphi
+- other microclimate diffs not considered
+- don't know specific anatomical change
 
-plant_info = read_rds("../data/plant-info.rds")
+Two outstanding areas for further research are 1) the specific anatomical changes, such as leaf thickness or mesophyll porosity, that increase the diffusion path length for CO$_2$ in sun leaves, and 2) the costs of amphistomy that could explain plasticity in \gmaxratio{} in populations from low-light habitats.
+Resolving the first question will require detailed anatomical measurements on a large number of shade and sun leaves with concomitant estimates of \aax.
+The fact that populations from low-light habitats are more plastic suggests that costs of amphistomy are greatest in these environments.
+(not finished yet)
 
-rh_curves = read_rds("../data/trimmed_rh_curves.rds")
+Not sure where this goes: We infer this from the fact that blocking gas exchange in pseudohypostomatous leaves reduced $A$ by X-X% depending on the population, light treatment, and light intensity (Table/figure).
+The AA is equivalent to an X-X% change in total $g_\text{sc}$ (see SI section gs equivalency).
+But whereas increasing $g_\text{sc}$ would increase water loss as a necessary by-product, amphistomy can increase $A$ without any appreciable affect on transpiration.
 
-stomata = read_rds("../data/stomata.rds")
+from Mott and Oleary: "However, Jones and Slatyer (5) reported a higher mesophyll resistance for CO, entering through the upper stomata than for the lower, and the data of Vaclavik (12) appear to support this conclusion"
 
-fit_aa = read_rds("../objects/fit_aa1.rds")
-dim_fit_aa = dim(fit_aa$fit)
-
-df_aa_pred1 = read_rds("../objects/df_aa_pred1.rds")
-# df_aa_pred2 = read_rds("../objects/df_aa_pred2.rds")
-df_aa_pred3 = read_rds("../objects/df_aa_pred3.rds")
-
-n_plant = length(unique(rh_curves$acc_id))
-n_acc = length(unique(rh_curves$acc))
-# average number of individual replicates for accession per light treatment
-mean_n_rep = rh_curves |>
-  summarize(
-    n = length(unique(acc_id)),
-    .by = c("acc", "light_treatment")
-  ) |>
-  pull(n) |>
-  mean() |>
-  prettyNum(digits = 3)
-
-licor_date_range = plant_info |>
-  filter(acc_id %in% unique(rh_curves$acc_id)) |>
-  dplyr::select(`1s_rh_response_date`, `2s_rh_response_date`) |>
-  pivot_longer(cols = everything()) |>
-  pull(value) |>
-  ymd() |>
-  range(na.rm = TRUE)
-
-n_weeks = prettyNum(as.numeric(difftime(licor_date_range[2], licor_date_range[1], units = "days")) / 7, digits = 3)
- # number of weeks for LICOR measurements
-start_licor_date = format(licor_date_range[1], "%B %d, %Y")
-end_licor_date = format(licor_date_range[2], "%B %d, %Y")
-
-# Environmental data
-df_germ_summary = read_rds("../data/df_germ_summary.rds") |>
-  ungroup() |>
-  summarize(value = median(value), .by = c("period", "variable", "parameter"))
-
-df_growth_summary = read_rds("../data/df_growth_summary.rds") |>
-  ungroup() |>
-  filter(!is.na(value))
-
-# environmental condition in germination room
-
-germ_temp_day = df_germ_summary |>
-  filter(period == "day", variable == "degC", parameter == "mu") |>
-  pull(value) |>
-  prettyNum(digits = 3)
-
-germ_temp_night = df_germ_summary |>
-  filter(period == "night", variable == "degC", parameter == "mu") |>
-  pull(value) |>
-  prettyNum(digits = 3)
-
-germ_rh_day = df_germ_summary |>
-  filter(period == "day", variable == "RH", parameter == "mu") |>
-  pull(value) |>
-  prettyNum(digits = 3)
-
-germ_rh_night = df_germ_summary |>
-  filter(period == "night", variable == "RH", parameter == "mu") |>
-  pull(value) |>
-  prettyNum(digits = 3)
-
-germ_ppfd = df_germ_summary |>
-  filter(period == "day", variable == "PAR", parameter == "mu") |>
-  pull(value) |>
-  prettyNum(digits = 3)
-
-# environmental conditions in growth room
-
-d1 = df_growth_summary |>
-  summarize(value = median(value),
-            .by = c("period", "rack", "variable", "parameter"))
-
-ppfd_sun = d1 |>
-  filter(period == "day", rack == "lower", variable == "PAR", parameter == "mu") |>
-  pull(value) |>
-  prettyNum(digits = 3)
-
-ppfd_shade = d1 |>
-  filter(period == "day", rack == "upper", variable == "PAR", parameter == "mu") |>
-  pull(value) |>
-  prettyNum(digits = 3)
-
-# degrees warmer that the sun treatment air was
-temp_diff = df_growth_summary |>
-  filter(period == "day", rack != "hobo", variable == "degC", parameter == "mu") |>
-  pivot_wider(names_from = "rack") |>
-  mutate(diff = lower - upper) |>
-  summarize(value = median(diff, na.rm = TRUE)) |>
-  pull(value) |>
-  prettyNum(digits = 3)
-
-# difference in RH
-rh_diff = df_growth_summary |>
-  filter(period == "day", rack != "hobo", variable == "RH", parameter == "mu") |>
-  pivot_wider(names_from = "rack") |>
-  mutate(diff = upper - lower) |>
-  summarize(value = median(diff, na.rm = TRUE)) |>
-  pull(value) |>
-  prettyNum(digits = 3)
-
-df_leaf_temp = plant_info |>
-  filter(!is.na(leaf_temperature_C)) |>
-  summarize(
-    leaf_temperature_C = mean(leaf_temperature_C),
-    n = n(),
-    .by = "light_treatment"
-  )
-
-# difference in leaf temperature between treatments
-leaf_temp_diff = df_leaf_temp$leaf_temperature_C |>
-  diff() |>
-  prettyNum(digits = 3)
-
-# number of leaves for which we have measured leaf temp under ambient conditions
-n_leaf_temp = sum(df_leaf_temp$n)
-
-# data on leaflet type
-df_leaflet_type = plant_info |>
-  summarize(n = n(), .by = "leaflet") |>
-  pivot_wider(names_from = "leaflet", values_from = "n") |>
-  mutate(
-    percent_terminal = prettyNum(terminal / (terminal + lateral) * 100, digits = 3),
-    percent_lateral = prettyNum(lateral / (terminal + lateral) * 100, digits = 3)
-  )
-
-percent_terminal = df_leaflet_type$percent_terminal
-percent_lateral = df_leaflet_type$percent_lateral
-# percent of time that anatomical traits were measured on different leaflet than LICOR measurements
-percent_change_leaflet = full_join(
-  plant_info |>
-    filter(acc_id %in% unique(rh_curves$acc_id)) |>
-    select(acc_id, licor = leaflet),
-  stomata |>
-    mutate(acc_id = str_c(accession, replicate, sep = "-")) |>
-    filter(acc_id %in% unique(rh_curves$acc_id)) |>
-    select(acc_id, stomata = leaflet),
-  by = join_by(acc_id)
-) |>
-  summarize(x = mean(licor != stomata, na.rm = TRUE)) |>
-  pull(x) |>
-  multiply_by(100) |>
-  prettyNum(digits = 3)
-
- # percent of times where AA and AQ curves were done on same leaflet
-percent_aa_and_aq = plant_info |>
-  filter(acc_id %in% unique(rh_curves$acc_id)) |>
-  mutate(same_leaflets = !switch_leaflets) |>
-  pull(same_leaflets) |>
-  mean() |>
-  multiply_by(100) |>
-  prettyNum(digits = 3)
-
-# data on chamber light conditions
-percent_red = rh_curves |>
-  mutate(p_red = Q_red / Q) |>
-  summarize(p_red = mean(p_red, na.rm = TRUE),
-            .by = c(acc_id, light_intensity)) |>
-  summarize(p_red = mean(p_red, na.rm = TRUE), .by = light_intensity) |>
-  pivot_wider(names_from = light_intensity, values_from = p_red)
-
-p_red_high = format(signif(100 * percent_red$`2000`, 3), nsmall = 1)
-p_blue_high = format(signif(100 * (1 - percent_red$`2000`), 3), nsmall = 1)
-p_red_low = format(signif(100 * percent_red$`150`, 3), nsmall = 1)
-p_blue_low = format(signif(100 * (1 - percent_red$`150`), 3), nsmall = 1)
-
-# percent measure LMA on alternate leaflet
-percent_lma_switch = plant_info |>
-  filter(acc_id %in% rh_curves$acc_id) |> 
-  summarize(x = mean(leaflet != lma_leaflet, na.rm = TRUE)) |>
-  pull(x) |>
-  multiply_by(100) |>
-  prettyNum(digits = 3)
-
- # percent leaves missing LMA
-percent_lma_missing = plant_info |>
-  filter(acc_id %in% rh_curves$acc_id) |> 
-  summarize(x = mean(is.na(lma_leaflet))) |>
-  pull(x) |>
-  multiply_by(100) |>
-  prettyNum(digits = 3)
-
-### Cleaning RH curves
-ca_min = aa_args$ca_min
-ca_max = aa_args$ca_max
-n_min_points_per_rh_curve = aa_args$n_min_points_per_rh_curve
-point_outlier_threshold = aa_args$point_outlier_threshold
-aa_outlier_threshold = aa_args$aa_outlier_threshold
-thinning_interval = aa_args$thinning_interval
-
-### Cleaning AQ curves
-aq_r2_threshold = aa_args$aq_r2_threshold
-
-### Bayesian data analysis
-n_model = 5 |> # i don't think I need this anymore
-  number_to_word()
-
-### Summarizing AA results
-aa_mean_range = df_aa_pred1 |>
-  ungroup() |>
-  mutate(signif = sign(.lower) == sign(.upper)) |>
-  summarize(
-    min = min(aa),
-    max = max(aa),
-    aa = mean(aa),
-    n_signif = sum(signif),
-    n_acc = n(),
-    .by = c(light_treatment, light_intensity)
-  ) |>
-  mutate(x = sprintf(
-    "%.3f (range: %.3f–%.3f; %.0f of %.0f accessions significant)",
-    aa,
-    min,
-    max,
-    n_signif,
-    n_acc
-  ))
-
-aa_summary3 = df_aa_pred3 |>
-  mutate(signif = (.upper < 0)) |>
-  summarize(n_signif = sum(signif), .by = light_treatment) |>
-  pull(n_signif, light_treatment)
-
-```
-
-## Notes
-
-Terminology to standardize:
-
--   the native light habitat or environment: the SPLASH paper refers to what they calculated as 'habitat' PPFD, so I am going to use that terminology
-
--   growth condition versus measurement conditions
-
--   ?
-
-## Authors
-
-definite: Wei Shen Lim, Dachuan Wang, Chris Muir
-
-question: Sam McKlin, Tom Buckley
+8. What it all means
+  * Our study reveals an important role for plasticity in explaining why amphi leaves are more common in sunny habitats
+  * Confirms that leaf structure of sun leaves incurs costs that amphistomy can mitigate
+  * It will be useful to consider coordination between leaf anatomy, stomatal density, and ratio in future studies of leaf structure-function and in crops (or something)
 
@@
 
-## Figures
-
-1. competing hypotheses and predictions (or predictions after methods?) for the benefit of amphistomy
-2. design of experiment to test competing hypotheses (with actual #'s of replicate, curves, etc.)
-3. Results on acclimatory, plastic, constitutive hypotheses
-4. Additional supporting figure on change in LMA, reduction of Amass, and relationship between LMA and AA? Modeling?
-
-## Abstract
-
-(use for GRC)
-
-Developmental plasticity to light intensity modulates amphistomy advantage
-
-The presence of stomata on both leaf surfaces (amphistomy) can increase photosynthesis in C$_3$ plants by reducing the path length for CO$_2$ diffusion between substomatal cavities and chloroplasts.
-Amphistomatous leaves are most common among herbaceous plants growing in sunny habitats, including many crop species.
-The distribution of amphistomatous leaves in nature may result from an increased photosynthetic benefit of amphistomy under high light intensity, either because of acclimatory, plastic, or constitutive variation in CO$_2$ supply or demand.
-We used a recently developed method to quantify amphistomy advantage, the photosynthetic rate of an amphistomatous leaf relative to an otherwise identical hypostomatous leaf, in 29 diverse populations representing 15 species of wild tomatoes (*Solanum* sect.
-*Lycopersicon* and sect.
-*Lycopersicoides*).
-Plant grown under high light intensity benefit more from amphostomy than those grown under low light, regardless of acclimation or native light intensity.
-Curvature of light response curves indicates that high-light leaves benefit more from amphistomy because of greater resistance to CO$_2$ diffusion within the mesophyll.
-Developmental plasticity in leaf thickness and/or cell arrangement to optimize light interception likely increases under varying light intensity modulates CO$_2$ diffusion within the leaf and, hence, amphistomy advantage as a byproduct.
-Developmental plasticity may therefore play an important and previously unnoticed role in explaining the adaptive significance of amphistomy and the distribution of amphistomatous leaves in nature.
-
 What's surprising: role of developmental plasticity in explaining comparative results; diffusional limitations through palisade ias is strong and needs to be considered with light gradients in understanding leaf strucuture- funciton
 
-## Intro
+big implications - if trees were amphi, more more carbon sink or reduced transpiration
 
-<!-- Paragraph 1 -->
-Stomata are microscopic pores on the surfaces of leaves and other photosynthetic organs formed by a pair of guard cells.
-They are essential for balancing carbon gained per unit water lost and permitted vascular plants to grow tall on land by enabling access to CO$_2$ for photosynthetic carbon assimilation while preventing hydraulic failure in variable environments [@raven_selection_2002; @mcadam_stomata_2021; @clark_origin_2022].
-Optimal stomatal function depends on both dynamic changes in aperture on the scale of minutes to hours, as well as static anatomy determined by developmental plasticity and constitutive genetic differences [@hetherington_role_2003; @de_boer_optimal_2016; @harrison_influence_2020].
-Understanding how stomata respond to environmental change over daily, developmental, and evolutionary time is important for understanding adaptation [@woodward_stomatal_1987; @buckley_modelling_2013; @raven_selection_2002; @haworth_co-ordination_2013; @liang_stomatal_2023; @chua_stomatal_2024; @lang_century-long_2024], predicting paleoclimate from fossil cuticles [@franks_new_2014; @mcelwain_paleoecology_2017; @the_cenozoic_co_proxy_integration_project_cencopip_consortium_toward_2023], and improving crops [@hofmann_impact_2025].
-Stomatal function contributes to global carbon and water cycles [@berry_stomata:_2010] and therefore predicting future climate [@franks_stomatal_2017].
-
-<!-- Paragraph 2 -->
-<!-- Parkhurst gradients 1988 -->
-<!-- Syvertsen et al 1995 - may be more here to check about gi in sun and shade-->
-<!-- "Farquhar and Raschke (2).
-However, using a technique similar to ours, Sharkey et al.
-(11) estimate r, in Xan:hium to be closer to 1.0 m2 s mol-', and state that they believe the estimate of 3.2 m2 s mol-' to be high" -->
-Despite extensive theoretical and empirical progress understanding stomata function and anatomy from molecular to ecosystem levels, the adaptive significance of amphistomatous leaves remains an important unsolved problem in leaf structure-function relationships [@grubb_leaf_1977; @parkhurst_adaptive_1978; @mott_adaptive_1982; @gibson_structure-function_1996; @smith_leaf_1997; @oguchi_leaf_2018; @drake_two_2019; @grubb_leaf_2020].
-Amphistomatous leaves develop abaxial and adaxial stomata whose aperture can be independently regulated [@pospisilova_environmental_1980; @mott_stomatal_1984; @reich_effects_1985; @mott_asymmetric_1993; @wall_stomata_2022] to control gas exchange through each surface.
-All else being equal, gas exchange through stomata on both surfaces increases CO$_2$ supply to chloroplasts by providing a second parallel pathway through leaf intercellular airspaces, enhancing photosynthesis [@parkhurst_adaptive_1978; @gutschick_photosynthesis_1984].
-The extent to which amphistomy increases CO$_2$ supply depends on resistance to diffusion in intercellular airspaces.
-This resistance can be low in thin, porous, amphistomatous leaves [@mott_stomatal_1984; @marquez_assessing_2023], but may be more substantial in thick, dense, hypostomatous leaves [@parkhurst_intercellular_1990].
-Amphistomatous leaves also lose more water through evaporation because of a second boundary layer conductance [@foster_influence_1986], but the additional carbon gain should be enough to offset this cost in most realistic scenarios [@muir_is_2019].
-
-<!-- Paragraph 3 -->
-The paradoxical fact is that, despite the photosynthetic benefit, most leaves are not amphistomatous.
-Many vertically oriented and/or isobilateral leaves are amphistomatous [@wood_physiology_1934; @howell_concerning_1945, @drake_two_2019].
-But among dorsiventral leaves, herbaceous plants in open, high light habitats tend to have have amphistomatous leaves [@salisbury_i_1928; @peat_comparative_1994; @gibson_structure-function_1996; @jordan_using_2014; @bucher_stomatal_2017; @muir_light_2018; @triplett_stomatal_2025].
-Most other leaves, except those from aquatic habitats, are hypostomatous, producing stomata only on the lower, abaxial surface.
-Even resupinate leaves develop stomata on the lower, albeit adaxial surface [@lyshede_comparative_2002], suggesting that leaf orientation (lower vs.
-upper) rather than leaf polarity (abaxial vs.
-adaxial) is causal.
-The covariation between stomatal density ratio and light habitat is both qualitative and quantitative.
-A higher proportion of sun leaves are amphistomatous [@salisbury_i_1928] and the proportion of stomata on the upper, adaxial surface increases with light [@bucher_stomatal_2017; @muir_light_2018].
-Resolving why high light intensity favors amphistomatous dorsivental leaves is an important first step toward understanding variation in stomatal density ratio and leaf structure-function relationships more generally.
-
-<!-- Paragraph 4+, hypotheses and predictions -->
-The overarching hypothesis is that leaves with greater stomatal density ratio are more common in open, sunny habitats because they increase photosynthesis most in those circumstances.
-An amphistomatous leaf increases photosynthetic carbon gain compared to an otherwise identical hypostomatous leaf by increasing conductance through the leaf intercellular airspaces and boudary layers; the additional water loss through a second boundary layer is typically small [@foster_influence_1986].
-We quantify this benefit as the amphistomy advantage (AA = log(Aamphi/Ahypo) [@parkhurst_adaptive_1978; @triplett_amphistomy_2024].
-Why would AA be greater in sun than shade?
-We consider three nonmutually exclusive hypotheses that we classify as 'acclimatory', 'plastic', and 'constitutive'.
-
-Acclimatory hypothesis: Photosynthetic induction to high light intensity typically involves increases in total leaf stomatal conductance (increased CO$_2$ supply), the concentration of active Rubisco, and electron transport capacity (increased CO$_2$ demand).
-A one-dimensional circuit model using the Farquhar-von Caemmerer-Berry biochemical model of C$_3$ photosynthesis [@farquhar_biochemical_1980] shows that both increased stomatal conductance and Rubisco activity should increase AA, all else being equal (Supporting Information).
-If the acclimatory hypothesis is correct, we predict that AAhigh \> AAlow for all species regardless of native habitat or growth environment.
-Plants adapted to sunny, open habitats will evolve greater stomatal density ratio to take advantage of regular exposure to high light intensity.
-
-Plastic hypothesis: Individuals of the same genotype often develop dramatically different leaves in sun and shade conditions [@poorter_metaanalysis_2019].
-Plastic responses are likely adaptations to optimize photosynthesis at different light intensities in variable environments [@givnish_common-garden_2014].
-Plastic changes in leaf anatomy and biochemistry could modulate AA as a byproduct.
-Thicker or less porous leaves, both of which are associated with high leaf mass per area (LMA), will have lower $g_\mathrm{ias}$; leaves with increased total stomatal density and photosynthetic capacity have greater potential CO$_2$ supply and demand.
-Under the plastic hypothesis, we predict that AAsun \> AAshade for all species and light intensities.
-Secondarily, AAsun and gsmax,sun should also be positively associated with native light habitat if genotypes adapted to sunny, open habitats if they can express a phenotype best adapted to that environment when leaves develop under high light intensity.
-Genotypes adapted to shaded, closed habitats may be plastic, but limits on the width of their reaction norms prevent them from developing traits optimal for conditions they do not regularly experience in nature.
-
-Constitutive hypothesis: In environments that are relatively constant or where environmental change cannot be anticipated by a reliable cue, natural selection will favor constitutive expression of optimal phenotypes.
-We therefore predict genotypes from more sunny, open habitats will have consistently greater AA under also measurement and growth light intensities.
-For herbaceous plants, light intensity is largely a function of the tree canopy [@engelbrecht_evaluation_2001].
-Herbs growing in the open will regularly experience high light intensity; herbs growing under a forest canopy will often experience low light intensity.
-
-The primary directional predictions for each hypothesis are summarized in Table X; detailed predictions for results that would indicate support for multiple hypotheses are in Table SX.
-
-CONCEPTUAL FIGURE HERE
-
-caption: Directional predictions associated with each hypothesis explaining why amphistomy advantage (AA) might be greater for leaves in sunny, open habitats.
-For each hypothesis, we make predictions for how native plant area index (PAI), growth light treatment, and measurement light intensity would affect AA.
-
-|  hypothesis  | measurement light intensity | growth light intensity  | native PAI |
-|:------------:|:---------------------------:|:------------------------:|:----------:|
-| acclimatory  | AA2000 > AA150             | AAsun = AAshade          | cor(PAI, AA) = 0 |
-| plastic      | AA2000 = AA150             | AAsun > AAshade          | cor(PAI, AA) = 0 |
-| constitutive | AA2000 = AA150             | AAsun = AAshade          | cor(PAI, AA) > 0 |
-
-<!-- Paragraph 5 (summarize methods) -->
-We tested these hypotheses by comparing AA among amphistomatous wild tomato species [@peralta_taxonomy_2008] from different native light habitats, grown under simulated sun and shade light treatments, and measured under contrasting light intensity (Figure of hypotheses and predictions).
-We measured AA on `r n_plant` individual plants from `r n_acc` accessions (average of `r mean_n_rep` replicates per light treatment) using a recently developed method [@triplett_amphistomy_2024].
-With this method, we directly compare the photosynthetic rate of an untreated amphistomatous leaf to that of the same leaf with gas exchange blocked through the adaxial (upper) surface by transparent plastic, which we refer to as 'pseudohypostomy'.
-To compare amphi- and pseudohypostomatous leaves at identical whole-leaf $g_\text{sc}$, we measure $A$ over a range of $g_\text{sc}$, inducing stomatal opening and closure by modulating humidity (see Materials and Methods for further details).
-We estimated 'amphistomy advantage' (\aax) *sensu* @parkhurst_adaptive_1978, but with modifications previously described in @triplett_amphistomy_2024 and here (Materials and Methods).
-The native light intensity was represented by plant area index (PAI $\unit{\meter\squared\per\meter\squared}$), estimated using a global gridded X-m2 resolution data set derived from the Global Ecosystem Dynamics Investigation [GEDI; @burns_multi-resolution_2024] and georeferenced accession collection information from the Tomato Genetics Resource Center.
-The growth light intensities were `r ppfd_sun` (sun treatment) and `r ppfd_shade` (shade treatment) while all other environment conditions were nearly identical (see @sec-methods for supporting detail).
-The high and low measurements intensities were \ppfdequals{2000} (`r p_red_high`:`r p_blue_high` red:blue) and \ppfdequals{150} (`r p_red_low`:`r p_blue_low` red:blue), respectively.
-
-<!-- Paragraph 6 (summarize main results) -->
-Consistent with biophysical theory of CO$_2$ diffusion within leaves, AA > 0 for all accessions (Figure X).
-AA varied substantially between measurement light intensities, growth light intensities, and among accessions (point to model comparison evidence).
-Measured under high light intensity, AA was consistently greater for sun plants.
-The average AA among accessions in the shade treatment was `r filter(aa_mean_range, light_treatment == "low", light_intensity == "2000")$x`; however, the same accessions grown at high light intensity showed a mean AA of `r filter(aa_mean_range, light_treatment == "high", light_intensity == "2000")$x`.
-Contrary to the predictions of the assimilatory hypothesis, AA was greater in all accessions under low measurement light intensity for both sun and shade grown plants.
-The effect of low light on AA was more pronounced in the sun-grown plants, where AA was significantly greater under low measurement light intensity in `r aa_summary3["high"]` of `r n_acc` accessions compared to to `r aa_summary3["low"]` accessions for shade-grown plants.
-The overall average AA of shade and sun grown plants measured under low light intensity was `r filter(aa_mean_range, light_treatment == "low", light_intensity == "150")$x` and `r filter(aa_mean_range, light_treatment == "high", light_intensity == "150")$x`, respectively.
-There was a slight tendency for accessions from more closed habitats (greater PAI), but AA varied widely among accessions from open habitats (low PAI) regardless of growth and measurement light intensities (Figure X).
-
-<!-- Paragraph 7 (summarize secondary results) -->
-The pattern of AA across wild tomatoes strongly supports the plasticity hypothesis, argues against the acclimatory hypothesis, and provides only weak support for the constitutive hypothesis.
-Plastic changes in leaf thickness and/or packing density, summarized by the bulk leaf mass per area (LMA), likely mediate the effect of growth light intensity on AA.
-LMA increased in sun grown plants in all accessions by an average of X% [CI], similar to plastic responses in many species [@poorter_metaanalysis_2019].
-When LMA is included in the model, the effect of growth light intensity on AA is no longer significant based on model comparison (Table ?
-- need to check out what the interaction with measurement light intensity means) and LMA is weakly positively correlated with AA (Figure X).
-In contrast, LMA did not mediate indirect effects of native PAI on AA (Table ?).
-Many anatomical traits underlie LMA [@john_anatomical_2017] and future research will be needed identify which particular traits, such as leaf thickness or mesophyll porosity, are responsible for mediating AA.
-Our study is also limited by the fact that we could not directly measure the stomatal conductance ratio and our method for estimating AA requires large sample sizes to overcome several sources of measurement error (see supporting information).
-Future experiments measuring AA with a dual sided chamber [@wall; @marquez] can overcome these limitations.
-
-Our results indicate that developmental plasticity, as opposed to acclimation or adaptation to open habitats, may explain the long-standing observation that amphistomatous leaves are more common in sunny habitats.
-If correct, this implies that high light intensity *per se* does not increase the benefit of amphistomy, but rather than anatomical and/or biochemical changes associated with higher light intensity modulate AA.
-We cannot yet attribute specific changes wrought by plasticity, but the correlation with laminar LMA suggests that changes in leaf thickness and/or density mediate the relationship between plasticity and AA.
-While we found little evidence for native light habitat in driving AA, unexplained constitutuve differences among accessions may be byproducts of adaptive evolutionary divergence.
-Hence, we cannot yet rule out an important role for environmental adaptation in modulating AA and this will require future research to resolve.
-
-#### MARKER
-Amphistomy likely plays a significant, but under appreciated, role in the carbon and water balance of plant groups where this trait is common, such as herbaceous plants in sunny habitats and woody species in arid biomes (cites).
-If our estimates of AA from the wild tomato clades
-big implications 
-- amphistomy plays role in carbon/water balance. do equivalent gs
-- if trees were amphi, more more carbon sink or reduced transpiration
-- comparative studies should consider plasticity
-
-## Outtro - I think this goes at the end??
 It is commonly assumed in comparative studies that most trait variation between species is constitutive and genetically determined, but our results suggest that plastic responses to light are necessary to explain why \gmaxratio{} increases with light habitat.
-Reaction norms to light intensity among wild tomatoes indicate that accessions from low light habitats express low \gmaxratio{} when grown in shade, whereas accessions from high light habitats have constitutively higher \gmaxratio{} (Fig.
+Reaction norms to light intensity among wild tomatoes indicate that populations from low light habitats express low \gmaxratio{} when grown in shade, whereas populations from high light habitats have constitutively higher \gmaxratio{} (Fig.
 rxn-norm).
-The larger response to light intensity is driven by greater plasticty in leaf structure in accessions from low-light habitats (Fig/table - I haven't tested this yet, so this is a placeholder.) We infer that these structural changes in sun leaves are costly because they increase the diffusion path length for CO$_2$ and reduce photosynthesis.
+The larger response to light intensity is driven by greater plasticty in leaf structure in populations from low-light habitats (Fig/table - I haven't tested this yet, so this is a placeholder.) We infer that these structural changes in sun leaves are costly because they increase the diffusion path length for CO$_2$ and reduce photosynthesis.
 As a consequence, sun leaves benefit more from amphistomy.
 This is the first demonstration that sun leaves benefit more from amphistomy because it compensates for developmental plasticity in leaf structure.
-
-### MANUSCRAPS
-Problems to bring up in discussion:
-- all our species were amphi, so they may all live in PAI below threshold to favor amphi
-- other microclimate diffs not considered
-- don't know specific anatomical change
-
-Two outstanding areas for further research are 1) the specific anatomical changes, such as leaf thickness or mesophyll porosity, that increase the diffusion path length for CO$_2$ in sun leaves, and 2) the costs of amphistomy that could explain plasticity in \gmaxratio{} in accessions from low-light habitats.
-Resolving the first question will require detailed anatomical measurements on a large number of shade and sun leaves with concomitant estimates of \aax.
-The fact that accessions from low-light habitats are more plastic suggests that costs of amphistomy are greatest in these environments.
-(not finished yet)
-
-Not sure where this goes: We infer this from the fact that blocking gas exchange in pseudohypostomatous leaves reduced $A$ by X-X% depending on the accession, light treatment, and light intensity (Table/figure).
-The AA is equivalent to an X-X% change in total $g_\text{sc}$ (see SI section gs equivalency).
-But whereas increasing $g_\text{sc}$ would increase water loss as a necessary by-product, amphistomy can increase $A$ without any appreciable affect on transpiration.
-
-from Mott and Oleary: "However, Jones and Slatyer (5) reported a higher mesophyll resistance for CO, entering through the upper stomata than for the lower, and the data of Vaclavik (12) appear to support this conclusion"
-
-8. What it all means
-  * Our study reveals an important role for plasticity in explaining why amphi leaves are more common in sunny habitats
-  * Confirms that leaf structure of sun leaves incurs costs that amphistomy can mitigate
-  * It will be useful to consider coordination between leaf anatomy, stomatal density, and ratio in future studies of leaf structure-function and in crops (or something)
-
-## Materials and Methods [#sec-methods]
-
-\[this will be moved to SI eventually. note that much of it also the same in solanum-aq project, so might change to cite that paper?\]
-
-(not sure where this goes)  * We acclimated the focal leaf to high light (\ppfdequals{2000}) and high relative humidity (\rhequals{70}) until $A$ and \gsw{} reach their maximum.
-After that, we decreased \rh{} to $\approx 10\%$ to induce rapid stomatal closure without biochemical down
-
-### Accessions
-
-We compared AA among `r n_acc` ecologically diverse accessions of wild tomato, including representatives of all described species of *Solanum* sect.
-*Lycopersicon* and sect.
-*Lycopersicoides* [@peralta_taxonomy_2008] and the cultivated tomato *S.
-lycopersicum* var.
-*lycopersicum* \autoref{tab:accessions}.
-Due to constraints on growth space and time, we spread out measurements over `r n_weeks` weeks from `r start_licor_date` to `r end_licor_date`.
-Replicates within accession were evenly spread out over this period to prevent confounding of temporal variation in growth conditions with accession.
-[anything else to say here?
-maybe explain accession selection and phylogeny?]
-
-```{r, accessions, eval=TRUE,echo=FALSE,message=FALSE,warning=FALSE}
-#| label: tbl-accessions
-#| tbl-cap: "Accessions of *Solanum* used in this study.
-The species name, accession number, collection latitude, longitude, elevation, and daily solar radiation estimated using the SPLASH algorithm (see 'Climate data').
-TGRC: Tomato Genetics Resource Center; PPFD: Photosynthetic Photon Flux Density."
-
-# Copied from solanum-aq project
-
-accession_info = read_rds("../data/accession-info.rds") |>
-  filter(accession %in% rh_curves$acc) |>
-  left_join(read_rds("../data/accession-gedi.rds") |>
-              select(accession, pai), by = join_by(accession)) |>
-  arrange(species, accession) |>
-  select(
-    Species = species,
-    `TGRC accession` = accession,
-    `Latitude` = lat,
-    `Longitude` = lon,
-    `Elevation (mas)` = elev,
-    `$\\mathrm{PAI}~(\\unit{\\meter\\squared\\per\\meter\\squared})$` = pai
-  )
-
-accession_info |>
-  knitr::kable(
-    format = "latex",
-    digits = c(rep(0, 2), 3, 3, 0, 2),
-    booktabs = TRUE,
-    longtable = TRUE,
-    escape = FALSE
-  ) |>
-  kableExtra::kable_styling(latex_options = c("striped", "no_vertical_lines")) |>
-  kableExtra::column_spec(1, width = "3.25cm", italic = TRUE)
-
-```
-
-### Plant growth conditions
-
-In all growth spaces, we recorded $\mathrm{PPFD}$ using full spectrum quantum sensors (SQ-500-SS, Apogee Instruments, Logan, Utah, USA); we recorded temperature, RH, and [CO$_2$] using an EE894 sensor (E+E Elektronik, Engerwitzdorf, Austria) protected by a radiation shield.
-All environmental measurements were taken every 10 minutes from the middle of plants racks at approximately the same height as the leaves we measured.
-We measured leaf temperature of focal leaves prior to measurement using an infrared radiometer (SI-111-SS, Apogee Instruments, Logan, Utah, USA).
-
-#### Germination and seedling stage
-
-Seeds provided by the Tomato Genetics Resource Center germinated on moist paper in plastic boxes after soaking for 30-60 minutes in a 50% (volume per volume) solution of household bleach and water, followed by a thorough rinse.
-We transferred seedlings to cell-pack flats containing Pro-Mix BX potting mix (Premier Tech, Rivière-du-Loup, Quebec, Canada) once cotyledons fully emerged, typically within 1-2 weeks of sowing.
-We grew seeds and seedlings for both sun and shade treatments under the same environmental conditions (12:12 h, `r germ_temp_day`:$\qty{`r germ_temp_night`}{\degreeCelsius}$, `r germ_rh_day`:`r germ_rh_night` RH day:night cycle).
-LED light provided $\mathrm{PPFD} = \qty{`r germ_ppfd`}{\micro\mol\raiseto{-2}\meter\raiseto{-1}\second}$ (Fluence RAZRx, Austin, Texas, USA).
-
-#### Light treatments
-
-Seedlings were randomly assigned in alternating order within accession to the sun or shade treatment during transplanting.
-After seedlings established in cell-pack flats for $\approx 2$ weeks, we transplanted them to 3.78 L plastic pots containing 60% Pro-Mix BX potting mix, 20% coral sand (Pro-Pak, Honolulu, Hawaiʻi, USA), and 20% cinders (Niu Nursery, Honolulu, Hawaiʻi, USA).
-Percentage composition is on a volume basis.
-The soil mixture contained slow release NPK fertilizer following manufacturer instructions (Osmocote Smart-Release Plant Food Flower & Vegetable, The Scotts Company, Marysville, Ohio, USA).
-We determined pot field capacity one week after transplanting using a scale (Ohaus V12P15 Valor 1000, Parsippany, New Jersey, USA) and watered to field capacity three times per week to prevent drought stress.
-
-We assigned sun and shade treatment to lower and upper racks of a $\qty{1.22}{\meter} \times \qty{2.44}{\meter}$ shelving unit in a climate-controlled growth room.
-We assigned the sun treatment to the lower rack to limit diffuse light from reaching the shade treatment.
-The average daytime \ppfd{} was \ppfdqty{`r ppfd_sun`} and \ppfdqty{`r ppfd_shade`} for sun and shade treatments, respectively.
-To isolate the effect of light intensity from quality, we used the same LED model with the the same spectrum (Fluence SPYDR 2i, Austin, Texas, USAS), but dimmed the lights in the shade treatment.
-To maintain homogeneous environmental conditions other than light, we mixed air within the growth room using an air circulator (Vornado 693DC, Andover, Kansas, USA) and within racks using a miniature oscillating air circulator (Vornado Atom 1, Andover, Kansas, USA).
-Despite these efforts, the air in the sun treatment was on average $\qty{`r temp_diff`}{\degreeCelsius}$ warmer and the average RH was consequently `r rh_diff` lower.
-However, because of evaporative cooling, the leaves in the sun treatment were only $\qty{`r leaf_temp_diff`}{\degreeCelsius}$ on average ($n = `r n_leaf_temp`$ leaves).
-
-### Leaf trait measurements
-
-We selected a fully expanded, unshaded leaf at least six leaves above the cotyledons during early vegetative growth.
-This typically meant that plants had grown in light treatments for $\approx$ 4 weeks, ensuring they had time to sense and respond developmentally to the light intensity of the treatment rather than the seedling conditions [@schoch_dependence_1980].
-Shade plants grew slower than sun plants, hence leaves at the same developmental stage were measured on chronologically older plants in the shade treatment.
-In some sun plants, we had to use leaves higher on the stem because short internodes made lower leaves inaccessible with the gas exchange equipment.
-We measured terminal leaflets in `r percent_terminal`\% of cases, but used the lateral leaflet closest to the terminal leaflet when it was damaged or difficult to clamp into the gas exchange chamber.
-When a leaflet was damaged during gas exchange measurements, we collected anatomical data from the nearest leaflet on the same leaf (`r percent_change_leaflet`)\%.
-We measured chlorophyll concentration index (CCI) using a chrolophyll concentration meter (MC-100, Apogee Instruments, Logan, Utah, USA) on the lamina of focal leaflets before gas exchange measurements at the same time we measured leaf temperature.
-
-#### Amphistomy advantage
-
-We estimated 'amphistomy advantage' (\aax) *sensu* @parkhurst_adaptive_1978, but with modifications previously described in @triplett_amphistomy_2024.
-\aax{} is calculated as the log-response ratio of $A$ compared at the same total \gsw:
-
-$$\mathrm{AA} = \mathrm{log}(A_{\mathrm{amphi}} / A_{\mathrm{hypo}})$$
-
-We measured the photosynthetic rate of an untreated amphistomatous leaf (\Aamphi) over a range of \gsw{} values.
-We refer to this as an \agcurve{} curve.
-We compared the \agcurve{} curve of the untreated leaf to the photosynthetic rate of pseudohypostomatous leaf (\Ahypo), which is the same leaf but with gas exchange through the upper surface blocked by a neutral density plastic (propafilm).
-
-We measured \agcurve{} curves using a portable infrared gas analyzer (LI-6800PF, LI-COR Biosciences, Lincoln, Nebraska, USA).
-Light-acclimated plants were placed under LEDs dimmed to match their light treatment during gas exchange measurements.
-We estimated the photosynthetic rate ($A$) and stomatal conductance to CO$_2$ (\gsw) at ambient CO$_2$ (\caequals{415}) and \tleafequals{25.0}.
-The irradiance of the light source in the pseudohypo leaf was higher because the propafilm reduces transmission.
-To compensate for reduced transmission, we increased incident \ppfd for pseudohypo leaves by a factor 1/0.91, the inverse of the measured transmissivity of the propafilm.
-We also set the stomatal conductance ratio, for purposes of calculating boundary layer conductance, to 0 for pseudohypo leaves following manufacturer directions.
-
-We collected four \agcurve{} curves per leaf, an amphi (untreated) curve and a pseudohypo (treated) curve at high light-intensity (\ppfdequals{2000}; `r p_red_high`:`r p_blue_high` red:blue) and low light-intensity (\ppfdequals{150}; `r p_red_low`:`r p_blue_low` red:blue).
-We always measured high light-intensity curves first because photosynthetic downregulation is faster than upregulation in these species.
-To control for order effects, we alternated between starting with amphi or pseudohypo leaf measurements.
-Unlike @triplett_amphistomy_2024, preliminary experiments with *Solanum* indicated a strong order effect in that $A$ declined in the second curve.
-Therefore, we made measurements over two days.
-On the first day, we measured high and low light-intensity curves for either amphi or pseudohypo leaves; on the second day, we measured high and low light-intensity curves on the other leaf type.
-
-In all cases, we acclimated the focal leaf to high light (\ppfdequals{2000}) and high relative humidity (\rhequals{70}) until $A$ and \gsw{} reach their maximum.
-After that, we decreased \rh{} to $\approx 10\%$ to induce rapid stomatal closure without biochemical downregulation.
-Hence, \Aamphi{} and \Ahypo{} were both measured at low chamber humidity after the leaf had acclimated to high humidity.
-All other environmental conditions in the leaf chamber remained the same.
-We logged data until \gsw{} reached its nadir.
-We then acclimated the leaf to low light (\ppfdequals{150}) and \rhequals{70} before inducing stomatal closure with low \rh and logging data as described above.
-
-#### Light-response (\aqcurve) curves
-
-In `r percent_aa_and_aq`\% of plants, we measured light-response (\aqcurve) curves on the same leaflets as \agcurve{} curves.
-However, when a leaflet was damaged during \agcurve{} curves, we used the next closest leaflet for \aqcurve{} curves.
-Leaves acclimated to high light-intensity (\ppfdequals{2000}), ambient CO$_2$ (\caequals{415}), \rhequals{50}, and \tleafequals{25}.
-After $A$ and \gsw{} stabilized, we measured $A$ at 20 light-intensity levels between $0$ and $\qty{2000}{\micro\mol\raiseto{-2}\meter\raiseto{-1}\second}$ in descending order.
-
-#### Stomatal anatomy
-
-<!-- are there any cases where we did not get stomatal anatomy data? -->
-
-We estimated the stomatal density and size on ab- and adaxial leaf surfaces from all leaves, using guard cell length as a proxy for stomatal size since it proportional to maximum conductance [@sack_developmental_2016].
-We made surface impressions of leaf lamina from the same area used for gas exchange measurements using a-silicone impression material (Zhermack elite HD+, light body, fast set, Rovigo, Italy).
-We applied clear nail polish to make positive replicas of the impression.
-After nail polish dried, we mounted replicas on a microscope slide using transparent tape [@mott_amphistomy_1991].
-We digitized a portion of each leaf surface replica using a brightfield microscope (Leica DM2000, Wetzlar, Germany).
-We counted and measured guard cell length on all stomata using the FIJI implementation of ImageJ2 version 2.3.0 [@schindelin_fiji_2012], then divided the count by the visible leaf area ($\qty{0.890}{\raiseto{2}\milli\meter}$) to estimate stomatal density.
-
-<!-- 
-#### Internal leaf anatomy
-
-Add information on internal leaf anatomy, if using 
--->
-
-#### Leaf mass per area
-
-Leaf mass per area (LMA) is the dry mass divided by the leaflet area.
-We scanned fresh leaflets on a flat bed scanner (Epson V600, Los Alamitos, California.
-USA) and measured leaflet area from digital images using the FIJI implementation of ImageJ2 version 2.3.0 [@schindelin_fiji_2012].
-We dried leaves for 72 hours at $\qty{74}{\degreeCelsius}$ in a food dehydrator (Cosori CP267-FD, Vesync Co., Anaheim, California, USA) and weighed using a benchtop analytical balance (Ohaus PR64 Analytical Balance, Parsippany, New Jersey, USA).
-In $`r percent_lma_switch`\%$ we measured LMA on the adjacent leaflet because the focal leaflet was damaged or wilted while making surface impressions and we could not reliably estimate area.
-LMA data are missing from $`r percent_lma_missing`\%$ of individuals because the area or mass was not recorded at all or recorded incorrectly.
-
-### Cleaning \agcurve{} curves
-
-The raw data set consisted of `r prettyNum(aa_stats$n_rh_curve0, big.mark = ",")` \agcurve{} curves with an average of `r formatC(aa_stats$n_point_per_rh_curve0, digits = 3)` points per curve.
-Manual curation of a data set this size in a principled, consistent manner is not feasible.
-Therefore, we automated data cleaning using custom *R* scripts.
-Cleaning is divided into six sequential steps (\autoref{tbl-cleaning1}).
-
-```{r, cleaning1, echo=FALSE,message=FALSE,warning=FALSE}
-#| label: tbl-cleaning1
-#| tbl-cap: "Six sequential steps for cleaning \\agcurve{} curves.
-The rationale and procedure for each step are described in the text.
-The rightmost columns summarize the number of curves and mean number of points per curve remaining after each step.
-For reference, there are four possible \\agcurve{} curves per replicate: all combinations of leaf type (amphi or pseudohypo) and light intensity (high or low)."
-
-step_description_col = c(
-  "1. remove unreliable and unusable data points",
-  "2. remove hysteretic portion of \\agcurve{} curves at low \\gsw{}",
-  "3. remove outliers within each \\agcurve{} curve",
-  "4. remove replicates with no overlap between amphi and pseudohypo \\agcurve{} curves",
-  "5. thin redundant data points within each \\agcurve{} curve",
-  "6. trim extreme \\aax{} values"
-)
-
-n_curve_col = unlist(aa_stats[str_detect(names(aa_stats), "^n_rh_curve[1-9]$")]) |>
-  prettyNum(big.mark = ",")
-
-n_point_col = unlist(aa_stats[str_detect(names(aa_stats), "^n_point_per_rh_curve[1-9]$")])
-
-tibble(
-  `Step: description` = step_description_col,
-  `Number of curves` = n_curve_col,
-  `Number of points per curve` = n_point_col
-) |>
-  knitr::kable(format = "latex",
-               align = c("l", "c", "c"),
-               digits = c(0, 0, 1),
-               booktabs = TRUE, escape = FALSE) |>
-  kableExtra::kable_styling(latex_options = c("striped", "no_vertical_lines")) |>
-  kableExtra::column_spec(1, width = "9cm") |>
-  kableExtra::column_spec(2, width = "2.5cm") |>
-  kableExtra::column_spec(3, width = "2.5cm")
-
-```
-
-#### Remove unreliable and unusable data points
-
-*Rationale*: Unreliable data points consisted of those where chamber [CO$_2$] was unstable and therefore measurements are not biologically meaningful.
-Unusable data points were those where $A < 0$ because the logarithm of a negative number is undefined.
-
-*Procedure*: We retained data points where \cabetween{`r ca_min`}{`r ca_max`} and $A> 0$.
-
-#### Remove hysteretic portion of \agcurve{} curves at low \gsw{}
-
-*Rationale*: In most \agcurve{} curves, we observed a hysteretic response at low \gsw.
-After \gsw{} and $A$ declined simultaneously, $A$ increased slightly as \gsw{} continued to decline or stabilize, indicating some leaf acclimation to low \rh.
-We removed this portion of the curve to focus curve-fitting on the primary domain where $A$ increases monotonically with \gsw{}.
-
-*Procedure*: For each curve, we removed data points after \gsw{} had reached its minimum unless there were fewer than `r n_min_points_per_rh_curve` data points remaining.
-
-#### Remove outliers within each \agcurve{} curve
-
-*Rationale*: Individual outliers within \agcurve{} curves, usually caused by transitory changes in chamber conditions, exert undue leverage on parameter estimates and cause bias and/or low precision in parameter estimates.
-
-*Procedure*: We fit provisional quadratic regressions for each curve using ordinary least squares with the `lm()` function in *R*.
-We sequentially removed data points with an absolute external studentized residual $> `r point_outlier_threshold`$ until none remained.
-
-#### Thin redundant data points within each \agcurve{} curve
-
-*Rationale*: Data points closely spaced along the \agcurve{} curve provide redundant information and may be highly correlated (i.e.
-pseudoreplication).
-This occurred because data was logged at a constant temporal interval, but the rate at which \gsw{} declined was not constant.
-Thinning reduces parameter estimation bias toward densely sampled regions of the curve which may not be the most biologically informative.
-
-*Procedure*: We retained the maxima and minima \gsw{} for each curve and thinned all but one point per thinning interval of $`r thinning_interval`~\log(\si{\mol\raiseto{-2}\meter\raiseto{-1}\second})$, retaining the point nearest the midpoint of the interval.
-
-#### Remove replicates with no overlap between amphi and pseudohypo \agcurve{} curves
-
-*Rationale*: We could not estimate \aax{} for replicates where amphi and pseudohypo \agcurve{} curves did not overlap.
-
-*Procedure*: We removed replicates where the range of \gsw{} values for amphi and pseudohypo \agcurve{} curves did not overlap.
-
-#### Trim extreme \aax{} values
-
-*Rationale*: Extreme \aax{} values were likely due to measurement error or leaf damage.
-Since amphi and pseudohypo \agcurve{} curves are measured on consecutive days, a poor calibration or a damaged leaf could cause a large difference in $A$ between days, which would appear as an extreme \aax{} value.
-
-*Procedure*: We provionsally estimated \aax{} for each replicate by integrating over the range of \gsw{} values where amphi and pseudohypo \agcurve{} curves overlap.
-In this procedure, curve parameters were provisionally estimated using ordinary least squares with the `lm()` function in *R*.
-We then used point estimates of \aax{} for each replicate as the response variable in a linear model with light treatment, light intensity, accession, and all interactions as explanatory variables.
-This model was also fit using ordinary least squares with the `lm()` function in *R*.
-We classified  extreme \aax{} values as those with an absolute internal studentized residual $> `r aa_outlier_threshold`$.
-Because these values likely indicate significant measurement error or leaf damage, we removed \agcurve{} curves at both light intensities if either was classified as extreme.
 
@@
 We therefore removed \aqcurve{} curves with $r^2 < `r aq_r2_threshold`$.
-
-### Bayesian data analysis in *Stan*
-
-We fit `r n_model` models to test predictions of competing hypotheses about why amphistomy advantage (\aax) might be greater for leaves in sunny, open habitats.
-This section provides an overview of differences among models (\autoref{tbl-models}).
-The next sections describe how we fit models in *Stan*, all model parameters and priors, and specific predictions about parameter values for each hypothesis.
-
-```{r, models, echo=FALSE,message=FALSE,warning=FALSE}
-#| label: tbl-models
-#| tbl-cap: "Summary of differences among competing models of how \\aax{} varies with light intensity, light treatment, and among accessions as a function of native \\ppfd.
-The models are numbered from simpler to more complex.
-All models include fixed effects of light intensity and light treatment; some models include interactions between.
-All models include a phylogenetic random effect of accession on \\aax; some models include varying effects of light intensity and light treatment among accessions.
-The last column indicates the accession-level \\aax{} variable we used as a response to native \\ppfd."
-
-tribble(
-    ~"Model", ~"Fixed effects", ~"Phylogenetic random effects", ~"Response to native \\ppfd",
-    "1", "light intensity", "varying intercept among accessions", "$\\mathrm{AA}_{0, \\text{acc}}$",
-    "1", "light treatment", "", "",
-    "2", "light intensity", "varying intercept among accessions", "$\\mathrm{AA}_{0, \\text{acc}}$",
-    "2", "light treatment", "", "",
-    "2", "intensity $\\times$ treatment", "", "",
-    "3", "light intensity", "varying intercept among accessions", "$\\mathrm{AA}_{2000, \\text{acc}}$",
-    "3", "light treatment", "varying effect of high light intensity among accessions", "",
-    "4", "light intensity", "varying intercept among accessions", "$\\mathrm{AA}_{\\text{sun, acc}}$",
-    "4", "light treatment", "varying effect of sun treatment among accessions", "",
-    "5", "light intensity", "varying intercept among accessions", "$\\mathrm{AA}_{2000, \\text{sun, acc}}$",
-    "5", "light treatment", "varying effect of high light intensity among accessions", "",
-    "5", "intensity $\\times$ treatment", "varying effect of sun treatment among accessions", ""
-  ) |>
-  knitr::kable(
-    format = "latex",
-    align = c("c", "l", "l", "l"),
-    booktabs = TRUE,
-    escape = FALSE,
-    longtable = FALSE
-  ) |>
-    kableExtra::collapse_rows(columns = 1, latex_hline = "major", row_group_label_position = "first") |>
-  kableExtra::column_spec(1, width = "0.3in") |>
-  kableExtra::column_spec(2, width = "1.3in") |>
-  kableExtra::column_spec(3, width = "3.4in") |>
-  kableExtra::column_spec(4, width = "1in")
-
-```
-
-#### Fitting models in *Stan* {#sec-fitting}
-
-We fit Bayesian models with MCMC sampling in the probabilistic programming language *Stan* [@stan_development_team_stan_2025] using the *R* package **brms** version `r fit_aa$version$brms` [@burkner_brms_2017].
-We used CmdStan version `r fit_aa$version$cmdstan` and **cmdstanr** version `r fit_aa$version$cmdstanr` [@gabry_cmdstanr_2025] to interface with *R* version `r getRversion()` [@r_core_team_r:_2025].
-We sampled the posterior distribution from `r dim_fit_aa[2]` chains with `r dim_fit_aa[1]` iterations each after `r dim_fit_aa[1]` warmup iterations per chain.
-We estimated parameters and confidence intervals as the median and 95% quantile intervals of the posterior, respectively.
-We chose the number of chains, warmup and sampling iterations, and maximum treedepth so that parameter estimates converged ($\hat{R} < `r aa_args['max_rhat']`$ [@gelman_inference_1992]) and the effective sample size (ESS) for each parameter was $> `r aa_args['min_ess']`$.
-
-#### Parameter estimation and priors {#sec-parameters}
-
-There were four levels of parameter estimation in our analysis:
-
-1. Estimate \agcurve{} curve parameters
-2. Estimate \aax{} for each light intensity with leaf using \agcurve{} curve parameters
-3. Estimate the effects of light intensity, light treatment, and accession on \aax{} (assimilatory and plasticity hypotheses)
-4. Estimate the effects of native light habitat on accession-level \aax{} (constitutive hypothesis)
-
-Although the higher-level parameter estimates depend on the lower-level parameter estimates, we fit all models simultaneously to ensure that the uncertainty in lower-level estimates propagated to higher-level estimates.
-
-```{r, parameters, echo=FALSE,message=FALSE,warning=FALSE}
-#| label: tbl-parameters
-#| tbl-cap: "Description of parameters estimated in the hierarchical Bayesian model.
-The `Parameter` column lists the parameter name as it appears in text.
-The `Description` column provides a brief description of the parameter."
-
-df_parameters = read_csv("../data/parameters1.csv", col_types = "ccccl") |>
-    dplyr::filter(include)
-
-df_parameters |>
-  dplyr::select(Parameter = latex, Description = description) |>
-  knitr::kable(
-    format = "latex",
-    align = c("l", "l"),
-    booktabs = TRUE,
-    escape = FALSE,
-    longtable = TRUE
-  ) |>
-  kableExtra::kable_styling(latex_options = c("striped", "no_vertical_lines")) |>
-  kableExtra::column_spec(1, width = "1in") |>
-  kableExtra::column_spec(2, width = "5in") |>
-  kableExtra::pack_rows("\\agcurve{} curve parameters",
-                        min(which(df_parameters$par_type == "ags")),
-                        max(which(df_parameters$par_type == "ags")),
-                        escape = FALSE) |>
-  kableExtra::pack_rows("\\aax{} for each light intensity with leaf using \\agcurve{} curve parameters",
-                        min(which(df_parameters$par_type == "aa_est")),
-                        max(which(df_parameters$par_type == "aa_est")),
-                        escape = FALSE) |>
-  kableExtra::pack_rows("effects of light intensity, light treatments, and accession on \\aax",
-                        min(which(df_parameters$par_type == "aa")),
-                        max(which(df_parameters$par_type == "aa")),
-                        escape = FALSE) |>
-  kableExtra::pack_rows("effects of native light habitat on accession-level \\aax",
-                        min(which(df_parameters$par_type == "ppfd_aa")),
-                        max(which(df_parameters$par_type == "ppfd_aa")),
-                        escape = FALSE)
-
-```
-
-##### \agcurve{} curve parameters
-
-We modeled \logA as a quadratic function of \loggsw for each leaf using the following equation:
-
-$$\log(A_{ij}) = (\beta_0 + b_{0,j}) + (\beta_1 + b_{1,j}) \log{g_{\text{sw},i}} + (\beta_2 + b_{2,j}) \log{g_{\text{sw},i}}^2 + \epsilon_{i}$$
-
-where $\beta_0$, $\beta_1$, and $\beta_2$ are the average intercept, linear, and quadratic coefficients, respectively.
-We used diffuse normal priors with mean $0$ and standard deviation $10$ on these parameters.
-We estimated random effects of curve $j$ on the intercept ($b_{0,j}$), linear ($b_{1,j}$), and quadratic ($b_{2,j}$) coefficients.
-We assumed that the $j \times 3$ array of coefficients was multivariate normal a mean vector of $\vec{0}$ and covariance $\Sigma_\text{curve}$.
-We used a weakly informative normal prior with mean $0$ and standard deviation $1$ on the log-transformed standard deviations (i.e.
-the diagonal of $\Sigma_\text{curve}$).
-We used a weakly informative $\mathrm{LJK}(2)$ prior on the correlation matrix.
-The off diagonal elements of $\Sigma_\text{curve}$ can be calculated from its diagonal elements and the correlation matrix.
-
-The residuals $\epsilon_{i}$ were modeled as a lag-1 autocorrelated time-series.
-We further assumed that the residual standard deviation of the $j^{\text{th}}$ curve ($\sigma_{\epsilon,j}$) was inversely proportional to the leaf surface area ($S_j$) within the chamber:
-
-$$\log (\sigma_{\epsilon,j}) = \log (\sigma_{\qty{6}{\raiseto{2}\centi\meter},\epsilon}) + (6 - S_j) \beta_{S,\epsilon}$$
-where $\sigma_{\qty{6}{\raiseto{2}\centi\meter},\epsilon}$ is the minimum residual standard deviation when the $\qty{6}{\raiseto{2}\centi\meter}$ chamber is completely filled.
-The residual standard deviation increases on log-linear scale by $\beta_{S,\epsilon}$.
-We used a weakly informative normal prior with mean $-3$ and standard deviation $5$ on $\log (\sigma_{\qty{6}{\raiseto{2}\centi\meter},\epsilon})$ and a weakly informative normal prior with mean $0$ and standard deviation $1$ on $\beta_{S,\epsilon}$.
-
-##### \aax{} for each light intensity with leaf using \agcurve{} curve parameters
-
-Within the $k^{\text{th}}$ leaf, we estimated \aax{} for each light intensity by integrating the difference in \logA{} between the amphi and pseudohypo \agcurve{} curves over the range of \gsw{} values where the curves overlap (from $\min(\log(g_\text{sw}))$ to $\max(\log(g_\text{sw}))$).
-The estimate of \aax{} for the $k^{\text{th}}$ leaf at light intensity $l$ in accession $m$ is:
-
-$$\widehat{\mathrm{AA}}_{klm} = \int_{\text{min(log(}g_\text{sw}))}^{\text{max(log(}g_\text{sw}))} \text{log}\bigg(\frac{\hat{A}_\text{amphi}(x; \theta_{klm,\text{amphi})}}{\hat{A}_\text{hypo}(x; \theta_{klm,\text{hypo})}}\bigg) dx$$
-where:
-
-$$\theta_\text{amphi} \in \{\hat{b}_{0, f(\text{amphi}, k,l,m)}, \hat{b}_{1, f(\text{amphi}, k,l,m)}, \hat{b}_{2, f(\text{amphi}, k,l,m)}\}, \text{and}$$
-
-$$\theta_\text{hypo} \in \{\hat{b}_{0, f(\text{hypo}, k,l,m)}, \hat{b}_{1, f(\text{hypo}, k,l,m)}, \hat{b}_{2, f(\text{hypo}, k,l,m)}\}.$$ 
-The function $f : \Theta_1 \rightarrow \Theta_2$ maps the set $\Theta_1$ indexed by leaf type (amphi or pseudohypo), leaf replicate, light intensity, and accession to set $\Theta_2$ indexed by individual \agcurve{} curve.
-This mapping is necessary because the random effects structure differs between models of \loggsw{} on \logA{} and that of models predicting \aax{} described in the next section.
-
-##### Effects of light intensity, light treatment, and accession on \aax
-
-We tested for effects of light intensity, light treatment, accession, and their interactions on \aax.
-All models included effects of light intensity and light treatment, as well as random effects of accession and replicate within accession.
-More complex models included interactions between light intensity and light treatment, as well as random effects of accession on the effects of light intensity and light treatment.
-We used a weakly informative normal prior with mean $0$ and standard deviation $10$ for fixed effects of light intensity and treatment.
-We used a weakly informative normal prior with mean $-3$ and standard deviation $5$ for the the random effect standard deviation of replicate within accession.
-We accounted for the phylogenetic structure among the random effects of accession using an Ornstein-Uhlenbeck (OU) process [@hansen_stabilizing_1997].
-The expected covariance between accessions $i$ and $j$ ($\text{Cov}(i, j)$ ) is:
-
-$$\text{Cov}(i, j) = \frac{\sigma^2}{2 \alpha} \exp(-\alpha D_{ij})$$
-where $\sigma^2$ is the variance of the random effect, $\alpha$ is the rate of decay of the covariance with phylogenetic distance, $D_{ij}$, the phylogenetic distance between accessions $i$ and $j$.
-We estimated $\sigma^2 / (2 \alpha)$ and $\alpha$ as a separate parameters and reparameterized them as $\sigma^2$ and $\alpha$.
-We used a weakly informative normal prior with mean $0$ and standard deviation $10$ on OU parameters.
-
-We modeled the residual standard deviation of \aax, (i.e.
-phylogenetically unstructured variation unaccounted for by explanatory variables) on a log-link scale with effects of light intensity and light treatment.
-We used a weakly informative normal prior with mean $-3$ and standard deviation $5$ on the residual standard deviation intercept and a weakly informative normal prior with mean $0$ and standard deviation $1$ on the effects of light intensity and light treatment on the residual standard deviation.
-
-##### Effects of native light habitat on accession-level \aax
-
-We tested a linear effect of native light habitat on accession-level \aax.
-In Models 1 and 2, we used the random intercept of \aax{} as a response variable; in model 3 we used the random intercept plus the random effect of accession at high light intensity; in model 4 we used the random intercept plus the random effect of accession in the sun treatment; in model 5 we used the random intercept plus the random effects of accession at high light intensity and sun treatment.
-We used a weakly informative normal prior with mean $0$ and standard deviation $1$ for the slope and intercept.
-We accounted for the phylogenetic structure among the model residuals using an (OU) process.
-We used a weakly informative normal prior with mean $0$ and standard deviation $10$ on OU parameters.
-
-#### Predictions {#sec-predictions}
-
-The assimilatory, plastic, and constitutive hypotheses make different predictions about the relationship between \aax{} and light intensity, light treatment, and native \ppfd{} among accessions.
-Since these hypotheses are not mutually exclusive, we describe how we assessed support for one, two, or all three hypotheses simultaneously in \autoref{tbl-predictions}.
-In general, the assimilatory hypothesis was supported if \aax{} was greater at high light intensity than low light intensity.
-The plastic hypothesis was supported if \aax{} was greater in sun leaves than shade leaves.
-The constitutive hypothesis was supported if accession-level \aax{} increased with native \ppfd.
-In interactive models, we only consider positively reinforcing interactions between high light intensity, sun leaves, and native \ppfd{} because these are the only interactions which could explain why amphistomatous leaves are advantageous in high light habitats.
-
-We evaluated predictions using a combination of parameter estimation and model selection.
-In all cases, we estimated parameters and confidence intervals as described in @sec-fitting.
-If the 95% confidence intervals for a parameter did not overlap zero, we considered the parameter to be significantly different from zero.
-We used the leave-one-out cross-validation information criterion (LOOIC) to compare the fit of models (\autoref{tbl-models}) using the *R* package **loo** version `r packageVersion("loo")` [@vehtari_practical_2017] to calculate LOOIC values.
-Note that LOOIC was calculated only from pointwise likelihood values of the submodel estimating effects of light intensity, light treatments, and accession on \aax{} (see @sec-parameters).
-We considered models with two standard errors of the mean lower LOOIC value to be a better fit to the data; we considered models with LOOIC values within two standard errors of the mean to be have similar support.
-
-```{r, predictions, eval=TRUE, echo=FALSE,message=FALSE,warning=FALSE}
-#| label: tbl-predictions
-#| tbl-cap: "Predictions of competing hypotheses about the relationship between \\aax{} and light intensity, light treatment, and native \\ppfd{} among accessions.
-The middle column lists specific directional predictions about parameter values and model fit according to leave-one-out cross-validation information criterion (LOOIC), where $\\text{LOOIC}_i$ is the LOOIC value for model $i$.
-The rightmost column describes the predictions in words and explains how accession-level \\aax{} is calculated in the relevant model."
-
-tribble(
-  ~"\\textbf{Hypothesis}", ~"\\textbf{Prediction(s)}", ~"\\textbf{Description}",
-  "Assimilatory", "$\\beta_{\\mathrm{AA},2000} > 0$", "Average \\aax{} at high light intensity is greater than that at low light intensity",
-  "Plastic", "$\\beta_{\\mathrm{AA},\\text{sun}} > 0$", "Average \\aax{} in sun leaves is greater than that in shade leaves",
-  "Constitutive", "$\\beta_{\\mathrm{PPFD,AA}} > 0$", "Accession-level \\aax{} ($\\mathrm{AA}_\\text{acc}$) increases with native \\ppfd",
-  "Constitutive", "", "$\\mathrm{AA}_\\text{acc} = \\beta_{\\mathrm{AA}, 0} + \\vec{\\beta}_{\\mathrm{AA}, \\text{acc}}$",
-  "Assimilatory $\\times$ Plastic", "$\\beta_{\\mathrm{AA},2000,\\text{sun}} > 0$", "Average \\aax{} is highest at high light intensity in sun leaves",
-  "Assimilatory $\\times$ Plastic", "$\\text{LOOIC}_\\text{1} > \\text{LOOIC}_\\text{2}$", "",
-  "Assimilatory $\\times$ Constitutive", "$\\beta_{\\mathrm{AA},2000} > 0$", "Average \\aax{} at high light intensity is greater than that at low light intensity",
-  "Assimilatory $\\times$ Constitutive", "$\\beta_{\\mathrm{PPFD,AA}} > 0$", "Accession-level \\aax{} at high light intensity ($\\mathrm{AA}_{\\text{acc},2000}$) increases with native \\ppfd",
-  "Assimilatory $\\times$ Constitutive", "$\\text{LOOIC}_\\text{1} > \\text{LOOIC}_\\text{3}$", "$\\mathrm{AA}_{\\text{acc},2000} = \\beta_{\\mathrm{AA}, 0} + \\vec{\\beta}_{\\mathrm{AA}, \\text{acc}} + \\vec{\\beta}_{\\mathrm{AA}, 2000, \\text{acc}}$",
-  "Plastic $\\times$ Constitutive", "$\\beta_{\\mathrm{AA},\\text{sun}} > 0$", "Average \\aax{} in sun leaves is greater than that in shade leaves",
-  "Plastic $\\times$ Constitutive", "$\\beta_{\\mathrm{PPFD,AA}} > 0$", "Accession-level \\aax{} in sun leaves ($\\mathrm{AA}_{\\text{acc},\\text{sun}}$) increases with native \\ppfd",
-  "Plastic $\\times$ Constitutive", "$\\text{LOOIC}_\\text{1} > \\text{LOOIC}_\\text{4}$", "$\\mathrm{AA}_{\\text{acc},\\text{sun}} = \\beta_{\\mathrm{AA}, 0} + \\vec{\\beta}_{\\mathrm{AA}, \\text{acc}} + \\vec{\\beta}_{\\mathrm{AA}, \\text{sun}, \\text{acc}}$",
-  "Assimilatory $\\times$ Plastic $\\times$ Constitutive", "$\\beta_{\\mathrm{AA},2000,\\text{sun}} > 0$", "Average \\aax{} is highest at high light intensity in sun leaves",
-  "Assimilatory $\\times$ Plastic $\\times$ Constitutive", "$\\beta_{\\mathrm{PPFD,AA}} > 0$", "Accession-level \\aax{} at high light intensity in sun leaves ($\\mathrm{AA}_{\\text{acc},2000,\\text{sun}}$) increases with native \\ppfd",
-  "Assimilatory $\\times$ Plastic $\\times$ Constitutive", "$\\text{LOOIC}_\\text{3} > \\text{LOOIC}_\\text{5}$", "$\\mathrm{AA}_{\\text{acc},2000,\\text{sun}} = \\beta_{\\mathrm{AA}, 0} + \\vec{\\beta}_{\\mathrm{AA}, \\text{acc}} + \\vec{\\beta}_{\\mathrm{AA}, 2000, \\text{acc}} + \\vec{\\beta}_{\\mathrm{AA}, \\text{sun}, \\text{acc}}$"
-
-) |>
-  knitr::kable(
-    format = "latex",
-    align = c("l", "l", "l"),
-    booktabs = TRUE,
-    escape = FALSE,
-    longtable = TRUE
-  ) |>
-  kableExtra::add_indent(positions = seq_len(15), level_of_indent = -1, target_cols = 3) |>
-  kableExtra::collapse_rows(1, latex_hline = "major", valign = "top",
-                            row_group_label_position = "first") |>
-  kableExtra::column_spec(1, width = "1in") |>
-  kableExtra::column_spec(2, width = "1.5in") |>
-  kableExtra::column_spec(3, width = "3in")
-
-```
-
-## Acknowledgements
-
-Justin Alter, Max Gatlin, Joana Kim, Jenna Matsuyama, Brandon Najarian, Kai Yasuda. Sarah Friedrich helped with figures.
-
-## References

### 2025-07-04 ms/ms.qmd
added aa_loo1 table to SI

@@
 
+aa_loo_table1 = read_rds("../objects/aa_loo_table1.rds")
+aa_loo_table2 = read_rds("../objects/aa_loo_table2.rds")
+
 n_plant = length(unique(rh_curves$acc_id))
@@
 We estimated 'amphistomy advantage' (\aax) *sensu* @parkhurst_adaptive_1978, but with modifications previously described in @triplett_amphistomy_2024 and here (Materials and Methods).
-The native light intensity was represented by plant area index (PAI $\unit{\meter\squared\per\meter\squared}$), estimated using a global gridded X-m2 resolution data set derived from the Global Ecosystem Dynamics Investigation [GEDI; @burns_multi-resolution_2024] and georeferenced accession collection information from the Tomato Genetics Resource Center.
+The native light intensity was represented by plant area index (PAI $\unit{\meter\squared\per\meter\squared}$), estimated using a global gridded data set derived from the Global Ecosystem Dynamics Investigation [GEDI; @burns_multi-resolution_2024] and georeferenced accession collection information from the Tomato Genetics Resource Center.
 The growth light intensities were \ppfdequals{`r ppfd_sun`} (sun treatment) and \ppfdqty{`r ppfd_shade`} (shade treatment) while all other environment conditions were nearly identical (see materials and methods in supplementary materials).
@@
 @fig-aa]A).
-\aax{} varied substantially between measurement light intensities, growth light intensities, and among populations (point to model comparison evidence).
+Bayesian phylogenetic mixed effects models that allowed \aax{} to vary between measurement light intensities, growth light intensities, and among populations outperformed simpler models based on their information criteria (@tbl-aa_loo1).
 Measured under high light intensity, \aax{} was consistently greater for sun plants.
@@
   ) |>
-  kableExtra::kable_styling(latex_options = c("striped", "no_vertical_lines")) |>
+  kableExtra::kable_styling(latex_options = c("striped", "no_vertical_lines", "repeat_header")) |>
   kableExtra::column_spec(1, width = "3.25cm", italic = TRUE)
@@
 
-### Cleaning \aqcurve{} curves
-
-The raw data set consisted of `r prettyNum(aa_stats$n_aq_curve0, big.mark = ",")` \aqcurve{} curves with an average of `r formatC(aa_stats$n_point_per_aq_curve0, digits = 3)`.
-Manual curation of a data set this size in a principled, consistent manner is not feasible.
-Therefore, we automated data cleaning using custom *R* scripts.
-Cleaning is divided into two sequential steps (\autoref{tbl-cleaning2}).
-
-```{r, cleaning2,echo=FALSE,message=FALSE,warning=FALSE}
-#| label: tbl-cleaning2
-#| tbl-cap: "Two sequential steps for cleaning \\agcurve{} curves.
-The rationale and procedure for each step are described in the text.
-The rightmost columns summarize the number of curves and mean number of points per curve remaining after each step."
-
-step_description_col = c(
-  "1. remove outliers within each \\aqcurve{} curve",
-  "2. remove \\aqcurve{} curves with poor fit"
-)
-
-n_curve_col = unlist(aa_stats[str_detect(names(aa_stats), "^n_aq_curve[1-9]$")]) |>
-  prettyNum(big.mark = ",")
-
-n_point_col = unlist(aa_stats[str_detect(names(aa_stats), "^n_point_per_aq_curve[1-9]$")])
-
-tibble(
-  `Step: description` = step_description_col,
-  `Number of curves` = n_curve_col,
-  `Number of points per curve` = n_point_col
-) |>
-  knitr::kable(format = "latex",
-               align = c("l", "c", "c"),
-               digits = c(0, 0, 1),
-               booktabs = TRUE, escape = FALSE) |>
-  kableExtra::kable_styling(latex_options = c("striped", "no_vertical_lines")) |>
-  kableExtra::column_spec(1, width = "9cm") |>
-  kableExtra::column_spec(2, width = "2.5cm") |>
-  kableExtra::column_spec(3, width = "2.5cm")
-
-```
-
-#### Remove outliers within each \aqcurve{} curve
-
-*Rationale*: Individual outliers within \agcurve{} curves, usually caused by transitory changes in chamber conditions, exert undue leverage on parameter estimates and cause bias and/or low precision in parameter estimates.
-
-*Procedure*: We fit provisional nonrectangular hyperbola [@marshall_model_1980] to each \aqcurve{} curve using nonlinear regression with the `nlsLM()` function from the *R* package **minpack.lm** version `r packageVersion("minpack.lm")` [@elzhov_minpacklm_2023].
-We sequentially removed data points with an absolute external studentized residual $> `r point_outlier_threshold`$ until none remained.
-
-#### Remove \aqcurve{} curves with poor fit
-
-*Rationale*: \aqcurve{} curves with a poor fit to the nonrectangular hyperbola most likely indicate systematic measurement error and/or the leaf was not fully acclimated to the chamber environment.
-
-*Procedure*: As described above, we fit provisional nonrectangular hyperbola to each \aqcurve{} curve and calculated the model $r^2$.
-There was a clear break between typical curves and poorly fitting curves where $r^2 < `r aq_r2_threshold`$.
-We therefore removed \aqcurve{} curves with $r^2 < `r aq_r2_threshold`$.
-
-### Bayesian data analysis in *Stan*
-
+### Bayesian phylogenetic mixed effects models in *Stan*
+#### MARKER - REVISE MODEL DESCRIPTION
 We fit `r n_model` models to test predictions of competing hypotheses about why amphistomy advantage (\aax) might be greater for leaves in sunny, open habitats.
@@
 
-{{< pagebreak >}}
-
-### MANUSCRAPS
-
-(not sure where this goes)  * We acclimated the focal leaf to high light (\ppfdequals{2000}) and high relative humidity (\rhequals{70}) until $A$ and \gsw{} reach their maximum.
-After that, we decreased \rh{} to $\approx 10\%$ to induce rapid stomatal closure without biochemical down
-
-Problems to bring up in discussion:
-- all our species were amphi, so they may all live in PAI below threshold to favor amphi
-- other microclimate diffs not considered
-- don't know specific anatomical change
-
-Two outstanding areas for further research are 1) the specific anatomical changes, such as leaf thickness or mesophyll porosity, that increase the diffusion path length for CO$_2$ in sun leaves, and 2) the costs of amphistomy that could explain plasticity in \gmaxratio{} in populations from low-light habitats.
-Resolving the first question will require detailed anatomical measurements on a large number of shade and sun leaves with concomitant estimates of \aax.
-The fact that populations from low-light habitats are more plastic suggests that costs of amphistomy are greatest in these environments.
-(not finished yet)
-
-Not sure where this goes: We infer this from the fact that blocking gas exchange in pseudohypostomatous leaves reduced $A$ by X-X% depending on the population, light treatment, and light intensity (Table/figure).
-The AA is equivalent to an X-X% change in total $g_\text{sc}$ (see SI section gs equivalency).
-But whereas increasing $g_\text{sc}$ would increase water loss as a necessary by-product, amphistomy can increase $A$ without any appreciable affect on transpiration.
-
-from Mott and Oleary: "However, Jones and Slatyer (5) reported a higher mesophyll resistance for CO, entering through the upper stomata than for the lower, and the data of Vaclavik (12) appear to support this conclusion"
-
-8. What it all means
-  * Our study reveals an important role for plasticity in explaining why amphi leaves are more common in sunny habitats
-  * Confirms that leaf structure of sun leaves incurs costs that amphistomy can mitigate
-  * It will be useful to consider coordination between leaf anatomy, stomatal density, and ratio in future studies of leaf structure-function and in crops (or something)
-
-I am using this approach to drafting the paper:
-
-https://blogs.nature.com/nyc/2011/08/10/how-to-write-a-paper-one-possible-answer
-
-## Punchline
-
-Plastic changes in the structure of sun leaves can increase the intercellular CO$_2$ diffusion path length.
-Gas exchange through stomata on both upper and lower leaf surfaces, amphistomy, may be a compensatory mechanism to reduce the impact of increased diffusion path length on photosynthesis without increasing water loss.
-We demonstrate for the first time that sun leaves benefit most from amphistomy because it compensates for plastic structural changes in leaf anatomy.
-
-What's surprising: role of developmental plasticity in explaining comparative results; diffusional limitations through palisade ias is strong and needs to be considered with light gradients in understanding leaf strucuture- funciton
-
-big implications - if trees were amphi, more more carbon sink or reduced transpiration
-
-It is commonly assumed in comparative studies that most trait variation between species is constitutive and genetically determined, but our results suggest that plastic responses to light are necessary to explain why \gmaxratio{} increases with light habitat.
-Reaction norms to light intensity among wild tomatoes indicate that populations from low light habitats express low \gmaxratio{} when grown in shade, whereas populations from high light habitats have constitutively higher \gmaxratio{} (Fig.
-rxn-norm).
-The larger response to light intensity is driven by greater plasticty in leaf structure in populations from low-light habitats (Fig/table - I haven't tested this yet, so this is a placeholder.) We infer that these structural changes in sun leaves are costly because they increase the diffusion path length for CO$_2$ and reduce photosynthesis.
-As a consequence, sun leaves benefit more from amphistomy.
-This is the first demonstration that sun leaves benefit more from amphistomy because it compensates for developmental plasticity in leaf structure.
+```{r, aa_loo1, echo=FALSE,message=FALSE,warning=FALSE}
+#| label: tbl-aa_loo1
+#| tbl-cap: 'Leave-one-out cross-validation information criterion (LOOIC) for models of \\aax{} as a function of light intensity, light treatment, and population.
+The model with the lowest LOOIC is considered the best fit to the data.
+The LOOIC column lists the LOOIC value for each model; the $\\Delta$LOOIC column lists the difference between the LOOIC of the corresponding model and the best model (lowest LOOIC).
+Checkmarks in each row indicate whether the corresponding model included affects of measurement light intensity ("Meas.") or growth light intensity ("Growth") as fixed effects, random effects, or distributional effect on the residual variance.
+We also test for fixed and random interaction effects ("Inter.") between measurement and growth light intensity.
+All models included phylogenetically structured random effects of population.'
+
+df1 = aa_loo_table1 |>
+  mutate(
+    fixed_meas = str_detect(fixed, "light_intensity"),
+    fixed_growth = str_detect(fixed, "light_treatment"),
+    fixed_intrxn = str_detect(fixed, "\\*"),
+    # random_acc = TRUE,
+    random_meas = str_detect(random, "light_intensity"),
+    random_growth = str_detect(random, "light_treatment"),
+    random_intrxn = str_detect(random, "\\*"),
+    sigma_meas = str_detect(sigma, "light_intensity"),
+    sigma_growth = str_detect(sigma, "light_treatment"),
+  ) |>
+  dplyr::select(
+    LOOIC = looic,
+    `$\\Delta$LOOIC` = delta_looic,
+    starts_with("fixed_"),
+    starts_with("random_"),
+    starts_with("sigma_")
+  ) |>
+  mutate(across(where(is.logical), ~ ifelse(.x, "$\\checkmark$", "")))
+
+c1 = colnames(df1)
+groupings = list(
+  " " = sum(str_detect(colnames(df1), "LOOIC")),
+  "Fixed" = sum(str_detect(colnames(df1), "fixed_")),
+  "Random" = sum(str_detect(colnames(df1), "random_")),
+  "Distributional" = sum(str_detect(colnames(df1), "sigma_"))
+)
+
+c2 = case_when(
+  str_detect(c1, "_meas") ~ "Meas.",
+  str_detect(c1, "_growth") ~ "Growth",
+  str_detect(c1, "_intrxn") ~ "Inter.",
+  str_detect(c1, "_acc") ~ "Population",
+  .default = c1
+)
+
+kbl(
+  df1,
+  col.names = c2,
+  booktabs = TRUE,
+  escape = FALSE,
+  align = "c",
+  longtable = TRUE,
+  digits = 2
+) |>
+  kable_styling(latex_options = c("striped", "repeat_header")) |>
+  add_header_above(groupings) |>
+  column_spec(1, width = "0.7in") |>
+  column_spec(2, width = "0.5in") |>
+  column_spec(3:10, width = "0.4in") |>
+  column_spec(c(5, 8), border_right = TRUE)
+
+```

### 2025-07-04 ms/ms.tex
added aa_loo1 table to SI

@@
 % Make \paragraph and \subparagraph free-standing
-\makeatletter
 \ifx\paragraph\undefined\else
   \let\oldparagraph\paragraph
-  \renewcommand{\paragraph}{
-    \@ifstar
-      \xxxParagraphStar
-      \xxxParagraphNoStar
-  }
-  \newcommand{\xxxParagraphStar}[1]{\oldparagraph*{#1}\mbox{}}
-  \newcommand{\xxxParagraphNoStar}[1]{\oldparagraph{#1}\mbox{}}
+  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
 \fi
@@
   \let\oldsubparagraph\subparagraph
-  \renewcommand{\subparagraph}{
-    \@ifstar
-      \xxxSubParagraphStar
-      \xxxSubParagraphNoStar
-  }
-  \newcommand{\xxxSubParagraphStar}[1]{\oldsubparagraph*{#1}\mbox{}}
-  \newcommand{\xxxSubParagraphNoStar}[1]{\oldsubparagraph{#1}\mbox{}}
+  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
 \fi
-\makeatother
 
@@
 \makeatletter
-\newsavebox\pandoc@box
-\newcommand*\pandocbounded[1]{% scales image to fit in text height/width
-  \sbox\pandoc@box{#1}%
-  \Gscale@div\@tempa{\textheight}{\dimexpr\ht\pandoc@box+\dp\pandoc@box\relax}%
-  \Gscale@div\@tempb{\linewidth}{\wd\pandoc@box}%
-  \ifdim\@tempb\p@<\@tempa\p@\let\@tempa\@tempb\fi% select the smaller of both
-  \ifdim\@tempa\p@<\p@\scalebox{\@tempa}{\usebox\pandoc@box}%
-  \else\usebox{\pandoc@box}%
-  \fi%
-}
+\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
+\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
+\makeatother
+% Scale images if necessary, so that they will not overflow the page
+% margins by default, and it is still possible to overwrite the defaults
+% using explicit options in \includegraphics[width, height, ...]{}
+\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
 % Set default figure placement to htbp
+\makeatletter
 \def\fps@figure{htbp}
@@
 \makeatother
-
+\ifLuaTeX
+  \usepackage{selnolig}  % disable illegal ligatures
+\fi
 \usepackage{bookmark}
@@
 
-\pandocbounded{\includegraphics[keepaspectratio]{../figures/amphisomy draft 2024-03-28.jpg}}
+\includegraphics[width=2.24in,height=\textheight]{../figures/amphisomy draft 2024-03-28.jpg}
 
@@
 index (PAI \(\unit{\meter\squared\per\meter\squared}\)), estimated using
-a global gridded X-m2 resolution data set derived from the Global
-Ecosystem Dynamics Investigation {[}GEDI; (\emph{52}){]} and
-georeferenced accession collection information from the Tomato Genetics
-Resource Center. The growth light intensities were \ppfdequals{761} (sun
-treatment) and \ppfdqty{115} (shade treatment) while all other
-environment conditions were nearly identical (see materials and methods
-in supplementary materials). The high and low measurements intensities
-were \ppfdequals{2000} (97.8:2.24 red:blue) and \ppfdequals{150}
-(87.0:13.0 red:blue), respectively.
+a global gridded data set derived from the Global Ecosystem Dynamics
+Investigation {[}GEDI; (\emph{52}){]} and georeferenced accession
+collection information from the Tomato Genetics Resource Center. The
+growth light intensities were \ppfdequals{761} (sun treatment) and
+\ppfdqty{115} (shade treatment) while all other environment conditions
+were nearly identical (see materials and methods in supplementary
+materials). The high and low measurements intensities were
+\ppfdequals{2000} (97.8:2.24 red:blue) and \ppfdequals{150} (87.0:13.0
+red:blue), respectively.
 
 Consistent with biophysical theory of CO\(_2\) diffusion within leaves,
-\(\aay{} > 0\) for all populations (Fig.~\ref{fig-aa}A). \aax{} varied
-substantially between measurement light intensities, growth light
-intensities, and among populations (point to model comparison evidence).
-Measured under high light intensity, \aax{} was consistently greater for
-sun plants. The average \aax{} among populations in the shade treatment
-was 0.039 (range: 0.012--0.109; 21 of 29 populations significant);
-however, the same populations grown at high light intensity showed a
-mean \aax{} of 0.050 (range: 0.024--0.120; 25 of 29 populations
-significant). Contrary to the predictions of the assimilatory
-hypothesis, \aax{} was greater in all populations under low measurement
-light intensity for both sun and shade grown plants. The effect of low
-light on \aax{} was more pronounced in the sun-grown plants, where
-\aax{} was significantly greater under low measurement light intensity
-in 28 of 29 populations compared to to 11 populations for shade-grown
-plants. The overall average \aax{} of shade and sun grown plants
-measured under low light intensity was 0.066 (range: 0.019--0.143; 27 of
-29 populations significant) and 0.098 (range: 0.051--0.175; 29 of 29
-populations significant), respectively. There was a slight tendency for
-populations from more closed habitats (greater PAI), but \aax{} varied
-widely among populations from open habitats (low PAI) regardless of
-growth and measurement light intensities (Fig.~\ref{fig-aa}B).
+\(\aay{} > 0\) for all populations (Fig.~\ref{fig-aa}A). Bayesian
+phylogenetic mixed effects models that allowed \aax{} to vary between
+measurement light intensities, growth light intensities, and among
+populations outperformed simpler models based on their information
+criteria (Table~\ref{tbl-aa_loo1}). Measured under high light intensity,
+\aax{} was consistently greater for sun plants. The average \aax{} among
+populations in the shade treatment was 0.039 (range: 0.012--0.109; 21 of
+29 populations significant); however, the same populations grown at high
+light intensity showed a mean \aax{} of 0.050 (range: 0.024--0.120; 25
+of 29 populations significant). Contrary to the predictions of the
+assimilatory hypothesis, \aax{} was greater in all populations under low
+measurement light intensity for both sun and shade grown plants. The
+effect of low light on \aax{} was more pronounced in the sun-grown
+plants, where \aax{} was significantly greater under low measurement
+light intensity in 28 of 29 populations compared to to 11 populations
+for shade-grown plants. The overall average \aax{} of shade and sun
+grown plants measured under low light intensity was 0.066 (range:
+0.019--0.143; 27 of 29 populations significant) and 0.098 (range:
+0.051--0.175; 29 of 29 populations significant), respectively. There was
+a slight tendency for populations from more closed habitats (greater
+PAI), but \aax{} varied widely among populations from open habitats (low
+PAI) regardless of growth and measurement light intensities
+(Fig.~\ref{fig-aa}B).
 
@@
 
-\pandocbounded{\includegraphics[keepaspectratio]{ms_files/figure-pdf/fig-aa-1.pdf}}
+\includegraphics{ms_files/figure-pdf/fig-aa-1.pdf}
 
@@
 
-\pandocbounded{\includegraphics[keepaspectratio]{ms_files/figure-pdf/fig-aacap-1.pdf}}
+\includegraphics{ms_files/figure-pdf/fig-aacap-1.pdf}
 
@@
 
-\pandocbounded{\includegraphics[keepaspectratio]{ms_files/figure-pdf/fig-lma_aa-1.pdf}}
+\includegraphics{ms_files/figure-pdf/fig-lma_aa-1.pdf}
 
@@
 differences.
+
+maybe this wording is better: Amphistomy likely plays a significant, but
+under appreciated, role in the carbon and water balance of plant groups
+where this trait is common, such as herbaceous plants in sunny habitats
+and woody species in arid biomes (cites). If our estimates of AA from
+the wild tomato clades
 
@@
 
-\bibitem[\citeproctext]{ref-marshall_model_1980}
+\bibitem[\citeproctext]{ref-stan_development_team_stan_2025}
 \CSLLeftMargin{59. }%
-\CSLRightInline{B. Marshall, P. V. Biscoe,
-\href{https://doi.org/10.1093/jxb/31.1.29}{A {Model} for
-{C}\(_{\textrm{3}}\) {Leaves} {Describing} the {Dependence} of {Net}
-{Photosynthesis} on {Irradiance}}. \emph{Journal of Experimental Botany}
-\textbf{31}, 29--39 (1980).}
-
-\bibitem[\citeproctext]{ref-elzhov_minpacklm_2023}
-\CSLLeftMargin{60. }%
-\CSLRightInline{T. V. Elzhov, K. M. Mullen, A.-N. Spiess, B. M. Bolker,
-\emph{Minpack.lm: {R} {Interface} to the {Levenberg}-{Marquardt}
-{Nonlinear} {Least}-{Squares} {Algorithm} {Found} in {MINPACK}, {Plus}
-{Support} for {Bounds}} (2023;
-\url{https://CRAN.R-project.org/package=minpack.lm}).}
-
-\bibitem[\citeproctext]{ref-stan_development_team_stan_2025}
-\CSLLeftMargin{61. }%
 \CSLRightInline{Stan Development Team, \emph{Stan {Modeling} {Language}
@@
 \bibitem[\citeproctext]{ref-burkner_brms_2017}
-\CSLLeftMargin{62. }%
+\CSLLeftMargin{60. }%
 \CSLRightInline{P.-C. Bürkner,
@@
 \bibitem[\citeproctext]{ref-gabry_cmdstanr_2025}
-\CSLLeftMargin{63. }%
+\CSLLeftMargin{61. }%
 \CSLRightInline{J. Gabry, R. Češnovar, A. Johnson, S. Bronder,
@@
 \bibitem[\citeproctext]{ref-r_core_team_r:_2025}
-\CSLLeftMargin{64. }%
+\CSLLeftMargin{62. }%
 \CSLRightInline{R Core Team, \emph{R: {A} {Language} and {Environment}
@@
 \bibitem[\citeproctext]{ref-gelman_inference_1992}
-\CSLLeftMargin{65. }%
+\CSLLeftMargin{63. }%
 \CSLRightInline{A. Gelman, D. B. Rubin, Inference from iterative
@@
 \bibitem[\citeproctext]{ref-hansen_stabilizing_1997}
-\CSLLeftMargin{66. }%
+\CSLLeftMargin{64. }%
 \CSLRightInline{T. F. Hansen,
@@
 \bibitem[\citeproctext]{ref-vehtari_practical_2017}
-\CSLLeftMargin{67. }%
+\CSLLeftMargin{65. }%
 \CSLRightInline{A. Vehtari, A. Gelman, J. Gabry,
@@
 \midrule
+\endfirsthead
+\multicolumn{6}{@{}l}{\textit{(continued)}}\\
+\toprule
+Species & TGRC accession & Latitude & Longitude & Elevation (mas) & $\mathrm{PAI}~(\unit{\meter\squared\per\meter\squared})$\\
+\midrule
+\endhead
+
+\endfoot
+\bottomrule
+\endlastfoot
 \em{\cellcolor{gray!10}{S. arcanum}} & \cellcolor{gray!10}{LA2172} & \cellcolor{gray!10}{-6.008} & \cellcolor{gray!10}{-78.858} & \cellcolor{gray!10}{662} & \cellcolor{gray!10}{0.70}\\
@@
 \em{S. pimpinellifolium} & LA2933 & -1.442 & -80.562 & 375 & 1.62\\
-\em{\cellcolor{gray!10}{S. sitiens}} & \cellcolor{gray!10}{LA4116} & \cellcolor{gray!10}{-22.159} & \cellcolor{gray!10}{-68.782} & \cellcolor{gray!10}{2960} & \cellcolor{gray!10}{0.06}\\
-\bottomrule
+\em{\cellcolor{gray!10}{S. sitiens}} & \cellcolor{gray!10}{LA4116} & \cellcolor{gray!10}{-22.159} & \cellcolor{gray!10}{-68.782} & \cellcolor{gray!10}{2960} & \cellcolor{gray!10}{0.06}\\*
 
@@
 
-\subsubsection{\texorpdfstring{Cleaning \aqcurve{}
-curves}{Cleaning  curves}}\label{cleaning-curves-1}
-
-The raw data set consisted of 658 \aqcurve{} curves with an average of
-19.4. Manual curation of a data set this size in a principled,
-consistent manner is not feasible. Therefore, we automated data cleaning
-using custom \emph{R} scripts. Cleaning is divided into two sequential
-steps (\autoref{tbl-cleaning2}).
-
-\begin{table}
-
-\caption{\label{tbl-cleaning2}Two sequential steps for cleaning
-\agcurve{} curves. The rationale and procedure for each step are
-described in the text. The rightmost columns summarize the number of
-curves and mean number of points per curve remaining after each step.}
-
-\centering{
-
-\centering
-\begin{tabular}{>{\raggedright\arraybackslash}p{9cm}>{\centering\arraybackslash}p{2.5cm}>{\centering\arraybackslash}p{2.5cm}}
-\toprule
-Step: description & Number of curves & Number of points per curve\\
-\midrule
-\cellcolor{gray!10}{1. remove outliers within each \aqcurve{} curve} & \cellcolor{gray!10}{658} & \cellcolor{gray!10}{19.1}\\
-2. remove \aqcurve{} curves with poor fit & 652 & 19.1\\
-\bottomrule
-\end{tabular}
-
-}
-
-\end{table}%
-
-\paragraph{\texorpdfstring{Remove outliers within each \aqcurve{}
-curve}{Remove outliers within each  curve}}\label{remove-outliers-within-each-curve-1}
-
-\emph{Rationale}: Individual outliers within \agcurve{} curves, usually
-caused by transitory changes in chamber conditions, exert undue leverage
-on parameter estimates and cause bias and/or low precision in parameter
-estimates.
-
-\emph{Procedure}: We fit provisional nonrectangular hyperbola
-(\emph{59}) to each \aqcurve{} curve using nonlinear regression with the
-\texttt{nlsLM()} function from the \emph{R} package \textbf{minpack.lm}
-version 1.2.4 (\emph{60}). We sequentially removed data points with an
-absolute external studentized residual \(> 3\) until none remained.
-
-\paragraph{\texorpdfstring{Remove \aqcurve{} curves with poor
-fit}{Remove  curves with poor fit}}\label{remove-curves-with-poor-fit}
-
-\emph{Rationale}: \aqcurve{} curves with a poor fit to the
-nonrectangular hyperbola most likely indicate systematic measurement
-error and/or the leaf was not fully acclimated to the chamber
-environment.
-
-\emph{Procedure}: As described above, we fit provisional nonrectangular
-hyperbola to each \aqcurve{} curve and calculated the model \(r^2\).
-There was a clear break between typical curves and poorly fitting curves
-where \(r^2 < 0.99\). We therefore removed \aqcurve{} curves with
-\(r^2 < 0.99\).
-
-\subsubsection{\texorpdfstring{Bayesian data analysis in
-\emph{Stan}}{Bayesian data analysis in Stan}}\label{bayesian-data-analysis-in-stan}
+\subsubsection{\texorpdfstring{Bayesian phylogenetic mixed effects
+models in
+\emph{Stan}}{Bayesian phylogenetic mixed effects models in Stan}}\label{bayesian-phylogenetic-mixed-effects-models-in-stan}
+
+\paragraph{MARKER - REVISE MODEL
+DESCRIPTION}\label{marker---revise-model-description}
 
@@
 We fit Bayesian models with MCMC sampling in the probabilistic
-programming language \emph{Stan} (\emph{61}) using the \emph{R} package
-\textbf{brms} version 2.22.0 (\emph{62}). We used \emph{CmdStan} version
-2.35.0 and \textbf{cmdstanr} version 0.9.0 (\emph{63}) to interface with
-\emph{R} version 4.5.0 (\emph{64}). We sampled the posterior
+programming language \emph{Stan} (\emph{59}) using the \emph{R} package
+\textbf{brms} version 2.22.0 (\emph{60}). We used \emph{CmdStan} version
+2.35.0 and \textbf{cmdstanr} version 0.9.0 (\emph{61}) to interface with
+\emph{R} version 4.5.0 (\emph{62}). We sampled the posterior
 distribution from 1 chains with 2000 iterations each after 2000 warmup
@@
 iterations, and maximum treedepth so that parameter estimates converged
-(\(\hat{R} < 1.01\) (\emph{65})) and the effective sample size (ESS) for
+(\(\hat{R} < 1.01\) (\emph{63})) and the effective sample size (ESS) for
 each parameter was \(> 400\).
@@
 phylogenetic structure among the random effects of population using an
-Ornstein-Uhlenbeck (OU) process (\emph{66}). The expected covariance
+Ornstein-Uhlenbeck (OU) process (\emph{64}). The expected covariance
 between populations \(i\) and \(j\) (\(\text{Cov}(i, j)\) ) is:
@@
 the fit of models (\autoref{tbl-models}) using the \emph{R} package
-\textbf{loo} version 2.8.0 (\emph{67}) to calculate LOOIC values. Note
+\textbf{loo} version 2.8.0 (\emph{65}) to calculate LOOIC values. Note
 that LOOIC was calculated only from pointwise likelihood values of the
@@
 
-\newpage{}
-
-\subsubsection{MANUSCRAPS}\label{manuscraps}
-
-(not sure where this goes) * We acclimated the focal leaf to high light
-(\ppfdequals{2000}) and high relative humidity (\rhequals{70}) until
-\(A\) and \gsw{} reach their maximum. After that, we decreased \rh{} to
-\(\approx 10\%\) to induce rapid stomatal closure without biochemical
-down
-
-Problems to bring up in discussion: - all our species were amphi, so
-they may all live in PAI below threshold to favor amphi - other
-microclimate diffs not considered - don't know specific anatomical
-change
-
-Two outstanding areas for further research are 1) the specific
-anatomical changes, such as leaf thickness or mesophyll porosity, that
-increase the diffusion path length for CO\(_2\) in sun leaves, and 2)
-the costs of amphistomy that could explain plasticity in \gmaxratio{} in
-populations from low-light habitats. Resolving the first question will
-require detailed anatomical measurements on a large number of shade and
-sun leaves with concomitant estimates of \aax. The fact that populations
-from low-light habitats are more plastic suggests that costs of
-amphistomy are greatest in these environments. (not finished yet)
-
-Not sure where this goes: We infer this from the fact that blocking gas
-exchange in pseudohypostomatous leaves reduced \(A\) by X-X\% depending
-on the population, light treatment, and light intensity (Table/figure).
-The AA is equivalent to an X-X\% change in total \(g_\text{sc}\) (see SI
-section gs equivalency). But whereas increasing \(g_\text{sc}\) would
-increase water loss as a necessary by-product, amphistomy can increase
-\(A\) without any appreciable affect on transpiration.
-
-from Mott and Oleary: ``However, Jones and Slatyer (5) reported a higher
-mesophyll resistance for CO, entering through the upper stomata than for
-the lower, and the data of Vaclavik (12) appear to support this
-conclusion''
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\setcounter{enumi}{7}
-\tightlist
-\item
-  What it all means
-\end{enumerate}
-
-\begin{itemize}
-\tightlist
-\item
-  Our study reveals an important role for plasticity in explaining why
-  amphi leaves are more common in sunny habitats
-\item
-  Confirms that leaf structure of sun leaves incurs costs that
-  amphistomy can mitigate
-\item
-  It will be useful to consider coordination between leaf anatomy,
-  stomatal density, and ratio in future studies of leaf
-  structure-function and in crops (or something)
-\end{itemize}
-
-I am using this approach to drafting the paper:
-
-https://blogs.nature.com/nyc/2011/08/10/how-to-write-a-paper-one-possible-answer
-
-\subsection{Punchline}\label{punchline}
-
-Plastic changes in the structure of sun leaves can increase the
-intercellular CO\(_2\) diffusion path length. Gas exchange through
-stomata on both upper and lower leaf surfaces, amphistomy, may be a
-compensatory mechanism to reduce the impact of increased diffusion path
-length on photosynthesis without increasing water loss. We demonstrate
-for the first time that sun leaves benefit most from amphistomy because
-it compensates for plastic structural changes in leaf anatomy.
-
-What's surprising: role of developmental plasticity in explaining
-comparative results; diffusional limitations through palisade ias is
-strong and needs to be considered with light gradients in understanding
-leaf strucuture- funciton
-
-big implications - if trees were amphi, more more carbon sink or reduced
-transpiration
-
-It is commonly assumed in comparative studies that most trait variation
-between species is constitutive and genetically determined, but our
-results suggest that plastic responses to light are necessary to explain
-why \gmaxratio{} increases with light habitat. Reaction norms to light
-intensity among wild tomatoes indicate that populations from low light
-habitats express low \gmaxratio{} when grown in shade, whereas
-populations from high light habitats have constitutively higher
-\gmaxratio{} (Fig. rxn-norm). The larger response to light intensity is
-driven by greater plasticty in leaf structure in populations from
-low-light habitats (Fig/table - I haven't tested this yet, so this is a
-placeholder.) We infer that these structural changes in sun leaves are
-costly because they increase the diffusion path length for CO\(_2\) and
-reduce photosynthesis. As a consequence, sun leaves benefit more from
-amphistomy. This is the first demonstration that sun leaves benefit more
-from amphistomy because it compensates for developmental plasticity in
-leaf structure.
+\begin{longtable}[t]{>{\centering\arraybackslash}p{0.7in}>{\centering\arraybackslash}p{0.5in}>{\centering\arraybackslash}p{0.4in}>{\centering\arraybackslash}p{0.4in}>{}c|>{\centering\arraybackslash}p{0.4in}>{\centering\arraybackslash}p{0.4in}>{}c|>{\centering\arraybackslash}p{0.4in}>{\centering\arraybackslash}p{0.4in}}
+
+\caption{\label{tbl-aa_loo1}Leave-one-out cross-validation information
+criterion (LOOIC) for models of \aax{} as a function of light intensity,
+light treatment, and population. The model with the lowest LOOIC is
+considered the best fit to the data. The LOOIC column lists the LOOIC
+value for each model; the \(\Delta\)LOOIC column lists the difference
+between the LOOIC of the corresponding model and the best model (lowest
+LOOIC). Checkmarks in each row indicate whether the corresponding model
+included affects of measurement light intensity (``Meas.'') or growth
+light intensity (``Growth'') as fixed effects, random effects, or
+distributional effect on the residual variance. We also test for fixed
+and random interaction effects (``Inter.'') between measurement and
+growth light intensity. All models included phylogenetically structured
+random effects of population.}
+
+\tabularnewline
+
+\toprule
+\multicolumn{2}{c}{ } & \multicolumn{3}{c}{Fixed} & \multicolumn{3}{c}{Random} & \multicolumn{2}{c}{Distributional} \\
+\cmidrule(l{3pt}r{3pt}){3-5} \cmidrule(l{3pt}r{3pt}){6-8} \cmidrule(l{3pt}r{3pt}){9-10}
+LOOIC & $\Delta$LOOIC & Meas. & Growth & Inter. & Meas. & Growth & Inter. & Meas. & Growth\\
+\midrule
+\endfirsthead
+\multicolumn{10}{@{}l}{\textit{(continued)}}\\
+\toprule
+LOOIC & $\Delta$LOOIC & Meas. & Growth & Inter. & Meas. & Growth & Inter. & Meas. & Growth\\
+\midrule
+\endhead
+
+\endfoot
+\bottomrule
+\endlastfoot
+\cellcolor{gray!10}{-1839.94} & \cellcolor{gray!10}{0.00} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$}\\
+-1839.19 & 0.76 &  &  &  & $\checkmark$ & $\checkmark$ & $\checkmark$ & $\checkmark$ & $\checkmark$\\
+\cellcolor{gray!10}{-1838.83} & \cellcolor{gray!10}{1.11} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$}\\
+-1838.44 & 1.50 & $\checkmark$ & $\checkmark$ & $\checkmark$ & $\checkmark$ & $\checkmark$ &  & $\checkmark$ & $\checkmark$\\
+\cellcolor{gray!10}{-1838.42} & \cellcolor{gray!10}{1.53} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$}\\
+\addlinespace
+-1838.33 & 1.61 & $\checkmark$ &  &  & $\checkmark$ & $\checkmark$ &  & $\checkmark$ & $\checkmark$\\
+\cellcolor{gray!10}{-1838.13} & \cellcolor{gray!10}{1.81} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$}\\
+-1838.05 & 1.90 &  & $\checkmark$ &  & $\checkmark$ &  &  & $\checkmark$ & $\checkmark$\\
+\cellcolor{gray!10}{-1837.62} & \cellcolor{gray!10}{2.32} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$}\\
+-1837.23 & 2.71 &  &  &  & $\checkmark$ & $\checkmark$ &  & $\checkmark$ & $\checkmark$\\
+\addlinespace
+\cellcolor{gray!10}{-1836.90} & \cellcolor{gray!10}{3.05} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$}\\
+-1836.76 & 3.19 & $\checkmark$ & $\checkmark$ & $\checkmark$ &  &  &  & $\checkmark$ & $\checkmark$\\
+\cellcolor{gray!10}{-1836.51} & \cellcolor{gray!10}{3.44} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$}\\
+-1836.46 & 3.48 & $\checkmark$ & $\checkmark$ &  & $\checkmark$ &  &  & $\checkmark$ & $\checkmark$\\
+\cellcolor{gray!10}{-1836.29} & \cellcolor{gray!10}{3.65} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$}\\
+\addlinespace
+-1835.53 & 4.41 & $\checkmark$ & $\checkmark$ &  &  & $\checkmark$ &  & $\checkmark$ & $\checkmark$\\
+\cellcolor{gray!10}{-1835.08} & \cellcolor{gray!10}{4.86} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$}\\
+-1834.95 & 4.99 & $\checkmark$ & $\checkmark$ &  &  &  &  & $\checkmark$ & $\checkmark$\\
+\cellcolor{gray!10}{-1831.01} & \cellcolor{gray!10}{8.93} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$}\\
+-1828.80 & 11.14 & $\checkmark$ &  &  & $\checkmark$ &  &  & $\checkmark$ & $\checkmark$\\
+\addlinespace
+\cellcolor{gray!10}{-1828.36} & \cellcolor{gray!10}{11.58} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$}\\
+-1828.35 & 11.59 & $\checkmark$ &  &  & $\checkmark$ & $\checkmark$ & $\checkmark$ &  & $\checkmark$\\
+\cellcolor{gray!10}{-1828.24} & \cellcolor{gray!10}{11.70} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$}\\
+-1827.57 & 12.37 & $\checkmark$ & $\checkmark$ &  & $\checkmark$ &  &  &  & $\checkmark$\\
+\cellcolor{gray!10}{-1827.43} & \cellcolor{gray!10}{12.51} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$}\\
+\addlinespace
+-1827.34 & 12.60 & $\checkmark$ &  &  & $\checkmark$ & $\checkmark$ &  &  & $\checkmark$\\
+\cellcolor{gray!10}{-1827.29} & \cellcolor{gray!10}{12.65} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$}\\
+-1827.05 & 12.89 &  & $\checkmark$ &  & $\checkmark$ & $\checkmark$ &  &  & $\checkmark$\\
+\cellcolor{gray!10}{-1826.92} & \cellcolor{gray!10}{13.03} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$}\\
+-1826.77 & 13.17 &  & $\checkmark$ &  & $\checkmark$ &  &  &  & $\checkmark$\\
+\addlinespace
+\cellcolor{gray!10}{-1826.69} & \cellcolor{gray!10}{13.25} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$}\\
+-1826.30 & 13.64 &  & $\checkmark$ &  & $\checkmark$ & $\checkmark$ & $\checkmark$ &  & $\checkmark$\\
+\cellcolor{gray!10}{-1826.26} & \cellcolor{gray!10}{13.68} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$}\\
+-1826.16 & 13.79 & $\checkmark$ & $\checkmark$ & $\checkmark$ &  & $\checkmark$ &  &  & $\checkmark$\\
+\cellcolor{gray!10}{-1826.00} & \cellcolor{gray!10}{13.94} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$}\\
+\addlinespace
+-1825.83 & 14.11 & $\checkmark$ & $\checkmark$ &  & $\checkmark$ & $\checkmark$ &  &  & $\checkmark$\\
+\cellcolor{gray!10}{-1824.27} & \cellcolor{gray!10}{15.68} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$}\\
+-1823.90 & 16.04 & $\checkmark$ & $\checkmark$ &  &  &  &  &  & $\checkmark$\\
+\cellcolor{gray!10}{-1823.59} & \cellcolor{gray!10}{16.36} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$}\\
+-1822.71 & 17.23 & $\checkmark$ &  &  & $\checkmark$ & $\checkmark$ & $\checkmark$ & $\checkmark$ & \\
+\addlinespace
+\cellcolor{gray!10}{-1822.60} & \cellcolor{gray!10}{17.34} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{}\\
+-1822.29 & 17.65 & $\checkmark$ & $\checkmark$ &  & $\checkmark$ & $\checkmark$ & $\checkmark$ & $\checkmark$ & \\
+\cellcolor{gray!10}{-1821.08} & \cellcolor{gray!10}{18.87} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{}\\
+-1821.06 & 18.88 &  & $\checkmark$ &  & $\checkmark$ & $\checkmark$ & $\checkmark$ & $\checkmark$ & \\
+\cellcolor{gray!10}{-1820.56} & \cellcolor{gray!10}{19.39} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{}\\
+\addlinespace
+-1819.66 & 20.28 &  &  &  & $\checkmark$ & $\checkmark$ & $\checkmark$ & $\checkmark$ & \\
+\cellcolor{gray!10}{-1819.58} & \cellcolor{gray!10}{20.37} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{}\\
+-1819.01 & 20.94 & $\checkmark$ & $\checkmark$ &  & $\checkmark$ & $\checkmark$ &  & $\checkmark$ & \\
+\cellcolor{gray!10}{-1818.51} & \cellcolor{gray!10}{21.43} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{}\\
+-1818.44 & 21.51 &  & $\checkmark$ &  & $\checkmark$ &  &  & $\checkmark$ & \\
+\addlinespace
+\cellcolor{gray!10}{-1818.43} & \cellcolor{gray!10}{21.51} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{}\\
+-1818.43 & 21.51 & $\checkmark$ & $\checkmark$ &  & $\checkmark$ &  &  & $\checkmark$ & \\
+\cellcolor{gray!10}{-1818.26} & \cellcolor{gray!10}{21.68} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{}\\
+-1818.23 & 21.71 & $\checkmark$ & $\checkmark$ &  &  &  &  & $\checkmark$ & \\
+\cellcolor{gray!10}{-1818.15} & \cellcolor{gray!10}{21.80} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{}\\
+\addlinespace
+-1817.85 & 22.09 & $\checkmark$ &  &  & $\checkmark$ &  &  &  & $\checkmark$\\
+\cellcolor{gray!10}{-1817.07} & \cellcolor{gray!10}{22.87} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$}\\
+-1816.97 & 22.98 & $\checkmark$ &  &  &  & $\checkmark$ &  & $\checkmark$ & \\
+\cellcolor{gray!10}{-1815.82} & \cellcolor{gray!10}{24.12} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{}\\
+-1814.29 & 25.65 & $\checkmark$ &  &  &  &  &  &  & $\checkmark$\\
+\addlinespace
+\cellcolor{gray!10}{-1811.97} & \cellcolor{gray!10}{27.97} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{}\\
+-1811.85 & 28.10 & $\checkmark$ &  &  & $\checkmark$ &  &  & $\checkmark$ & \\
+\cellcolor{gray!10}{-1811.44} & \cellcolor{gray!10}{28.50} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{}\\
+-1811.06 & 28.88 & $\checkmark$ &  &  & $\checkmark$ & $\checkmark$ & $\checkmark$ &  & \\
+\cellcolor{gray!10}{-1810.70} & \cellcolor{gray!10}{29.24} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{}\\
+\addlinespace
+-1809.72 & 30.23 & $\checkmark$ & $\checkmark$ &  & $\checkmark$ & $\checkmark$ & $\checkmark$ &  & \\
+\cellcolor{gray!10}{-1809.67} & \cellcolor{gray!10}{30.27} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{}\\
+-1809.41 & 30.54 & $\checkmark$ &  &  & $\checkmark$ & $\checkmark$ &  &  & \\
+\cellcolor{gray!10}{-1809.15} & \cellcolor{gray!10}{30.79} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{}\\
+-1809.09 & 30.86 & $\checkmark$ & $\checkmark$ & $\checkmark$ & $\checkmark$ & $\checkmark$ &  &  & \\
+\addlinespace
+\cellcolor{gray!10}{-1809.03} & \cellcolor{gray!10}{30.91} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{}\\
+-1808.89 & 31.06 &  &  &  & $\checkmark$ & $\checkmark$ &  &  & \\
+\cellcolor{gray!10}{-1808.76} & \cellcolor{gray!10}{31.18} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{}\\
+-1807.88 & 32.06 &  & $\checkmark$ &  & $\checkmark$ &  &  &  & \\
+\cellcolor{gray!10}{-1807.84} & \cellcolor{gray!10}{32.10} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{}\\
+\addlinespace
+-1807.16 & 32.78 & $\checkmark$ & $\checkmark$ &  & $\checkmark$ &  &  &  & \\
+\cellcolor{gray!10}{-1806.33} & \cellcolor{gray!10}{33.61} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{}\\
+-1806.04 & 33.90 &  & $\checkmark$ &  &  &  &  & $\checkmark$ & $\checkmark$\\
+\cellcolor{gray!10}{-1805.87} & \cellcolor{gray!10}{34.07} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{}\\
+-1805.61 & 34.33 & $\checkmark$ & $\checkmark$ & $\checkmark$ &  & $\checkmark$ &  &  & \\
+\addlinespace
+\cellcolor{gray!10}{-1805.31} & \cellcolor{gray!10}{34.64} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$}\\
+-1804.67 & 35.27 & $\checkmark$ & $\checkmark$ &  &  &  &  &  & \\
+\cellcolor{gray!10}{-1804.53} & \cellcolor{gray!10}{35.41} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{}\\
+-1804.49 & 35.45 &  &  &  &  & $\checkmark$ &  & $\checkmark$ & $\checkmark$\\
+\cellcolor{gray!10}{-1800.46} & \cellcolor{gray!10}{39.49} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{}\\
+\addlinespace
+-1799.42 & 40.52 &  &  &  & $\checkmark$ &  &  &  & \\
+\cellcolor{gray!10}{-1797.93} & \cellcolor{gray!10}{42.01} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$}\\
+-1795.85 & 44.09 & $\checkmark$ &  &  &  &  &  &  & \\
+\cellcolor{gray!10}{-1794.84} & \cellcolor{gray!10}{45.10} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$}\\
+-1793.22 & 46.72 &  &  &  &  & $\checkmark$ &  &  & $\checkmark$\\
+\addlinespace
+\cellcolor{gray!10}{-1792.49} & \cellcolor{gray!10}{47.45} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$}\\
+-1787.14 & 52.80 &  &  &  &  &  &  &  & $\checkmark$\\
+\cellcolor{gray!10}{-1783.17} & \cellcolor{gray!10}{56.77} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{}\\
+-1781.97 & 57.97 &  &  &  &  & $\checkmark$ &  & $\checkmark$ & \\
+\cellcolor{gray!10}{-1780.61} & \cellcolor{gray!10}{59.33} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{}\\
+\addlinespace
+-1777.91 & 62.04 &  &  &  &  &  &  & $\checkmark$ & \\
+\cellcolor{gray!10}{-1771.89} & \cellcolor{gray!10}{68.06} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{}\\
+-1771.87 & 68.08 &  &  &  &  & $\checkmark$ &  &  & \\
+\cellcolor{gray!10}{-1770.84} & \cellcolor{gray!10}{69.10} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{}\\
+-1766.11 & 73.83 &  &  &  &  &  &  &  & \\*
+
+\end{longtable}
 

### 2025-07-04 r/204_select-model1.R

@@ -36,7 +36,10 @@ model_forms = expand.grid(fixed = fixed_effects,
                           random = random_effects,
                           sigma = sigma_effects,
                           stringsAsFactors = FALSE) %>%
-  mutate(seed = sample(1e9, nrow(.)))
+  mutate(
+    seed = sample(1e9, nrow(.)),
+    model = paste0("model_", row_number())
+  )
 
 # Build and fit each model
 plan(multisession, workers = 19)
@@ -81,8 +84,10 @@ aa_loo_table = tibble(
   mutate(
     delta_looic = looic - min(looic),
     best_model = delta_looic == 0
-  )
+  ) |>
+  left_join(model_forms, by = join_by(model))
 
+# Write results
 write_rds(aa_loo_table, "objects/aa_loo_table1.rds")
 best_model_index = str_extract(aa_loo_table[1, "model"], "\\d+") |>
   as.integer()

### 2025-07-06 ms/ms.qmd
tidying up fitting and plotting

@@
 ### Bayesian data analysis
-n_model = 5 |> # i don't think I need this anymore
-  number_to_word()
+n_model1 = nrow(aa_loo_table1)
+n_model2 = nrow(aa_loo_table2)
 
@@
 
+note: estimates of AA are less than that described by Parkhurst and Mott (1990) and also less than that of Xiong and Flexas (2021). What does this tell us?
+
 A second major conclusion is that the significant photosynthetic gain of amphistomatous leaves implies that the many hypostomatous leaves that dominate mesic to wet forests globally are giving up 'free' carbon that would require little to no additional water loss.
@@
 
-### Bayesian phylogenetic mixed effects models in *Stan*
-#### MARKER - REVISE MODEL DESCRIPTION
-We fit `r n_model` models to test predictions of competing hypotheses about why amphistomy advantage (\aax) might be greater for leaves in sunny, open habitats.
-This section provides an overview of differences among models (\autoref{tbl-models}).
-The next sections describe how we fit models in *Stan*, all model parameters and priors, and specific predictions about parameter values for each hypothesis.
-
-```{r, models, echo=FALSE,message=FALSE,warning=FALSE}
-#| label: tbl-models
-#| tbl-cap: "Summary of differences among competing models of how \\aax{} varies with light intensity, light treatment, and among populations as a function of native \\ppfd.
-The models are numbered from simpler to more complex.
-All models include fixed effects of light intensity and light treatment; some models include interactions between.
-All models include a phylogenetic random effect of population on \\aax; some models include varying effects of light intensity and light treatment among populations.
-The last column indicates the population-level \\aax{} variable we used as a response to native \\ppfd."
-
-tribble(
-    ~"Model", ~"Fixed effects", ~"Phylogenetic random effects", ~"Response to native \\ppfd",
-    "1", "light intensity", "varying intercept among populations", "$\\mathrm{AA}_{0, \\text{acc}}$",
-    "1", "light treatment", "", "",
-    "2", "light intensity", "varying intercept among populations", "$\\mathrm{AA}_{0, \\text{acc}}$",
-    "2", "light treatment", "", "",
-    "2", "intensity $\\times$ treatment", "", "",
-    "3", "light intensity", "varying intercept among populations", "$\\mathrm{AA}_{2000, \\text{acc}}$",
-    "3", "light treatment", "varying effect of high light intensity among populations", "",
-    "4", "light intensity", "varying intercept among populations", "$\\mathrm{AA}_{\\text{sun, acc}}$",
-    "4", "light treatment", "varying effect of sun treatment among populations", "",
-    "5", "light intensity", "varying intercept among populations", "$\\mathrm{AA}_{2000, \\text{sun, acc}}$",
-    "5", "light treatment", "varying effect of high light intensity among populations", "",
-    "5", "intensity $\\times$ treatment", "varying effect of sun treatment among populations", ""
-  ) |>
-  knitr::kable(
-    format = "latex",
-    align = c("c", "l", "l", "l"),
-    booktabs = TRUE,
-    escape = FALSE,
-    longtable = FALSE
-  ) |>
-    kableExtra::collapse_rows(columns = 1, latex_hline = "major", row_group_label_position = "first") |>
-  kableExtra::column_spec(1, width = "0.3in") |>
-  kableExtra::column_spec(2, width = "1.3in") |>
-  kableExtra::column_spec(3, width = "3.4in") |>
-  kableExtra::column_spec(4, width = "1in")
-
-```
-
-#### Fitting models in *Stan* {#sec-fitting}
-
-We fit Bayesian models with MCMC sampling in the probabilistic programming language *Stan* [@stan_development_team_stan_2025] using the *R* package **brms** version `r fit_aa1$version$brms` [@burkner_brms_2017].
-We used *CmdStan* version `r fit_aa1$version$cmdstan` and **cmdstanr** version `r fit_aa1$version$cmdstanr` [@gabry_cmdstanr_2025] to interface with *R* version `r getRversion()` [@r_core_team_r:_2025].
-We sampled the posterior distribution from `r dim_fit_aa[2]` chains with `r dim_fit_aa[1]` iterations each after `r dim_fit_aa[1]` warmup iterations per chain.
-We estimated parameters and confidence intervals as the median and 95% quantile intervals of the posterior, respectively.
-We chose the number of chains, warmup and sampling iterations, and maximum treedepth so that parameter estimates converged ($\hat{R} < `r aa_args['max_rhat']`$ [@gelman_inference_1992]) and the effective sample size (ESS) for each parameter was $> `r aa_args['min_ess']`$.
-
-#### Parameter estimation and priors {#sec-parameters}
-
-There were four levels of parameter estimation in our analysis:
-
-1. Estimate \agcurve{} curve parameters
-2. Estimate \aax{} for each light intensity with leaf using \agcurve{} curve parameters
+### Bayesian data analysis with *Stan* {#sec-analysis}
+
+Our analysis pipeline consisted of four main steps:
+
+1. Estimate \agcurve{} curve parameters for each leaf
+2. Estimate \aax{} for both measurement light intensities using \agcurve{} curve parameters
 3. Estimate the effects of light intensity, light treatment, and population on \aax{} (assimilatory and plasticity hypotheses)
@@
 
-Although the higher-level parameter estimates depend on the lower-level parameter estimates, we fit all models simultaneously to ensure that the uncertainty in lower-level estimates propagated to higher-level estimates.
-
-```{r, parameters, echo=FALSE,message=FALSE,warning=FALSE}
-#| label: tbl-parameters
-#| tbl-cap: "Description of parameters estimated in the hierarchical Bayesian model.
-The `Parameter` column lists the parameter name as it appears in text.
-The `Description` column provides a brief description of the parameter."
-
-df_parameters = read_csv("../data/parameters1.csv", col_types = "ccccl") |>
-    dplyr::filter(include)
-
-df_parameters |>
-  dplyr::select(Parameter = latex, Description = description) |>
-  knitr::kable(
-    format = "latex",
-    align = c("l", "l"),
-    booktabs = TRUE,
-    escape = FALSE,
-    longtable = TRUE
-  ) |>
-  kableExtra::kable_styling(latex_options = c("striped", "no_vertical_lines")) |>
-  kableExtra::column_spec(1, width = "1in") |>
-  kableExtra::column_spec(2, width = "5in") |>
-  kableExtra::pack_rows("\\agcurve{} curve parameters",
-                        min(which(df_parameters$par_type == "ags")),
-                        max(which(df_parameters$par_type == "ags")),
-                        escape = FALSE) |>
-  kableExtra::pack_rows("\\aax{} for each light intensity with leaf using \\agcurve{} curve parameters",
-                        min(which(df_parameters$par_type == "aa_est")),
-                        max(which(df_parameters$par_type == "aa_est")),
-                        escape = FALSE) |>
-  kableExtra::pack_rows("effects of light intensity, light treatments, and population on \\aax",
-                        min(which(df_parameters$par_type == "aa")),
-                        max(which(df_parameters$par_type == "aa")),
-                        escape = FALSE) |>
-  kableExtra::pack_rows("effects of native light habitat on population-level \\aax",
-                        min(which(df_parameters$par_type == "ppfd_aa")),
-                        max(which(df_parameters$par_type == "ppfd_aa")),
-                        escape = FALSE)
-
-```
-
-##### \agcurve{} curve parameters
-
-We modeled \logA as a quadratic function of \loggsw for each leaf using the following equation:
-
-$$\log(A_{ij}) = (\beta_0 + b_{0,j}) + (\beta_1 + b_{1,j}) \log{g_{\text{sw},i}} + (\beta_2 + b_{2,j}) \log{g_{\text{sw},i}}^2 + \epsilon_{i}$$
-
-where $\beta_0$, $\beta_1$, and $\beta_2$ are the average intercept, linear, and quadratic coefficients, respectively.
-We used diffuse normal priors with mean $0$ and standard deviation $10$ on these parameters.
-We estimated random effects of curve $j$ on the intercept ($b_{0,j}$), linear ($b_{1,j}$), and quadratic ($b_{2,j}$) coefficients.
-We assumed that the $j \times 3$ array of coefficients was multivariate normal a mean vector of $\vec{0}$ and covariance $\Sigma_\text{curve}$.
-We used a weakly informative normal prior with mean $0$ and standard deviation $1$ on the log-transformed standard deviations (i.e.
-the diagonal of $\Sigma_\text{curve}$).
-We used a weakly informative $\mathrm{LJK}(2)$ prior on the correlation matrix.
-The off diagonal elements of $\Sigma_\text{curve}$ can be calculated from its diagonal elements and the correlation matrix.
-
-The residuals $\epsilon_{i}$ were modeled as a lag-1 autocorrelated time-series.
-We further assumed that the residual standard deviation of the $j^{\text{th}}$ curve ($\sigma_{\epsilon,j}$) was inversely proportional to the leaf surface area ($S_j$) within the chamber:
-
-$$\log (\sigma_{\epsilon,j}) = \log (\sigma_{\qty{6}{\raiseto{2}\centi\meter},\epsilon}) + (6 - S_j) \beta_{S,\epsilon}$$
-where $\sigma_{\qty{6}{\raiseto{2}\centi\meter},\epsilon}$ is the minimum residual standard deviation when the $\qty{6}{\raiseto{2}\centi\meter}$ chamber is completely filled.
-The residual standard deviation increases on log-linear scale by $\beta_{S,\epsilon}$.
-We used a weakly informative normal prior with mean $-3$ and standard deviation $5$ on $\log (\sigma_{\qty{6}{\raiseto{2}\centi\meter},\epsilon})$ and a weakly informative normal prior with mean $0$ and standard deviation $1$ on $\beta_{S,\epsilon}$.
-
-##### \aax{} for each light intensity with leaf using \agcurve{} curve parameters
-
-Within the $k^{\text{th}}$ leaf, we estimated \aax{} for each light intensity by integrating the difference in \logA{} between the amphi and pseudohypo \agcurve{} curves over the range of \gsw{} values where the curves overlap (from $\min(\log(g_\text{sw}))$ to $\max(\log(g_\text{sw}))$).
-The estimate of \aax{} for the $k^{\text{th}}$ leaf at light intensity $l$ in population $m$ is:
-
-$$\widehat{\mathrm{AA}}_{klm} = \int_{\text{min(log(}g_\text{sw}))}^{\text{max(log(}g_\text{sw}))} \text{log}\bigg(\frac{\hat{A}_\text{amphi}(x; \theta_{klm,\text{amphi})}}{\hat{A}_\text{hypo}(x; \theta_{klm,\text{hypo})}}\bigg) dx$$
-where:
-
-$$\theta_\text{amphi} \in \{\hat{b}_{0, f(\text{amphi}, k,l,m)}, \hat{b}_{1, f(\text{amphi}, k,l,m)}, \hat{b}_{2, f(\text{amphi}, k,l,m)}\}, \text{and}$$
-
-$$\theta_\text{hypo} \in \{\hat{b}_{0, f(\text{hypo}, k,l,m)}, \hat{b}_{1, f(\text{hypo}, k,l,m)}, \hat{b}_{2, f(\text{hypo}, k,l,m)}\}.$$ 
-The function $f : \Theta_1 \rightarrow \Theta_2$ maps the set $\Theta_1$ indexed by leaf type (amphi or pseudohypo), leaf replicate, light intensity, and population to set $\Theta_2$ indexed by individual \agcurve{} curve.
-This mapping is necessary because the random effects structure differs between models of \loggsw{} on \logA{} and that of models predicting \aax{} described in the next section.
+All models were using a Bayesian model with HMC sampling in the probabilistic programming language *Stan* [@stan_development_team_stan_2025] using the *R* package **brms** version `r fit_aa1$version$brms` [@burkner_brms_2017].
+We used *CmdStan* version `r fit_aa1$version$cmdstan` and **cmdstanr** version `r fit_aa1$version$cmdstanr` [@gabry_cmdstanr_2025] to interface with *R* version `r getRversion()` [@r_core_team_r:_2025].
+We sampled the posterior distribution from a single chain with a minimum of `r aa_args$n_iter_init / 2` iterations after `r aa_args$n_iter_init / 2` warmup iterations.
+If necessary, we refit models with more iterations to decrease the Gelman-Rubin convergence statistic ($\hat{R}$) [@gelman_inference_1992] to less than $`aa_stats[["max_rhat"]]`$ and the bulk effective sample size (ESS) to greater than $`aa_stats[["min_ess"]]`$ for all model parameters.
+We also increased the the `adapt_delta` parameter until there were fewer than $`aa_stats[["n_divergent"]]`$ divergent transitions.
+
+#### \agcurve{} curve parameters
+
+We modeled \logA as a quadratic function of \loggsw for each leaf, measured at low and high light intensity, in both amphistomatous (untreated) and pseudohypostomatous (propafilm treatment) states.
+At minimum, each model was fit using four chains with 2000 warmup and sampling iterations per chain.
+In cases where we encountered divergent transitions while fitting, we increased the number of sampling iterations and `adapt_delta` parameter until there were fewer than 10 divergences.
+
+#### \aax{} for each light intensity within leaf using \agcurve{} curve parameters
+##### MARKER - add figure of this
+We estimated \aax{} for each light intensity by integrating the difference in \logA{} between the amphi and pseudohypo \agcurve{} curves over the range of \gsw{} values where the curves overlap (from $\min(\log(g_\text{sw}))$ to $\max(\log(g_\text{sw}))$).
+
+$$\widehat{\mathrm{AA}} = \int_{\text{min(log(}g_\text{sw}))}^{\text{max(log(}g_\text{sw}))} \text{log}\bigg(\frac{\hat{A}_\text{amphi}(x; \theta_{\text{amphi}})}{\hat{A}_\text{hypo}(x; \theta_{\text{hypo}})}\bigg) dx$$
+where $\theta_\text{amphi}$ and $\theta_\text{hypo}$ are the quadratic parameters of the amphi and pseudohypo \agcurve{} curves, respectively.
+We calculated $\widehat{\mathrm{AA}}$ from each draw of the posterior for the paired amphi and pseudohypo \agcurve{} curves to obtain the posterior distribution of \aax{} for each leaf at each light intensity.
+We then summarized the posterior distribution of \aax{} for each leaf at each light intensity by the posterior median, which became our estimate of \aax{} for subsequent analyses.
+
+### Bayesian phylogenetic mixed effects models
+
+We fit Bayesian mixed effects models with phylogenetically structured random effects to test predictions of competing hypotheses about why amphistomy advantage (\aax) might be greater for leaves in sunny, open habitats.
+We fit all models combinations of measurement light intensity, growth light intensity, and their interaction as both fixed effects and random effects among populations.
+In other words, we test whether there were main effects of light intensity treatments and/or whether populations varied in their response to light intensity treatments.
+In all models, we included population as a phylogenetically structured random effect.
+We crossed those fixed and random effect structures with fixed effects of measurement light intensity and/or growth light intensity on the distributional parameter, $\sigma$, the residual variance in \aax{}.
+In total, we fit `r n_model1` models, each with a different combination of fixed, random, and distributional effects.
+We used the RAxML whole-transcriptome concatenated phylogeny based on 2,745 100-kb genomic windows from @pease_phylogenomics_2016.
+Two of our populations were not in this tree.
+We used accession LA1044 in place of LA3909, two populations of *S.
+galapagnse*; LA0750 was added as sister to LA0716, two closely replated populations of *S.
+pennellii*.
+The node was placed half-way between the next deepest node.
+
+NEED TO EDIT THIS: The next sections describe how we fit models in *Stan*, all model parameters and priors, and specific predictions about parameter values for each hypothesis.
+Each curve was fit For initial fits, we sampled the posterior distribution from a single chain with `r aa_args$n_iter_init / 2` iterations after `r aa_args$n_iter_init / 2` warmup iterations.
+After model selection (see below), we refit the selected model with more iterations if necessary to decrease the Gelman-Rubin convergence statistic ($\hat{R}$) [@gelman_inference_1992] to less than $`aa_stats[["max_rhat"]]`$ and the bulk effective sample size (ESS) to greater than $`aa_stats[["min_ess"]]`$ for all model parameters.
+
+#### Fitting models in *Stan* using **brms** interface {#sec-fitting}
+
+We fit Bayesian models with HMC sampling in the probabilistic programming language *Stan* [@stan_development_team_stan_2025] using the *R* package **brms** version `r fit_aa1$version$brms` [@burkner_brms_2017].
+We used *CmdStan* version `r fit_aa1$version$cmdstan` and **cmdstanr** version `r fit_aa1$version$cmdstanr` [@gabry_cmdstanr_2025] to interface with *R* version `r getRversion()` [@r_core_team_r:_2025].
+For initial fits, we sampled the posterior distribution from a single chain with `r aa_args$n_iter_init / 2` iterations after `r aa_args$n_iter_init / 2` warmup iterations.
+After model selection (see below), we refit the selected model with more iterations if necessary to decrease the Gelman-Rubin convergence statistic ($\hat{R}$) [@gelman_inference_1992] to less than $`aa_stats[["max_rhat"]]`$ and the bulk effective sample size (ESS) to greater than $`aa_stats[["min_ess"]]`$ for all model parameters.
 
@@
 * Table S9: Calculation of alternate scenario where amphi leaves can get same A with less R
+* Each A-gs figure with fitted curves, R2, and AA
 

### 2025-07-06 r/203_prepare-stan-data.R

@@ -14,7 +14,7 @@ trimmed_amphi_first = read_rds("data/trimmed_amphi_first.rds") |>
 phy = read_rds("data/phylogeny.rds")
 
 aa_summary = read_rds("objects/aa_summary.rds") |>
-  dplyr::select(acc_id, light_intensity, aa = median) |>
+  dplyr::select(acc_id, light_intensity, aa = median, se_aa = sd) |>
   mutate(scaled_aa = (aa - mean(aa)) / sd(aa)) |>
   left_join(plant_info, by = join_by(acc_id)) |>
   left_join(trimmed_amphi_first, by = join_by(acc_id)) 

### 2025-07-06 r/204_select-model1.R

@@ -1,4 +1,4 @@
-# Fit phylogenetic model on posterior estimate of AA 
+# Fit phylogenetic model on posterior estimate of AA
 source("r/header.R")
 
 d1 = read_rds("objects/stan_data_df.rds")
@@ -23,39 +23,70 @@ random_effects = c(
   "(1 + light_treatment * light_intensity | gr(acc, cov = A))"
 )
 
-sigma_effects = c(
-  "1",
-  "light_treatment",
-  "light_intensity",
-  "light_treatment + light_intensity"
-)
+sigma_effects = c("1",
+                  "light_treatment",
+                  "light_intensity",
+                  "light_treatment + light_intensity")
 
 # Cross all combinations
 set.seed(622232590)
-model_forms = expand.grid(fixed = fixed_effects,
-                          random = random_effects,
-                          sigma = sigma_effects,
-                          stringsAsFactors = FALSE) %>%
-  mutate(
-    seed = sample(1e9, nrow(.)),
-    model = paste0("model_", row_number())
-  )
+model_forms = expand.grid(
+  fixed = fixed_effects,
+  random = random_effects,
+  sigma = sigma_effects,
+  stringsAsFactors = FALSE
+) %>%
+  mutate(seed = sample(1e9, nrow(.)),
+         model = paste0("model_", row_number()))
 
 # Build and fit each model
-plan(multisession, workers = 19)
+plan(multisession, workers = 9)
 
-aa_models = future_pmap(model_forms, function(fixed, random, sigma, seed) {
-  fml = bf(as.formula(paste("aa ~", fixed, "+", random)), as.formula(paste("sigma ~", sigma)))
-  brm(
-    formula = fml,
-    data = d1,
-    data2 = list(A = A),
-    backend = "cmdstanr",
-    chains = 1,
-    family = student(),
-    save_pars = save_pars(all = TRUE),
-    seed = seed
-  )
+aa_models = model_forms |>
+  dlpyr::select(fixed, random, sigma, seed) |>
+  future_pmap(function(fixed, random, sigma, seed) {
+  fml = bf(as.formula(paste(
+    "aa | se(se_aa, sigma = TRUE) ~", fixed, "+", random
+  )), as.formula(paste("sigma ~", sigma)))
+  
+  crit = 0
+  x = 1
+  ad = 0.8
+  
+  while (crit == 0 & x < 24) {
+    m = brm(
[…]

### 2025-07-06 r/211_plot-curves.R

@@ -1,0 +1,263 @@
+# Plot quadratic fits to each curve
+source("r/header.R")
+
+# set plotting parameters
+fig_width = 6
+fig_height = 5
+fontsize_pt = 10
+hjust = -0.1
+vjust = 0.5
+
+# load data
+accession_info = read_rds("data/accession-info.rds")
+plant_info = read_rds("data/plant-info.rds")
+aa_summary = read_rds("objects/aa_summary.rds")
+fit_curve_diagnostics = read_rds("objects/fit_curve_diagnostics.rds") |>
+  mutate(
+    acc_id = str_extract(file, "LA[0-9]{4}A*-[A-Z]{1}[AB]{0,1}"),
+    `Curve type` = str_extract(file, "amphi|pseudohypo"),
+    light_intensity = str_extract(file, "150|2000")
+  )
+
+# checks
+assert_true(all(!is.na(fit_curve_diagnostics$acc_id))) # Check for NA values in acc_id
+assert_true(all(!is.na(fit_curve_diagnostics$`Curve type`))) # Check for NA values in `Curve type`
+assert_true(all(!is.na(fit_curve_diagnostics$light_intensity))) # Check for NA values in light_intensity
+
+plan(multisession, workers = 9)
+ags_curves = future_map(
+  unique(aa_summary$acc_id),
+  \(.x) {
+    # Extract fit, r2, data, and predictions for each curve type and light intensity
+    df1 = filter(aa_summary, acc_id == .x) |>
+      crossing(curve_type = c("amphi", "pseudohypo")) |>
+      mutate(
+        file = glue(
+          "objects/curve-fits/{acc_id}_{curve_type}_{light_intensity}.rds"
+        ),
+        fit = map(file, read_rds),
+        r2 = map(fit, bayes_R2),
+        data = map(fit, ~ .x$data),
+        pred = map2(fit, data, \(.x, .y) {
+          d1 = tibble(log_gsw = seq(min(.y$log_gsw), max(.y$log_gsw), length.out = 100))
+          predict(.x, newdata = d1) |>
+            as_tibble() |>
+            bind_cols(d1) |>
+            rename(log_A = Estimate)
+          
+        })
+      )
+    
+    # Data points
+    df_points = pmap_dfr(dplyr::select(df1, acc_id, light_intensity, curve_type, data),
+                         \(acc_id, light_intensity, curve_type, data) {
+                           tibble(
+                             acc_id = acc_id,
+                             light_intensity = light_intensity,
[…]

### 2025-07-07 ms/ms.qmd
updating select model 2

@@
 Contrary to the predictions of the assimilatory hypothesis, \aax{} was greater in all populations under low measurement light intensity for both sun and shade grown plants.
-The effect of low light on \aax{} was more pronounced in the sun-grown plants, where \aax{} was significantly greater under low measurement light intensity in `r aa_summary3["high"]` of `r n_acc` populations compared to to `r aa_summary3["low"]` populations for shade-grown plants.
+The effect of low light on \aax{} was more pronounced in the sun-grown plants, where \aax{} was significantly greater under low measurement light intensity in `r aa_summary3["high"]` of `r n_acc` populations compared to `r aa_summary3["low"]` populations for shade-grown plants.
 The overall average \aax{} of shade and sun grown plants measured under low light intensity was `r filter(aa_mean_range, light_treatment == "low", light_intensity == "150")$x` and `r filter(aa_mean_range, light_treatment == "high", light_intensity == "150")$x`, respectively.
@@
 We modeled \logA as a quadratic function of \loggsw for each leaf, measured at low and high light intensity, in both amphistomatous (untreated) and pseudohypostomatous (propafilm treatment) states.
-At minimum, each model was fit using four chains with 2000 warmup and sampling iterations per chain.
-In cases where we encountered divergent transitions while fitting, we increased the number of sampling iterations and `adapt_delta` parameter until there were fewer than 10 divergences.
 

### 2025-07-07 ms/ms.tex
updating select model 2

--- 2025-07-04 ms/ms.tex
@@
 index (PAI \(\unit{\meter\squared\per\meter\squared}\)), estimated using
-a global gridded X-m2 resolution data set derived from the Global
-Ecosystem Dynamics Investigation {[}GEDI; (\emph{52}){]} and
-georeferenced accession collection information from the Tomato Genetics
-Resource Center. The growth light intensities were \ppfdequals{761} (sun
-treatment) and \ppfdqty{115} (shade treatment) while all other
-environment conditions were nearly identical (see materials and methods
-in supplementary materials). The high and low measurements intensities
-were \ppfdequals{2000} (97.8:2.24 red:blue) and \ppfdequals{150}
-(87.0:13.0 red:blue), respectively.
+a global gridded data set derived from the Global Ecosystem Dynamics
+Investigation {[}GEDI; (\emph{52}){]} and georeferenced accession
+collection information from the Tomato Genetics Resource Center. The
+growth light intensities were \ppfdequals{761} (sun treatment) and
+\ppfdqty{115} (shade treatment) while all other environment conditions
+were nearly identical (see materials and methods in supplementary
+materials). The high and low measurements intensities were
+\ppfdequals{2000} (97.8:2.24 red:blue) and \ppfdequals{150} (87.0:13.0
+red:blue), respectively.
 
 Consistent with biophysical theory of CO\(_2\) diffusion within leaves,
-\(\aay{} > 0\) for all populations (Fig.~\ref{fig-aa}A). \aax{} varied
-substantially between measurement light intensities, growth light
-intensities, and among populations (point to model comparison evidence).
-Measured under high light intensity, \aax{} was consistently greater for
-sun plants. The average \aax{} among populations in the shade treatment
-was 0.039 (range: 0.012--0.109; 21 of 29 populations significant);
-however, the same populations grown at high light intensity showed a
-mean \aax{} of 0.050 (range: 0.024--0.120; 25 of 29 populations
-significant). Contrary to the predictions of the assimilatory
-hypothesis, \aax{} was greater in all populations under low measurement
-light intensity for both sun and shade grown plants. The effect of low
-light on \aax{} was more pronounced in the sun-grown plants, where
-\aax{} was significantly greater under low measurement light intensity
-in 28 of 29 populations compared to to 11 populations for shade-grown
-plants. The overall average \aax{} of shade and sun grown plants
-measured under low light intensity was 0.066 (range: 0.019--0.143; 27 of
-29 populations significant) and 0.098 (range: 0.051--0.175; 29 of 29
-populations significant), respectively. There was a slight tendency for
-populations from more closed habitats (greater PAI), but \aax{} varied
-widely among populations from open habitats (low PAI) regardless of
-growth and measurement light intensities (Fig.~\ref{fig-aa}B).
+\(\aay{} > 0\) for all populations (Fig.~\ref{fig-aa}A). Bayesian
+phylogenetic mixed effects models that allowed \aax{} to vary between
+measurement light intensities, growth light intensities, and among
+populations outperformed simpler models based on their information
+criteria (Table~\ref{tbl-aa_loo1}). Measured under high light intensity,
+\aax{} was consistently greater for sun plants. The average \aax{} among
+populations in the shade treatment was 0.041 (range: 0.010--0.113; 21 of
+29 populations significant); however, the same populations grown at high
+light intensity showed a mean \aax{} of 0.053 (range: 0.021--0.125; 19
+of 29 populations significant). Contrary to the predictions of the
+assimilatory hypothesis, \aax{} was greater in all populations under low
+measurement light intensity for both sun and shade grown plants. The
+effect of low light on \aax{} was more pronounced in the sun-grown
+plants, where \aax{} was significantly greater under low measurement
+light intensity in 11 of 29 populations compared to to 4 populations for
+shade-grown plants. The overall average \aax{} of shade and sun grown
+plants measured under low light intensity was 0.064 (range:
+0.017--0.132; 28 of 29 populations significant) and 0.094 (range:
+0.036--0.197; 24 of 29 populations significant), respectively. There was
+a slight tendency for populations from more closed habitats (greater
+PAI), but \aax{} varied widely among populations from open habitats (low
+PAI) regardless of growth and measurement light intensities
+(Fig.~\ref{fig-aa}B).
 
@@
 
+maybe this wording is better: Amphistomy likely plays a significant, but
+under appreciated, role in the carbon and water balance of plant groups
+where this trait is common, such as herbaceous plants in sunny habitats
+and woody species in arid biomes (cites). If our estimates of AA from
+the wild tomato clades
+
+note: estimates of AA are less than that described by Parkhurst and Mott
+(1990) and also less than that of Xiong and Flexas (2021). What does
+this tell us?
+
 A second major conclusion is that the significant photosynthetic gain of
@@
 
-\subsubsection{\texorpdfstring{Cleaning \aqcurve{}
-curves}{Cleaning  curves}}\label{cleaning-curves-1}
-
-The raw data set consisted of 658 \aqcurve{} curves with an average of
-19.4. Manual curation of a data set this size in a principled,
-consistent manner is not feasible. Therefore, we automated data cleaning
-using custom \emph{R} scripts. Cleaning is divided into two sequential
-steps (\autoref{tbl-cleaning2}).
-
-\begin{table}
-
-\caption{\label{tbl-cleaning2}Two sequential steps for cleaning
-\agcurve{} curves. The rationale and procedure for each step are
-described in the text. The rightmost columns summarize the number of
-curves and mean number of points per curve remaining after each step.}
-
-\centering{
-
-\centering
-\begin{tabular}{>{\raggedright\arraybackslash}p{9cm}>{\centering\arraybackslash}p{2.5cm}>{\centering\arraybackslash}p{2.5cm}}
-\toprule
-Step: description & Number of curves & Number of points per curve\\
-\midrule
-\cellcolor{gray!10}{1. remove outliers within each \aqcurve{} curve} & \cellcolor{gray!10}{658} & \cellcolor{gray!10}{19.1}\\
-2. remove \aqcurve{} curves with poor fit & 652 & 19.1\\
-\bottomrule
-\end{tabular}
-
-}
-
-\end{table}%
-
-\paragraph{\texorpdfstring{Remove outliers within each \aqcurve{}
-curve}{Remove outliers within each  curve}}\label{remove-outliers-within-each-curve-1}
-
-\emph{Rationale}: Individual outliers within \agcurve{} curves, usually
-caused by transitory changes in chamber conditions, exert undue leverage
-on parameter estimates and cause bias and/or low precision in parameter
-estimates.
-
-\emph{Procedure}: We fit provisional nonrectangular hyperbola
-(\emph{59}) to each \aqcurve{} curve using nonlinear regression with the
-\texttt{nlsLM()} function from the \emph{R} package \textbf{minpack.lm}
-version 1.2.4 (\emph{60}). We sequentially removed data points with an
-absolute external studentized residual \(> 3\) until none remained.
-
-\paragraph{\texorpdfstring{Remove \aqcurve{} curves with poor
-fit}{Remove  curves with poor fit}}\label{remove-curves-with-poor-fit}
-
-\emph{Rationale}: \aqcurve{} curves with a poor fit to the
-nonrectangular hyperbola most likely indicate systematic measurement
-error and/or the leaf was not fully acclimated to the chamber
-environment.
-
-\emph{Procedure}: As described above, we fit provisional nonrectangular
-hyperbola to each \aqcurve{} curve and calculated the model \(r^2\).
-There was a clear break between typical curves and poorly fitting curves
-where \(r^2 < 0.99\). We therefore removed \aqcurve{} curves with
-\(r^2 < 0.99\).
-
-\subsubsection{\texorpdfstring{Bayesian data analysis in
-\emph{Stan}}{Bayesian data analysis in Stan}}\label{bayesian-data-analysis-in-stan}
-
-We fit five models to test predictions of competing hypotheses about why
-amphistomy advantage (\aax) might be greater for leaves in sunny, open
-habitats. This section provides an overview of differences among models
-(\autoref{tbl-models}). The next sections describe how we fit models in
-\emph{Stan}, all model parameters and priors, and specific predictions
-about parameter values for each hypothesis.
-
-\begin{table}
-
-\caption{\label{tbl-models}Summary of differences among competing models
-of how \aax{} varies with light intensity, light treatment, and among
-populations as a function of native \ppfd. The models are numbered from
-simpler to more complex. All models include fixed effects of light
-intensity and light treatment; some models include interactions between.
-All models include a phylogenetic random effect of population on \aax;
-some models include varying effects of light intensity and light
-treatment among populations. The last column indicates the
-population-level \aax{} variable we used as a response to native \ppfd.}
-
-\centering{
-
-\begin{tabular}{>{\centering\arraybackslash}p{0.3in}>{\raggedright\arraybackslash}p{1.3in}>{\raggedright\arraybackslash}p{3.4in}>{\raggedright\arraybackslash}p{1in}}
-\toprule
-Model & Fixed effects & Phylogenetic random effects & Response to native \ppfd\\
-\midrule
-1 & light intensity & varying intercept among populations & $\mathrm{AA}_{0, \text{acc}}$\\
-
- & light treatment &  & \\
-\cmidrule{1-4}
-2 & light intensity & varying intercept among populations & $\mathrm{AA}_{0, \text{acc}}$\\
-
- & light treatment &  & \\
-
- & intensity $\times$ treatment &  & \\
-\cmidrule{1-4}
-3 & light intensity & varying intercept among populations & $\mathrm{AA}_{2000, \text{acc}}$\\
-
- & light treatment & varying effect of high light intensity among populations & \\
-\cmidrule{1-4}
-4 & light intensity & varying intercept among populations & $\mathrm{AA}_{\text{sun, acc}}$\\
-
- & light treatment & varying effect of sun treatment among populations & \\
-\cmidrule{1-4}
-5 & light intensity & varying intercept among populations & $\mathrm{AA}_{2000, \text{sun, acc}}$\\
-
- & light treatment & varying effect of high light intensity among populations & \\
-
- & intensity $\times$ treatment & varying effect of sun treatment among populations & \\
-\bottomrule
-\end{tabular}
-
-}
-
-\end{table}%
-
-\paragraph{\texorpdfstring{Fitting models in
-\emph{Stan}}{Fitting models in Stan}}\label{sec-fitting}
-
-We fit Bayesian models with MCMC sampling in the probabilistic
-programming language \emph{Stan} (\emph{61}) using the \emph{R} package
-\textbf{brms} version 2.22.0 (\emph{62}). We used \emph{CmdStan} version
-2.35.0 and \textbf{cmdstanr} version 0.9.0 (\emph{63}) to interface with
-\emph{R} version 4.5.0 (\emph{64}). We sampled the posterior
-distribution from 1 chains with 2000 iterations each after 2000 warmup
-iterations per chain. We estimated parameters and confidence intervals
-as the median and 95\% quantile intervals of the posterior,
-respectively. We chose the number of chains, warmup and sampling
-iterations, and maximum treedepth so that parameter estimates converged
-(\(\hat{R} < 1.01\) (\emph{65})) and the effective sample size (ESS) for
-each parameter was \(> 400\).
-
-\paragraph{Parameter estimation and priors}\label{sec-parameters}
-
-There were four levels of parameter estimation in our analysis:
+\subsubsection{\texorpdfstring{Bayesian data analysis with
+\emph{Stan}}{Bayesian data analysis with Stan}}\label{sec-analysis}
+
+Our analysis pipeline consisted of four main steps:
 
@@
 \item
-  Estimate \agcurve{} curve parameters
+  Estimate \agcurve{} curve parameters for each leaf
 \item
-  Estimate \aax{} for each light intensity with leaf using \agcurve{}
-  curve parameters
+  Estimate \aax{} for both measurement light intensities using
+  \agcurve{} curve parameters
 \item
@@
 
-Although the higher-level parameter estimates depend on the lower-level
-parameter estimates, we fit all models simultaneously to ensure that the
-uncertainty in lower-level estimates propagated to higher-level
-estimates.
-
-\begin{longtable}{>{\raggedright\arraybackslash}p{1in}>{\raggedright\arraybackslash}p{5in}}
-
-\caption{\label{tbl-parameters}Description of parameters estimated in
-the hierarchical Bayesian model. The \texttt{Parameter} column lists the
-parameter name as it appears in text. The \texttt{Description} column
-provides a brief description of the parameter.}
-
-\tabularnewline
-
-\toprule
-Parameter & Description\\
-\midrule
-\addlinespace[0.3em]
-\multicolumn{2}{l}{\textbf{\agcurve{} curve parameters}}\\
-\hspace{1em}\cellcolor{gray!10}{$\mathbf{B}_\text{curve}$} & \cellcolor{gray!10}{$n_\text{curve} \times 3$ array of random \agcurve{} curve-level coefficients ($b_{0,j}$, $b_{1,j}$, $b_{2,j}$); $\mathbf{B}_\text{curve} \sim \text{MVN}(\vec{0}, \symbf{\Sigma}_\text{curve})$}\\
-\hspace{1em}$\vec{\beta}_\text{curve}$ & vector of mean quadratic coefficients ($\beta_0$, $\beta_1$, $\beta_2$)\\
-\hspace{1em}\cellcolor{gray!10}{$\symbf{\Sigma}_\text{curve}$} & \cellcolor{gray!10}{$3 \times 3$ covariance matrix of curve-level coefficients}\\
-\hspace{1em}$\sigma_{\qty{6}{\raiseto{2}\centi\meter},\epsilon}$ & minimum residual standard deviation when the measured leaf surface area is $\qty{6}{\raiseto{2}\centi\meter}$\\
-\hspace{1em}\cellcolor{gray!10}{$\beta_{S,\epsilon}$} & \cellcolor{gray!10}{slope of the relationship between residual standard deviation and measured leaf surface area (log-link scale)}\\
-\hspace{1em}$\rho_{\epsilon}$ & lag-1 residual autocorrelation\\
-\addlinespace[0.3em]
-\multicolumn{2}{l}{\textbf{\aax{} for each light intensity with leaf using agcurve{} curve parameters}}\\
-\hspace{1em}\cellcolor{gray!10}{$\widehat{\mathrm{AA}}_{klm}$} & \cellcolor{gray!10}{estimate of \aax{} for the $k^{\text{th}}$ leaf at light intensity $l$ in accession $m$}\\
-\addlinespace[0.3em]
-\multicolumn{2}{l}{\textbf{effects of light intensity, light treatments, and population on \aax}}\\
-\hspace{1em}$\beta_{\mathrm{AA},0}$ & intercept of \aax{} at low light intensity in shade treatment\\
-\hspace{1em}\cellcolor{gray!10}{$\beta_{\mathrm{AA},2000}$} & \cellcolor{gray!10}{effect of high light intensity at \ppfdequals{2000} on \aax}\\
-\hspace{1em}$\beta_{\mathrm{AA},\text{sun}}$ & effect of sun treatment on \aax{}\\
-\hspace{1em}\cellcolor{gray!10}{$\beta_{\mathrm{AA},2000,\text{sun}}$} & \cellcolor{gray!10}{effect of high light intensity at \ppfdequals{2000} on \aax{} in sun treatment}\\
-\hspace{1em}$\vec{\beta}_{\mathrm{AA},\text{acc}}$ & vector of $n_\text{acc}$ phylogenetically structured random accession-level effects on \aax; $\vec{\beta}_{\mathrm{AA},\text{acc}} \sim \text{MVN}(\vec{0}, \symbf{\Sigma}_{\text{AA,acc}})$\\
-\hspace{1em}\cellcolor{gray!10}{$\symbf{\Sigma}_{\text{AA,acc}}$} & \cellcolor{gray!10}{$n_\text{acc} \times n_\text{acc}$ covariance matrix of phylogenetically structured random accession-level effects on \aax;}\\
-\hspace{1em}$\vec{\beta}_{\mathrm{AA},\text{rep}}$ & vector of $n_\text{rep}$ random replicate-level effects on \aax; $\vec{\beta}_{\mathrm{AA},\text{rep}} \sim \text{Normal}(0, \sigma_{\mathrm{AA},\text{rep}})$\\
-\hspace{1em}\cellcolor{gray!10}{$\vec{\beta}_{\mathrm{AA},\text{2000,acc}}$} & \cellcolor{gray!10}{vector of $n_\text{acc}$ phylogenetically structured random accession-level effects of high light intensity at \ppfdequals{2000} on \aax; $\vec{\beta}_{\mathrm{AA},\text{2000,acc}} \sim \text{MVN}(\vec{0}, \symbf{\Sigma}_{\mathrm{AA},\text{2000,acc}})$}\\
-\hspace{1em}$\vec{\beta}_{\mathrm{AA},\text{sun,acc}}$ & vector of $n_\text{acc}$ phylogenetically structured random accession-level effects of sun treatment on \aax; $\vec{\beta}_{\mathrm{AA},\text{sun,acc}} \sim \text{MVN}(\vec{0}, \symbf{\Sigma}_{\mathrm{AA},\text{sun,acc}})$\\
-\hspace{1em}\cellcolor{gray!10}{$\sigma_{\mathrm{AA},\epsilon,0}$} & \cellcolor{gray!10}{intercept of phylogenetically unstructured residual standard deviation of \aax}\\
-\hspace{1em}$\beta_{\mathrm{AA},\epsilon,2000}$ & effect of high light intensity at \ppfdequals{2000} on phylogenetically unstructured residual standard deviation of \aax{} (log-link scale)\\
-\hspace{1em}\cellcolor{gray!10}{$\beta_{\mathrm{AA},\epsilon,\text{sun}}$} & \cellcolor{gray!10}{effect of sun treatment on phylogenetically unstructured residual standard deviation of \aax{} (log-link scale)}\\
-\hspace{1em}$\sigma_{\mathrm{AA},\text{rep}}$ & standard deviation of random replicate-level effects on \aax\\
-\hspace{1em}\cellcolor{gray!10}{$\alpha_{\mathrm{AA},\text{acc}}$} & \cellcolor{gray!10}{decay rate of phylogenetic covariance in random accession-level effects on \aax}\\
-\hspace{1em}$\sigma^2_{\mathrm{AA},\text{acc}}$ & phylogenetic diffusion rate in random accession-level effects on \aax\\
-\hspace{1em}\cellcolor{gray!10}{$\alpha_{\mathrm{AA},\text{2000,acc}}$} & \cellcolor{gray!10}{decay rate of phylogenetic covariance in random accession-level effects of high light intensity at \ppfdequals{2000} on \aax}\\
-\hspace{1em}$\sigma^2_{\mathrm{AA},\text{2000,acc}}$ & phylogenetic diffusion rate in random accession-level effects of high light intensity at \ppfdequals{2000} on \aax\\
-\hspace{1em}\cellcolor{gray!10}{$\alpha_{\mathrm{AA},\text{sun,acc}}$} & \cellcolor{gray!10}{decay rate of phylogenetic covariance in random accession-level effects of sun treatment on \aax}\\
-\hspace{1em}$\sigma^2_{\mathrm{AA},\text{sun,acc}}$ & phylogenetic diffusion rate in random accession-level effects of sun treatment on \aax\\
-\addlinespace[0.3em]
-\multicolumn{2}{l}{\textbf{effects of native light habitat on population-level \aax}}\\
-\hspace{1em}\hspace{1em}\cellcolor{gray!10}{$\beta_{\mathrm{AA},\text{PPFD},0}$} & \cellcolor{gray!10}{intercept of accession-level \aax{} when native \ppfdequals{0}}\\
-\hspace{1em}\hspace{1em}$\beta_{\mathrm{AA},\text{PPFD},1}$ & slope of native \ppfd{} on accession-level \aax{}\\
-\hspace{1em}\cellcolor{gray!10}{$\alpha_{\mathrm{AA},\text{PPFD}}$} & \cellcolor{gray!10}{decay rate of phylogenetic covariance in residuals of model testing effect of native \ppfd{} on accession-level \aax}\\
-\hspace{1em}$\sigma^2_{\mathrm{AA},\text{PPFD}}$ & phylogenetic diffusion rate in residuals of model testing effect of native \ppfd{} on accession-level \aax\\
-\bottomrule
-
-\end{longtable}
-
-\subparagraph{\texorpdfstring{\agcurve{} curve
+All models were using a Bayesian model with HMC sampling in the
+probabilistic programming language \emph{Stan} (\emph{59}) using the
+\emph{R} package \textbf{brms} version 2.22.0 (\emph{60}). We used
+\emph{CmdStan} version 2.36.0 and \textbf{cmdstanr} version 0.9.0
+(\emph{61}) to interface with \emph{R} version 4.5.0 (\emph{62}). We
+sampled the posterior distribution from a single chain with a minimum of
+1000 iterations after 1000 warmup iterations. If necessary, we refit
+models with more iterations to decrease the Gelman-Rubin convergence
+statistic (\(\hat{R}\)) (\emph{63}) to less than
+\(`aa_stats[["max_rhat"]]`\) and the bulk effective sample size (ESS) to
+greater than \(`aa_stats[["min_ess"]]`\) for all model parameters. We
+also increased the the \texttt{adapt\_delta} parameter until there were
+fewer than \(`aa_stats[["n_divergent"]]`\) divergent transitions.
+
+\paragraph{\texorpdfstring{\agcurve{} curve
 parameters}{ curve parameters}}\label{curve-parameters}
 
-We modeled \logA as a quadratic function of \loggsw for each leaf using
-the following equation:
-
-\[\log(A_{ij}) = (\beta_0 + b_{0,j}) + (\beta_1 + b_{1,j}) \log{g_{\text{sw},i}} + (\beta_2 + b_{2,j}) \log{g_{\text{sw},i}}^2 + \epsilon_{i}\]
-
-where \(\beta_0\), \(\beta_1\), and \(\beta_2\) are the average
-intercept, linear, and quadratic coefficients, respectively. We used
-diffuse normal priors with mean \(0\) and standard deviation \(10\) on
-these parameters. We estimated random effects of curve \(j\) on the
-intercept (\(b_{0,j}\)), linear (\(b_{1,j}\)), and quadratic
-(\(b_{2,j}\)) coefficients. We assumed that the \(j \times 3\) array of
-coefficients was multivariate normal a mean vector of \(\vec{0}\) and
-covariance \(\Sigma_\text{curve}\). We used a weakly informative normal
-prior with mean \(0\) and standard deviation \(1\) on the
-log-transformed standard deviations (i.e.~the diagonal of
-\(\Sigma_\text{curve}\)). We used a weakly informative
-\(\mathrm{LJK}(2)\) prior on the correlation matrix. The off diagonal
-elements of \(\Sigma_\text{curve}\) can be calculated from its diagonal
-elements and the correlation matrix.
-
-The residuals \(\epsilon_{i}\) were modeled as a lag-1 autocorrelated
-time-series. We further assumed that the residual standard deviation of
-the \(j^{\text{th}}\) curve (\(\sigma_{\epsilon,j}\)) was inversely
-proportional to the leaf surface area (\(S_j\)) within the chamber:
-
-\[\log (\sigma_{\epsilon,j}) = \log (\sigma_{\qty{6}{\raiseto{2}\centi\meter},\epsilon}) + (6 - S_j) \beta_{S,\epsilon}\]
-where \(\sigma_{\qty{6}{\raiseto{2}\centi\meter},\epsilon}\) is the
-minimum residual standard deviation when the
-\(\qty{6}{\raiseto{2}\centi\meter}\) chamber is completely filled. The
-residual standard deviation increases on log-linear scale by
-\(\beta_{S,\epsilon}\). We used a weakly informative normal prior with
-mean \(-3\) and standard deviation \(5\) on
-\(\log (\sigma_{\qty{6}{\raiseto{2}\centi\meter},\epsilon})\) and a
-weakly informative normal prior with mean \(0\) and standard deviation
-\(1\) on \(\beta_{S,\epsilon}\).
-
-\subparagraph{\texorpdfstring{\aax{} for each light intensity with leaf
+We modeled \logA as a quadratic function of \loggsw for each leaf,
+measured at low and high light intensity, in both amphistomatous
+(untreated) and pseudohypostomatous (propafilm treatment) states.
+
+\paragraph{\texorpdfstring{\aax{} for each light intensity within leaf
 using \agcurve{} curve
-parameters}{ for each light intensity with leaf using  curve parameters}}\label{for-each-light-intensity-with-leaf-using-curve-parameters}
-
-Within the \(k^{\text{th}}\) leaf, we estimated \aax{} for each light
-intensity by integrating the difference in \logA{} between the amphi and
-pseudohypo \agcurve{} curves over the range of \gsw{} values where the
-curves overlap (from \(\min(\log(g_\text{sw}))\) to
-\(\max(\log(g_\text{sw}))\)). The estimate of \aax{} for the
-\(k^{\text{th}}\) leaf at light intensity \(l\) in population \(m\) is:
-
-\[\widehat{\mathrm{AA}}_{klm} = \int_{\text{min(log(}g_\text{sw}))}^{\text{max(log(}g_\text{sw}))} \text{log}\bigg(\frac{\hat{A}_\text{amphi}(x; \theta_{klm,\text{amphi})}}{\hat{A}_\text{hypo}(x; \theta_{klm,\text{hypo})}}\bigg) dx\]
-where:
-
-\[\theta_\text{amphi} \in \{\hat{b}_{0, f(\text{amphi}, k,l,m)}, \hat{b}_{1, f(\text{amphi}, k,l,m)}, \hat{b}_{2, f(\text{amphi}, k,l,m)}\}, \text{and}\]
-
-\[\theta_\text{hypo} \in \{\hat{b}_{0, f(\text{hypo}, k,l,m)}, \hat{b}_{1, f(\text{hypo}, k,l,m)}, \hat{b}_{2, f(\text{hypo}, k,l,m)}\}.\]
-The function \(f : \Theta_1 \rightarrow \Theta_2\) maps the set
-\(\Theta_1\) indexed by leaf type (amphi or pseudohypo), leaf replicate,
-light intensity, and population to set \(\Theta_2\) indexed by
-individual \agcurve{} curve. This mapping is necessary because the
-random effects structure differs between models of \loggsw{} on \logA{}
-and that of models predicting \aax{} described in the next section.
+parameters}{ for each light intensity within leaf using  curve parameters}}\label{for-each-light-intensity-within-leaf-using-curve-parameters}
+
+\subparagraph{MARKER - add figure of
+this}\label{marker---add-figure-of-this}
+
+We estimated \aax{} for each light intensity by integrating the
+difference in \logA{} between the amphi and pseudohypo \agcurve{} curves
+over the range of \gsw{} values where the curves overlap (from
+\(\min(\log(g_\text{sw}))\) to \(\max(\log(g_\text{sw}))\)).
+
+\[\widehat{\mathrm{AA}} = \int_{\text{min(log(}g_\text{sw}))}^{\text{max(log(}g_\text{sw}))} \text{log}\bigg(\frac{\hat{A}_\text{amphi}(x; \theta_{\text{amphi}})}{\hat{A}_\text{hypo}(x; \theta_{\text{hypo}})}\bigg) dx\]
+where \(\theta_\text{amphi}\) and \(\theta_\text{hypo}\) are the
+quadratic parameters of the amphi and pseudohypo \agcurve{} curves,
+respectively. We calculated \(\widehat{\mathrm{AA}}\) from each draw of
+the posterior for the paired amphi and pseudohypo \agcurve{} curves to
+obtain the posterior distribution of \aax{} for each leaf at each light
+intensity. We then summarized the posterior distribution of \aax{} for
+each leaf at each light intensity by the posterior median, which became
+our estimate of \aax{} for subsequent analyses.
+
+\subsubsection{Bayesian phylogenetic mixed effects
+models}\label{bayesian-phylogenetic-mixed-effects-models}
+
+We fit Bayesian mixed effects models with phylogenetically structured
+random effects to test predictions of competing hypotheses about why
+amphistomy advantage (\aax) might be greater for leaves in sunny, open
+habitats. We fit all models combinations of measurement light intensity,
+growth light intensity, and their interaction as both fixed effects and
+random effects among populations. In other words, we test whether there
+were main effects of light intensity treatments and/or whether
+populations varied in their response to light intensity treatments. In
+all models, we included population as a phylogenetically structured
+random effect. We crossed those fixed and random effect structures with
+fixed effects of measurement light intensity and/or growth light
+intensity on the distributional parameter, \(\sigma\), the residual
+variance in \aax{}. In total, we fit 100 models, each with a different
+combination of fixed, random, and distributional effects. We used the
+RAxML whole-transcriptome concatenated phylogeny based on 2,745 100-kb
+genomic windows from (\textbf{pease\_phylogenomics\_2016?}). Two of our
+populations were not in this tree. We used accession LA1044 in place of
+LA3909, two populations of \emph{S. galapagnse}; LA0750 was added as
+sister to LA0716, two closely replated populations of \emph{S.
+pennellii}. The node was placed half-way between the next deepest node.
+
+NEED TO EDIT THIS: The next sections describe how we fit models in
+\emph{Stan}, all model parameters and priors, and specific predictions
+about parameter values for each hypothesis. Each curve was fit For
+initial fits, we sampled the posterior distribution from a single chain
+with 1000 iterations after 1000 warmup iterations. After model selection
+(see below), we refit the selected model with more iterations if
+necessary to decrease the Gelman-Rubin convergence statistic
+(\(\hat{R}\)) (\emph{63}) to less than \(`aa_stats[["max_rhat"]]`\) and
+the bulk effective sample size (ESS) to greater than
+\(`aa_stats[["min_ess"]]`\) for all model parameters.
+
+\paragraph{\texorpdfstring{Fitting models in \emph{Stan} using
+\textbf{brms}
+interface}{Fitting models in Stan using brms interface}}\label{sec-fitting}
+
+We fit Bayesian models with HMC sampling in the probabilistic
+programming language \emph{Stan} (\emph{59}) using the \emph{R} package
+\textbf{brms} version 2.22.0 (\emph{60}). We used \emph{CmdStan} version
+2.36.0 and \textbf{cmdstanr} version 0.9.0 (\emph{61}) to interface with
+\emph{R} version 4.5.0 (\emph{62}). For initial fits, we sampled the
+posterior distribution from a single chain with 1000 iterations after
+1000 warmup iterations. After model selection (see below), we refit the
+selected model with more iterations if necessary to decrease the
+Gelman-Rubin convergence statistic (\(\hat{R}\)) (\emph{63}) to less
+than \(`aa_stats[["max_rhat"]]`\) and the bulk effective sample size
+(ESS) to greater than \(`aa_stats[["min_ess"]]`\) for all model
+parameters.
 
@@
 the fit of models (\autoref{tbl-models}) using the \emph{R} package
-\textbf{loo} version 2.8.0 (\emph{67}) to calculate LOOIC values. Note
+\textbf{loo} version 2.8.0 (\emph{65}) to calculate LOOIC values. Note
 that LOOIC was calculated only from pointwise likelihood values of the
 submodel estimating effects of light intensity, light treatments, and
-population on \aax{} (see Section~\ref{sec-parameters}). We considered
+population on \aax{} (see \textbf{?@sec-parameters}). We considered
 models with two standard errors of the mean lower LOOIC value to be a
@@
   same A with less R
+\item
+  Each A-gs figure with fitted curves, R2, and AA
 \end{itemize}
 
-\newpage{}
-
-\subsubsection{MANUSCRAPS}\label{manuscraps}
-
-(not sure where this goes) * We acclimated the focal leaf to high light
-(\ppfdequals{2000}) and high relative humidity (\rhequals{70}) until
-\(A\) and \gsw{} reach their maximum. After that, we decreased \rh{} to
-\(\approx 10\%\) to induce rapid stomatal closure without biochemical
-down
-
-Problems to bring up in discussion: - all our species were amphi, so
-they may all live in PAI below threshold to favor amphi - other
-microclimate diffs not considered - don't know specific anatomical
-change
-
-Two outstanding areas for further research are 1) the specific
-anatomical changes, such as leaf thickness or mesophyll porosity, that
-increase the diffusion path length for CO\(_2\) in sun leaves, and 2)
-the costs of amphistomy that could explain plasticity in \gmaxratio{} in
-populations from low-light habitats. Resolving the first question will
-require detailed anatomical measurements on a large number of shade and
-sun leaves with concomitant estimates of \aax. The fact that populations
-from low-light habitats are more plastic suggests that costs of
-amphistomy are greatest in these environments. (not finished yet)
-
-Not sure where this goes: We infer this from the fact that blocking gas
-exchange in pseudohypostomatous leaves reduced \(A\) by X-X\% depending
-on the population, light treatment, and light intensity (Table/figure).
-The AA is equivalent to an X-X\% change in total \(g_\text{sc}\) (see SI
-section gs equivalency). But whereas increasing \(g_\text{sc}\) would
-increase water loss as a necessary by-product, amphistomy can increase
-\(A\) without any appreciable affect on transpiration.
-
-from Mott and Oleary: ``However, Jones and Slatyer (5) reported a higher
-mesophyll resistance for CO, entering through the upper stomata than for
-the lower, and the data of Vaclavik (12) appear to support this
-conclusion''
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\setcounter{enumi}{7}
-\tightlist
-\item
-  What it all means
-\end{enumerate}
-
-\begin{itemize}
-\tightlist
-\item
-  Our study reveals an important role for plasticity in explaining why
-  amphi leaves are more common in sunny habitats
-\item
-  Confirms that leaf structure of sun leaves incurs costs that
-  amphistomy can mitigate
-\item
-  It will be useful to consider coordination between leaf anatomy,
-  stomatal density, and ratio in future studies of leaf
-  structure-function and in crops (or something)
-\end{itemize}
-
-I am using this approach to drafting the paper:
-
-https://blogs.nature.com/nyc/2011/08/10/how-to-write-a-paper-one-possible-answer
-
-\subsection{Punchline}\label{punchline}
-
-Plastic changes in the structure of sun leaves can increase the
-intercellular CO\(_2\) diffusion path length. Gas exchange through
-stomata on both upper and lower leaf surfaces, amphistomy, may be a
-compensatory mechanism to reduce the impact of increased diffusion path
-length on photosynthesis without increasing water loss. We demonstrate
-for the first time that sun leaves benefit most from amphistomy because
-it compensates for plastic structural changes in leaf anatomy.
-
-What's surprising: role of developmental plasticity in explaining
-comparative results; diffusional limitations through palisade ias is
-strong and needs to be considered with light gradients in understanding
-leaf strucuture- funciton
-
-big implications - if trees were amphi, more more carbon sink or reduced
-transpiration
-
-It is commonly assumed in comparative studies that most trait variation
-between species is constitutive and genetically determined, but our
-results suggest that plastic responses to light are necessary to explain
-why \gmaxratio{} increases with light habitat. Reaction norms to light
-intensity among wild tomatoes indicate that populations from low light
-habitats express low \gmaxratio{} when grown in shade, whereas
-populations from high light habitats have constitutively higher
-\gmaxratio{} (Fig. rxn-norm). The larger response to light intensity is
-driven by greater plasticty in leaf structure in populations from
-low-light habitats (Fig/table - I haven't tested this yet, so this is a
-placeholder.) We infer that these structural changes in sun leaves are
-costly because they increase the diffusion path length for CO\(_2\) and
-reduce photosynthesis. As a consequence, sun leaves benefit more from
-amphistomy. This is the first demonstration that sun leaves benefit more
-from amphistomy because it compensates for developmental plasticity in
-leaf structure.
+\begin{longtable}[t]{>{\centering\arraybackslash}p{0.7in}>{\centering\arraybackslash}p{0.5in}>{\centering\arraybackslash}p{0.4in}>{\centering\arraybackslash}p{0.4in}>{}c|>{\centering\arraybackslash}p{0.4in}>{\centering\arraybackslash}p{0.4in}>{}c|>{\centering\arraybackslash}p{0.4in}>{\centering\arraybackslash}p{0.4in}}
+
+\caption{\label{tbl-aa_loo1}Leave-one-out cross-validation information
+criterion (LOOIC) for models of \aax{} as a function of light intensity,
+light treatment, and population. The model with the lowest LOOIC is
+considered the best fit to the data. The LOOIC column lists the LOOIC
+value for each model; the \(\Delta\)LOOIC column lists the difference
+between the LOOIC of the corresponding model and the best model (lowest
+LOOIC). Checkmarks in each row indicate whether the corresponding model
+included affects of measurement light intensity (``Meas.'') or growth
+light intensity (``Growth'') as fixed effects, random effects, or
+distributional effect on the residual variance. We also test for fixed
+and random interaction effects (``Inter.'') between measurement and
+growth light intensity. All models included phylogenetically structured
+random effects of population.}
+
+\tabularnewline
+
+\toprule
+\multicolumn{2}{c}{ } & \multicolumn{3}{c}{Fixed} & \multicolumn{3}{c}{Random} & \multicolumn{2}{c}{Distributional} \\
+\cmidrule(l{3pt}r{3pt}){3-5} \cmidrule(l{3pt}r{3pt}){6-8} \cmidrule(l{3pt}r{3pt}){9-10}
+LOOIC & $\Delta$LOOIC & Meas. & Growth & Inter. & Meas. & Growth & Inter. & Meas. & Growth\\
+\midrule
+\endfirsthead
+\multicolumn{10}{@{}l}{\textit{(continued)}}\\
+\toprule
+LOOIC & $\Delta$LOOIC & Meas. & Growth & Inter. & Meas. & Growth & Inter. & Meas. & Growth\\
+\midrule
+\endhead
+
+\endfoot
+\bottomrule
+\endlastfoot
+\cellcolor{gray!10}{-1869.80} & \cellcolor{gray!10}{0.00} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$}\\
+-1868.94 & 0.86 & $\checkmark$ & $\checkmark$ & $\checkmark$ & $\checkmark$ & $\checkmark$ & $\checkmark$ & $\checkmark$ & $\checkmark$\\
+\cellcolor{gray!10}{-1868.68} & \cellcolor{gray!10}{1.12} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$}\\
+-1867.89 & 1.92 &  & $\checkmark$ &  & $\checkmark$ & $\checkmark$ & $\checkmark$ & $\checkmark$ & $\checkmark$\\
+\cellcolor{gray!10}{-1867.44} & \cellcolor{gray!10}{2.36} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$}\\
+\addlinespace
+-1867.06 & 2.74 & $\checkmark$ & $\checkmark$ &  & $\checkmark$ & $\checkmark$ & $\checkmark$ & $\checkmark$ & $\checkmark$\\
+\cellcolor{gray!10}{-1867.00} & \cellcolor{gray!10}{2.81} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$}\\
+-1866.86 & 2.94 & $\checkmark$ & $\checkmark$ & $\checkmark$ & $\checkmark$ &  &  & $\checkmark$ & $\checkmark$\\
+\cellcolor{gray!10}{-1866.71} & \cellcolor{gray!10}{3.09} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$}\\
+-1866.36 & 3.44 &  &  &  & $\checkmark$ & $\checkmark$ &  & $\checkmark$ & $\checkmark$\\
+\addlinespace
+\cellcolor{gray!10}{-1866.19} & \cellcolor{gray!10}{3.61} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$}\\
+-1865.08 & 4.72 &  & $\checkmark$ &  & $\checkmark$ &  &  & $\checkmark$ & $\checkmark$\\
+\cellcolor{gray!10}{-1864.70} & \cellcolor{gray!10}{5.10} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$}\\
+-1864.47 & 5.34 & $\checkmark$ & $\checkmark$ & $\checkmark$ &  & $\checkmark$ &  & $\checkmark$ & $\checkmark$\\
+\cellcolor{gray!10}{-1863.52} & \cellcolor{gray!10}{6.28} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$}\\
+\addlinespace
+-1863.27 & 6.54 & $\checkmark$ & $\checkmark$ & $\checkmark$ &  &  &  & $\checkmark$ & $\checkmark$\\
+\cellcolor{gray!10}{-1862.59} & \cellcolor{gray!10}{7.22} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$}\\
+-1862.42 & 7.39 & $\checkmark$ & $\checkmark$ &  &  & $\checkmark$ &  & $\checkmark$ & $\checkmark$\\
+\cellcolor{gray!10}{-1862.06} & \cellcolor{gray!10}{7.74} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$}\\
+-1861.93 & 7.87 & $\checkmark$ &  &  & $\checkmark$ & $\checkmark$ & $\checkmark$ &  & $\checkmark$\\
+\addlinespace
+\cellcolor{gray!10}{-1861.71} & \cellcolor{gray!10}{8.10} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$}\\
+-1861.45 & 8.36 & $\checkmark$ & $\checkmark$ & $\checkmark$ & $\checkmark$ & $\checkmark$ &  &  & $\checkmark$\\
+\cellcolor{gray!10}{-1860.69} & \cellcolor{gray!10}{9.11} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$}\\
+-1860.41 & 9.39 & $\checkmark$ & $\checkmark$ &  & $\checkmark$ & $\checkmark$ & $\checkmark$ &  & $\checkmark$\\
+\cellcolor{gray!10}{-1859.96} & \cellcolor{gray!10}{9.84} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$}\\
+\addlinespace
+-1858.68 & 11.12 & $\checkmark$ & $\checkmark$ & $\checkmark$ & $\checkmark$ &  &  &  & $\checkmark$\\
+\cellcolor{gray!10}{-1858.53} & \cellcolor{gray!10}{11.27} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$}\\
+-1858.18 & 11.62 &  & $\checkmark$ &  & $\checkmark$ & $\checkmark$ &  &  & $\checkmark$\\
+\cellcolor{gray!10}{-1858.10} & \cellcolor{gray!10}{11.70} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$}\\
+-1857.84 & 11.96 & $\checkmark$ & $\checkmark$ &  & $\checkmark$ & $\checkmark$ &  &  & $\checkmark$\\
+\addlinespace
+\cellcolor{gray!10}{-1857.64} & \cellcolor{gray!10}{12.16} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$}\\
+-1857.06 & 12.75 & $\checkmark$ &  &  & $\checkmark$ &  &  & $\checkmark$ & $\checkmark$\\
+\cellcolor{gray!10}{-1856.48} & \cellcolor{gray!10}{13.32} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$}\\
+-1856.40 & 13.40 & $\checkmark$ & $\checkmark$ & $\checkmark$ &  & $\checkmark$ &  &  & $\checkmark$\\
+\cellcolor{gray!10}{-1856.24} & \cellcolor{gray!10}{13.56} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{}\\
+\addlinespace
+-1855.98 & 13.82 &  &  &  & $\checkmark$ &  &  & $\checkmark$ & $\checkmark$\\
+\cellcolor{gray!10}{-1855.89} & \cellcolor{gray!10}{13.92} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{}\\
+-1855.37 & 14.43 & $\checkmark$ & $\checkmark$ & $\checkmark$ & $\checkmark$ & $\checkmark$ & $\checkmark$ & $\checkmark$ & \\
+\cellcolor{gray!10}{-1855.36} & \cellcolor{gray!10}{14.44} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$}\\
+-1855.05 & 14.75 & $\checkmark$ & $\checkmark$ &  & $\checkmark$ & $\checkmark$ & $\checkmark$ & $\checkmark$ & \\
+\addlinespace
+\cellcolor{gray!10}{-1854.54} & \cellcolor{gray!10}{15.26} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$}\\
+-1854.41 & 15.39 &  & $\checkmark$ &  & $\checkmark$ & $\checkmark$ & $\checkmark$ & $\checkmark$ & \\
+\cellcolor{gray!10}{-1853.87} & \cellcolor{gray!10}{15.94} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$}\\
+-1853.55 & 16.25 & $\checkmark$ &  &  &  &  &  & $\checkmark$ & $\checkmark$\\
+\cellcolor{gray!10}{-1853.43} & \cellcolor{gray!10}{16.37} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{}\\
+\addlinespace
+-1852.90 & 16.91 & $\checkmark$ &  &  & $\checkmark$ & $\checkmark$ &  & $\checkmark$ & \\
+\cellcolor{gray!10}{-1851.37} & \cellcolor{gray!10}{18.43} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{}\\
+-1851.18 & 18.62 &  &  &  & $\checkmark$ & $\checkmark$ &  & $\checkmark$ & \\
+\cellcolor{gray!10}{-1851.03} & \cellcolor{gray!10}{18.77} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{}\\
+-1851.03 & 18.77 &  & $\checkmark$ &  & $\checkmark$ &  &  & $\checkmark$ & \\
+\addlinespace
+\cellcolor{gray!10}{-1850.89} & \cellcolor{gray!10}{18.91} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{}\\
+-1850.69 & 19.12 & $\checkmark$ & $\checkmark$ &  & $\checkmark$ & $\checkmark$ &  & $\checkmark$ & \\
+\cellcolor{gray!10}{-1850.37} & \cellcolor{gray!10}{19.43} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{}\\
+-1850.04 & 19.76 & $\checkmark$ & $\checkmark$ &  & $\checkmark$ &  &  & $\checkmark$ & \\
+\cellcolor{gray!10}{-1849.45} & \cellcolor{gray!10}{20.35} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{}\\
+\addlinespace
+-1848.68 & 21.12 & $\checkmark$ &  &  &  & $\checkmark$ &  & $\checkmark$ & \\
+\cellcolor{gray!10}{-1848.64} & \cellcolor{gray!10}{21.16} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{}\\
+-1847.73 & 22.07 & $\checkmark$ & $\checkmark$ & $\checkmark$ &  &  &  & $\checkmark$ & \\
+\cellcolor{gray!10}{-1847.72} & \cellcolor{gray!10}{22.08} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{}\\
+-1847.47 & 22.33 & $\checkmark$ & $\checkmark$ &  & $\checkmark$ & $\checkmark$ & $\checkmark$ &  & \\
+\addlinespace
+\cellcolor{gray!10}{-1847.40} & \cellcolor{gray!10}{22.40} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$}\\
+-1847.36 & 22.44 & $\checkmark$ &  &  & $\checkmark$ &  &  &  & $\checkmark$\\
+\cellcolor{gray!10}{-1847.14} & \cellcolor{gray!10}{22.66} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{}\\
+-1846.28 & 23.52 & $\checkmark$ & $\checkmark$ & $\checkmark$ & $\checkmark$ & $\checkmark$ &  &  & \\
+\cellcolor{gray!10}{-1845.82} & \cellcolor{gray!10}{23.98} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{}\\
+\addlinespace
+-1845.58 & 24.22 & $\checkmark$ & $\checkmark$ & $\checkmark$ & $\checkmark$ &  &  &  & \\
+\cellcolor{gray!10}{-1845.55} & \cellcolor{gray!10}{24.26} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{}\\
+-1845.31 & 24.49 & $\checkmark$ & $\checkmark$ &  &  &  &  & $\checkmark$ & \\
+\cellcolor{gray!10}{-1845.29} & \cellcolor{gray!10}{24.51} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$}\\
+-1845.28 & 24.52 & $\checkmark$ &  &  & $\checkmark$ & $\checkmark$ &  &  & \\
+\addlinespace
+\cellcolor{gray!10}{-1844.09} & \cellcolor{gray!10}{25.71} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{}\\
+-1844.00 & 25.80 & $\checkmark$ & $\checkmark$ &  & $\checkmark$ & $\checkmark$ &  &  & \\
+\cellcolor{gray!10}{-1843.72} & \cellcolor{gray!10}{26.08} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{}\\
+-1841.91 & 27.89 & $\checkmark$ & $\checkmark$ & $\checkmark$ &  & $\checkmark$ &  &  & \\
+\cellcolor{gray!10}{-1841.65} & \cellcolor{gray!10}{28.16} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{}\\
+\addlinespace
+-1841.19 & 28.61 &  &  &  & $\checkmark$ &  &  & $\checkmark$ & \\
+\cellcolor{gray!10}{-1840.69} & \cellcolor{gray!10}{29.12} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{}\\
+-1839.82 & 29.98 & $\checkmark$ & $\checkmark$ &  &  &  &  &  & \\
+\cellcolor{gray!10}{-1839.81} & \cellcolor{gray!10}{29.99} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{}\\
+-1839.04 & 30.76 & $\checkmark$ & $\checkmark$ &  &  & $\checkmark$ &  &  & \\
+\addlinespace
+\cellcolor{gray!10}{-1838.36} & \cellcolor{gray!10}{31.44} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{}\\
+-1834.02 & 35.78 &  &  &  &  & $\checkmark$ &  & $\checkmark$ & $\checkmark$\\
+\cellcolor{gray!10}{-1833.87} & \cellcolor{gray!10}{35.93} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{}\\
+-1833.86 & 35.94 & $\checkmark$ &  &  & $\checkmark$ &  &  &  & \\
+\cellcolor{gray!10}{-1833.29} & \cellcolor{gray!10}{36.51} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$}\\
+\addlinespace
+-1832.82 & 36.99 &  & $\checkmark$ &  &  & $\checkmark$ &  & $\checkmark$ & $\checkmark$\\
+\cellcolor{gray!10}{-1829.41} & \cellcolor{gray!10}{40.39} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{}\\
+-1826.86 & 42.94 &  &  &  &  & $\checkmark$ &  &  & $\checkmark$\\
+\cellcolor{gray!10}{-1826.67} & \cellcolor{gray!10}{43.13} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$}\\
+-1825.19 & 44.61 &  & $\checkmark$ &  &  & $\checkmark$ &  &  & $\checkmark$\\
+\addlinespace
+\cellcolor{gray!10}{-1824.95} & \cellcolor{gray!10}{44.86} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$}\\
+-1818.14 & 51.66 &  &  &  &  &  &  &  & $\checkmark$\\
+\cellcolor{gray!10}{-1814.64} & \cellcolor{gray!10}{55.16} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{}\\
+-1814.51 & 55.29 &  &  &  &  & $\checkmark$ &  & $\checkmark$ & \\
+\cellcolor{gray!10}{-1814.23} & \cellcolor{gray!10}{55.57} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{}\\
+\addlinespace
+-1808.39 & 61.41 &  &  &  &  &  &  & $\checkmark$ & \\
+\cellcolor{gray!10}{-1808.13} & \cellcolor{gray!10}{61.67} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{}\\
+-1807.22 & 62.58 &  & $\checkmark$ &  &  & $\checkmark$ &  &  & \\
+\cellcolor{gray!10}{-1806.92} & \cellcolor{gray!10}{62.88} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{}\\
+-1799.31 & 70.49 &  &  &  &  &  &  &  & \\*
+
+\end{longtable}
 

### 2025-07-07 r/205_select-model2.R

@@ -31,39 +31,75 @@ random_effects = c(
   "(1 + light_treatment * light_intensity | gr(acc, cov = A))"
 )
 
-sigma_effects = c(
-  "1",
-  "light_treatment",
-  "light_intensity",
-  "light_treatment + light_intensity"
-)
+sigma_effects = c("1",
+                  "light_treatment",
+                  "light_intensity",
+                  "light_treatment + light_intensity")
 
 # Cross all combinations
 set.seed(602275201)
-model_forms = expand.grid(fixed = fixed_effects,
-                          random = random_effects,
-                          sigma = sigma_effects,
-                          stringsAsFactors = FALSE) %>%
-  mutate(seed = sample(1e9, nrow(.)))
+model_forms = expand.grid(
+  fixed = fixed_effects,
+  random = random_effects,
+  sigma = sigma_effects,
+  stringsAsFactors = FALSE
+) %>%
+  mutate(seed = sample(1e9, nrow(.)),
+         model = paste0("model_", row_number()))
 
 # Build and fit each model
 plan(multisession, workers = 19)
 
-aa_models = future_pmap(model_forms, function(fixed, random, sigma, seed) {
-  fml = bf(as.formula(paste("aa ~", fixed, "+", random)), as.formula(paste("sigma ~", sigma)))
-  brm(
-    formula = fml + bf(llma |
-                         mi() ~ light_treatment + (1 + light_treatment |
-                                                     gr(acc, cov = A))),
-    data = d1,
-    data2 = list(A = A),
-    backend = "cmdstanr",
-    chains = 1,
-    family = student(),
-    save_pars = save_pars(all = TRUE),
-    seed = seed
-  )
-}, .progress = TRUE, .options = furrr_options(seed = TRUE))
+aa_models = model_forms |>
+  dplyr::select(fixed, random, sigma, seed) |>
+  future_pmap(function(fixed, random, sigma, seed) {
+    fml = bf(as.formula(paste(
+      "aa | se(se_aa, sigma = TRUE) ~", fixed, "+", random
+    )), as.formula(paste("sigma ~", sigma)))
+    
+    crit = 0
+    x = 1
+    ad = 0.8
+    
+    while (crit == 0 & x < 24) {
+      m = brm(
+        formula = fml + bf(llma |
+                             mi() ~ light_treatment + (1 + light_treatment |
+                                                         gr(acc, cov = A))),
[…]

### 2025-07-07 r/207_plot-aa.R

@@ -7,7 +7,8 @@ d1 = fit_aa$data
 df_new = crossing(
   acc = unique(d1$acc),
   light_treatment = c("low", "high"),
-  light_intensity = c("150", "2000")
+  light_intensity = c("150", "2000"),
+  se_aa = 0
 ) |>
   mutate(row = row_number())
 
@@ -40,7 +41,8 @@ df_aa_pred1 = posterior_epred(fit_aa, newdata = df_new) |>
 df_new = crossing(
   acc = first(d1$acc),
   light_treatment = c("low", "high"),
-  light_intensity = c("150", "2000")
+  light_intensity = c("150", "2000"),
+  se_aa = 0
 ) |>
   mutate(row = row_number())
 
@@ -62,7 +64,8 @@ df_aa_pred2 = posterior_epred(fit_aa, newdata = df_new) |>
 df_new = crossing(
   acc = unique(d1$acc),
   light_treatment = c("low", "high"),
-  light_intensity = c("150", "2000")
+  light_intensity = c("150", "2000"),
+  se_aa = 0
 ) |>
   mutate(row = row_number())
 
@@ -132,112 +135,3 @@ ggsave(
 write_rds(df_aa_pred1, "objects/df_aa_pred1.rds")
 write_rds(df_aa_pred2, "objects/df_aa_pred2.rds")
 write_rds(df_aa_pred3, "objects/df_aa_pred3.rds")
-
-# code I am not sure I need
-df_pred2 = posterior_epred(fit_aa, newdata = df_new) |>
-  t() |>
-  as_draws_df() |>
-  mutate(row = row_number()) |>
-  pivot_longer(cols = -row, names_to = "draw", values_to = "aa") |>
-  full_join(df_new, by = "row") |>
-  dplyr::select(-row) |>
-  pivot_wider(names_from = light_treatment, values_from = aa) |>
-  mutate(d_aa = high - low) |>
-  group_by(acc, light_intensity) |>
-  point_interval(d_aa) |>   
-  group_by(light_intensity) |>
-  arrange(d_aa) |>
-  mutate(x = order(d_aa))
-
-ggplot(df_pred2, aes(x, d_aa, ymin = .lower, ymax = .upper, group = acc)) +
-  facet_grid(light_intensity ~ .) +
-  geom_pointinterval() +
-  geom_hline(yintercept = 0, linetype = "dashed") 
-
-df_pred2 |>
-  ungroup() |>
-  dplyr::select(acc, light_intensity, d_aa) |>
[…]

### 2025-07-09 ms/manuscraps.qmd

@@
 We therefore removed \aqcurve{} curves with $r^2 < `r aq_r2_threshold`$.
+
+## table describing 5 original competing models
+
+```{r, models, echo=FALSE,message=FALSE,warning=FALSE}
+#| label: tbl-models
+#| tbl-cap: "Summary of differences among competing models of how \\aax{} varies with light intensity, light treatment, and among populations as a function of native \\ppfd.
+The models are numbered from simpler to more complex.
+All models include fixed effects of light intensity and light treatment; some models include interactions between.
+All models include a phylogenetic random effect of population on \\aax; some models include varying effects of light intensity and light treatment among populations.
+The last column indicates the population-level \\aax{} variable we used as a response to native \\ppfd."
+
+tribble(
+    ~"Model", ~"Fixed effects", ~"Phylogenetic random effects", ~"Response to native \\ppfd",
+    "1", "light intensity", "varying intercept among populations", "$\\mathrm{AA}_{0, \\text{acc}}$",
+    "1", "light treatment", "", "",
+    "2", "light intensity", "varying intercept among populations", "$\\mathrm{AA}_{0, \\text{acc}}$",
+    "2", "light treatment", "", "",
+    "2", "intensity $\\times$ treatment", "", "",
+    "3", "light intensity", "varying intercept among populations", "$\\mathrm{AA}_{2000, \\text{acc}}$",
+    "3", "light treatment", "varying effect of high light intensity among populations", "",
+    "4", "light intensity", "varying intercept among populations", "$\\mathrm{AA}_{\\text{sun, acc}}$",
+    "4", "light treatment", "varying effect of sun treatment among populations", "",
+    "5", "light intensity", "varying intercept among populations", "$\\mathrm{AA}_{2000, \\text{sun, acc}}$",
+    "5", "light treatment", "varying effect of high light intensity among populations", "",
+    "5", "intensity $\\times$ treatment", "varying effect of sun treatment among populations", ""
+  ) |>
+  knitr::kable(
+    format = "latex",
+    align = c("c", "l", "l", "l"),
+    booktabs = TRUE,
+    escape = FALSE,
+    longtable = FALSE
+  ) |>
+    kableExtra::collapse_rows(columns = 1, latex_hline = "major", row_group_label_position = "first") |>
+  kableExtra::column_spec(1, width = "0.3in") |>
+  kableExtra::column_spec(2, width = "1.3in") |>
+  kableExtra::column_spec(3, width = "3.4in") |>
+  kableExtra::column_spec(4, width = "1in")
+
+```
+
+## table of all parameters
+
+```{r, parameters, echo=FALSE,message=FALSE,warning=FALSE}
+#| label: tbl-parameters
+#| tbl-cap: "Description of parameters estimated in the hierarchical Bayesian model.
+The `Parameter` column lists the parameter name as it appears in text.
+The `Description` column provides a brief description of the parameter."
+
+df_parameters = read_csv("../data/parameters1.csv", col_types = "ccccl") |>
+    dplyr::filter(include)
+
+df_parameters |>
+  dplyr::select(Parameter = latex, Description = description) |>
+  knitr::kable(
+    format = "latex",
+    align = c("l", "l"),
+    booktabs = TRUE,
+    escape = FALSE,
+    longtable = TRUE
+  ) |>
+  kableExtra::kable_styling(latex_options = c("striped", "no_vertical_lines")) |>
+  kableExtra::column_spec(1, width = "1in") |>
+  kableExtra::column_spec(2, width = "5in") |>
+  kableExtra::pack_rows("\\agcurve{} curve parameters",
+                        min(which(df_parameters$par_type == "ags")),
+                        max(which(df_parameters$par_type == "ags")),
+                        escape = FALSE) |>
+  kableExtra::pack_rows("\\aax{} for each light intensity with leaf using \\agcurve{} curve parameters",
+                        min(which(df_parameters$par_type == "aa_est")),
+                        max(which(df_parameters$par_type == "aa_est")),
+                        escape = FALSE) |>
+  kableExtra::pack_rows("effects of light intensity, light treatments, and population on \\aax",
+                        min(which(df_parameters$par_type == "aa")),
+                        max(which(df_parameters$par_type == "aa")),
+                        escape = FALSE) |>
+  kableExtra::pack_rows("effects of native light habitat on population-level \\aax",
+                        min(which(df_parameters$par_type == "ppfd_aa")),
+                        max(which(df_parameters$par_type == "ppfd_aa")),
+                        escape = FALSE)
+
+```
+
+# This was the old description based on custom model
+
+#### \agcurve{} curve parameters
+
+We modeled \logA as a quadratic function of \loggsw for each leaf using the following equation:
+
+$$\log(A_{ij}) = (\beta_0 + b_{0,j}) + (\beta_1 + b_{1,j}) \log{g_{\text{sw},i}} + (\beta_2 + b_{2,j}) \log{g_{\text{sw},i}}^2 + \epsilon_{i}$$
+
+where $\beta_0$, $\beta_1$, and $\beta_2$ are the average intercept, linear, and quadratic coefficients, respectively.
+We used diffuse normal priors with mean $0$ and standard deviation $10$ on these parameters.
+We estimated random effects of curve $j$ on the intercept ($b_{0,j}$), linear ($b_{1,j}$), and quadratic ($b_{2,j}$) coefficients.
+We assumed that the $j \times 3$ array of coefficients was multivariate normal a mean vector of $\vec{0}$ and covariance $\Sigma_\text{curve}$.
+We used a weakly informative normal prior with mean $0$ and standard deviation $1$ on the log-transformed standard deviations (i.e.
+the diagonal of $\Sigma_\text{curve}$).
+We used a weakly informative $\mathrm{LJK}(2)$ prior on the correlation matrix.
+The off diagonal elements of $\Sigma_\text{curve}$ can be calculated from its diagonal elements and the correlation matrix.
+
+The residuals $\epsilon_{i}$ were modeled as a lag-1 autocorrelated time-series.
+We further assumed that the residual standard deviation of the $j^{\text{th}}$ curve ($\sigma_{\epsilon,j}$) was inversely proportional to the leaf surface area ($S_j$) within the chamber:
+
+$$\log (\sigma_{\epsilon,j}) = \log (\sigma_{\qty{6}{\raiseto{2}\centi\meter},\epsilon}) + (6 - S_j) \beta_{S,\epsilon}$$
+where $\sigma_{\qty{6}{\raiseto{2}\centi\meter},\epsilon}$ is the minimum residual standard deviation when the $\qty{6}{\raiseto{2}\centi\meter}$ chamber is completely filled.
+The residual standard deviation increases on log-linear scale by $\beta_{S,\epsilon}$.
+We used a weakly informative normal prior with mean $-3$ and standard deviation $5$ on $\log (\sigma_{\qty{6}{\raiseto{2}\centi\meter},\epsilon})$ and a weakly informative normal prior with mean $0$ and standard deviation $1$ on $\beta_{S,\epsilon}$.
+
+#### \aax{} for each light intensity within leaf using \agcurve{} curve parameters
+
+We estimated \aax{} for each light intensity by integrating the difference in \logA{} between the amphi and pseudohypo \agcurve{} curves over the range of \gsw{} values where the curves overlap (from $\min(\log(g_\text{sw}))$ to $\max(\log(g_\text{sw}))$).
+The estimate of \aax{} for the $k^{\text{th}}$ leaf at light intensity $l$ in population $m$ is:
+
+$$\widehat{\mathrm{AA}}_{klm} = \int_{\text{min(log(}g_\text{sw}))}^{\text{max(log(}g_\text{sw}))} \text{log}\bigg(\frac{\hat{A}_\text{amphi}(x; \theta_{klm,\text{amphi})}}{\hat{A}_\text{hypo}(x; \theta_{klm,\text{hypo})}}\bigg) dx$$
+where:
+
+$$\theta_\text{amphi} \in \{\hat{b}_{0, f(\text{amphi}, k,l,m)}, \hat{b}_{1, f(\text{amphi}, k,l,m)}, \hat{b}_{2, f(\text{amphi}, k,l,m)}\}, \text{and}$$
+
+$$\theta_\text{hypo} \in \{\hat{b}_{0, f(\text{hypo}, k,l,m)}, \hat{b}_{1, f(\text{hypo}, k,l,m)}, \hat{b}_{2, f(\text{hypo}, k,l,m)}\}.$$ 
+The function $f : \Theta_1 \rightarrow \Theta_2$ maps the set $\Theta_1$ indexed by leaf type (amphi or pseudohypo), leaf replicate, light intensity, and population to set $\Theta_2$ indexed by individual \agcurve{} curve.
+This mapping is necessary because the random effects structure differs between models of \loggsw{} on \logA{} and that of models predicting \aax{} described in the next section.

### 2025-07-20 ms/manuscraps.qmd
example-aa figure

@@
 This mapping is necessary because the random effects structure differs between models of \loggsw{} on \logA{} and that of models predicting \aax{} described in the next section.
+
+##### Effects of light intensity, light treatment, and population on \aax
+
+We tested for effects of light intensity, light treatment, population, and their interactions on \aax.
+All models included effects of light intensity and light treatment, as well as random effects of population and replicate within population.
+More complex models included interactions between light intensity and light treatment, as well as random effects of population on the effects of light intensity and light treatment.
+We used a weakly informative normal prior with mean $0$ and standard deviation $10$ for fixed effects of light intensity and treatment.
+We used a weakly informative normal prior with mean $-3$ and standard deviation $5$ for the the random effect standard deviation of replicate within population.
+We accounted for the phylogenetic structure among the random effects of population using an Ornstein-Uhlenbeck (OU) process [@hansen_stabilizing_1997].
+The expected covariance between populations $i$ and $j$ ($\text{Cov}(i, j)$ ) is:
+
+$$\text{Cov}(i, j) = \frac{\sigma^2}{2 \alpha} \exp(-\alpha D_{ij})$$
+where $\sigma^2$ is the variance of the random effect, $\alpha$ is the rate of decay of the covariance with phylogenetic distance, $D_{ij}$, the phylogenetic distance between populations $i$ and $j$.
+We estimated $\sigma^2 / (2 \alpha)$ and $\alpha$ as a separate parameters and reparameterized them as $\sigma^2$ and $\alpha$.
+We used a weakly informative normal prior with mean $0$ and standard deviation $10$ on OU parameters.
+
+We modeled the residual standard deviation of \aax, (i.e.
+phylogenetically unstructured variation unaccounted for by explanatory variables) on a log-link scale with effects of light intensity and light treatment.
+We used a weakly informative normal prior with mean $-3$ and standard deviation $5$ on the residual standard deviation intercept and a weakly informative normal prior with mean $0$ and standard deviation $1$ on the effects of light intensity and light treatment on the residual standard deviation.
+
+##### Effects of native light habitat on population-level \aax
+
+We tested a linear effect of native light habitat on population-level \aax.
+In Models 1 and 2, we used the random intercept of \aax{} as a response variable; in model 3 we used the random intercept plus the random effect of population at high light intensity; in model 4 we used the random intercept plus the random effect of population in the sun treatment; in model 5 we used the random intercept plus the random effects of population at high light intensity and sun treatment.
+We used a weakly informative normal prior with mean $0$ and standard deviation $1$ for the slope and intercept.
+We accounted for the phylogenetic structure among the model residuals using an (OU) process.
+We used a weakly informative normal prior with mean $0$ and standard deviation $10$ on OU parameters.
+
+## note on LOOIC
+
+We considered models with two standard errors of the mean lower LOOIC value to be a better fit to the data; we considered models with LOOIC values within two standard errors of the mean to be have similar support.

### 2025-07-20 ms/ms.qmd
example-aa figure

@@
 
-#### \agcurve{} curve parameters
+#### \agcurve{} curve parameters {#sec-agcurve-pars}
 
@@
 
-#### \aax{} for each light intensity within leaf using \agcurve{} curve parameters
-##### MARKER - add figure of this
+#### \aax{} for each light intensity within leaf using \agcurve{} curve parameters {#sec-aa-est}
+
 We estimated \aax{} for each light intensity by integrating the difference in \logA{} between the amphi and pseudohypo \agcurve{} curves over the range of \gsw{} values where the curves overlap (from $\min(\log(g_\text{sw}))$ to $\max(\log(g_\text{sw}))$).
@@
 $$\widehat{\mathrm{AA}} = \int_{\text{min(log(}g_\text{sw}))}^{\text{max(log(}g_\text{sw}))} \text{log}\bigg(\frac{\hat{A}_\text{amphi}(x; \theta_{\text{amphi}})}{\hat{A}_\text{hypo}(x; \theta_{\text{hypo}})}\bigg) dx$$
-where $\theta_\text{amphi}$ and $\theta_\text{hypo}$ are the quadratic parameters of the amphi and pseudohypo \agcurve{} curves, respectively.
+where $\theta_\text{amphi}$ and $\theta_\text{hypo}$ are the quadratic parameters of the amphi and pseudohypo \agcurve{} curves, respectively (@fig-example-aa).
 We calculated $\widehat{\mathrm{AA}}$ from each draw of the posterior for the paired amphi and pseudohypo \agcurve{} curves to obtain the posterior distribution of \aax{} for each leaf at each light intensity.
-We then summarized the posterior distribution of \aax{} for each leaf at each light intensity by the posterior median, which became our estimate of \aax{} for subsequent analyses.
-
-### Bayesian phylogenetic mixed effects models
+We then summarized the posterior distribution of \aax{} for each leaf at each light intensity by the posterior median and standard deviation, which became our point estimate of, and uncertainty in, \aax{} for the phylogenetic mixed effects model described below.
+
+```{r, example-aa, echo=FALSE,message=FALSE,warning=FALSE}
+#| fig-align: center
+#| fig-width: 6.5
+#| fig-height: 3.5
+#| fig-cap: "Example amphistomy advantage (\\aax) calculation from \\agcurve{} curves in an amphistomatous leaf with gas exchange through both surfaces (blue) and the same leaf with gas exchange blocked on the upper surface only (orange).
+We integrate over the gray region between the curves to estimate \\aax{}.
+The \\aax{} is low (left facet) when the photosynthetic rate of an amphistomatous leaf is similar to a pseudohypostomatous leaf at the same total stomatal conductance ($x$-axes); large \\aax{} (right facet) is indicated when an amphistomatous leaf has a higher photosynthetic rate than a pseudohypostomatous leaf."
+read_rds("../figures/example-aa.rds")
+```
+
+#### Phylogenetic mixed effects models {#sec-phylo}
 
@@
 In all models, we included population as a phylogenetically structured random effect.
-We crossed those fixed and random effect structures with fixed effects of measurement light intensity and/or growth light intensity on the distributional parameter, $\sigma$, the residual variance in \aax{}.
+We accounted for uncertainty in estimates of \aax{} using the standard deviation of the posterior distribution of each \aax{} estimate (@sec-aa-est).
+We crossed fixed and random effect structures with fixed effects of measurement light intensity and/or growth light intensity on the distributional parameter, $\sigma$, the residual variance in \aax{}.
 In total, we fit `r n_model1` models, each with a different combination of fixed, random, and distributional effects.
@@
 We used accession LA1044 in place of LA3909, two populations of *S.
-galapagnse*; LA0750 was added as sister to LA0716, two closely replated populations of *S.
+galapagnse*; LA0750 was added as sister to LA0716, two closely related populations of *S.
 pennellii*.
-The node was placed half-way between the next deepest node.
-
-NEED TO EDIT THIS: The next sections describe how we fit models in *Stan*, all model parameters and priors, and specific predictions about parameter values for each hypothesis.
-Each curve was fit For initial fits, we sampled the posterior distribution from a single chain with `r aa_args$n_iter_init / 2` iterations after `r aa_args$n_iter_init / 2` warmup iterations.
-After model selection (see below), we refit the selected model with more iterations if necessary to decrease the Gelman-Rubin convergence statistic ($\hat{R}$) [@gelman_inference_1992] to less than $`aa_stats[["max_rhat"]]`$ and the bulk effective sample size (ESS) to greater than $`aa_stats[["min_ess"]]`$ for all model parameters.
-
-#### Fitting models in *Stan* using **brms** interface {#sec-fitting}
-
-We fit Bayesian models with HMC sampling in the probabilistic programming language *Stan* [@stan_development_team_stan_2025] using the *R* package **brms** version `r fit_aa1$version$brms` [@burkner_brms_2017].
-We used *CmdStan* version `r fit_aa1$version$cmdstan` and **cmdstanr** version `r fit_aa1$version$cmdstanr` [@gabry_cmdstanr_2025] to interface with *R* version `r getRversion()` [@r_core_team_r:_2025].
-For initial fits, we sampled the posterior distribution from a single chain with `r aa_args$n_iter_init / 2` iterations after `r aa_args$n_iter_init / 2` warmup iterations.
-After model selection (see below), we refit the selected model with more iterations if necessary to decrease the Gelman-Rubin convergence statistic ($\hat{R}$) [@gelman_inference_1992] to less than $`aa_stats[["max_rhat"]]`$ and the bulk effective sample size (ESS) to greater than $`aa_stats[["min_ess"]]`$ for all model parameters.
-
-##### Effects of light intensity, light treatment, and population on \aax
-
-We tested for effects of light intensity, light treatment, population, and their interactions on \aax.
-All models included effects of light intensity and light treatment, as well as random effects of population and replicate within population.
-More complex models included interactions between light intensity and light treatment, as well as random effects of population on the effects of light intensity and light treatment.
-We used a weakly informative normal prior with mean $0$ and standard deviation $10$ for fixed effects of light intensity and treatment.
-We used a weakly informative normal prior with mean $-3$ and standard deviation $5$ for the the random effect standard deviation of replicate within population.
-We accounted for the phylogenetic structure among the random effects of population using an Ornstein-Uhlenbeck (OU) process [@hansen_stabilizing_1997].
-The expected covariance between populations $i$ and $j$ ($\text{Cov}(i, j)$ ) is:
-
-$$\text{Cov}(i, j) = \frac{\sigma^2}{2 \alpha} \exp(-\alpha D_{ij})$$
-where $\sigma^2$ is the variance of the random effect, $\alpha$ is the rate of decay of the covariance with phylogenetic distance, $D_{ij}$, the phylogenetic distance between populations $i$ and $j$.
-We estimated $\sigma^2 / (2 \alpha)$ and $\alpha$ as a separate parameters and reparameterized them as $\sigma^2$ and $\alpha$.
-We used a weakly informative normal prior with mean $0$ and standard deviation $10$ on OU parameters.
-
-We modeled the residual standard deviation of \aax, (i.e.
-phylogenetically unstructured variation unaccounted for by explanatory variables) on a log-link scale with effects of light intensity and light treatment.
-We used a weakly informative normal prior with mean $-3$ and standard deviation $5$ on the residual standard deviation intercept and a weakly informative normal prior with mean $0$ and standard deviation $1$ on the effects of light intensity and light treatment on the residual standard deviation.
-
-##### Effects of native light habitat on population-level \aax
-
-We tested a linear effect of native light habitat on population-level \aax.
-In Models 1 and 2, we used the random intercept of \aax{} as a response variable; in model 3 we used the random intercept plus the random effect of population at high light intensity; in model 4 we used the random intercept plus the random effect of population in the sun treatment; in model 5 we used the random intercept plus the random effects of population at high light intensity and sun treatment.
-We used a weakly informative normal prior with mean $0$ and standard deviation $1$ for the slope and intercept.
-We accounted for the phylogenetic structure among the model residuals using an (OU) process.
-We used a weakly informative normal prior with mean $0$ and standard deviation $10$ on OU parameters.
-
-#### Predictions {#sec-predictions}
+The node separating LA0750 from LA0716 was placed half-way between the next deepest node.
+We used the leave-one-out cross-validation information criterion (LOOIC) to compare the fit of models (\autoref{tbl-models}) using the *R* package **loo** version `r packageVersion("loo")` [@vehtari_practical_2017] to calculate LOOIC values.
+We selected the model with the lowest LOOIC value to generate posterior predictions for hypothesis testing.
+
+#### Posterior predictions {#sec-predictions}
 
@@
 
-We evaluated predictions using a combination of parameter estimation and model selection.
-In all cases, we estimated parameters and confidence intervals as described in @sec-fitting.
-If the 95% confidence intervals for a parameter did not overlap zero, we considered the parameter to be significantly different from zero.
-We used the leave-one-out cross-validation information criterion (LOOIC) to compare the fit of models (\autoref{tbl-models}) using the *R* package **loo** version `r packageVersion("loo")` [@vehtari_practical_2017] to calculate LOOIC values.
-Note that LOOIC was calculated only from pointwise likelihood values of the submodel estimating effects of light intensity, light treatments, and population on \aax{} (see @sec-parameters).
-We considered models with two standard errors of the mean lower LOOIC value to be a better fit to the data; we considered models with LOOIC values within two standard errors of the mean to be have similar support.
+From the posterior of the selected model (@sec-phylo), we predicted the posterior distribution of the expected \aax{} for each population at each measurement light intensity and growth light intensity.
+If the 95% confidence intervals for \aax{} did not overlap zero, we considered \aax{} to be significantly different from zero.
+We also calculated posterior distribution for the average \aax{} among populations in each condition.
+We tested a linear effect of native light habitat on population-level \aax{} in each treatment condition, integrating over the entire posterior distribution.
+We calculated the posterior distribution of the slope.
+If the 95\% confidence interval for the slope did not overlap zero, we considered the slope to be significant.
+
+## MARKER - need to update table
 
@@
 ```
+
+THIS COULD BE USED FOR LMA MODEL?
+Note that LOOIC was calculated only from pointwise likelihood values of the submodel estimating effects of light intensity, light treatments, and population on \aax{}.
 

### 2025-07-20 ms/ms.tex
example-aa figure

@@
 
-\pandocbounded{\includegraphics[keepaspectratio]{../figures/amphisomy draft 2024-03-28.jpg}}
+\includegraphics[width=2.24in,height=\textheight,keepaspectratio]{../figures/amphisomy draft 2024-03-28.jpg}
 
@@
 plants, where \aax{} was significantly greater under low measurement
-light intensity in 11 of 29 populations compared to to 4 populations for
+light intensity in 11 of 29 populations compared to 4 populations for
 shade-grown plants. The overall average \aax{} of shade and sun grown
@@
 intensity on \aax{}. LMA increased in sun grown plants in all
-populations by an average of 123\% {[}95\% CI: 50.5 to 245\%{]}, similar
+populations by an average of 119\% {[}95\% CI: 47 to 240\%{]}, similar
 to plastic responses in many species (\emph{48}). When LMA is included
@@
 
-\bibitem[\citeproctext]{ref-hansen_stabilizing_1997}
+\bibitem[\citeproctext]{ref-pease_phylogenomics_2016}
 \CSLLeftMargin{64. }%
-\CSLRightInline{T. F. Hansen,
-\href{https://doi.org/10.1111/j.1558-5646.1997.tb01457.x}{Stabilizing
-selection and the comparative analysis of adaptation}. \emph{Evolution}
-\textbf{51}, 1341--1351 (1997).}
+\CSLRightInline{J. B. Pease, D. C. Haak, M. W. Hahn, L. C. Moyle,
+\href{https://doi.org/10.1371/journal.pbio.1002379}{Phylogenomics
+reveals three sources of adaptive variation during a rapid radiation}.
+\emph{PLOS Biology} \textbf{14}, e1002379 (2016).}
 
@@
 \paragraph{\texorpdfstring{\agcurve{} curve
-parameters}{ curve parameters}}\label{curve-parameters}
+parameters}{ curve parameters}}\label{sec-agcurve-pars}
 
@@
 using \agcurve{} curve
-parameters}{ for each light intensity within leaf using  curve parameters}}\label{for-each-light-intensity-within-leaf-using-curve-parameters}
-
-\subparagraph{MARKER - add figure of
-this}\label{marker---add-figure-of-this}
+parameters}{ for each light intensity within leaf using  curve parameters}}\label{sec-aa-est}
 
@@
 quadratic parameters of the amphi and pseudohypo \agcurve{} curves,
-respectively. We calculated \(\widehat{\mathrm{AA}}\) from each draw of
-the posterior for the paired amphi and pseudohypo \agcurve{} curves to
-obtain the posterior distribution of \aax{} for each leaf at each light
-intensity. We then summarized the posterior distribution of \aax{} for
-each leaf at each light intensity by the posterior median, which became
-our estimate of \aax{} for subsequent analyses.
-
-\subsubsection{Bayesian phylogenetic mixed effects
-models}\label{bayesian-phylogenetic-mixed-effects-models}
+respectively (\textbf{?@fig-example-aa}). We calculated
+\(\widehat{\mathrm{AA}}\) from each draw of the posterior for the paired
+amphi and pseudohypo \agcurve{} curves to obtain the posterior
+distribution of \aax{} for each leaf at each light intensity. We then
+summarized the posterior distribution of \aax{} for each leaf at each
+light intensity by the posterior median and standard deviation, which
+became our point estimate of, and uncertainty in, \aax{} for the
+phylogenetic mixed effects model described below.
+
+\begin{figure}[H]
+
+{\centering \pandocbounded{\includegraphics[keepaspectratio]{ms_files/figure-pdf/example-aa-1.pdf}}
+
+}
+
+\caption{Example amphistomy advantage (\aax) calculation from \agcurve{}
+curves in an amphistomatous leaf with gas exchange through both surfaces
+(blue) and the same leaf with gas exchange blocked on the upper surface
+only (orange). We integrate over the gray region between the curves to
+estimate \aax{}. The \aax{} is low (left facet) when the photosynthetic
+rate of an amphistomatous leaf is similar to a pseudohypostomatous leaf
+at the same total stomatal conductance (\(x\)-axes); large \aax{} (right
+facet) is indicated when an amphistomatous leaf has a higher
+photosynthetic rate than a pseudohypostomatous leaf.}
+
+\end{figure}%
+
+\paragraph{Phylogenetic mixed effects models}\label{sec-phylo}
 
@@
 all models, we included population as a phylogenetically structured
-random effect. We crossed those fixed and random effect structures with
-fixed effects of measurement light intensity and/or growth light
-intensity on the distributional parameter, \(\sigma\), the residual
-variance in \aax{}. In total, we fit 100 models, each with a different
-combination of fixed, random, and distributional effects. We used the
-RAxML whole-transcriptome concatenated phylogeny based on 2,745 100-kb
-genomic windows from (\textbf{pease\_phylogenomics\_2016?}). Two of our
-populations were not in this tree. We used accession LA1044 in place of
-LA3909, two populations of \emph{S. galapagnse}; LA0750 was added as
-sister to LA0716, two closely replated populations of \emph{S.
-pennellii}. The node was placed half-way between the next deepest node.
-
-NEED TO EDIT THIS: The next sections describe how we fit models in
-\emph{Stan}, all model parameters and priors, and specific predictions
-about parameter values for each hypothesis. Each curve was fit For
-initial fits, we sampled the posterior distribution from a single chain
-with 1000 iterations after 1000 warmup iterations. After model selection
-(see below), we refit the selected model with more iterations if
-necessary to decrease the Gelman-Rubin convergence statistic
-(\(\hat{R}\)) (\emph{63}) to less than \(`aa_stats[["max_rhat"]]`\) and
-the bulk effective sample size (ESS) to greater than
-\(`aa_stats[["min_ess"]]`\) for all model parameters.
-
-\paragraph{\texorpdfstring{Fitting models in \emph{Stan} using
-\textbf{brms}
-interface}{Fitting models in Stan using brms interface}}\label{sec-fitting}
-
-We fit Bayesian models with HMC sampling in the probabilistic
-programming language \emph{Stan} (\emph{59}) using the \emph{R} package
-\textbf{brms} version 2.22.0 (\emph{60}). We used \emph{CmdStan} version
-2.36.0 and \textbf{cmdstanr} version 0.9.0 (\emph{61}) to interface with
-\emph{R} version 4.5.0 (\emph{62}). For initial fits, we sampled the
-posterior distribution from a single chain with 1000 iterations after
-1000 warmup iterations. After model selection (see below), we refit the
-selected model with more iterations if necessary to decrease the
-Gelman-Rubin convergence statistic (\(\hat{R}\)) (\emph{63}) to less
-than \(`aa_stats[["max_rhat"]]`\) and the bulk effective sample size
-(ESS) to greater than \(`aa_stats[["min_ess"]]`\) for all model
-parameters.
-
-\subparagraph{\texorpdfstring{Effects of light intensity, light
-treatment, and population on
-\aax}{Effects of light intensity, light treatment, and population on }}\label{effects-of-light-intensity-light-treatment-and-population-on}
-
-We tested for effects of light intensity, light treatment, population,
-and their interactions on \aax. All models included effects of light
-intensity and light treatment, as well as random effects of population
-and replicate within population. More complex models included
-interactions between light intensity and light treatment, as well as
-random effects of population on the effects of light intensity and light
-treatment. We used a weakly informative normal prior with mean \(0\) and
-standard deviation \(10\) for fixed effects of light intensity and
-treatment. We used a weakly informative normal prior with mean \(-3\)
-and standard deviation \(5\) for the the random effect standard
-deviation of replicate within population. We accounted for the
-phylogenetic structure among the random effects of population using an
-Ornstein-Uhlenbeck (OU) process (\emph{64}). The expected covariance
-between populations \(i\) and \(j\) (\(\text{Cov}(i, j)\) ) is:
-
-\[\text{Cov}(i, j) = \frac{\sigma^2}{2 \alpha} \exp(-\alpha D_{ij})\]
-where \(\sigma^2\) is the variance of the random effect, \(\alpha\) is
-the rate of decay of the covariance with phylogenetic distance,
-\(D_{ij}\), the phylogenetic distance between populations \(i\) and
-\(j\). We estimated \(\sigma^2 / (2 \alpha)\) and \(\alpha\) as a
-separate parameters and reparameterized them as \(\sigma^2\) and
-\(\alpha\). We used a weakly informative normal prior with mean \(0\)
-and standard deviation \(10\) on OU parameters.
-
-We modeled the residual standard deviation of \aax,
-(i.e.~phylogenetically unstructured variation unaccounted for by
-explanatory variables) on a log-link scale with effects of light
-intensity and light treatment. We used a weakly informative normal prior
-with mean \(-3\) and standard deviation \(5\) on the residual standard
-deviation intercept and a weakly informative normal prior with mean
-\(0\) and standard deviation \(1\) on the effects of light intensity and
-light treatment on the residual standard deviation.
-
-\subparagraph{\texorpdfstring{Effects of native light habitat on
-population-level
-\aax}{Effects of native light habitat on population-level }}\label{effects-of-native-light-habitat-on-population-level}
-
-We tested a linear effect of native light habitat on population-level
-\aax. In Models 1 and 2, we used the random intercept of \aax{} as a
-response variable; in model 3 we used the random intercept plus the
-random effect of population at high light intensity; in model 4 we used
-the random intercept plus the random effect of population in the sun
-treatment; in model 5 we used the random intercept plus the random
-effects of population at high light intensity and sun treatment. We used
-a weakly informative normal prior with mean \(0\) and standard deviation
-\(1\) for the slope and intercept. We accounted for the phylogenetic
-structure among the model residuals using an (OU) process. We used a
-weakly informative normal prior with mean \(0\) and standard deviation
-\(10\) on OU parameters.
-
-\paragraph{Predictions}\label{sec-predictions}
+random effect. We accounted for uncertainty in estimates of \aax{} using
+the standard deviation of the posterior distribution of each \aax{}
+estimate (Section~\ref{sec-aa-est}). We crossed fixed and random effect
+structures with fixed effects of measurement light intensity and/or
+growth light intensity on the distributional parameter, \(\sigma\), the
+residual variance in \aax{}. In total, we fit 100 models, each with a
+different combination of fixed, random, and distributional effects. We
+used the RAxML whole-transcriptome concatenated phylogeny based on 2,745
+100-kb genomic windows from (\emph{64}). Two of our populations were not
+in this tree. We used accession LA1044 in place of LA3909, two
+populations of \emph{S. galapagnse}; LA0750 was added as sister to
+LA0716, two closely related populations of \emph{S. pennellii}. The node
+separating LA0750 from LA0716 was placed half-way between the next
+deepest node. We used the leave-one-out cross-validation information
+criterion (LOOIC) to compare the fit of models (\autoref{tbl-models})
+using the \emph{R} package \textbf{loo} version 2.8.0 (\emph{65}) to
+calculate LOOIC values. We selected the model with the lowest LOOIC
+value to generate posterior predictions for hypothesis testing.
+
+\paragraph{Posterior predictions}\label{sec-predictions}
 
@@
 
-We evaluated predictions using a combination of parameter estimation and
-model selection. In all cases, we estimated parameters and confidence
-intervals as described in Section~\ref{sec-fitting}. If the 95\%
-confidence intervals for a parameter did not overlap zero, we considered
-the parameter to be significantly different from zero. We used the
-leave-one-out cross-validation information criterion (LOOIC) to compare
-the fit of models (\autoref{tbl-models}) using the \emph{R} package
-\textbf{loo} version 2.8.0 (\emph{65}) to calculate LOOIC values. Note
-that LOOIC was calculated only from pointwise likelihood values of the
-submodel estimating effects of light intensity, light treatments, and
-population on \aax{} (see \textbf{?@sec-parameters}). We considered
-models with two standard errors of the mean lower LOOIC value to be a
-better fit to the data; we considered models with LOOIC values within
-two standard errors of the mean to be have similar support.
+From the posterior of the selected model (Section~\ref{sec-phylo}), we
+predicted the posterior distribution of the expected \aax{} for each
+population at each measurement light intensity and growth light
+intensity. If the 95\% confidence intervals for \aax{} did not overlap
+zero, we considered \aax{} to be significantly different from zero. We
+also calculated posterior distribution for the average \aax{} among
+populations in each condition. We tested a linear effect of native light
+habitat on population-level \aax{} in each treatment condition,
+integrating over the entire posterior distribution. We calculated the
+posterior distribution of the slope. If the 95\% confidence interval for
+the slope did not overlap zero, we considered the slope to be
+significant.
+
+\subsection{MARKER - need to update
+table}\label{marker---need-to-update-table}
 
@@
 
+THIS COULD BE USED FOR LMA MODEL? Note that LOOIC was calculated only
+from pointwise likelihood values of the submodel estimating effects of
+light intensity, light treatments, and population on \aax{}.
+
 \subsection{Results}\label{results}

### 2025-07-20 r/212_plot-example-aa.R

@@ -110,15 +110,16 @@ gp_aa = ggplot(df_segment,
     axis.text = element_blank(),
     axis.ticks = element_blank(),
     axis.title = element_text(size = 12),
+    legend.key.height = unit(1.5, "cm"),
     title = element_text(size = 12)
   )
 
-gp_aa
+write_rds(gp_aa, "figures/example-aa.rds")
 
-ggsave(
-  "figures/example-aa.pdf",
-  gp_aa,
-  width = 6.5,
-  height = 3.5,
-  device = cairo_pdf
-)
+# ggsave(
+#   "figures/example-aa.pdf",
+#   gp_aa,
+#   width = 6.5,
+#   height = 3.5,
+#   device = cairo_pdf
+# )

### 2025-07-21 ms/manuscraps.qmd
updated pai analysis

@@
 We considered models with two standard errors of the mean lower LOOIC value to be a better fit to the data; we considered models with LOOIC values within two standard errors of the mean to be have similar support.
+
+#### Light-response (\aqcurve) curves
+
+In `r percent_aa_and_aq`\% of plants, we measured light-response (\aqcurve) curves on the same leaflets as \agcurve{} curves.
+However, when a leaflet was damaged during \agcurve{} curves, we used the next closest leaflet for \aqcurve{} curves.
+Leaves acclimated to high light-intensity (\ppfdequals{2000}), ambient CO$_2$ (\caequals{415}), \rhequals{50}, and \tleafequals{25}.
+After $A$ and \gsw{} stabilized, we measured $A$ at 20 light-intensity levels between $0$ and $\qty{2000}{\micro\mol\raiseto{-2}\meter\raiseto{-1}\second}$ in descending order.

### 2025-07-21 ms/ms.qmd
updated pai analysis

@@
 
-#### Light-response (\aqcurve) curves
-
-In `r percent_aa_and_aq`\% of plants, we measured light-response (\aqcurve) curves on the same leaflets as \agcurve{} curves.
-However, when a leaflet was damaged during \agcurve{} curves, we used the next closest leaflet for \aqcurve{} curves.
+#### Light-saturated photosynthetic rate
+
+In `r percent_aa_and_aq`\% of plants, we measured the maximum photosynthetic rate ($A_\mathrm{max}$) on the same leaflets as \agcurve{} curves.
+However, when a leaflet was damaged during \agcurve{} curves, we used the next closest leaflet for $A_\mathrm{max}$.
 Leaves acclimated to high light-intensity (\ppfdequals{2000}), ambient CO$_2$ (\caequals{415}), \rhequals{50}, and \tleafequals{25}.
-After $A$ and \gsw{} stabilized, we measured $A$ at 20 light-intensity levels between $0$ and $\qty{2000}{\micro\mol\raiseto{-2}\meter\raiseto{-1}\second}$ in descending order.
+After $A$ and \gsw{} stabilized, we measured $A$ at $\qty{2000}{\micro\mol\raiseto{-2}\meter\raiseto{-1}\second}$.
 
@@
 where $\theta_\text{amphi}$ and $\theta_\text{hypo}$ are the quadratic parameters of the amphi and pseudohypo \agcurve{} curves, respectively (@fig-example-aa).
-We calculated $\widehat{\mathrm{AA}}$ from each draw of the posterior for the paired amphi and pseudohypo \agcurve{} curves to obtain the posterior distribution of \aax{} for each leaf at each light intensity.
-We then summarized the posterior distribution of \aax{} for each leaf at each light intensity by the posterior median and standard deviation, which became our point estimate of, and uncertainty in, \aax{} for the phylogenetic mixed effects model described below.
+We calculated $\widehat{\mathrm{AA}}$ from each draw of the posterior for the paired amphi and pseudohypo \agcurve{} curves to obtain the posterior distribution of $\widehat{\mathrm{AA}}$ for each leaf at each light intensity.
+We then summarized the posterior distribution of $\widehat{\mathrm{AA}}$ for each leaf at each light intensity by the posterior median and standard deviation, which became our point estimate of, and uncertainty in, $\widehat{\mathrm{AA}}$ for the phylogenetic mixed effects model described below.
 
 ```{r, example-aa, echo=FALSE,message=FALSE,warning=FALSE}
+#| label: fig-example-aa
 #| fig-align: center
@@
 In all models, we included population as a phylogenetically structured random effect.
-We accounted for uncertainty in estimates of \aax{} using the standard deviation of the posterior distribution of each \aax{} estimate (@sec-aa-est).
+We accounted for uncertainty in $\widehat{\mathrm{AA}}$ using the standard deviation of the posterior distribution of each $\widehat{\mathrm{AA}}$ estimate (@sec-aa-est).
 We crossed fixed and random effect structures with fixed effects of measurement light intensity and/or growth light intensity on the distributional parameter, $\sigma$, the residual variance in \aax{}.
@@
 The node separating LA0750 from LA0716 was placed half-way between the next deepest node.
-We used the leave-one-out cross-validation information criterion (LOOIC) to compare the fit of models (\autoref{tbl-models}) using the *R* package **loo** version `r packageVersion("loo")` [@vehtari_practical_2017] to calculate LOOIC values.
+We used the leave-one-out cross-validation information criterion (LOOIC) to compare the fit of models (\autoref{tbl-aa_loo1}) using the *R* package **loo** version `r packageVersion("loo")` [@vehtari_practical_2017] to calculate LOOIC values.
 We selected the model with the lowest LOOIC value to generate posterior predictions for hypothesis testing.
@@
 
-## MARKER - need to update table
-
 ```{r, predictions, eval=TRUE, echo=FALSE,message=FALSE,warning=FALSE}
 #| label: tbl-predictions
-#| tbl-cap: "Predictions of competing hypotheses about the relationship between \\aax{} and light intensity, light treatment, and native \\ppfd{} among populations.
-The middle column lists specific directional predictions about parameter values and model fit according to leave-one-out cross-validation information criterion (LOOIC), where $\\text{LOOIC}_i$ is the LOOIC value for model $i$.
+#| tbl-cap: "Predictions of competing hypotheses about the relationship between \\aax{} and light intensity, light treatment, and native plant area index (PAI) among populations.
+The middle column lists specific directional predictions about parameter values.
 The rightmost column describes the predictions in words and explains how population-level \\aax{} is calculated in the relevant model."
@@
   ~"\\textbf{Hypothesis}", ~"\\textbf{Prediction(s)}", ~"\\textbf{Description}",
-  "Assimilatory", "$\\beta_{\\mathrm{AA},2000} > 0$", "Average \\aax{} at high light intensity is greater than that at low light intensity",
-  "Plastic", "$\\beta_{\\mathrm{AA},\\text{sun}} > 0$", "Average \\aax{} in sun leaves is greater than that in shade leaves",
-  "Constitutive", "$\\beta_{\\mathrm{PPFD,AA}} > 0$", "Population-level \\aax{} ($\\mathrm{AA}_\\text{acc}$) increases with native \\ppfd",
+  "Assimilatory", "$\\beta_{\\mathrm{AA},2000} > 0$", "\\aax{} at high light intensity is greater than that at low light intensity",
+  "Plastic", "$\\beta_{\\mathrm{AA},\\text{sun}} > 0$", "\\aax{} in sun leaves is greater than that in shade leaves",
+  "Constitutive", "$\\beta_{\\mathrm{PPFD,AA}} > 0$", "Population-level \\aax{} ($\\mathrm{AA}_\\text{acc}$) decreases with native PAI",
   "Constitutive", "", "$\\mathrm{AA}_\\text{acc} = \\beta_{\\mathrm{AA}, 0} + \\vec{\\beta}_{\\mathrm{AA}, \\text{acc}}$",
-  "Assimilatory $\\times$ Plastic", "$\\beta_{\\mathrm{AA},2000,\\text{sun}} > 0$", "Average \\aax{} is highest at high light intensity in sun leaves",
-  "Assimilatory $\\times$ Plastic", "$\\text{LOOIC}_\\text{1} > \\text{LOOIC}_\\text{2}$", "",
-  "Assimilatory $\\times$ Constitutive", "$\\beta_{\\mathrm{AA},2000} > 0$", "Average \\aax{} at high light intensity is greater than that at low light intensity",
-  "Assimilatory $\\times$ Constitutive", "$\\beta_{\\mathrm{PPFD,AA}} > 0$", "Population-level \\aax{} at high light intensity ($\\mathrm{AA}_{\\text{acc},2000}$) increases with native \\ppfd",
-  "Assimilatory $\\times$ Constitutive", "$\\text{LOOIC}_\\text{1} > \\text{LOOIC}_\\text{3}$", "$\\mathrm{AA}_{\\text{acc},2000} = \\beta_{\\mathrm{AA}, 0} + \\vec{\\beta}_{\\mathrm{AA}, \\text{acc}} + \\vec{\\beta}_{\\mathrm{AA}, 2000, \\text{acc}}$",
-  "Plastic $\\times$ Constitutive", "$\\beta_{\\mathrm{AA},\\text{sun}} > 0$", "Average \\aax{} in sun leaves is greater than that in shade leaves",
-  "Plastic $\\times$ Constitutive", "$\\beta_{\\mathrm{PPFD,AA}} > 0$", "Population-level \\aax{} in sun leaves ($\\mathrm{AA}_{\\text{acc},\\text{sun}}$) increases with native \\ppfd",
-  "Plastic $\\times$ Constitutive", "$\\text{LOOIC}_\\text{1} > \\text{LOOIC}_\\text{4}$", "$\\mathrm{AA}_{\\text{acc},\\text{sun}} = \\beta_{\\mathrm{AA}, 0} + \\vec{\\beta}_{\\mathrm{AA}, \\text{acc}} + \\vec{\\beta}_{\\mathrm{AA}, \\text{sun}, \\text{acc}}$",
-  "Assimilatory $\\times$ Plastic $\\times$ Constitutive", "$\\beta_{\\mathrm{AA},2000,\\text{sun}} > 0$", "Average \\aax{} is highest at high light intensity in sun leaves",
-  "Assimilatory $\\times$ Plastic $\\times$ Constitutive", "$\\beta_{\\mathrm{PPFD,AA}} > 0$", "Population-level \\aax{} at high light intensity in sun leaves ($\\mathrm{AA}_{\\text{acc},2000,\\text{sun}}$) increases with native \\ppfd",
-  "Assimilatory $\\times$ Plastic $\\times$ Constitutive", "$\\text{LOOIC}_\\text{3} > \\text{LOOIC}_\\text{5}$", "$\\mathrm{AA}_{\\text{acc},2000,\\text{sun}} = \\beta_{\\mathrm{AA}, 0} + \\vec{\\beta}_{\\mathrm{AA}, \\text{acc}} + \\vec{\\beta}_{\\mathrm{AA}, 2000, \\text{acc}} + \\vec{\\beta}_{\\mathrm{AA}, \\text{sun}, \\text{acc}}$"
+  "Assimilatory $\\times$ Plastic", "$\\beta_{\\mathrm{AA},2000,\\text{sun}} > 0$", "\\aax{} is highest at high light intensity in sun leaves",
+  "Assimilatory $\\times$ Plastic", "", "",
+  "Assimilatory $\\times$ Constitutive", "$\\beta_{\\mathrm{AA},2000} > 0$", "\\aax{} at high light intensity is greater than that at low light intensity",
+  "Assimilatory $\\times$ Constitutive", "$\\beta_{\\mathrm{PPFD,AA}} > 0$", "Population-level \\aax{} at high light intensity ($\\mathrm{AA}_{\\text{acc},2000}$) decreases with native PAI",
+  "Assimilatory $\\times$ Constitutive", "", "$\\mathrm{AA}_{\\text{acc},2000} = \\beta_{\\mathrm{AA}, 0} + \\vec{\\beta}_{\\mathrm{AA}, \\text{acc}} + \\vec{\\beta}_{\\mathrm{AA}, 2000, \\text{acc}}$",
+  "Plastic $\\times$ Constitutive", "$\\beta_{\\mathrm{AA},\\text{sun}} > 0$", "\\aax{} in sun leaves is greater than that in shade leaves",
+  "Plastic $\\times$ Constitutive", "$\\beta_{\\mathrm{PPFD,AA}} > 0$", "Population-level \\aax{} in sun leaves ($\\mathrm{AA}_{\\text{acc},\\text{sun}}$) decreases with native PAI",
+  "Plastic $\\times$ Constitutive", "", "$\\mathrm{AA}_{\\text{acc},\\text{sun}} = \\beta_{\\mathrm{AA}, 0} + \\vec{\\beta}_{\\mathrm{AA}, \\text{acc}} + \\vec{\\beta}_{\\mathrm{AA}, \\text{sun}, \\text{acc}}$",
+  "Assimilatory $\\times$ Plastic $\\times$ Constitutive", "$\\beta_{\\mathrm{AA},2000,\\text{sun}} > 0$", "\\aax{} is highest at high light intensity in sun leaves",
+  "Assimilatory $\\times$ Plastic $\\times$ Constitutive", "$\\beta_{\\mathrm{PPFD,AA}} > 0$", "Population-level \\aax{} at high light intensity in sun leaves ($\\mathrm{AA}_{\\text{acc},2000,\\text{sun}}$) decreases with native PAI",
+  "Assimilatory $\\times$ Plastic $\\times$ Constitutive", "", "$\\mathrm{AA}_{\\text{acc},2000,\\text{sun}} = \\beta_{\\mathrm{AA}, 0} + \\vec{\\beta}_{\\mathrm{AA}, \\text{acc}} + \\vec{\\beta}_{\\mathrm{AA}, 2000, \\text{acc}} + \\vec{\\beta}_{\\mathrm{AA}, \\text{sun}, \\text{acc}}$"
 

### 2025-07-21 ms/ms.tex
updated pai analysis

@@
 
-\paragraph{\texorpdfstring{Light-response (\aqcurve)
-curves}{Light-response () curves}}\label{light-response-curves}
-
-In 91.3\% of plants, we measured light-response (\aqcurve) curves on the
-same leaflets as \agcurve{} curves. However, when a leaflet was damaged
-during \agcurve{} curves, we used the next closest leaflet for
-\aqcurve{} curves. Leaves acclimated to high light-intensity
-(\ppfdequals{2000}), ambient CO\(_2\) (\caequals{415}), \rhequals{50},
-and \tleafequals{25}. After \(A\) and \gsw{} stabilized, we measured
-\(A\) at 20 light-intensity levels between \(0\) and
-\(\qty{2000}{\micro\mol\raiseto{-2}\meter\raiseto{-1}\second}\) in
-descending order.
+\paragraph{Light-saturated photosynthetic
+rate}\label{light-saturated-photosynthetic-rate}
+
+In 91.3\% of plants, we measured the maximum photosynthetic rate
+(\(A_\mathrm{max}\)) on the same leaflets as \agcurve{} curves. However,
+when a leaflet was damaged during \agcurve{} curves, we used the next
+closest leaflet for \(A_\mathrm{max}\). Leaves acclimated to high
+light-intensity (\ppfdequals{2000}), ambient CO\(_2\) (\caequals{415}),
+\rhequals{50}, and \tleafequals{25}. After \(A\) and \gsw{} stabilized,
+we measured \(A\) at
+\(\qty{2000}{\micro\mol\raiseto{-2}\meter\raiseto{-1}\second}\).
 
@@
 quadratic parameters of the amphi and pseudohypo \agcurve{} curves,
-respectively (\textbf{?@fig-example-aa}). We calculated
+respectively (Figure~\ref{fig-example-aa}). We calculated
 \(\widehat{\mathrm{AA}}\) from each draw of the posterior for the paired
 amphi and pseudohypo \agcurve{} curves to obtain the posterior
-distribution of \aax{} for each leaf at each light intensity. We then
-summarized the posterior distribution of \aax{} for each leaf at each
-light intensity by the posterior median and standard deviation, which
-became our point estimate of, and uncertainty in, \aax{} for the
-phylogenetic mixed effects model described below.
-
-\begin{figure}[H]
-
-{\centering \pandocbounded{\includegraphics[keepaspectratio]{ms_files/figure-pdf/example-aa-1.pdf}}
+distribution of \(\widehat{\mathrm{AA}}\) for each leaf at each light
+intensity. We then summarized the posterior distribution of
+\(\widehat{\mathrm{AA}}\) for each leaf at each light intensity by the
+posterior median and standard deviation, which became our point estimate
+of, and uncertainty in, \(\widehat{\mathrm{AA}}\) for the phylogenetic
+mixed effects model described below.
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{ms_files/figure-pdf/fig-example-aa-1.pdf}}
 
@@
 
-\caption{Example amphistomy advantage (\aax) calculation from \agcurve{}
-curves in an amphistomatous leaf with gas exchange through both surfaces
-(blue) and the same leaf with gas exchange blocked on the upper surface
-only (orange). We integrate over the gray region between the curves to
-estimate \aax{}. The \aax{} is low (left facet) when the photosynthetic
-rate of an amphistomatous leaf is similar to a pseudohypostomatous leaf
-at the same total stomatal conductance (\(x\)-axes); large \aax{} (right
-facet) is indicated when an amphistomatous leaf has a higher
-photosynthetic rate than a pseudohypostomatous leaf.}
+\caption{\label{fig-example-aa}Example amphistomy advantage (\aax)
+calculation from \agcurve{} curves in an amphistomatous leaf with gas
+exchange through both surfaces (blue) and the same leaf with gas
+exchange blocked on the upper surface only (orange). We integrate over
+the gray region between the curves to estimate \aax{}. The \aax{} is low
+(left facet) when the photosynthetic rate of an amphistomatous leaf is
+similar to a pseudohypostomatous leaf at the same total stomatal
+conductance (\(x\)-axes); large \aax{} (right facet) is indicated when
+an amphistomatous leaf has a higher photosynthetic rate than a
+pseudohypostomatous leaf.}
 
@@
 all models, we included population as a phylogenetically structured
-random effect. We accounted for uncertainty in estimates of \aax{} using
-the standard deviation of the posterior distribution of each \aax{}
-estimate (Section~\ref{sec-aa-est}). We crossed fixed and random effect
-structures with fixed effects of measurement light intensity and/or
-growth light intensity on the distributional parameter, \(\sigma\), the
-residual variance in \aax{}. In total, we fit 100 models, each with a
-different combination of fixed, random, and distributional effects. We
-used the RAxML whole-transcriptome concatenated phylogeny based on 2,745
-100-kb genomic windows from (\emph{64}). Two of our populations were not
-in this tree. We used accession LA1044 in place of LA3909, two
-populations of \emph{S. galapagnse}; LA0750 was added as sister to
-LA0716, two closely related populations of \emph{S. pennellii}. The node
-separating LA0750 from LA0716 was placed half-way between the next
-deepest node. We used the leave-one-out cross-validation information
-criterion (LOOIC) to compare the fit of models (\autoref{tbl-models})
-using the \emph{R} package \textbf{loo} version 2.8.0 (\emph{65}) to
-calculate LOOIC values. We selected the model with the lowest LOOIC
-value to generate posterior predictions for hypothesis testing.
+random effect. We accounted for uncertainty in \(\widehat{\mathrm{AA}}\)
+using the standard deviation of the posterior distribution of each
+\(\widehat{\mathrm{AA}}\) estimate (Section~\ref{sec-aa-est}). We
+crossed fixed and random effect structures with fixed effects of
+measurement light intensity and/or growth light intensity on the
+distributional parameter, \(\sigma\), the residual variance in \aax{}.
+In total, we fit 100 models, each with a different combination of fixed,
+random, and distributional effects. We used the RAxML
+whole-transcriptome concatenated phylogeny based on 2,745 100-kb genomic
+windows from (\emph{64}). Two of our populations were not in this tree.
+We used accession LA1044 in place of LA3909, two populations of \emph{S.
+galapagnse}; LA0750 was added as sister to LA0716, two closely related
+populations of \emph{S. pennellii}. The node separating LA0750 from
+LA0716 was placed half-way between the next deepest node. We used the
+leave-one-out cross-validation information criterion (LOOIC) to compare
+the fit of models (\autoref{tbl-aa_loo1}) using the \emph{R} package
+\textbf{loo} version 2.8.0 (\emph{65}) to calculate LOOIC values. We
+selected the model with the lowest LOOIC value to generate posterior
+predictions for hypothesis testing.
 
@@
 about the relationship between \aax{} and light intensity, light
-treatment, and native \ppfd{} among populations. The middle column lists
-specific directional predictions about parameter values and model fit
-according to leave-one-out cross-validation information criterion
-(LOOIC), where \(\text{LOOIC}_i\) is the LOOIC value for model \(i\).
-The rightmost column describes the predictions in words and explains how
-population-level \aax{} is calculated in the relevant model.}
+treatment, and native plant area index (PAI) among populations. The
+middle column lists specific directional predictions about parameter
+values. The rightmost column describes the predictions in words and
+explains how population-level \aax{} is calculated in the relevant
+model.}
 
@@
 \midrule
-Assimilatory & $\beta_{\mathrm{AA},2000} > 0$ & \hspace{-1em}Average \aax{} at high light intensity is greater than that at low light intensity\\
+Assimilatory & $\beta_{\mathrm{AA},2000} > 0$ & \hspace{-1em}\aax{} at high light intensity is greater than that at low light intensity\\
 \cmidrule{1-3}\pagebreak[0]
-Plastic & $\beta_{\mathrm{AA},\text{sun}} > 0$ & \hspace{-1em}Average \aax{} in sun leaves is greater than that in shade leaves\\
+Plastic & $\beta_{\mathrm{AA},\text{sun}} > 0$ & \hspace{-1em}\aax{} in sun leaves is greater than that in shade leaves\\
 \cmidrule{1-3}\pagebreak[0]
-Constitutive & $\beta_{\mathrm{PPFD,AA}} > 0$ & \hspace{-1em}Population-level \aax{} ($\mathrm{AA}_\text{acc}$) increases with native \ppfd\\
+Constitutive & $\beta_{\mathrm{PPFD,AA}} > 0$ & \hspace{-1em}Population-level \aax{} ($\mathrm{AA}_\text{acc}$) decreases with native PAI\\
 \nopagebreak
@@
 \cmidrule{1-3}\pagebreak[0]
-Assimilatory $\times$ Plastic & $\beta_{\mathrm{AA},2000,\text{sun}} > 0$ & \hspace{-1em}Average \aax{} is highest at high light intensity in sun leaves\\
+Assimilatory $\times$ Plastic & $\beta_{\mathrm{AA},2000,\text{sun}} > 0$ & \hspace{-1em}\aax{} is highest at high light intensity in sun leaves\\
 \nopagebreak
- & $\text{LOOIC}_\text{1} > \text{LOOIC}_\text{2}$ & \hspace{-1em}\\
+ &  & \hspace{-1em}\\
 \cmidrule{1-3}\pagebreak[0]
-Assimilatory $\times$ Constitutive & $\beta_{\mathrm{AA},2000} > 0$ & \hspace{-1em}Average \aax{} at high light intensity is greater than that at low light intensity\\
+Assimilatory $\times$ Constitutive & $\beta_{\mathrm{AA},2000} > 0$ & \hspace{-1em}\aax{} at high light intensity is greater than that at low light intensity\\
 \nopagebreak
- & $\beta_{\mathrm{PPFD,AA}} > 0$ & \hspace{-1em}Population-level \aax{} at high light intensity ($\mathrm{AA}_{\text{acc},2000}$) increases with native \ppfd\\
+ & $\beta_{\mathrm{PPFD,AA}} > 0$ & \hspace{-1em}Population-level \aax{} at high light intensity ($\mathrm{AA}_{\text{acc},2000}$) decreases with native PAI\\
 \nopagebreak
- & $\text{LOOIC}_\text{1} > \text{LOOIC}_\text{3}$ & \hspace{-1em}$\mathrm{AA}_{\text{acc},2000} = \beta_{\mathrm{AA}, 0} + \vec{\beta}_{\mathrm{AA}, \text{acc}} + \vec{\beta}_{\mathrm{AA}, 2000, \text{acc}}$\\
+ &  & \hspace{-1em}$\mathrm{AA}_{\text{acc},2000} = \beta_{\mathrm{AA}, 0} + \vec{\beta}_{\mathrm{AA}, \text{acc}} + \vec{\beta}_{\mathrm{AA}, 2000, \text{acc}}$\\
 \cmidrule{1-3}\pagebreak[0]
-Plastic $\times$ Constitutive & $\beta_{\mathrm{AA},\text{sun}} > 0$ & \hspace{-1em}Average \aax{} in sun leaves is greater than that in shade leaves\\
+Plastic $\times$ Constitutive & $\beta_{\mathrm{AA},\text{sun}} > 0$ & \hspace{-1em}\aax{} in sun leaves is greater than that in shade leaves\\
 \nopagebreak
- & $\beta_{\mathrm{PPFD,AA}} > 0$ & \hspace{-1em}Population-level \aax{} in sun leaves ($\mathrm{AA}_{\text{acc},\text{sun}}$) increases with native \ppfd\\
+ & $\beta_{\mathrm{PPFD,AA}} > 0$ & \hspace{-1em}Population-level \aax{} in sun leaves ($\mathrm{AA}_{\text{acc},\text{sun}}$) decreases with native PAI\\
 \nopagebreak
- & $\text{LOOIC}_\text{1} > \text{LOOIC}_\text{4}$ & \hspace{-1em}$\mathrm{AA}_{\text{acc},\text{sun}} = \beta_{\mathrm{AA}, 0} + \vec{\beta}_{\mathrm{AA}, \text{acc}} + \vec{\beta}_{\mathrm{AA}, \text{sun}, \text{acc}}$\\
+ &  & \hspace{-1em}$\mathrm{AA}_{\text{acc},\text{sun}} = \beta_{\mathrm{AA}, 0} + \vec{\beta}_{\mathrm{AA}, \text{acc}} + \vec{\beta}_{\mathrm{AA}, \text{sun}, \text{acc}}$\\
 \cmidrule{1-3}\pagebreak[0]
-Assimilatory $\times$ Plastic $\times$ Constitutive & $\beta_{\mathrm{AA},2000,\text{sun}} > 0$ & \hspace{-1em}Average \aax{} is highest at high light intensity in sun leaves\\
+Assimilatory $\times$ Plastic $\times$ Constitutive & $\beta_{\mathrm{AA},2000,\text{sun}} > 0$ & \hspace{-1em}\aax{} is highest at high light intensity in sun leaves\\
 \nopagebreak
- & $\beta_{\mathrm{PPFD,AA}} > 0$ & \hspace{-1em}Population-level \aax{} at high light intensity in sun leaves ($\mathrm{AA}_{\text{acc},2000,\text{sun}}$) increases with native \ppfd\\
+ & $\beta_{\mathrm{PPFD,AA}} > 0$ & \hspace{-1em}Population-level \aax{} at high light intensity in sun leaves ($\mathrm{AA}_{\text{acc},2000,\text{sun}}$) decreases with native PAI\\
 \nopagebreak
- & $\text{LOOIC}_\text{3} > \text{LOOIC}_\text{5}$ & \hspace{-1em}$\mathrm{AA}_{\text{acc},2000,\text{sun}} = \beta_{\mathrm{AA}, 0} + \vec{\beta}_{\mathrm{AA}, \text{acc}} + \vec{\beta}_{\mathrm{AA}, 2000, \text{acc}} + \vec{\beta}_{\mathrm{AA}, \text{sun}, \text{acc}}$\\
+ &  & \hspace{-1em}$\mathrm{AA}_{\text{acc},2000,\text{sun}} = \beta_{\mathrm{AA}, 0} + \vec{\beta}_{\mathrm{AA}, \text{acc}} + \vec{\beta}_{\mathrm{AA}, 2000, \text{acc}} + \vec{\beta}_{\mathrm{AA}, \text{sun}, \text{acc}}$\\
 \bottomrule

### 2025-07-21 r/208_plot-pai.R

@@ -7,13 +7,12 @@ accession_gedi = read_rds("data/accession-gedi.rds") |>
   rename(acc = accession)
 d1 = fit_aa1$data
 
-# Some redundancy with 206_plot-rxn-norms. Need to streamline and condense.
 df_new = crossing(
   acc = unique(d1$acc),
   light_treatment = c("low", "high"),
   light_intensity = c("150", "2000")
 ) |>
-  mutate(row = row_number())
+  mutate(row = row_number(), se_aa = 0)
 
 df_aa_pred1 = posterior_epred(fit_aa1, newdata = df_new) |>
   t() |>
@@ -22,7 +21,27 @@ df_aa_pred1 = posterior_epred(fit_aa1, newdata = df_new) |>
   pivot_longer(cols = -row,
                names_to = "draw",
                values_to = "aa") |>
-  full_join(df_new, by = "row") |>
+  full_join(df_new, by = "row") 
+
+df_coef = df_aa_pred1 |>
+  left_join(accession_gedi, by = join_by(acc)) |>
+  split(~ draw + light_treatment + light_intensity) |>
+  map(\(.x) {
+    lm(aa ~ log(pai), data = .x)
+  }) |>
+  map(coef) |>
+  map(as_tibble) |>
+  map(mutate, term = c("intercept", "slope")) |>
+  imap_dfr(\(.x, .y) {
+    mutate(.x, light_treatment = str_extract(.y, "low|high"),
+           light_intensity = str_extract(.y, "150|2000"))
+  }) 
+
+df_coef_summary = df_coef |>
+  group_by(term, light_treatment, light_intensity) |>
+  point_interval(value) 
+
+df_aa_pred2 = df_aa_pred1 |>
   group_by(acc, light_treatment, light_intensity) |>
   point_interval(aa) |>
   left_join(accession_gedi, by = join_by(acc)) |>
@@ -36,7 +55,7 @@ df_aa_pred1 = posterior_epred(fit_aa1, newdata = df_new) |>
       fct_recode(low = "150", high = "2000")
   )
 
-fig_aa_pai = ggplot(df_aa_pred1, aes(pai, aa, color = Growth, shape = Measurement)) +
+fig_aa_pai = ggplot(df_aa_pred2, aes(pai, aa, color = Growth, shape = Measurement)) +
   facet_grid(Measurement ~ Growth) +
   geom_point() +
   scale_color_manual(values = c("shade" = "tomato4", "sun" = "tomato")) +

### 2025-07-21 ms/ms.qmd
added lineribbon to pai plot

@@
 We sampled the posterior distribution from a single chain with a minimum of `r aa_args$n_iter_init / 2` iterations after `r aa_args$n_iter_init / 2` warmup iterations.
-If necessary, we refit models with more iterations to decrease the Gelman-Rubin convergence statistic ($\hat{R}$) [@gelman_inference_1992] to less than $`aa_stats[["max_rhat"]]`$ and the bulk effective sample size (ESS) to greater than $`aa_stats[["min_ess"]]`$ for all model parameters.
-We also increased the the `adapt_delta` parameter until there were fewer than $`aa_stats[["n_divergent"]]`$ divergent transitions.
+If necessary, we refit models with more iterations to decrease the Gelman-Rubin convergence statistic ($\hat{R}$) [@gelman_inference_1992] to less than $`aa_stats[["max_rhat"]]`$ and the bulk effective sample size (ESS) to greater than $`aa_stats[["min_ess"]]`$ for all model parameters, and there were fewer than $`aa_stats[["n_divergent"]]`$ divergent transitions.
 

### 2025-07-21 r/204_select-model1.R

@@ -51,8 +51,7 @@ aa_models = model_forms |>
   
   crit = 0
   x = 1
-  ad = 0.8
-  
+
   while (crit == 0 & x < 24) {
     m = brm(
       formula = fml,
@@ -81,8 +80,7 @@ aa_models = model_forms |>
       pull(c4)
     
     x = x + 1
-    ad = min(ad * 1.1, 0.99)
-    
+
   }
   
   return(m)

### 2025-07-21 r/205_select-model2.R

@@ -59,8 +59,7 @@ aa_models = model_forms |>
     
     crit = 0
     x = 1
-    ad = 0.8
-    
+
     while (crit == 0 & x < 24) {
       m = brm(
         formula = fml + bf(llma |
@@ -91,8 +90,7 @@ aa_models = model_forms |>
         pull(c4)
       
       x = x + 1
-      ad = min(ad * 1.1, 0.99)
-      
+
     }
     
     return(m)

### 2025-07-21 r/208_plot-pai.R

@@ -18,10 +18,11 @@ df_aa_pred1 = posterior_epred(fit_aa1, newdata = df_new) |>
   t() |>
   as_draws_df() |>
   mutate(row = row_number()) |>
+  dplyr::select(-.chain, -.iteration, -.draw) |>
   pivot_longer(cols = -row,
                names_to = "draw",
                values_to = "aa") |>
-  full_join(df_new, by = "row") 
+  full_join(df_new, by = "row")
 
 df_coef = df_aa_pred1 |>
   left_join(accession_gedi, by = join_by(acc)) |>
@@ -32,14 +33,20 @@ df_coef = df_aa_pred1 |>
   map(coef) |>
   map(as_tibble) |>
   map(mutate, term = c("intercept", "slope")) |>
+  map(pivot_wider, names_from = term) |>
   imap_dfr(\(.x, .y) {
-    mutate(.x, light_treatment = str_extract(.y, "low|high"),
-           light_intensity = str_extract(.y, "150|2000"))
-  }) 
+    mutate(
+      .x,
+      light_treatment = str_extract(.y, "low|high"),
+      light_intensity = str_extract(.y, "150|2000")
+    )
+  })  |>
+  mutate(draw = row_number())
 
 df_coef_summary = df_coef |>
+  pivot_longer(intercept:slope, names_to = "term", values_to = "value") |>
   group_by(term, light_treatment, light_intensity) |>
-  point_interval(value) 
+  point_interval(value)
 
 df_aa_pred2 = df_aa_pred1 |>
   group_by(acc, light_treatment, light_intensity) |>
@@ -55,13 +62,40 @@ df_aa_pred2 = df_aa_pred1 |>
       fct_recode(low = "150", high = "2000")
   )
 
+df_lineribbon = df_coef |>
+  crossing(log_pai = log(seq(
+    min(accession_gedi$pai),
+    max(accession_gedi$pai),
+    length.out = 10
+  ))) |>
+  mutate(aa = intercept + slope * log_pai) |>
+  group_by(log_pai, light_treatment, light_intensity) |>
+  point_interval(aa) |>
+  mutate(
+    pai = exp(log_pai),
+    Growth = light_treatment |>
+      factor(levels = c("low", "high")) |>
+      fct_recode(sun = "high", shade = "low"),
+    Measurement = light_intensity |>
+      factor(levels = c("150", "2000")) |>
[…]

### 2025-07-22 ms/ms.qmd
updated pai-aa figure

@@
 #| fig-height: 0.1
-#| fig-cap: "\\textbf{Amphistomy advantage (\\aax{}) in wild tomatoes is amplified under simulated sunny growth conditions.} \\aax{} ($y$-axis) is the log-response ratio of photosynthesis in an amphistomatous leaf compared to an otherwise identical pseudohypostomatous leaf.
-In both panels, estimates are shown for plants grown under simulated shade or sun and measured under high light intensity (\\ppfdequals{2000}) and low light intensity (\\ppfdequals{150}).
+#| fig-cap: "\\textbf{Amphistomy advantage (\\aax{}) in greater wild tomatoes from open habitats and is further amplified by developmental plasticity under sunny growth conditions.} \\aax{} ($y$-axis) is the log-response ratio of photosynthesis in an amphistomatous leaf compared to an otherwise identical pseudohypostomatous leaf.
+In both panels, estimates are shown for plants grown under simulated shade (brown) or sun (orange) and measured under high light intensity (\\ppfdequals{2000}; open circles) and low light intensity (\\ppfdequals{150}; solid circles).
 The dashed line at zero indicates no difference in photosynthesis between amphi- and pseudohypostomatous leaves.
@@
 (\\textbf{B}) The same estimates of \\aax{} in each combination of growth and measurement light intensity ploted against native plant area index (PAI; $x$-axis, log-scale).
-The confidence intervals were omitted for visual clarity."
+The confidence intervals were omitted for visual clarity.
+The median relationship between native PAI ($x$-axis; log-scale) and \\aax{} is shown as a solid line with the shaded region between dashed lines showing the 95% confidence ribbon.
+The * in the upper left corner of each facet indicates the slope of the linear regression is significantly different from zero; n.s.
+indicates not significant."
 

### 2025-07-22 ms/ms.tex
updated pai-aa figure

@@
 
-\caption{\label{fig-aacap}\textbf{Amphistomy advantage (\aax{}) in wild tomatoes is amplified under simulated sunny growth conditions.}
+\caption{\label{fig-aacap}\textbf{Amphistomy advantage (\aax{}) in greater wild tomatoes from open habitats and is further amplified by developmental plasticity under sunny growth conditions.}
 \aax{} (\(y\)-axis) is the log-response ratio of photosynthesis in an
@@
 pseudohypostomatous leaf. In both panels, estimates are shown for plants
-grown under simulated shade or sun and measured under high light
-intensity (\ppfdequals{2000}) and low light intensity
-(\ppfdequals{150}). The dashed line at zero indicates no difference in
-photosynthesis between amphi- and pseudohypostomatous leaves.
-(\textbf{A}) The points are the posterior median \aax{} for each
-population, with error bars showing 95\% confidence intervals. Within
-each facet, the populations are arranged by \aax{} estimated in that
-growth and measurement condition. (\textbf{B}) The same estimates of
-\aax{} in each combination of growth and measurement light intensity
-ploted against native plant area index (PAI; \(x\)-axis, log-scale). The
-confidence intervals were omitted for visual clarity.}
+grown under simulated shade (brown) or sun (orange) and measured under
+high light intensity (\ppfdequals{2000}; open circles) and low light
+intensity (\ppfdequals{150}; solid circles). The dashed line at zero
+indicates no difference in photosynthesis between amphi- and
+pseudohypostomatous leaves. (\textbf{A}) The points are the posterior
+median \aax{} for each population, with error bars showing 95\%
+confidence intervals. Within each facet, the populations are arranged by
+\aax{} estimated in that growth and measurement condition. (\textbf{B})
+The same estimates of \aax{} in each combination of growth and
+measurement light intensity ploted against native plant area index (PAI;
+\(x\)-axis, log-scale). The confidence intervals were omitted for visual
+clarity. The median relationship between native PAI (\(x\)-axis;
+log-scale) and \aax{} is shown as a solid line with the shaded region
+between dashed lines showing the 95\% confidence ribbon. The * in the
+upper left corner of each facet indicates the slope of the linear
+regression is significantly different from zero; n.s. indicates not
+significant.}
 
@@
 \(`aa_stats[["max_rhat"]]`\) and the bulk effective sample size (ESS) to
-greater than \(`aa_stats[["min_ess"]]`\) for all model parameters. We
-also increased the the \texttt{adapt\_delta} parameter until there were
-fewer than \(`aa_stats[["n_divergent"]]`\) divergent transitions.
+greater than \(`aa_stats[["min_ess"]]`\) for all model parameters, and
+there were fewer than \(`aa_stats[["n_divergent"]]`\) divergent
+transitions.
 
@@
 significant.
-
-\subsection{MARKER - need to update
-table}\label{marker---need-to-update-table}
 

### 2025-07-22 r/208_plot-pai.R

@@ -26,7 +26,7 @@ df_aa_pred1 = posterior_epred(fit_aa1, newdata = df_new) |>
 
 df_coef = df_aa_pred1 |>
   left_join(accession_gedi, by = join_by(acc)) |>
-  split(~ draw + light_treatment + light_intensity) |>
+  split( ~ draw + light_treatment + light_intensity) |>
   map(\(.x) {
     lm(aa ~ log(pai), data = .x)
   }) |>
@@ -52,15 +52,8 @@ df_aa_pred2 = df_aa_pred1 |>
   group_by(acc, light_treatment, light_intensity) |>
   point_interval(aa) |>
   left_join(accession_gedi, by = join_by(acc)) |>
-  mutate(
-    x = order(aa),
-    Growth = light_treatment |>
-      factor(levels = c("low", "high")) |>
-      fct_recode(sun = "high", shade = "low"),
-    Measurement = light_intensity |>
-      factor(levels = c("150", "2000")) |>
-      fct_recode(low = "150", high = "2000")
-  )
+  mutate(x = order(aa)) |>
+  refactor_for_figure()
 
 df_lineribbon = df_coef |>
   crossing(log_pai = log(seq(
@@ -71,16 +64,22 @@ df_lineribbon = df_coef |>
   mutate(aa = intercept + slope * log_pai) |>
   group_by(log_pai, light_treatment, light_intensity) |>
   point_interval(aa) |>
-  mutate(
-    pai = exp(log_pai),
-    Growth = light_treatment |>
-      factor(levels = c("low", "high")) |>
-      fct_recode(sun = "high", shade = "low"),
-    Measurement = light_intensity |>
-      factor(levels = c("150", "2000")) |>
-      fct_recode(low = "150", high = "2000")
-  )
+  mutate(pai = exp(log_pai)) |>
+  refactor_for_figure()
 
+df_text = df_coef_summary |>
+  filter(term == "slope") |>
+  refactor_for_figure() |>
+  mutate(
+    # across(value:.upper, \(.x) sprintf("%.3f", -.x)),
+    # label = glue("-{term}: {value}\n95% CI: {.upper} – {.lower}"),
+    label = case_when(
+      sign(.lower) == sign(.upper) ~ "*",
+      sign(.lower) != sign(.upper) ~ "n.s."
+    ),
+    pai = max(accession_gedi$pai),
+    aa = 0.2
+  )
 
[…]

### 2025-07-22 r/functions.R

@@ -120,3 +120,18 @@ get_panel_dim = function(gp, fig_width, fig_height) {
     panel_width = convertUnit(panel_width, "in")
   )
 }
+
+## Function to refactor treatments for figures
+refactor_for_figure = function(.x) {
+  
+  mutate(
+    .x,
+    Growth = light_treatment |>
+      factor(levels = c("low", "high")) |>
+      fct_recode(sun = "high", shade = "low"),
+    Measurement = light_intensity |>
+      factor(levels = c("150", "2000")) |>
+      fct_recode(low = "150", high = "2000")
+  )
+  
+}

### 2025-07-22 ms/ms.qmd
ph2d stuff

@@
 Within each facet, the populations are arranged by \\aax{} estimated in that growth and measurement condition.
+The average \\aax{} among populations for each treatment group is written in the top left corner.
 (\\textbf{B}) The same estimates of \\aax{} in each combination of growth and measurement light intensity ploted against native plant area index (PAI; $x$-axis, log-scale).
@@
 The median relationship between native PAI ($x$-axis; log-scale) and \\aax{} is shown as a solid line with the shaded region between dashed lines showing the 95% confidence ribbon.
-The * in the upper left corner of each facet indicates the slope of the linear regression is significantly different from zero; n.s.
+The * in the top left corner of each facet indicates the slope of the linear regression is significantly different from zero; n.s.
 indicates not significant."

### 2025-07-22 ms/ms.tex
ph2d stuff

@@
 
-\includegraphics[width=2.24in,height=\textheight,keepaspectratio]{../figures/amphisomy draft 2024-03-28.jpg}
+\includegraphics[width=2.24in,height=\textheight]{../figures/amphisomy draft 2024-03-28.jpg}
 
@@
 
-\pandocbounded{\includegraphics[keepaspectratio]{ms_files/figure-pdf/fig-example-aa-1.pdf}}
+\includegraphics{ms_files/figure-pdf/fig-example-aa-1.pdf}
 

### 2025-07-22 r/207_plot-aa.R

@@ -86,9 +86,9 @@ df_aa_pred3 = posterior_epred(fit_aa, newdata = df_new) |>
 df_aa_text = df_aa_pred1 |>
   ungroup() |>
   summarize(
-    x = median(x),
+    x = 0,
     label = round(mean(aa), 3),
-    aa = 0.20,
+    aa = 0.30,
     .by = c("Growth", "Measurement")
   )
 
@@ -109,7 +109,7 @@ fig_aa = ggplot(
   geom_hline(yintercept = 0, linetype = "dashed") +
   geom_text(data = df_aa_text,
             aes(x, aa, label = label),
-            inherit.aes = FALSE) +
+            inherit.aes = FALSE, hjust = 0, vjust = 1) +
   scale_color_manual(values = c("shade" = "tomato4", "sun" = "tomato")) +
   scale_shape_manual(values = c("low" = 19, "high" = 21)) +
   ylab("amphi advantage") +

### 2025-07-22 r/213_model-ph2d.R

@@ -1,0 +1,51 @@
+source("r/header.R")
+source("r/photo_2d_pm.R")
+
+
+plan(multisession, workers = 10)
+
+handlers(global = TRUE)
+handlers("cli")
+
+# U / 2 %in% c(169, 53, 17) corresponds to approximately 10, 100, 1000 mm^2
+# Stomatal density (number per mm^2)
+# c(10, 100, 1000) |>
+#   # convert to per um^2
+#   multiply_by(1e-6) |>
+#   D2U() |>
+#   divide_by(2)
+
+# Palisade porosity based on Lundgren et al. (2019), fig 5I
+
+input = crossing(
+  n_x = 53,
+  n_z = c(101, 201, 401),
+  stomatal_distribution = c("amphi", "hypo"),
+  I_0 = round(seq(0.00005, 0.002, length.out = 10), 5),
+  phi_pal = 0.1
+) %>%
+  # J_max prop to I_0
+  mutate(J_max = I_0) %>%
+  split(~ seq_len(nrow(.)))
+
+run_photo_2d_pm = function(input, fit_rubisco) {
+  
+  p = progressor(along = input)
+  out = future_map(input, \(.x, fit_rubisco) {
+    
+    r = with(.x, list(I_0 = I_0, n_x = n_x, n_z = n_z, phi_pal = phi_pal))
+    ph2d = photo_2d_pm(.x$stomatal_distribution, fit_rubisco = fit_rubisco, replace = r)
+    ph2d$parms$stomatal_distribution = .x$stomatal_distribution
+    ph2d$parms$A_n = calc_2d_pm_An(ph2d)
+    p()
+    ph2d
+    
+  }, fit_rubisco = fit_rubisco)
+  
+  out
+  
+}
+
+ph2d = run_photo_2d_pm(input, fit_rubisco)
+
+write_rds(ph2d, "objects/ph2d.rds")

### 2025-07-22 r/214_plot-ph2d.R

@@ -1,0 +1,38 @@
+source("r/header.R")
+source("r/photo_2d_pm.R")
+
+ph2d = read_rds("objects/ph2d.rds")
+
+vars = c("stomatal_distribution", "n_z", "n_x", "I_0", "A_n", "phi_pal")
+
+ph2d_vars = ph2d |>
+  map("parms") |>
+  map_dfr(\(.x) as_tibble(.x[vars]))
+
+# tikz("ms/figures/model_summary.tex", standAlone = TRUE, width = 5, height = 5)
+# options(tikzLatexPackages = c(getOption("tikzLatexPackages")))
+
+ph2d_vars |>
+  pivot_wider(names_from = "stomatal_distribution", values_from = "A_n") |>
+  # amphi advantage
+  mutate(
+    ca = log(amphi / hypo),
+    U = glue("$U = {U}~\\mu \\mathrm{{m}}$", U = n_x * 2),
+    T_leaf = glue("$T_\\mathrm{{leaf}} = {T_leaf}~\\mu \\mathrm{{m}}$", T_leaf = n_z - 1),
+    U = fct_reorder(U, n_x),
+    T_leaf = fct_reorder(T_leaf, n_z),
+    `$\\varphi_\\mathrm{pal}$` = as.factor(phi_pal)
+  ) |>
+  ggplot(aes(1e6 * I_0, ca, color = `$\\varphi_\\mathrm{pal}$`)) +
+  facet_grid(T_leaf ~ U) +
+  geom_line() +
+  geom_point() +
+  # scale_x_continuous(limits = c(0, 1000), breaks = c(0, 500, 1000)) +
+  xlab("$I_0~[\\mu \\mathrm{mol}~\\mathrm{m}^{-2}~\\mathrm{s}^{-1}]$") +
+  ylab("amphi advantage") +
+  theme(legend.position = "top")
+
+# dev.off()
+# system("pdflatex -output-directory=ms/figures/ ms/figures/model_summary.tex")
+
+# file.remove(paste0("ms/figures/model_summary.", c("aux", "log")))

### 2025-07-22 r/215_plot-ph2d-example.R

@@ -1,0 +1,37 @@
+source("r/header.R")
+source("r/photo_2d_pm.R")
+
+r = list(I_0 = 0.0001, n_x = 53, n_z = 201, phi_pal = 0.1)
+ph2d_amphi = photo_2d_pm("amphi", fit_rubisco = fit_rubisco, replace = r)
+ph2d_hypo = photo_2d_pm("hypo", fit_rubisco = fit_rubisco, replace = r)
+
+# Plot results
+df_C = expand.grid(
+  z = seq_len(ph2d_amphi$parms[["n_z"]]),
+  x = seq_len(ph2d_amphi$parms[["n_x"]]),
+  name = c("italic(C)[ias]", "italic(C)[liq]"),
+  distribution = c("amphi", "hypo")
+) |>
+  mutate(value = c(ph2d_amphi$fit$y, ph2d_hypo$fit$y))
+
+df_points = crossing(
+  name = c("italic(C)[ias]", "italic(C)[liq]"),
+  distribution = c("amphi", "hypo"),
+  z = c(0, r$n_z)
+) |>
+  mutate(x = 0 + (z > 0 & distribution == "amphi") * r$n_x, value = numeric(8))
+
+ggplot(df_C, aes(x, z)) +
+  facet_grid(distribution ~ name, labeller = label_parsed) +
+  scale_x_continuous(breaks = c(0, 50, 100), expand = expansion(mult = rep(0.1, 2))) +
+  scale_y_continuous(breaks = c(0, 100, 200)) +
+  geom_contour_filled(mapping = aes(z = value), breaks = seq(0.004, 0.015, 0.0005)) +
+  geom_point( data = df_points) +
+  xlab(expression(distance~from~abaxial~stomate~bgroup("[", paste(mu, 'm'), "]"))) +
+  ylab(expression(distance~from~abaxial~epidermis~bgroup("[", paste(mu, 'm'), "]"))) +
+  coord_equal() +
+  guides(fill = guide_legend(expression(group('[', CO[2], ']')~mol~m^-3))) +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1))
+
+ggsave("ms/figures/ph2d-example.pdf", width = 6.5, height = 4.5)
+write_rds(ph2d_offset, "objects/ph2d_offset.rds")

### 2025-07-22 r/header.R

@@ -15,11 +15,14 @@ library(grid)
 library(loo)
 library(lubridate)
 library(magrittr)
+library(mgcv)
 library(mvnfast)
 library(phangorn)
+library(progressr)
 library(purrr)
 library(readr)
 library(rlang)
+library(rootSolve)
 library(scales)
 library(stringr)
 library(tibble)
@@ -28,10 +31,10 @@ library(tidyr)
 library(TreeTools)
 
 source("r/functions.R")
-source("r/licor-functions.R")
+# source("r/licor-functions.R") # I don't think I am using anymore
 
 # format of acceptable IDs
-id_string = "^(LA[0-9]{4}A*|nelsonii|sandwicense)-[A-Z]{1}[AB]{0,1}$"
+id_string = "^(LA[0-9]{4}A*)-[A-Z]{1}[AB]{0,1}$"
 
 theme_set(theme_cowplot())
 
@@ -47,3 +50,20 @@ aa_args = read_rds("objects/aa_args.rds") |>
   )
 
 write_rds(aa_args, "objects/aa_args1.rds")
+
+# Fit Rubisco model
+x_depth = c(145, 830.5)
+y_depth = c(0, 700)
+x_rubisco = c(558, 56.5)
+y_rubisco = c(0, 100)
+nishio_carbon_1993_fig5 = read_csv("data/nishio_carbon_1993_fig5.csv",
+                                   show_col_types = FALSE) |>
+  mutate(
+    depth = raw_depth * (diff(y_depth) / diff(x_depth)) - x_depth[1],
+    rubisco = raw_rubisco * (diff(y_rubisco) / diff(x_rubisco)) -
+      x_rubisco[1] * (diff(y_rubisco) / diff(x_rubisco)),
+    # Reversing so that 0 is at abaxial, 1 is at adaxial, to match Earles et al.
+    rel_depth = rev(depth / (min(depth) + max(depth)))
+  )
+
+fit_rubisco = gam(rubisco ~ s(rel_depth), data = nishio_carbon_1993_fig5)

### 2025-07-22 r/photo_2d_pm.R

@@ -1,0 +1,310 @@
+# Model photosynthesis using a 2D porous medium approximation
+#
+# Note that n_z = 0 is bottom (abaxial) side of leaf
+
+# 2D photosynthesis model using a porous medium approximation
+photo_2d_pm = function(stomatal_distribution, replace = NULL, ...) {
+  
+  # Set parameters
+  parms = set_2d_pm_parms(replace) |>
+    derive_2d_pm_parms(fit_rubisco = fit_rubisco) |>
+    make_2d_pm_mat() |>
+    add_stomatal_position(stomatal_distribution)
+  
+  # Find solution
+  C_init = rep(parms[["C_stom"]], 2 * parms[["n_z"]] * parms[["n_x"]])
+  fit = steady.2D(C_init, func = rd_2p_pm, parms = parms,
+                  pos = FALSE, dimens = c(parms[["n_z"]], parms[["n_x"]]),
+                  nspec = 2, lrw = 1e8, atol = 1e-10, rtol = 1e-10, ctol = 1e-10)
+  
+  # Structure output
+  list(fit = fit, parms = parms)
+  
+}
+
+# Function to add stomatal position to parms
+add_stomatal_position = function(parms, stomatal_distribution) {
+  
+  parms[["x_abaxial_stomata"]] = switch(
+    stomatal_distribution,
+    amphi = 1,
+    hypo = c(1, parms[["n_x"]])
+  )
+  parms[["x_adaxial_stomata"]] = switch(
+    stomatal_distribution,
+    amphi = parms[["n_x"]],
+    hypo = NULL
+  )
+  parms
+  
+}
+
+# Set parameters for photo_2d_pm()
+set_2d_pm_parms = function(replace, ...) {
+  
+  parms = get_2d_pm_default_parms()
+  if (!is.null(replace)) {
+    parms[names(replace)] = replace[names(replace)]
+  }
+  parms
+  
+}
+
+# Get set of parameters for photo_2d_pm()
+get_2d_pm_default_parms = function(...) {
+  
+  read_csv("data/2d-pm-parameters.csv", col_types = "cccdccl") |>
+    filter(!calculated) |>
+    select(r, default_value) |>
+    with(split(default_value, r))
+  
+}
+
[…]

### 2025-07-22 ms/ms.qmd
updating PAI results in text

@@
 
+## PAI results
+df_coef_summary = read_rds("../objects/df_coef_summary.rds") |>
+  filter(term == "slope") |>
+  mutate(signif = (sign(.lower) == sign(.upper))) |>
+  summarize(n_signif= sum(signif),
+            n_total = n()) |>
+  as.list()
+
 ```
@@
 <!-- Paragraph 5 (summarize methods) -->
-We tested these hypotheses by comparing \aax{} among amphistomatous wild tomato species [@peralta_taxonomy_2008] from different native light habitats, grown under simulated sun and shade light treatments, and measured under contrasting light intensity ().
-We measured AA on `r n_plant` individual plants from `r n_acc` populations (average of `r mean_n_rep` replicates per light treatment) using a recently developed method [@triplett_amphistomy_2024].
+We tested these hypotheses by comparing \aax{} among amphistomatous wild tomato species [@peralta_taxonomy_2008] from different native light habitats, grown under simulated sun and shade light treatments, and measured under contrasting light intensities (low and high).
+We measured \aax{} on `r n_plant` individual plants from `r n_acc` populations (average of `r mean_n_rep` replicates per light treatment) using a recently developed method [@triplett_amphistomy_2024].
 With this method, we directly compare the photosynthetic rate of an untreated amphistomatous leaf to that of the same leaf with gas exchange blocked through the adaxial (upper) surface by transparent plastic, which we refer to as 'pseudohypostomy'.
@@
 The native light intensity was represented by plant area index (PAI $\unit{\meter\squared\per\meter\squared}$), estimated using a global gridded data set derived from the Global Ecosystem Dynamics Investigation [GEDI; @burns_multi-resolution_2024] and georeferenced accession collection information from the Tomato Genetics Resource Center.
-The growth light intensities were \ppfdequals{`r ppfd_sun`} (sun treatment) and \ppfdqty{`r ppfd_shade`} (shade treatment) while all other environment conditions were nearly identical (see materials and methods in supplementary materials).
-The high and low measurements intensities were \ppfdequals{2000} (`r p_red_high`:`r p_blue_high` red:blue) and \ppfdequals{150} (`r p_red_low`:`r p_blue_low` red:blue), respectively.
+The growth light intensities were \ppfdequals{`r ppfd_sun`} (sun treatment) and \ppfdqty{`r ppfd_shade`} (shade treatment) while all other environment conditions were nearly identical (see Materials and Methods).
+The high and low measurement light intensities were \ppfdequals{2000} (`r p_red_high`:`r p_blue_high` red:blue) and \ppfdequals{150} (`r p_red_low`:`r p_blue_low` red:blue), respectively.
 
@@
 @fig-aa]A).
-Bayesian phylogenetic mixed effects models that allowed \aax{} to vary between measurement light intensities, growth light intensities, and among populations outperformed simpler models based on their information criteria (@tbl-aa_loo1).
+Bayesian phylogenetic mixed effects models that allowed \aax{} to vary between measurement light intensities, growth light intensities, and among populations outperformed simpler models based on information criteria (@tbl-aa_loo1).
 Measured under high light intensity, \aax{} was consistently greater for sun plants.
@@
 The overall average \aax{} of shade and sun grown plants measured under low light intensity was `r filter(aa_mean_range, light_treatment == "low", light_intensity == "150")$x` and `r filter(aa_mean_range, light_treatment == "high", light_intensity == "150")$x`, respectively.
-There was a slight tendency for populations from more closed habitats (greater PAI), but \aax{} varied widely among populations from open habitats (low PAI) regardless of growth and measurement light intensities ([Fig.
+There was a modest tendency for populations from more open habitats (lower PAI) to exhibit greater \aax{} and the slope was significantly different than 0 in `r df_coef_summary[['n_signif']]` of the `r df_coef_summary[['n_total']]` treatment combinations ([Fig.
 @fig-aa]B).

### 2025-07-22 ms/ms.tex
updating PAI results in text

@@
 under simulated sun and shade light treatments, and measured under
-contrasting light intensity (). We measured AA on 572 individual plants
-from 29 populations (average of 9.86 replicates per light treatment)
-using a recently developed method (\emph{46}). With this method, we
-directly compare the photosynthetic rate of an untreated amphistomatous
-leaf to that of the same leaf with gas exchange blocked through the
-adaxial (upper) surface by transparent plastic, which we refer to as
-`pseudohypostomy'. To compare amphi- and pseudohypostomatous leaves at
-identical whole-leaf \(g_\text{sc}\), we measure \(A\) over a range of
-\(g_\text{sc}\), inducing stomatal opening and closure by modulating
-humidity (see Materials and Methods for further details). We estimated
-`amphistomy advantage' (\aax) \emph{sensu} (\emph{20}), but with
-modifications previously described in (\emph{46}) and here (Materials
-and Methods). The native light intensity was represented by plant area
-index (PAI \(\unit{\meter\squared\per\meter\squared}\)), estimated using
-a global gridded data set derived from the Global Ecosystem Dynamics
-Investigation {[}GEDI; (\emph{52}){]} and georeferenced accession
-collection information from the Tomato Genetics Resource Center. The
-growth light intensities were \ppfdequals{761} (sun treatment) and
-\ppfdqty{115} (shade treatment) while all other environment conditions
-were nearly identical (see materials and methods in supplementary
-materials). The high and low measurements intensities were
+contrasting light intensities (low and high). We measured \aax{} on 572
+individual plants from 29 populations (average of 9.86 replicates per
+light treatment) using a recently developed method (\emph{46}). With
+this method, we directly compare the photosynthetic rate of an untreated
+amphistomatous leaf to that of the same leaf with gas exchange blocked
+through the adaxial (upper) surface by transparent plastic, which we
+refer to as `pseudohypostomy'. To compare amphi- and pseudohypostomatous
+leaves at identical whole-leaf \(g_\text{sc}\), we measure \(A\) over a
+range of \(g_\text{sc}\), inducing stomatal opening and closure by
+modulating humidity (see Materials and Methods for further details). We
+estimated `amphistomy advantage' (\aax) \emph{sensu} (\emph{20}), but
+with modifications previously described in (\emph{46}) and here
+(Materials and Methods). The native light intensity was represented by
+plant area index (PAI \(\unit{\meter\squared\per\meter\squared}\)),
+estimated using a global gridded data set derived from the Global
+Ecosystem Dynamics Investigation {[}GEDI; (\emph{52}){]} and
+georeferenced accession collection information from the Tomato Genetics
+Resource Center. The growth light intensities were \ppfdequals{761} (sun
+treatment) and \ppfdqty{115} (shade treatment) while all other
+environment conditions were nearly identical (see Materials and
+Methods). The high and low measurement light intensities were
 \ppfdequals{2000} (97.8:2.24 red:blue) and \ppfdequals{150} (87.0:13.0
@@
 measurement light intensities, growth light intensities, and among
-populations outperformed simpler models based on their information
-criteria (Table~\ref{tbl-aa_loo1}). Measured under high light intensity,
-\aax{} was consistently greater for sun plants. The average \aax{} among
+populations outperformed simpler models based on information criteria
+(Table~\ref{tbl-aa_loo1}). Measured under high light intensity, \aax{}
+was consistently greater for sun plants. The average \aax{} among
 populations in the shade treatment was 0.041 (range: 0.010--0.113; 21 of
@@
 0.036--0.197; 24 of 29 populations significant), respectively. There was
-a slight tendency for populations from more closed habitats (greater
-PAI), but \aax{} varied widely among populations from open habitats (low
-PAI) regardless of growth and measurement light intensities
-(Fig.~\ref{fig-aa}B).
+a modest tendency for populations from more open habitats (lower PAI) to
+exhibit greater \aax{} and the slope was significantly different than 0
+in 3 of the 4 treatment combinations (Fig.~\ref{fig-aa}B).
 
@@
 confidence intervals. Within each facet, the populations are arranged by
-\aax{} estimated in that growth and measurement condition. (\textbf{B})
-The same estimates of \aax{} in each combination of growth and
-measurement light intensity ploted against native plant area index (PAI;
-\(x\)-axis, log-scale). The confidence intervals were omitted for visual
-clarity. The median relationship between native PAI (\(x\)-axis;
-log-scale) and \aax{} is shown as a solid line with the shaded region
-between dashed lines showing the 95\% confidence ribbon. The * in the
-upper left corner of each facet indicates the slope of the linear
-regression is significantly different from zero; n.s. indicates not
-significant.}
+\aax{} estimated in that growth and measurement condition. The average
+\aax{} among populations for each treatment group is written in the top
+left corner. (\textbf{B}) The same estimates of \aax{} in each
+combination of growth and measurement light intensity ploted against
+native plant area index (PAI; \(x\)-axis, log-scale). The confidence
+intervals were omitted for visual clarity. The median relationship
+between native PAI (\(x\)-axis; log-scale) and \aax{} is shown as a
+solid line with the shaded region between dashed lines showing the 95\%
+confidence ribbon. The * in the top left corner of each facet indicates
+the slope of the linear regression is significantly different from zero;
+n.s. indicates not significant.}
 

### 2025-07-22 r/208_plot-pai.R

@@ -120,6 +120,8 @@ ggsave(
   bg = "transparent"
 )
 
+write_rds(df_coef_summary, "objects/df_coef_summary.rds")
+
 df_new = crossing(acc = unique(d1$acc),
                   light_treatment = c("low", "high")) |>
   mutate(row = row_number())

### 2025-07-23 ms/ms.qmd
loo table 2

@@
 bibliography: solanum-aa.bib
-csl: https://www.zotero.org/styles/science
+csl: science.csl
 ---
@@
 
+### MARKER - where is evidence for relationship between PAI and LMA, AA?
+
 <!-- Paragraph 7 (summarize secondary results) -->
-The pattern of \aax{} across wild tomatoes strongly supports the plasticity hypothesis, argues against the acclimatory hypothesis, and provides only weak support for the constitutive hypothesis.
-Plastic changes in leaf thickness and/or packing density, summarized by the bulk leaf mass per area (LMA), likely mediate the effect of growth light intensity on \aax{}.
+The pattern of \aax{} across wild tomatoes strongly supports the plasticity hypothesis, argues against the acclimatory hypothesis, and provides modest support for the constitutive hypothesis.
+Plastic changes in leaf thickness and/or packing density, summarized by the bulk leaf mass per area (LMA), may mediate the effect of growth light intensity on \aax{}.
 LMA increased in sun grown plants in all populations by an average of `r signif(100 * (exp(post_lma["Estimate"]) - 1), 3)`% [95% CI: `r signif(100 * (exp(post_lma["Q2.5"]) - 1), 3)` to `r signif(100 * (exp(post_lma["Q97.5"]) - 1), 3)`%], similar to plastic responses in many species [@poorter_metaanalysis_2019].
-When LMA is included in the model, the effect of growth light intensity on \aax{} is no longer significant based on model comparison (Table S8) and LMA is weakly positively correlated with \aax{} ([Fig.
-@fig-lma_aa]).
-In contrast, LMA did not mediate an indirect effect of native PAI on \aax{} (Table S8).
+While LMA is weakly, albeit significantly, associated with individual-level \aax{} ([Fig.
+@fig-lma_aa]), the effect of growth light intensity on \aax{} is still predictive based on model comparison (@tbl-aa_loo1).
+In contrast, LMA did not mediate an indirect effect of native PAI on \aax{} (Table/Figure).
 Many anatomical traits underlie LMA [@john_anatomical_2017] and future research will be needed identify which particular traits, such as leaf thickness or mesophyll porosity, are responsible for mediating \aax{}.
@@
 At low light intensity, the bulk properties of the leaf are weighed toward the upper palisade where most light is intercepted.
-Hence, amphistomy may, expectedly, be particularly beneficial for sun leaves experiencing intermittant shade or cloud cover.
+Hence, amphistomy may, unexpectedly, be particularly beneficial for sun leaves experiencing intermittent shade or cloud cover.
 Our study is limited in testing this because we could not directly measure the stomatal conductance ratio and intercellular resistance on each surface.
@@
 ```
+
+### MARKER - make other LOOIC table
+
+```{r, aa_loo2, eval=TRUE,echo=FALSE,message=FALSE,warning=FALSE}
+#| label: tbl-aa_loo2
+#| tbl-cap: 'Leave-one-out cross-validation information criterion (LOOIC) for models of \\aax{} as a function of light intensity, light treatment, leaf mass per area (LMA), and population.
+The model with the lowest LOOIC is considered the best fit to the data.
+The LOOIC column lists the LOOIC value for each model; the $\\Delta$LOOIC column lists the difference between the LOOIC of the corresponding model and the best model (lowest LOOIC).
+Checkmarks in each row indicate whether the corresponding model included affects of measurement light intensity ("Meas."), growth light intensity ("Growth"), or LMA as fixed effects, random effects, or distributional effect on the residual variance.
+We test for fixed interaction effects between measurement and growth light intensity and LMA.
+We also test for random interaction effects between measurement and growth light intensity.
+All models included phylogenetically structured random effects of population.'
+
+df1 = aa_loo_table2 |>
+  mutate(
+    fixed_meas = str_detect(fixed, "light_intensity"),
+    fixed_growth = str_detect(fixed, "light_treatment"),
+    fixed_lma = str_detect(fixed, "mi\\(llma\\)"),
+    fixed_intrxn123 = str_detect(fixed, "light_treatment \\* light_intensity \\* mi\\(llma\\)"),
+    fixed_intrxn12 = str_detect(fixed, "light_treatment \\* light_intensity") | fixed_intrxn123,
+    fixed_intrxn13 = str_detect(fixed, "light_treatment \\* mi\\(llma\\)") | fixed_intrxn123,
+    fixed_intrxn23 = str_detect(fixed, "light_intensity \\* mi\\(llma\\)") | fixed_intrxn123,
+    # random_acc = TRUE,
+    random_meas = str_detect(random, "light_intensity"),
+    random_growth = str_detect(random, "light_treatment"),
+    random_intrxn = str_detect(random, "\\*"),
+    sigma_meas = str_detect(sigma, "light_intensity"),
+    sigma_growth = str_detect(sigma, "light_treatment"),
+  ) |>
+  dplyr::select(
+    LOOIC = looic,
+    `$\\Delta$LOOIC` = delta_looic,
+    starts_with("fixed_"),
+    starts_with("random_"),
+    starts_with("sigma_")
+  ) |>
+  mutate(across(where(is.logical), ~ ifelse(.x, "$\\checkmark$", "")))
+
+c1 = colnames(df1)
+groupings = list(
+  " " = sum(str_detect(colnames(df1), "LOOIC")),
+  "Fixed" = sum(str_detect(colnames(df1), "fixed_")),
+  "Random" = sum(str_detect(colnames(df1), "random_")),
+  "Distributional" = sum(str_detect(colnames(df1), "sigma_"))
+)
+
+c2 = case_when(
+  str_detect(c1, "_meas") ~ "Meas.",
+  str_detect(c1, "_growth") ~ "Growth",
+  str_detect(c1, "_lma") ~ "LMA",
+  str_detect(c1, "_intrxn12") ~ "Meas. x Growth",
+  str_detect(c1, "_intrxn13") ~ "Meas. x LMA",
+  str_detect(c1, "_intrxn23") ~ "Growth x LMA",
+  str_detect(c1, "_intrxn123") ~ "Meas. x Growth x LMA",
+  str_detect(c1, "_acc") ~ "Population",
+  .default = c1
+)
+
+kbl(
+  df1,
+  col.names = c2,
+  booktabs = TRUE,
+  escape = FALSE,
+  align = "c",
+  longtable = TRUE,
+  digits = 2
+) |>
+  kable_styling(latex_options = c("striped", "repeat_header")) |>
+  add_header_above(groupings) |>
+  column_spec(1, width = "0.7in") |>
+  column_spec(2, width = "0.5in") |>
+  column_spec(3:10, width = "0.4in") |>
+  column_spec(c(5, 8), border_right = TRUE)
+
+```

### 2025-07-23 ms/ms.tex
loo table 2

@@
 
+\subsubsection{MARKER - where is evidence for relationship between PAI
+and LMA,
+AA?}\label{marker---where-is-evidence-for-relationship-between-pai-and-lma-aa}
+
 The pattern of \aax{} across wild tomatoes strongly supports the
 plasticity hypothesis, argues against the acclimatory hypothesis, and
-provides only weak support for the constitutive hypothesis. Plastic
-changes in leaf thickness and/or packing density, summarized by the bulk
-leaf mass per area (LMA), likely mediate the effect of growth light
-intensity on \aax{}. LMA increased in sun grown plants in all
-populations by an average of 119\% {[}95\% CI: 47 to 240\%{]}, similar
-to plastic responses in many species (\emph{48}). When LMA is included
-in the model, the effect of growth light intensity on \aax{} is no
-longer significant based on model comparison (Table S8) and LMA is
-weakly positively correlated with \aax{} (Fig.~\ref{fig-lma_aa}). In
-contrast, LMA did not mediate an indirect effect of native PAI on \aax{}
-(Table S8). Many anatomical traits underlie LMA (\emph{53}) and future
-research will be needed identify which particular traits, such as leaf
-thickness or mesophyll porosity, are responsible for mediating \aax{}.
-The fact that the \aax{} of sun plants was greater under low measurement
-light intensity supports a long-standing hypothesis that resistance to
-CO\(_2\) diffusion is greater in the upper than lower portions of the
-leaf interior (\emph{54}). At low light intensity, the bulk properties
-of the leaf are weighed toward the upper palisade where most light is
-intercepted. Hence, amphistomy may, expectedly, be particularly
-beneficial for sun leaves experiencing intermittant shade or cloud
-cover. Our study is limited in testing this because we could not
-directly measure the stomatal conductance ratio and intercellular
-resistance on each surface. Future experiments measuring \aax{} with a
-dual sided chamber (\emph{31}, \emph{33}) can overcome these
-limitations.
+provides modest support for the constitutive hypothesis. Plastic changes
+in leaf thickness and/or packing density, summarized by the bulk leaf
+mass per area (LMA), may mediate the effect of growth light intensity on
+\aax{}. LMA increased in sun grown plants in all populations by an
+average of 119\% {[}95\% CI: 47 to 240\%{]}, similar to plastic
+responses in many species (\emph{48}). While LMA is weakly, albeit
+significantly, associated with individual-level \aax{}
+(Fig.~\ref{fig-lma_aa}), the effect of growth light intensity on \aax{}
+is still predictive based on model comparison (Table~\ref{tbl-aa_loo1}).
+In contrast, LMA did not mediate an indirect effect of native PAI on
+\aax{} (Table/Figure). Many anatomical traits underlie LMA (\emph{53})
+and future research will be needed identify which particular traits,
+such as leaf thickness or mesophyll porosity, are responsible for
+mediating \aax{}. The fact that the \aax{} of sun plants was greater
+under low measurement light intensity supports a long-standing
+hypothesis that resistance to CO\(_2\) diffusion is greater in the upper
+than lower portions of the leaf interior (\emph{54}). At low light
+intensity, the bulk properties of the leaf are weighed toward the upper
+palisade where most light is intercepted. Hence, amphistomy may,
+unexpectedly, be particularly beneficial for sun leaves experiencing
+intermittent shade or cloud cover. Our study is limited in testing this
+because we could not directly measure the stomatal conductance ratio and
+intercellular resistance on each surface. Future experiments measuring
+\aax{} with a dual sided chamber (\emph{31}, \emph{33}) can overcome
+these limitations.
 
@@
 
+\subsubsection{MARKER - make other LOOIC
+table}\label{marker---make-other-looic-table}
+
+\begin{table}
+
+\caption{\label{tbl-aa_loo2}Leave-one-out cross-validation information
+criterion (LOOIC) for models of \aax{} as a function of light intensity,
+light treatment, leaf mass per area (LMA), and population. The model
+with the lowest LOOIC is considered the best fit to the data. The LOOIC
+column lists the LOOIC value for each model; the \(\Delta\)LOOIC column
+lists the difference between the LOOIC of the corresponding model and
+the best model (lowest LOOIC). Checkmarks in each row indicate whether
+the corresponding model included affects of measurement light intensity
+(``Meas.''), growth light intensity (``Growth''), or LMA as fixed
+effects, random effects, or distributional effect on the residual
+variance. We test for fixed interaction effects between measurement and
+growth light intensity and LMA. We also test for random interaction
+effects between measurement and growth light intensity. All models
+included phylogenetically structured random effects of population.}
+
+\centering{
+
+}
+
+\end{table}%
+
 \end{document}

### 2025-07-23 r/205_select-model2.R

@@ -125,7 +125,8 @@ aa_loo_table = tibble(
   mutate(
     delta_looic = looic - min(looic),
     best_model = delta_looic == 0
-  )
+  ) |>
+  left_join(model_forms, by = join_by(model))
 
 write_rds(aa_loo_table, "objects/aa_loo_table2.rds")
 best_model_index = str_extract(aa_loo_table[1, "model"], "\\d+") |>

### 2025-07-23 r/scratch - fit overlap only.R

@@ -1,73 +1,0 @@
-# I'm not sure if this is necessary...I think the idea was just to fit curves where they overlap, but I don't think this is necessary.
-source("r/header.R")
-
-aa_summary = read_rds("objects/aa_summary.rds")
-
-# Summarize RH curves to get overlap of log_gsw values ----
-trimmed_rh_curves = read_rds("data/trimmed_rh_curves.rds") 
-
-df1 = trimmed_rh_curves |>
-  summarize(
-    min_log_gsw = min(log_gsw),
-    max_log_gsw = max(log_gsw),
-    .by = c(
-      "acc",
-      "acc_id",
-      "light_treatment",
-      "light_intensity",
-      "leaf_type"
-    )
-  ) |>
-  reframe(
-    p = seq_len(20),
-    log_gsw = seq(
-      min_log_gsw,
-      max_log_gsw,
-      length.out = 20
-    ),
-    .by = c(
-      "acc",
-      "acc_id",
-      "light_treatment",
-      "light_intensity",
-      "leaf_type"
-    )
-  )
-
-# Draws from the curve fits ----
-curve_fits_draws = read_rds("objects/curve-fits-draws.rds") |>
-  rename(b0 = b_Intercept, b1 = b_polylog_gsw2rawEQTRUE1, b2 = b_polylog_gsw2rawEQTRUE2) |>
-  dplyr::select(starts_with("."), matches("b[0-2]"), file) |>
-  mutate(accid_leaftype_lightintensity = str_remove(file, ".rds"),
-         .keep = "unused") |>
-  separate_wider_delim(
-    accid_leaftype_lightintensity,
-    delim = "_",
-    names = c("acc_id", "leaf_type", "light_intensity")
-  )
-
-df2 = full_join(df1, curve_fits_draws, by = join_by(acc_id, light_intensity, leaf_type)) |>
-  mutate(log_A = b0 + b1 * log_gsw + b2 * log_gsw^2) |>
-  select(-b0, -b1, -b2) |>
-  group_by(acc, acc_id, light_treatment, light_intensity, leaf_type, p) |>
-  point_interval(log_A) |>
-  left_join(df1,
-            by = join_by(acc, acc_id, light_treatment, light_intensity, leaf_type, p))
-
-
-.x =  "LA0436-P"
-aa_summary |>
-  arrange(desc(sd))
-
[…]

### 2025-07-23 r/scratch new aa function.R

@@ -1,27 +1,0 @@
-# FIGURE OUT FUNCTION TO PULL OUT REGION FOR FITTING
-source("r/header.R")
-
-rh_curves = read_rds("data/rh_hi_curves.rds")
-
-df1 = filter(rh_curves, acc_id == "nelsonii-S", light_intensity == "2000")
-
-ggplot(df1, aes(gsw, A)) +
-  geom_point()
-
-# FUNCTION STARTS HERE
-# function(df1, min_n = 20, ...) {
-ranges = df1 |>
-  summarize(
-    min_gsw = min(gsw, na.rm = TRUE),
-    max_gsw = max(gsw, na.rm = TRUE),
-    .by = "curve_type"
-  )
-
-df2 = filter(df1, gsw >= max(ranges$min_gsw) &
-               gsw <= min(ranges$max_gsw))
-
-# Check for minimum number of points
-summarize(df2, n = n(), .by = curve_type)
-
-ggplot(df2, aes(gsw, A)) +
-  geom_point()

### 2025-07-23 ms/ms.qmd
cleaned up calculate-aa-posterior

@@
 #| tbl-cap: 'Leave-one-out cross-validation information criterion (LOOIC) for models of \\aax{} as a function of light intensity, light treatment, and population.
+Only the top models ($\\Delta~\\text{LOOIC} < 2$) are shown.
 The model with the lowest LOOIC is considered the best fit to the data.
@@
 df1 = aa_loo_table1 |>
+   filter(delta_looic < 2) |>
    mutate(
@@
 
-```{r, aa_loo2, eval=TRUE,echo=FALSE,message=FALSE,warning=FALSE}
+```{r, aa_loo2, results = "asis", eval=TRUE,echo=FALSE,message=FALSE,warning=FALSE}
 #| label: tbl-aa_loo2
 #| tbl-cap: 'Leave-one-out cross-validation information criterion (LOOIC) for models of \\aax{} as a function of light intensity, light treatment, leaf mass per area (LMA), and population.
+Only the top models ($\\Delta~\\text{LOOIC} < 2$) are shown.
 The model with the lowest LOOIC is considered the best fit to the data.
@@
 df1 = aa_loo_table2 |>
+  filter(delta_looic < 2) |>
   mutate(
@@
   str_detect(c1, "_lma") ~ "LMA",
-  str_detect(c1, "_intrxn12") ~ "Meas. x Growth",
-  str_detect(c1, "_intrxn13") ~ "Meas. x LMA",
-  str_detect(c1, "_intrxn23") ~ "Growth x LMA",
-  str_detect(c1, "_intrxn123") ~ "Meas. x Growth x LMA",
+  str_detect(c1, "_intrxn12") ~ "Meas. $\\times$ Growth",
+  str_detect(c1, "_intrxn13") ~ "Meas. $\\times$ LMA",
+  str_detect(c1, "_intrxn23") ~ "Growth $\\times$ LMA",
+  str_detect(c1, "_intrxn123") ~ "Meas. $\\times$ Growth $\\times$ LMA",
+  str_detect(c1, "_intrxn$") ~ "Meas. $\\times$ Growth",
   str_detect(c1, "_acc") ~ "Population",
@@
 )
+
+cat("\\begin{landscape}")
 
@@
   add_header_above(groupings) |>
+  column_spec(c(9, 12), border_right = TRUE) |>
   column_spec(1, width = "0.7in") |>
   column_spec(2, width = "0.5in") |>
-  column_spec(3:10, width = "0.4in") |>
-  column_spec(c(5, 8), border_right = TRUE)
-
-```
+  column_spec(3:14, width = "0.4in")
+cat("\\end{landscape}")
+```

### 2025-07-23 ms/ms.tex
cleaned up calculate-aa-posterior

@@
 % Make \paragraph and \subparagraph free-standing
+\makeatletter
 \ifx\paragraph\undefined\else
   \let\oldparagraph\paragraph
-  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
+  \renewcommand{\paragraph}{
+    \@ifstar
+      \xxxParagraphStar
+      \xxxParagraphNoStar
+  }
+  \newcommand{\xxxParagraphStar}[1]{\oldparagraph*{#1}\mbox{}}
+  \newcommand{\xxxParagraphNoStar}[1]{\oldparagraph{#1}\mbox{}}
 \fi
@@
 
-\includegraphics[width=2.24in,height=\textheight]{../figures/amphisomy draft 2024-03-28.jpg}
+\includegraphics[width=2.24in,height=\textheight,keepaspectratio]{../figures/amphisomy draft 2024-03-28.jpg}
 
@@
 
-\includegraphics{ms_files/figure-pdf/fig-aa-1.pdf}
+\pandocbounded{\includegraphics[keepaspectratio]{ms_files/figure-pdf/fig-aa-1.pdf}}
 
@@
 
-\includegraphics{ms_files/figure-pdf/fig-aacap-1.pdf}
+\pandocbounded{\includegraphics[keepaspectratio]{ms_files/figure-pdf/fig-aacap-1.pdf}}
 
@@
 
-\includegraphics{ms_files/figure-pdf/fig-lma_aa-1.pdf}
+\pandocbounded{\includegraphics[keepaspectratio]{ms_files/figure-pdf/fig-lma_aa-1.pdf}}
 
@@
 
-\includegraphics{ms_files/figure-pdf/fig-example-aa-1.pdf}
+\pandocbounded{\includegraphics[keepaspectratio]{ms_files/figure-pdf/fig-example-aa-1.pdf}}
 
@@
 criterion (LOOIC) for models of \aax{} as a function of light intensity,
-light treatment, and population. The model with the lowest LOOIC is
-considered the best fit to the data. The LOOIC column lists the LOOIC
+light treatment, and population. Only the top models
+(\(\Delta~\text{LOOIC} < 2\)) are shown. The model with the lowest LOOIC
+is considered the best fit to the data. The LOOIC column lists the LOOIC
 value for each model; the \(\Delta\)LOOIC column lists the difference
@@
 \cellcolor{gray!10}{-1868.68} & \cellcolor{gray!10}{1.12} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$}\\
--1867.89 & 1.92 &  & $\checkmark$ &  & $\checkmark$ & $\checkmark$ & $\checkmark$ & $\checkmark$ & $\checkmark$\\
-\cellcolor{gray!10}{-1867.44} & \cellcolor{gray!10}{2.36} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$}\\
-\addlinespace
--1867.06 & 2.74 & $\checkmark$ & $\checkmark$ &  & $\checkmark$ & $\checkmark$ & $\checkmark$ & $\checkmark$ & $\checkmark$\\
-\cellcolor{gray!10}{-1867.00} & \cellcolor{gray!10}{2.81} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$}\\
--1866.86 & 2.94 & $\checkmark$ & $\checkmark$ & $\checkmark$ & $\checkmark$ &  &  & $\checkmark$ & $\checkmark$\\
-\cellcolor{gray!10}{-1866.71} & \cellcolor{gray!10}{3.09} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$}\\
--1866.36 & 3.44 &  &  &  & $\checkmark$ & $\checkmark$ &  & $\checkmark$ & $\checkmark$\\
-\addlinespace
-\cellcolor{gray!10}{-1866.19} & \cellcolor{gray!10}{3.61} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$}\\
--1865.08 & 4.72 &  & $\checkmark$ &  & $\checkmark$ &  &  & $\checkmark$ & $\checkmark$\\
-\cellcolor{gray!10}{-1864.70} & \cellcolor{gray!10}{5.10} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$}\\
--1864.47 & 5.34 & $\checkmark$ & $\checkmark$ & $\checkmark$ &  & $\checkmark$ &  & $\checkmark$ & $\checkmark$\\
-\cellcolor{gray!10}{-1863.52} & \cellcolor{gray!10}{6.28} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$}\\
-\addlinespace
--1863.27 & 6.54 & $\checkmark$ & $\checkmark$ & $\checkmark$ &  &  &  & $\checkmark$ & $\checkmark$\\
-\cellcolor{gray!10}{-1862.59} & \cellcolor{gray!10}{7.22} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$}\\
--1862.42 & 7.39 & $\checkmark$ & $\checkmark$ &  &  & $\checkmark$ &  & $\checkmark$ & $\checkmark$\\
-\cellcolor{gray!10}{-1862.06} & \cellcolor{gray!10}{7.74} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$}\\
--1861.93 & 7.87 & $\checkmark$ &  &  & $\checkmark$ & $\checkmark$ & $\checkmark$ &  & $\checkmark$\\
-\addlinespace
-\cellcolor{gray!10}{-1861.71} & \cellcolor{gray!10}{8.10} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$}\\
--1861.45 & 8.36 & $\checkmark$ & $\checkmark$ & $\checkmark$ & $\checkmark$ & $\checkmark$ &  &  & $\checkmark$\\
-\cellcolor{gray!10}{-1860.69} & \cellcolor{gray!10}{9.11} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$}\\
--1860.41 & 9.39 & $\checkmark$ & $\checkmark$ &  & $\checkmark$ & $\checkmark$ & $\checkmark$ &  & $\checkmark$\\
-\cellcolor{gray!10}{-1859.96} & \cellcolor{gray!10}{9.84} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$}\\
-\addlinespace
--1858.68 & 11.12 & $\checkmark$ & $\checkmark$ & $\checkmark$ & $\checkmark$ &  &  &  & $\checkmark$\\
-\cellcolor{gray!10}{-1858.53} & \cellcolor{gray!10}{11.27} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$}\\
--1858.18 & 11.62 &  & $\checkmark$ &  & $\checkmark$ & $\checkmark$ &  &  & $\checkmark$\\
-\cellcolor{gray!10}{-1858.10} & \cellcolor{gray!10}{11.70} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$}\\
--1857.84 & 11.96 & $\checkmark$ & $\checkmark$ &  & $\checkmark$ & $\checkmark$ &  &  & $\checkmark$\\
-\addlinespace
-\cellcolor{gray!10}{-1857.64} & \cellcolor{gray!10}{12.16} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$}\\
--1857.06 & 12.75 & $\checkmark$ &  &  & $\checkmark$ &  &  & $\checkmark$ & $\checkmark$\\
-\cellcolor{gray!10}{-1856.48} & \cellcolor{gray!10}{13.32} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$}\\
--1856.40 & 13.40 & $\checkmark$ & $\checkmark$ & $\checkmark$ &  & $\checkmark$ &  &  & $\checkmark$\\
-\cellcolor{gray!10}{-1856.24} & \cellcolor{gray!10}{13.56} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{}\\
-\addlinespace
--1855.98 & 13.82 &  &  &  & $\checkmark$ &  &  & $\checkmark$ & $\checkmark$\\
-\cellcolor{gray!10}{-1855.89} & \cellcolor{gray!10}{13.92} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{}\\
--1855.37 & 14.43 & $\checkmark$ & $\checkmark$ & $\checkmark$ & $\checkmark$ & $\checkmark$ & $\checkmark$ & $\checkmark$ & \\
-\cellcolor{gray!10}{-1855.36} & \cellcolor{gray!10}{14.44} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$}\\
--1855.05 & 14.75 & $\checkmark$ & $\checkmark$ &  & $\checkmark$ & $\checkmark$ & $\checkmark$ & $\checkmark$ & \\
-\addlinespace
-\cellcolor{gray!10}{-1854.54} & \cellcolor{gray!10}{15.26} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$}\\
--1854.41 & 15.39 &  & $\checkmark$ &  & $\checkmark$ & $\checkmark$ & $\checkmark$ & $\checkmark$ & \\
-\cellcolor{gray!10}{-1853.87} & \cellcolor{gray!10}{15.94} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$}\\
--1853.55 & 16.25 & $\checkmark$ &  &  &  &  &  & $\checkmark$ & $\checkmark$\\
-\cellcolor{gray!10}{-1853.43} & \cellcolor{gray!10}{16.37} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{}\\
-\addlinespace
--1852.90 & 16.91 & $\checkmark$ &  &  & $\checkmark$ & $\checkmark$ &  & $\checkmark$ & \\
-\cellcolor{gray!10}{-1851.37} & \cellcolor{gray!10}{18.43} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{}\\
--1851.18 & 18.62 &  &  &  & $\checkmark$ & $\checkmark$ &  & $\checkmark$ & \\
-\cellcolor{gray!10}{-1851.03} & \cellcolor{gray!10}{18.77} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{}\\
--1851.03 & 18.77 &  & $\checkmark$ &  & $\checkmark$ &  &  & $\checkmark$ & \\
-\addlinespace
-\cellcolor{gray!10}{-1850.89} & \cellcolor{gray!10}{18.91} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{}\\
--1850.69 & 19.12 & $\checkmark$ & $\checkmark$ &  & $\checkmark$ & $\checkmark$ &  & $\checkmark$ & \\
-\cellcolor{gray!10}{-1850.37} & \cellcolor{gray!10}{19.43} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{}\\
--1850.04 & 19.76 & $\checkmark$ & $\checkmark$ &  & $\checkmark$ &  &  & $\checkmark$ & \\
-\cellcolor{gray!10}{-1849.45} & \cellcolor{gray!10}{20.35} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{}\\
-\addlinespace
--1848.68 & 21.12 & $\checkmark$ &  &  &  & $\checkmark$ &  & $\checkmark$ & \\
-\cellcolor{gray!10}{-1848.64} & \cellcolor{gray!10}{21.16} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{}\\
--1847.73 & 22.07 & $\checkmark$ & $\checkmark$ & $\checkmark$ &  &  &  & $\checkmark$ & \\
-\cellcolor{gray!10}{-1847.72} & \cellcolor{gray!10}{22.08} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{}\\
--1847.47 & 22.33 & $\checkmark$ & $\checkmark$ &  & $\checkmark$ & $\checkmark$ & $\checkmark$ &  & \\
-\addlinespace
-\cellcolor{gray!10}{-1847.40} & \cellcolor{gray!10}{22.40} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$}\\
--1847.36 & 22.44 & $\checkmark$ &  &  & $\checkmark$ &  &  &  & $\checkmark$\\
-\cellcolor{gray!10}{-1847.14} & \cellcolor{gray!10}{22.66} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{}\\
--1846.28 & 23.52 & $\checkmark$ & $\checkmark$ & $\checkmark$ & $\checkmark$ & $\checkmark$ &  &  & \\
-\cellcolor{gray!10}{-1845.82} & \cellcolor{gray!10}{23.98} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{}\\
-\addlinespace
--1845.58 & 24.22 & $\checkmark$ & $\checkmark$ & $\checkmark$ & $\checkmark$ &  &  &  & \\
-\cellcolor{gray!10}{-1845.55} & \cellcolor{gray!10}{24.26} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{}\\
--1845.31 & 24.49 & $\checkmark$ & $\checkmark$ &  &  &  &  & $\checkmark$ & \\
-\cellcolor{gray!10}{-1845.29} & \cellcolor{gray!10}{24.51} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$}\\
--1845.28 & 24.52 & $\checkmark$ &  &  & $\checkmark$ & $\checkmark$ &  &  & \\
-\addlinespace
-\cellcolor{gray!10}{-1844.09} & \cellcolor{gray!10}{25.71} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{}\\
--1844.00 & 25.80 & $\checkmark$ & $\checkmark$ &  & $\checkmark$ & $\checkmark$ &  &  & \\
-\cellcolor{gray!10}{-1843.72} & \cellcolor{gray!10}{26.08} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{}\\
--1841.91 & 27.89 & $\checkmark$ & $\checkmark$ & $\checkmark$ &  & $\checkmark$ &  &  & \\
-\cellcolor{gray!10}{-1841.65} & \cellcolor{gray!10}{28.16} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{}\\
-\addlinespace
--1841.19 & 28.61 &  &  &  & $\checkmark$ &  &  & $\checkmark$ & \\
-\cellcolor{gray!10}{-1840.69} & \cellcolor{gray!10}{29.12} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{}\\
--1839.82 & 29.98 & $\checkmark$ & $\checkmark$ &  &  &  &  &  & \\
-\cellcolor{gray!10}{-1839.81} & \cellcolor{gray!10}{29.99} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{}\\
--1839.04 & 30.76 & $\checkmark$ & $\checkmark$ &  &  & $\checkmark$ &  &  & \\
-\addlinespace
-\cellcolor{gray!10}{-1838.36} & \cellcolor{gray!10}{31.44} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{}\\
--1834.02 & 35.78 &  &  &  &  & $\checkmark$ &  & $\checkmark$ & $\checkmark$\\
-\cellcolor{gray!10}{-1833.87} & \cellcolor{gray!10}{35.93} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{}\\
--1833.86 & 35.94 & $\checkmark$ &  &  & $\checkmark$ &  &  &  & \\
-\cellcolor{gray!10}{-1833.29} & \cellcolor{gray!10}{36.51} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$}\\
-\addlinespace
--1832.82 & 36.99 &  & $\checkmark$ &  &  & $\checkmark$ &  & $\checkmark$ & $\checkmark$\\
-\cellcolor{gray!10}{-1829.41} & \cellcolor{gray!10}{40.39} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{}\\
--1826.86 & 42.94 &  &  &  &  & $\checkmark$ &  &  & $\checkmark$\\
-\cellcolor{gray!10}{-1826.67} & \cellcolor{gray!10}{43.13} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$}\\
--1825.19 & 44.61 &  & $\checkmark$ &  &  & $\checkmark$ &  &  & $\checkmark$\\
-\addlinespace
-\cellcolor{gray!10}{-1824.95} & \cellcolor{gray!10}{44.86} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$}\\
--1818.14 & 51.66 &  &  &  &  &  &  &  & $\checkmark$\\
-\cellcolor{gray!10}{-1814.64} & \cellcolor{gray!10}{55.16} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{}\\
--1814.51 & 55.29 &  &  &  &  & $\checkmark$ &  & $\checkmark$ & \\
-\cellcolor{gray!10}{-1814.23} & \cellcolor{gray!10}{55.57} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{}\\
-\addlinespace
--1808.39 & 61.41 &  &  &  &  &  &  & $\checkmark$ & \\
-\cellcolor{gray!10}{-1808.13} & \cellcolor{gray!10}{61.67} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{}\\
--1807.22 & 62.58 &  & $\checkmark$ &  &  & $\checkmark$ &  &  & \\
-\cellcolor{gray!10}{-1806.92} & \cellcolor{gray!10}{62.88} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{}\\
--1799.31 & 70.49 &  &  &  &  &  &  &  & \\*
+-1867.89 & 1.92 &  & $\checkmark$ &  & $\checkmark$ & $\checkmark$ & $\checkmark$ & $\checkmark$ & $\checkmark$\\*
 
@@
 
-\begin{table}
+\begin{landscape}
+
+\begin{longtable}[t]{>{\centering\arraybackslash}p{0.7in}>{\centering\arraybackslash}p{0.5in}>{\centering\arraybackslash}p{0.4in}>{\centering\arraybackslash}p{0.4in}>{\centering\arraybackslash}p{0.4in}>{\centering\arraybackslash}p{0.4in}>{\centering\arraybackslash}p{0.4in}>{\centering\arraybackslash}p{0.4in}>{\centering\arraybackslash}p{0.4in}>{\centering\arraybackslash}p{0.4in}>{\centering\arraybackslash}p{0.4in}>{\centering\arraybackslash}p{0.4in}>{\centering\arraybackslash}p{0.4in}>{\centering\arraybackslash}p{0.4in}}
 
@@
 criterion (LOOIC) for models of \aax{} as a function of light intensity,
-light treatment, leaf mass per area (LMA), and population. The model
-with the lowest LOOIC is considered the best fit to the data. The LOOIC
-column lists the LOOIC value for each model; the \(\Delta\)LOOIC column
-lists the difference between the LOOIC of the corresponding model and
-the best model (lowest LOOIC). Checkmarks in each row indicate whether
-the corresponding model included affects of measurement light intensity
+light treatment, leaf mass per area (LMA), and population. Only the top
+models (\(\Delta~\text{LOOIC} < 2\)) are shown. The model with the
+lowest LOOIC is considered the best fit to the data. The LOOIC column
+lists the LOOIC value for each model; the \(\Delta\)LOOIC column lists
+the difference between the LOOIC of the corresponding model and the best
+model (lowest LOOIC). Checkmarks in each row indicate whether the
+corresponding model included affects of measurement light intensity
 (``Meas.''), growth light intensity (``Growth''), or LMA as fixed
@@
 
-\centering{
-
-}
-
-\end{table}%
+\tabularnewline
+
+\toprule
+\multicolumn{2}{c}{ } & \multicolumn{7}{c}{Fixed} & \multicolumn{3}{c}{Random} & \multicolumn{2}{c}{Distributional} \\
+\cmidrule(l{3pt}r{3pt}){3-9} \cmidrule(l{3pt}r{3pt}){10-12} \cmidrule(l{3pt}r{3pt}){13-14}
+LOOIC & $\Delta$LOOIC & Meas. & Growth & LMA & Meas. $\times$ Growth & Meas. $\times$ Growth & Meas. $\times$ LMA & Growth $\times$ LMA & Meas. & Growth & Meas. $\times$ Growth & Meas. & Growth\\
+\midrule
+\endfirsthead
+\multicolumn{14}{@{}l}{\textit{(continued)}}\\
+\toprule
+LOOIC & $\Delta$LOOIC & Meas. & Growth & LMA & Meas. $\times$ Growth & Meas. $\times$ Growth & Meas. $\times$ LMA & Growth $\times$ LMA & Meas. & Growth & Meas. $\times$ Growth & Meas. & Growth\\
+\midrule
+\endhead
+
+\endfoot
+\bottomrule
+\endlastfoot
+\cellcolor{gray!10}{-1872.05} & \cellcolor{gray!10}{0.00} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$}\\
+-1871.15 & 0.89 & $\checkmark$ &  & $\checkmark$ &  &  &  &  & $\checkmark$ &  &  & $\checkmark$ & $\checkmark$\\
+\cellcolor{gray!10}{-1871.03} & \cellcolor{gray!10}{1.01} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$}\\
+-1871.03 & 1.02 &  &  & $\checkmark$ &  &  &  &  & $\checkmark$ &  &  & $\checkmark$ & $\checkmark$\\
+\cellcolor{gray!10}{-1870.79} & \cellcolor{gray!10}{1.26} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$}\\
+\addlinespace
+-1870.74 & 1.31 &  &  & $\checkmark$ &  &  &  &  & $\checkmark$ & $\checkmark$ & $\checkmark$ & $\checkmark$ & $\checkmark$\\
+\cellcolor{gray!10}{-1870.06} & \cellcolor{gray!10}{1.99} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$}\\*
+
+\end{longtable}
+
+\end{landscape}
 

### 2025-07-23 r/202_calculate-aa-posterior.R

@@ -17,10 +17,6 @@ trimmed_rh_curves = read_rds("data/trimmed_rh_curves.rds") |>
 
 curve_fits_summary = read_rds("objects/curve-fits-summary.rds")
 
-# I need to decide if any of these require re-fitting
-curve_fits_summary |>
-  filter(rhat > 1.05 | ess_bulk < 100)
-
 # Draws from the curve fits ----
 curve_fits_draws = read_rds("objects/curve-fits-draws.rds") |>
   rename(b0 = b_Intercept, b1 = b_polylog_gsw2rawEQTRUE1, b2 = b_polylog_gsw2rawEQTRUE2) |>
@@ -63,17 +59,6 @@ aa_post = full_join(curve_fits_draws,
     aa = (upper_int - lower_int) / (log_gsw_pseudohypo - log_gsw_amphi)
   )
 
-# Export figures ----
-tmp = full_join(curve_fits_draws,
-          trimmed_rh_curves,
-          by = join_by(acc_id, leaf_type, light_intensity)) |>
-  mutate(log_gsw = min_log_gsw * (leaf_type == "amphi") + max_log_gsw * (leaf_type == "pseudohypo")) |>
-  filter(.draw < 11, acc == "LA0107") |>
-  split(~ acc_id)
-
-tmp$max_log_gsw
-
-
 # Check that AA estimates converged ----
 aa_summary = aa_post |>
   dplyr::select(starts_with("."), acc_id, aa, light_intensity) |>
@@ -86,10 +71,6 @@ aa_summary = aa_post |>
       mutate(acc_id = .x$acc_id[1],
              light_intensity = .x$light_intensity[1])
   }, .progress = TRUE)
-
-# I need to decide if any of these require re-fitting
-aa_summary |>
-  filter(rhat > 1.05 | ess_bulk < 100)
 
 # Write posterior of AA estimates ----
 write_rds(aa_post, "objects/aa_post.rds")

### 2025-07-24 ms/ms.tex
cont reorg and rerun

@@
 was consistently greater for sun plants. The average \aax{} among
-populations in the shade treatment was 0.041 (range: 0.010--0.113; 21 of
+populations in the shade treatment was 0.041 (range: 0.007--0.113; 19 of
 29 populations significant); however, the same populations grown at high
-light intensity showed a mean \aax{} of 0.053 (range: 0.021--0.125; 19
+light intensity showed a mean \aax{} of 0.052 (range: 0.020--0.120; 20
 of 29 populations significant). Contrary to the predictions of the
@@
 plants, where \aax{} was significantly greater under low measurement
-light intensity in 11 of 29 populations compared to 4 populations for
+light intensity in 13 of 29 populations compared to 12 populations for
 shade-grown plants. The overall average \aax{} of shade and sun grown
 plants measured under low light intensity was 0.064 (range:
-0.017--0.132; 28 of 29 populations significant) and 0.094 (range:
-0.036--0.197; 24 of 29 populations significant), respectively. There was
+0.022--0.137; 28 of 29 populations significant) and 0.100 (range:
+0.049--0.206; 27 of 29 populations significant), respectively. There was
 a modest tendency for populations from more open habitats (lower PAI) to
@@
 \aax{}. LMA increased in sun grown plants in all populations by an
-average of 119\% {[}95\% CI: 47 to 240\%{]}, similar to plastic
+average of 123\% {[}95\% CI: 42.9 to 256\%{]}, similar to plastic
 responses in many species (\emph{48}). While LMA is weakly, albeit
@@
 \cellcolor{gray!10}{3. remove outliers within each \agcurve{} curve} & \cellcolor{gray!10}{2,360} & \cellcolor{gray!10}{58.7}\\
-4. remove replicates with no overlap between amphi and pseudohypo \agcurve{} curves & 2,268 & 58.5\\
+4. remove replicates with no overlap between amphi and pseudohypo \agcurve{} curves & 2,268 & 58.4\\
 \cellcolor{gray!10}{5. thin redundant data points within each \agcurve{} curve} & \cellcolor{gray!10}{2,268} & \cellcolor{gray!10}{28.1}\\
@@
 \emph{CmdStan} version 2.36.0 and \textbf{cmdstanr} version 0.9.0
-(\emph{61}) to interface with \emph{R} version 4.5.0 (\emph{62}). We
+(\emph{61}) to interface with \emph{R} version 4.5.1 (\emph{62}). We
 sampled the posterior distribution from a single chain with a minimum of
@@
 \endlastfoot
-\cellcolor{gray!10}{-1869.80} & \cellcolor{gray!10}{0.00} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$}\\
--1868.94 & 0.86 & $\checkmark$ & $\checkmark$ & $\checkmark$ & $\checkmark$ & $\checkmark$ & $\checkmark$ & $\checkmark$ & $\checkmark$\\
-\cellcolor{gray!10}{-1868.68} & \cellcolor{gray!10}{1.12} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$}\\
--1867.89 & 1.92 &  & $\checkmark$ &  & $\checkmark$ & $\checkmark$ & $\checkmark$ & $\checkmark$ & $\checkmark$\\*
+\cellcolor{gray!10}{-1882.56} & \cellcolor{gray!10}{0.00} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$}\\
+-1881.70 & 0.86 &  &  &  & $\checkmark$ & $\checkmark$ & $\checkmark$ & $\checkmark$ & $\checkmark$\\
+\cellcolor{gray!10}{-1881.00} & \cellcolor{gray!10}{1.56} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$}\\
+-1880.91 & 1.65 & $\checkmark$ & $\checkmark$ &  & $\checkmark$ & $\checkmark$ & $\checkmark$ & $\checkmark$ & $\checkmark$\\
+\cellcolor{gray!10}{-1880.63} & \cellcolor{gray!10}{1.93} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$}\\*
 
@@
 \endlastfoot
-\cellcolor{gray!10}{-1872.05} & \cellcolor{gray!10}{0.00} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$}\\
--1871.15 & 0.89 & $\checkmark$ &  & $\checkmark$ &  &  &  &  & $\checkmark$ &  &  & $\checkmark$ & $\checkmark$\\
-\cellcolor{gray!10}{-1871.03} & \cellcolor{gray!10}{1.01} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$}\\
--1871.03 & 1.02 &  &  & $\checkmark$ &  &  &  &  & $\checkmark$ &  &  & $\checkmark$ & $\checkmark$\\
-\cellcolor{gray!10}{-1870.79} & \cellcolor{gray!10}{1.26} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$}\\
+\cellcolor{gray!10}{-1883.84} & \cellcolor{gray!10}{0.00} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$}\\
+-1883.80 & 0.04 & $\checkmark$ &  & $\checkmark$ &  &  &  &  & $\checkmark$ & $\checkmark$ & $\checkmark$ & $\checkmark$ & $\checkmark$\\
+\cellcolor{gray!10}{-1883.67} & \cellcolor{gray!10}{0.17} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$}\\
+-1883.16 & 0.67 & $\checkmark$ &  & $\checkmark$ &  &  &  &  & $\checkmark$ &  &  & $\checkmark$ & $\checkmark$\\
+\cellcolor{gray!10}{-1883.05} & \cellcolor{gray!10}{0.78} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$}\\
 \addlinespace
--1870.74 & 1.31 &  &  & $\checkmark$ &  &  &  &  & $\checkmark$ & $\checkmark$ & $\checkmark$ & $\checkmark$ & $\checkmark$\\
-\cellcolor{gray!10}{-1870.06} & \cellcolor{gray!10}{1.99} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$}\\*
+-1883.05 & 0.78 & $\checkmark$ &  & $\checkmark$ &  &  &  & $\checkmark$ & $\checkmark$ & $\checkmark$ &  & $\checkmark$ & $\checkmark$\\
+\cellcolor{gray!10}{-1882.93} & \cellcolor{gray!10}{0.90} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$}\\
+-1882.87 & 0.97 & $\checkmark$ & $\checkmark$ & $\checkmark$ &  & $\checkmark$ &  &  & $\checkmark$ & $\checkmark$ & $\checkmark$ & $\checkmark$ & $\checkmark$\\
+\cellcolor{gray!10}{-1882.80} & \cellcolor{gray!10}{1.04} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$}\\
+-1882.55 & 1.29 & $\checkmark$ & $\checkmark$ & $\checkmark$ &  & $\checkmark$ &  &  & $\checkmark$ &  &  & $\checkmark$ & $\checkmark$\\
+\addlinespace
+\cellcolor{gray!10}{-1882.29} & \cellcolor{gray!10}{1.55} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$}\\
+-1882.27 & 1.56 & $\checkmark$ & $\checkmark$ & $\checkmark$ &  &  & $\checkmark$ &  & $\checkmark$ & $\checkmark$ & $\checkmark$ & $\checkmark$ & $\checkmark$\\
+\cellcolor{gray!10}{-1882.27} & \cellcolor{gray!10}{1.56} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$}\\
+-1882.08 & 1.76 &  &  & $\checkmark$ &  &  &  &  & $\checkmark$ & $\checkmark$ & $\checkmark$ & $\checkmark$ & $\checkmark$\\
+\cellcolor{gray!10}{-1881.86} & \cellcolor{gray!10}{1.98} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$}\\*
 

### 2025-07-24 r/11_plot-aa.R

@@ -87,7 +87,7 @@ df_aa_text = df_aa_pred1 |>
   ungroup() |>
   summarize(
     x = 0,
-    label = round(mean(aa), 3),
+    label = sprintf("%.3f", round(mean(aa), 3)),
     aa = 0.30,
     .by = c("Growth", "Measurement")
   )
@@ -137,22 +137,22 @@ write_rds(df_aa_pred2, "objects/df_aa_pred2.rds")
 write_rds(df_aa_pred3, "objects/df_aa_pred3.rds")
 
 # Plot for amphistomy-proposal
-set.seed(07102025)
-amphi_advantage = ggplot(df_aa_pred1, aes(light_intensity, aa, color = light_treatment, shape = light_intensity)) +
-  facet_wrap(~Growth) +
-  geom_jitter(width = 0.1, height = 0, size = 2, fill = "white") +
-  stat_summary(
-    fun.data = mean_cl_boot,
-    fun.args = list(conf.int = 0.95),
-    geom = "pointinterval",
-    position = position_nudge(x = -0.3),
-    fill = "white", size = 12, linewidth = 1
-  ) +
-  geom_hline(yintercept = 0, linetype = "dashed") +
-  scale_color_manual(values = c("low" = "tomato4", "high" = "tomato")) +
-  scale_shape_manual(values = c("150" = 19, "2000" = 21)) +
-  xlab(expression(Light~intensity~(mu*mol~m^-2~s^-1))) +
-  ylab("Amphistomy advantage") +
-  theme(legend.position = "none", strip.background = element_blank())
-
-write_rds(amphi_advantage, "objects/amphi_advantage.rds")
+# set.seed(07102025)
+# amphi_advantage = ggplot(df_aa_pred1, aes(light_intensity, aa, color = light_treatment, shape = light_intensity)) +
+#   facet_wrap(~Growth) +
+#   geom_jitter(width = 0.1, height = 0, size = 2, fill = "white") +
+#   stat_summary(
+#     fun.data = mean_cl_boot,
+#     fun.args = list(conf.int = 0.95),
+#     geom = "pointinterval",
+#     position = position_nudge(x = -0.3),
+#     fill = "white", size = 12, linewidth = 1
+#   ) +
+#   geom_hline(yintercept = 0, linetype = "dashed") +
+#   scale_color_manual(values = c("low" = "tomato4", "high" = "tomato")) +
[…]

### 2025-07-24 r/12_plot-pai.R

@@ -107,7 +107,6 @@ fig_aa_pai = ggplot(df_aa_pred2, aes(pai, aa, color = Growth, shape = Measuremen
   ylim(0, 0.2) +
   geom_hline(yintercept = 0, linetype = "dashed") +
   theme(legend.position = "none")
-fig_aa_pai
 
 write_rds(fig_aa_pai, "objects/fig_aa_pai.rds")
 

### 2025-07-24 r/13_plot-lma.R

@@ -90,68 +90,3 @@ fig_lma_aa = ggplot(df1, aes(x = lma, y = aa, shape = Measurement)) +
 
 write_rds(fig_lma_aa, "objects/fig_lma_aa.rds")
 ggsave("figures/lma-aa.pdf", fig_lma_aa, width = 6, height = 4)
-
-
-# other code. not using at the moment
-fit_aa$data |>
-  mutate(
-    lma = exp(llma),
-    Growth = light_treatment |>
-      factor(levels = c("low", "high")) |>
-      fct_recode(sun = "high", shade = "low"),
-    Measurement = light_intensity |>
-      factor(levels = c("150", "2000")) |>
-      fct_recode(low = "150", high = "2000")
-  ) |>
-  ggplot(aes(
-    x = lma,
-    y = aa,
-    color = Growth,
-    shape = Measurement
-  )) +
-  geom_ribbon(
-    data = df2,
-    mapping = aes(ymin = .lower, ymax = .upper, group = Measurement),
-    alpha = 0.5
-  ) +
-  geom_line(data = df2,
-            aes(group = Measurement),
-            linewidth = 2) +
-  # geom_interval(aes(ymin = aa_lower, ymax = aa_upper), linewidth = 0.5) +
-  # geom_interval(aes(xmin = lma_lower, xmax = lma_upper), linewidth = 0.5) +
-  geom_point(fill = "white") +
-  scale_color_manual(values = c("shade" = "tomato4", "sun" = "tomato")) +
-  scale_shape_manual(values = c("low" = 19, "high" = 21)) +
-  labs(x = expression(paste("leaf mass per area [g ", m^-2, "]")), , y = "amphi advantage") +
-  scale_x_log10()
-
-
-a1 = predict(fit_aa) |>
-  as_tibble()
-
-a1 |>
-  dplyr::select(starts_with("Estimate")) |>
-  rename_with(\(.x) str_remove(.x, "^Estimate\\."), .cols = everything()) |>
-  bind_cols(dplyr::select(fit_aa$data, light_intensity, light_treatment)) |>
-  mutate(
-    lma = exp(llma),
-    Growth = light_treatment |>
-      factor(levels = c("low", "high")) |>
-      fct_recode(sun = "high", shade = "low"),
-    Measurement = light_intensity |>
-      factor(levels = c("150", "2000")) |>
-      fct_recode(low = "150", high = "2000")
-  ) |>
[…]

### 2025-07-24 r/17_plot-example-aa.R

@@ -1,7 +1,4 @@
 source("r/header.R")
-# library(gridExtra)
-# library(grid)
-# library(gridtext)
 
 scenarios = c("No amphistomy\nadvantage",
               "Significant amphistomy\nadvantage")

### 2025-07-24 ms/manuscraps.qmd
editing main text

@@
 ### MANUSCRAPS
+
+from abstract: Developmental plasticity in leaf thickness and/or density, an adaptive response to optimize light interception, likely affects CO$_2$ diffusion within the leaf, altering amphistomy advantage as a byproduct.
+
+From plasticity hypothesis: Secondarily, $\aasun{}$ and \gmaxsun{} should also be positively associated with native light habitat if genotypes adapted to sunny, open habitats if they can express a phenotype best adapted to that environment when leaves develop under high light intensity.
 
@@
 After $A$ and \gsw{} stabilized, we measured $A$ at 20 light-intensity levels between $0$ and $\qty{2000}{\micro\mol\raiseto{-2}\meter\raiseto{-1}\second}$ in descending order.
+
+### From leaf traits
+
+ We measured chlorophyll concentration index (CCI) using a chrolophyll concentration meter (MC-100, Apogee Instruments, Logan, Utah, USA) on the lamina of focal leaflets before gas exchange measurements at the same time we measured leaf temperature.
+
+ ### From hypothesis section
+
+ A one-dimensional circuit model using the Farquhar-von Caemmerer-Berry biochemical model of C$_3$ photosynthesis [@farquhar_biochemical_1980] shows that both increased stomatal conductance and Rubisco activity should increase \aax{}, all else being equal (Supporting Information)
+
+ ### results
+
+ The effect of low light on \aax{} was more pronounced in the sun-grown plants, where \aax{} was significantly greater under low measurement light intensity in `r aa_summary3["high"]` of `r n_acc` populations compared to `r aa_summary3["low"]` populations for shade-grown plants.

### 2025-07-24 ms/ms.qmd
editing main text

@@
 ---
-title: Developmental plasticity to high light intensity amplifies the advantage of amphistomous leaves
+title: Plasticity and adaptation to high light intensity amplify the advantage of amphistomous leaves
 format:
@@
   \newcommand{\cabetween}[2]{#1 < $C_\mathrm{a} < \qty{#2}{\micro\mol\raiseto{-1}\mol}$}
+  \newcommand{\gias}{$g_\text{ias}$}
+  \newcommand{\gmax}{$g_\text{max}$}
   \newcommand{\gmaxratio}{$g_\text{max,ratio}$}
@@
 *Lycopersicoides*).
-Plant grown under high light intensity benefit more in terms of CO$_2$ assimilated for a given stomatal conductance than plants grown under low light, regardless of measurement light intensity or native light habitat.
-Developmental plasticity in leaf thickness and/or density, an adaptive response to optimize light interception, likely affects CO$_2$ diffusion within the leaf, altering amphistomy advantage as a byproduct.
-Developmental plasticity may therefore play an important and previously unnoticed role in explaining the adaptive significance of amphistomy and the distribution of amphistomatous leaves in nature.
+Plants grown under high light intensity benefit more in terms of CO$_2$ assimilated for a given stomatal conductance than plants grown under low light, regardless of measurement light intensity.
+Furthermore, populations native to open habitats tend to benefit more from amphistomy compared to those from more closed habitats.
+Developmental plasticity and adaptation of leaves to high light intensity likely play a combined role in explaining the adaptive significance and distribution of amphistomy in nature.
 
@@
 Stomata are microscopic pores on the surfaces of leaves and other photosynthetic organs formed by a pair of guard cells.
-They are essential for balancing carbon gained per unit water lost and permitted vascular plants to grow tall on land by enabling access to CO$_2$ for photosynthetic carbon assimilation while preventing hydraulic failure in variable environments [@raven_selection_2002; @mcadam_stomata_2021; @clark_origin_2022].
+They are essential for balancing carbon gained per unit water lost and were an essential innovation in vascular plants, permitting them to grow tall on land by enabling access to CO$_2$ for photosynthetic carbon assimilation while preventing hydraulic failure in variable environments [@raven_selection_2002; @mcadam_stomata_2021; @clark_origin_2022].
 Optimal stomatal function depends on both dynamic changes in aperture on the scale of minutes to hours, as well as static anatomy determined by developmental plasticity and constitutive genetic differences [@hetherington_role_2003; @de_boer_optimal_2016; @harrison_influence_2020].
@@
 Amphistomatous leaves develop abaxial and adaxial stomata whose aperture can be independently regulated [@pospisilova_environmental_1980; @mott_stomatal_1984; @reich_effects_1985; @mott_asymmetric_1993; @wall_stomata_2022] to control gas exchange through each surface.
-All else being equal, gas exchange through stomata on both surfaces increases CO$_2$ supply to chloroplasts by providing a second parallel pathway through leaf intercellular airspaces, enhancing photosynthesis [@parkhurst_adaptive_1978; @gutschick_photosynthesis_1984].
+All else being equal, simultaneous gas exchange through stomata on both surfaces increases CO$_2$ supply to chloroplasts by providing a second parallel pathway through leaf intercellular airspaces, enhancing photosynthesis [@parkhurst_adaptive_1978; @gutschick_photosynthesis_1984].
 The extent to which amphistomy increases CO$_2$ supply depends on resistance to diffusion in intercellular airspaces.
 This resistance can be low in thin, porous, amphistomatous leaves [@mott_stomatal_1984; @marquez_assessing_2023], but may be more substantial in thick, dense, hypostomatous leaves [@parkhurst_intercellular_1990].
+We refer to the intercellular airspace conductance (\gias), the inverse of resistance.
 Amphistomatous leaves also lose more water through evaporation because of a second boundary layer conductance [@foster_influence_1986], but the additional carbon gain should be enough to offset this cost in most realistic scenarios [@muir_is_2019].
@@
 Many vertically oriented and/or isobilateral leaves are amphistomatous [@wood_physiology_1934; @howell_concerning_1945, @drake_two_2019].
-But among dorsiventral leaves, herbaceous plants in open, high light habitats tend to have have amphistomatous leaves [@salisbury_i_1928; @peat_comparative_1994; @gibson_structure-function_1996; @jordan_using_2014; @bucher_stomatal_2017; @muir_light_2018; @triplett_stomatal_2025].
+But among dorsiventral leaves, it is primarily herbaceous plants in open, high light habitats that tend to have amphistomatous leaves [@salisbury_i_1928; @peat_comparative_1994; @gibson_structure-function_1996; @jordan_using_2014; @bucher_stomatal_2017; @muir_light_2018; @triplett_stomatal_2025].
 Most other leaves, except those from aquatic habitats, are hypostomatous, producing stomata only on the lower, abaxial surface.
@@
 The overarching hypothesis is that leaves with greater stomatal density ratio are more common in open, sunny habitats because they increase photosynthesis most in those circumstances.
-An amphistomatous leaf increases photosynthetic carbon gain compared to an otherwise identical hypostomatous leaf by increasing conductance through the leaf intercellular airspaces and boundary layers; the additional water loss through a second boundary layer is typically small [@foster_influence_1986].
+An amphistomatous leaf increases photosynthetic carbon gain compared to an otherwise identical hypostomatous leaf by increasing conductance through the leaf intercellular airspaces and boundary layers.
 We quantify this benefit as the amphistomy advantage (\aax{}), the log-response ratio of photosynthesis in an amphistomatous leaf compared to an otherwise identical pseudohypostomatous leaf [@parkhurst_adaptive_1978; @triplett_amphistomy_2024].
 Why would \aax{} be greater in sun than shade?
-We consider three nonmutually exclusive hypotheses that we classify as 'acclimatory', 'plastic', and 'constitutive' ([Fig.
+We consider three nonmutually exclusive hypotheses that we classify as 'acclimation', 'plasticity', and 'constitutive' ([Fig.
 @fig-concept]).
 
-*Acclimatory hypothesis*: Photosynthetic induction to high light intensity typically involves increases in total leaf stomatal conductance (increased CO$_2$ supply), the concentration of active Rubisco, and electron transport capacity (increased CO$_2$ demand).
-A one-dimensional circuit model using the Farquhar-von Caemmerer-Berry biochemical model of C$_3$ photosynthesis [@farquhar_biochemical_1980] shows that both increased stomatal conductance and Rubisco activity should increase \aax{}, all else being equal (Supporting Information).
-If the acclimatory hypothesis is correct, we predict that $\aahigh{} > \aalow{}$ for all populations regardless of native habitat or growth environment.
+*Acclimation hypothesis*: Photosynthetic induction to high light intensity typically involves increases in total leaf stomatal conductance (increased CO$_2$ supply), the concentration of active Rubisco, and electron transport capacity (increased CO$_2$ demand).
+If the acclimation hypothesis is correct, we predict that $\aahigh{} > \aalow{}$ for all populations regardless of native habitat or growth environment.
 
@@
 Plastic changes in leaf anatomy and biochemistry could modulate \aax{} as a byproduct.
-Thicker or less porous leaves, both of which are associated with high leaf mass per area (LMA), will have lower $g_\mathrm{ias}$; leaves with increased total stomatal density and photosynthetic capacity have greater potential CO$_2$ supply and demand.
-Under the plastic hypothesis, we predict that $\aasun{} > \aashade{}$ for all populations and light intensities.
-Secondarily, $\aasun{}$ and \gmaxsun{} should also be positively associated with native light habitat if genotypes adapted to sunny, open habitats if they can express a phenotype best adapted to that environment when leaves develop under high light intensity.
+Thicker or less porous leaves, both of which are associated with high leaf mass per area (LMA), will have lower \gias; leaves with increased total stomatal density and photosynthetic capacity have greater potential CO$_2$ supply and demand.
+Under the plasticity hypothesis, we predict that $\aasun{} > \aashade{}$ for all populations and light intensities.
 
 *Constitutive hypothesis*: In environments that are relatively constant or where environmental change cannot be anticipated by a reliable cue, natural selection will favor constitutive expression of optimal phenotypes.
-We therefore predict genotypes from more sunny, open habitats will have consistently greater \aax under also measurement and growth light intensities.
+We therefore predict genotypes from more sunny, open habitats will have consistently greater \aax{} under all measurement and growth light intensities.
 For herbaceous plants, light intensity is largely a function of the tree canopy [@engelbrecht_evaluation_2001].
@@
 
-The primary directional predictions for each hypothesis are summarized in @tbl-hypotheses; detailed predictions for results that would indicate support for multiple hypotheses are in @tbl-predictions.
+The primary directional predictions for each hypothesis are summarized in @tbl-hypotheses; detailed predictions for results that would indicate simultaneous support for multiple hypotheses are in @tbl-predictions.
 
@@
   data.frame(
-    hypothesis = c("acclimatory", "plastic", "constitutive"),
+    hypothesis = c("acclimation", "plastic", "constitutive"),
     measurement = c(
@@
 <!-- Paragraph 5 (summarize methods) -->
-We tested these hypotheses by comparing \aax{} among amphistomatous wild tomato species [@peralta_taxonomy_2008] from different native light habitats, grown under simulated sun and shade light treatments, and measured under contrasting light intensities (low and high).
+We tested these hypotheses by comparing \aax{} among amphistomatous wild tomato species (@tbl-populations; @fig-sr-rxnorm; [@peralta_taxonomy_2008]) from different native light habitats, grown under simulated sun and shade light treatments, and measured under contrasting light intensities (low and high).
 We measured \aax{} on `r n_plant` individual plants from `r n_acc` populations (average of `r mean_n_rep` replicates per light treatment) using a recently developed method [@triplett_amphistomy_2024].
 With this method, we directly compare the photosynthetic rate of an untreated amphistomatous leaf to that of the same leaf with gas exchange blocked through the adaxial (upper) surface by transparent plastic, which we refer to as 'pseudohypostomy'.
-To compare amphi- and pseudohypostomatous leaves at identical whole-leaf $g_\text{sc}$, we measure $A$ over a range of $g_\text{sc}$, inducing stomatal opening and closure by modulating humidity (see Materials and Methods for further details).
+To compare amphi- and pseudohypostomatous leaves at identical whole-leaf stomatal conductance (\gsw), we measure $A$ over a range of \gsw, inducing stomatal opening and closure by modulating humidity (see Materials and Methods for further details).
 We estimated 'amphistomy advantage' (\aax) *sensu* @parkhurst_adaptive_1978, but with modifications previously described in @triplett_amphistomy_2024 and here (Materials and Methods).
-The native light intensity was represented by plant area index (PAI $\unit{\meter\squared\per\meter\squared}$), estimated using a global gridded data set derived from the Global Ecosystem Dynamics Investigation [GEDI; @burns_multi-resolution_2024] and georeferenced accession collection information from the Tomato Genetics Resource Center.
+The native light intensity was represented by plant area index (PAI $\unit{\meter\squared\per\meter\squared}$), estimated using a global gridded data set derived from the Global Ecosystem Dynamics Investigation [GEDI; @burns_multi-resolution_2024] and georeferenced accession collection information from the Tomato Genetics Resource Center (@tbl-populations).
 The growth light intensities were \ppfdequals{`r ppfd_sun`} (sun treatment) and \ppfdqty{`r ppfd_shade`} (shade treatment) while all other environment conditions were nearly identical (see Materials and Methods).
@@
 The average \aax{} among populations in the shade treatment was `r filter(aa_mean_range, light_treatment == "low", light_intensity == "2000")$x`; however, the same populations grown at high light intensity showed a mean \aax{} of `r filter(aa_mean_range, light_treatment == "high", light_intensity == "2000")$x`.
-Contrary to the predictions of the assimilatory hypothesis, \aax{} was greater in all populations under low measurement light intensity for both sun and shade grown plants.
-The effect of low light on \aax{} was more pronounced in the sun-grown plants, where \aax{} was significantly greater under low measurement light intensity in `r aa_summary3["high"]` of `r n_acc` populations compared to `r aa_summary3["low"]` populations for shade-grown plants.
+Contrary to the predictions of the acclimation hypothesis, \aax{} was greater in all populations under low measurement light intensity for both sun and shade grown plants.
 The overall average \aax{} of shade and sun grown plants measured under low light intensity was `r filter(aa_mean_range, light_treatment == "low", light_intensity == "150")$x` and `r filter(aa_mean_range, light_treatment == "high", light_intensity == "150")$x`, respectively.
@@
 <!-- Paragraph 7 (summarize secondary results) -->
-The pattern of \aax{} across wild tomatoes strongly supports the plasticity hypothesis, argues against the acclimatory hypothesis, and provides modest support for the constitutive hypothesis.
+The pattern of \aax{} across wild tomatoes strongly supports the plasticity hypothesis, contradicts the acclimation hypothesis, and provides modest support for the constitutive hypothesis.
 Plastic changes in leaf thickness and/or packing density, summarized by the bulk leaf mass per area (LMA), may mediate the effect of growth light intensity on \aax{}.
-LMA increased in sun grown plants in all populations by an average of `r signif(100 * (exp(post_lma["Estimate"]) - 1), 3)`% [95% CI: `r signif(100 * (exp(post_lma["Q2.5"]) - 1), 3)` to `r signif(100 * (exp(post_lma["Q97.5"]) - 1), 3)`%], similar to plastic responses in many species [@poorter_metaanalysis_2019].
+LMA increased in sun grown plants in all populations by an average of `r signif(100 * (exp(post_lma["Estimate"]) - 1), 3)`% [95% CI: `r signif(100 * (exp(post_lma["Q2.5"]) - 1), 3)` to `r signif(100 * (exp(post_lma["Q97.5"]) - 1), 3)`%], qualitatively similar to plastic responses in many species [@poorter_metaanalysis_2019].
 While LMA is weakly, albeit significantly, associated with individual-level \aax{} ([Fig.
@@
 The solid line is the posterior median of the linear regression of individual-level LMA on $\\aahigh$  with the shaded region showing the 95% confidence ribbon; the dashed line is the same for $\\aalow$.
-Confidence intervals for each population estimate and individual-level estimates are omitted for visual clarity."
+Confidence intervals for each population estimate were omitted for visual clarity."
 
@@
 
-We conclude that developmental plasticity, as opposed to acclimation or adaptation to open habitats, may explain the explain the long-standing observation that amphistomatous leaves are more common in sunny habitats, at least among herbaceous plants including crop relatives.
+We conclude that developmental plasticity, as opposed to acclimation or adaptation to open habitats, may explain the long-standing observation that amphistomatous leaves are more common in sunny habitats, at least among herbaceous plants including crop relatives.
 Our results advance existing explanations by showing high light intensity *per se* does not increase the benefit of amphistomy, but rather that anatomical and biochemical change associated with higher light intensity modulate \aax{}.
@@
 Under realistic assumptions, we estimate that hypostomatous leaves could achieve the same $A$ with X to X% less water loss if they were amphistomatous (Table S9).
-Ours is the first experimental demonstration that amphistomy is beneficial for sun leaves because of plastic changes, but future research needs to estimates the tradeoffs associated with amphistomatous leaves in order to fully explain their rarity among many woody plants and shade-tolerant herbs.
+Ours is the first experimental demonstration that amphistomy is beneficial for sun leaves because of plastic changes, but future research needs to estimate the tradeoffs associated with amphistomatous leaves in order to fully explain their rarity among many woody plants and shade-tolerant herbs.
 
@@
 #| tbl-cap: "Accession information of *Solanum* populations used in this study.
-The species name, accession number, collection latitude, longitude, elevation, and daily solar radiation estimated using the SPLASH algorithm (see 'Climate data').
-TGRC: Tomato Genetics Resource Center; PAI: Plant area index."
+The species name, accession number, collection latitude, longitude, elevation, and plant area index (PAI) from the Global Ecosystem Dynamics Investigation (see 'Climate data').
+TGRC: Tomato Genetics Resource Center."
 
@@
 We measured terminal leaflets in `r percent_terminal`\% of cases, but used the lateral leaflet closest to the terminal leaflet when it was damaged or difficult to clamp into the gas exchange chamber.
-When a leaflet was damaged during gas exchange measurements, we collected anatomical data from the nearest leaflet on the same leaf (`r percent_change_leaflet`)\%.
-We measured chlorophyll concentration index (CCI) using a chrolophyll concentration meter (MC-100, Apogee Instruments, Logan, Utah, USA) on the lamina of focal leaflets before gas exchange measurements at the same time we measured leaf temperature.
+When a leaflet was damaged during gas exchange measurements, we collected anatomical data from the nearest leaflet on the same leaf (`r percent_change_leaflet`\% of leaves).
 
@@
 The irradiance of the light source in the pseudohypo leaf was higher because the propafilm reduces transmission.
-To compensate for reduced transmission, we increased incident \ppfd for pseudohypo leaves by a factor 1/0.91, the inverse of the measured transmissivity of the propafilm.
+To compensate for reduced transmission, we increased incident \ppfd{} for pseudohypo leaves by a factor 1/0.91, the inverse of the measured transmissivity of the propafilm.
 We also set the stomatal conductance ratio, for purposes of calculating boundary layer conductance, to 0 for pseudohypo leaves following manufacturer directions.
@@
 We logged data until \gsw{} reached its nadir.
-We then acclimated the leaf to low light (\ppfdequals{150}) and \rhequals{70} before inducing stomatal closure with low \rh and logging data as described above.
+We then acclimated the leaf to low light (\ppfdequals{150}) and \rhequals{70} before inducing stomatal closure with low \rh{} and logging data as described above.
 
@@
 
-We estimated the stomatal density and size on ab- and adaxial leaf surfaces from all leaves, using guard cell length as a proxy for stomatal size since it proportional to maximum conductance [@sack_developmental_2016].
+We estimated the stomatal density and size on ab- and adaxial leaf surfaces from all leaves, using guard cell length as a proxy for stomatal size since it is proportional to maximum conductance [@sack_developmental_2016].
 We made surface impressions of leaf lamina from the same area used for gas exchange measurements using a-silicone impression material (Zhermack elite HD+, light body, fast set, Rovigo, Italy).
@@
 We counted and measured guard cell length on all stomata using the FIJI implementation of ImageJ2 version 2.3.0 [@schindelin_fiji_2012], then divided the count by the visible leaf area ($\qty{0.890}{\raiseto{2}\milli\meter}$) to estimate stomatal density.
+For each surface we calculated the anatomical maximum stomatal conductance (\gmax) at a reference leaf temperature of $\qty{25}{\degreeCelsius}$ following @sack_developmental_2016 as:
+
+$$g_\mathrm{max} = b m D_\mathrm{S} A_\mathrm{S}.$$
+
+The biophysical and morphological constants $b$ and $m$ are:
+
+$$b = \frac{D_\mathrm{wv}}{v},~\text{and}$$	
+$$m = \frac{\pi c ^ 2}{j^{0.5} (4 h j + \pi c)},$$
+
+where $D_\mathrm{wv}$ is the diffusion coefficient of water vapor in air and $v$ is by the kinematic viscosity of dry air.
+We assumed $D_\mathrm{wv} = \qty{2.49e-5}{\meter\squared\per\second}$ and $v = \qty{2.24e-2}{\raiseto{3}\meter\per\mol}$ [@monteith_principles_2013].
+For kidney-shaped guard cells like those in wild tomatoes, $c = h = j = 0.5$.
+The \gmaxratio{} is the \gmax{} of the adaxial surface divided by the sum total \gmax{} of both surfaces.
 
@@
 We dried leaves for 72 hours at $\qty{74}{\degreeCelsius}$ in a food dehydrator (Cosori CP267-FD, Vesync Co., Anaheim, California, USA) and weighed using a benchtop analytical balance (Ohaus PR64 Analytical Balance, Parsippany, New Jersey, USA).
-In $`r percent_lma_switch`\%$ we measured LMA on the adjacent leaflet because the focal leaflet was damaged or wilted while making surface impressions and we could not reliably estimate area.
+In $`r percent_lma_switch`\%$ of leaves we measured LMA on the adjacent leaflet because the focal leaflet was damaged or wilted while making surface impressions and we could not reliably estimate area.
 LMA data are missing from $`r percent_lma_missing`\%$ of individuals because the area or mass was not recorded at all or recorded incorrectly.
@@
 
-*Procedure*: We provionsally estimated \aax{} for each replicate by integrating over the range of \gsw{} values where amphi and pseudohypo \agcurve{} curves overlap.
+*Procedure*: We provisionally estimated \aax{} for each replicate by integrating over the range of \gsw{} values where amphi and pseudohypo \agcurve{} curves overlap.
 In this procedure, curve parameters were provisionally estimated using ordinary least squares with the `lm()` function in *R*.
@@
 2. Estimate \aax{} for both measurement light intensities using \agcurve{} curve parameters
-3. Estimate the effects of light intensity, light treatment, and population on \aax{} (assimilatory and plasticity hypotheses)
+3. Estimate the effects of light intensity, light treatment, and population on \aax{} (acclimation and plasticity hypotheses)
 4. Estimate the effects of native light habitat on population-level \aax{} (constitutive hypothesis)
@@
 We sampled the posterior distribution from a single chain with a minimum of `r aa_args$n_iter_init / 2` iterations after `r aa_args$n_iter_init / 2` warmup iterations.
-If necessary, we refit models with more iterations to decrease the Gelman-Rubin convergence statistic ($\hat{R}$) [@gelman_inference_1992] to less than $`aa_stats[["max_rhat"]]`$ and the bulk effective sample size (ESS) to greater than $`aa_stats[["min_ess"]]`$ for all model parameters, and there were fewer than $`aa_stats[["n_divergent"]]`$ divergent transitions.
+If necessary, we refit models with more iterations to decrease the Gelman-Rubin convergence statistic ($\hat{R}$) [@gelman_inference_1992] to less than `r aa_args[["max_rhat"]]` and the bulk effective sample size (ESS) to greater than `r aa_args[["min_ess"]]` for all model parameters, and there were fewer than `r aa_args[["max_divergent"]]` divergent transitions.
 
@@
 
-We modeled \logA as a quadratic function of \loggsw for each leaf, measured at low and high light intensity, in both amphistomatous (untreated) and pseudohypostomatous (propafilm treatment) states.
+We modeled \logA{} as a quadratic function of \loggsw{} for each leaf, measured at low and high light intensity, in both amphistomatous (untreated) and pseudohypostomatous (propafilm treatment) states.
 
@@
 
-The assimilatory, plastic, and constitutive hypotheses make different predictions about the relationship between \aax{} and light intensity, light treatment, and native \ppfd{} among populations.
+The acclimation, plasticity, and constitutive hypotheses make different predictions about the relationship between \aax{} and light intensity, light treatment, and native \ppfd{} among populations.
 Since these hypotheses are not mutually exclusive, we describe how we assessed support for one, two, or all three hypotheses simultaneously in \autoref{tbl-predictions}.
-In general, the assimilatory hypothesis was supported if \aax{} was greater at high light intensity than low light intensity.
+In general, the acclimation hypothesis was supported if \aax{} was greater at high light intensity than low light intensity.
 The plastic hypothesis was supported if \aax{} was greater in sun leaves than shade leaves.
@@
   ~"\\textbf{Hypothesis}", ~"\\textbf{Prediction(s)}", ~"\\textbf{Description}",
-  "Assimilatory", "$\\beta_{\\mathrm{AA},2000} > 0$", "\\aax{} at high light intensity is greater than that at low light intensity",
-  "Plastic", "$\\beta_{\\mathrm{AA},\\text{sun}} > 0$", "\\aax{} in sun leaves is greater than that in shade leaves",
-  "Constitutive", "$\\beta_{\\mathrm{PPFD,AA}} > 0$", "Population-level \\aax{} ($\\mathrm{AA}_\\text{acc}$) decreases with native PAI",
-  "Constitutive", "", "$\\mathrm{AA}_\\text{acc} = \\beta_{\\mathrm{AA}, 0} + \\vec{\\beta}_{\\mathrm{AA}, \\text{acc}}$",
-  "Assimilatory $\\times$ Plastic", "$\\beta_{\\mathrm{AA},2000,\\text{sun}} > 0$", "\\aax{} is highest at high light intensity in sun leaves",
-  "Assimilatory $\\times$ Plastic", "", "",
-  "Assimilatory $\\times$ Constitutive", "$\\beta_{\\mathrm{AA},2000} > 0$", "\\aax{} at high light intensity is greater than that at low light intensity",
-  "Assimilatory $\\times$ Constitutive", "$\\beta_{\\mathrm{PPFD,AA}} > 0$", "Population-level \\aax{} at high light intensity ($\\mathrm{AA}_{\\text{acc},2000}$) decreases with native PAI",
-  "Assimilatory $\\times$ Constitutive", "", "$\\mathrm{AA}_{\\text{acc},2000} = \\beta_{\\mathrm{AA}, 0} + \\vec{\\beta}_{\\mathrm{AA}, \\text{acc}} + \\vec{\\beta}_{\\mathrm{AA}, 2000, \\text{acc}}$",
-  "Plastic $\\times$ Constitutive", "$\\beta_{\\mathrm{AA},\\text{sun}} > 0$", "\\aax{} in sun leaves is greater than that in shade leaves",
-  "Plastic $\\times$ Constitutive", "$\\beta_{\\mathrm{PPFD,AA}} > 0$", "Population-level \\aax{} in sun leaves ($\\mathrm{AA}_{\\text{acc},\\text{sun}}$) decreases with native PAI",
-  "Plastic $\\times$ Constitutive", "", "$\\mathrm{AA}_{\\text{acc},\\text{sun}} = \\beta_{\\mathrm{AA}, 0} + \\vec{\\beta}_{\\mathrm{AA}, \\text{acc}} + \\vec{\\beta}_{\\mathrm{AA}, \\text{sun}, \\text{acc}}$",
-  "Assimilatory $\\times$ Plastic $\\times$ Constitutive", "$\\beta_{\\mathrm{AA},2000,\\text{sun}} > 0$", "\\aax{} is highest at high light intensity in sun leaves",
-  "Assimilatory $\\times$ Plastic $\\times$ Constitutive", "$\\beta_{\\mathrm{PPFD,AA}} > 0$", "Population-level \\aax{} at high light intensity in sun leaves ($\\mathrm{AA}_{\\text{acc},2000,\\text{sun}}$) decreases with native PAI",
-  "Assimilatory $\\times$ Plastic $\\times$ Constitutive", "", "$\\mathrm{AA}_{\\text{acc},2000,\\text{sun}} = \\beta_{\\mathrm{AA}, 0} + \\vec{\\beta}_{\\mathrm{AA}, \\text{acc}} + \\vec{\\beta}_{\\mathrm{AA}, 2000, \\text{acc}} + \\vec{\\beta}_{\\mathrm{AA}, \\text{sun}, \\text{acc}}$"
+  "Acclimation", "$\\mathrm{AA}_{2000} > 0$", "\\aax{} at high light intensity is greater than that at low light intensity",
+  "Plasticity", "$\\mathrm{AA}_{\\text{sun}} > 0$", "\\aax{} in sun leaves is greater than that in shade leaves",
+  "Constitutive", "$\\beta_{\\mathrm{PAI,AA}} > 0$", "Population-level \\aax{} ($\\mathrm{AA}_\\text{pop}$) decreases with native PAI",
+  "Constitutive", "", "$\\mathrm{AA}_\\text{pop} = \\beta_{\\mathrm{AA}, 0} + \\vec{\\beta}_{\\mathrm{AA}, \\text{pop}}$",
+  "Acclimation $\\times$ Plasticity", "$\\mathrm{AA}_{2000,\\text{sun}} > 0$", "\\aax{} is highest at high light intensity in sun leaves",
+  "Acclimation $\\times$ Plasticity", "", "",
+  "Acclimation $\\times$ Constitutive", "$\\mathrm{AA}_{2000} > 0$", "\\aax{} at high light intensity is greater than that at low light intensity",
+  "Acclimation $\\times$ Constitutive", "$\\beta_{\\mathrm{PAI,AA}} > 0$", "Population-level \\aax{} at high light intensity ($\\mathrm{AA}_{\\text{pop},2000}$) decreases with native PAI",
+  "Acclimation $\\times$ Constitutive", "", "$\\mathrm{AA}_{\\text{pop},2000} = \\beta_{\\mathrm{AA}, 0} + \\vec{\\beta}_{\\mathrm{AA}, \\text{pop}} + \\vec{\\beta}_{\\mathrm{AA}, 2000, \\text{pop}}$",
+  "Plasticity $\\times$ Constitutive", "$\\mathrm{AA}_{\\text{sun}} > 0$", "\\aax{} in sun leaves is greater than that in shade leaves",
+  "Plasticity $\\times$ Constitutive", "$\\beta_{\\mathrm{PAI,AA}} > 0$", "Population-level \\aax{} in sun leaves ($\\mathrm{AA}_{\\text{pop},\\text{sun}}$) decreases with native PAI",
+  "Plasticity $\\times$ Constitutive", "", "$\\mathrm{AA}_{\\text{pop},\\text{sun}} = \\beta_{\\mathrm{AA}, 0} + \\vec{\\beta}_{\\mathrm{AA}, \\text{pop}} + \\vec{\\beta}_{\\mathrm{AA}, \\text{sun}, \\text{pop}}$",
+  "Acclimation $\\times$ Plasticity $\\times$ Constitutive", "$\\mathrm{AA}_{2000,\\text{sun}} > 0$", "\\aax{} is highest at high light intensity in sun leaves",
+  "Acclimation $\\times$ Plasticity $\\times$ Constitutive", "$\\beta_{\\mathrm{PAI,AA}} > 0$", "Population-level \\aax{} at high light intensity in sun leaves ($\\mathrm{AA}_{\\text{pop},2000,\\text{sun}}$) decreases with native PAI",
+  "Acclimation $\\times$ Plasticity $\\times$ Constitutive", "", "$\\mathrm{AA}_{\\text{pop},2000,\\text{sun}} = \\beta_{\\mathrm{AA}, 0} + \\vec{\\beta}_{\\mathrm{AA}, \\text{pop}} + \\vec{\\beta}_{\\mathrm{AA}, 2000, \\text{pop}} + \\vec{\\beta}_{\\mathrm{AA}, \\text{sun}, \\text{pop}}$"
 
@@
 
-* Variation in stomatal ratio among pops and light treatments
-* Table S7: Model comparison LOOIC tables
-* Table S8: Model comparison LOOIC tables with LMA
+### MARKER - table of selected model 1
+
+### MARKER - table of selected model 2
+
 * Table S9: Calculation of alternate scenario where amphi leaves can get same A with less R
 * Each A-gs figure with fitted curves, R2, and AA
+
+```{r, sr-rxnorm, echo=FALSE,message=FALSE,warning=FALSE}
+#| label: fig-sr-rxnorm
+#| fig-align: center
+#| fig-width: 4
+#| fig-height: 4
+#| fig-cap: "Developmental plasticity in anatomical maximum stomatal conductance ratio (\\gmaxratio; $y$-axis) to growth light intensity ($x$-axis) among `r n_acc` wild tomato populations.
+All populations are amphistomatous, but vary in the stomatal density and size on each surface, which determines \\gmaxratio.
+Each point is the average per population in that treatment; black lines connect the same population across treatments."
+read_rds("../figures/sr-rxnorm.rds")
+```
 
@@
 #| tbl-cap: 'Leave-one-out cross-validation information criterion (LOOIC) for models of \\aax{} as a function of light intensity, light treatment, and population.
-Only the top models ($\\Delta~\\text{LOOIC} < 2$) are shown.
+Only the top models ($\\Delta \\text{LOOIC} < 2$) are shown.
 The model with the lowest LOOIC is considered the best fit to the data.
@@
 
-### MARKER - make other LOOIC table
-
 ```{r, aa_loo2, results = "asis", eval=TRUE,echo=FALSE,message=FALSE,warning=FALSE}
@@
 The LOOIC column lists the LOOIC value for each model; the $\\Delta$LOOIC column lists the difference between the LOOIC of the corresponding model and the best model (lowest LOOIC).
-Checkmarks in each row indicate whether the corresponding model included affects of measurement light intensity ("Meas."), growth light intensity ("Growth"), or LMA as fixed effects, random effects, or distributional effect on the residual variance.
+Checkmarks in each row indicate whether the corresponding model included affects of measurement light intensity ("Meas."), growth light intensity ("Growth"), or LMA as fixed or random effects.
 We test for fixed interaction effects between measurement and growth light intensity and LMA.
@@
     random_intrxn = str_detect(random, "\\*"),
-    sigma_meas = str_detect(sigma, "light_intensity"),
-    sigma_growth = str_detect(sigma, "light_treatment"),
+    # sigma_meas = str_detect(sigma, "light_intensity"),
+    # sigma_growth = str_detect(sigma, "light_treatment"),
   ) |>
@@
     starts_with("random_"),
-    starts_with("sigma_")
+    # starts_with("sigma_")
   ) |>
@@
   "Fixed" = sum(str_detect(colnames(df1), "fixed_")),
-  "Random" = sum(str_detect(colnames(df1), "random_")),
-  "Distributional" = sum(str_detect(colnames(df1), "sigma_"))
+  "Random" = sum(str_detect(colnames(df1), "random_"))
+  # "Distributional" = sum(str_detect(colnames(df1), "sigma_"))
 )
@@
   add_header_above(groupings) |>
-  column_spec(c(9, 12), border_right = TRUE) |>
   column_spec(1, width = "0.7in") |>
   column_spec(2, width = "0.5in") |>
-  column_spec(3:14, width = "0.4in")
+  column_spec(3:12, width = "0.4in")
 cat("\\end{landscape}")

### 2025-07-24 ms/ms.tex
editing main text

@@
 \newcommand{\cabetween}[2]{#1 < $C_\mathrm{a} < \qty{#2}{\micro\mol\raiseto{-1}\mol}$}
+\newcommand{\gias}{$g_\text{ias}$}
+\newcommand{\gmax}{$g_\text{max}$}
 \newcommand{\gmaxratio}{$g_\text{max,ratio}$}
@@
 \hypersetup{
-  pdftitle={Developmental plasticity to high light intensity amplifies the advantage of amphistomous leaves},
+  pdftitle={Plasticity and adaptation to high light intensity amplify the advantage of amphistomous leaves},
   pdfauthor={Christopher D. Muir\^{}\{1,2*\}; Wei Shen Lim\^{}\{1\}; Dachuan Wang\^{}\{1,2\}},
@@
 
-\title{Developmental plasticity to high light intensity amplifies the
+\title{Plasticity and adaptation to high light intensity amplify the
 advantage of amphistomous leaves}
@@
 populations representing 14 species of wild tomatoes (\emph{Solanum}
-sect. \emph{Lycopersicon} and sect. \emph{Lycopersicoides}). Plant grown
-under high light intensity benefit more in terms of CO\(_2\) assimilated
-for a given stomatal conductance than plants grown under low light,
-regardless of measurement light intensity or native light habitat.
-Developmental plasticity in leaf thickness and/or density, an adaptive
-response to optimize light interception, likely affects CO\(_2\)
-diffusion within the leaf, altering amphistomy advantage as a byproduct.
-Developmental plasticity may therefore play an important and previously
-unnoticed role in explaining the adaptive significance of amphistomy and
-the distribution of amphistomatous leaves in nature.
+sect. \emph{Lycopersicon} and sect. \emph{Lycopersicoides}). Plants
+grown under high light intensity benefit more in terms of CO\(_2\)
+assimilated for a given stomatal conductance than plants grown under low
+light, regardless of measurement light intensity. Furthermore,
+populations native to open habitats tend to benefit more from amphistomy
+compared to those from more closed habitats. Developmental plasticity
+and adaptation of leaves to high light intensity likely play a combined
+role in explaining the adaptive significance and distribution of
+amphistomy in nature.
 
@@
 photosynthetic organs formed by a pair of guard cells. They are
-essential for balancing carbon gained per unit water lost and permitted
-vascular plants to grow tall on land by enabling access to CO\(_2\) for
-photosynthetic carbon assimilation while preventing hydraulic failure in
-variable environments (\emph{1}--\emph{3}). Optimal stomatal function
-depends on both dynamic changes in aperture on the scale of minutes to
-hours, as well as static anatomy determined by developmental plasticity
-and constitutive genetic differences (\emph{4}--\emph{6}). Understanding
-how stomata respond to environmental change over daily, developmental,
-and evolutionary time is important for understanding adaptation
-(\emph{1}, \emph{7}--\emph{12}), predicting paleoclimate from fossil
-cuticles (\emph{13}--\emph{15}), and improving crops (\emph{16}).
-Stomatal function contributes to global carbon and water cycles
-(\emph{17}) and therefore predicting future climate (\emph{18}).
+essential for balancing carbon gained per unit water lost and were an
+essential innovation in vascular plants, permitting them to grow tall on
+land by enabling access to CO\(_2\) for photosynthetic carbon
+assimilation while preventing hydraulic failure in variable environments
+(\emph{1}--\emph{3}). Optimal stomatal function depends on both dynamic
+changes in aperture on the scale of minutes to hours, as well as static
+anatomy determined by developmental plasticity and constitutive genetic
+differences (\emph{4}--\emph{6}). Understanding how stomata respond to
+environmental change over daily, developmental, and evolutionary time is
+important for understanding adaptation (\emph{1}, \emph{7}--\emph{12}),
+predicting paleoclimate from fossil cuticles (\emph{13}--\emph{15}), and
+improving crops (\emph{16}). Stomatal function contributes to global
+carbon and water cycles (\emph{17}) and therefore predicting future
+climate (\emph{18}).
 
@@
 (\emph{27}--\emph{31}) to control gas exchange through each surface. All
-else being equal, gas exchange through stomata on both surfaces
-increases CO\(_2\) supply to chloroplasts by providing a second parallel
-pathway through leaf intercellular airspaces, enhancing photosynthesis
-(\emph{20}, \emph{32}). The extent to which amphistomy increases
-CO\(_2\) supply depends on resistance to diffusion in intercellular
-airspaces. This resistance can be low in thin, porous, amphistomatous
-leaves (\emph{28}, \emph{33}), but may be more substantial in thick,
-dense, hypostomatous leaves (\emph{34}). Amphistomatous leaves also lose
-more water through evaporation because of a second boundary layer
-conductance (\emph{35}), but the additional carbon gain should be enough
-to offset this cost in most realistic scenarios (\emph{36}).
+else being equal, simultaneous gas exchange through stomata on both
+surfaces increases CO\(_2\) supply to chloroplasts by providing a second
+parallel pathway through leaf intercellular airspaces, enhancing
+photosynthesis (\emph{20}, \emph{32}). The extent to which amphistomy
+increases CO\(_2\) supply depends on resistance to diffusion in
+intercellular airspaces. This resistance can be low in thin, porous,
+amphistomatous leaves (\emph{28}, \emph{33}), but may be more
+substantial in thick, dense, hypostomatous leaves (\emph{34}). We refer
+to the intercellular airspace conductance (\gias), the inverse of
+resistance. Amphistomatous leaves also lose more water through
+evaporation because of a second boundary layer conductance (\emph{35}),
+but the additional carbon gain should be enough to offset this cost in
+most realistic scenarios (\emph{36}).
 
@@
 isobilateral leaves are amphistomatous (\emph{25}). But among
-dorsiventral leaves, herbaceous plants in open, high light habitats tend
-to have have amphistomatous leaves (\emph{22}, \emph{39}--\emph{44}).
-Most other leaves, except those from aquatic habitats, are
-hypostomatous, producing stomata only on the lower, abaxial surface.
-Even resupinate leaves develop stomata on the lower, albeit adaxial
-surface (\emph{45}), suggesting that leaf orientation (lower vs.~upper)
-rather than leaf polarity (abaxial vs.~adaxial) is causal. The
-covariation between stomatal density ratio and light habitat is both
-qualitative and quantitative. A higher proportion of sun leaves are
-amphistomatous (\emph{39}) and the proportion of stomata on the upper,
-adaxial surface increases with light (\emph{42}, \emph{43}). Resolving
-why high light intensity favors amphistomatous dorsivental leaves is an
-important first step toward understanding variation in stomatal density
-ratio and leaf structure-function relationships more generally.
+dorsiventral leaves, it is primarily herbaceous plants in open, high
+light habitats that tend to have amphistomatous leaves (\emph{22},
+\emph{39}--\emph{44}). Most other leaves, except those from aquatic
+habitats, are hypostomatous, producing stomata only on the lower,
+abaxial surface. Even resupinate leaves develop stomata on the lower,
+albeit adaxial surface (\emph{45}), suggesting that leaf orientation
+(lower vs.~upper) rather than leaf polarity (abaxial vs.~adaxial) is
+causal. The covariation between stomatal density ratio and light habitat
+is both qualitative and quantitative. A higher proportion of sun leaves
+are amphistomatous (\emph{39}) and the proportion of stomata on the
+upper, adaxial surface increases with light (\emph{42}, \emph{43}).
+Resolving why high light intensity favors amphistomatous dorsivental
+leaves is an important first step toward understanding variation in
+stomatal density ratio and leaf structure-function relationships more
+generally.
 
@@
 hypostomatous leaf by increasing conductance through the leaf
-intercellular airspaces and boundary layers; the additional water loss
-through a second boundary layer is typically small (\emph{35}). We
-quantify this benefit as the amphistomy advantage (\aax{}), the
-log-response ratio of photosynthesis in an amphistomatous leaf compared
-to an otherwise identical pseudohypostomatous leaf (\emph{20},
-\emph{46}). Why would \aax{} be greater in sun than shade? We consider
-three nonmutually exclusive hypotheses that we classify as
-`acclimatory', `plastic', and `constitutive' (Fig.~\ref{fig-concept}).
-
-\emph{Acclimatory hypothesis}: Photosynthetic induction to high light
+intercellular airspaces and boundary layers. We quantify this benefit as
+the amphistomy advantage (\aax{}), the log-response ratio of
+photosynthesis in an amphistomatous leaf compared to an otherwise
+identical pseudohypostomatous leaf (\emph{20}, \emph{46}). Why would
+\aax{} be greater in sun than shade? We consider three nonmutually
+exclusive hypotheses that we classify as `acclimation', `plasticity',
+and `constitutive' (Fig.~\ref{fig-concept}).
+
+\emph{Acclimation hypothesis}: Photosynthetic induction to high light
 intensity typically involves increases in total leaf stomatal
 conductance (increased CO\(_2\) supply), the concentration of active
-Rubisco, and electron transport capacity (increased CO\(_2\) demand). A
-one-dimensional circuit model using the Farquhar-von Caemmerer-Berry
-biochemical model of C\(_3\) photosynthesis (\emph{47}) shows that both
-increased stomatal conductance and Rubisco activity should increase
-\aax{}, all else being equal (Supporting Information). If the
-acclimatory hypothesis is correct, we predict that
+Rubisco, and electron transport capacity (increased CO\(_2\) demand). If
+the acclimation hypothesis is correct, we predict that
 \(\aahigh{} > \aalow{}\) for all populations regardless of native
@@
 develop dramatically different leaves in sun and shade conditions
-(\emph{48}). Plastic responses are likely adaptations to optimize
+(\emph{47}). Plastic responses are likely adaptations to optimize
 photosynthesis at different light intensities in variable environments
-(\emph{49}). Plastic changes in leaf anatomy and biochemistry could
+(\emph{48}). Plastic changes in leaf anatomy and biochemistry could
 modulate \aax{} as a byproduct. Thicker or less porous leaves, both of
 which are associated with high leaf mass per area (LMA), will have lower
-\(g_\mathrm{ias}\); leaves with increased total stomatal density and
-photosynthetic capacity have greater potential CO\(_2\) supply and
-demand. Under the plastic hypothesis, we predict that
-\(\aasun{} > \aashade{}\) for all populations and light intensities.
-Secondarily, \(\aasun{}\) and \gmaxsun{} should also be positively
-associated with native light habitat if genotypes adapted to sunny, open
-habitats if they can express a phenotype best adapted to that
-environment when leaves develop under high light intensity.
+\gias; leaves with increased total stomatal density and photosynthetic
+capacity have greater potential CO\(_2\) supply and demand. Under the
+plasticity hypothesis, we predict that \(\aasun{} > \aashade{}\) for all
+populations and light intensities.
 
@@
 optimal phenotypes. We therefore predict genotypes from more sunny, open
-habitats will have consistently greater \aax under also measurement and
+habitats will have consistently greater \aax{} under all measurement and
 growth light intensities. For herbaceous plants, light intensity is
-largely a function of the tree canopy (\emph{50}). Herbs growing in the
+largely a function of the tree canopy (\emph{49}). Herbs growing in the
 open will regularly experience high light intensity; herbs growing under
@@
 in Table~\ref{tbl-hypotheses}; detailed predictions for results that
-would indicate support for multiple hypotheses are in
+would indicate simultaneous support for multiple hypotheses are in
 Table~\ref{tbl-predictions}.
@@
 \midrule
-acclimatory & $\mathrm{AA}_{2000} > \mathrm{AA}_{150}$ & $\mathrm{AA}_{\text{sun}} = \mathrm{AA}_{\text{shade}}$ & $\operatorname{cor}(\mathrm{PAI}, \mathrm{AA}) = 0$\\
+acclimation & $\mathrm{AA}_{2000} > \mathrm{AA}_{150}$ & $\mathrm{AA}_{\text{sun}} = \mathrm{AA}_{\text{shade}}$ & $\operatorname{cor}(\mathrm{PAI}, \mathrm{AA}) = 0$\\
 plastic & $\mathrm{AA}_{2000} = \mathrm{AA}_{150}$ & $\mathrm{AA}_{\text{sun}} > \mathrm{AA}_{\text{shade}}$ & $\operatorname{cor}(\mathrm{PAI}, \mathrm{AA}) = 0$\\
@@
 We tested these hypotheses by comparing \aax{} among amphistomatous wild
-tomato species (\emph{51}) from different native light habitats, grown
-under simulated sun and shade light treatments, and measured under
-contrasting light intensities (low and high). We measured \aax{} on 572
-individual plants from 29 populations (average of 9.86 replicates per
-light treatment) using a recently developed method (\emph{46}). With
-this method, we directly compare the photosynthetic rate of an untreated
-amphistomatous leaf to that of the same leaf with gas exchange blocked
-through the adaxial (upper) surface by transparent plastic, which we
-refer to as `pseudohypostomy'. To compare amphi- and pseudohypostomatous
-leaves at identical whole-leaf \(g_\text{sc}\), we measure \(A\) over a
-range of \(g_\text{sc}\), inducing stomatal opening and closure by
-modulating humidity (see Materials and Methods for further details). We
-estimated `amphistomy advantage' (\aax) \emph{sensu} (\emph{20}), but
-with modifications previously described in (\emph{46}) and here
-(Materials and Methods). The native light intensity was represented by
-plant area index (PAI \(\unit{\meter\squared\per\meter\squared}\)),
-estimated using a global gridded data set derived from the Global
-Ecosystem Dynamics Investigation {[}GEDI; (\emph{52}){]} and
-georeferenced accession collection information from the Tomato Genetics
-Resource Center. The growth light intensities were \ppfdequals{761} (sun
-treatment) and \ppfdqty{115} (shade treatment) while all other
-environment conditions were nearly identical (see Materials and
-Methods). The high and low measurement light intensities were
-\ppfdequals{2000} (97.8:2.24 red:blue) and \ppfdequals{150} (87.0:13.0
-red:blue), respectively.
+tomato species (Table~\ref{tbl-populations}; Figure~\ref{fig-sr-rxnorm};
+(\emph{50})) from different native light habitats, grown under simulated
+sun and shade light treatments, and measured under contrasting light
+intensities (low and high). We measured \aax{} on 572 individual plants
+from 29 populations (average of 9.86 replicates per light treatment)
+using a recently developed method (\emph{46}). With this method, we
+directly compare the photosynthetic rate of an untreated amphistomatous
+leaf to that of the same leaf with gas exchange blocked through the
+adaxial (upper) surface by transparent plastic, which we refer to as
+`pseudohypostomy'. To compare amphi- and pseudohypostomatous leaves at
+identical whole-leaf stomatal conductance (\gsw), we measure \(A\) over
+a range of \gsw, inducing stomatal opening and closure by modulating
+humidity (see Materials and Methods for further details). We estimated
+`amphistomy advantage' (\aax) \emph{sensu} (\emph{20}), but with
+modifications previously described in (\emph{46}) and here (Materials
+and Methods). The native light intensity was represented by plant area
+index (PAI \(\unit{\meter\squared\per\meter\squared}\)), estimated using
+a global gridded data set derived from the Global Ecosystem Dynamics
+Investigation {[}GEDI; (\emph{51}){]} and georeferenced accession
+collection information from the Tomato Genetics Resource Center
+(Table~\ref{tbl-populations}). The growth light intensities were
+\ppfdequals{761} (sun treatment) and \ppfdqty{115} (shade treatment)
+while all other environment conditions were nearly identical (see
+Materials and Methods). The high and low measurement light intensities
+were \ppfdequals{2000} (97.8:2.24 red:blue) and \ppfdequals{150}
+(87.0:13.0 red:blue), respectively.
 
@@
 of 29 populations significant). Contrary to the predictions of the
-assimilatory hypothesis, \aax{} was greater in all populations under low
+acclimation hypothesis, \aax{} was greater in all populations under low
 measurement light intensity for both sun and shade grown plants. The
-effect of low light on \aax{} was more pronounced in the sun-grown
-plants, where \aax{} was significantly greater under low measurement
-light intensity in 13 of 29 populations compared to 12 populations for
-shade-grown plants. The overall average \aax{} of shade and sun grown
-plants measured under low light intensity was 0.064 (range:
-0.022--0.137; 28 of 29 populations significant) and 0.100 (range:
-0.049--0.206; 27 of 29 populations significant), respectively. There was
-a modest tendency for populations from more open habitats (lower PAI) to
-exhibit greater \aax{} and the slope was significantly different than 0
-in 3 of the 4 treatment combinations (Fig.~\ref{fig-aa}B).
+overall average \aax{} of shade and sun grown plants measured under low
+light intensity was 0.064 (range: 0.022--0.137; 28 of 29 populations
+significant) and 0.100 (range: 0.049--0.206; 27 of 29 populations
+significant), respectively. There was a modest tendency for populations
+from more open habitats (lower PAI) to exhibit greater \aax{} and the
+slope was significantly different than 0 in 3 of the 4 treatment
+combinations (Fig.~\ref{fig-aa}B).
 
@@
 The pattern of \aax{} across wild tomatoes strongly supports the
-plasticity hypothesis, argues against the acclimatory hypothesis, and
+plasticity hypothesis, contradicts the acclimation hypothesis, and
 provides modest support for the constitutive hypothesis. Plastic changes
@@
 \aax{}. LMA increased in sun grown plants in all populations by an
-average of 123\% {[}95\% CI: 42.9 to 256\%{]}, similar to plastic
-responses in many species (\emph{48}). While LMA is weakly, albeit
-significantly, associated with individual-level \aax{}
+average of 123\% {[}95\% CI: 42.9 to 256\%{]}, qualitatively similar to
+plastic responses in many species (\emph{47}). While LMA is weakly,
+albeit significantly, associated with individual-level \aax{}
 (Fig.~\ref{fig-lma_aa}), the effect of growth light intensity on \aax{}
@@
 In contrast, LMA did not mediate an indirect effect of native PAI on
-\aax{} (Table/Figure). Many anatomical traits underlie LMA (\emph{53})
+\aax{} (Table/Figure). Many anatomical traits underlie LMA (\emph{52})
 and future research will be needed identify which particular traits,
@@
 hypothesis that resistance to CO\(_2\) diffusion is greater in the upper
-than lower portions of the leaf interior (\emph{54}). At low light
+than lower portions of the leaf interior (\emph{53}). At low light
 intensity, the bulk properties of the leaf are weighed toward the upper
@@
 the dashed line is the same for \(\aalow\). Confidence intervals for
-each population estimate and individual-level estimates are omitted for
-visual clarity.}
+each population estimate were omitted for visual clarity.}
 
@@
 We conclude that developmental plasticity, as opposed to acclimation or
-adaptation to open habitats, may explain the explain the long-standing
-observation that amphistomatous leaves are more common in sunny
-habitats, at least among herbaceous plants including crop relatives. Our
-results advance existing explanations by showing high light intensity
-\emph{per se} does not increase the benefit of amphistomy, but rather
-that anatomical and biochemical change associated with higher light
-intensity modulate \aax{}. Developmental plasticity needs to be
-considered when interpreting phylogenetic comparative analyses of
-adaptation, because plastic differences may be mistaken for genetic
-differences.
+adaptation to open habitats, may explain the long-standing observation
+that amphistomatous leaves are more common in sunny habitats, at least
+among herbaceous plants including crop relatives. Our results advance
+existing explanations by showing high light intensity \emph{per se} does
+not increase the benefit of amphistomy, but rather that anatomical and
+biochemical change associated with higher light intensity modulate
+\aax{}. Developmental plasticity needs to be considered when
+interpreting phylogenetic comparative analyses of adaptation, because
+plastic differences may be mistaken for genetic differences.
 
@@
 is beneficial for sun leaves because of plastic changes, but future
-research needs to estimates the tradeoffs associated with amphistomatous
+research needs to estimate the tradeoffs associated with amphistomatous
 leaves in order to fully explain their rarity among many woody plants
@@
 
-\bibitem[\citeproctext]{ref-farquhar_biochemical_1980}
+\bibitem[\citeproctext]{ref-poorter_metaanalysis_2019}
 \CSLLeftMargin{47. }%
-\CSLRightInline{G. D. Farquhar, S. von Caemmerer, J. A. Berry,
-\href{https://doi.org/10.1007/BF00386231}{A biochemical model of
-photosynthetic {CO}\(_{\textrm{2}}\) assimilation in leaves of
-{C}\(_{\textrm{3}}\) species}. \emph{Planta} \textbf{149}, 78--90
-(1980).}
-
-\bibitem[\citeproctext]{ref-poorter_metaanalysis_2019}
-\CSLLeftMargin{48. }%
 \CSLRightInline{H. Poorter, Ü. Niinemets, N. Ntagkas, A. Siebenkäs, M.
@@
 \bibitem[\citeproctext]{ref-givnish_common-garden_2014}
-\CSLLeftMargin{49. }%
+\CSLLeftMargin{48. }%
 \CSLRightInline{T. J. Givnish, R. A. Montgomery,
@@
 \bibitem[\citeproctext]{ref-engelbrecht_evaluation_2001}
-\CSLLeftMargin{50. }%
+\CSLLeftMargin{49. }%
 \CSLRightInline{B. M. J. Engelbrecht, H. M. Herz,
@@
 \bibitem[\citeproctext]{ref-peralta_taxonomy_2008}
-\CSLLeftMargin{51. }%
+\CSLLeftMargin{50. }%
 \CSLRightInline{I. E. Peralta, D. M. Spooner, S. Knapp, Taxonomy of wild
@@
 \bibitem[\citeproctext]{ref-burns_multi-resolution_2024}
-\CSLLeftMargin{52. }%
+\CSLLeftMargin{51. }%
 \CSLRightInline{P. Burns, C. R. Hakkenberg, S. J. Goetz,
@@
 \bibitem[\citeproctext]{ref-john_anatomical_2017}
-\CSLLeftMargin{53. }%
+\CSLLeftMargin{52. }%
 \CSLRightInline{G. P. John, C. Scoffoni, T. N. Buckley, R. Villar, H.
@@
 \bibitem[\citeproctext]{ref-jones_effects_1972}
-\CSLLeftMargin{54. }%
+\CSLLeftMargin{53. }%
 \CSLRightInline{H. G. Jones, R. O. Slatyer,
@@
 \bibitem[\citeproctext]{ref-schoch_dependence_1980}
-\CSLLeftMargin{55. }%
+\CSLLeftMargin{54. }%
 \CSLRightInline{P.-G. Schoch, C. Zinsou, M. Sibi,
@@
 \bibitem[\citeproctext]{ref-sack_developmental_2016}
-\CSLLeftMargin{56. }%
+\CSLLeftMargin{55. }%
 \CSLRightInline{L. Sack, T. N. Buckley,
@@
 \bibitem[\citeproctext]{ref-mott_amphistomy_1991}
-\CSLLeftMargin{57. }%
+\CSLLeftMargin{56. }%
 \CSLRightInline{K. A. Mott, O. Michaelson, Amphistomy as an adaptation
@@
 \bibitem[\citeproctext]{ref-schindelin_fiji_2012}
-\CSLLeftMargin{58. }%
+\CSLLeftMargin{57. }%
 \CSLRightInline{J. Schindelin, I. Arganda-Carreras, E. Frise, V. Kaynig,
@@
 Methods} \textbf{9}, 676--682 (2012).}
+
+\bibitem[\citeproctext]{ref-monteith_principles_2013}
+\CSLLeftMargin{58. }%
+\CSLRightInline{J. L. Monteith, M. H. Unsworth, \emph{Principles of
+Environmental Physics: Plants, Animals, and the Atmosphere}
+(Elsevier/Academic Press, Amsterdam ; Boston, 4th ed., 2013).}
 
@@
 \emph{Solanum} sect. \emph{Lycopersicon} and sect.
-\emph{Lycopersicoides} (\emph{51}) and the cultivated tomato \emph{S.
+\emph{Lycopersicoides} (\emph{50}) and the cultivated tomato \emph{S.
 lycopersicum} var. \emph{lycopersicum} (Table~\ref{tbl-populations}).
@@
 populations used in this study. The species name, accession number,
-collection latitude, longitude, elevation, and daily solar radiation
-estimated using the SPLASH algorithm (see `Climate data'). TGRC: Tomato
-Genetics Resource Center; PAI: Plant area index.}
+collection latitude, longitude, elevation, and plant area index (PAI)
+from the Global Ecosystem Dynamics Investigation (see `Climate data').
+TGRC: Tomato Genetics Resource Center.}
 
@@
 intensity of the treatment rather than the seedling conditions
-(\emph{55}). Shade plants grew slower than sun plants, hence leaves at
+(\emph{54}). Shade plants grew slower than sun plants, hence leaves at
 the same developmental stage were measured on chronologically older
@@
 measurements, we collected anatomical data from the nearest leaflet on
-the same leaf (1.58)\%. We measured chlorophyll concentration index
-(CCI) using a chrolophyll concentration meter (MC-100, Apogee
-Instruments, Logan, Utah, USA) on the lamina of focal leaflets before
-gas exchange measurements at the same time we measured leaf temperature.
+the same leaf (1.58\% of leaves).
 
@@
 the propafilm reduces transmission. To compensate for reduced
-transmission, we increased incident \ppfd for pseudohypo leaves by a
+transmission, we increased incident \ppfd{} for pseudohypo leaves by a
 factor 1/0.91, the inverse of the measured transmissivity of the
@@
 to low light (\ppfdequals{150}) and \rhequals{70} before inducing
-stomatal closure with low \rh and logging data as described above.
+stomatal closure with low \rh{} and logging data as described above.
 
@@
 surfaces from all leaves, using guard cell length as a proxy for
-stomatal size since it proportional to maximum conductance (\emph{56}).
-We made surface impressions of leaf lamina from the same area used for
-gas exchange measurements using a-silicone impression material (Zhermack
-elite HD+, light body, fast set, Rovigo, Italy). We applied clear nail
-polish to make positive replicas of the impression. After nail polish
-dried, we mounted replicas on a microscope slide using transparent tape
-(\emph{57}). We digitized a portion of each leaf surface replica using a
-brightfield microscope (Leica DM2000, Wetzlar, Germany). We counted and
-measured guard cell length on all stomata using the FIJI implementation
-of ImageJ2 version 2.3.0 (\emph{58}), then divided the count by the
-visible leaf area (\(\qty{0.890}{\raiseto{2}\milli\meter}\)) to estimate
-stomatal density.
+stomatal size since it is proportional to maximum conductance
+(\emph{55}). We made surface impressions of leaf lamina from the same
+area used for gas exchange measurements using a-silicone impression
+material (Zhermack elite HD+, light body, fast set, Rovigo, Italy). We
+applied clear nail polish to make positive replicas of the impression.
+After nail polish dried, we mounted replicas on a microscope slide using
+transparent tape (\emph{56}). We digitized a portion of each leaf
+surface replica using a brightfield microscope (Leica DM2000, Wetzlar,
+Germany). We counted and measured guard cell length on all stomata using
+the FIJI implementation of ImageJ2 version 2.3.0 (\emph{57}), then
+divided the count by the visible leaf area
+(\(\qty{0.890}{\raiseto{2}\milli\meter}\)) to estimate stomatal density.
+For each surface we calculated the anatomical maximum stomatal
+conductance (\gmax) at a reference leaf temperature of
+\(\qty{25}{\degreeCelsius}\) following (\emph{55}) as:
+
+\[g_\mathrm{max} = b m D_\mathrm{S} A_\mathrm{S}.\]
+
+The biophysical and morphological constants \(b\) and \(m\) are:
+
+\[b = \frac{D_\mathrm{wv}}{v},~\text{and}\]
+\[m = \frac{\pi c ^ 2}{j^{0.5} (4 h j + \pi c)},\]
+
+where \(D_\mathrm{wv}\) is the diffusion coefficient of water vapor in
+air and \(v\) is by the kinematic viscosity of dry air. We assumed
+\(D_\mathrm{wv} = \qty{2.49e-5}{\meter\squared\per\second}\) and
+\(v = \qty{2.24e-2}{\raiseto{3}\meter\per\mol}\) (\emph{58}). For
+kidney-shaped guard cells like those in wild tomatoes,
+\(c = h = j = 0.5\). The \gmaxratio{} is the \gmax{} of the adaxial
+surface divided by the sum total \gmax{} of both surfaces.
 
@@
 California. USA) and measured leaflet area from digital images using the
-FIJI implementation of ImageJ2 version 2.3.0 (\emph{58}). We dried
+FIJI implementation of ImageJ2 version 2.3.0 (\emph{57}). We dried
 leaves for 72 hours at \(\qty{74}{\degreeCelsius}\) in a food dehydrator
@@
 using a benchtop analytical balance (Ohaus PR64 Analytical Balance,
-Parsippany, New Jersey, USA). In \(10.5\%\) we measured LMA on the
-adjacent leaflet because the focal leaflet was damaged or wilted while
-making surface impressions and we could not reliably estimate area. LMA
-data are missing from \(2.97\%\) of individuals because the area or mass
-was not recorded at all or recorded incorrectly.
+Parsippany, New Jersey, USA). In \(10.5\%\) of leaves we measured LMA on
+the adjacent leaflet because the focal leaflet was damaged or wilted
+while making surface impressions and we could not reliably estimate
+area. LMA data are missing from \(2.97\%\) of individuals because the
+area or mass was not recorded at all or recorded incorrectly.
 
@@
 
-\emph{Procedure}: We provionsally estimated \aax{} for each replicate by
-integrating over the range of \gsw{} values where amphi and pseudohypo
-\agcurve{} curves overlap. In this procedure, curve parameters were
-provisionally estimated using ordinary least squares with the
-\texttt{lm()} function in \emph{R}. We then used point estimates of
-\aax{} for each replicate as the response variable in a linear model
-with light treatment, light intensity, population, and all interactions
-as explanatory variables. This model was also fit using ordinary least
-squares with the \texttt{lm()} function in \emph{R}. We classified
-extreme \aax{} values as those with an absolute internal studentized
-residual \(> 3\). Because these values likely indicate significant
-measurement error or leaf damage, we removed \agcurve{} curves at both
-light intensities if either was classified as extreme.
+\emph{Procedure}: We provisionally estimated \aax{} for each replicate
+by integrating over the range of \gsw{} values where amphi and
+pseudohypo \agcurve{} curves overlap. In this procedure, curve
+parameters were provisionally estimated using ordinary least squares
+with the \texttt{lm()} function in \emph{R}. We then used point
+estimates of \aax{} for each replicate as the response variable in a
+linear model with light treatment, light intensity, population, and all
+interactions as explanatory variables. This model was also fit using
+ordinary least squares with the \texttt{lm()} function in \emph{R}. We
+classified extreme \aax{} values as those with an absolute internal
+studentized residual \(> 3\). Because these values likely indicate
+significant measurement error or leaf damage, we removed \agcurve{}
+curves at both light intensities if either was classified as extreme.
 
@@
   Estimate the effects of light intensity, light treatment, and
-  population on \aax{} (assimilatory and plasticity hypotheses)
+  population on \aax{} (acclimation and plasticity hypotheses)
 \item
@@
 models with more iterations to decrease the Gelman-Rubin convergence
-statistic (\(\hat{R}\)) (\emph{63}) to less than
-\(`aa_stats[["max_rhat"]]`\) and the bulk effective sample size (ESS) to
-greater than \(`aa_stats[["min_ess"]]`\) for all model parameters, and
-there were fewer than \(`aa_stats[["n_divergent"]]`\) divergent
-transitions.
+statistic (\(\hat{R}\)) (\emph{63}) to less than 1.01 and the bulk
+effective sample size (ESS) to greater than 400 for all model
+parameters, and there were fewer than 10 divergent transitions.
 
@@
 
-We modeled \logA as a quadratic function of \loggsw for each leaf,
+We modeled \logA{} as a quadratic function of \loggsw{} for each leaf,
 measured at low and high light intensity, in both amphistomatous
@@
 
-The assimilatory, plastic, and constitutive hypotheses make different
+The acclimation, plasticity, and constitutive hypotheses make different
 predictions about the relationship between \aax{} and light intensity,
@@
 support for one, two, or all three hypotheses simultaneously in
-\autoref{tbl-predictions}. In general, the assimilatory hypothesis was
+\autoref{tbl-predictions}. In general, the acclimation hypothesis was
 supported if \aax{} was greater at high light intensity than low light
@@
 \midrule
-Assimilatory & $\beta_{\mathrm{AA},2000} > 0$ & \hspace{-1em}\aax{} at high light intensity is greater than that at low light intensity\\
+Acclimation & $\mathrm{AA}_{2000} > 0$ & \hspace{-1em}\aax{} at high light intensity is greater than that at low light intensity\\
 \cmidrule{1-3}\pagebreak[0]
-Plastic & $\beta_{\mathrm{AA},\text{sun}} > 0$ & \hspace{-1em}\aax{} in sun leaves is greater than that in shade leaves\\
+Plasticity & $\mathrm{AA}_{\text{sun}} > 0$ & \hspace{-1em}\aax{} in sun leaves is greater than that in shade leaves\\
 \cmidrule{1-3}\pagebreak[0]
-Constitutive & $\beta_{\mathrm{PPFD,AA}} > 0$ & \hspace{-1em}Population-level \aax{} ($\mathrm{AA}_\text{acc}$) decreases with native PAI\\
+Constitutive & $\beta_{\mathrm{PAI,AA}} > 0$ & \hspace{-1em}Population-level \aax{} ($\mathrm{AA}_\text{pop}$) decreases with native PAI\\
 \nopagebreak
- &  & \hspace{-1em}$\mathrm{AA}_\text{acc} = \beta_{\mathrm{AA}, 0} + \vec{\beta}_{\mathrm{AA}, \text{acc}}$\\
+ &  & \hspace{-1em}$\mathrm{AA}_\text{pop} = \beta_{\mathrm{AA}, 0} + \vec{\beta}_{\mathrm{AA}, \text{pop}}$\\
 \cmidrule{1-3}\pagebreak[0]
-Assimilatory $\times$ Plastic & $\beta_{\mathrm{AA},2000,\text{sun}} > 0$ & \hspace{-1em}\aax{} is highest at high light intensity in sun leaves\\
+Acclimation $\times$ Plasticity & $\mathrm{AA}_{2000,\text{sun}} > 0$ & \hspace{-1em}\aax{} is highest at high light intensity in sun leaves\\
 \nopagebreak
@@
 \cmidrule{1-3}\pagebreak[0]
-Assimilatory $\times$ Constitutive & $\beta_{\mathrm{AA},2000} > 0$ & \hspace{-1em}\aax{} at high light intensity is greater than that at low light intensity\\
+Acclimation $\times$ Constitutive & $\mathrm{AA}_{2000} > 0$ & \hspace{-1em}\aax{} at high light intensity is greater than that at low light intensity\\
 \nopagebreak
- & $\beta_{\mathrm{PPFD,AA}} > 0$ & \hspace{-1em}Population-level \aax{} at high light intensity ($\mathrm{AA}_{\text{acc},2000}$) decreases with native PAI\\
+ & $\beta_{\mathrm{PAI,AA}} > 0$ & \hspace{-1em}Population-level \aax{} at high light intensity ($\mathrm{AA}_{\text{pop},2000}$) decreases with native PAI\\
 \nopagebreak
- &  & \hspace{-1em}$\mathrm{AA}_{\text{acc},2000} = \beta_{\mathrm{AA}, 0} + \vec{\beta}_{\mathrm{AA}, \text{acc}} + \vec{\beta}_{\mathrm{AA}, 2000, \text{acc}}$\\
+ &  & \hspace{-1em}$\mathrm{AA}_{\text{pop},2000} = \beta_{\mathrm{AA}, 0} + \vec{\beta}_{\mathrm{AA}, \text{pop}} + \vec{\beta}_{\mathrm{AA}, 2000, \text{pop}}$\\
 \cmidrule{1-3}\pagebreak[0]
-Plastic $\times$ Constitutive & $\beta_{\mathrm{AA},\text{sun}} > 0$ & \hspace{-1em}\aax{} in sun leaves is greater than that in shade leaves\\
+Plasticity $\times$ Constitutive & $\mathrm{AA}_{\text{sun}} > 0$ & \hspace{-1em}\aax{} in sun leaves is greater than that in shade leaves\\
 \nopagebreak
- & $\beta_{\mathrm{PPFD,AA}} > 0$ & \hspace{-1em}Population-level \aax{} in sun leaves ($\mathrm{AA}_{\text{acc},\text{sun}}$) decreases with native PAI\\
+ & $\beta_{\mathrm{PAI,AA}} > 0$ & \hspace{-1em}Population-level \aax{} in sun leaves ($\mathrm{AA}_{\text{pop},\text{sun}}$) decreases with native PAI\\
 \nopagebreak
- &  & \hspace{-1em}$\mathrm{AA}_{\text{acc},\text{sun}} = \beta_{\mathrm{AA}, 0} + \vec{\beta}_{\mathrm{AA}, \text{acc}} + \vec{\beta}_{\mathrm{AA}, \text{sun}, \text{acc}}$\\
+ &  & \hspace{-1em}$\mathrm{AA}_{\text{pop},\text{sun}} = \beta_{\mathrm{AA}, 0} + \vec{\beta}_{\mathrm{AA}, \text{pop}} + \vec{\beta}_{\mathrm{AA}, \text{sun}, \text{pop}}$\\
 \cmidrule{1-3}\pagebreak[0]
-Assimilatory $\times$ Plastic $\times$ Constitutive & $\beta_{\mathrm{AA},2000,\text{sun}} > 0$ & \hspace{-1em}\aax{} is highest at high light intensity in sun leaves\\
+Acclimation $\times$ Plasticity $\times$ Constitutive & $\mathrm{AA}_{2000,\text{sun}} > 0$ & \hspace{-1em}\aax{} is highest at high light intensity in sun leaves\\
 \nopagebreak
- & $\beta_{\mathrm{PPFD,AA}} > 0$ & \hspace{-1em}Population-level \aax{} at high light intensity in sun leaves ($\mathrm{AA}_{\text{acc},2000,\text{sun}}$) decreases with native PAI\\
+ & $\beta_{\mathrm{PAI,AA}} > 0$ & \hspace{-1em}Population-level \aax{} at high light intensity in sun leaves ($\mathrm{AA}_{\text{pop},2000,\text{sun}}$) decreases with native PAI\\
 \nopagebreak
- &  & \hspace{-1em}$\mathrm{AA}_{\text{acc},2000,\text{sun}} = \beta_{\mathrm{AA}, 0} + \vec{\beta}_{\mathrm{AA}, \text{acc}} + \vec{\beta}_{\mathrm{AA}, 2000, \text{acc}} + \vec{\beta}_{\mathrm{AA}, \text{sun}, \text{acc}}$\\
+ &  & \hspace{-1em}$\mathrm{AA}_{\text{pop},2000,\text{sun}} = \beta_{\mathrm{AA}, 0} + \vec{\beta}_{\mathrm{AA}, \text{pop}} + \vec{\beta}_{\mathrm{AA}, 2000, \text{pop}} + \vec{\beta}_{\mathrm{AA}, \text{sun}, \text{pop}}$\\
 \bottomrule
@@
 \tightlist
-\item
-  Variation in stomatal ratio among pops and light treatments
-\item
-  Table S7: Model comparison LOOIC tables
-\item
-  Table S8: Model comparison LOOIC tables with LMA
 \item
@@
 
+\begin{figure}
+
+\centering{
+
+\includegraphics{ms_files/figure-pdf/fig-sr-rxnorm-1.pdf}
+
+}
+
+\caption{\label{fig-sr-rxnorm}Developmental plasticity in anatomical
+maximum stomatal conductance ratio (\gmaxratio; \(y\)-axis) to growth
+light intensity (\(x\)-axis) among \texttt{r\ n\_acc} wild tomato
+populations. All populations are amphistomatous, but vary in the
+stomatal density and size on each surface, which determines \gmaxratio.
+Each point is the average per population in that treatment; black lines
+connect the same population across treatments.}
+
+\end{figure}%
+
 \begin{longtable}[t]{>{\centering\arraybackslash}p{0.7in}>{\centering\arraybackslash}p{0.5in}>{\centering\arraybackslash}p{0.4in}>{\centering\arraybackslash}p{0.4in}>{}c|>{\centering\arraybackslash}p{0.4in}>{\centering\arraybackslash}p{0.4in}>{}c|>{\centering\arraybackslash}p{0.4in}>{\centering\arraybackslash}p{0.4in}}
@@
 light treatment, and population. Only the top models
-(\(\Delta~\text{LOOIC} < 2\)) are shown. The model with the lowest LOOIC
+(\(\Delta \text{LOOIC} < 2\)) are shown. The model with the lowest LOOIC
 is considered the best fit to the data. The LOOIC column lists the LOOIC
@@
 
-\subsubsection{MARKER - make other LOOIC
-table}\label{marker---make-other-looic-table}
-
 \begin{landscape}
 
-\begin{longtable}[t]{>{\centering\arraybackslash}p{0.7in}>{\centering\arraybackslash}p{0.5in}>{\centering\arraybackslash}p{0.4in}>{\centering\arraybackslash}p{0.4in}>{\centering\arraybackslash}p{0.4in}>{\centering\arraybackslash}p{0.4in}>{\centering\arraybackslash}p{0.4in}>{\centering\arraybackslash}p{0.4in}>{\centering\arraybackslash}p{0.4in}>{\centering\arraybackslash}p{0.4in}>{\centering\arraybackslash}p{0.4in}>{\centering\arraybackslash}p{0.4in}>{\centering\arraybackslash}p{0.4in}>{\centering\arraybackslash}p{0.4in}}
+\begin{longtable}[t]{>{\centering\arraybackslash}p{0.7in}>{\centering\arraybackslash}p{0.5in}>{\centering\arraybackslash}p{0.4in}>{\centering\arraybackslash}p{0.4in}>{\centering\arraybackslash}p{0.4in}>{\centering\arraybackslash}p{0.4in}>{\centering\arraybackslash}p{0.4in}>{\centering\arraybackslash}p{0.4in}>{\centering\arraybackslash}p{0.4in}>{\centering\arraybackslash}p{0.4in}>{\centering\arraybackslash}p{0.4in}>{\centering\arraybackslash}p{0.4in}}
 
@@
 corresponding model included affects of measurement light intensity
-(``Meas.''), growth light intensity (``Growth''), or LMA as fixed
-effects, random effects, or distributional effect on the residual
-variance. We test for fixed interaction effects between measurement and
-growth light intensity and LMA. We also test for random interaction
-effects between measurement and growth light intensity. All models
-included phylogenetically structured random effects of population.}
+(``Meas.''), growth light intensity (``Growth''), or LMA as fixed or
+random effects. We test for fixed interaction effects between
+measurement and growth light intensity and LMA. We also test for random
+interaction effects between measurement and growth light intensity. All
+models included phylogenetically structured random effects of
+population.}
 
@@
 \toprule
-\multicolumn{2}{c}{ } & \multicolumn{7}{c}{Fixed} & \multicolumn{3}{c}{Random} & \multicolumn{2}{c}{Distributional} \\
-\cmidrule(l{3pt}r{3pt}){3-9} \cmidrule(l{3pt}r{3pt}){10-12} \cmidrule(l{3pt}r{3pt}){13-14}
-LOOIC & $\Delta$LOOIC & Meas. & Growth & LMA & Meas. $\times$ Growth & Meas. $\times$ Growth & Meas. $\times$ LMA & Growth $\times$ LMA & Meas. & Growth & Meas. $\times$ Growth & Meas. & Growth\\
+\multicolumn{2}{c}{ } & \multicolumn{7}{c}{Fixed} & \multicolumn{3}{c}{Random} \\
+\cmidrule(l{3pt}r{3pt}){3-9} \cmidrule(l{3pt}r{3pt}){10-12}
+LOOIC & $\Delta$LOOIC & Meas. & Growth & LMA & Meas. $\times$ Growth & Meas. $\times$ Growth & Meas. $\times$ LMA & Growth $\times$ LMA & Meas. & Growth & Meas. $\times$ Growth\\
 \midrule
 \endfirsthead
-\multicolumn{14}{@{}l}{\textit{(continued)}}\\
+\multicolumn{12}{@{}l}{\textit{(continued)}}\\
 \toprule
-LOOIC & $\Delta$LOOIC & Meas. & Growth & LMA & Meas. $\times$ Growth & Meas. $\times$ Growth & Meas. $\times$ LMA & Growth $\times$ LMA & Meas. & Growth & Meas. $\times$ Growth & Meas. & Growth\\
+LOOIC & $\Delta$LOOIC & Meas. & Growth & LMA & Meas. $\times$ Growth & Meas. $\times$ Growth & Meas. $\times$ LMA & Growth $\times$ LMA & Meas. & Growth & Meas. $\times$ Growth\\
 \midrule
@@
 \endlastfoot
-\cellcolor{gray!10}{-1883.84} & \cellcolor{gray!10}{0.00} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$}\\
--1883.80 & 0.04 & $\checkmark$ &  & $\checkmark$ &  &  &  &  & $\checkmark$ & $\checkmark$ & $\checkmark$ & $\checkmark$ & $\checkmark$\\
-\cellcolor{gray!10}{-1883.67} & \cellcolor{gray!10}{0.17} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$}\\
--1883.16 & 0.67 & $\checkmark$ &  & $\checkmark$ &  &  &  &  & $\checkmark$ &  &  & $\checkmark$ & $\checkmark$\\
-\cellcolor{gray!10}{-1883.05} & \cellcolor{gray!10}{0.78} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$}\\
+\cellcolor{gray!10}{-1883.84} & \cellcolor{gray!10}{0.00} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$}\\
+-1883.80 & 0.04 & $\checkmark$ &  & $\checkmark$ &  &  &  &  & $\checkmark$ & $\checkmark$ & $\checkmark$\\
+\cellcolor{gray!10}{-1883.67} & \cellcolor{gray!10}{0.17} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{}\\
+-1883.16 & 0.67 & $\checkmark$ &  & $\checkmark$ &  &  &  &  & $\checkmark$ &  & \\
+\cellcolor{gray!10}{-1883.05} & \cellcolor{gray!10}{0.78} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{}\\
 \addlinespace
--1883.05 & 0.78 & $\checkmark$ &  & $\checkmark$ &  &  &  & $\checkmark$ & $\checkmark$ & $\checkmark$ &  & $\checkmark$ & $\checkmark$\\
-\cellcolor{gray!10}{-1882.93} & \cellcolor{gray!10}{0.90} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$}\\
--1882.87 & 0.97 & $\checkmark$ & $\checkmark$ & $\checkmark$ &  & $\checkmark$ &  &  & $\checkmark$ & $\checkmark$ & $\checkmark$ & $\checkmark$ & $\checkmark$\\
-\cellcolor{gray!10}{-1882.80} & \cellcolor{gray!10}{1.04} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$}\\
--1882.55 & 1.29 & $\checkmark$ & $\checkmark$ & $\checkmark$ &  & $\checkmark$ &  &  & $\checkmark$ &  &  & $\checkmark$ & $\checkmark$\\
+-1883.05 & 0.78 & $\checkmark$ &  & $\checkmark$ &  &  &  & $\checkmark$ & $\checkmark$ & $\checkmark$ & \\
+\cellcolor{gray!10}{-1882.93} & \cellcolor{gray!10}{0.90} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{}\\
+-1882.87 & 0.97 & $\checkmark$ & $\checkmark$ & $\checkmark$ &  & $\checkmark$ &  &  & $\checkmark$ & $\checkmark$ & $\checkmark$\\
+\cellcolor{gray!10}{-1882.80} & \cellcolor{gray!10}{1.04} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$}\\
+-1882.55 & 1.29 & $\checkmark$ & $\checkmark$ & $\checkmark$ &  & $\checkmark$ &  &  & $\checkmark$ &  & \\
 \addlinespace
-\cellcolor{gray!10}{-1882.29} & \cellcolor{gray!10}{1.55} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$}\\
--1882.27 & 1.56 & $\checkmark$ & $\checkmark$ & $\checkmark$ &  &  & $\checkmark$ &  & $\checkmark$ & $\checkmark$ & $\checkmark$ & $\checkmark$ & $\checkmark$\\
-\cellcolor{gray!10}{-1882.27} & \cellcolor{gray!10}{1.56} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$}\\
--1882.08 & 1.76 &  &  & $\checkmark$ &  &  &  &  & $\checkmark$ & $\checkmark$ & $\checkmark$ & $\checkmark$ & $\checkmark$\\
-\cellcolor{gray!10}{-1881.86} & \cellcolor{gray!10}{1.98} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$}\\*
+\cellcolor{gray!10}{-1882.29} & \cellcolor{gray!10}{1.55} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$}\\
+-1882.27 & 1.56 & $\checkmark$ & $\checkmark$ & $\checkmark$ &  &  & $\checkmark$ &  & $\checkmark$ & $\checkmark$ & $\checkmark$\\
+\cellcolor{gray!10}{-1882.27} & \cellcolor{gray!10}{1.56} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$}\\
+-1882.08 & 1.76 &  &  & $\checkmark$ &  &  &  &  & $\checkmark$ & $\checkmark$ & $\checkmark$\\
+\cellcolor{gray!10}{-1881.86} & \cellcolor{gray!10}{1.98} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{}\\*
 

### 2025-07-24 r/14_plot-stomata.R

@@ -38,7 +38,7 @@ df_stomata_pred = c("lower_sd", "upper_sd", "lower_gcl", "upper_gcl") |>
            factor(levels = c("low", "high")) |>
            fct_recode(sun = "high", shade = "low"))
 
-ggplot(df_stomata_pred, aes(Growth, gmax_ratio, group = acc, color = Growth)) +
+sr_rxnorm = ggplot(df_stomata_pred, aes(Growth, gmax_ratio, group = acc, color = Growth)) +
   geom_line(color = "black") +
   geom_point() +
   scale_color_manual(values = c("shade" = "tomato4", "sun" = "tomato")) +
@@ -54,107 +54,6 @@ ggsave(
   "figures/sr-rxnorm.pdf",
   width = 4,
   height = 4,
-  device = cairo_pdf,
-  bg = "transparent"
 )
 
-
-# STUFF OF PLOTTING STOMATAL TRAITS FOR AA AND CLIMATE. NOT SURE WHAT I AM GOING TO USE
-# Posterior of AA for each accession in each light treatment and light intensity
-
-full_join(
-  df_stomata_pred |>
-    dplyr::select(acc, light_treatment, sr),
-  read_rds("objects/df_aa_pred1.rds") |>
-    ungroup() |>
-    dplyr::select(acc, aa, light_treatment, light_intensity),
-  by = join_by(acc, light_treatment)
-) |>
-  ggplot(aes(sr, aa, color = light_treatment)) +
-  facet_grid(light_intensity ~ .) +
-  geom_point() +
-  xlab("stomatal ratio") +
-  ylab("amphi advantage") 
-
-
-
-df_stomata = read_rds("data/stomata.rds") |>
-  left_join(plant_info, by = join_by(accession, replicate)) |>
-  rename(acc = accession) |>
-  filter(acc %in% unique(d1$acc))
-
-df_gmax_ratio = df_stomata |>
-  ungroup() |>
-  summarize(gmax_ratio = mean(gmax_ratio, na.rm = TRUE),
-            .by = c("acc", "light_treatment")) |>
-  mutate(
-    Growth = light_treatment |>
-      factor(levels = c("low", "high")) |>
-      fct_recode(sun = "high", shade = "low")
-  )
-
-df_sd = df_stomata |>
-  ungroup() |>
-  summarize(total_stomatal_density_mm2 = mean(total_stomatal_density_mm2, na.rm = TRUE),
[…]

### 2025-07-25 ms/ms.qmd
added model summary tables

@@
 We crossed fixed and random effect structures with fixed effects of measurement light intensity and/or growth light intensity on the distributional parameter, $\sigma$, the residual variance in \aax{}.
+We used a robust regression approach by assuming $t$-distributed residuals, which makes estimates less susceptible to extreme values.
 In total, we fit `r n_model1` models, each with a different combination of fixed, random, and distributional effects.
@@
 
-### MARKER - table of selected model 1
-
-### MARKER - table of selected model 2
+```{r fit_aa1, results = 'asis', echo=FALSE, eval=TRUE, message=FALSE, warning=FALSE}
+#| label: tbl-fit_aa1
+#| tbl-cap: "Parameter estimates and 95% confidence intervals (CIs) from the posterior distribution of the selected model.
+Models potentially include fixed, random, and distributional effects of growth light intensity (shade and sun), measurement light intensity (low and high), and their interaction on amphistomy advantage ($\\mathrm{AA}$).
+The distributional effects refer to the effect of factors on residual variation in $\\mathrm{AA}$, denoted $\\sigma$, on a log-link scale.
+For phylogenetically structured random effects, we report the estimated standard deviation (SD)."
+
+make_brms_summary_table = function(fit,
+                                   ci_level = 0.95,
+                                   digits = 2) {
+  # Get posterior summary
+  summ = posterior_summary(fit, prob = c((1 - ci_level) / 2, 1 - (1 - ci_level) / 2))
+  summ_df = as.data.frame(summ)
+  summ_df$param = rownames(summ_df)
+  colnames(summ_df) = c("Estimate", "Est.Error", "CI_low", "CI_high", "Parameter")
+
+  # Classify parameters
+  summ_df = summ_df |>
+    mutate(
+      Type = case_when(
+        grepl("^b_", Parameter) &
+          !grepl("sigma|Intercept", Parameter) ~ "fixed effects",
+        grepl("^b_Intercept$", Parameter) ~ "fixed effects",
+        grepl("^b_", Parameter) &
+          grepl("sigma", Parameter) ~ "distributional parameters",
+        grepl("^sd_", Parameter) ~ "random effect SDs",
+        TRUE ~ "Other"
+      )
+    )
+
+  # Clean parameter names
+  summ_df = summ_df |>
+    mutate(
+      Parameter = case_when(
+        Parameter == "b_Intercept" ~ "$\\mathrm{AA}$ intercept (shade, low light)",
+        Parameter == "b_light_treatmenthigh" ~ "effect of sun on $\\mathrm{AA}$",
+        Parameter == "b_light_intensity2000" ~ "effect of high light on $\\mathrm{AA}$",
+        Parameter == "b_light_treatmenthigh:light_intensity2000" ~ "effect of sun $\\times$ high light interaction on $\\mathrm{AA}$",
+        Parameter == "b_sigma_Intercept" ~ "$\\log \\sigma$ intercept (shade, low light)",
+        Parameter == "b_sigma_light_treatmenthigh" ~ "effect of sun on $\\log \\sigma$",
+        Parameter == "b_sigma_light_intensity2000" ~ "effect of high light on $\\log \\sigma$",
+        Parameter == "sd_acc__Intercept" ~ "$\\mathrm{AA}$ among populations",
+        Parameter == "sd_acc__light_treatmenthigh" ~ "effect of sun on $\\mathrm{AA}$ among populations",
+        Parameter == "sd_acc__light_intensity2000" ~ "effect of high light on $\\mathrm{AA}$ among populations",
+        Parameter == "sd_acc__light_treatmenthigh:light_intensity2000" ~ "effect of sun $\\times$ high light interaction on $\\mathrm{AA}$ among populations"
+      )
+    )
+
+  # Select and format
+  summ_df |>
+    filter(Type != "Other") |>
+    mutate(across(c(Estimate, CI_low, CI_high), ~ round(., digits))) |>
+    mutate(`Estimate [95% CI]` = paste0(Estimate, " [", CI_low, ", ", CI_high, "]")) |>
+    select(Type, Parameter, `Estimate [95% CI]`) |>
+    arrange(match(
+      Type,
+      c(
+        "fixed effects",
+        "distributional parameters",
+        "random effect SDs"
+      )
+    )) |>
+    as_tibble()
+}
+
+summary_tbl = make_brms_summary_table(fit_aa1)
+
+tbl_fixed = summary_tbl |> filter(Type == "fixed effects") |> select(-Type)
+tbl_dist  = summary_tbl |> filter(Type == "distributional parameters") |> select(-Type)
+tbl_rand  = summary_tbl |> filter(Type == "random effect SDs") |> select(-Type)
+
+bind_rows(tbl_fixed, tbl_dist, tbl_rand) |>
+  kable(
+    col.names = c("Parameter", "Estimate [95\\% CI]"),
+    booktabs = TRUE,
+    format = "latex",
+    escape = FALSE
+  ) |>
+  kable_styling(latex_options = c("striped")) |>
+  pack_rows("Fixed effects",
+            1,
+            nrow(tbl_fixed),
+            bold = FALSE,
+            italic = TRUE) |>
+  pack_rows(
+    "Distributional parameters",
+    nrow(tbl_fixed) + 1,
+    nrow(tbl_fixed) + nrow(tbl_dist),
+    bold = FALSE,
+    italic = TRUE
+  ) |>
+  pack_rows(
+    "Random effect SDs",
+    nrow(tbl_fixed) + nrow(tbl_dist) + 1,
+    nrow(tbl_fixed) + nrow(tbl_dist) + nrow(tbl_rand),
+    bold = FALSE,
+    italic = TRUE
+  ) |>
+  column_spec(1, width = "4in") |>
+  column_spec(2, width = "2in")
+
+```
+
+```{r fit_aa2, results = 'asis', echo=FALSE, eval=TRUE, message=FALSE, warning=FALSE}
+#| label: tbl-fit_aa2
+#| tbl-cap: "Parameter estimates and 95% confidence intervals (CIs) from the posterior distribution of the selected model with leaf mass per area (LMA) as potential mediator variable.
+Models potentially include fixed, random, and distributional effects of growth light intensity (shade and sun), measurement light intensity (low and high), $\\log \\mathrm{LMA}$, and their interactions on amphistomy advantage ($\\mathrm{AA}$).
+The distributional effects refer to the effect of factors on residual variation in $\\mathrm{AA}$, denoted $\\sigma$, on a log-link scale.
+LMA can also be affected by growth light intensity and population.
+For phylogenetically structured random effects, we report the estimated standard deviation (SD)."
+
+make_brms_summary_table = function(fit,
+                                   ci_level = 0.95,
+                                   digits = 2) {
+  # Get posterior summary
+  summ = posterior_summary(fit, prob = c((1 - ci_level) / 2, 1 - (1 - ci_level) / 2))
+  summ_df = as.data.frame(summ)
+  summ_df$param = rownames(summ_df)
+  colnames(summ_df) = c("Estimate", "Est.Error", "CI_low", "CI_high", "Parameter")
+
+  # Classify parameters
+  summ_df = summ_df |>
+    mutate(
+      Type = case_when(
+        grepl("^b_", Parameter) &
+          !grepl("sigma|Intercept", Parameter) ~ "fixed effects",
+        grepl("^b_(aa|llma)_Intercept$", Parameter) ~ "fixed effects",
+        grepl("^b_", Parameter) &
+          grepl("sigma", Parameter) ~ "distributional parameters",
+        grepl("^sd_", Parameter) ~ "random effect SDs",
+        TRUE ~ "Other",
+      ),
+      resp = case_when(
+        str_detect(Parameter, "_aa_") ~ "aa",
+        str_detect(Parameter, "_llma_") ~ "lma",
+        TRUE ~ NA_character_
+      )
+    )
+
+  # Clean parameter names
+  summ_df <- summ_df %>%
+    mutate(
+      Parameter = case_when(
+        # AA
+        Parameter == "b_aa_Intercept" ~ "$\\mathrm{AA}$ intercept (shade, low light)",
+        Parameter == "b_aa_light_treatmenthigh" ~ "effect of sun on $\\mathrm{AA}$",
+        Parameter == "b_aa_light_intensity2000" ~ "effect of high light on $\\mathrm{AA}$",
+        Parameter == "b_aa_light_treatmenthigh:light_intensity2000" ~ "effect of sun $\\times$ high light interaction on $\\mathrm{AA}$",
+        Parameter == "b_sigma_aa_Intercept" ~ "$\\log \\sigma$ intercept (shade, low light)",
+        Parameter == "b_sigma_aa_light_treatmenthigh" ~ "effect of sun on $\\log \\sigma$",
+        Parameter == "b_sigma_aa_light_intensity2000" ~ "effect of high light on $\\log \\sigma$",
+        Parameter == "sd_acc__aa_Intercept" ~ "$\\mathrm{AA}$ among populations",
+        Parameter == "sd_acc__aa_light_treatmenthigh" ~ "effect of sun on $\\mathrm{AA}$ among populations",
+        Parameter == "sd_acc__aa_light_intensity2000" ~ "effect of high light on $\\mathrm{AA}$ among populations",
+        Parameter == "sd_acc__aa_light_treatmenthigh:light_intensity2000" ~ "effect of sun $\\times$ high light interaction on $\\mathrm{AA}$ among populations",
+
+        # LMA
+        Parameter == "b_llma_Intercept" ~ "$\\log \\mathrm{LMA}$ intercept (shade)",
+        Parameter == "b_llma_light_treatmenthigh" ~ "effect of sun on $\\log \\mathrm{LMA}$",
+        Parameter == "b_sigma_llma_Intercept" ~ "$\\log \\sigma$ intercept (shade)",
+        Parameter == "b_sigma_llma_light_treatmenthigh" ~ "effect of sun on $\\log \\sigma$",
+        Parameter == "sd_acc__llma_Intercept" ~ "$\\log \\mathrm{LMA}$ among populations",
+        Parameter == "sd_acc__llma_light_treatmenthigh" ~ "effect of sun on $\\log \\mathrm{LMA}$ among populations"
+      )
+    )
+
+  # Select and format
+  summ_df |>
+    filter(Type != "Other") |>
+    mutate(across(c(Estimate, CI_low, CI_high), ~ round(., digits))) |>
+    mutate(`Estimate [95% CI]` = paste0(Estimate, " [", CI_low, ", ", CI_high, "]")) |>
+    select(resp, Type, Parameter, `Estimate [95% CI]`) |>
+    arrange(resp, match(
+      Type,
+      c(
+        "fixed effects",
+        "distributional parameters",
+        "random effect SDs"
+      )
+    )) |>
+    as_tibble()
+}
+
+summary_tbl = make_brms_summary_table(fit_aa2)
+
+tbl_aa_fixed = summary_tbl |> filter(Type == "fixed effects", resp == "aa") |> select(-Type)
+tbl_aa_dist  = summary_tbl |> filter(Type == "distributional parameters", resp == "aa") |> select(-Type)
+tbl_aa_rand  = summary_tbl |> filter(Type == "random effect SDs", resp == "aa") |> select(-Type)
+tbl_lma_fixed = summary_tbl |> filter(Type == "fixed effects", resp == "lma") |> select(-Type)
+tbl_lma_rand  = summary_tbl |> filter(Type == "random effect SDs", resp == "lma") |> select(-Type)
+
+bind_rows(tbl_aa_fixed,
+          tbl_aa_dist,
+          tbl_aa_rand,
+          tbl_lma_fixed,
+          tbl_lma_rand) |>
+  select(-resp) |>
+  kable(
+    col.names = c("Parameter", "Estimate [95\\% CI]"),
+    booktabs = TRUE,
+    format = "latex",
+    escape = FALSE
+  ) |>
+  kable_styling(latex_options = c("striped")) |>
+  pack_rows(
+    "Response: $\\mathrm{AA}$",
+    1,
+    nrow(tbl_aa_fixed) + nrow(tbl_aa_dist) + nrow(tbl_aa_rand),
+    escape = FALSE
+  ) |>
+  pack_rows(
+    "~~Fixed effects",
+    1,
+    nrow(tbl_aa_fixed),
+    escape = FALSE,
+    bold = FALSE,
+    italic = TRUE
+  ) |>
+  pack_rows(
+    "~~Distributional parameters",
+    nrow(tbl_aa_fixed) + 1,
+    nrow(tbl_aa_fixed) + nrow(tbl_aa_dist),
+    escape = FALSE,
+    bold = FALSE,
+    italic = TRUE
+  ) |>
+  pack_rows(
+    "~~Random effect SDs",
+    nrow(tbl_aa_fixed) + nrow(tbl_aa_dist) + 1,
+    nrow(tbl_aa_fixed) + nrow(tbl_aa_dist) + nrow(tbl_aa_rand),
+    escape = FALSE,
+    bold = FALSE,
+    italic = TRUE
+  ) |>
+  pack_rows(
+    "Response: $\\log~\\\\mathrm{LMA}$",
+    nrow(tbl_aa_fixed) + nrow(tbl_aa_dist) + nrow(tbl_aa_rand) + 1,
+    nrow(tbl_aa_fixed) + nrow(tbl_aa_dist) + nrow(tbl_aa_rand) + nrow(tbl_lma_fixed) + nrow(tbl_lma_rand),
+    escape = FALSE
+  ) |>
+  pack_rows(
+    "~~Fixed effects",
+    nrow(tbl_aa_fixed) + nrow(tbl_aa_dist) + nrow(tbl_aa_rand) + 1,
+    nrow(tbl_aa_fixed) + nrow(tbl_aa_dist) + nrow(tbl_aa_rand) + nrow(tbl_lma_fixed),
+    escape = FALSE,
+    bold = FALSE,
+    italic = TRUE
+  ) |>
+  pack_rows(
+    "~~Random effect SDs",
+    nrow(tbl_aa_fixed) + nrow(tbl_aa_dist) + nrow(tbl_aa_rand) + nrow(tbl_lma_fixed) + 1,
+    nrow(tbl_aa_fixed) + nrow(tbl_aa_dist) + nrow(tbl_aa_rand) + nrow(tbl_lma_fixed) + nrow(tbl_lma_rand),
+    escape = FALSE,
+    bold = FALSE,
+    italic = TRUE
+  ) |>
+  column_spec(1, width = "4.5in") |>
+  column_spec(2, width = "2in")
+
+```
 

### 2025-07-25 ms/ms.tex
added model summary tables

@@
 distributional parameter, \(\sigma\), the residual variance in \aax{}.
-In total, we fit 100 models, each with a different combination of fixed,
+We used a robust regression approach by assuming \(t\)-distributed
+residuals, which makes estimates less susceptible to extreme values. In
+total, we fit 100 models, each with a different combination of fixed,
 random, and distributional effects. We used the RAxML
@@
 
+\begin{table}
+
+\caption{\label{tbl-fit_aa1}Parameter estimates and 95\% confidence
+intervals (CIs) from the posterior distribution of the selected model.
+Models potentially include fixed, random, and distributional effects of
+growth light intensity (shade and sun), measurement light intensity (low
+and high), and their interaction on amphistomy advantage
+(\(\mathrm{AA}\)). The distributional effects refer to the effect of
+factors on residual variation in \(\mathrm{AA}\), denoted \(\sigma\), on
+a log-link scale. For phylogenetically structured random effects, we
+report the estimated standard deviation (SD).}
+
+\centering{
+
+\centering
+\begin{tabular}{>{\raggedright\arraybackslash}p{4in}>{\raggedright\arraybackslash}p{2in}}
+\toprule
+Parameter & Estimate [95\% CI]\\
+\midrule
+\addlinespace[0.3em]
+\multicolumn{2}{l}{\textit{Fixed effects}}\\
+\hspace{1em}\cellcolor{gray!10}{$\mathrm{AA}$ intercept (shade, low light)} & \cellcolor{gray!10}{0.08 [0.01, 0.16]}\\
+\hspace{1em}effect of sun on $\mathrm{AA}$ & 0.04 [-0.02, 0.12]\\
+\hspace{1em}\cellcolor{gray!10}{effect of high light on $\mathrm{AA}$} & \cellcolor{gray!10}{-0.03 [-0.06, 0]}\\
+\hspace{1em}effect of sun $\times$ high light interaction on $\mathrm{AA}$ & -0.03 [-0.1, 0.03]\\
+\addlinespace[0.3em]
+\multicolumn{2}{l}{\textit{Distributional parameters}}\\
+\hspace{1em}\cellcolor{gray!10}{$\log \sigma$ intercept (shade, low light)} & \cellcolor{gray!10}{-2.51 [-2.61, -2.41]}\\
+\hspace{1em}effect of sun on $\log \sigma$ & 0.23 [0.12, 0.33]\\
+\hspace{1em}\cellcolor{gray!10}{effect of high light on $\log \sigma$} & \cellcolor{gray!10}{-0.19 [-0.3, -0.09]}\\
+\addlinespace[0.3em]
+\multicolumn{2}{l}{\textit{Random effect SDs}}\\
+\hspace{1em}$\mathrm{AA}$ among populations & 0.06 [0.04, 0.1]\\
+\hspace{1em}\cellcolor{gray!10}{effect of sun on $\mathrm{AA}$ among populations} & \cellcolor{gray!10}{0.05 [0, 0.1]}\\
+\hspace{1em}effect of high light on $\mathrm{AA}$ among populations & 0.02 [0, 0.05]\\
+\hspace{1em}\cellcolor{gray!10}{effect of sun $\times$ high light interaction on $\mathrm{AA}$ among populations} & \cellcolor{gray!10}{0.04 [0, 0.1]}\\
+\bottomrule
+\end{tabular}
+
+}
+
+\end{table}%
+
+\begin{table}
+
+\caption{\label{tbl-fit_aa2}Parameter estimates and 95\% confidence
+intervals (CIs) from the posterior distribution of the selected model
+with leaf mass per area (LMA) as potential mediator variable. Models
+potentially include fixed, random, and distributional effects of growth
+light intensity (shade and sun), measurement light intensity (low and
+high), \(\log \mathrm{LMA}\), and their interactions on amphistomy
+advantage (\(\mathrm{AA}\)). The distributional effects refer to the
+effect of factors on residual variation in \(\mathrm{AA}\), denoted
+\(\sigma\), on a log-link scale. LMA can also be affected by growth
+light intensity and population. For phylogenetically structured random
+effects, we report the estimated standard deviation (SD).}
+
+\centering{
+
+\centering
+\begin{tabular}{>{\raggedright\arraybackslash}p{4.5in}>{\raggedright\arraybackslash}p{2in}}
+\toprule
+Parameter & Estimate [95\% CI]\\
+\midrule
+\addlinespace[0.3em]
+\multicolumn{2}{l}{\textbf{Response: $\mathrm{AA}$}}\\
+\addlinespace[0.3em]
+\multicolumn{2}{l}{\textit{~~Fixed effects}}\\
+\hspace{1em}\hspace{1em}\cellcolor{gray!10}{$\mathrm{AA}$ intercept (shade, low light)} & \cellcolor{gray!10}{-0.01 [-0.12, 0.09]}\\
+\hspace{1em}\hspace{1em}effect of sun on $\mathrm{AA}$ & 0.01 [-0.05, 0.06]\\
+\hspace{1em}\hspace{1em}\cellcolor{gray!10}{effect of high light on $\mathrm{AA}$} & \cellcolor{gray!10}{-0.03 [-0.07, 0]}\\
+\addlinespace[0.3em]
+\multicolumn{2}{l}{\textit{~~Distributional parameters}}\\
+\hspace{1em}\hspace{1em}$\log \sigma$ intercept (shade, low light) & -2.5 [-2.61, -2.4]\\
+\hspace{1em}\hspace{1em}\cellcolor{gray!10}{effect of sun on $\log \sigma$} & \cellcolor{gray!10}{0.23 [0.13, 0.34]}\\
+\hspace{1em}\hspace{1em}effect of high light on $\log \sigma$ & -0.19 [-0.29, -0.08]\\
+\addlinespace[0.3em]
+\multicolumn{2}{l}{\textit{~~Random effect SDs}}\\
+\hspace{1em}\hspace{1em}\cellcolor{gray!10}{$\mathrm{AA}$ among populations} & \cellcolor{gray!10}{0.06 [0.04, 0.1]}\\
+\hspace{1em}\hspace{1em}effect of sun on $\mathrm{AA}$ among populations & 0.04 [0, 0.1]\\
+\hspace{1em}\hspace{1em}\cellcolor{gray!10}{effect of high light on $\mathrm{AA}$ among populations} & \cellcolor{gray!10}{0.02 [0, 0.05]}\\
+\hspace{1em}\hspace{1em}effect of sun $\times$ high light interaction on $\mathrm{AA}$ among populations & 0.04 [0, 0.09]\\
+\addlinespace[0.3em]
+\multicolumn{2}{l}{\textbf{Response: $\log~\mathrm{LMA}$}}\\
+\addlinespace[0.3em]
+\multicolumn{2}{l}{\textit{~~Fixed effects}}\\
+\hspace{1em}\hspace{1em}\cellcolor{gray!10}{$\log \mathrm{LMA}$ intercept (shade)} & \cellcolor{gray!10}{2.99 [2.42, 3.52]}\\
+\hspace{1em}\hspace{1em}effect of sun on $\log \mathrm{LMA}$ & 0.8 [0.36, 1.27]\\
+\addlinespace[0.3em]
+\multicolumn{2}{l}{\textit{~~Random effect SDs}}\\
+\hspace{1em}\hspace{1em}\cellcolor{gray!10}{$\log \mathrm{LMA}$ among populations} & \cellcolor{gray!10}{0.48 [0.35, 0.66]}\\
+\hspace{1em}\hspace{1em}effect of sun on $\log \mathrm{LMA}$ among populations & 0.37 [0.23, 0.6]\\
+\bottomrule
+\end{tabular}
+
+}
+
+\end{table}%
+
 \begin{itemize}
@@
 
-\includegraphics{ms_files/figure-pdf/fig-sr-rxnorm-1.pdf}
+\pandocbounded{\includegraphics[keepaspectratio]{ms_files/figure-pdf/fig-sr-rxnorm-1.pdf}}
 

### 2025-07-25 ms/test.qmd
added model summary tables

--- 2024-04-09 ms/test.qmd
@@
 
-```{r load-objects, echo=FALSE, eval=TRUE}
+```{r load-objects, echo=FALSE, eval=TRUE, message=FALSE, warning=FALSE}
+library(brms)
+library(cowplot)
 library(dplyr)
+library(ggplot2)
+library(kableExtra)
+library(knitr)
+library(lubridate)
+library(magrittr)
 library(readr)
+library(stringr)
 library(tibble)
+library(tidyr)
+
+theme_set(theme_cowplot())
+
+source("../r/functions.R")
+
+aa_args = read_rds("../objects/aa_args1.rds")
+aa_stats = read_rds("../objects/aa_stats.rds")
+# sort list by name
+aa_stats = aa_stats[order(names(aa_stats))]
+
+plant_info = read_rds("../data/plant-info.rds")
+
+accession_info = read_rds("../data/accession-info.rds") |>
+  dplyr::select(species, accession)
+
+rh_curves = read_rds("../data/trimmed_rh_curves.rds")
+
+stomata = read_rds("../data/stomata.rds")
+
+fit_aa2 = read_rds("../objects/fit_aa2.rds")
 ```
 
-blah \autoref{tbl-predictions}.
+```{r fit_aa2, results = 'asis', echo=FALSE, eval=TRUE, message=FALSE, warning=FALSE}
+#| label: tbl-fit_aa2
+#| tbl-cap: "Parameter estimates and 95% confidence intervals (CIs) from the posterior distribution of the selected model with leaf mass per area (LMA) as potential mediator variable.
+Models potentially include fixed, random, and distributional effects of growth light intensity (shade and sun), measurement light intensity (low and high), $\\log \\mathrm{LMA}$, and their interactions on amphistomy advantage ($\\mathrm{AA}$).
+The distributional effects refer to the effect of factors on residual variation in $\\mathrm{AA}$, denoted $\\sigma$, on a log-link scale.
+LMA can also be affected by growth light intensity and population.
+For phylogenetically structured random effects, we report the estimated standard deviation (SD)."
 
-```{r, predictions, echo=FALSE,message=FALSE,warning=FALSE}
-#| label: tbl-predictions
-#| tbl-cap: "caption"
+make_brms_summary_table = function(fit,
+                                   ci_level = 0.95,
+                                   digits = 2) {
+  # Get posterior summary
+  summ = posterior_summary(fit, prob = c((1 - ci_level) / 2, 1 - (1 - ci_level) / 2))
+  summ_df = as.data.frame(summ)
+  summ_df$param = rownames(summ_df)
+  colnames(summ_df) = c("Estimate", "Est.Error", "CI_low", "CI_high", "Parameter")
 
-df_parameters |>
-  tribble(
-    Parameter = latex, Description = description) |>
-  knitr::kable(
+  # Classify parameters
+  summ_df = summ_df |>
+    mutate(
+      Type = case_when(
+        grepl("^b_", Parameter) &
+          !grepl("sigma|Intercept", Parameter) ~ "fixed effects",
+        grepl("^b_(aa|llma)_Intercept$", Parameter) ~ "fixed effects",
+        grepl("^b_", Parameter) &
+          grepl("sigma", Parameter) ~ "distributional parameters",
+        grepl("^sd_", Parameter) ~ "random effect SDs",
+        TRUE ~ "Other",
+      ),
+      resp = case_when(
+        str_detect(Parameter, "_aa_") ~ "aa",
+        str_detect(Parameter, "_llma_") ~ "lma",
+        TRUE ~ NA_character_
+      )
+    )
+
+  # Clean parameter names
+  summ_df <- summ_df %>%
+    mutate(Parameter = case_when(
+
+      # AA
+      Parameter == "b_aa_Intercept" ~ "$\\mathrm{AA}$ intercept (shade, low light)",
+      Parameter == "b_aa_light_treatmenthigh" ~ "effect of sun on $\\mathrm{AA}$",
+      Parameter == "b_aa_light_intensity2000" ~ "effect of high light on $\\mathrm{AA}$",
+      Parameter == "b_aa_light_treatmenthigh:light_intensity2000" ~ "effect of sun $\\times$ high light interaction on $\\mathrm{AA}$",
+      Parameter == "b_sigma_aa_Intercept" ~ "$\\log \\sigma$ intercept (shade, low light)",
+      Parameter == "b_sigma_aa_light_treatmenthigh" ~ "effect of sun on $\\log \\sigma$",
+      Parameter == "b_sigma_aa_light_intensity2000" ~ "effect of high light on $\\log \\sigma$",
+      Parameter == "sd_acc__aa_Intercept" ~ "$\\mathrm{AA}$ among populations",
+      Parameter == "sd_acc__aa_light_treatmenthigh" ~ "effect of sun on $\\mathrm{AA}$ among populations",
+      Parameter == "sd_acc__aa_light_intensity2000" ~ "effect of high light on $\\mathrm{AA}$ among populations",
+      Parameter == "sd_acc__aa_light_treatmenthigh:light_intensity2000" ~ "effect of sun $\\times$ high light interaction on $\\mathrm{AA}$ among populations",
+
+      # LMA
+      Parameter == "b_llma_Intercept" ~ "$\\log \\mathrm{LMA}$ intercept (shade)",
+      Parameter == "b_llma_light_treatmenthigh" ~ "effect of sun on $\\log \\mathrm{LMA}$",
+      Parameter == "b_sigma_llma_Intercept" ~ "$\\log \\sigma$ intercept (shade)",
+      Parameter == "b_sigma_llma_light_treatmenthigh" ~ "effect of sun on $\\log \\sigma$",
+      Parameter == "sd_acc__llma_Intercept" ~ "$\\log \\mathrm{LMA}$ among populations",
+      Parameter == "sd_acc__llma_light_treatmenthigh" ~ "effect of sun on $\\log \\mathrm{LMA}$ among populations"
+    ))
+
+  # Select and format
+  summ_df |>
+    filter(Type != "Other") |>
+    mutate(across(c(Estimate, CI_low, CI_high), ~ round(., digits))) |>
+    mutate(`Estimate [95% CI]` = paste0(Estimate, " [", CI_low, ", ", CI_high, "]")) |>
+    select(resp, Type, Parameter, `Estimate [95% CI]`) |>
+    arrange(resp,
+            match(
+      Type,
+      c(
+        "fixed effects",
+        "distributional parameters",
+        "random effect SDs"
+      )
+    )) |>
+    as_tibble()
+}
+
+summary_tbl = make_brms_summary_table(fit_aa2)
+
+tbl_aa_fixed = summary_tbl |> filter(Type == "fixed effects", resp == "aa") |> select(-Type)
+tbl_aa_dist  = summary_tbl |> filter(Type == "distributional parameters", resp == "aa") |> select(-Type)
+tbl_aa_rand  = summary_tbl |> filter(Type == "random effect SDs", resp == "aa") |> select(-Type)
+tbl_lma_fixed = summary_tbl |> filter(Type == "fixed effects", resp == "lma") |> select(-Type)
+tbl_lma_rand  = summary_tbl |> filter(Type == "random effect SDs", resp == "lma") |> select(-Type)
+
+bind_rows(tbl_aa_fixed, tbl_aa_dist, tbl_aa_rand, tbl_lma_fixed, tbl_lma_rand) |>
+  select(-resp) |>
+  kable(
+    col.names = c("Parameter", "Estimate [95\\% CI]"),
+    booktabs = TRUE,
     format = "latex",
-    align = c("l", "l"),
-    booktabs = TRUE,
-    escape = FALSE,
-    longtable = TRUE
+    escape = FALSE
   ) |>
-  kableExtra::kable_styling(latex_options = c("striped", "no_vertical_lines")) |>
-  kableExtra::column_spec(1, width = "1in") |>
-  kableExtra::column_spec(2, width = "5in") |>
-  kableExtra::pack_rows("\\agcurve{} curve parameters",
-                        min(which(df_parameters$par_type == "ags")),
-                        max(which(df_parameters$par_type == "ags")),
-                        escape = FALSE) |>
-  kableExtra::pack_rows("\\aax{} for each light intensity with leaf using \\agcurve{} curve parameters",
-                        min(which(df_parameters$par_type == "aa_est")),
-                        max(which(df_parameters$par_type == "aa_est")),
-                        escape = FALSE) |>
-  kableExtra::pack_rows("effects light intensity, light treatments, and accession on \\aax",
-                        min(which(df_parameters$par_type == "aa")),
-                        max(which(df_parameters$par_type == "aa")),
-                        escape = FALSE) |>
-  kableExtra::pack_rows("effects of native light habitat on accession-level \\aax",
-                        min(which(df_parameters$par_type == "ppfd_aa")),
-                        max(which(df_parameters$par_type == "ppfd_aa")),
-                        escape = FALSE)
+  kable_styling(latex_options = c("striped")) |>
+  pack_rows("Response: $\\mathrm{AA}$", 1, nrow(tbl_aa_fixed) + nrow(tbl_aa_dist) + nrow(tbl_aa_rand), escape = FALSE) |>
+  pack_rows("~~Fixed effects", 1, nrow(tbl_aa_fixed), escape = FALSE, bold = FALSE, italic = TRUE) |>
+  pack_rows("~~Distributional parameters",
+            nrow(tbl_aa_fixed) + 1,
+            nrow(tbl_aa_fixed) + nrow(tbl_aa_dist), escape = FALSE, bold = FALSE, italic = TRUE) |>
+  pack_rows(
+    "~~Random effect SDs",
+    nrow(tbl_aa_fixed) + nrow(tbl_aa_dist) + 1,
+    nrow(tbl_aa_fixed) + nrow(tbl_aa_dist) + nrow(tbl_aa_rand), escape = FALSE, bold = FALSE, italic = TRUE
+  ) |>
+    pack_rows("Response: $\\log~\\\\mathrm{LMA}$", nrow(tbl_aa_fixed) + nrow(tbl_aa_dist) + nrow(tbl_aa_rand) + 1, nrow(tbl_aa_fixed) + nrow(tbl_aa_dist) + nrow(tbl_aa_rand) + nrow(tbl_lma_fixed) + nrow(tbl_lma_rand), escape = FALSE) |>
+  pack_rows("~~Fixed effects", nrow(tbl_aa_fixed) + nrow(tbl_aa_dist) + nrow(tbl_aa_rand) + 1, nrow(tbl_aa_fixed) + nrow(tbl_aa_dist) + nrow(tbl_aa_rand) + nrow(tbl_lma_fixed), escape = FALSE, bold = FALSE, italic = TRUE) |>
+ pack_rows(
+    "~~Random effect SDs",
+    nrow(tbl_aa_fixed) + nrow(tbl_aa_dist) + nrow(tbl_aa_rand) + nrow(tbl_lma_fixed) + 1,
+    nrow(tbl_aa_fixed) + nrow(tbl_aa_dist) + nrow(tbl_aa_rand) + nrow(tbl_lma_fixed) + nrow(tbl_lma_rand), escape = FALSE, bold = FALSE, italic = TRUE
+  ) |>
+  column_spec(1, width = "4.5in") |>
+  column_spec(2, width = "2in")
 

### 2025-07-25 ms/manuscraps.qmd
finalizing draft ms

@@
  The effect of low light on \aax{} was more pronounced in the sun-grown plants, where \aax{} was significantly greater under low measurement light intensity in `r aa_summary3["high"]` of `r n_acc` populations compared to `r aa_summary3["low"]` populations for shade-grown plants.
+
+ ## conclusion
+
+ maybe this wording is better: Amphistomy likely plays a significant, but under appreciated, role in the carbon and water balance of plant groups where this trait is common, such as herbaceous plants in sunny habitats and woody species in arid biomes (cites).
+If our estimates of AA from the wild tomato clades

### 2025-07-25 ms/ms.qmd
finalizing draft ms

@@
 library(ggplot2)
+library(glue)
 library(kableExtra)
@@
 
-### MARKER - where is evidence for relationship between PAI and LMA, AA?
-
 <!-- Paragraph 7 (summarize secondary results) -->
@@
 LMA increased in sun grown plants in all populations by an average of `r signif(100 * (exp(post_lma["Estimate"]) - 1), 3)`% [95% CI: `r signif(100 * (exp(post_lma["Q2.5"]) - 1), 3)` to `r signif(100 * (exp(post_lma["Q97.5"]) - 1), 3)`%], qualitatively similar to plastic responses in many species [@poorter_metaanalysis_2019].
-While LMA is weakly, albeit significantly, associated with individual-level \aax{} ([Fig.
+While LMA is weakly, albeit significantly (@tbl-fit_aa2), associated with individual-level \aax{} ([Fig.
 @fig-lma_aa]), the effect of growth light intensity on \aax{} is still predictive based on model comparison (@tbl-aa_loo1).
-In contrast, LMA did not mediate an indirect effect of native PAI on \aax{} (Table/Figure).
+However, the direct effect of the sun treatment on \aax{} was weaker when LMA was included in the model (@tbl-fit_aa1 vs.
+@tbl-fit_aa2), suggestive of a mediating role.
 Many anatomical traits underlie LMA [@john_anatomical_2017] and future research will be needed identify which particular traits, such as leaf thickness or mesophyll porosity, are responsible for mediating \aax{}.
@@
 
-We conclude that developmental plasticity, as opposed to acclimation or adaptation to open habitats, may explain the long-standing observation that amphistomatous leaves are more common in sunny habitats, at least among herbaceous plants including crop relatives.
-Our results advance existing explanations by showing high light intensity *per se* does not increase the benefit of amphistomy, but rather that anatomical and biochemical change associated with higher light intensity modulate \aax{}.
-Developmental plasticity needs to be considered when interpreting phylogenetic comparative analyses of adaptation, because plastic differences may be mistaken for genetic differences.
-
-maybe this wording is better: Amphistomy likely plays a significant, but under appreciated, role in the carbon and water balance of plant groups where this trait is common, such as herbaceous plants in sunny habitats and woody species in arid biomes (cites).
-If our estimates of AA from the wild tomato clades
-
-note: estimates of AA are less than that described by Parkhurst and Mott (1990) and also less than that of Xiong and Flexas (2021). What does this tell us?
-
-A second major conclusion is that the significant photosynthetic gain of amphistomatous leaves implies that the many hypostomatous leaves that dominate mesic to wet forests globally are giving up 'free' carbon that would require little to no additional water loss.
-Under realistic assumptions, we estimate that hypostomatous leaves could achieve the same $A$ with X to X% less water loss if they were amphistomatous (Table S9).
-Ours is the first experimental demonstration that amphistomy is beneficial for sun leaves because of plastic changes, but future research needs to estimate the tradeoffs associated with amphistomatous leaves in order to fully explain their rarity among many woody plants and shade-tolerant herbs.
+We conclude that both developmental plasticity and adaptation to open habitats, but not acclimation, may explain the long-standing observation that amphistomatous leaves are more common in sunny habitats, at least among herbaceous plants including crop relatives.
+This result changes our understanding by showing high light intensity *per se* does not increase the benefit of amphistomy, but rather arise as a byproduct of anatomical and biochemical changes caused by plasticity and adaptation to higher light intensity modulate \aax{}.
+To gain more precise understanding of when amphistomy is most beneficial will require further research on its leaf anataomical and biochemical basis.
+
+The magnitude of \aax{} we document across populations is also noteworthy, as it implies that hypostomatous leaves predominant in mesic to wet forests globally are giving up 'free' carbon that would require little to no additional water loss.
+Back-of-the-envelope calculations based on our data suggest that leaves could use X-X% less water for the same carbon gain if they were amphistomatous (see Supporting Information for details of calculations).
+As this is the first quantitative comparison of \aax{} across species, estimation from a broader range of species and environments will be required to fully understand the ecological causes and consequences of stomatal distributions on leaf surfaces.
+
+<!-- note: estimates of AA are less than that described by Parkhurst and Mott (1990) and also less than that of Xiong and Flexas (2021). What does this tell us? -->
 
@@
 Justin Alter, Max Gatlin, Joana Kim, Jenna Matsuyama, Brandon Najarian, and Kai Yasuda contributed to data collection.
-Sarah Friedrich helped with conceptual figure.
+Sarah Friedrich helped with graphics.
 
@@
 We used the leave-one-out cross-validation information criterion (LOOIC) to compare the fit of models (\autoref{tbl-aa_loo1}) using the *R* package **loo** version `r packageVersion("loo")` [@vehtari_practical_2017] to calculate LOOIC values.
-We selected the model with the lowest LOOIC value to generate posterior predictions for hypothesis testing.
+We selected the model with the lowest LOOIC value (@tbl-aa_loo1) to generate posterior predictions for hypothesis testing.
+Parameter estimates from the selected model are shown in @tbl-fit_aa1.
+
+To test whether LMA mediated the relationship between growth light intensity and \aax{}, we repeated a similar model comparison procedure but included $\log \mathrm{LMA}$ as an additional fixed effect in the model, as well its interaction with measurement and growth light intensity.
+The model included a direct effect of growth light intensity on $\log \mathrm{LMA}$ and random phylogeneticlly structured effects of population on $\log \mathrm{LMA}$ and effects of sun treatment on $\log \mathrm{LMA}$.
+This enabled us to disentangle whether $\log \mathrm{LMA}$ effects \aax{} directly or indirectly because both traits respond to growth light intensity.
+We integrated over missing LMA data using the `mi()` function in **brms**.
+Because the evidence strong supported including effects of growth and measurement light intensity on the residual variance in $\mathrm{AA}$, we included these in terms in all models.
+In total, we fit `r n_model2` models, each with a different combination of fixed and random effect structures.
+We selected the model with the lowest LOOIC value (@tbl-aa_loo2) to generate posterior predictions for hypothesis testing.
+Parameter estimates from the selected model are shown in @tbl-fit_aa2.
 
@@
         grepl("^b_(aa|llma)_Intercept$", Parameter) ~ "fixed effects",
+        grepl("^bsp_", Parameter) ~ "fixed effects",
         grepl("^b_", Parameter) &
@@
         Parameter == "b_aa_light_treatmenthigh:light_intensity2000" ~ "effect of sun $\\times$ high light interaction on $\\mathrm{AA}$",
+        Parameter == "bsp_aa_millma" ~ "effect of $\\log \\mathrm{LMA}$ on $\\mathrm{AA}$",
         Parameter == "b_sigma_aa_Intercept" ~ "$\\log \\sigma$ intercept (shade, low light)",
@@
 
+```{r, sr-rxnorm-cap, echo=FALSE,message=FALSE,warning=FALSE}
+
+sr_rxnorm_cap = glue::glue("Developmental plasticity in anatomical maximum stomatal conductance ratio (\\gmaxratio; $y$-axis) to growth light intensity ($x$-axis) among {n_acc} wild tomato populations.
+All populations are amphistomatous, but vary in the stomatal density and size on each surface, which determines \\gmaxratio.
+Each point is the average per population in that treatment; black lines connect the same population across treatments.")
+
+```
+
 ```{r, sr-rxnorm, echo=FALSE,message=FALSE,warning=FALSE}
@@
 #| fig-height: 4
-#| fig-cap: "Developmental plasticity in anatomical maximum stomatal conductance ratio (\\gmaxratio; $y$-axis) to growth light intensity ($x$-axis) among `r n_acc` wild tomato populations.
-All populations are amphistomatous, but vary in the stomatal density and size on each surface, which determines \\gmaxratio.
-Each point is the average per population in that treatment; black lines connect the same population across treatments."
+#| fig-cap: !expr sr_rxnorm_cap
 read_rds("../figures/sr-rxnorm.rds")

### 2025-07-25 ms/ms.tex
finalizing draft ms

--- 2025-07-24 ms/ms.tex
@@
 plastic responses in many species (\emph{47}). While LMA is weakly,
-albeit significantly, associated with individual-level \aax{}
-(Fig.~\ref{fig-lma_aa}), the effect of growth light intensity on \aax{}
-is still predictive based on model comparison (Table~\ref{tbl-aa_loo1}).
-In contrast, LMA did not mediate an indirect effect of native PAI on
-\aax{} (Table/Figure). Many anatomical traits underlie LMA (\emph{52})
-and future research will be needed identify which particular traits,
-such as leaf thickness or mesophyll porosity, are responsible for
-mediating \aax{}. The fact that the \aax{} of sun plants was greater
-under low measurement light intensity supports a long-standing
-hypothesis that resistance to CO\(_2\) diffusion is greater in the upper
-than lower portions of the leaf interior (\emph{53}). At low light
-intensity, the bulk properties of the leaf are weighed toward the upper
-palisade where most light is intercepted. Hence, amphistomy may,
-unexpectedly, be particularly beneficial for sun leaves experiencing
-intermittent shade or cloud cover. Our study is limited in testing this
-because we could not directly measure the stomatal conductance ratio and
-intercellular resistance on each surface. Future experiments measuring
-\aax{} with a dual sided chamber (\emph{31}, \emph{33}) can overcome
-these limitations.
+albeit significantly (Table~\ref{tbl-fit_aa2}), associated with
+individual-level \aax{} (Fig.~\ref{fig-lma_aa}), the effect of growth
+light intensity on \aax{} is still predictive based on model comparison
+(Table~\ref{tbl-aa_loo1}). In contrast, LMA did not mediate an indirect
+effect of native PAI on \aax{} (Table/Figure). Many anatomical traits
+underlie LMA (\emph{52}) and future research will be needed identify
+which particular traits, such as leaf thickness or mesophyll porosity,
+are responsible for mediating \aax{}. The fact that the \aax{} of sun
+plants was greater under low measurement light intensity supports a
+long-standing hypothesis that resistance to CO\(_2\) diffusion is
+greater in the upper than lower portions of the leaf interior
+(\emph{53}). At low light intensity, the bulk properties of the leaf are
+weighed toward the upper palisade where most light is intercepted.
+Hence, amphistomy may, unexpectedly, be particularly beneficial for sun
+leaves experiencing intermittent shade or cloud cover. Our study is
+limited in testing this because we could not directly measure the
+stomatal conductance ratio and intercellular resistance on each surface.
+Future experiments measuring \aax{} with a dual sided chamber
+(\emph{31}, \emph{33}) can overcome these limitations.
 
@@
 \textbf{loo} version 2.8.0 (\emph{65}) to calculate LOOIC values. We
-selected the model with the lowest LOOIC value to generate posterior
-predictions for hypothesis testing.
+selected the model with the lowest LOOIC value (Table~\ref{tbl-aa_loo1})
+to generate posterior predictions for hypothesis testing. Parameter
+estimates from the selected model are shown in Table~\ref{tbl-fit_aa1}.
+
+To test whether LMA mediated the relationship between growth light
+intensity and \aax{}, we repeated a similar model comparison procedure
+but included \(\log \mathrm{LMA}\) as an additional fixed effect in the
+model, as well its interaction with measurement and growth light
+intensity. The model included a direct effect of growth light intensity
+on \(\log \mathrm{LMA}\) and random phylogeneticlly structured effects
+of population on \(\log \mathrm{LMA}\) and effects of sun treatment on
+\(\log \mathrm{LMA}\). This enabled us to disentangle whether
+\(\log \mathrm{LMA}\) effects \aax{} directly or indirectly because both
+traits respond to growth light intensity. We integrated over missing LMA
+data using the \texttt{mi()} function in \textbf{brms}. Because the
+evidence strong supported including effects of growth and measurement
+light intensity on the residual variance in \(\mathrm{AA}\), we included
+these in terms in all models. In total, we fit 75 models, each with a
+different combination of fixed and random effect structures. We selected
+the model with the lowest LOOIC value (Table~\ref{tbl-aa_loo2}) to
+generate posterior predictions for hypothesis testing. Parameter
+estimates from the selected model are shown in Table~\ref{tbl-fit_aa2}.
 
@@
 
+\begin{table}
+
+\caption{\label{tbl-fit_aa1}Parameter estimates and 95\% confidence
+intervals (CIs) from the posterior distribution of the selected model.
+Models potentially include fixed, random, and distributional effects of
+growth light intensity (shade and sun), measurement light intensity (low
+and high), and their interaction on amphistomy advantage
+(\(\mathrm{AA}\)). The distributional effects refer to the effect of
+factors on residual variation in \(\mathrm{AA}\), denoted \(\sigma\), on
+a log-link scale. For phylogenetically structured random effects, we
+report the estimated standard deviation (SD).}
+
+\centering{
+
+\centering
+\begin{tabular}{>{\raggedright\arraybackslash}p{4in}>{\raggedright\arraybackslash}p{2in}}
+\toprule
+Parameter & Estimate [95\% CI]\\
+\midrule
+\addlinespace[0.3em]
+\multicolumn{2}{l}{\textit{Fixed effects}}\\
+\hspace{1em}\cellcolor{gray!10}{$\mathrm{AA}$ intercept (shade, low light)} & \cellcolor{gray!10}{0.08 [0.01, 0.16]}\\
+\hspace{1em}effect of sun on $\mathrm{AA}$ & 0.04 [-0.02, 0.12]\\
+\hspace{1em}\cellcolor{gray!10}{effect of high light on $\mathrm{AA}$} & \cellcolor{gray!10}{-0.03 [-0.06, 0]}\\
+\hspace{1em}effect of sun $\times$ high light interaction on $\mathrm{AA}$ & -0.03 [-0.1, 0.03]\\
+\addlinespace[0.3em]
+\multicolumn{2}{l}{\textit{Distributional parameters}}\\
+\hspace{1em}\cellcolor{gray!10}{$\log \sigma$ intercept (shade, low light)} & \cellcolor{gray!10}{-2.51 [-2.61, -2.41]}\\
+\hspace{1em}effect of sun on $\log \sigma$ & 0.23 [0.12, 0.33]\\
+\hspace{1em}\cellcolor{gray!10}{effect of high light on $\log \sigma$} & \cellcolor{gray!10}{-0.19 [-0.3, -0.09]}\\
+\addlinespace[0.3em]
+\multicolumn{2}{l}{\textit{Random effect SDs}}\\
+\hspace{1em}$\mathrm{AA}$ among populations & 0.06 [0.04, 0.1]\\
+\hspace{1em}\cellcolor{gray!10}{effect of sun on $\mathrm{AA}$ among populations} & \cellcolor{gray!10}{0.05 [0, 0.1]}\\
+\hspace{1em}effect of high light on $\mathrm{AA}$ among populations & 0.02 [0, 0.05]\\
+\hspace{1em}\cellcolor{gray!10}{effect of sun $\times$ high light interaction on $\mathrm{AA}$ among populations} & \cellcolor{gray!10}{0.04 [0, 0.1]}\\
+\bottomrule
+\end{tabular}
+
+}
+
+\end{table}%
+
+\begin{table}
+
+\caption{\label{tbl-fit_aa2}Parameter estimates and 95\% confidence
+intervals (CIs) from the posterior distribution of the selected model
+with leaf mass per area (LMA) as potential mediator variable. Models
+potentially include fixed, random, and distributional effects of growth
+light intensity (shade and sun), measurement light intensity (low and
+high), \(\log \mathrm{LMA}\), and their interactions on amphistomy
+advantage (\(\mathrm{AA}\)). The distributional effects refer to the
+effect of factors on residual variation in \(\mathrm{AA}\), denoted
+\(\sigma\), on a log-link scale. LMA can also be affected by growth
+light intensity and population. For phylogenetically structured random
+effects, we report the estimated standard deviation (SD).}
+
+\centering{
+
+\centering
+\begin{tabular}{>{\raggedright\arraybackslash}p{4.5in}>{\raggedright\arraybackslash}p{2in}}
+\toprule
+Parameter & Estimate [95\% CI]\\
+\midrule
+\addlinespace[0.3em]
+\multicolumn{2}{l}{\textbf{Response: $\mathrm{AA}$}}\\
+\addlinespace[0.3em]
+\multicolumn{2}{l}{\textit{~~Fixed effects}}\\
+\hspace{1em}\hspace{1em}\cellcolor{gray!10}{$\mathrm{AA}$ intercept (shade, low light)} & \cellcolor{gray!10}{-0.01 [-0.12, 0.09]}\\
+\hspace{1em}\hspace{1em}effect of sun on $\mathrm{AA}$ & 0.01 [-0.05, 0.06]\\
+\hspace{1em}\hspace{1em}\cellcolor{gray!10}{effect of high light on $\mathrm{AA}$} & \cellcolor{gray!10}{-0.03 [-0.07, 0]}\\
+\hspace{1em}\hspace{1em}effect of $\log \mathrm{LMA}$ on $\mathrm{AA}$ & 0.03 [0, 0.05]\\
+\addlinespace[0.3em]
+\multicolumn{2}{l}{\textit{~~Distributional parameters}}\\
+\hspace{1em}\hspace{1em}\cellcolor{gray!10}{$\log \sigma$ intercept (shade, low light)} & \cellcolor{gray!10}{-2.5 [-2.61, -2.4]}\\
+\hspace{1em}\hspace{1em}effect of sun on $\log \sigma$ & 0.23 [0.13, 0.34]\\
+\hspace{1em}\hspace{1em}\cellcolor{gray!10}{effect of high light on $\log \sigma$} & \cellcolor{gray!10}{-0.19 [-0.29, -0.08]}\\
+\addlinespace[0.3em]
+\multicolumn{2}{l}{\textit{~~Random effect SDs}}\\
+\hspace{1em}\hspace{1em}$\mathrm{AA}$ among populations & 0.06 [0.04, 0.1]\\
+\hspace{1em}\hspace{1em}\cellcolor{gray!10}{effect of sun on $\mathrm{AA}$ among populations} & \cellcolor{gray!10}{0.04 [0, 0.1]}\\
+\hspace{1em}\hspace{1em}effect of high light on $\mathrm{AA}$ among populations & 0.02 [0, 0.05]\\
+\hspace{1em}\hspace{1em}\cellcolor{gray!10}{effect of sun $\times$ high light interaction on $\mathrm{AA}$ among populations} & \cellcolor{gray!10}{0.04 [0, 0.09]}\\
+\addlinespace[0.3em]
+\multicolumn{2}{l}{\textbf{Response: $\log~\mathrm{LMA}$}}\\
+\addlinespace[0.3em]
+\multicolumn{2}{l}{\textit{~~Fixed effects}}\\
+\hspace{1em}\hspace{1em}$\log \mathrm{LMA}$ intercept (shade) & 2.99 [2.42, 3.52]\\
+\hspace{1em}\hspace{1em}\cellcolor{gray!10}{effect of sun on $\log \mathrm{LMA}$} & \cellcolor{gray!10}{0.8 [0.36, 1.27]}\\
+\addlinespace[0.3em]
+\multicolumn{2}{l}{\textit{~~Random effect SDs}}\\
+\hspace{1em}\hspace{1em}$\log \mathrm{LMA}$ among populations & 0.48 [0.35, 0.66]\\
+\hspace{1em}\hspace{1em}\cellcolor{gray!10}{effect of sun on $\log \mathrm{LMA}$ among populations} & \cellcolor{gray!10}{0.37 [0.23, 0.6]}\\
+\bottomrule
+\end{tabular}
+
+}
+
+\end{table}%
+
 \begin{itemize}
@@
 maximum stomatal conductance ratio (\gmaxratio; \(y\)-axis) to growth
-light intensity (\(x\)-axis) among \texttt{r\ n\_acc} wild tomato
-populations. All populations are amphistomatous, but vary in the
-stomatal density and size on each surface, which determines \gmaxratio.
-Each point is the average per population in that treatment; black lines
-connect the same population across treatments.}
+light intensity (\(x\)-axis) among 29 wild tomato populations. All
+populations are amphistomatous, but vary in the stomatal density and
+size on each surface, which determines \gmaxratio. Each point is the
+average per population in that treatment; black lines connect the same
+population across treatments.}
 

### 2025-07-25 r/scratch - calcs.R

@@ -1,0 +1,18 @@
+# simplified back-of the envelope calculation for how much water could be save by switching to amphistomy
+source("r/header.R")
+
+rh_curves = read_rds("data/trimmed_rh_curves.rds")
+tr = read_rds("data/phylogeny.rds")
+A = vcv(tr, corr = TRUE)
+
+fit = brm(
+  log_A ~ leaf_type + light_intensity * log_gsw + (light_intensity * log_gsw | acc_id),
+  data = filter(rh_curves, acc == "LA0716"),
+  # data2 = list(A = A),
+  backend = "cmdstanr",
+  chains = 1,
+  family = student()
+)
+
+summary(fit)
+conditional_effects(fit)

### 2025-07-27 ms/ms.qmd
draft sent for feedback

@@
 Developmental plasticity and adaptation of leaves to high light intensity likely play a combined role in explaining the adaptive significance and distribution of amphistomy in nature.
+We estimate that hypostomatous leaves would expend 10-65\% more water to achieve the same photosynthetic rate as a comparable amphistomatous leaf, implying large fitness costs of amphistomy in  ecological contexts where this trait is rare.
 
@@
 The magnitude of \aax{} we document across populations is also noteworthy, as it implies that hypostomatous leaves predominant in mesic to wet forests globally are giving up 'free' carbon that would require little to no additional water loss.
-Back-of-the-envelope calculations based on our data suggest that leaves could use X-X% less water for the same carbon gain if they were amphistomatous (see Supporting Information for details of calculations).
+The cost of hypostomy can be quantified as the difference in total \gsw{} that an amphistomatous leaf would require to achieve the same photosynthetic rate.
+This cost can be locally approximated on a log-ratio scale as $\mathrm{AA} / \varepsilon_g$ (Supporting Information), where $\varepsilon_g$ is the elasticity of $A$ to \gsw.
+For $\mathrm{AA} = 0.05$, the water cost of hypostomy would be 0.5 (64.9\%) at low elasticities ($\varepsilon_g = 0.1$) and 0.1 (10.5\%) at high elasticities ($\varepsilon_g = 0.5$).
+For plants to expend this much extra water implies a large fitness cost of upper stomata in certain ecological contexts.
 As this is the first quantitative comparison of \aax{} across species, estimation from a broader range of species and environments will be required to fully understand the ecological causes and consequences of stomatal distributions on leaf surfaces.
@@
 We selected the model with the lowest LOOIC value (@tbl-aa_loo2) to generate posterior predictions for hypothesis testing.
+Note that LOOIC was calculated only from pointwise likelihood values of the submodel estimating effects of light intensity, light treatments, and population on \aax{}.
 Parameter estimates from the selected model are shown in @tbl-fit_aa2.
@@
 
-THIS COULD BE USED FOR LMA MODEL?
-Note that LOOIC was calculated only from pointwise likelihood values of the submodel estimating effects of light intensity, light treatments, and population on \aax{}.
+### Estimating the water cost of hypostomy
+
+Amphistomy advantage implies that hypostomatous leaves would need to lose more water by having higher \gsw{} to achieve the same photosynthetic rate.
+We estimated this cost of hypostomy approximately.
+For simpliticity, we focus on a sufficiently small region of the \agcurve{} curve such that we can approximate the relationship between \gsw{} and $A$ as log-log linear.
+We also assume that slope is same in both amphistomatous and hypostomatous leaves, which was often approximately the case in our experiment (@fig-curves).
+With these assumptions, the approximate water cost of hypostomy can be derived from the equations:
+
+$$\log(A_{\mathrm{amphi}}) = \beta_\text{amphi} + \varepsilon_g \log(g_\text{sw}),~\text{and}$$
+$$\log(A_{\mathrm{hypo}}) = \beta_\text{hypo} + \varepsilon_g \log(g_\text{sw}).$$
+Solving for the difference in $\log{g_\text{sw}}$ at a given $\log{A_{\mathrm{hypo}}}$ is:
+
+$$\log(g_\text{sw,amphi}) - \log(g_\text{sw,hypo}) = \frac{\beta_\text{amphi} - \beta_\text{hypo}}{\varepsilon_g}.$$
+Noting that because the curves are parallel, $\mathrm{AA} = \log(A_\text{amphi} /A_\text{hypo}) = \beta_\text{amphi} - \beta_\text{hypo}$, this reveals that the negative log-ratio of the \gsw{} is related to \aax{} divided by the slope:
+
+$$-\log\left(\frac{g_\text{sw,amphi}}{g_\text{sw,hypo}}\right) = \frac{\mathrm{AA}}{\varepsilon_g}.$$
+We note that the slope in this case is an elasticity of of $A$ to \gsw{} because both are log-transformed.
+This quantity will be proportional the water cost of hypostomy assuming that VPD and other environmental variables are held constant and boundary layer conductance are high enough to be negligible.
+
+<!-- estimated the average value of this quantity for each population under each measurement light treatment by regressing \gsw{} against $A$ on a log-log scale, with curve type (amphistomy versus pseudohypostomy) and replicate within population as a random effect using **brms**.
+We followed the fitting procedures described above for other models.
+Because of the different assumptions, the estimates of \aax{} in this model will not be identical to that described above, but this approximate method enables a simpler, back-of-the-envelope calculation of the water cost of hypostomy.
+-->
 
@@
 
-* Table S9: Calculation of alternate scenario where amphi leaves can get same A with less R
-* Each A-gs figure with fitted curves, R2, and AA
-
 ```{r, sr-rxnorm-cap, echo=FALSE,message=FALSE,warning=FALSE}
@@
 ```
+
+```{r, curves, echo=FALSE,message=FALSE,warning=FALSE}
+#| label: fig-curves
+#| fig-cap: "(Figures in separate file) The \\agcurve{} curves, fitted lines, and \\aax{} estimates for every individual plant.
+The title provides the Tomato Genetics Resource Center accession number, replicate letter, and species names.
+The subtitle indicates the growth light intensity (sun or shade).
+Points are raw data, lines are fitted curves, and shaded regions are 95% confidence ribbons for the fitted curves.
+The Bayesian correlation coefficient $r^2$ is shown to the right of each curve, along with the estimate of \\aax{} $\\pm$ two standard errors for each pair of amphi (orange) and pseudohypo (blue) curves."
+ggplot()
+```

### 2025-07-27 ms/ms.tex
draft sent for feedback

@@
 role in explaining the adaptive significance and distribution of
-amphistomy in nature.
+amphistomy in nature. We estimate that hypostomatous leaves would expend
+10-65\% more water to achieve the same photosynthetic rate as a
+comparable amphistomatous leaf, implying large fitness costs of
+amphistomy in ecological contexts where this trait is rare.
 
@@
 
-\subsubsection{MARKER - where is evidence for relationship between PAI
-and LMA,
-AA?}\label{marker---where-is-evidence-for-relationship-between-pai-and-lma-aa}
-
 The pattern of \aax{} across wild tomatoes strongly supports the
@@
 light intensity on \aax{} is still predictive based on model comparison
-(Table~\ref{tbl-aa_loo1}). In contrast, LMA did not mediate an indirect
-effect of native PAI on \aax{} (Table/Figure). Many anatomical traits
-underlie LMA (\emph{52}) and future research will be needed identify
-which particular traits, such as leaf thickness or mesophyll porosity,
-are responsible for mediating \aax{}. The fact that the \aax{} of sun
-plants was greater under low measurement light intensity supports a
-long-standing hypothesis that resistance to CO\(_2\) diffusion is
-greater in the upper than lower portions of the leaf interior
-(\emph{53}). At low light intensity, the bulk properties of the leaf are
-weighed toward the upper palisade where most light is intercepted.
-Hence, amphistomy may, unexpectedly, be particularly beneficial for sun
-leaves experiencing intermittent shade or cloud cover. Our study is
-limited in testing this because we could not directly measure the
-stomatal conductance ratio and intercellular resistance on each surface.
-Future experiments measuring \aax{} with a dual sided chamber
-(\emph{31}, \emph{33}) can overcome these limitations.
+(Table~\ref{tbl-aa_loo1}). However, the direct effect of the sun
+treatment on \aax{} was weaker when LMA was included in the model
+(Table~\ref{tbl-fit_aa1} vs. Table~\ref{tbl-fit_aa2}), suggestive of a
+mediating role. Many anatomical traits underlie LMA (\emph{52}) and
+future research will be needed identify which particular traits, such as
+leaf thickness or mesophyll porosity, are responsible for mediating
+\aax{}. The fact that the \aax{} of sun plants was greater under low
+measurement light intensity supports a long-standing hypothesis that
+resistance to CO\(_2\) diffusion is greater in the upper than lower
+portions of the leaf interior (\emph{53}). At low light intensity, the
+bulk properties of the leaf are weighed toward the upper palisade where
+most light is intercepted. Hence, amphistomy may, unexpectedly, be
+particularly beneficial for sun leaves experiencing intermittent shade
+or cloud cover. Our study is limited in testing this because we could
+not directly measure the stomatal conductance ratio and intercellular
+resistance on each surface. Future experiments measuring \aax{} with a
+dual sided chamber (\emph{31}, \emph{33}) can overcome these
+limitations.
 
@@
 
-We conclude that developmental plasticity, as opposed to acclimation or
-adaptation to open habitats, may explain the long-standing observation
+We conclude that both developmental plasticity and adaptation to open
+habitats, but not acclimation, may explain the long-standing observation
 that amphistomatous leaves are more common in sunny habitats, at least
-among herbaceous plants including crop relatives. Our results advance
-existing explanations by showing high light intensity \emph{per se} does
-not increase the benefit of amphistomy, but rather that anatomical and
-biochemical change associated with higher light intensity modulate
-\aax{}. Developmental plasticity needs to be considered when
-interpreting phylogenetic comparative analyses of adaptation, because
-plastic differences may be mistaken for genetic differences.
-
-maybe this wording is better: Amphistomy likely plays a significant, but
-under appreciated, role in the carbon and water balance of plant groups
-where this trait is common, such as herbaceous plants in sunny habitats
-and woody species in arid biomes (cites). If our estimates of AA from
-the wild tomato clades
-
-note: estimates of AA are less than that described by Parkhurst and Mott
-(1990) and also less than that of Xiong and Flexas (2021). What does
-this tell us?
-
-A second major conclusion is that the significant photosynthetic gain of
-amphistomatous leaves implies that the many hypostomatous leaves that
-dominate mesic to wet forests globally are giving up `free' carbon that
-would require little to no additional water loss. Under realistic
-assumptions, we estimate that hypostomatous leaves could achieve the
-same \(A\) with X to X\% less water loss if they were amphistomatous
-(Table S9). Ours is the first experimental demonstration that amphistomy
-is beneficial for sun leaves because of plastic changes, but future
-research needs to estimate the tradeoffs associated with amphistomatous
-leaves in order to fully explain their rarity among many woody plants
-and shade-tolerant herbs.
+among herbaceous plants including crop relatives. This result changes
+our understanding by showing high light intensity \emph{per se} does not
+increase the benefit of amphistomy, but rather arise as a byproduct of
+anatomical and biochemical changes caused by plasticity and adaptation
+to higher light intensity modulate \aax{}. To gain more precise
+understanding of when amphistomy is most beneficial will require further
+research on its leaf anataomical and biochemical basis.
+
+The magnitude of \aax{} we document across populations is also
+noteworthy, as it implies that hypostomatous leaves predominant in mesic
+to wet forests globally are giving up `free' carbon that would require
+little to no additional water loss. The cost of hypostomy can be
+quantified as the difference in total \gsw{} that an amphistomatous leaf
+would require to achieve the same photosynthetic rate. This cost can be
+locally approximated on a log-ratio scale as
+\(\mathrm{AA} / \varepsilon_g\) (Supporting Information), where
+\(\varepsilon_g\) is the elasticity of \(A\) to \gsw. For
+\(\mathrm{AA} = 0.05\), the water cost of hypostomy would be 0.5
+(64.9\%) at low elasticities (\(\varepsilon_g = 0.1\)) and 0.1 (10.5\%)
+at high elasticities (\(\varepsilon_g = 0.5\)). For plants to expend
+this much extra water implies a large fitness cost of upper stomata in
+certain ecological contexts. As this is the first quantitative
+comparison of \aax{} across species, estimation from a broader range of
+species and environments will be required to fully understand the
+ecological causes and consequences of stomatal distributions on leaf
+surfaces.
 
@@
 Yasuda contributed to data collection. Sarah Friedrich helped with
-conceptual figure.
+graphics.
 
@@
 the model with the lowest LOOIC value (Table~\ref{tbl-aa_loo2}) to
-generate posterior predictions for hypothesis testing. Parameter
-estimates from the selected model are shown in Table~\ref{tbl-fit_aa2}.
+generate posterior predictions for hypothesis testing. Note that LOOIC
+was calculated only from pointwise likelihood values of the submodel
+estimating effects of light intensity, light treatments, and population
+on \aax{}. Parameter estimates from the selected model are shown in
+Table~\ref{tbl-fit_aa2}.
 
@@
 
-THIS COULD BE USED FOR LMA MODEL? Note that LOOIC was calculated only
-from pointwise likelihood values of the submodel estimating effects of
-light intensity, light treatments, and population on \aax{}.
+\subsubsection{Estimating the water cost of
+hypostomy}\label{estimating-the-water-cost-of-hypostomy}
+
+Amphistomy advantage implies that hypostomatous leaves would need to
+lose more water by having higher \gsw{} to achieve the same
+photosynthetic rate. We estimated this cost of hypostomy approximately.
+For simpliticity, we focus on a sufficiently small region of the
+\agcurve{} curve such that we can approximate the relationship between
+\gsw{} and \(A\) as log-log linear. We also assume that slope is same in
+both amphistomatous and hypostomatous leaves, which was often
+approximately the case in our experiment (Figure~\ref{fig-curves}). With
+these assumptions, the approximate water cost of hypostomy can be
+derived from the equations:
+
+\[\log(A_{\mathrm{amphi}}) = \beta_\text{amphi} + \varepsilon_g \log(g_\text{sw}),~\text{and}\]
+\[\log(A_{\mathrm{hypo}}) = \beta_\text{hypo} + \varepsilon_g \log(g_\text{sw}).\]
+Solving for the difference in \(\log{g_\text{sw}}\) at a given
+\(\log{A_{\mathrm{hypo}}}\) is:
+
+\[\log(g_\text{sw,amphi}) - \log(g_\text{sw,hypo}) = \frac{\beta_\text{amphi} - \beta_\text{hypo}}{\varepsilon_g}.\]
+Noting that because the curves are parallel,
+\(\mathrm{AA} = \log(A_\text{amphi} /A_\text{hypo}) = \beta_\text{amphi} - \beta_\text{hypo}\),
+this reveals that the negative log-ratio of the \gsw{} is related to
+\aax{} divided by the slope:
+
+\[-\log\left(\frac{g_\text{sw,amphi}}{g_\text{sw,hypo}}\right) = \frac{\mathrm{AA}}{\varepsilon_g}.\]
+We note that the slope in this case is an elasticity of of \(A\) to
+\gsw{} because both are log-transformed. This quantity will be
+proportional the water cost of hypostomy assuming that VPD and other
+environmental variables are held constant and boundary layer conductance
+are high enough to be negligible.
 
@@
 
-\begin{itemize}
-\tightlist
-\item
-  Table S9: Calculation of alternate scenario where amphi leaves can get
-  same A with less R
-\item
-  Each A-gs figure with fitted curves, R2, and AA
-\end{itemize}
-
 \begin{figure}
@@
 
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{ms_files/figure-pdf/fig-curves-1.pdf}}
+
+}
+
+\caption{\label{fig-curves}(Figures in separate file) The \agcurve{}
+curves, fitted lines, and \aax{} estimates for every individual plant.
+The title provides the Tomato Genetics Resource Center accession number,
+replicate letter, and species names. The subtitle indicates the growth
+light intensity (sun or shade). Points are raw data, lines are fitted
+curves, and shaded regions are 95\% confidence ribbons for the fitted
+curves. The Bayesian correlation coefficient \(r^2\) is shown to the
+right of each curve, along with the estimate of \aax{} \(\pm\) two
+standard errors for each pair of amphi (orange) and pseudohypo (blue)
+curves.}
+
+\end{figure}%
+
 \end{document}

### 2025-07-27 r/scratch - calcs.R

@@ -5,7 +5,7 @@ rh_curves = read_rds("data/trimmed_rh_curves.rds")
 tr = read_rds("data/phylogeny.rds")
 A = vcv(tr, corr = TRUE)
 
-plan(multisession, workers = 19)
+plan(multisession, workers = 9)
 
 set.seed(125691028)
 tibble(acc = unique(rh_curves$acc), seed = sample(1e9, length(unique(rh_curves$acc)))) |>
@@ -16,7 +16,7 @@ tibble(acc = unique(rh_curves$acc), seed = sample(1e9, length(unique(rh_curves$a
     
     while (crit == 0 & x < 24) {
       m = brm(
-        formula = log_A ~ leaf_type + light_intensity * log_gsw + (leaf_type + light_intensity * log_gsw |
+        formula = log_A ~ leaf_type + leaf_type:light_intensity + light_intensity * log_gsw + (leaf_type + leaf_type:light_intensity + light_intensity * log_gsw |
                                                                      acc_id),
         
         data = filter(rh_curves, acc == .acc),

### 2025-08-14 ms/ms.qmd
addressing k8's comments

@@
 ---
-title: Plasticity and adaptation to high light intensity amplify the advantage of amphistomous leaves
+title: Plasticity and adaptation to high light intensity amplify the advantage of amphistomatous leaves
 format:
@@
 
+<!-- Abstracts of Research Articles should explain to the general reader why the research was done, what was found and why the results are important.
+They should start with some brief BACKGROUND information: a sentence giving a broad introduction to the field comprehensible to the general reader, and then a sentence of more detailed background specific to your study.
+This should be followed by an explanation of the OBJECTIVES/METHODS and then the RESULTS.
+The final sentence should outline the main CONCLUSIONS of the study, in terms that will be comprehensible to all readers.
+The Abstract is distinct from the main body of the text and thus should not be the only source of background information critical to understanding the manuscript.
+Please do not include citations or abbreviations in the Abstract.
+The abstract should be 125 words or less.
+For Perspectives and Policy Forums, please include a one-sentence abstract.
+-->
+
 The presence of stomata on both leaf surfaces (amphistomy) can increase photosynthesis in C$_3$ plants by reducing the path length for CO$_2$ diffusion between substomatal cavities and chloroplasts.
-Amphistomatous leaves are common among herbaceous plants growing in sunny habitats, including many crop species.
-This distribution of amphistomatous leaves in nature may result from an increased photosynthetic benefit of amphistomy under high light intensity, either because of acclimatory, plastic, or constitutive variation in CO$_2$ supply or demand.
+Paradoxically, despite this photosynthetic advantage, most leaves are hypostomatous (stomata on lower surface only).
+Based on the observation that amphistomatous leaves are most common among herbaceous plants growing in sunny habitats, including many crop species, it has long been hypothesized that traits affecting CO$_2$ supply or demand amplify the advantage of amphistomy under high light intensity.
 We used a recently developed method to quantify amphistomy advantage, the photosynthetic rate of an amphistomatous leaf relative to an otherwise identical hypostomatous leaf, in `r n_acc` diverse populations representing `r n_sp` species of wild tomatoes (*Solanum* sect.
@@
 *Lycopersicoides*).
-Plants grown under high light intensity benefit more in terms of CO$_2$ assimilated for a given stomatal conductance than plants grown under low light, regardless of measurement light intensity.
+Plants grown under high light intensity benefit more from amphistomy in terms of CO$_2$ assimilated for a given stomatal conductance than plants grown under low light, regardless of measurement light intensity.
 Furthermore, populations native to open habitats tend to benefit more from amphistomy compared to those from more closed habitats.
@@
 Optimal stomatal function depends on both dynamic changes in aperture on the scale of minutes to hours, as well as static anatomy determined by developmental plasticity and constitutive genetic differences [@hetherington_role_2003; @de_boer_optimal_2016; @harrison_influence_2020].
-Understanding how stomata respond to environmental change over daily, developmental, and evolutionary time is important for understanding adaptation [@woodward_stomatal_1987; @buckley_modelling_2013; @raven_selection_2002; @haworth_co-ordination_2013; @liang_stomatal_2023; @chua_stomatal_2024; @lang_century-long_2024], predicting paleoclimate from fossil cuticles [@franks_new_2014; @mcelwain_paleoecology_2017; @the_cenozoic_co_proxy_integration_project_cencopip_consortium_toward_2023], and improving crops [@hofmann_impact_2025].
+Understanding how stomata respond to environmental change over daily, developmental, and evolutionary time is important for studying adaptation [@buckley_modelling_2013; @raven_selection_2002; @haworth_co-ordination_2013], inferring paleoclimate from fossil cuticles [@franks_new_2014; @mcelwain_paleoecology_2017; @the_cenozoic_co_proxy_integration_project_cencopip_consortium_toward_2023], predicting responses to climate change [@woodward_stomatal_1987; @liang_stomatal_2023; @chua_stomatal_2024; @lang_century-long_2024], and improving crops [@hofmann_impact_2025].
 Stomatal function contributes to global carbon and water cycles [@berry_stomata:_2010] and therefore predicting future climate [@franks_stomatal_2017].
@@
 adaxial) is causal.
+Stomatal density ratio, defined as the ratio of upper to total stomatal density, is a quantitative metric of stomatal distribution.
 The covariation between stomatal density ratio and light habitat is both qualitative and quantitative.
@@
 #| fig-align: center
-#| fig-cap: "\\textbf{Conceptual outline of three nonmutually exclusive hypotheses explaining why amphistomy advantage (\\aax{}) might be greater for leaves in sunny, open habitats.} The acclimatory hypothesis predicts that \\aax{} is greater under high measurement light intensity (\\ppfdqty{2000}) than low measurement light intensity (\\ppfdequals{150}), regardless of growth light intensity or native plant area index (PAI).
+#| fig-cap: "\\textbf{Conceptual outline of three nonmutually exclusive hypotheses explaining why amphistomy advantage (\\aax{}) might be greater for leaves in sunny, open habitats.} The acclimatopn hypothesis predicts that \\aax{} is greater under high measurement light intensity (\\ppfdqty{2000}) than low measurement light intensity (\\ppfdequals{150}), regardless of growth light intensity or native plant area index (PAI).
 The plasticity hypothesis predicts that \\aax{} is greater for plants grown in sun than shade, regardless of measurement light intensity or native PAI.
@@
 #| fig-height: 0.1
-#| fig-cap: "\\textbf{Amphistomy advantage (\\aax{}) in greater wild tomatoes from open habitats and is further amplified by developmental plasticity under sunny growth conditions.} \\aax{} ($y$-axis) is the log-response ratio of photosynthesis in an amphistomatous leaf compared to an otherwise identical pseudohypostomatous leaf.
+#| fig-cap: "\\textbf{Amphistomy advantage (\\aax{}) is greater in wild tomatoes from open habitats and is further amplified by developmental plasticity under sunny growth conditions.} \\aax{} ($y$-axis) is the log-response ratio of photosynthesis in an amphistomatous leaf compared to an otherwise identical pseudohypostomatous leaf.
 In both panels, estimates are shown for plants grown under simulated shade (brown) or sun (orange) and measured under high light intensity (\\ppfdequals{2000}; open circles) and low light intensity (\\ppfdequals{150}; solid circles).
@@
 The average \\aax{} among populations for each treatment group is written in the top left corner.
-(\\textbf{B}) The same estimates of \\aax{} in each combination of growth and measurement light intensity ploted against native plant area index (PAI; $x$-axis, log-scale).
+(\\textbf{B}) The same estimates of \\aax{} in each combination of growth and measurement light intensity plotted against native plant area index (PAI; $x$-axis, log-scale).
 The confidence intervals were omitted for visual clarity.
@@
 @tbl-fit_aa2), suggestive of a mediating role.
-Many anatomical traits underlie LMA [@john_anatomical_2017] and future research will be needed identify which particular traits, such as leaf thickness or mesophyll porosity, are responsible for mediating \aax{}.
+Many anatomical traits underlie LMA [@john_anatomical_2017] and future research will be needed to  identify which particular traits, such as leaf thickness or mesophyll porosity, are responsible for mediating \aax{}.
 The fact that the \aax{} of sun plants was greater under low measurement light intensity supports a long-standing hypothesis that resistance to CO$_2$ diffusion is greater in the upper than lower portions of the leaf interior [@jones_effects_1972].
-At low light intensity, the bulk properties of the leaf are weighed toward the upper palisade where most light is intercepted.
-Hence, amphistomy may, unexpectedly, be particularly beneficial for sun leaves experiencing intermittent shade or cloud cover.
+At low light intensity, photosynthesis is weighted toward the upper palisade where most light is intercepted.
+If resistance to CO$_2$ diffusion is high in the upper palisade, amphistomy may, unexpectedly, be particularly beneficial for sun leaves experiencing intermittent shade or cloud cover.
 Our study is limited in testing this because we could not directly measure the stomatal conductance ratio and intercellular resistance on each surface.
@@
 We conclude that both developmental plasticity and adaptation to open habitats, but not acclimation, may explain the long-standing observation that amphistomatous leaves are more common in sunny habitats, at least among herbaceous plants including crop relatives.
-This result changes our understanding by showing high light intensity *per se* does not increase the benefit of amphistomy, but rather arise as a byproduct of anatomical and biochemical changes caused by plasticity and adaptation to higher light intensity modulate \aax{}.
-To gain more precise understanding of when amphistomy is most beneficial will require further research on its leaf anataomical and biochemical basis.
+This result changes our understanding by showing that high light intensity *per se* does not increase the benefit of amphistomy.
+Instead, the benefit increases as a byproduct of anatomical and biochemical changes caused by plasticity and adaptation to higher light intensity.
+To gain more precise understanding of when amphistomy is most beneficial will require further research on its leaf anatomical and biochemical basis.
 
@@
 For plants to expend this much extra water implies a large fitness cost of upper stomata in certain ecological contexts.
-As this is the first quantitative comparison of \aax{} across species, estimation from a broader range of species and environments will be required to fully understand the ecological causes and consequences of stomatal distributions on leaf surfaces.
+As this is the first quantitative comparison of \aax{} across species, estimates from a broader range of species and more stressful environments will be required to fully understand the ecological causes and consequences of stomatal distributions on leaf surfaces.
 
@@
 *lycopersicum* (@tbl-populations).
+These populations were selected in a previous study because they encompass the breadth of climatic variation in the wild tomato clade [@pease_phylogenomics_2016].
 Due to constraints on growth space and time, we spread out measurements over `r n_weeks` weeks from `r start_licor_date` to `r end_licor_date`.
 Replicates within population were evenly spread out over this period to prevent confounding of temporal variation in growth conditions with accession.
-[anything else to say here?
-maybe explain population selection and phylogeny?]
 
@@
 
-In all growth spaces, we recorded $\mathrm{PPFD}$ using full spectrum quantum sensors (SQ-500-SS, Apogee Instruments, Logan, Utah, USA); we recorded temperature, RH, and [CO$_2$] using an EE894 sensor (E+E Elektronik, Engerwitzdorf, Austria) protected by a radiation shield.
+We recorded $\mathrm{PPFD}$ using full spectrum quantum sensors (SQ-500-SS, Apogee Instruments, Logan, Utah, USA); we recorded temperature, RH, and [CO$_2$] using an EE894 sensor (E+E Elektronik, Engerwitzdorf, Austria) protected by a radiation shield.
 All environmental measurements were taken every 10 minutes from the middle of plants racks at approximately the same height as the leaves we measured.
@@
 Despite these efforts, the air in the sun treatment was on average $\qty{`r temp_diff`}{\degreeCelsius}$ warmer and the average RH was consequently `r rh_diff`% lower.
-However, because of evaporative cooling, the leaves in the sun treatment were only $\qty{`r leaf_temp_diff`}{\degreeCelsius}$ on average ($n = `r n_leaf_temp`$ leaves).
+However, because of evaporative cooling, the leaves in the sun treatment were only $\qty{`r leaf_temp_diff`}{\degreeCelsius}$ warmer on average ($n = `r n_leaf_temp`$ leaves).
 
@@
 
-All models were using a Bayesian model with HMC sampling in the probabilistic programming language *Stan* [@stan_development_team_stan_2025] using the *R* package **brms** version `r fit_aa1$version$brms` [@burkner_brms_2017].
+All models were fit using a Bayesian model with HMC sampling in the probabilistic programming language *Stan* [@stan_development_team_stan_2025] using the *R* package **brms** version `r fit_aa1$version$brms` [@burkner_brms_2017].
 We used *CmdStan* version `r fit_aa1$version$cmdstan` and **cmdstanr** version `r fit_aa1$version$cmdstanr` [@gabry_cmdstanr_2025] to interface with *R* version `r getRversion()` [@r_core_team_r:_2025].

### 2025-08-14 ms/ms.tex
addressing k8's comments

@@
 \hypersetup{
-  pdftitle={Plasticity and adaptation to high light intensity amplify the advantage of amphistomous leaves},
+  pdftitle={Plasticity and adaptation to high light intensity amplify the advantage of amphistomatous leaves},
   pdfauthor={Christopher D. Muir\^{}\{1,2*\}; Wei Shen Lim\^{}\{1\}; Dachuan Wang\^{}\{1,2\}},
@@
 \title{Plasticity and adaptation to high light intensity amplify the
-advantage of amphistomous leaves}
+advantage of amphistomatous leaves}
 \author{Christopher D. Muir\(^{1,2*}\) \and Wei Shen
@@
 CO\(_2\) diffusion between substomatal cavities and chloroplasts.
-Amphistomatous leaves are common among herbaceous plants growing in
-sunny habitats, including many crop species. This distribution of
-amphistomatous leaves in nature may result from an increased
-photosynthetic benefit of amphistomy under high light intensity, either
-because of acclimatory, plastic, or constitutive variation in CO\(_2\)
-supply or demand. We used a recently developed method to quantify
-amphistomy advantage, the photosynthetic rate of an amphistomatous leaf
-relative to an otherwise identical hypostomatous leaf, in 29 diverse
-populations representing 14 species of wild tomatoes (\emph{Solanum}
-sect. \emph{Lycopersicon} and sect. \emph{Lycopersicoides}). Plants
-grown under high light intensity benefit more in terms of CO\(_2\)
-assimilated for a given stomatal conductance than plants grown under low
-light, regardless of measurement light intensity. Furthermore,
-populations native to open habitats tend to benefit more from amphistomy
-compared to those from more closed habitats. Developmental plasticity
-and adaptation of leaves to high light intensity likely play a combined
-role in explaining the adaptive significance and distribution of
-amphistomy in nature. We estimate that hypostomatous leaves would expend
-10-65\% more water to achieve the same photosynthetic rate as a
-comparable amphistomatous leaf, implying large fitness costs of
-amphistomy in ecological contexts where this trait is rare.
+Paradoxically, despite this photosynthetic advantage, most leaves are
+hypostomatous (stomata on lower surface only). Based on the observation
+that amphistomatous leaves are most common among herbaceous plants
+growing in sunny habitats, including many crop species, it has long been
+hypothesized that traits affecting CO\(_2\) supply or demand amplify the
+advantage of amphistomy under high light intensity. We used a recently
+developed method to quantify amphistomy advantage, the photosynthetic
+rate of an amphistomatous leaf relative to an otherwise identical
+hypostomatous leaf, in 29 diverse populations representing 14 species of
+wild tomatoes (\emph{Solanum} sect. \emph{Lycopersicon} and sect.
+\emph{Lycopersicoides}). Plants grown under high light intensity benefit
+more from amphistomy in terms of CO\(_2\) assimilated for a given
+stomatal conductance than plants grown under low light, regardless of
+measurement light intensity. Furthermore, populations native to open
+habitats tend to benefit more from amphistomy compared to those from
+more closed habitats. Developmental plasticity and adaptation of leaves
+to high light intensity likely play a combined role in explaining the
+adaptive significance and distribution of amphistomy in nature. We
+estimate that hypostomatous leaves would expend 10-65\% more water to
+achieve the same photosynthetic rate as a comparable amphistomatous
+leaf, implying large fitness costs of amphistomy in ecological contexts
+where this trait is rare.
 
@@
 environmental change over daily, developmental, and evolutionary time is
-important for understanding adaptation (\emph{1}, \emph{7}--\emph{12}),
-predicting paleoclimate from fossil cuticles (\emph{13}--\emph{15}), and
+important for studying adaptation (\emph{1}, \emph{7}, \emph{8}),
+inferring paleoclimate from fossil cuticles (\emph{9}--\emph{11}),
+predicting responses to climate change (\emph{12}--\emph{15}), and
 improving crops (\emph{16}). Stomatal function contributes to global
@@
 (lower vs.~upper) rather than leaf polarity (abaxial vs.~adaxial) is
-causal. The covariation between stomatal density ratio and light habitat
-is both qualitative and quantitative. A higher proportion of sun leaves
-are amphistomatous (\emph{39}) and the proportion of stomata on the
-upper, adaxial surface increases with light (\emph{42}, \emph{43}).
-Resolving why high light intensity favors amphistomatous dorsivental
-leaves is an important first step toward understanding variation in
-stomatal density ratio and leaf structure-function relationships more
-generally.
+causal. Stomatal density ratio, defined as the ratio of upper to total
+stomatal density, is a quantitative metric of stomatal distribution. The
+covariation between stomatal density ratio and light habitat is both
+qualitative and quantitative. A higher proportion of sun leaves are
+amphistomatous (\emph{39}) and the proportion of stomata on the upper,
+adaxial surface increases with light (\emph{42}, \emph{43}). Resolving
+why high light intensity favors amphistomatous dorsivental leaves is an
+important first step toward understanding variation in stomatal density
+ratio and leaf structure-function relationships more generally.
 
@@
 \caption{\label{fig-concept}\textbf{Conceptual outline of three nonmutually exclusive hypotheses explaining why amphistomy advantage (\aax{}) might be greater for leaves in sunny, open habitats.}
-The acclimatory hypothesis predicts that \aax{} is greater under high
+The acclimatopn hypothesis predicts that \aax{} is greater under high
 measurement light intensity (\ppfdqty{2000}) than low measurement light
@@
 
-\caption{\label{fig-aacap}\textbf{Amphistomy advantage (\aax{}) in greater wild tomatoes from open habitats and is further amplified by developmental plasticity under sunny growth conditions.}
+\caption{\label{fig-aacap}\textbf{Amphistomy advantage (\aax{}) is greater in wild tomatoes from open habitats and is further amplified by developmental plasticity under sunny growth conditions.}
 \aax{} (\(y\)-axis) is the log-response ratio of photosynthesis in an
@@
 left corner. (\textbf{B}) The same estimates of \aax{} in each
-combination of growth and measurement light intensity ploted against
+combination of growth and measurement light intensity plotted against
 native plant area index (PAI; \(x\)-axis, log-scale). The confidence
@@
 mediating role. Many anatomical traits underlie LMA (\emph{52}) and
-future research will be needed identify which particular traits, such as
-leaf thickness or mesophyll porosity, are responsible for mediating
+future research will be needed to identify which particular traits, such
+as leaf thickness or mesophyll porosity, are responsible for mediating
 \aax{}. The fact that the \aax{} of sun plants was greater under low
@@
 resistance to CO\(_2\) diffusion is greater in the upper than lower
-portions of the leaf interior (\emph{53}). At low light intensity, the
-bulk properties of the leaf are weighed toward the upper palisade where
-most light is intercepted. Hence, amphistomy may, unexpectedly, be
-particularly beneficial for sun leaves experiencing intermittent shade
-or cloud cover. Our study is limited in testing this because we could
-not directly measure the stomatal conductance ratio and intercellular
-resistance on each surface. Future experiments measuring \aax{} with a
-dual sided chamber (\emph{31}, \emph{33}) can overcome these
-limitations.
+portions of the leaf interior (\emph{53}). At low light intensity,
+photosynthesis is weighted toward the upper palisade where most light is
+intercepted. If resistance to CO\(_2\) diffusion is high in the upper
+palisade, amphistomy may, unexpectedly, be particularly beneficial for
+sun leaves experiencing intermittent shade or cloud cover. Our study is
+limited in testing this because we could not directly measure the
+stomatal conductance ratio and intercellular resistance on each surface.
+Future experiments measuring \aax{} with a dual sided chamber
+(\emph{31}, \emph{33}) can overcome these limitations.
 
@@
 among herbaceous plants including crop relatives. This result changes
-our understanding by showing high light intensity \emph{per se} does not
-increase the benefit of amphistomy, but rather arise as a byproduct of
-anatomical and biochemical changes caused by plasticity and adaptation
-to higher light intensity modulate \aax{}. To gain more precise
-understanding of when amphistomy is most beneficial will require further
-research on its leaf anataomical and biochemical basis.
+our understanding by showing that high light intensity \emph{per se}
+does not increase the benefit of amphistomy. Instead, the benefit
+increases as a byproduct of anatomical and biochemical changes caused by
+plasticity and adaptation to higher light intensity. To gain more
+precise understanding of when amphistomy is most beneficial will require
+further research on its leaf anatomical and biochemical basis.
 
@@
 certain ecological contexts. As this is the first quantitative
-comparison of \aax{} across species, estimation from a broader range of
-species and environments will be required to fully understand the
-ecological causes and consequences of stomatal distributions on leaf
-surfaces.
+comparison of \aax{} across species, estimates from a broader range of
+species and more stressful environments will be required to fully
+understand the ecological causes and consequences of stomatal
+distributions on leaf surfaces.
 
@@
 
-\bibitem[\citeproctext]{ref-woodward_stomatal_1987}
+\bibitem[\citeproctext]{ref-buckley_modelling_2013}
 \CSLLeftMargin{7. }%
-\CSLRightInline{F. I. Woodward,
-\href{https://doi.org/10.1038/327617a0}{Stomatal numbers are sensitive
-to increases in {CO2} from pre-industrial levels}. \emph{Nature}
-\textbf{327}, 617--618 (1987).}
-
-\bibitem[\citeproctext]{ref-buckley_modelling_2013}
-\CSLLeftMargin{8. }%
 \CSLRightInline{T. N. Buckley, K. A. Mott,
@@
 \bibitem[\citeproctext]{ref-haworth_co-ordination_2013}
-\CSLLeftMargin{9. }%
+\CSLLeftMargin{8. }%
 \CSLRightInline{M. Haworth, C. Elliott-Kingston, J. C. McElwain,
@@
 
-\bibitem[\citeproctext]{ref-liang_stomatal_2023}
-\CSLLeftMargin{10. }%
-\CSLRightInline{X. Liang, D. Wang, Q. Ye, J. Zhang, M. Liu, H. Liu, K.
-Yu, Y. Wang, E. Hou, B. Zhong, L. Xu, T. Lv, S. Peng, H. Lu, P. Sicard,
-A. Anav, D. S. Ellsworth,
-\href{https://doi.org/10.1038/s41467-023-37934-7}{Stomatal responses of
-terrestrial plants to global change}. \emph{Nature Communications}
-\textbf{14}, 2188 (2023).}
-
-\bibitem[\citeproctext]{ref-chua_stomatal_2024}
-\CSLLeftMargin{11. }%
-\CSLRightInline{L. C. Chua, O. S. Lau,
-\href{https://doi.org/10.1242/dev.202681}{Stomatal development in the
-changing climate}. \emph{Development} \textbf{151}, dev202681 (2024).}
-
-\bibitem[\citeproctext]{ref-lang_century-long_2024}
-\CSLLeftMargin{12. }%
-\CSLRightInline{P. L. M. Lang, J. M. Erberich, L. Lopez, C. L. Weiß, G.
-Amador, H. F. Fung, S. M. Latorre, J. R. Lasky, H. A. Burbano, M.
-Expósito-Alonso, D. C. Bergmann,
-\href{https://doi.org/10.1038/s41559-024-02481-x}{Century-long timelines
-of herbarium genomes predict plant stomatal response to climate change}.
-\emph{Nature Ecology \& Evolution} \textbf{8}, 1641--1653 (2024).}
-
 \bibitem[\citeproctext]{ref-franks_new_2014}
-\CSLLeftMargin{13. }%
+\CSLLeftMargin{9. }%
 \CSLRightInline{P. J. Franks, D. L. Royer, D. J. Beerling, P. K. Van de
@@
 \bibitem[\citeproctext]{ref-mcelwain_paleoecology_2017}
-\CSLLeftMargin{14. }%
+\CSLLeftMargin{10. }%
 \CSLRightInline{J. C. McElwain, M. Steinthorsdottir, Paleoecology,
@@
 \bibitem[\citeproctext]{ref-the_cenozoic_co_proxy_integration_project_cencopip_consortium_toward_2023}
-\CSLLeftMargin{15. }%
+\CSLLeftMargin{11. }%
 \CSLRightInline{The Cenozoic CO Proxy Integration Project (CenCOPIP)
@@
 \emph{Science} \textbf{382}, eadi5177 (2023).}
+
+\bibitem[\citeproctext]{ref-woodward_stomatal_1987}
+\CSLLeftMargin{12. }%
+\CSLRightInline{F. I. Woodward,
+\href{https://doi.org/10.1038/327617a0}{Stomatal numbers are sensitive
+to increases in {CO2} from pre-industrial levels}. \emph{Nature}
+\textbf{327}, 617--618 (1987).}
+
+\bibitem[\citeproctext]{ref-liang_stomatal_2023}
+\CSLLeftMargin{13. }%
+\CSLRightInline{X. Liang, D. Wang, Q. Ye, J. Zhang, M. Liu, H. Liu, K.
+Yu, Y. Wang, E. Hou, B. Zhong, L. Xu, T. Lv, S. Peng, H. Lu, P. Sicard,
+A. Anav, D. S. Ellsworth,
+\href{https://doi.org/10.1038/s41467-023-37934-7}{Stomatal responses of
+terrestrial plants to global change}. \emph{Nature Communications}
+\textbf{14}, 2188 (2023).}
+
+\bibitem[\citeproctext]{ref-chua_stomatal_2024}
+\CSLLeftMargin{14. }%
+\CSLRightInline{L. C. Chua, O. S. Lau,
+\href{https://doi.org/10.1242/dev.202681}{Stomatal development in the
+changing climate}. \emph{Development} \textbf{151}, dev202681 (2024).}
+
+\bibitem[\citeproctext]{ref-lang_century-long_2024}
+\CSLLeftMargin{15. }%
+\CSLRightInline{P. L. M. Lang, J. M. Erberich, L. Lopez, C. L. Weiß, G.
+Amador, H. F. Fung, S. M. Latorre, J. R. Lasky, H. A. Burbano, M.
+Expósito-Alonso, D. C. Bergmann,
+\href{https://doi.org/10.1038/s41559-024-02481-x}{Century-long timelines
+of herbarium genomes predict plant stomatal response to climate change}.
+\emph{Nature Ecology \& Evolution} \textbf{8}, 1641--1653 (2024).}
 
@@
 
-In all growth spaces, we recorded \(\mathrm{PPFD}\) using full spectrum
-quantum sensors (SQ-500-SS, Apogee Instruments, Logan, Utah, USA); we
-recorded temperature, RH, and {[}CO\(_2\){]} using an EE894 sensor (E+E
+We recorded \(\mathrm{PPFD}\) using full spectrum quantum sensors
+(SQ-500-SS, Apogee Instruments, Logan, Utah, USA); we recorded
+temperature, RH, and {[}CO\(_2\){]} using an EE894 sensor (E+E
 Elektronik, Engerwitzdorf, Austria) protected by a radiation shield. All
@@
 consequently 5.75\% lower. However, because of evaporative cooling, the
-leaves in the sun treatment were only \(\qty{0.886}{\degreeCelsius}\) on
-average (\(n = 699\) leaves).
+leaves in the sun treatment were only \(\qty{0.886}{\degreeCelsius}\)
+warmer on average (\(n = 699\) leaves).
 
@@
 
-\pandocbounded{\includegraphics[keepaspectratio]{ms_files/figure-pdf/fig-sr-rxnorm-1.pdf}}
+\includegraphics{ms_files/figure-pdf/fig-sr-rxnorm-1.pdf}
 
@@
 
-\pandocbounded{\includegraphics[keepaspectratio]{ms_files/figure-pdf/fig-curves-1.pdf}}
+\includegraphics{ms_files/figure-pdf/fig-curves-1.pdf}
 

### 2025-08-14 ms/ms.qmd
proofread supporting materials

@@
 These populations were selected in a previous study because they encompass the breadth of climatic variation in the wild tomato clade [@pease_phylogenomics_2016].
+In one case, we substituted a congeric population of the species (*S.
+galapagense*) because of difficult growing the focal population.
+We also added an a population of *S.
+pennellii*.
 Due to constraints on growth space and time, we spread out measurements over `r n_weeks` weeks from `r start_licor_date` to `r end_licor_date`.
-Replicates within population were evenly spread out over this period to prevent confounding of temporal variation in growth conditions with accession.
+Replicates within population were evenly spread out over this period to prevent confounding of temporal variation in growth conditions with variation among populations.
 
@@
 We recorded $\mathrm{PPFD}$ using full spectrum quantum sensors (SQ-500-SS, Apogee Instruments, Logan, Utah, USA); we recorded temperature, RH, and [CO$_2$] using an EE894 sensor (E+E Elektronik, Engerwitzdorf, Austria) protected by a radiation shield.
-All environmental measurements were taken every 10 minutes from the middle of plants racks at approximately the same height as the leaves we measured.
+All environmental measurements were taken every 10 minutes from the middle of plant racks at approximately the same height as the leaves we measured.
 We measured leaf temperature of focal leaves prior to measurement using an infrared radiometer (SI-111-SS, Apogee Instruments, Logan, Utah, USA).
@@
 We grew seeds and seedlings for both sun and shade treatments under the same environmental conditions (12:12 h, `r germ_temp_day`:$\qty{`r germ_temp_night`}{\degreeCelsius}$, `r germ_rh_day`:`r germ_rh_night` RH day:night cycle).
-LED light provided $\mathrm{PPFD} = \qty{`r germ_ppfd`}{\micro\mol\raiseto{-2}\meter\raiseto{-1}\second}$ (Fluence RAZRx, Austin, Texas, USA).
+LED light provided $\mathrm{PPFD} = \qty{`r germ_ppfd`}{\micro\mol\raiseto{-2}\meter\raiseto{-1}\second}$ (Fluence RAZRx, Austin, Texas, USA) during the germination and seedling stages.
 
@@
 
-#### Light-saturated photosynthetic rate
-
-In `r percent_aa_and_aq`\% of plants, we measured the maximum photosynthetic rate ($A_\mathrm{max}$) on the same leaflets as \agcurve{} curves.
+<!-- Not using this currently  -->
+<!-- #### Light-saturated photosynthetic rate -->
+
+<!-- In `r percent_aa_and_aq`\% of plants, we measured the maximum photosynthetic rate ($A_\mathrm{max}$) on the same leaflets as \agcurve{} curves.
 However, when a leaflet was damaged during \agcurve{} curves, we used the next closest leaflet for $A_\mathrm{max}$.
@@
 After $A$ and \gsw{} stabilized, we measured $A$ at $\qty{2000}{\micro\mol\raiseto{-2}\meter\raiseto{-1}\second}$.
+-->
 
@@
 
-1. Estimate \agcurve{} curve parameters for each leaf
+1. Estimate \agcurve{} curve parameters
 2. Estimate \aax{} for both measurement light intensities using \agcurve{} curve parameters
@@
 
-All models were fit using a Bayesian model with HMC sampling in the probabilistic programming language *Stan* [@stan_development_team_stan_2025] using the *R* package **brms** version `r fit_aa1$version$brms` [@burkner_brms_2017].
+All models were fit using a Bayesian HMC sampling in the probabilistic programming language *Stan* [@stan_development_team_stan_2025] using the *R* package **brms** version `r fit_aa1$version$brms` [@burkner_brms_2017].
 We used *CmdStan* version `r fit_aa1$version$cmdstan` and **cmdstanr** version `r fit_aa1$version$cmdstanr` [@gabry_cmdstanr_2025] to interface with *R* version `r getRversion()` [@r_core_team_r:_2025].
@@
 In all models, we included population as a phylogenetically structured random effect.
-We accounted for uncertainty in $\widehat{\mathrm{AA}}$ using the standard deviation of the posterior distribution of each $\widehat{\mathrm{AA}}$ estimate (@sec-aa-est).
+We accounted for uncertainty in $\widehat{\mathrm{AA}}$ using the standard deviation of the posterior distribution of each $\widehat{\mathrm{AA}}$ estimate (see above).
 We crossed fixed and random effect structures with fixed effects of measurement light intensity and/or growth light intensity on the distributional parameter, $\sigma$, the residual variance in \aax{}.
@@
 We used accession LA1044 in place of LA3909, two populations of *S.
-galapagnse*; LA0750 was added as sister to LA0716, two closely related populations of *S.
+galapagense*; LA0750 was added as sister to LA0716, two closely related populations of *S.
 pennellii*.
@@
 We integrated over missing LMA data using the `mi()` function in **brms**.
-Because the evidence strong supported including effects of growth and measurement light intensity on the residual variance in $\mathrm{AA}$, we included these in terms in all models.
+Because the evidence from the first set of models strongly supported including effects of growth and measurement light intensity on the residual variance in $\mathrm{AA}$, we included these in terms in all models.
 In total, we fit `r n_model2` models, each with a different combination of fixed and random effect structures.
@@
 
-The acclimation, plasticity, and constitutive hypotheses make different predictions about the relationship between \aax{} and light intensity, light treatment, and native \ppfd{} among populations.
+The acclimation, plasticity, and constitutive hypotheses make different predictions about the relationship between \aax{} and light intensity, light treatment, and native $\mathrm{PAI}$ among populations.
 Since these hypotheses are not mutually exclusive, we describe how we assessed support for one, two, or all three hypotheses simultaneously in \autoref{tbl-predictions}.
@@
 
-From the posterior of the selected model (@sec-phylo), we predicted the posterior distribution of the expected \aax{} for each population at each measurement light intensity and growth light intensity.
+From the model selected using LOOIC, we predicted the posterior distribution of the expected \aax{} for each population at each measurement light intensity and growth light intensity.
 If the 95% confidence intervals for \aax{} did not overlap zero, we considered \aax{} to be significantly different from zero.
 We also calculated posterior distribution for the average \aax{} among populations in each condition.
-We tested a linear effect of native light habitat on population-level \aax{} in each treatment condition, integrating over the entire posterior distribution.
-We calculated the posterior distribution of the slope.
+We tested a linear effect of native light habitat on population-level \aax{} in each treatment condition, integrating over the entire posterior distribution to calculate a posterior distribution of the slope.
 If the 95\% confidence interval for the slope did not overlap zero, we considered the slope to be significant.
@@
   "Plasticity", "$\\mathrm{AA}_{\\text{sun}} > 0$", "\\aax{} in sun leaves is greater than that in shade leaves",
-  "Constitutive", "$\\beta_{\\mathrm{PAI,AA}} > 0$", "Population-level \\aax{} ($\\mathrm{AA}_\\text{pop}$) decreases with native PAI",
+  "Constitutive", "$\\beta_{\\mathrm{PAI,AA}} < 0$", "Population-level \\aax{} ($\\mathrm{AA}_\\text{pop}$) decreases with native PAI",
   "Constitutive", "", "$\\mathrm{AA}_\\text{pop} = \\beta_{\\mathrm{AA}, 0} + \\vec{\\beta}_{\\mathrm{AA}, \\text{pop}}$",
@@
   "Acclimation $\\times$ Constitutive", "$\\mathrm{AA}_{2000} > 0$", "\\aax{} at high light intensity is greater than that at low light intensity",
-  "Acclimation $\\times$ Constitutive", "$\\beta_{\\mathrm{PAI,AA}} > 0$", "Population-level \\aax{} at high light intensity ($\\mathrm{AA}_{\\text{pop},2000}$) decreases with native PAI",
+  "Acclimation $\\times$ Constitutive", "$\\beta_{\\mathrm{PAI,AA}} < 0$", "Population-level \\aax{} at high light intensity ($\\mathrm{AA}_{\\text{pop},2000}$) decreases with native PAI",
   "Acclimation $\\times$ Constitutive", "", "$\\mathrm{AA}_{\\text{pop},2000} = \\beta_{\\mathrm{AA}, 0} + \\vec{\\beta}_{\\mathrm{AA}, \\text{pop}} + \\vec{\\beta}_{\\mathrm{AA}, 2000, \\text{pop}}$",
   "Plasticity $\\times$ Constitutive", "$\\mathrm{AA}_{\\text{sun}} > 0$", "\\aax{} in sun leaves is greater than that in shade leaves",
-  "Plasticity $\\times$ Constitutive", "$\\beta_{\\mathrm{PAI,AA}} > 0$", "Population-level \\aax{} in sun leaves ($\\mathrm{AA}_{\\text{pop},\\text{sun}}$) decreases with native PAI",
+  "Plasticity $\\times$ Constitutive", "$\\beta_{\\mathrm{PAI,AA}} < 0$", "Population-level \\aax{} in sun leaves ($\\mathrm{AA}_{\\text{pop},\\text{sun}}$) decreases with native PAI",
   "Plasticity $\\times$ Constitutive", "", "$\\mathrm{AA}_{\\text{pop},\\text{sun}} = \\beta_{\\mathrm{AA}, 0} + \\vec{\\beta}_{\\mathrm{AA}, \\text{pop}} + \\vec{\\beta}_{\\mathrm{AA}, \\text{sun}, \\text{pop}}$",
@@
 
-### Estimating the water cost of hypostomy
-
-Amphistomy advantage implies that hypostomatous leaves would need to lose more water by having higher \gsw{} to achieve the same photosynthetic rate.
+### The water cost of hypostomy
+
+Amphistomy advantage implies that hypostomatous leaves would need to lose more water by having higher \gsw{} to achieve the same photosynthetic rate as an otherwise identical amphistomatous leaf.
 We estimated this cost of hypostomy approximately.
-For simpliticity, we focus on a sufficiently small region of the \agcurve{} curve such that we can approximate the relationship between \gsw{} and $A$ as log-log linear.
-We also assume that slope is same in both amphistomatous and hypostomatous leaves, which was often approximately the case in our experiment (@fig-curves).
+For simplicity, we focus on a sufficiently small region of the \agcurve{} curve such that we can approximate the relationship between \gsw{} and $A$ as log-log linear.
+We also assume that the slope is the same in both amphistomatous and hypostomatous leaves, which was often approximately the case in our experiment (@fig-curves).
 With these assumptions, the approximate water cost of hypostomy can be derived from the equations:
@@
 $$-\log\left(\frac{g_\text{sw,amphi}}{g_\text{sw,hypo}}\right) = \frac{\mathrm{AA}}{\varepsilon_g}.$$
-We note that the slope in this case is an elasticity of of $A$ to \gsw{} because both are log-transformed.
-This quantity will be proportional the water cost of hypostomy assuming that VPD and other environmental variables are held constant and boundary layer conductance are high enough to be negligible.
+We note that the slope in this case is the elasticity of $A$ to \gsw{} because both are log-transformed.
+This quantity will be proportional the water cost of hypostomy assuming that VPD and other environmental variables are held constant and the boundary layer resistance is negligible.
 
@@
 -->
-
-## Results
 
@@
 #| tbl-cap: "Parameter estimates and 95% confidence intervals (CIs) from the posterior distribution of the selected model with leaf mass per area (LMA) as potential mediator variable.
-Models potentially include fixed, random, and distributional effects of growth light intensity (shade and sun), measurement light intensity (low and high), $\\log \\mathrm{LMA}$, and their interactions on amphistomy advantage ($\\mathrm{AA}$).
-The distributional effects refer to the effect of factors on residual variation in $\\mathrm{AA}$, denoted $\\sigma$, on a log-link scale.
+Models potentially include fixed and random effects of growth light intensity (shade and sun), measurement light intensity (low and high), $\\log \\mathrm{LMA}$, and their interactions on amphistomy advantage ($\\mathrm{AA}$).
 LMA can also be affected by growth light intensity and population.
@@
 Only the top models ($\\Delta \\text{LOOIC} < 2$) are shown.
-The model with the lowest LOOIC is considered the best fit to the data.
+The model with the lowest LOOIC was selected for generating posterior predictions to test hypotheses.
 The LOOIC column lists the LOOIC value for each model; the $\\Delta$LOOIC column lists the difference between the LOOIC of the corresponding model and the best model (lowest LOOIC).
@@
 Only the top models ($\\Delta~\\text{LOOIC} < 2$) are shown.
-The model with the lowest LOOIC is considered the best fit to the data.
+The model with the lowest LOOIC was selected for generating posterior predictions to test hypotheses.
 The LOOIC column lists the LOOIC value for each model; the $\\Delta$LOOIC column lists the difference between the LOOIC of the corresponding model and the best model (lowest LOOIC).
@@
 Points are raw data, lines are fitted curves, and shaded regions are 95% confidence ribbons for the fitted curves.
-The Bayesian correlation coefficient $r^2$ is shown to the right of each curve, along with the estimate of \\aax{} $\\pm$ two standard errors for each pair of amphi (orange) and pseudohypo (blue) curves."
+The Bayesian correlation coefficient, $r^2$, is shown to the right of each curve, along with the estimate of \\aax{} $\\pm$ two standard errors for each pair of amphi (orange) and pseudohypo (blue) curves."
 ggplot()

### 2025-08-14 ms/ms.tex
proofread supporting materials

@@
 
+\bibitem[\citeproctext]{ref-pease_phylogenomics_2016}
+\CSLLeftMargin{54. }%
+\CSLRightInline{J. B. Pease, D. C. Haak, M. W. Hahn, L. C. Moyle,
+\href{https://doi.org/10.1371/journal.pbio.1002379}{Phylogenomics
+reveals three sources of adaptive variation during a rapid radiation}.
+\emph{PLOS Biology} \textbf{14}, e1002379 (2016).}
+
 \bibitem[\citeproctext]{ref-schoch_dependence_1980}
-\CSLLeftMargin{54. }%
+\CSLLeftMargin{55. }%
 \CSLRightInline{P.-G. Schoch, C. Zinsou, M. Sibi,
@@
 \bibitem[\citeproctext]{ref-sack_developmental_2016}
-\CSLLeftMargin{55. }%
+\CSLLeftMargin{56. }%
 \CSLRightInline{L. Sack, T. N. Buckley,
@@
 \bibitem[\citeproctext]{ref-mott_amphistomy_1991}
-\CSLLeftMargin{56. }%
+\CSLLeftMargin{57. }%
 \CSLRightInline{K. A. Mott, O. Michaelson, Amphistomy as an adaptation
@@
 \bibitem[\citeproctext]{ref-schindelin_fiji_2012}
-\CSLLeftMargin{57. }%
+\CSLLeftMargin{58. }%
 \CSLRightInline{J. Schindelin, I. Arganda-Carreras, E. Frise, V. Kaynig,
@@
 \bibitem[\citeproctext]{ref-monteith_principles_2013}
-\CSLLeftMargin{58. }%
+\CSLLeftMargin{59. }%
 \CSLRightInline{J. L. Monteith, M. H. Unsworth, \emph{Principles of
@@
 \bibitem[\citeproctext]{ref-stan_development_team_stan_2025}
-\CSLLeftMargin{59. }%
+\CSLLeftMargin{60. }%
 \CSLRightInline{Stan Development Team, \emph{Stan {Modeling} {Language}
@@
 \bibitem[\citeproctext]{ref-burkner_brms_2017}
-\CSLLeftMargin{60. }%
+\CSLLeftMargin{61. }%
 \CSLRightInline{P.-C. Bürkner,
@@
 \bibitem[\citeproctext]{ref-gabry_cmdstanr_2025}
-\CSLLeftMargin{61. }%
+\CSLLeftMargin{62. }%
 \CSLRightInline{J. Gabry, R. Češnovar, A. Johnson, S. Bronder,
@@
 \bibitem[\citeproctext]{ref-r_core_team_r:_2025}
-\CSLLeftMargin{62. }%
+\CSLLeftMargin{63. }%
 \CSLRightInline{R Core Team, \emph{R: {A} {Language} and {Environment}
@@
 \bibitem[\citeproctext]{ref-gelman_inference_1992}
-\CSLLeftMargin{63. }%
+\CSLLeftMargin{64. }%
 \CSLRightInline{A. Gelman, D. B. Rubin, Inference from iterative
@@
 \textbf{7}, 457--472 (1992).}
-
-\bibitem[\citeproctext]{ref-pease_phylogenomics_2016}
-\CSLLeftMargin{64. }%
-\CSLRightInline{J. B. Pease, D. C. Haak, M. W. Hahn, L. C. Moyle,
-\href{https://doi.org/10.1371/journal.pbio.1002379}{Phylogenomics
-reveals three sources of adaptive variation during a rapid radiation}.
-\emph{PLOS Biology} \textbf{14}, e1002379 (2016).}
 
@@
 lycopersicum} var. \emph{lycopersicum} (Table~\ref{tbl-populations}).
-Due to constraints on growth space and time, we spread out measurements
-over 61.1 weeks from August 29, 2022 to October 31, 2023. Replicates
-within population were evenly spread out over this period to prevent
-confounding of temporal variation in growth conditions with accession.
-{[}anything else to say here? maybe explain population selection and
-phylogeny?{]}
+These populations were selected in a previous study because they
+encompass the breadth of climatic variation in the wild tomato clade
+(\emph{54}). In one case, we substituted a congeric population of the
+species (\emph{S. galapagense}) because of difficult growing the focal
+population. We also added an a population of \emph{S. pennellii}. Due to
+constraints on growth space and time, we spread out measurements over
+61.1 weeks from August 29, 2022 to October 31, 2023. Replicates within
+population were evenly spread out over this period to prevent
+confounding of temporal variation in growth conditions with variation
+among populations.
 
@@
 environmental measurements were taken every 10 minutes from the middle
-of plants racks at approximately the same height as the leaves we
+of plant racks at approximately the same height as the leaves we
 measured. We measured leaf temperature of focal leaves prior to
@@
 \(\mathrm{PPFD} = \qty{267}{\micro\mol\raiseto{-2}\meter\raiseto{-1}\second}\)
-(Fluence RAZRx, Austin, Texas, USA).
+(Fluence RAZRx, Austin, Texas, USA) during the germination and seedling
+stages.
 
@@
 intensity of the treatment rather than the seedling conditions
-(\emph{54}). Shade plants grew slower than sun plants, hence leaves at
+(\emph{55}). Shade plants grew slower than sun plants, hence leaves at
 the same developmental stage were measured on chronologically older
@@
 
-\paragraph{Light-saturated photosynthetic
-rate}\label{light-saturated-photosynthetic-rate}
-
-In 91.3\% of plants, we measured the maximum photosynthetic rate
-(\(A_\mathrm{max}\)) on the same leaflets as \agcurve{} curves. However,
-when a leaflet was damaged during \agcurve{} curves, we used the next
-closest leaflet for \(A_\mathrm{max}\). Leaves acclimated to high
-light-intensity (\ppfdequals{2000}), ambient CO\(_2\) (\caequals{415}),
-\rhequals{50}, and \tleafequals{25}. After \(A\) and \gsw{} stabilized,
-we measured \(A\) at
-\(\qty{2000}{\micro\mol\raiseto{-2}\meter\raiseto{-1}\second}\).
-
 \paragraph{Stomatal anatomy}\label{stomatal-anatomy}
@@
 stomatal size since it is proportional to maximum conductance
-(\emph{55}). We made surface impressions of leaf lamina from the same
+(\emph{56}). We made surface impressions of leaf lamina from the same
 area used for gas exchange measurements using a-silicone impression
@@
 After nail polish dried, we mounted replicas on a microscope slide using
-transparent tape (\emph{56}). We digitized a portion of each leaf
+transparent tape (\emph{57}). We digitized a portion of each leaf
 surface replica using a brightfield microscope (Leica DM2000, Wetzlar,
 Germany). We counted and measured guard cell length on all stomata using
-the FIJI implementation of ImageJ2 version 2.3.0 (\emph{57}), then
+the FIJI implementation of ImageJ2 version 2.3.0 (\emph{58}), then
 divided the count by the visible leaf area
@@
 conductance (\gmax) at a reference leaf temperature of
-\(\qty{25}{\degreeCelsius}\) following (\emph{55}) as:
+\(\qty{25}{\degreeCelsius}\) following (\emph{56}) as:
 
@@
 \(D_\mathrm{wv} = \qty{2.49e-5}{\meter\squared\per\second}\) and
-\(v = \qty{2.24e-2}{\raiseto{3}\meter\per\mol}\) (\emph{58}). For
+\(v = \qty{2.24e-2}{\raiseto{3}\meter\per\mol}\) (\emph{59}). For
 kidney-shaped guard cells like those in wild tomatoes,
@@
 California. USA) and measured leaflet area from digital images using the
-FIJI implementation of ImageJ2 version 2.3.0 (\emph{57}). We dried
+FIJI implementation of ImageJ2 version 2.3.0 (\emph{58}). We dried
 leaves for 72 hours at \(\qty{74}{\degreeCelsius}\) in a food dehydrator
@@
 
-All models were using a Bayesian model with HMC sampling in the
-probabilistic programming language \emph{Stan} (\emph{59}) using the
-\emph{R} package \textbf{brms} version 2.22.0 (\emph{60}). We used
-\emph{CmdStan} version 2.36.0 and \textbf{cmdstanr} version 0.9.0
-(\emph{61}) to interface with \emph{R} version 4.5.1 (\emph{62}). We
-sampled the posterior distribution from a single chain with a minimum of
-1000 iterations after 1000 warmup iterations. If necessary, we refit
-models with more iterations to decrease the Gelman-Rubin convergence
-statistic (\(\hat{R}\)) (\emph{63}) to less than 1.01 and the bulk
-effective sample size (ESS) to greater than 400 for all model
-parameters, and there were fewer than 10 divergent transitions.
+All models were fit using a Bayesian HMC sampling in the probabilistic
+programming language \emph{Stan} (\emph{60}) using the \emph{R} package
+\textbf{brms} version 2.22.0 (\emph{61}). We used \emph{CmdStan} version
+2.36.0 and \textbf{cmdstanr} version 0.9.0 (\emph{62}) to interface with
+\emph{R} version 4.5.1 (\emph{63}). We sampled the posterior
+distribution from a single chain with a minimum of 1000 iterations after
+1000 warmup iterations. If necessary, we refit models with more
+iterations to decrease the Gelman-Rubin convergence statistic
+(\(\hat{R}\)) (\emph{64}) to less than 1.01 and the bulk effective
+sample size (ESS) to greater than 400 for all model parameters, and
+there were fewer than 10 divergent transitions.
 
@@
 using the standard deviation of the posterior distribution of each
-\(\widehat{\mathrm{AA}}\) estimate (Section~\ref{sec-aa-est}). We
-crossed fixed and random effect structures with fixed effects of
-measurement light intensity and/or growth light intensity on the
-distributional parameter, \(\sigma\), the residual variance in \aax{}.
-We used a robust regression approach by assuming \(t\)-distributed
-residuals, which makes estimates less susceptible to extreme values. In
-total, we fit 100 models, each with a different combination of fixed,
-random, and distributional effects. We used the RAxML
-whole-transcriptome concatenated phylogeny based on 2,745 100-kb genomic
-windows from (\emph{64}). Two of our populations were not in this tree.
-We used accession LA1044 in place of LA3909, two populations of \emph{S.
-galapagnse}; LA0750 was added as sister to LA0716, two closely related
-populations of \emph{S. pennellii}. The node separating LA0750 from
-LA0716 was placed half-way between the next deepest node. We used the
-leave-one-out cross-validation information criterion (LOOIC) to compare
-the fit of models (\autoref{tbl-aa_loo1}) using the \emph{R} package
-\textbf{loo} version 2.8.0 (\emph{65}) to calculate LOOIC values. We
-selected the model with the lowest LOOIC value (Table~\ref{tbl-aa_loo1})
-to generate posterior predictions for hypothesis testing. Parameter
-estimates from the selected model are shown in Table~\ref{tbl-fit_aa1}.
+\(\widehat{\mathrm{AA}}\) estimate (see above). We crossed fixed and
+random effect structures with fixed effects of measurement light
+intensity and/or growth light intensity on the distributional parameter,
+\(\sigma\), the residual variance in \aax{}. We used a robust regression
+approach by assuming \(t\)-distributed residuals, which makes estimates
+less susceptible to extreme values. In total, we fit 100 models, each
+with a different combination of fixed, random, and distributional
+effects. We used the RAxML whole-transcriptome concatenated phylogeny
+based on 2,745 100-kb genomic windows from (\emph{54}). Two of our
+populations were not in this tree. We used accession LA1044 in place of
+LA3909, two populations of \emph{S. galapagense}; LA0750 was added as
+sister to LA0716, two closely related populations of \emph{S.
+pennellii}. The node separating LA0750 from LA0716 was placed half-way
+between the next deepest node. We used the leave-one-out
+cross-validation information criterion (LOOIC) to compare the fit of
+models (\autoref{tbl-aa_loo1}) using the \emph{R} package \textbf{loo}
+version 2.8.0 (\emph{65}) to calculate LOOIC values. We selected the
+model with the lowest LOOIC value (Table~\ref{tbl-aa_loo1}) to generate
+posterior predictions for hypothesis testing. Parameter estimates from
+the selected model are shown in Table~\ref{tbl-fit_aa1}.
 
@@
 data using the \texttt{mi()} function in \textbf{brms}. Because the
-evidence strong supported including effects of growth and measurement
-light intensity on the residual variance in \(\mathrm{AA}\), we included
-these in terms in all models. In total, we fit 75 models, each with a
-different combination of fixed and random effect structures. We selected
-the model with the lowest LOOIC value (Table~\ref{tbl-aa_loo2}) to
-generate posterior predictions for hypothesis testing. Note that LOOIC
-was calculated only from pointwise likelihood values of the submodel
-estimating effects of light intensity, light treatments, and population
-on \aax{}. Parameter estimates from the selected model are shown in
-Table~\ref{tbl-fit_aa2}.
+evidence from the first set of models strongly supported including
+effects of growth and measurement light intensity on the residual
+variance in \(\mathrm{AA}\), we included these in terms in all models.
+In total, we fit 75 models, each with a different combination of fixed
+and random effect structures. We selected the model with the lowest
+LOOIC value (Table~\ref{tbl-aa_loo2}) to generate posterior predictions
+for hypothesis testing. Note that LOOIC was calculated only from
+pointwise likelihood values of the submodel estimating effects of light
+intensity, light treatments, and population on \aax{}. Parameter
+estimates from the selected model are shown in Table~\ref{tbl-fit_aa2}.
 
@@
 predictions about the relationship between \aax{} and light intensity,
-light treatment, and native \ppfd{} among populations. Since these
-hypotheses are not mutually exclusive, we describe how we assessed
+light treatment, and native \(\mathrm{PAI}\) among populations. Since
+these hypotheses are not mutually exclusive, we describe how we assessed
 support for one, two, or all three hypotheses simultaneously in
@@
 
-From the posterior of the selected model (Section~\ref{sec-phylo}), we
-predicted the posterior distribution of the expected \aax{} for each
-population at each measurement light intensity and growth light
-intensity. If the 95\% confidence intervals for \aax{} did not overlap
-zero, we considered \aax{} to be significantly different from zero. We
-also calculated posterior distribution for the average \aax{} among
-populations in each condition. We tested a linear effect of native light
-habitat on population-level \aax{} in each treatment condition,
-integrating over the entire posterior distribution. We calculated the
-posterior distribution of the slope. If the 95\% confidence interval for
-the slope did not overlap zero, we considered the slope to be
-significant.
+From the model selected using LOOIC, we predicted the posterior
+distribution of the expected \aax{} for each population at each
+measurement light intensity and growth light intensity. If the 95\%
+confidence intervals for \aax{} did not overlap zero, we considered
+\aax{} to be significantly different from zero. We also calculated
+posterior distribution for the average \aax{} among populations in each
+condition. We tested a linear effect of native light habitat on
+population-level \aax{} in each treatment condition, integrating over
+the entire posterior distribution to calculate a posterior distribution
+of the slope. If the 95\% confidence interval for the slope did not
+overlap zero, we considered the slope to be significant.
 
@@
 \cmidrule{1-3}\pagebreak[0]
-Constitutive & $\beta_{\mathrm{PAI,AA}} > 0$ & \hspace{-1em}Population-level \aax{} ($\mathrm{AA}_\text{pop}$) decreases with native PAI\\
+Constitutive & $\beta_{\mathrm{PAI,AA}} < 0$ & \hspace{-1em}Population-level \aax{} ($\mathrm{AA}_\text{pop}$) decreases with native PAI\\
 \nopagebreak
@@
 \nopagebreak
- & $\beta_{\mathrm{PAI,AA}} > 0$ & \hspace{-1em}Population-level \aax{} at high light intensity ($\mathrm{AA}_{\text{pop},2000}$) decreases with native PAI\\
+ & $\beta_{\mathrm{PAI,AA}} < 0$ & \hspace{-1em}Population-level \aax{} at high light intensity ($\mathrm{AA}_{\text{pop},2000}$) decreases with native PAI\\
 \nopagebreak
@@
 \nopagebreak
- & $\beta_{\mathrm{PAI,AA}} > 0$ & \hspace{-1em}Population-level \aax{} in sun leaves ($\mathrm{AA}_{\text{pop},\text{sun}}$) decreases with native PAI\\
+ & $\beta_{\mathrm{PAI,AA}} < 0$ & \hspace{-1em}Population-level \aax{} in sun leaves ($\mathrm{AA}_{\text{pop},\text{sun}}$) decreases with native PAI\\
 \nopagebreak
@@
 
-\subsubsection{Estimating the water cost of
-hypostomy}\label{estimating-the-water-cost-of-hypostomy}
+\subsubsection{The water cost of
+hypostomy}\label{the-water-cost-of-hypostomy}
 
@@
 lose more water by having higher \gsw{} to achieve the same
-photosynthetic rate. We estimated this cost of hypostomy approximately.
-For simpliticity, we focus on a sufficiently small region of the
-\agcurve{} curve such that we can approximate the relationship between
-\gsw{} and \(A\) as log-log linear. We also assume that slope is same in
-both amphistomatous and hypostomatous leaves, which was often
-approximately the case in our experiment (Figure~\ref{fig-curves}). With
-these assumptions, the approximate water cost of hypostomy can be
-derived from the equations:
+photosynthetic rate as an otherwise identical amphistomatous leaf. We
+estimated this cost of hypostomy approximately. For simplicity, we focus
+on a sufficiently small region of the \agcurve{} curve such that we can
+approximate the relationship between \gsw{} and \(A\) as log-log linear.
+We also assume that the slope is the same in both amphistomatous and
+hypostomatous leaves, which was often approximately the case in our
+experiment (Figure~\ref{fig-curves}). With these assumptions, the
+approximate water cost of hypostomy can be derived from the equations:
 
@@
 \[-\log\left(\frac{g_\text{sw,amphi}}{g_\text{sw,hypo}}\right) = \frac{\mathrm{AA}}{\varepsilon_g}.\]
-We note that the slope in this case is an elasticity of of \(A\) to
-\gsw{} because both are log-transformed. This quantity will be
-proportional the water cost of hypostomy assuming that VPD and other
-environmental variables are held constant and boundary layer conductance
-are high enough to be negligible.
-
-\subsection{Results}\label{results}
+We note that the slope in this case is the elasticity of \(A\) to \gsw{}
+because both are log-transformed. This quantity will be proportional the
+water cost of hypostomy assuming that VPD and other environmental
+variables are held constant and the boundary layer resistance is
+negligible.
 
@@
 with leaf mass per area (LMA) as potential mediator variable. Models
-potentially include fixed, random, and distributional effects of growth
-light intensity (shade and sun), measurement light intensity (low and
-high), \(\log \mathrm{LMA}\), and their interactions on amphistomy
-advantage (\(\mathrm{AA}\)). The distributional effects refer to the
-effect of factors on residual variation in \(\mathrm{AA}\), denoted
-\(\sigma\), on a log-link scale. LMA can also be affected by growth
-light intensity and population. For phylogenetically structured random
-effects, we report the estimated standard deviation (SD).}
+potentially include fixed and random effects of growth light intensity
+(shade and sun), measurement light intensity (low and high),
+\(\log \mathrm{LMA}\), and their interactions on amphistomy advantage
+(\(\mathrm{AA}\)). LMA can also be affected by growth light intensity
+and population. For phylogenetically structured random effects, we
+report the estimated standard deviation (SD).}
 
@@
 (\(\Delta \text{LOOIC} < 2\)) are shown. The model with the lowest LOOIC
-is considered the best fit to the data. The LOOIC column lists the LOOIC
-value for each model; the \(\Delta\)LOOIC column lists the difference
-between the LOOIC of the corresponding model and the best model (lowest
-LOOIC). Checkmarks in each row indicate whether the corresponding model
-included affects of measurement light intensity (``Meas.'') or growth
-light intensity (``Growth'') as fixed effects, random effects, or
-distributional effect on the residual variance. We also test for fixed
-and random interaction effects (``Inter.'') between measurement and
-growth light intensity. All models included phylogenetically structured
-random effects of population.}
+was selected for generating posterior predictions to test hypotheses.
+The LOOIC column lists the LOOIC value for each model; the
+\(\Delta\)LOOIC column lists the difference between the LOOIC of the
+corresponding model and the best model (lowest LOOIC). Checkmarks in
+each row indicate whether the corresponding model included affects of
+measurement light intensity (``Meas.'') or growth light intensity
+(``Growth'') as fixed effects, random effects, or distributional effect
+on the residual variance. We also test for fixed and random interaction
+effects (``Inter.'') between measurement and growth light intensity. All
+models included phylogenetically structured random effects of
+population.}
 
@@
 models (\(\Delta~\text{LOOIC} < 2\)) are shown. The model with the
-lowest LOOIC is considered the best fit to the data. The LOOIC column
-lists the LOOIC value for each model; the \(\Delta\)LOOIC column lists
-the difference between the LOOIC of the corresponding model and the best
-model (lowest LOOIC). Checkmarks in each row indicate whether the
-corresponding model included affects of measurement light intensity
-(``Meas.''), growth light intensity (``Growth''), or LMA as fixed or
-random effects. We test for fixed interaction effects between
-measurement and growth light intensity and LMA. We also test for random
-interaction effects between measurement and growth light intensity. All
-models included phylogenetically structured random effects of
-population.}
+lowest LOOIC was selected for generating posterior predictions to test
+hypotheses. The LOOIC column lists the LOOIC value for each model; the
+\(\Delta\)LOOIC column lists the difference between the LOOIC of the
+corresponding model and the best model (lowest LOOIC). Checkmarks in
+each row indicate whether the corresponding model included affects of
+measurement light intensity (``Meas.''), growth light intensity
+(``Growth''), or LMA as fixed or random effects. We test for fixed
+interaction effects between measurement and growth light intensity and
+LMA. We also test for random interaction effects between measurement and
+growth light intensity. All models included phylogenetically structured
+random effects of population.}
 
@@
 
-\includegraphics{ms_files/figure-pdf/fig-curves-1.pdf}
+\pandocbounded{\includegraphics[keepaspectratio]{ms_files/figure-pdf/fig-curves-1.pdf}}
 
@@
 curves, and shaded regions are 95\% confidence ribbons for the fitted
-curves. The Bayesian correlation coefficient \(r^2\) is shown to the
+curves. The Bayesian correlation coefficient, \(r^2\), is shown to the
 right of each curve, along with the estimate of \aax{} \(\pm\) two

### 2025-08-14 r/400_fit-sk.R

@@ -1,227 +1,0 @@
-# modeling ideas (move elsewhere, material below is good)
-# based on eq. A2 in Buckley et al. (2023)
-# X is osmolality gradient (delta pi in buckley paper)
-# X_f is target osmolality
-# dX / dt = a (X - X_f)  # buckley paper, results in Vico eqn I think
-#
-# Now suppose that a is a function of X
-# a = a0 + a1 * X
-# dX / dt = (a0 + a1 * X) * (X - X_f)
-# this only has implicit soln according to ChatGpt.
-# alternate forms usually result in sigmoidal
-# 
-# what if we go back to dX / dt = a (X - X_f) but assume that d X_f / dt = b (X_f - X_inf)
-# we get this according to chatGPT:
-solve_X_dynamics <- function(t, X0, Xf0, X_inf, a, b) {
-  if (a == b) {
-    # Special case: a == b
-    term1 <- X_inf
-    term2 <- a * (Xf0 - X_inf) * t * exp(-a * t)
-    term3 <- (X0 - X_inf) * exp(a * t)
-    return(term1 + term2 + term3)
-  } else {
-    # General case: a != b
-    term1 <- X_inf
-    term2 <- (a * (Xf0 - X_inf)) / (b - a) * exp((b - 2 * a) * t)
-    term3 <- (X0 - X_inf + (a * (Xf0 - X_inf)) / (b - a)) * exp(a * t)
-    return(term1 - term2 + term3)
-  }
-}
-
-# Fit CDWeibulll mode to each curve separately
-source("r/header.R")
-
-sk_dir = "objects/sk-curves/"
-if (!dir.exists(sk_dir)) {
-  dir.create(sk_dir)
-}
-
-rh_curves = read_rds("data/trimmed_rh_curves.rds") |>
-  mutate(ci = as.numeric(as.factor(curve))) |>
-  mutate(t_sec = elapsed - min(elapsed), .by = ci)
-  
-form_vico = gsw ~ gf + dg * exp(-(t_sec / tau))
-form_cdweibull = gsw ~ gf + dg * exp(-(t_sec / tau) ^ lambda)
-
-bform_vico = bf(
-  form_vico,
-  gf ~ 1,
-  logdg ~ 1,
-  tau ~ 1,
-  nl = TRUE
-)
-
-bform_cdweibull = bf(
-  form_cdweibull,
-  gf ~ 1,
-  logdg ~ 1,
[…]

### 2025-08-14 r/401_tbl-parameters.R

@@ -1,21 +1,0 @@
-source("r/header.R")
-
-
-df_parameters = read_csv("data/parameters1.csv", col_types = "cccc", n_max = 8) 
-
-min(which(df_parameters$par_type == "ags"))
-
-read_csv("data/parameters1.csv", col_types = "cccc", n_max = 6) 
-
-  
-  # 1. Estimate \agcurve{} curve parameters for each leaf
-    # B_curve, n_curve x 3 array of random \agcurve{} curve-level coefficients (b_{0,j}, b_{1,j}, b_{2,j}). B_curve ~ MVN(0, S_curve)
-    # Mu_curve, vector of mean quadratic coefficients (\Beta_0, \Beta_1, \Beta_2) 
-    # 3 S_curve, covariance matrix of B_curve                  
-    # 5 b0_log_sigma_resid       
-    # 6 b_log_sigma_resid_S      
-    # 7 rho_resid                
-    2. Estimate \aax{} for each light intensity with leaf using \agcurve{} curve parameters
-  3. Estimate the effects of light intensity, light treatment, and accession on \aax{} (assimilatory and plasticity hypotheses)
-  4. Estimate the effects of native light habitat on accession-level \aax{} (constitutive hypothesis)
-  

### 2025-08-19 ms/ms.qmd
cleaning up

@@
 We consider three nonmutually exclusive hypotheses that we classify as 'acclimation', 'plasticity', and 'constitutive' ([Fig.
-@fig-concept]).
+@fig-hypotheses]).
 
@@
 
-```{r, concept, echo = FALSE}
-#| label: fig-concept
+```{r, hypotheses, echo = FALSE}
+#| label: fig-hypotheses
 #| fig-align: center
@@
 
-knitr::include_graphics("../figures/amphisomy draft 2024-03-28.jpg")
+knitr::include_graphics("../figures/hypotheses.pdf")
 ```

### 2025-08-19 ms/ms.tex
cleaning up

@@
 exclusive hypotheses that we classify as `acclimation', `plasticity',
-and `constitutive' (Fig.~\ref{fig-concept}).
+and `constitutive' (Fig.~\ref{fig-hypotheses}).
 
@@
 
-\includegraphics[width=2.24in,height=\textheight,keepaspectratio]{../figures/amphisomy draft 2024-03-28.jpg}
+\pandocbounded{\includegraphics[keepaspectratio]{../figures/hypotheses.pdf}}
 
@@
 
-\caption{\label{fig-concept}\textbf{Conceptual outline of three nonmutually exclusive hypotheses explaining why amphistomy advantage (\aax{}) might be greater for leaves in sunny, open habitats.}
+\caption{\label{fig-hypotheses}\textbf{Conceptual outline of three nonmutually exclusive hypotheses explaining why amphistomy advantage (\aax{}) might be greater for leaves in sunny, open habitats.}
 The acclimatopn hypothesis predicts that \aax{} is greater under high

### 2025-08-19 r/02_make-phy.R

@@ -1,4 +1,4 @@
-# Prepare data for analysis
+# Make phylogeny for analysis
 source("r/header.R")
 
 read_rds("data/thinned_rh_curves.rds") |>

### 2025-08-19 r/06_calculate-aa-posterior.R

@@ -60,7 +60,7 @@ aa_post = full_join(curve_fits_draws,
 # Check that AA estimates converged ----
 aa_summary = aa_post |>
   dplyr::select(starts_with("."), acc_id, aa, light_intensity) |>
-  split(~ acc_id + light_intensity) |>
+  split( ~ acc_id + light_intensity) |>
   map_dfr(\(.x) {
     .x |>
       select(-acc_id, -light_intensity) |>

### 2025-08-19 r/10_fit-stomata.R

@@ -13,11 +13,9 @@ plant_info = read_rds("data/plant-info.rds") |>
 df_stomata = read_rds("data/stomata.rds") |>
   left_join(plant_info, by = join_by(accession, replicate)) |>
   rename(acc = accession) |>
-  filter(
-    acc %in% unique(d1$acc),
-    if_all(contains("stomatal_density"), ~ !is.na(.x)),
-    if_all(contains("guard_cell_length"), ~ !is.na(.x))
-  ) |>
+  filter(acc %in% unique(d1$acc),
+         if_all(contains("stomatal_density"), ~ !is.na(.x)),
+         if_all(contains("guard_cell_length"), ~ !is.na(.x))) |>
   mutate(
     lower_sd = log(lower_stomatal_density_mm2),
     upper_sd = log(upper_stomatal_density_mm2),
@@ -42,7 +40,7 @@ model_forms = expand.grid(
   mutate(seed = sample(1e9, nrow(.)))
 
 # Build and fit each model
-plan(multisession, workers = 8)
+plan(multisession, workers = 19)
 
 stomata_models = future_pmap(model_forms, function(fixed, random, sigma, seed) {
   fml = bf(as.formula(
@@ -88,10 +86,7 @@ stomata_loo_table = tibble(
   se = sapply(loos, \(x) x$estimates["looic", "SE"])
 ) |>
   arrange(looic) |>
-  mutate(
-    delta_looic = looic - min(looic),
-    best_model = delta_looic == 0
-  )
+  mutate(delta_looic = looic - min(looic), best_model = delta_looic == 0)
 
 write_rds(stomata_loo_table, "objects/stomata_loo_table.rds")
 best_model_index = str_extract(stomata_loo_table[1, "model"], "\\d+") |>

### 2025-08-19 r/11_plot-aa.R

@@ -135,24 +135,3 @@ ggsave(
 write_rds(df_aa_pred1, "objects/df_aa_pred1.rds")
 write_rds(df_aa_pred2, "objects/df_aa_pred2.rds")
 write_rds(df_aa_pred3, "objects/df_aa_pred3.rds")
-
-# Plot for amphistomy-proposal
-# set.seed(07102025)
-# amphi_advantage = ggplot(df_aa_pred1, aes(light_intensity, aa, color = light_treatment, shape = light_intensity)) +
-#   facet_wrap(~Growth) +
-#   geom_jitter(width = 0.1, height = 0, size = 2, fill = "white") +
-#   stat_summary(
-#     fun.data = mean_cl_boot,
-#     fun.args = list(conf.int = 0.95),
-#     geom = "pointinterval",
-#     position = position_nudge(x = -0.3),
-#     fill = "white", size = 12, linewidth = 1
-#   ) +
-#   geom_hline(yintercept = 0, linetype = "dashed") +
-#   scale_color_manual(values = c("low" = "tomato4", "high" = "tomato")) +
-#   scale_shape_manual(values = c("150" = 19, "2000" = 21)) +
-#   xlab(expression(Light~intensity~(mu*mol~m^-2~s^-1))) +
-#   ylab("Amphistomy advantage") +
-#   theme(legend.position = "none", strip.background = element_blank())
-# 
-# write_rds(amphi_advantage, "objects/amphi_advantage.rds")

### 2025-08-19 r/15_plot-curves.R

@@ -26,7 +26,7 @@ assert_true(all(!is.na(fit_curve_diagnostics$acc_id))) # Check for NA values in 
 assert_true(all(!is.na(fit_curve_diagnostics$`Curve type`))) # Check for NA values in `Curve type`
 assert_true(all(!is.na(fit_curve_diagnostics$light_intensity))) # Check for NA values in light_intensity
 
-plan(multisession, workers = 19)
+plan(multisession, workers = 1)
 ags_curves = future_map(
   unique(aa_summary$acc_id),
   \(.x) {

### 2025-08-19 r/header.R

@@ -1,7 +1,6 @@
 rm(list = ls())
 
 library(ape)
-library(bayesplot)
 library(brms)
 library(checkmate)
 library(cmdstanr)
@@ -15,17 +14,14 @@ library(grid)
 library(loo)
 library(lubridate)
 library(magrittr)
-library(mgcv)
-library(mvnfast)
 library(phangorn)
-library(progressr)
+library(posterior)
 library(purrr)
 library(readr)
 library(rlang)
-library(rootSolve)
+library(rstan)
 library(scales)
 library(stringr)
-library(tibble)
 library(tidybayes)
 library(tidyr)
 library(TreeTools)
@@ -49,20 +45,3 @@ aa_args = read_rds("objects/aa_args.rds") |>
   )
 
 write_rds(aa_args, "objects/aa_args1.rds")
-
-# Fit Rubisco model
-x_depth = c(145, 830.5)
-y_depth = c(0, 700)
-x_rubisco = c(558, 56.5)
-y_rubisco = c(0, 100)
-nishio_carbon_1993_fig5 = read_csv("data/nishio_carbon_1993_fig5.csv",
-                                   show_col_types = FALSE) |>
-  mutate(
-    depth = raw_depth * (diff(y_depth) / diff(x_depth)) - x_depth[1],
-    rubisco = raw_rubisco * (diff(y_rubisco) / diff(x_rubisco)) -
-      x_rubisco[1] * (diff(y_rubisco) / diff(x_rubisco)),
-    # Reversing so that 0 is at abaxial, 1 is at adaxial, to match Earles et al.
-    rel_depth = rev(depth / (min(depth) + max(depth)))
-  )
-
-fit_rubisco = gam(rubisco ~ s(rel_depth), data = nishio_carbon_1993_fig5)

### 2025-08-19 r/install-packages.R

@@ -1,0 +1,49 @@
+# install_packages.R
+
+# Combine both sets of packages and remove duplicates
+pkgs <- unique(c(
+  "ape",
+  "brms",
+  "checkmate",
+  "cmdstanr",
+  "cowplot",
+  "dplyr",
+  "forcats",
+  "furrr",
+  "ggplot2",
+  "glue",
+  "loo",
+  "lubridate",
+  "magrittr",
+  "phangorn",
+  "posterior",
+  "purrr",
+  "readr",
+  "rlang",
+  "rstan",
+  "scales",
+  "stringr",
+  "tidybayes",
+  "tidyr",
+  "TreeTools",
+  "kableExtra",
+  "knitr",
+  "tibble"
+))
+
+# Install any missing packages from CRAN
+install_if_missing <- function(pkg) {
+  if (!requireNamespace(pkg, quietly = TRUE)) {
+    install.packages(pkg)
+  }
+}
+
+# Run install
+invisible(lapply(pkgs, install_if_missing))
+
+# Special handling for cmdstanr (on CRAN but often recommended via GitHub for latest)
+if (!requireNamespace("cmdstanr", quietly = TRUE)) {
+  install.packages("cmdstanr", repos = c("https://mc-stan.org/r-packages/", getOption("repos")))
+}
+
+message("All requested packages are installed.")

### 2025-08-19 r/scratch - calcs.R

@@ -1,53 +1,0 @@
-# simplified back-of the envelope calculation for how much water could be save by switching to amphistomy
-source("r/header.R")
-
-rh_curves = read_rds("data/trimmed_rh_curves.rds")
-tr = read_rds("data/phylogeny.rds")
-A = vcv(tr, corr = TRUE)
-
-plan(multisession, workers = 9)
-
-set.seed(125691028)
-tibble(acc = unique(rh_curves$acc), seed = sample(1e9, length(unique(rh_curves$acc)))) |>
-  future_pwalk(function(acc, seed) {
-    .acc = acc
-    crit = 0
-    x = 1
-    
-    while (crit == 0 & x < 24) {
-      m = brm(
-        formula = log_A ~ leaf_type + leaf_type:light_intensity + light_intensity * log_gsw + (leaf_type + leaf_type:light_intensity + light_intensity * log_gsw |
-                                                                     acc_id),
-        
-        data = filter(rh_curves, acc == .acc),
-        backend = "cmdstanr",
-        chains = 1,
-        iter = x * aa_args$n_iter_init,
-        thin = x,
-        family = student(),
-        save_pars = save_pars(all = TRUE),
-        seed = seed
-      )
-      
-      crit = summarise_draws(m) |>
-        summarize(c1 = (max(rhat, na.rm = TRUE) < aa_args$max_rhat),
-                  c2 = (min(ess_bulk, na.rm = TRUE) > aa_args$min_ess)) |>
-        mutate(
-          n_divergent = nuts_params(m) |>
-            subset(Parameter == "divergent__") |>
-            pull(Value) |>
-            sum(),
-          c3 = (n_divergent < aa_args$max_divergent),
-          c4 = c1 * c2 * c3
-        ) |>
-        pull(c4)
-      
-      x = x + 1
-      
-    }
-    
-    write_rds(m, glue("objects/simple-Ags-fits/{acc}.rds"))
-    
-  },
-  .progress = TRUE,
-  .options = furrr_options(seed = TRUE))

### 2025-08-19 ms/ms.qmd
migrating main text to word

@@
 \begin{center}
-$^1$School of Life Sciences, University of Hawaiʻi, Mānoa, Hawaiʻi, USA\\
-$^2$Department of Botany, University of Wisconsin, Madison, Wisconsin, USA\\
+$^1$School of Life Sciences, University of Hawaiʻi; Mānoa, Hawaiʻi, USA\\
+$^2$Department of Botany, University of Wisconsin, Madison; Wisconsin, USA\\
 \medskip
@@
 
-The presence of stomata on both leaf surfaces (amphistomy) can increase photosynthesis in C$_3$ plants by reducing the path length for CO$_2$ diffusion between substomatal cavities and chloroplasts.
+The presence of stomata on both leaf surfaces (amphistomy) increases photosynthesis by reducing the distance for CO$_2$ diffusion between stomata and chloroplasts.
+Paradoxically, most leaves are hypostomatous (stomata on lower surface), despite the photosynthetic advantage of amphistomy.
+Across `r n_acc` diverse populations of "wild tomatoes", leaves developed under high light intensity benefit more from amphistomy in terms of CO$_2$ assimilated for a given stomatal conductance than plants developed under low light.
+Furthermore, populations native to open habitats benefit more from amphistomy than those from more closed habitats.
+Thus, plasticity and adaptation together may explain why amphistomatous leaves are prevalent in sunny, open habitats, including many crops.
+Contrary to common assumptions, amphistomy can save water because hypostomatous leaves evaporate 10-65\% more to achieve the same photosynthetic rate.
+
+ <!-- LONG FORM - save in case of resubmission -->
+<!-- The presence of stomata on both leaf surfaces (amphistomy) can increase photosynthesis in C$_3$ plants by reducing the path length for CO$_2$ diffusion between substomatal cavities and chloroplasts.
 Paradoxically, despite this photosynthetic advantage, most leaves are hypostomatous (stomata on lower surface only).
@@
 We estimate that hypostomatous leaves would expend 10-65\% more water to achieve the same photosynthetic rate as a comparable amphistomatous leaf, implying large fitness costs of amphistomy in ecological contexts where this trait is rare.
+-->
 
@@
 Understanding how stomata respond to environmental change over daily, developmental, and evolutionary time is important for studying adaptation [@buckley_modelling_2013; @raven_selection_2002; @haworth_co-ordination_2013], inferring paleoclimate from fossil cuticles [@franks_new_2014; @mcelwain_paleoecology_2017; @the_cenozoic_co_proxy_integration_project_cencopip_consortium_toward_2023], predicting responses to climate change [@woodward_stomatal_1987; @liang_stomatal_2023; @chua_stomatal_2024; @lang_century-long_2024], and improving crops [@hofmann_impact_2025].
-Stomatal function contributes to global carbon and water cycles [@berry_stomata:_2010] and therefore predicting future climate [@franks_stomatal_2017].
+Stomatal function contributes to global carbon and water cycles [@berry_stomata:_2010] and therefore to predicting future climate [@franks_stomatal_2017].
 
@@
 If the acclimation hypothesis is correct, we predict that $\aahigh{} > \aalow{}$ for all populations regardless of native habitat or growth environment.
+$\aahigh{}$ is the \aax{} measured under high light intensity (\ppfdequals{2000}); $\aalow{}$ is the \aax{} measured under low light intensity (\ppfdequals{150}).
 
@@
 Under the plasticity hypothesis, we predict that $\aasun{} > \aashade{}$ for all populations and light intensities.
+$\aasun{}$ is the \aax{} measured on sun leaves; $\aashade{}$ is the \aax{} measured on shade leaves.
 
@@
 #| fig-align: center
-#| fig-cap: "\\textbf{Conceptual outline of three nonmutually exclusive hypotheses explaining why amphistomy advantage (\\aax{}) might be greater for leaves in sunny, open habitats.} The acclimatopn hypothesis predicts that \\aax{} is greater under high measurement light intensity (\\ppfdqty{2000}) than low measurement light intensity (\\ppfdequals{150}), regardless of growth light intensity or native plant area index (PAI).
+#| fig-cap: "\\textbf{Conceptual outline of three nonmutually exclusive hypotheses explaining why amphistomy advantage (\\aax{}) might be greater for leaves in sunny, open habitats.} The acclimation hypothesis predicts that \\aax{} is greater under high measurement light intensity (\\ppfdequals{2000}) than low measurement light intensity (\\ppfdequals{150}), regardless of growth light intensity or native plant area index (PAI).
 The plasticity hypothesis predicts that \\aax{} is greater for plants grown in sun than shade, regardless of measurement light intensity or native PAI.
@@
 With this method, we directly compare the photosynthetic rate of an untreated amphistomatous leaf to that of the same leaf with gas exchange blocked through the adaxial (upper) surface by transparent plastic, which we refer to as 'pseudohypostomy'.
-To compare amphi- and pseudohypostomatous leaves at identical whole-leaf stomatal conductance (\gsw), we measure $A$ over a range of \gsw, inducing stomatal opening and closure by modulating humidity (see Materials and Methods for further details).
-We estimated 'amphistomy advantage' (\aax) *sensu* @parkhurst_adaptive_1978, but with modifications previously described in @triplett_amphistomy_2024 and here (Materials and Methods).
+To compare amphi- and pseudohypostomatous leaves at identical whole-leaf stomatal conductance (\gsw), we measure $A$ over a range of \gsw, inducing stomatal opening and closure by modulating humidity (see supplementary text for further details).
+We estimated 'amphistomy advantage' (\aax) *sensu* @parkhurst_adaptive_1978, but with modifications previously described in @triplett_amphistomy_2024 and here (Supplementary Materials).
 The native light intensity was represented by plant area index (PAI $\unit{\meter\squared\per\meter\squared}$), estimated using a global gridded data set derived from the Global Ecosystem Dynamics Investigation [GEDI; @burns_multi-resolution_2024] and georeferenced accession collection information from the Tomato Genetics Resource Center (@tbl-populations).
@@
 While LMA is weakly, albeit significantly (@tbl-fit_aa2), associated with individual-level \aax{} ([Fig.
-@fig-lma_aa]), the effect of growth light intensity on \aax{} is still predictive based on model comparison (@tbl-aa_loo1).
+@fig-lma_aa]), the effect of growth light intensity on \aax{} is still predictive based on model comparison using information criteria (@tbl-aa_loo1).
 However, the direct effect of the sun treatment on \aax{} was weaker when LMA was included in the model (@tbl-fit_aa1 vs.
@@
 #| fig-height: 4
-#| fig-cap: "\\textbf{Plastic changes in leaf mass per area (LMA) mediate the effect of sun- to shade-grown plants on amphistomy advantage (\\aax{}).} Each point is the estimated population-level LMA ($x$-axis) and \\aax{} ($y$-axis) for each population grown under high light intensity (\\ppfdequals{2000}) and low light intensity (\\ppfdequals{150}).
+#| fig-cap: "\\textbf{Plastic changes in leaf mass per area (LMA) partially mediate the effect of sun- to shade-grown plants on amphistomy advantage (\\aax{}).} Each point is the estimated LMA ($x$-axis) and \\aax{} ($y$-axis) for each population grown under high light intensity (\\ppfdequals{2000}) and low light intensity (\\ppfdequals{150}).
 The points are colored by growth light treatment.

### 2025-08-19 ms/ms.tex
migrating main text to word

@@
 
-The presence of stomata on both leaf surfaces (amphistomy) can increase
-photosynthesis in C\(_3\) plants by reducing the path length for
-CO\(_2\) diffusion between substomatal cavities and chloroplasts.
-Paradoxically, despite this photosynthetic advantage, most leaves are
-hypostomatous (stomata on lower surface only). Based on the observation
-that amphistomatous leaves are most common among herbaceous plants
-growing in sunny habitats, including many crop species, it has long been
-hypothesized that traits affecting CO\(_2\) supply or demand amplify the
-advantage of amphistomy under high light intensity. We used a recently
-developed method to quantify amphistomy advantage, the photosynthetic
-rate of an amphistomatous leaf relative to an otherwise identical
-hypostomatous leaf, in 29 diverse populations representing 14 species of
-wild tomatoes (\emph{Solanum} sect. \emph{Lycopersicon} and sect.
-\emph{Lycopersicoides}). Plants grown under high light intensity benefit
-more from amphistomy in terms of CO\(_2\) assimilated for a given
-stomatal conductance than plants grown under low light, regardless of
-measurement light intensity. Furthermore, populations native to open
-habitats tend to benefit more from amphistomy compared to those from
-more closed habitats. Developmental plasticity and adaptation of leaves
-to high light intensity likely play a combined role in explaining the
-adaptive significance and distribution of amphistomy in nature. We
-estimate that hypostomatous leaves would expend 10-65\% more water to
-achieve the same photosynthetic rate as a comparable amphistomatous
-leaf, implying large fitness costs of amphistomy in ecological contexts
-where this trait is rare.
+The presence of stomata on both leaf surfaces (amphistomy) increases
+photosynthesis by reducing the distance for CO\(_2\) diffusion between
+stomata and chloroplasts. Paradoxically, most leaves are hypostomatous
+(stomata on lower surface), despite the photosynthetic advantage of
+amphistomy. Across 29 diverse populations of ``wild tomatoes'', leaves
+developed under high light intensity benefit more from amphistomy in
+terms of CO\(_2\) assimilated for a given stomatal conductance than
+plants developed under low light. Furthermore, populations native to
+open habitats benefit more from amphistomy than those from more closed
+habitats. Thus, plasticity and adaptation together may explain why
+amphistomatous leaves are prevalent in sunny, open habitats, including
+many crops. Contrary to common assumptions, amphistomy can save water
+because hypostomatous leaves evaporate 10-65\% more to achieve the same
+photosynthetic rate.
 
@@
 improving crops (\emph{16}). Stomatal function contributes to global
-carbon and water cycles (\emph{17}) and therefore predicting future
+carbon and water cycles (\emph{17}) and therefore to predicting future
 climate (\emph{18}).
@@
 \(\aahigh{} > \aalow{}\) for all populations regardless of native
-habitat or growth environment.
+habitat or growth environment. \(\aahigh{}\) is the \aax{} measured
+under high light intensity (\ppfdequals{2000}); \(\aalow{}\) is the
+\aax{} measured under low light intensity (\ppfdequals{150}).
 
@@
 plasticity hypothesis, we predict that \(\aasun{} > \aashade{}\) for all
-populations and light intensities.
+populations and light intensities. \(\aasun{}\) is the \aax{} measured
+on sun leaves; \(\aashade{}\) is the \aax{} measured on shade leaves.
 
@@
 \caption{\label{fig-hypotheses}\textbf{Conceptual outline of three nonmutually exclusive hypotheses explaining why amphistomy advantage (\aax{}) might be greater for leaves in sunny, open habitats.}
-The acclimatopn hypothesis predicts that \aax{} is greater under high
-measurement light intensity (\ppfdqty{2000}) than low measurement light
-intensity (\ppfdequals{150}), regardless of growth light intensity or
-native plant area index (PAI). The plasticity hypothesis predicts that
-\aax{} is greater for plants grown in sun than shade, regardless of
+The acclimation hypothesis predicts that \aax{} is greater under high
+measurement light intensity (\ppfdequals{2000}) than low measurement
+light intensity (\ppfdequals{150}), regardless of growth light intensity
+or native plant area index (PAI). The plasticity hypothesis predicts
+that \aax{} is greater for plants grown in sun than shade, regardless of
 measurement light intensity or native PAI. The constitutive hypothesis
@@
 a range of \gsw, inducing stomatal opening and closure by modulating
-humidity (see Materials and Methods for further details). We estimated
+humidity (see supplementary text for further details). We estimated
 `amphistomy advantage' (\aax) \emph{sensu} (\emph{20}), but with
-modifications previously described in (\emph{46}) and here (Materials
-and Methods). The native light intensity was represented by plant area
-index (PAI \(\unit{\meter\squared\per\meter\squared}\)), estimated using
-a global gridded data set derived from the Global Ecosystem Dynamics
-Investigation {[}GEDI; (\emph{51}){]} and georeferenced accession
-collection information from the Tomato Genetics Resource Center
-(Table~\ref{tbl-populations}). The growth light intensities were
-\ppfdequals{761} (sun treatment) and \ppfdqty{115} (shade treatment)
-while all other environment conditions were nearly identical (see
-Materials and Methods). The high and low measurement light intensities
-were \ppfdequals{2000} (97.8:2.24 red:blue) and \ppfdequals{150}
-(87.0:13.0 red:blue), respectively.
+modifications previously described in (\emph{46}) and here
+(Supplementary Materials). The native light intensity was represented by
+plant area index (PAI \(\unit{\meter\squared\per\meter\squared}\)),
+estimated using a global gridded data set derived from the Global
+Ecosystem Dynamics Investigation {[}GEDI; (\emph{51}){]} and
+georeferenced accession collection information from the Tomato Genetics
+Resource Center (Table~\ref{tbl-populations}). The growth light
+intensities were \ppfdequals{761} (sun treatment) and \ppfdqty{115}
+(shade treatment) while all other environment conditions were nearly
+identical (see Materials and Methods). The high and low measurement
+light intensities were \ppfdequals{2000} (97.8:2.24 red:blue) and
+\ppfdequals{150} (87.0:13.0 red:blue), respectively.
 
@@
 light intensity on \aax{} is still predictive based on model comparison
-(Table~\ref{tbl-aa_loo1}). However, the direct effect of the sun
-treatment on \aax{} was weaker when LMA was included in the model
-(Table~\ref{tbl-fit_aa1} vs. Table~\ref{tbl-fit_aa2}), suggestive of a
-mediating role. Many anatomical traits underlie LMA (\emph{52}) and
-future research will be needed to identify which particular traits, such
-as leaf thickness or mesophyll porosity, are responsible for mediating
-\aax{}. The fact that the \aax{} of sun plants was greater under low
-measurement light intensity supports a long-standing hypothesis that
-resistance to CO\(_2\) diffusion is greater in the upper than lower
-portions of the leaf interior (\emph{53}). At low light intensity,
-photosynthesis is weighted toward the upper palisade where most light is
-intercepted. If resistance to CO\(_2\) diffusion is high in the upper
-palisade, amphistomy may, unexpectedly, be particularly beneficial for
-sun leaves experiencing intermittent shade or cloud cover. Our study is
-limited in testing this because we could not directly measure the
-stomatal conductance ratio and intercellular resistance on each surface.
-Future experiments measuring \aax{} with a dual sided chamber
-(\emph{31}, \emph{33}) can overcome these limitations.
+using information criteria (Table~\ref{tbl-aa_loo1}). However, the
+direct effect of the sun treatment on \aax{} was weaker when LMA was
+included in the model (Table~\ref{tbl-fit_aa1} vs.
+Table~\ref{tbl-fit_aa2}), suggestive of a mediating role. Many
+anatomical traits underlie LMA (\emph{52}) and future research will be
+needed to identify which particular traits, such as leaf thickness or
+mesophyll porosity, are responsible for mediating \aax{}. The fact that
+the \aax{} of sun plants was greater under low measurement light
+intensity supports a long-standing hypothesis that resistance to
+CO\(_2\) diffusion is greater in the upper than lower portions of the
+leaf interior (\emph{53}). At low light intensity, photosynthesis is
+weighted toward the upper palisade where most light is intercepted. If
+resistance to CO\(_2\) diffusion is high in the upper palisade,
+amphistomy may, unexpectedly, be particularly beneficial for sun leaves
+experiencing intermittent shade or cloud cover. Our study is limited in
+testing this because we could not directly measure the stomatal
+conductance ratio and intercellular resistance on each surface. Future
+experiments measuring \aax{} with a dual sided chamber (\emph{31},
+\emph{33}) can overcome these limitations.
 
@@
 
-\caption{\label{fig-lma_aa}\textbf{Plastic changes in leaf mass per area (LMA) mediate the effect of sun- to shade-grown plants on amphistomy advantage (\aax{}).}
-Each point is the estimated population-level LMA (\(x\)-axis) and \aax{}
-(\(y\)-axis) for each population grown under high light intensity
-(\ppfdequals{2000}) and low light intensity (\ppfdequals{150}). The
-points are colored by growth light treatment. The solid line is the
-posterior median of the linear regression of individual-level LMA on
-\(\aahigh\) with the shaded region showing the 95\% confidence ribbon;
-the dashed line is the same for \(\aalow\). Confidence intervals for
-each population estimate were omitted for visual clarity.}
+\caption{\label{fig-lma_aa}\textbf{Plastic changes in leaf mass per area (LMA) partially mediate the effect of sun- to shade-grown plants on amphistomy advantage (\aax{}).}
+Each point is the estimated LMA (\(x\)-axis) and \aax{} (\(y\)-axis) for
+each population grown under high light intensity (\ppfdequals{2000}) and
+low light intensity (\ppfdequals{150}). The points are colored by growth
+light treatment. The solid line is the posterior median of the linear
+regression of individual-level LMA on \(\aahigh\) with the shaded region
+showing the 95\% confidence ribbon; the dashed line is the same for
+\(\aalow\). Confidence intervals for each population estimate were
+omitted for visual clarity.}
 

### 2025-08-19 r/13_plot-lma.R

@@ -89,7 +89,7 @@ fig_lma_aa = ggplot(df1, aes(x = lma, y = aa, shape = Measurement)) +
   ) +
   geom_line(data = df2,
             aes(linetype = Measurement),
-            linewidth = 1.1) +
+            linewidth = 1.1, show.legend = FALSE) +
   # geom_interval(aes(ymin = aa_lower, ymax = aa_upper), linewidth = 0.5) +
   # geom_interval(aes(xmin = lma_lower, xmax = lma_upper), linewidth = 0.5) +
   geom_point(

### 2025-08-19 ms/ms.qmd
biorxiv submission

@@
       "$\\operatorname{cor}(\\mathrm{PAI}, \\mathrm{AA}) = 0$",
-      "$\\operatorname{cor}(\\mathrm{PAI}, \\mathrm{AA}) > 0$"
+      "$\\operatorname{cor}(\\mathrm{PAI}, \\mathrm{AA}) < 0$"
     )
@@
 Justin Alter, Max Gatlin, Joana Kim, Jenna Matsuyama, Brandon Najarian, and Kai Yasuda contributed to data collection.
+Kate McCulloh provided feedback on an earlier version of this manuscript.
 Sarah Friedrich helped with graphics.
+We used Copilot and ChatGPT to assist with coding.
 
@@
 
-US National Science Foundation OIA-1929167 to C.D.M.
+US National Science Foundation OIA-1929167 (C.D.M.)
 
@@
 \setcounter{equation}{0}
+\setcounter{page}{1}
 ```

### 2025-08-19 ms/ms.tex
biorxiv submission

@@
 
-\pandocbounded{\includegraphics[keepaspectratio]{../figures/hypotheses.pdf}}
+\includegraphics{../figures/hypotheses.pdf}
 
@@
 plastic & $\mathrm{AA}_{2000} = \mathrm{AA}_{150}$ & $\mathrm{AA}_{\text{sun}} > \mathrm{AA}_{\text{shade}}$ & $\operatorname{cor}(\mathrm{PAI}, \mathrm{AA}) = 0$\\
-constitutive & $\mathrm{AA}_{2000} = \mathrm{AA}_{150}$ & $\mathrm{AA}_{\text{sun}} = \mathrm{AA}_{\text{shade}}$ & $\operatorname{cor}(\mathrm{PAI}, \mathrm{AA}) > 0$\\
+constitutive & $\mathrm{AA}_{2000} = \mathrm{AA}_{150}$ & $\mathrm{AA}_{\text{sun}} = \mathrm{AA}_{\text{shade}}$ & $\operatorname{cor}(\mathrm{PAI}, \mathrm{AA}) < 0$\\
 \bottomrule
@@
 Alter, Max Gatlin, Joana Kim, Jenna Matsuyama, Brandon Najarian, and Kai
-Yasuda contributed to data collection. Sarah Friedrich helped with
-graphics.
+Yasuda contributed to data collection. Kate McCulloh provided feedback
+on an earlier version of this manuscript. Sarah Friedrich helped with
+graphics. We used Copilot and ChatGPT to assist with coding.
 
@@
 \setcounter{equation}{0}
+\setcounter{page}{1}
 

### 2025-08-19 r/03_trim-outliers.R

@@ -80,11 +80,3 @@ do.call(
 rh_curves |>
   filter(!(acc_id %in% acc_id_outlier)) |>
   write_rds("data/trimmed_rh_curves.rds")
-
-rh_curves1 |>
-  filter(!(acc_id %in% acc_id_outlier)) |>
-  summarize(
-    amphi_first = first(amphi_first),
-    .by = c("acc_id", "light_treatment", "light_intensity")
-  ) |>
-  write_rds("data/trimmed_amphi_first.rds")

### 2025-08-19 r/07_prepare-stan-data.R

@@ -5,19 +5,12 @@ plant_info = read_rds("data/plant-info.rds") |>
   mutate(llma = log(lma_gm2)) |>
   dplyr::select(accession, acc_id, light_treatment, llma) 
 
-trimmed_amphi_first = read_rds("data/trimmed_amphi_first.rds") |>
-  summarize(
-    amphi_first = unique(amphi_first),
-    .by = c("acc_id")
-  )
-
 phy = read_rds("data/phylogeny.rds")
 
 aa_summary = read_rds("objects/aa_summary.rds") |>
   dplyr::select(acc_id, light_intensity, aa = median, se_aa = sd) |>
   mutate(scaled_aa = (aa - mean(aa)) / sd(aa)) |>
-  left_join(plant_info, by = join_by(acc_id)) |>
-  left_join(trimmed_amphi_first, by = join_by(acc_id)) 
+  left_join(plant_info, by = join_by(acc_id))
 
 accession_gedi = read_rds("data/accession-gedi.rds") |>
   dplyr::select(accession, pai) |>

### 2025-08-19 r/14_plot-stomata.R

@@ -4,8 +4,6 @@ fit_aa = read_rds("objects/fit_aa1.rds")
 fit_stomata = read_rds("objects/fit_stomata.rds")
 accession_gedi = read_rds("data/accession-gedi.rds") |>
   rename(acc = accession)
-accession_climate = read_rds("data/accession-climate.rds") |>
-  rename(acc = accession)
 d1 = fit_aa$data
 
 plant_info = read_rds("data/plant-info.rds") |>

### 2025-08-19 r/15_plot-curves.R

@@ -26,7 +26,7 @@ assert_true(all(!is.na(fit_curve_diagnostics$acc_id))) # Check for NA values in 
 assert_true(all(!is.na(fit_curve_diagnostics$`Curve type`))) # Check for NA values in `Curve type`
 assert_true(all(!is.na(fit_curve_diagnostics$light_intensity))) # Check for NA values in light_intensity
 
-plan(multisession, workers = 1)
+plan(multisession, workers = 19)
 ags_curves = future_map(
   unique(aa_summary$acc_id),
   \(.x) {

### 2025-09-15 ms/ms.qmd
reformat ms

@@
   \newcommand{\tleafequals}[1]{$T_\mathrm{leaf} = \qty{#1}{\degreeCelsius}$}
+  \usepackage{setspace}
+  \usepackage[modulo]{lineno}
 mainfont: Times New Roman
 bibliography: solanum-aa.bib
-csl: science.csl
+csl: https://www.zotero.org/styles/pnas
 ---
@@
 ORCID: Christopher D. Muir — \href{https://orcid.org/0000-0003-2555-3878}{0000-0003-2555-3878}\\
+\medskip
+Classification: Biological Sciences, Ecology\\
+\medskip
+Keywords: amphistomy, photosynthesis, \textit{Solanum}, stomata
 \end{center}
@@
 
+```{=latex}
+\newpage
+\begin{linenumbers}
+\begin{doublespace}
+```
+
 ## Abstract
@@
 -->
+
+## Significance statement
+
+Stomata, the "mouths" of plants, regulate CO$_2$ available for photosynthesis.
+They are ubiquitous on leaves and essential to understanding why vascular plants transformed Earth and how our crops grow.
+Yet the adaptive significance of why many leaves only have stomata on their lower surface is a mystery.
+Many assume that this limits water loss, but we show the opposite is true.
+By manipulating which stomata a leaf can use, we show that leaves with stomata on both surfaces gain more CO$_2$ for the same amount of water lost, increasing water use efficiency.
+Furthermore, species from open canopies grown under high light benefit the most, which may explain why leaves with stomata on both sides are most prevalent in these habitats.
 
@@
 
+```{=latex}
+\end{doublespace}
+\end{linenumbers}
+```
+
 ## References

### 2025-09-15 ms/ms.tex
reformat ms

@@
 \newcommand{\tleafequals}[1]{$T_\mathrm{leaf} = \qty{#1}{\degreeCelsius}$}
+\usepackage{setspace}
+\usepackage[modulo]{lineno}
 \makeatletter
@@
 ORCID: Christopher D. Muir — \href{https://orcid.org/0000-0003-2555-3878}{0000-0003-2555-3878}\\
+\medskip
+Classification: Biological Sciences, Ecology\\
+\medskip
+Keywords: amphistomy, photosynthesis, \textit{Solanum}, stomata
 \end{center}
+
+\newpage
+\begin{linenumbers}
+\begin{doublespace}
 
@@
 
+\subsection{Significance statement}\label{significance-statement}
+
+Stomata, the ``mouths'' of plants, regulate CO\(_2\) available for
+photosynthesis. They are ubiquitous on leaves and essential to
+understanding why vascular plants transformed Earth and how our crops
+grow. Yet the adaptive significance of why many leaves only have stomata
+on their lower surface is a mystery. Many assume that this limits water
+loss, but we show the opposite is true. By manipulating which stomata a
+leaf can use, we show that leaves with stomata on both surfaces gain
+more CO\(_2\) for the same amount of water lost, increasing water use
+efficiency. Furthermore, species from open canopies grown under high
+light benefit the most, which may explain why leaves with stomata on
+both sides are most prevalent in these habitats.
+
 \subsection{Main text}\label{main-text}
@@
 assimilation while preventing hydraulic failure in variable environments
-(\emph{1}--\emph{3}). Optimal stomatal function depends on both dynamic
-changes in aperture on the scale of minutes to hours, as well as static
-anatomy determined by developmental plasticity and constitutive genetic
-differences (\emph{4}--\emph{6}). Understanding how stomata respond to
-environmental change over daily, developmental, and evolutionary time is
-important for studying adaptation (\emph{1}, \emph{7}, \emph{8}),
-inferring paleoclimate from fossil cuticles (\emph{9}--\emph{11}),
-predicting responses to climate change (\emph{12}--\emph{15}), and
-improving crops (\emph{16}). Stomatal function contributes to global
-carbon and water cycles (\emph{17}) and therefore to predicting future
-climate (\emph{18}).
+(1--3). Optimal stomatal function depends on both dynamic changes in
+aperture on the scale of minutes to hours, as well as static anatomy
+determined by developmental plasticity and constitutive genetic
+differences (4--6). Understanding how stomata respond to environmental
+change over daily, developmental, and evolutionary time is important for
+studying adaptation (1, 7, 8), inferring paleoclimate from fossil
+cuticles (9--11), predicting responses to climate change (12--15), and
+improving crops (16). Stomatal function contributes to global carbon and
+water cycles (17) and therefore to predicting future climate (18).
 
@@
 adaptive significance of amphistomatous leaves remains an important
-unsolved problem in leaf structure-function relationships
-(\emph{19}--\emph{26}). Amphistomatous leaves develop abaxial and
-adaxial stomata whose aperture can be independently regulated
-(\emph{27}--\emph{31}) to control gas exchange through each surface. All
-else being equal, simultaneous gas exchange through stomata on both
-surfaces increases CO\(_2\) supply to chloroplasts by providing a second
-parallel pathway through leaf intercellular airspaces, enhancing
-photosynthesis (\emph{20}, \emph{32}). The extent to which amphistomy
-increases CO\(_2\) supply depends on resistance to diffusion in
-intercellular airspaces. This resistance can be low in thin, porous,
-amphistomatous leaves (\emph{28}, \emph{33}), but may be more
-substantial in thick, dense, hypostomatous leaves (\emph{34}). We refer
-to the intercellular airspace conductance (\gias), the inverse of
-resistance. Amphistomatous leaves also lose more water through
-evaporation because of a second boundary layer conductance (\emph{35}),
-but the additional carbon gain should be enough to offset this cost in
-most realistic scenarios (\emph{36}).
+unsolved problem in leaf structure-function relationships (19--26).
+Amphistomatous leaves develop abaxial and adaxial stomata whose aperture
+can be independently regulated (27--31) to control gas exchange through
+each surface. All else being equal, simultaneous gas exchange through
+stomata on both surfaces increases CO\(_2\) supply to chloroplasts by
+providing a second parallel pathway through leaf intercellular
+airspaces, enhancing photosynthesis (20, 32). The extent to which
+amphistomy increases CO\(_2\) supply depends on resistance to diffusion
+in intercellular airspaces. This resistance can be low in thin, porous,
+amphistomatous leaves (28, 33), but may be more substantial in thick,
+dense, hypostomatous leaves (34). We refer to the intercellular airspace
+conductance (\gias), the inverse of resistance. Amphistomatous leaves
+also lose more water through evaporation because of a second boundary
+layer conductance (35), but the additional carbon gain should be enough
+to offset this cost in most realistic scenarios (36).
 
@@
 leaves are not amphistomatous. Many vertically oriented and/or
-isobilateral leaves are amphistomatous (\emph{25}). But among
-dorsiventral leaves, it is primarily herbaceous plants in open, high
-light habitats that tend to have amphistomatous leaves (\emph{22},
-\emph{39}--\emph{44}). Most other leaves, except those from aquatic
-habitats, are hypostomatous, producing stomata only on the lower,
-abaxial surface. Even resupinate leaves develop stomata on the lower,
-albeit adaxial surface (\emph{45}), suggesting that leaf orientation
-(lower vs.~upper) rather than leaf polarity (abaxial vs.~adaxial) is
-causal. Stomatal density ratio, defined as the ratio of upper to total
-stomatal density, is a quantitative metric of stomatal distribution. The
-covariation between stomatal density ratio and light habitat is both
-qualitative and quantitative. A higher proportion of sun leaves are
-amphistomatous (\emph{39}) and the proportion of stomata on the upper,
-adaxial surface increases with light (\emph{42}, \emph{43}). Resolving
-why high light intensity favors amphistomatous dorsivental leaves is an
-important first step toward understanding variation in stomatal density
-ratio and leaf structure-function relationships more generally.
+isobilateral leaves are amphistomatous (25). But among dorsiventral
+leaves, it is primarily herbaceous plants in open, high light habitats
+that tend to have amphistomatous leaves (22, 39--44). Most other leaves,
+except those from aquatic habitats, are hypostomatous, producing stomata
+only on the lower, abaxial surface. Even resupinate leaves develop
+stomata on the lower, albeit adaxial surface (45), suggesting that leaf
+orientation (lower vs.~upper) rather than leaf polarity (abaxial
+vs.~adaxial) is causal. Stomatal density ratio, defined as the ratio of
+upper to total stomatal density, is a quantitative metric of stomatal
+distribution. The covariation between stomatal density ratio and light
+habitat is both qualitative and quantitative. A higher proportion of sun
+leaves are amphistomatous (39) and the proportion of stomata on the
+upper, adaxial surface increases with light (42, 43). Resolving why high
+light intensity favors amphistomatous dorsivental leaves is an important
+first step toward understanding variation in stomatal density ratio and
+leaf structure-function relationships more generally.
 
@@
 photosynthesis in an amphistomatous leaf compared to an otherwise
-identical pseudohypostomatous leaf (\emph{20}, \emph{46}). Why would
-\aax{} be greater in sun than shade? We consider three nonmutually
-exclusive hypotheses that we classify as `acclimation', `plasticity',
-and `constitutive' (Fig.~\ref{fig-hypotheses}).
+identical pseudohypostomatous leaf (20, 46). Why would \aax{} be greater
+in sun than shade? We consider three nonmutually exclusive hypotheses
+that we classify as `acclimation', `plasticity', and `constitutive'
+(Fig.~\ref{fig-hypotheses}).
 
@@
 \emph{Plasticity hypothesis}: Individuals of the same genotype often
-develop dramatically different leaves in sun and shade conditions
-(\emph{47}). Plastic responses are likely adaptations to optimize
-photosynthesis at different light intensities in variable environments
-(\emph{48}). Plastic changes in leaf anatomy and biochemistry could
-modulate \aax{} as a byproduct. Thicker or less porous leaves, both of
-which are associated with high leaf mass per area (LMA), will have lower
-\gias; leaves with increased total stomatal density and photosynthetic
-capacity have greater potential CO\(_2\) supply and demand. Under the
-plasticity hypothesis, we predict that \(\aasun{} > \aashade{}\) for all
+develop dramatically different leaves in sun and shade conditions (47).
+Plastic responses are likely adaptations to optimize photosynthesis at
+different light intensities in variable environments (48). Plastic
+changes in leaf anatomy and biochemistry could modulate \aax{} as a
+byproduct. Thicker or less porous leaves, both of which are associated
+with high leaf mass per area (LMA), will have lower \gias; leaves with
+increased total stomatal density and photosynthetic capacity have
+greater potential CO\(_2\) supply and demand. Under the plasticity
+hypothesis, we predict that \(\aasun{} > \aashade{}\) for all
 populations and light intensities. \(\aasun{}\) is the \aax{} measured
@@
 growth light intensities. For herbaceous plants, light intensity is
-largely a function of the tree canopy (\emph{49}). Herbs growing in the
-open will regularly experience high light intensity; herbs growing under
-a forest canopy will often experience low light intensity.
+largely a function of the tree canopy (49). Herbs growing in the open
+will regularly experience high light intensity; herbs growing under a
+forest canopy will often experience low light intensity.
 
@@
 tomato species (Table~\ref{tbl-populations}; Figure~\ref{fig-sr-rxnorm};
-(\emph{50})) from different native light habitats, grown under simulated
-sun and shade light treatments, and measured under contrasting light
+(50)) from different native light habitats, grown under simulated sun
+and shade light treatments, and measured under contrasting light
 intensities (low and high). We measured \aax{} on 572 individual plants
 from 29 populations (average of 9.86 replicates per light treatment)
-using a recently developed method (\emph{46}). With this method, we
-directly compare the photosynthetic rate of an untreated amphistomatous
-leaf to that of the same leaf with gas exchange blocked through the
-adaxial (upper) surface by transparent plastic, which we refer to as
+using a recently developed method (46). With this method, we directly
+compare the photosynthetic rate of an untreated amphistomatous leaf to
+that of the same leaf with gas exchange blocked through the adaxial
+(upper) surface by transparent plastic, which we refer to as
 `pseudohypostomy'. To compare amphi- and pseudohypostomatous leaves at
@@
 humidity (see supplementary text for further details). We estimated
-`amphistomy advantage' (\aax) \emph{sensu} (\emph{20}), but with
-modifications previously described in (\emph{46}) and here
-(Supplementary Materials). The native light intensity was represented by
-plant area index (PAI \(\unit{\meter\squared\per\meter\squared}\)),
-estimated using a global gridded data set derived from the Global
-Ecosystem Dynamics Investigation {[}GEDI; (\emph{51}){]} and
-georeferenced accession collection information from the Tomato Genetics
-Resource Center (Table~\ref{tbl-populations}). The growth light
-intensities were \ppfdequals{761} (sun treatment) and \ppfdqty{115}
-(shade treatment) while all other environment conditions were nearly
-identical (see Materials and Methods). The high and low measurement
-light intensities were \ppfdequals{2000} (97.8:2.24 red:blue) and
-\ppfdequals{150} (87.0:13.0 red:blue), respectively.
+`amphistomy advantage' (\aax) \emph{sensu} (20), but with modifications
+previously described in (46) and here (Supplementary Materials). The
+native light intensity was represented by plant area index (PAI
+\(\unit{\meter\squared\per\meter\squared}\)), estimated using a global
+gridded data set derived from the Global Ecosystem Dynamics
+Investigation {[}GEDI; (51){]} and georeferenced accession collection
+information from the Tomato Genetics Resource Center
+(Table~\ref{tbl-populations}). The growth light intensities were
+\ppfdequals{761} (sun treatment) and \ppfdqty{115} (shade treatment)
+while all other environment conditions were nearly identical (see
+Materials and Methods). The high and low measurement light intensities
+were \ppfdequals{2000} (97.8:2.24 red:blue) and \ppfdequals{150}
+(87.0:13.0 red:blue), respectively.
 
@@
 average of 123\% {[}95\% CI: 42.9 to 256\%{]}, qualitatively similar to
-plastic responses in many species (\emph{47}). While LMA is weakly,
-albeit significantly (Table~\ref{tbl-fit_aa2}), associated with
+plastic responses in many species (47). While LMA is weakly, albeit
+significantly (Table~\ref{tbl-fit_aa2}), associated with
 individual-level \aax{} (Fig.~\ref{fig-lma_aa}), the effect of growth
@@
 Table~\ref{tbl-fit_aa2}), suggestive of a mediating role. Many
-anatomical traits underlie LMA (\emph{52}) and future research will be
-needed to identify which particular traits, such as leaf thickness or
-mesophyll porosity, are responsible for mediating \aax{}. The fact that
-the \aax{} of sun plants was greater under low measurement light
-intensity supports a long-standing hypothesis that resistance to
-CO\(_2\) diffusion is greater in the upper than lower portions of the
-leaf interior (\emph{53}). At low light intensity, photosynthesis is
-weighted toward the upper palisade where most light is intercepted. If
-resistance to CO\(_2\) diffusion is high in the upper palisade,
-amphistomy may, unexpectedly, be particularly beneficial for sun leaves
-experiencing intermittent shade or cloud cover. Our study is limited in
-testing this because we could not directly measure the stomatal
-conductance ratio and intercellular resistance on each surface. Future
-experiments measuring \aax{} with a dual sided chamber (\emph{31},
-\emph{33}) can overcome these limitations.
+anatomical traits underlie LMA (52) and future research will be needed
+to identify which particular traits, such as leaf thickness or mesophyll
+porosity, are responsible for mediating \aax{}. The fact that the \aax{}
+of sun plants was greater under low measurement light intensity supports
+a long-standing hypothesis that resistance to CO\(_2\) diffusion is
+greater in the upper than lower portions of the leaf interior (53). At
+low light intensity, photosynthesis is weighted toward the upper
+palisade where most light is intercepted. If resistance to CO\(_2\)
+diffusion is high in the upper palisade, amphistomy may, unexpectedly,
+be particularly beneficial for sun leaves experiencing intermittent
+shade or cloud cover. Our study is limited in testing this because we
+could not directly measure the stomatal conductance ratio and
+intercellular resistance on each surface. Future experiments measuring
+\aax{} with a dual sided chamber (31, 33) can overcome these
+limitations.
 
@@
 
+\end{doublespace}
+\end{linenumbers}
+
 \subsection{References}\label{references}
@@
 \CSLLeftMargin{2. }%
-\CSLRightInline{S. A. M. McAdam, J. G. Duckett, F. C. Sussmilch, S.
-Pressel, K. S. Renzaglia, R. Hedrich, T. J. Brodribb, A. Merced,
+\CSLRightInline{S. A. M. McAdam, \emph{et al.},
 \href{https://doi.org/10.1002/ajb2.1619}{Stomata: The holey grail of
@@
 \CSLLeftMargin{3. }%
-\CSLRightInline{J. W. Clark, B. J. Harris, A. J. Hetherington, N.
-Hurtado-Castano, R. A. Brench, S. Casson, T. A. Williams, J. E. Gray, A.
-M. Hetherington, \href{https://doi.org/10.1016/j.cub.2022.04.040}{The
-origin and evolution of stomata}. \emph{Current Biology} \textbf{32},
-R539--R553 (2022).}
+\CSLRightInline{J. W. Clark, \emph{et al.},
+\href{https://doi.org/10.1016/j.cub.2022.04.040}{The origin and
+evolution of stomata}. \emph{Current Biology} \textbf{32}, R539--R553
+(2022).}
 
@@
 \CSLLeftMargin{5. }%
-\CSLRightInline{H. J. de Boer, C. A. Price, F. Wagner‐Cremer, S. C.
-Dekker, P. J. Franks, E. J. Veneklaas,
+\CSLRightInline{H. J. de Boer, \emph{et al.},
 \href{https://doi.org/10.1111/nph.13929}{Optimal allocation of leaf
@@
 \CSLLeftMargin{9. }%
-\CSLRightInline{P. J. Franks, D. L. Royer, D. J. Beerling, P. K. Van de
-Water, D. J. Cantrill, M. M. Barbour, J. A. Berry,
+\CSLRightInline{P. J. Franks, \emph{et al.},
 \href{https://doi.org/10.1002/2014GL060457}{New constraints on
@@
 \CSLRightInline{The Cenozoic CO Proxy Integration Project (CenCOPIP)
-Consortium*†, B. Hönisch, D. L. Royer, D. O. Breecker, P. J. Polissar,
-G. J. Bowen, M. J. Henehan, Y. Cui, M. Steinthorsdottir, J. C. McElwain,
-M. J. Kohn, A. Pearson, S. R. Phelps, K. T. Uno, A. Ridgwell, E.
-Anagnostou, J. Austermann, M. P. S. Badger, R. S. Barclay, P. K. Bijl,
-T. B. Chalk, C. R. Scotese, E. De La Vega, R. M. DeConto, K. A. Dyez, V.
-Ferrini, P. J. Franks, C. F. Giulivi, M. Gutjahr, D. T. Harper, L. L.
-Haynes, M. Huber, K. E. Snell, B. A. Keisling, W. Konrad, T. K.
-Lowenstein, A. Malinverno, M. Guillermic, L. M. Mejía, J. N. Milligan,
-J. J. Morton, L. Nordt, R. Whiteford, A. Roth-Nebelsick, J. K. C.
-Rugenstein, M. F. Schaller, N. D. Sheldon, S. Sosdian, E. B. Wilkes, C.
-R. Witkowski, Y. G. Zhang, L. Anderson, D. J. Beerling, C. Bolton, T. E.
-Cerling, J. M. Cotton, J. Da, D. D. Ekart, G. L. Foster, D. R.
-Greenwood, E. G. Hyland, E. A. Jagniecki, J. P. Jasper, J. B. Kowalczyk,
-L. Kunzmann, W. M. Kürschner, C. E. Lawrence, C. H. Lear, M. A.
-Martínez-Botí, D. P. Maxbauer, P. Montagna, B. D. A. Naafs, J. W. B.
-Rae, M. Raitzsch, G. J. Retallack, S. J. Ring, O. Seki, J. Sepúlveda, A.
-Sinha, T. F. Tesfamichael, A. Tripati, J. Van Der Burgh, J. Yu, J. C.
-Zachos, L. Zhang, \href{https://doi.org/10.1126/science.adi5177}{Toward
-a {Cenozoic} history of atmospheric {CO}\(_{\textrm{2}}\)}.
-\emph{Science} \textbf{382}, eadi5177 (2023).}
+Consortium*†, \emph{et al.},
+\href{https://doi.org/10.1126/science.adi5177}{Toward a {Cenozoic}
+history of atmospheric {CO}\(_{\textrm{2}}\)}. \emph{Science}
+\textbf{382}, eadi5177 (2023).}
 
@@
 \CSLLeftMargin{13. }%
-\CSLRightInline{X. Liang, D. Wang, Q. Ye, J. Zhang, M. Liu, H. Liu, K.
-Yu, Y. Wang, E. Hou, B. Zhong, L. Xu, T. Lv, S. Peng, H. Lu, P. Sicard,
-A. Anav, D. S. Ellsworth,
+\CSLRightInline{X. Liang, \emph{et al.},
 \href{https://doi.org/10.1038/s41467-023-37934-7}{Stomatal responses of
@@
 \CSLLeftMargin{15. }%
-\CSLRightInline{P. L. M. Lang, J. M. Erberich, L. Lopez, C. L. Weiß, G.
-Amador, H. F. Fung, S. M. Latorre, J. R. Lasky, H. A. Burbano, M.
-Expósito-Alonso, D. C. Bergmann,
+\CSLRightInline{P. L. M. Lang, \emph{et al.},
 \href{https://doi.org/10.1038/s41559-024-02481-x}{Century-long timelines
@@
 \CSLLeftMargin{16. }%
-\CSLRightInline{T. A. Hofmann, W. Atkinson, M. Fan, A. J. Simkin, P.
-Jindal, T. Lawson, \href{https://doi.org/10.1098/rstb.2024.0244}{Impact
-of climate-driven changes in temperature on stomatal anatomy and
-physiology}. \emph{Philosophical Transactions of the Royal Society B:
-Biological Sciences} \textbf{380}, 20240244 (2025).}
+\CSLRightInline{T. A. Hofmann, \emph{et al.},
+\href{https://doi.org/10.1098/rstb.2024.0244}{Impact of climate-driven
+changes in temperature on stomatal anatomy and physiology}.
+\emph{Philosophical Transactions of the Royal Society B: Biological
+Sciences} \textbf{380}, 20240244 (2025).}
 
@@
 \emph{The Encyclopedia of Ignorance}, R. Duncan, M. Weston-Smith, Eds.
-(Pergamon, Oxford, 1977)vol. 2, pp. 317--330.}
+(Pergamon, 1977), pp. 317--330.}
 
@@
 \CSLLeftMargin{22. }%
-\CSLRightInline{A. C. Gibson, \emph{Structure-{Function} {Relations} of
-{Warm} {Desert} {Plants}} (Springer Berlin / Heidelberg, Berlin,
-Heidelberg, 1996;
-\url{http://public.eblib.com/choice/PublicFullRecord.aspx?p=6495247}).}
+\CSLRightInline{A. C. Gibson,
+\emph{\href{http://public.eblib.com/choice/PublicFullRecord.aspx?p=6495247}{Structure-{Function}
+{Relations} of {Warm} {Desert} {Plants}}} (Springer Berlin / Heidelberg,
+1996).}
 
@@
 \CSLLeftMargin{24. }%
-\CSLRightInline{R. Oguchi, Y. Onoda, I. Terashima, D. Tholen, {``Leaf
-{Anatomy} and {Function}''} in \emph{The {Leaf}: {A} {Platform} for
-{Performing} {Photosynthesis}}, W. W. Adams III, I. Terashima, Eds.
-(Springer International Publishing, Cham, 2018;
-\url{https://doi.org/10.1007/978-3-319-93594-2_5})\emph{Advances in
-{Photosynthesis} and {Respiration}}, pp. 97--139.}
+\CSLRightInline{R. Oguchi, Y. Onoda, I. Terashima, D. Tholen,
+{``\href{https://doi.org/10.1007/978-3-319-93594-2_5}{Leaf {Anatomy} and
+{Function}}''} in \emph{The {Leaf}: {A} {Platform} for {Performing}
+{Photosynthesis}}, Advances in {Photosynthesis} and {Respiration}., W.
+W. Adams III, I. Terashima, Eds. (Springer International Publishing,
+2018), pp. 97--139.}
 
@@
 \emph{Unsolved {Problems} in {Ecology}}, A. Dobson, D. Tilman, R. D.
-Holt, Eds. (Princeton University Press, Princeton, 2020), pp. 124--144.}
+Holt, Eds. (Princeton University Press, 2020), pp. 124--144.}
 
@@
 \CSLLeftMargin{31. }%
-\CSLRightInline{S. Wall, S. Vialet‐Chabrand, P. Davey, J. Van Rie, A.
-Galle, J. Cockram, T. Lawson,
+\CSLRightInline{S. Wall, \emph{et al.},
 \href{https://doi.org/10.1111/nph.18257}{Stomata on the abaxial and
@@
 \CSLLeftMargin{42. }%
-\CSLRightInline{S. F. Bucher, K. Auerswald, C. Grün-Wenzel, S. I.
-Higgins, J. Garcia Jorge, C. Römermann,
+\CSLRightInline{S. F. Bucher, \emph{et al.},
 \href{https://doi.org/10.1016/j.flora.2017.02.011}{Stomatal traits
@@
 \CSLLeftMargin{44. }%
-\CSLRightInline{G. Triplett, A. S. David,
-\href{https://doi.org/10.1002/ajb2.70050}{Stomatal distribution and
+\CSLRightInline{G. Triplett, A. S. David, Stomatal distribution and
 post‐fire recovery: {Intra}‐ and interspecific variation in plants of
-the pyrogenic {Florida} scrub}. \emph{American Journal of Botany},
-e70050 (2025).}
+the pyrogenic {Florida} scrub. \emph{American Journal of Botany} e70050
+(2025). \url{https://doi.org/10.1002/ajb2.70050}.}
 
@@
 \CSLLeftMargin{47. }%
-\CSLRightInline{H. Poorter, Ü. Niinemets, N. Ntagkas, A. Siebenkäs, M.
-Mäenpää, S. Matsubara, T. L. Pons,
+\CSLRightInline{H. Poorter, \emph{et al.},
 \href{https://doi.org/10.1111/nph.15754}{A meta‐analysis of plant
@@
 \CSLLeftMargin{52. }%
-\CSLRightInline{G. P. John, C. Scoffoni, T. N. Buckley, R. Villar, H.
-Poorter, L. Sack, \href{https://doi.org/10.1111/ele.12739}{The
-anatomical and compositional basis of leaf mass per area}. \emph{Ecology
-Letters} \textbf{20}, 412--425 (2017).}
+\CSLRightInline{G. P. John, \emph{et al.},
+\href{https://doi.org/10.1111/ele.12739}{The anatomical and
+compositional basis of leaf mass per area}. \emph{Ecology Letters}
+\textbf{20}, 412--425 (2017).}
 
@@
 \CSLLeftMargin{58. }%
-\CSLRightInline{J. Schindelin, I. Arganda-Carreras, E. Frise, V. Kaynig,
-M. Longair, T. Pietzsch, S. Preibisch, C. Rueden, S. Saalfeld, B.
-Schmid, J.-Y. Tinevez, D. J. White, V. Hartenstein, K. Eliceiri, P.
-Tomancak, A. Cardona, \href{https://doi.org/10.1038/nmeth.2019}{Fiji: An
-open-source platform for biological-image analysis}. \emph{Nature
-Methods} \textbf{9}, 676--682 (2012).}
+\CSLRightInline{J. Schindelin, \emph{et al.},
+\href{https://doi.org/10.1038/nmeth.2019}{Fiji: An open-source platform
+for biological-image analysis}. \emph{Nature Methods} \textbf{9},
+676--682 (2012).}
 
@@
 \CSLRightInline{J. L. Monteith, M. H. Unsworth, \emph{Principles of
-Environmental Physics: Plants, Animals, and the Atmosphere}
-(Elsevier/Academic Press, Amsterdam ; Boston, 4th ed., 2013).}
+environmental physics: Plants, animals, and the atmosphere}, 4th ed
+(Elsevier/Academic Press, 2013).}
 
@@
 \CSLLeftMargin{60. }%
-\CSLRightInline{Stan Development Team, \emph{Stan {Modeling} {Language}
-{Users} {Guide} and {Reference} {Manual}} (2025;
-\url{https://mc-stan.org}).}
+\CSLRightInline{Stan Development Team,
+\emph{\href{https://mc-stan.org}{Stan {Modeling} {Language} {Users}
+{Guide} and {Reference} {Manual}}} (2025).}
 
@@
 \CSLRightInline{J. Gabry, R. Češnovar, A. Johnson, S. Bronder,
-\emph{Cmdstanr: {R} {Interface} to '{CmdStan}'} (2025;
-\href{https://mc-stan.org/cmdstanr,\%20https://discourse.mc-stan.org}{https://mc-stan.org/cmdstanr,
-https://discourse.mc-stan.org}).}
+\emph{\href{https://mc-stan.org/cmdstanr,\%20https://discourse.mc-stan.org}{Cmdstanr:
+{R} {Interface} to '{CmdStan}'}} (2025).}
 
@@
 \CSLLeftMargin{63. }%
-\CSLRightInline{R Core Team, \emph{R: {A} {Language} and {Environment}
-for {Statistical} {Computing}} (R Foundation for Statistical Computing,
-Vienna, Austria, 2025; \url{http://www.R-project.org/}).}
+\CSLRightInline{R Core Team, \emph{\href{http://www.R-project.org/}{R:
+{A} {Language} and {Environment} for {Statistical} {Computing}}} (R
+Foundation for Statistical Computing, 2025).}
 
@@
 \emph{Solanum} sect. \emph{Lycopersicon} and sect.
-\emph{Lycopersicoides} (\emph{50}) and the cultivated tomato \emph{S.
+\emph{Lycopersicoides} (50) and the cultivated tomato \emph{S.
 lycopersicum} var. \emph{lycopersicum} (Table~\ref{tbl-populations}).
@@
 encompass the breadth of climatic variation in the wild tomato clade
-(\emph{54}). In one case, we substituted a congeric population of the
-species (\emph{S. galapagense}) because of difficult growing the focal
+(54). In one case, we substituted a congeric population of the species
+(\emph{S. galapagense}) because of difficult growing the focal
 population. We also added an a population of \emph{S. pennellii}. Due to
@@
 they had time to sense and respond developmentally to the light
-intensity of the treatment rather than the seedling conditions
-(\emph{55}). Shade plants grew slower than sun plants, hence leaves at
-the same developmental stage were measured on chronologically older
-plants in the shade treatment. In some sun plants, we had to use leaves
-higher on the stem because short internodes made lower leaves
-inaccessible with the gas exchange equipment. We measured terminal
-leaflets in 82.6\% of cases, but used the lateral leaflet closest to the
-terminal leaflet when it was damaged or difficult to clamp into the gas
-exchange chamber. When a leaflet was damaged during gas exchange
-measurements, we collected anatomical data from the nearest leaflet on
-the same leaf (1.58\% of leaves).
+intensity of the treatment rather than the seedling conditions (55).
+Shade plants grew slower than sun plants, hence leaves at the same
+developmental stage were measured on chronologically older plants in the
+shade treatment. In some sun plants, we had to use leaves higher on the
+stem because short internodes made lower leaves inaccessible with the
+gas exchange equipment. We measured terminal leaflets in 82.6\% of
+cases, but used the lateral leaflet closest to the terminal leaflet when
+it was damaged or difficult to clamp into the gas exchange chamber. When
+a leaflet was damaged during gas exchange measurements, we collected
+anatomical data from the nearest leaflet on the same leaf (1.58\% of
+leaves).
 
@@
 
-We estimated `amphistomy advantage' (\aax) \emph{sensu} (\emph{20}), but
-with modifications previously described in (\emph{46}). \aax{} is
-calculated as the log-response ratio of \(A\) compared at the same total
-\gsw:
+We estimated `amphistomy advantage' (\aax) \emph{sensu} (20), but with
+modifications previously described in (46). \aax{} is calculated as the
+log-response ratio of \(A\) compared at the same total \gsw:
 
@@
 we alternated between starting with amphi or pseudohypo leaf
-measurements. Unlike (\emph{46}), preliminary experiments with
-\emph{Solanum} indicated a strong order effect in that \(A\) declined in
-the second curve. Therefore, we made measurements over two days. On the
-first day, we measured high and low light-intensity curves for either
-amphi or pseudohypo leaves; on the second day, we measured high and low
+measurements. Unlike (46), preliminary experiments with \emph{Solanum}
+indicated a strong order effect in that \(A\) declined in the second
+curve. Therefore, we made measurements over two days. On the first day,
+we measured high and low light-intensity curves for either amphi or
+pseudohypo leaves; on the second day, we measured high and low
 light-intensity curves on the other leaf type.
@@
 surfaces from all leaves, using guard cell length as a proxy for
-stomatal size since it is proportional to maximum conductance
-(\emph{56}). We made surface impressions of leaf lamina from the same
-area used for gas exchange measurements using a-silicone impression
-material (Zhermack elite HD+, light body, fast set, Rovigo, Italy). We
-applied clear nail polish to make positive replicas of the impression.
-After nail polish dried, we mounted replicas on a microscope slide using
-transparent tape (\emph{57}). We digitized a portion of each leaf
-surface replica using a brightfield microscope (Leica DM2000, Wetzlar,
-Germany). We counted and measured guard cell length on all stomata using
-the FIJI implementation of ImageJ2 version 2.3.0 (\emph{58}), then
-divided the count by the visible leaf area
-(\(\qty{0.890}{\raiseto{2}\milli\meter}\)) to estimate stomatal density.
-For each surface we calculated the anatomical maximum stomatal
-conductance (\gmax) at a reference leaf temperature of
-\(\qty{25}{\degreeCelsius}\) following (\emph{56}) as:
+stomatal size since it is proportional to maximum conductance (56). We
+made surface impressions of leaf lamina from the same area used for gas
+exchange measurements using a-silicone impression material (Zhermack
+elite HD+, light body, fast set, Rovigo, Italy). We applied clear nail
+polish to make positive replicas of the impression. After nail polish
+dried, we mounted replicas on a microscope slide using transparent tape
+(57). We digitized a portion of each leaf surface replica using a
+brightfield microscope (Leica DM2000, Wetzlar, Germany). We counted and
+measured guard cell length on all stomata using the FIJI implementation
+of ImageJ2 version 2.3.0 (58), then divided the count by the visible
+leaf area (\(\qty{0.890}{\raiseto{2}\milli\meter}\)) to estimate
+stomatal density. For each surface we calculated the anatomical maximum
+stomatal conductance (\gmax) at a reference leaf temperature of
+\(\qty{25}{\degreeCelsius}\) following (56) as:
 
@@
 \(D_\mathrm{wv} = \qty{2.49e-5}{\meter\squared\per\second}\) and
-\(v = \qty{2.24e-2}{\raiseto{3}\meter\per\mol}\) (\emph{59}). For
-kidney-shaped guard cells like those in wild tomatoes,
-\(c = h = j = 0.5\). The \gmaxratio{} is the \gmax{} of the adaxial
-surface divided by the sum total \gmax{} of both surfaces.
+\(v = \qty{2.24e-2}{\raiseto{3}\meter\per\mol}\) (59). For kidney-shaped
+guard cells like those in wild tomatoes, \(c = h = j = 0.5\). The
+\gmaxratio{} is the \gmax{} of the adaxial surface divided by the sum
+total \gmax{} of both surfaces.
 
@@
 California. USA) and measured leaflet area from digital images using the
-FIJI implementation of ImageJ2 version 2.3.0 (\emph{58}). We dried
-leaves for 72 hours at \(\qty{74}{\degreeCelsius}\) in a food dehydrator
-(Cosori CP267-FD, Vesync Co., Anaheim, California, USA) and weighed
-using a benchtop analytical balance (Ohaus PR64 Analytical Balance,
-Parsippany, New Jersey, USA). In \(10.5\%\) of leaves we measured LMA on
-the adjacent leaflet because the focal leaflet was damaged or wilted
-while making surface impressions and we could not reliably estimate
-area. LMA data are missing from \(2.97\%\) of individuals because the
-area or mass was not recorded at all or recorded incorrectly.
+FIJI implementation of ImageJ2 version 2.3.0 (58). We dried leaves for
+72 hours at \(\qty{74}{\degreeCelsius}\) in a food dehydrator (Cosori
+CP267-FD, Vesync Co., Anaheim, California, USA) and weighed using a
+benchtop analytical balance (Ohaus PR64 Analytical Balance, Parsippany,
+New Jersey, USA). In \(10.5\%\) of leaves we measured LMA on the
+adjacent leaflet because the focal leaflet was damaged or wilted while
+making surface impressions and we could not reliably estimate area. LMA
+data are missing from \(2.97\%\) of individuals because the area or mass
+was not recorded at all or recorded incorrectly.
 
@@
 All models were fit using a Bayesian HMC sampling in the probabilistic
-programming language \emph{Stan} (\emph{60}) using the \emph{R} package
-\textbf{brms} version 2.22.0 (\emph{61}). We used \emph{CmdStan} version
-2.36.0 and \textbf{cmdstanr} version 0.9.0 (\emph{62}) to interface with
-\emph{R} version 4.5.1 (\emph{63}). We sampled the posterior
-distribution from a single chain with a minimum of 1000 iterations after
-1000 warmup iterations. If necessary, we refit models with more
-iterations to decrease the Gelman-Rubin convergence statistic
-(\(\hat{R}\)) (\emph{64}) to less than 1.01 and the bulk effective
-sample size (ESS) to greater than 400 for all model parameters, and
-there were fewer than 10 divergent transitions.
+programming language \emph{Stan} (60) using the \emph{R} package
+\textbf{brms} version 2.22.0 (61). We used \emph{CmdStan} version 2.36.0
+and \textbf{cmdstanr} version 0.9.0 (62) to interface with \emph{R}
+version 4.5.1 (63). We sampled the posterior distribution from a single
+chain with a minimum of 1000 iterations after 1000 warmup iterations. If
+necessary, we refit models with more iterations to decrease the
+Gelman-Rubin convergence statistic (\(\hat{R}\)) (64) to less than 1.01
+and the bulk effective sample size (ESS) to greater than 400 for all
+model parameters, and there were fewer than 10 divergent transitions.
 
@@
 effects. We used the RAxML whole-transcriptome concatenated phylogeny
-based on 2,745 100-kb genomic windows from (\emph{54}). Two of our
-populations were not in this tree. We used accession LA1044 in place of
-LA3909, two populations of \emph{S. galapagense}; LA0750 was added as
-sister to LA0716, two closely related populations of \emph{S.
-pennellii}. The node separating LA0750 from LA0716 was placed half-way
-between the next deepest node. We used the leave-one-out
-cross-validation information criterion (LOOIC) to compare the fit of
-models (\autoref{tbl-aa_loo1}) using the \emph{R} package \textbf{loo}
-version 2.8.0 (\emph{65}) to calculate LOOIC values. We selected the
-model with the lowest LOOIC value (Table~\ref{tbl-aa_loo1}) to generate
-posterior predictions for hypothesis testing. Parameter estimates from
-the selected model are shown in Table~\ref{tbl-fit_aa1}.
+based on 2,745 100-kb genomic windows from (54). Two of our populations
+were not in this tree. We used accession LA1044 in place of LA3909, two
+populations of \emph{S. galapagense}; LA0750 was added as sister to
+LA0716, two closely related populations of \emph{S. pennellii}. The node
+separating LA0750 from LA0716 was placed half-way between the next
+deepest node. We used the leave-one-out cross-validation information
+criterion (LOOIC) to compare the fit of models (\autoref{tbl-aa_loo1})
+using the \emph{R} package \textbf{loo} version 2.8.0 (65) to calculate
+LOOIC values. We selected the model with the lowest LOOIC value
+(Table~\ref{tbl-aa_loo1}) to generate posterior predictions for
+hypothesis testing. Parameter estimates from the selected model are
+shown in Table~\ref{tbl-fit_aa1}.
 

### 2025-10-17 ms/ms.qmd
adding info to SI

@@
 Therefore, we automated data cleaning using custom *R* scripts.
-Cleaning is divided into six sequential steps (\autoref{tbl-cleaning1}).
+Cleaning is divided into seven sequential steps (\autoref{tbl-cleaning1}).
 
@@
 #| label: tbl-cleaning1
-#| tbl-cap: "Six sequential steps for cleaning \\agcurve{} curves.
+#| tbl-cap: "Seven sequential steps for cleaning \\agcurve{} curves.
 The rationale and procedure for each step are described in the text.
@@
   "1. remove unreliable and unusable data points",
-  "2. remove hysteretic portion of \\agcurve{} curves at low \\gsw{}",
-  "3. remove outliers within each \\agcurve{} curve",
-  "4. remove replicates with no overlap between amphi and pseudohypo \\agcurve{} curves",
-  "5. thin redundant data points within each \\agcurve{} curve",
-  "6. trim extreme \\aax{} values"
+  "2. remove points not at steady state",
+  "3. remove hysteretic portion of \\agcurve{} curves at low \\gsw{}",
+  "4. remove outliers within each \\agcurve{} curve",
+  "5. remove replicates with no overlap between amphi and pseudohypo \\agcurve{} curves",
+  "6. thin redundant data points within each \\agcurve{} curve",
+  "7. trim extreme \\aax{} values"
 )
@@
 ```
+
+#### Remove points not at steady state
+
+*Rationale*: Recent methods have been developed to quantify dynamic (i.e.
+unsteady state) assimilation and transpiration rates during gas exchange measurements [@saathoff_gas_2021].
+These methods potentially obviate the need to wait minutes to assure steady state conditions, enabling rapid estimation of dynamic gas exchange.
+The equations for dynamic assimilation and transpiration can also be used to assess whether steady state conditions have been reached.
+If dynamic and steady state estimates are similar, this indicates that the system is in steady state [@saathoff_gas_2021].
+To be conservative, we did not use dynamic assimilation or transpiration rates, but instead used the comparison between dynamic and steady state rates to identify and remove points that were not at steady state.
+
+*Procedure*: We removed all points where $\left| \log{A_\mathrm{sty} / A_\mathrm{dyn}} \right| < `r aa_args[['steadystate_cutoff']]`$ and $\left| \log{E_\mathrm{sty} / E_\mathrm{dyn}} \right| < `r aa_args[['steadystate_cutoff']]`$.
+The subscripts 'sty' and 'dyn' refer to steady state and dynamic estimates, respectively; the assimilation and transpiration rates are denoted by $A$ and $E$, respectively.
 

### 2025-10-17 r/04_fit-steadystate.R

@@ -71,3 +71,9 @@ fit_steadystate_diagnostics = list.files(dir, full.names = TRUE) |>
   }, .progress = TRUE)
 
 write_rds(fit_steadystate_diagnostics, "objects/fit_steadystate_diagnostics.rds")
+
+zip::zip(
+  "objects/curve-fits/steadystate.zip",
+  files = list.files("objects/curve-fits/steadystate", full.names = TRUE),
+  mode = "cherry-pick"
+)

### 2025-10-19 ms/ms.qmd
updated ms with new analysis

@@
 
-fit_aa1 = read_rds("../objects/fit_aa1.rds")
-fit_aa2 = read_rds("../objects/fit_aa2.rds")
+fit_aa1 = read_rds("../objects/fit_aa1_sty.rds")
+fit_aa2 = read_rds("../objects/fit_aa2_sty.rds")
 dim_fit_aa = dim(fit_aa1$fit)
@@
 
-aa_loo_table1 = read_rds("../objects/aa_loo_table1.rds")
-aa_loo_table2 = read_rds("../objects/aa_loo_table2.rds")
+aa_loo_table1 = read_rds("../objects/aa_sty_loo_table1.rds")
+aa_loo_table2 = read_rds("../objects/aa_sty_loo_table2.rds")
 
@@
 Stomata are microscopic pores on the surfaces of leaves and other photosynthetic organs formed by a pair of guard cells.
-They are essential for balancing carbon gained per unit water lost and were an essential innovation in vascular plants, permitting them to grow tall on land by enabling access to CO$_2$ for photosynthetic carbon assimilation while preventing hydraulic failure in variable environments [@raven_selection_2002; @mcadam_stomata_2021; @clark_origin_2022].
+They are essential for balancing carbon gained per unit water lost and were a key innovation in vascular plants, permitting them to grow tall on land by enabling access to CO$_2$ for photosynthetic carbon assimilation while preventing hydraulic failure in variable environments [@raven_selection_2002; @mcadam_stomata_2021; @clark_origin_2022].
 Optimal stomatal function depends on both dynamic changes in aperture on the scale of minutes to hours, as well as static anatomy determined by developmental plasticity and constitutive genetic differences [@hetherington_role_2003; @de_boer_optimal_2016; @harrison_influence_2020].
@@
 (11) estimate r, in Xan:hium to be closer to 1.0 m2 s mol-', and state that they believe the estimate of 3.2 m2 s mol-' to be high" -->
-Despite extensive theoretical and empirical progress understanding stomata function and anatomy from molecular to ecosystem levels, the adaptive significance of amphistomatous leaves remains an important unsolved problem in leaf structure-function relationships [@grubb_leaf_1977; @parkhurst_adaptive_1978; @mott_adaptive_1982; @gibson_structure-function_1996; @smith_leaf_1997; @oguchi_leaf_2018; @drake_two_2019; @grubb_leaf_2020].
+Despite extensive theoretical and empirical progress understanding stomatal function and anatomy from molecular to ecosystem levels, the adaptive significance of amphistomatous leaves remains an important unsolved problem in leaf structure-function relationships [@grubb_leaf_1977; @parkhurst_adaptive_1978; @mott_adaptive_1982; @gibson_structure-function_1996; @smith_leaf_1997; @oguchi_leaf_2018; @drake_two_2019; @grubb_leaf_2020].
 Amphistomatous leaves develop abaxial and adaxial stomata whose aperture can be independently regulated [@pospisilova_environmental_1980; @mott_stomatal_1984; @reich_effects_1985; @mott_asymmetric_1993; @wall_stomata_2022] to control gas exchange through each surface.
@@
 <!-- Paragraph 6 (summarize main results) -->
+
+```{r, echo = FALSE}
+x = if (df_coef_summary[['n_signif']] < df_coef_summary[['n_total']]) {
+  glue("{n1} of the {n2}", n1 = df_coef_summary[['n_signif']], n2 = df_coef_summary[['n_total']])
+} else {
+  "all"
+}
+```
+
 Consistent with biophysical theory of CO$_2$ diffusion within leaves, $\aay{} > 0$ for all populations ([Fig.
 @fig-aa]A).
-Bayesian phylogenetic mixed effects models that allowed \aax{} to vary between measurement light intensities, growth light intensities, and among populations outperformed simpler models based on information criteria (@tbl-aa_loo1).
+We tested whether \aax{} varied among treatments by comparing the fit of competing Bayesian phylogenetic mixed effects models using information criteria (see Materials and Methods).
+Models including measurement and growth light intensity consistently outperformed simpler models (@tbl-aa_loo1).
+There was equivocal support for interactions between factors and population because several models including these terms performed similarly to those without them (@tbl-aa_loo1).
+We focus on results that were robust to model structure and addressed our *a priori* hypotheses using parameter estimates from the best model.
+
 Measured under high light intensity, \aax{} was consistently greater for sun plants.
@@
 The overall average \aax{} of shade and sun grown plants measured under low light intensity was `r filter(aa_mean_range, light_treatment == "low", light_intensity == "150")$x` and `r filter(aa_mean_range, light_treatment == "high", light_intensity == "150")$x`, respectively.
-There was a modest tendency for populations from more open habitats (lower PAI) to exhibit greater \aax{} and the slope was significantly different than 0 in `r df_coef_summary[['n_signif']]` of the `r df_coef_summary[['n_total']]` treatment combinations ([Fig.
+There was a modest tendency for populations from more open habitats (lower PAI) to exhibit greater \aax{} and the slope was significantly different than 0 in `r x` treatment combinations ([Fig.
 @fig-aa]B).
@@
 The median relationship between native PAI ($x$-axis; log-scale) and \\aax{} is shown as a solid line with the shaded region between dashed lines showing the 95% confidence ribbon.
-The * in the top left corner of each facet indicates the slope of the linear regression is significantly different from zero; n.s.
-indicates not significant."
+The * in the top left corner of each facet indicates the slope of the linear regression is significantly different from zero."
+
+# Add to caption if needed
+# "; n.s. indicates not significant."
 
@@
 <!-- Paragraph 7 (summarize secondary results) -->
-The pattern of \aax{} across wild tomatoes strongly supports the plasticity hypothesis, contradicts the acclimation hypothesis, and provides modest support for the constitutive hypothesis.
-Plastic changes in leaf thickness and/or packing density, summarized by the bulk leaf mass per area (LMA), may mediate the effect of growth light intensity on \aax{}.
-LMA increased in sun grown plants in all populations by an average of `r signif(100 * (exp(post_lma["Estimate"]) - 1), 3)`% [95% CI: `r signif(100 * (exp(post_lma["Q2.5"]) - 1), 3)` to `r signif(100 * (exp(post_lma["Q97.5"]) - 1), 3)`%], qualitatively similar to plastic responses in many species [@poorter_metaanalysis_2019].
+
+<!-- Text from previous version where effect of growth light intensity was retained in LMA model:
+
 While LMA is weakly, albeit significantly (@tbl-fit_aa2), associated with individual-level \aax{} ([Fig.
@@
 @tbl-fit_aa2), suggestive of a mediating role.
-Many anatomical traits underlie LMA [@john_anatomical_2017] and future research will be needed to  identify which particular traits, such as leaf thickness or mesophyll porosity, are responsible for mediating \aax{}.
+
+-->
+
+The pattern of \aax{} across wild tomatoes strongly supports the plasticity hypothesis, contradicts the acclimation hypothesis, and provides modest support for the constitutive hypothesis.
+Plastic changes in leaf thickness and/or packing density, summarized by the bulk leaf mass per area (LMA), may mediate the effect of growth light intensity on \aax{}.
+LMA increased in sun grown plants in all populations by an average of `r signif(100 * (exp(post_lma["Estimate"]) - 1), 3)`% [95% CI: `r signif(100 * (exp(post_lma["Q2.5"]) - 1), 3)` to `r signif(100 * (exp(post_lma["Q97.5"]) - 1), 3)`%], qualitatively similar to plastic responses in many species [@poorter_metaanalysis_2019].
+When LMA is included in statistical models, the effect growth light intensity on \aax{} is no longer supported, indicating that plastic changes in LMA mediate the effect of growth light intensity (@tbl-fit_aa2).
+The effect of LMA on individual-level \aax{} ([Fig.
+@fig-lma_aa]) is weak, albeit significantly (@tbl-fit_aa2).
+This is likely because many anatomical traits underlie LMA [@john_anatomical_2017] and future research will be needed to identify which particular traits, such as leaf thickness or mesophyll porosity, are responsible for mediating \aax{}.
 The fact that the \aax{} of sun plants was greater under low measurement light intensity supports a long-standing hypothesis that resistance to CO$_2$ diffusion is greater in the upper than lower portions of the leaf interior [@jones_effects_1972].
@@
 As this is the first quantitative comparison of \aax{} across species, estimates from a broader range of species and more stressful environments will be required to fully understand the ecological causes and consequences of stomatal distributions on leaf surfaces.
-
-<!-- note: estimates of AA are less than that described by Parkhurst and Mott (1990) and also less than that of Xiong and Flexas (2021). What does this tell us? -->
 
@@
 
-<!-- Not using this currently  -->
-<!-- #### Light-saturated photosynthetic rate -->
-
-<!-- In `r percent_aa_and_aq`\% of plants, we measured the maximum photosynthetic rate ($A_\mathrm{max}$) on the same leaflets as \agcurve{} curves.
-However, when a leaflet was damaged during \agcurve{} curves, we used the next closest leaflet for $A_\mathrm{max}$.
-Leaves acclimated to high light-intensity (\ppfdequals{2000}), ambient CO$_2$ (\caequals{415}), \rhequals{50}, and \tleafequals{25}.
-After $A$ and \gsw{} stabilized, we measured $A$ at $\qty{2000}{\micro\mol\raiseto{-2}\meter\raiseto{-1}\second}$.
--->
-
 #### Stomatal anatomy
-
-<!-- are there any cases where we did not get stomatal anatomy data? -->
 
@@
 The \gmaxratio{} is the \gmax{} of the adaxial surface divided by the sum total \gmax{} of both surfaces.
-
-<!-- 
-#### Internal leaf anatomy
-
-Add information on internal leaf anatomy, if using 
--->
 

### 2025-10-19 ms/ms.tex
updated ms with new analysis

@@
+% Options for packages loaded elsewhere
 % Options for packages loaded elsewhere
@@
   numbers=noendperiod]{scrartcl}
-
+\usepackage{xcolor}
 \usepackage{amsmath,amssymb}
+\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
 \usepackage{iftex}
@@
 \else % if luatex or xetex
-  \usepackage{unicode-math}
+  \usepackage{unicode-math} % this also loads fontspec
   \defaultfontfeatures{Scale=MatchLowercase}
@@
 \makeatother
-\usepackage{xcolor}
-\setlength{\emergencystretch}{3em} % prevent overfull lines
-\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
 % Make \paragraph and \subparagraph free-standing
+\makeatletter
 \ifx\paragraph\undefined\else
   \let\oldparagraph\paragraph
-  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
+  \renewcommand{\paragraph}{
+    \@ifstar
+      \xxxParagraphStar
+      \xxxParagraphNoStar
+  }
+  \newcommand{\xxxParagraphStar}[1]{\oldparagraph*{#1}\mbox{}}
+  \newcommand{\xxxParagraphNoStar}[1]{\oldparagraph{#1}\mbox{}}
 \fi
@@
   \let\oldsubparagraph\subparagraph
-  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
+  \renewcommand{\subparagraph}{
+    \@ifstar
+      \xxxSubParagraphStar
+      \xxxSubParagraphNoStar
+  }
+  \newcommand{\xxxSubParagraphStar}[1]{\oldsubparagraph*{#1}\mbox{}}
+  \newcommand{\xxxSubParagraphNoStar}[1]{\oldsubparagraph{#1}\mbox{}}
 \fi
-
-\providecommand{\tightlist}{%
-  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
+\makeatother
+
+\usepackage{longtable,booktabs,array}
 \usepackage{calc} % for calculating minipage widths
@@
 \makeatletter
-\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
-\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
-\makeatother
-% Scale images if necessary, so that they will not overflow the page
-% margins by default, and it is still possible to overwrite the defaults
-% using explicit options in \includegraphics[width, height, ...]{}
-\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
+\newsavebox\pandoc@box
+\newcommand*\pandocbounded[1]{% scales image to fit in text height/width
+  \sbox\pandoc@box{#1}%
+  \Gscale@div\@tempa{\textheight}{\dimexpr\ht\pandoc@box+\dp\pandoc@box\relax}%
+  \Gscale@div\@tempb{\linewidth}{\wd\pandoc@box}%
+  \ifdim\@tempb\p@<\@tempa\p@\let\@tempa\@tempb\fi% select the smaller of both
+  \ifdim\@tempa\p@<\p@\scalebox{\@tempa}{\usebox\pandoc@box}%
+  \else\usebox{\pandoc@box}%
+  \fi%
+}
 % Set default figure placement to htbp
-\makeatletter
 \def\fps@figure{htbp}
 \makeatother
+
 % definitions for citeproc citations
@@
 \newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
+
+\setlength{\emergencystretch}{3em} % prevent overfull lines
+
+\providecommand{\tightlist}{%
+  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
 
@@
 \makeatother
-\ifLuaTeX
-  \usepackage{selnolig}  % disable illegal ligatures
-\fi
 \usepackage{bookmark}
-
 \IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
-\urlstyle{same} % disable monospaced font for URLs
+\urlstyle{same}
 \hypersetup{
@@
 \date{}
-
 \begin{document}
@@
 photosynthetic organs formed by a pair of guard cells. They are
-essential for balancing carbon gained per unit water lost and were an
-essential innovation in vascular plants, permitting them to grow tall on
-land by enabling access to CO\(_2\) for photosynthetic carbon
-assimilation while preventing hydraulic failure in variable environments
-(1--3). Optimal stomatal function depends on both dynamic changes in
-aperture on the scale of minutes to hours, as well as static anatomy
-determined by developmental plasticity and constitutive genetic
-differences (4--6). Understanding how stomata respond to environmental
-change over daily, developmental, and evolutionary time is important for
-studying adaptation (1, 7, 8), inferring paleoclimate from fossil
-cuticles (9--11), predicting responses to climate change (12--15), and
-improving crops (16). Stomatal function contributes to global carbon and
-water cycles (17) and therefore to predicting future climate (18).
+essential for balancing carbon gained per unit water lost and were a key
+innovation in vascular plants, permitting them to grow tall on land by
+enabling access to CO\(_2\) for photosynthetic carbon assimilation while
+preventing hydraulic failure in variable environments (1--3). Optimal
+stomatal function depends on both dynamic changes in aperture on the
+scale of minutes to hours, as well as static anatomy determined by
+developmental plasticity and constitutive genetic differences (4--6).
+Understanding how stomata respond to environmental change over daily,
+developmental, and evolutionary time is important for studying
+adaptation (1, 7, 8), inferring paleoclimate from fossil cuticles
+(9--11), predicting responses to climate change (12--15), and improving
+crops (16). Stomatal function contributes to global carbon and water
+cycles (17) and therefore to predicting future climate (18).
 
 Despite extensive theoretical and empirical progress understanding
-stomata function and anatomy from molecular to ecosystem levels, the
+stomatal function and anatomy from molecular to ecosystem levels, the
 adaptive significance of amphistomatous leaves remains an important
@@
 
-\includegraphics{../figures/hypotheses.pdf}
+\pandocbounded{\includegraphics[keepaspectratio]{../figures/hypotheses.pdf}}
 
@@
 and shade light treatments, and measured under contrasting light
-intensities (low and high). We measured \aax{} on 572 individual plants
-from 29 populations (average of 9.86 replicates per light treatment)
+intensities (low and high). We measured \aax{} on 567 individual plants
+from 29 populations (average of 9.78 replicates per light treatment)
 using a recently developed method (46). With this method, we directly
@@
 Consistent with biophysical theory of CO\(_2\) diffusion within leaves,
-\(\aay{} > 0\) for all populations (Fig.~\ref{fig-aa}A). Bayesian
-phylogenetic mixed effects models that allowed \aax{} to vary between
-measurement light intensities, growth light intensities, and among
-populations outperformed simpler models based on information criteria
-(Table~\ref{tbl-aa_loo1}). Measured under high light intensity, \aax{}
-was consistently greater for sun plants. The average \aax{} among
-populations in the shade treatment was 0.041 (range: 0.007--0.113; 19 of
-29 populations significant); however, the same populations grown at high
-light intensity showed a mean \aax{} of 0.052 (range: 0.020--0.120; 20
-of 29 populations significant). Contrary to the predictions of the
-acclimation hypothesis, \aax{} was greater in all populations under low
-measurement light intensity for both sun and shade grown plants. The
-overall average \aax{} of shade and sun grown plants measured under low
-light intensity was 0.064 (range: 0.022--0.137; 28 of 29 populations
-significant) and 0.100 (range: 0.049--0.206; 27 of 29 populations
-significant), respectively. There was a modest tendency for populations
-from more open habitats (lower PAI) to exhibit greater \aax{} and the
-slope was significantly different than 0 in 3 of the 4 treatment
-combinations (Fig.~\ref{fig-aa}B).
+\(\aay{} > 0\) for all populations (Fig.~\ref{fig-aa}A). We tested
+whether \aax{} varied among treatments by comparing the fit of competing
+Bayesian phylogenetic mixed effects models using information criteria
+(see Materials and Methods). Models including measurement and growth
+light intensity consistently outperformed simpler models
+(Table~\ref{tbl-aa_loo1}). There was equivocal support for interactions
+between factors and population because several models including these
+terms performed similarly to those without them
+(Table~\ref{tbl-aa_loo1}). We focus on results that were robust to model
+structure and addressed our \emph{a priori} hypotheses using parameter
+estimates from the best model.
+
+Measured under high light intensity, \aax{} was consistently greater for
+sun plants. The average \aax{} among populations in the shade treatment
+was 0.039 (range: 0.012--0.098; 23 of 29 populations significant);
+however, the same populations grown at high light intensity showed a
+mean \aax{} of 0.055 (range: 0.028--0.114; 28 of 29 populations
+significant). Contrary to the predictions of the acclimation hypothesis,
+\aax{} was greater in all populations under low measurement light
+intensity for both sun and shade grown plants. The overall average
+\aax{} of shade and sun grown plants measured under low light intensity
+was 0.074 (range: 0.032--0.141; 28 of 29 populations significant) and
+0.090 (range: 0.049--0.158; 29 of 29 populations significant),
+respectively. There was a modest tendency for populations from more open
+habitats (lower PAI) to exhibit greater \aax{} and the slope was
+significantly different than 0 in all treatment combinations
+(Fig.~\ref{fig-aa}B).
 
@@
 confidence ribbon. The * in the top left corner of each facet indicates
-the slope of the linear regression is significantly different from zero;
-n.s. indicates not significant.}
+the slope of the linear regression is significantly different from
+zero.}
 
@@
 \aax{}. LMA increased in sun grown plants in all populations by an
-average of 123\% {[}95\% CI: 42.9 to 256\%{]}, qualitatively similar to
-plastic responses in many species (47). While LMA is weakly, albeit
-significantly (Table~\ref{tbl-fit_aa2}), associated with
-individual-level \aax{} (Fig.~\ref{fig-lma_aa}), the effect of growth
-light intensity on \aax{} is still predictive based on model comparison
-using information criteria (Table~\ref{tbl-aa_loo1}). However, the
-direct effect of the sun treatment on \aax{} was weaker when LMA was
-included in the model (Table~\ref{tbl-fit_aa1} vs.
-Table~\ref{tbl-fit_aa2}), suggestive of a mediating role. Many
-anatomical traits underlie LMA (52) and future research will be needed
-to identify which particular traits, such as leaf thickness or mesophyll
-porosity, are responsible for mediating \aax{}. The fact that the \aax{}
-of sun plants was greater under low measurement light intensity supports
-a long-standing hypothesis that resistance to CO\(_2\) diffusion is
-greater in the upper than lower portions of the leaf interior (53). At
-low light intensity, photosynthesis is weighted toward the upper
-palisade where most light is intercepted. If resistance to CO\(_2\)
-diffusion is high in the upper palisade, amphistomy may, unexpectedly,
-be particularly beneficial for sun leaves experiencing intermittent
-shade or cloud cover. Our study is limited in testing this because we
-could not directly measure the stomatal conductance ratio and
+average of 123\% {[}95\% CI: 50.8 to 233\%{]}, qualitatively similar to
+plastic responses in many species (47). When LMA is included in
+statistical models, the effect growth light intensity on \aax{} is no
+longer supported, indicating that plastic changes in LMA mediate the
+effect of growth light intensity (Table~\ref{tbl-fit_aa2}). The effect
+of LMA on individual-level \aax{} (Fig.~\ref{fig-lma_aa}) is weak,
+albeit significantly (Table~\ref{tbl-fit_aa2}). This is likely because
+many anatomical traits underlie LMA (52) and future research will be
+needed to identify which particular traits, such as leaf thickness or
+mesophyll porosity, are responsible for mediating \aax{}. The fact that
+the \aax{} of sun plants was greater under low measurement light
+intensity supports a long-standing hypothesis that resistance to
+CO\(_2\) diffusion is greater in the upper than lower portions of the
+leaf interior (53). At low light intensity, photosynthesis is weighted
+toward the upper palisade where most light is intercepted. If resistance
+to CO\(_2\) diffusion is high in the upper palisade, amphistomy may,
+unexpectedly, be particularly beneficial for sun leaves experiencing
+intermittent shade or cloud cover. Our study is limited in testing this
+because we could not directly measure the stomatal conductance ratio and
 intercellular resistance on each surface. Future experiments measuring
@@
 
+\bibitem[\citeproctext]{ref-saathoff_gas_2021}
+\CSLLeftMargin{60. }%
+\CSLRightInline{A. J. Saathoff, J. Welles,
+\href{https://doi.org/10.1111/pce.14178}{Gas exchange measurements in
+the unsteady state}. \emph{Plant, Cell \& Environment} \textbf{44},
+3509--3523 (2021).}
+
 \bibitem[\citeproctext]{ref-stan_development_team_stan_2025}
-\CSLLeftMargin{60. }%
+\CSLLeftMargin{61. }%
 \CSLRightInline{Stan Development Team,
@@
 \bibitem[\citeproctext]{ref-burkner_brms_2017}
-\CSLLeftMargin{61. }%
+\CSLLeftMargin{62. }%
 \CSLRightInline{P.-C. Bürkner,
@@
 \bibitem[\citeproctext]{ref-gabry_cmdstanr_2025}
-\CSLLeftMargin{62. }%
+\CSLLeftMargin{63. }%
 \CSLRightInline{J. Gabry, R. Češnovar, A. Johnson, S. Bronder,
@@
 \bibitem[\citeproctext]{ref-r_core_team_r:_2025}
-\CSLLeftMargin{63. }%
+\CSLLeftMargin{64. }%
 \CSLRightInline{R Core Team, \emph{\href{http://www.R-project.org/}{R:
@@
 \bibitem[\citeproctext]{ref-gelman_inference_1992}
-\CSLLeftMargin{64. }%
+\CSLLeftMargin{65. }%
 \CSLRightInline{A. Gelman, D. B. Rubin, Inference from iterative
@@
 \bibitem[\citeproctext]{ref-vehtari_practical_2017}
-\CSLLeftMargin{65. }%
+\CSLLeftMargin{66. }%
 \CSLRightInline{A. Vehtari, A. Gelman, J. Gabry,
@@
 a leaflet was damaged during gas exchange measurements, we collected
-anatomical data from the nearest leaflet on the same leaf (1.58\% of
+anatomical data from the nearest leaflet on the same leaf (1.6\% of
 leaves).
@@
 benchtop analytical balance (Ohaus PR64 Analytical Balance, Parsippany,
-New Jersey, USA). In \(10.5\%\) of leaves we measured LMA on the
+New Jersey, USA). In \(10.3\%\) of leaves we measured LMA on the
 adjacent leaflet because the focal leaflet was damaged or wilted while
 making surface impressions and we could not reliably estimate area. LMA
-data are missing from \(2.97\%\) of individuals because the area or mass
+data are missing from \(2.82\%\) of individuals because the area or mass
 was not recorded at all or recorded incorrectly.
@@
 data cleaning using custom \emph{R} scripts. Cleaning is divided into
-six sequential steps (\autoref{tbl-cleaning1}).
+seven sequential steps (\autoref{tbl-cleaning1}).
 
@@
 
-\caption{\label{tbl-cleaning1}Six sequential steps for cleaning
+\caption{\label{tbl-cleaning1}Seven sequential steps for cleaning
 \agcurve{} curves. The rationale and procedure for each step are
@@
 \cellcolor{gray!10}{1. remove unreliable and unusable data points} & \cellcolor{gray!10}{2,361} & \cellcolor{gray!10}{63.0}\\
-2. remove hysteretic portion of \agcurve{} curves at low \gsw{} & 2,360 & 59.2\\
-\cellcolor{gray!10}{3. remove outliers within each \agcurve{} curve} & \cellcolor{gray!10}{2,360} & \cellcolor{gray!10}{58.7}\\
-4. remove replicates with no overlap between amphi and pseudohypo \agcurve{} curves & 2,268 & 58.4\\
-\cellcolor{gray!10}{5. thin redundant data points within each \agcurve{} curve} & \cellcolor{gray!10}{2,268} & \cellcolor{gray!10}{28.1}\\
+2. remove points not at steady state & 2,359 & 59.3\\
+\cellcolor{gray!10}{3. remove hysteretic portion of \agcurve{} curves at low \gsw{}} & \cellcolor{gray!10}{2,359} & \cellcolor{gray!10}{55.6}\\
+4. remove outliers within each \agcurve{} curve & 2,349 & 55.4\\
+\cellcolor{gray!10}{5. remove replicates with no overlap between amphi and pseudohypo \agcurve{} curves} & \cellcolor{gray!10}{2,224} & \cellcolor{gray!10}{55.2}\\
 \addlinespace
-6. trim extreme \aax{} values & 2,214 & 28.1\\
+6. thin redundant data points within each \agcurve{} curve & 2,224 & 27.0\\
+\cellcolor{gray!10}{7. trim extreme \aax{} values} & \cellcolor{gray!10}{2,162} & \cellcolor{gray!10}{27.0}\\
 \bottomrule
@@
 \end{table}%
+
+\paragraph{Remove points not at steady
+state}\label{remove-points-not-at-steady-state}
+
+\emph{Rationale}: Recent methods have been developed to quantify dynamic
+(i.e.~unsteady state) assimilation and transpiration rates during gas
+exchange measurements (60). These methods potentially obviate the need
+to wait minutes to assure steady state conditions, enabling rapid
+estimation of dynamic gas exchange. The equations for dynamic
+assimilation and transpiration can also be used to assess whether steady
+state conditions have been reached. If dynamic and steady state
+estimates are similar, this indicates that the system is in steady state
+(60). To be conservative, we did not use dynamic assimilation or
+transpiration rates, but instead used the comparison between dynamic and
+steady state rates to identify and remove points that were not at steady
+state.
+
+\emph{Procedure}: We removed all points where
+\(\left| \log{A_\mathrm{sty} / A_\mathrm{dyn}} \right| < 0.05\) and
+\(\left| \log{E_\mathrm{sty} / E_\mathrm{dyn}} \right| < 0.05\). The
+subscripts `sty' and `dyn' refer to steady state and dynamic estimates,
+respectively; the assimilation and transpiration rates are denoted by
+\(A\) and \(E\), respectively.
 
@@
 All models were fit using a Bayesian HMC sampling in the probabilistic
-programming language \emph{Stan} (60) using the \emph{R} package
-\textbf{brms} version 2.22.0 (61). We used \emph{CmdStan} version 2.36.0
-and \textbf{cmdstanr} version 0.9.0 (62) to interface with \emph{R}
-version 4.5.1 (63). We sampled the posterior distribution from a single
+programming language \emph{Stan} (61) using the \emph{R} package
+\textbf{brms} version 2.23.0 (62). We used \emph{CmdStan} version 2.37.0
+and \textbf{cmdstanr} version 0.9.0 (63) to interface with \emph{R}
+version 4.5.1 (64). We sampled the posterior distribution from a single
 chain with a minimum of 1000 iterations after 1000 warmup iterations. If
 necessary, we refit models with more iterations to decrease the
-Gelman-Rubin convergence statistic (\(\hat{R}\)) (64) to less than 1.01
+Gelman-Rubin convergence statistic (\(\hat{R}\)) (65) to less than 1.01
 and the bulk effective sample size (ESS) to greater than 400 for all
@@
 criterion (LOOIC) to compare the fit of models (\autoref{tbl-aa_loo1})
-using the \emph{R} package \textbf{loo} version 2.8.0 (65) to calculate
+using the \emph{R} package \textbf{loo} version 2.8.0 (66) to calculate
 LOOIC values. We selected the model with the lowest LOOIC value
@@
 \multicolumn{2}{l}{\textit{Fixed effects}}\\
-\hspace{1em}\cellcolor{gray!10}{$\mathrm{AA}$ intercept (shade, low light)} & \cellcolor{gray!10}{0.08 [0.01, 0.16]}\\
-\hspace{1em}effect of sun on $\mathrm{AA}$ & 0.04 [-0.02, 0.12]\\
-\hspace{1em}\cellcolor{gray!10}{effect of high light on $\mathrm{AA}$} & \cellcolor{gray!10}{-0.03 [-0.06, 0]}\\
-\hspace{1em}effect of sun $\times$ high light interaction on $\mathrm{AA}$ & -0.03 [-0.1, 0.03]\\
+\hspace{1em}\cellcolor{gray!10}{$\mathrm{AA}$ intercept (shade, low light)} & \cellcolor{gray!10}{0.09 [0.02, 0.16]}\\
+\hspace{1em}effect of sun on $\mathrm{AA}$ & 0.02 [0.01, 0.03]\\
+\hspace{1em}\cellcolor{gray!10}{effect of high light on $\mathrm{AA}$} & \cellcolor{gray!10}{-0.04 [-0.07, -0.01]}\\
 \addlinespace[0.3em]
 \multicolumn{2}{l}{\textit{Distributional parameters}}\\
-\hspace{1em}\cellcolor{gray!10}{$\log \sigma$ intercept (shade, low light)} & \cellcolor{gray!10}{-2.51 [-2.61, -2.41]}\\
-\hspace{1em}effect of sun on $\log \sigma$ & 0.23 [0.12, 0.33]\\
-\hspace{1em}\cellcolor{gray!10}{effect of high light on $\log \sigma$} & \cellcolor{gray!10}{-0.19 [-0.3, -0.09]}\\
+\hspace{1em}$\log \sigma$ intercept (shade, low light) & -2.49 [-2.6, -2.39]\\
+\hspace{1em}\cellcolor{gray!10}{effect of sun on $\log \sigma$} & \cellcolor{gray!10}{0.23 [0.11, 0.33]}\\
+\hspace{1em}effect of high light on $\log \sigma$ & -0.19 [-0.3, -0.07]\\
 \addlinespace[0.3em]
 \multicolumn{2}{l}{\textit{Random effect SDs}}\\
-\hspace{1em}$\mathrm{AA}$ among populations & 0.06 [0.04, 0.1]\\
-\hspace{1em}\cellcolor{gray!10}{effect of sun on $\mathrm{AA}$ among populations} & \cellcolor{gray!10}{0.05 [0, 0.1]}\\
+\hspace{1em}\cellcolor{gray!10}{$\mathrm{AA}$ among populations} & \cellcolor{gray!10}{0.06 [0.03, 0.09]}\\
 \hspace{1em}effect of high light on $\mathrm{AA}$ among populations & 0.02 [0, 0.05]\\
-\hspace{1em}\cellcolor{gray!10}{effect of sun $\times$ high light interaction on $\mathrm{AA}$ among populations} & \cellcolor{gray!10}{0.04 [0, 0.1]}\\
 \bottomrule
@@
 \multicolumn{2}{l}{\textit{~~Fixed effects}}\\
-\hspace{1em}\hspace{1em}\cellcolor{gray!10}{$\mathrm{AA}$ intercept (shade, low light)} & \cellcolor{gray!10}{-0.01 [-0.12, 0.09]}\\
-\hspace{1em}\hspace{1em}effect of sun on $\mathrm{AA}$ & 0.01 [-0.05, 0.06]\\
-\hspace{1em}\hspace{1em}\cellcolor{gray!10}{effect of high light on $\mathrm{AA}$} & \cellcolor{gray!10}{-0.03 [-0.07, 0]}\\
-\hspace{1em}\hspace{1em}effect of $\log \mathrm{LMA}$ on $\mathrm{AA}$ & 0.03 [0, 0.05]\\
+\hspace{1em}\hspace{1em}\cellcolor{gray!10}{$\mathrm{AA}$ intercept (shade, low light)} & \cellcolor{gray!10}{0.03 [-0.07, 0.11]}\\
+\hspace{1em}\hspace{1em}effect of high light on $\mathrm{AA}$ & -0.04 [-0.08, 0]\\
+\hspace{1em}\hspace{1em}\cellcolor{gray!10}{effect of $\log \mathrm{LMA}$ on $\mathrm{AA}$} & \cellcolor{gray!10}{0.02 [0.01, 0.03]}\\
 \addlinespace[0.3em]
 \multicolumn{2}{l}{\textit{~~Distributional parameters}}\\
-\hspace{1em}\hspace{1em}\cellcolor{gray!10}{$\log \sigma$ intercept (shade, low light)} & \cellcolor{gray!10}{-2.5 [-2.61, -2.4]}\\
-\hspace{1em}\hspace{1em}effect of sun on $\log \sigma$ & 0.23 [0.13, 0.34]\\
-\hspace{1em}\hspace{1em}\cellcolor{gray!10}{effect of high light on $\log \sigma$} & \cellcolor{gray!10}{-0.19 [-0.29, -0.08]}\\
+\hspace{1em}\hspace{1em}$\log \sigma$ intercept (shade, low light) & -2.51 [-2.62, -2.41]\\
+\hspace{1em}\hspace{1em}\cellcolor{gray!10}{effect of sun on $\log \sigma$} & \cellcolor{gray!10}{0.23 [0.12, 0.34]}\\
+\hspace{1em}\hspace{1em}effect of high light on $\log \sigma$ & -0.17 [-0.29, -0.06]\\
 \addlinespace[0.3em]
 \multicolumn{2}{l}{\textit{~~Random effect SDs}}\\
-\hspace{1em}\hspace{1em}$\mathrm{AA}$ among populations & 0.06 [0.04, 0.1]\\
-\hspace{1em}\hspace{1em}\cellcolor{gray!10}{effect of sun on $\mathrm{AA}$ among populations} & \cellcolor{gray!10}{0.04 [0, 0.1]}\\
-\hspace{1em}\hspace{1em}effect of high light on $\mathrm{AA}$ among populations & 0.02 [0, 0.05]\\
-\hspace{1em}\hspace{1em}\cellcolor{gray!10}{effect of sun $\times$ high light interaction on $\mathrm{AA}$ among populations} & \cellcolor{gray!10}{0.04 [0, 0.09]}\\
+\hspace{1em}\hspace{1em}\cellcolor{gray!10}{$\mathrm{AA}$ among populations} & \cellcolor{gray!10}{0.06 [0.04, 0.1]}\\
+\hspace{1em}\hspace{1em}effect of high light on $\mathrm{AA}$ among populations & 0.02 [0, 0.06]\\
 \addlinespace[0.3em]
@@
 \multicolumn{2}{l}{\textit{~~Fixed effects}}\\
-\hspace{1em}\hspace{1em}$\log \mathrm{LMA}$ intercept (shade) & 2.99 [2.42, 3.52]\\
-\hspace{1em}\hspace{1em}\cellcolor{gray!10}{effect of sun on $\log \mathrm{LMA}$} & \cellcolor{gray!10}{0.8 [0.36, 1.27]}\\
+\hspace{1em}\hspace{1em}\cellcolor{gray!10}{$\log \mathrm{LMA}$ intercept (shade)} & \cellcolor{gray!10}{3 [2.5, 3.52]}\\
+\hspace{1em}\hspace{1em}effect of sun on $\log \mathrm{LMA}$ & 0.8 [0.41, 1.2]\\
 \addlinespace[0.3em]
 \multicolumn{2}{l}{\textit{~~Random effect SDs}}\\
-\hspace{1em}\hspace{1em}$\log \mathrm{LMA}$ among populations & 0.48 [0.35, 0.66]\\
-\hspace{1em}\hspace{1em}\cellcolor{gray!10}{effect of sun on $\log \mathrm{LMA}$ among populations} & \cellcolor{gray!10}{0.37 [0.23, 0.6]}\\
+\hspace{1em}\hspace{1em}\cellcolor{gray!10}{$\log \mathrm{LMA}$ among populations} & \cellcolor{gray!10}{0.47 [0.34, 0.67]}\\
+\hspace{1em}\hspace{1em}effect of sun on $\log \mathrm{LMA}$ among populations & 0.36 [0.22, 0.57]\\
 \bottomrule
@@
 \endlastfoot
-\cellcolor{gray!10}{-1882.56} & \cellcolor{gray!10}{0.00} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$}\\
--1881.70 & 0.86 &  &  &  & $\checkmark$ & $\checkmark$ & $\checkmark$ & $\checkmark$ & $\checkmark$\\
-\cellcolor{gray!10}{-1881.00} & \cellcolor{gray!10}{1.56} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$}\\
--1880.91 & 1.65 & $\checkmark$ & $\checkmark$ &  & $\checkmark$ & $\checkmark$ & $\checkmark$ & $\checkmark$ & $\checkmark$\\
-\cellcolor{gray!10}{-1880.63} & \cellcolor{gray!10}{1.93} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$}\\*
+\cellcolor{gray!10}{-1789.54} & \cellcolor{gray!10}{0.00} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$}\\
+-1789.47 & 0.07 & $\checkmark$ &  &  & $\checkmark$ & $\checkmark$ &  & $\checkmark$ & $\checkmark$\\
+\cellcolor{gray!10}{-1789.25} & \cellcolor{gray!10}{0.29} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$}\\
+-1789.18 & 0.36 & $\checkmark$ & $\checkmark$ &  & $\checkmark$ & $\checkmark$ & $\checkmark$ & $\checkmark$ & $\checkmark$\\
+\cellcolor{gray!10}{-1788.95} & \cellcolor{gray!10}{0.60} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$}\\
+\addlinespace
+-1788.67 & 0.88 & $\checkmark$ &  &  &  & $\checkmark$ &  & $\checkmark$ & $\checkmark$\\
+\cellcolor{gray!10}{-1788.33} & \cellcolor{gray!10}{1.21} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$}\\
+-1788.33 & 1.22 & $\checkmark$ & $\checkmark$ &  &  & $\checkmark$ &  & $\checkmark$ & $\checkmark$\\
+\cellcolor{gray!10}{-1788.21} & \cellcolor{gray!10}{1.34} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$}\\
+-1788.19 & 1.35 & $\checkmark$ & $\checkmark$ &  &  &  &  & $\checkmark$ & $\checkmark$\\
+\addlinespace
+\cellcolor{gray!10}{-1788.17} & \cellcolor{gray!10}{1.38} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$}\\
+-1788.17 & 1.38 & $\checkmark$ & $\checkmark$ & $\checkmark$ &  & $\checkmark$ &  & $\checkmark$ & $\checkmark$\\
+\cellcolor{gray!10}{-1787.91} & \cellcolor{gray!10}{1.64} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$}\\
+-1787.75 & 1.80 &  &  &  & $\checkmark$ & $\checkmark$ &  & $\checkmark$ & $\checkmark$\\*
 
@@
 \endlastfoot
-\cellcolor{gray!10}{-1883.84} & \cellcolor{gray!10}{0.00} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$}\\
--1883.80 & 0.04 & $\checkmark$ &  & $\checkmark$ &  &  &  &  & $\checkmark$ & $\checkmark$ & $\checkmark$\\
-\cellcolor{gray!10}{-1883.67} & \cellcolor{gray!10}{0.17} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{}\\
--1883.16 & 0.67 & $\checkmark$ &  & $\checkmark$ &  &  &  &  & $\checkmark$ &  & \\
-\cellcolor{gray!10}{-1883.05} & \cellcolor{gray!10}{0.78} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{}\\
+\cellcolor{gray!10}{-1799.06} & \cellcolor{gray!10}{0.00} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{}\\
+-1798.39 & 0.67 & $\checkmark$ &  & $\checkmark$ &  &  &  &  &  &  & \\
+\cellcolor{gray!10}{-1798.21} & \cellcolor{gray!10}{0.85} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{}\\
+-1797.91 & 1.15 &  &  & $\checkmark$ &  &  &  &  & $\checkmark$ &  & \\
+\cellcolor{gray!10}{-1797.77} & \cellcolor{gray!10}{1.29} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$}\\
 \addlinespace
--1883.05 & 0.78 & $\checkmark$ &  & $\checkmark$ &  &  &  & $\checkmark$ & $\checkmark$ & $\checkmark$ & \\
-\cellcolor{gray!10}{-1882.93} & \cellcolor{gray!10}{0.90} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{}\\
--1882.87 & 0.97 & $\checkmark$ & $\checkmark$ & $\checkmark$ &  & $\checkmark$ &  &  & $\checkmark$ & $\checkmark$ & $\checkmark$\\
-\cellcolor{gray!10}{-1882.80} & \cellcolor{gray!10}{1.04} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$}\\
--1882.55 & 1.29 & $\checkmark$ & $\checkmark$ & $\checkmark$ &  & $\checkmark$ &  &  & $\checkmark$ &  & \\
-\addlinespace
-\cellcolor{gray!10}{-1882.29} & \cellcolor{gray!10}{1.55} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$}\\
--1882.27 & 1.56 & $\checkmark$ & $\checkmark$ & $\checkmark$ &  &  & $\checkmark$ &  & $\checkmark$ & $\checkmark$ & $\checkmark$\\
-\cellcolor{gray!10}{-1882.27} & \cellcolor{gray!10}{1.56} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$}\\
--1882.08 & 1.76 &  &  & $\checkmark$ &  &  &  &  & $\checkmark$ & $\checkmark$ & $\checkmark$\\
-\cellcolor{gray!10}{-1881.86} & \cellcolor{gray!10}{1.98} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{}\\*
+-1797.44 & 1.62 & $\checkmark$ & $\checkmark$ & $\checkmark$ &  & $\checkmark$ &  &  & $\checkmark$ & $\checkmark$ & $\checkmark$\\
+\cellcolor{gray!10}{-1797.17} & \cellcolor{gray!10}{1.89} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{$\checkmark$} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{} & \cellcolor{gray!10}{}\\
+-1797.11 & 1.95 & $\checkmark$ &  & $\checkmark$ &  &  &  &  & $\checkmark$ & $\checkmark$ & \\*
 

### 2025-10-19 r/14_plot-aa.R

@@ -88,7 +88,7 @@ df_aa_text = df_aa_pred1 |>
   summarize(
     x = 0,
     label = sprintf("%.3f", round(mean(aa), 3)),
-    aa = 0.30,
+    aa = 0.20,
     .by = c("Growth", "Measurement")
   )
 

### 2025-10-19 r/15_plot-pai.R

@@ -2,7 +2,7 @@
 source("r/header.R")
 
 fit_aa1 = read_rds("objects/fit_aa1_sty.rds")
-fit_aa2 = read_rds("objects/fit_aa2.rds")
+fit_aa2 = read_rds("objects/fit_aa2_sty.rds")
 d1 = fit_aa1$data
 accession_gedi = read_rds("data/accession-gedi.rds") |>
   rename(acc = accession) |>

### 2025-10-19 r/16_plot-lma.R

@@ -1,7 +1,7 @@
 # Plot effect LMA on AA
 source("r/header.R")
 
-fit_aa = read_rds("objects/fit_aa2.rds")
+fit_aa = read_rds("objects/fit_aa2_sty.rds")
 
 # Estimated llma for each acc, light_intensity, and light_treatment
 df_new1 = crossing(
